### 2023-02-02 Thesis.Rmd
initial commit

---
title: "Manual/Thesis"
author: "Lukas Gamp"
date: "2023-02-02"
output: 
  pdf_document:
    number_sections: TRUE
geometry: "left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm"
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{sectsty}
  - \usepackage{paralist}
  - \usepackage{setspace}\spacing{1.5}
  - \usepackage{fancyhdr}
  - \usepackage{lastpage}
  - \usepackage{dcolumn}
  - \usepackage{natbib}\bibliographystyle{agsm}
  - \usepackage[nottoc, numbib]{tocbibind}
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install dependencies
if (!requireNamespace("tinytex", quietly=TRUE))
  install.packages("tinytex")

library(tinytex)

```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-02-02 markdown/Thesis.Rmd
structure of repo changed

---
title: " "
output: 
  pdf_document:
    number_sections: TRUE
geometry: "left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm"
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{sectsty}
  - \usepackage{paralist}
  - \usepackage{setspace}\spacing{1.5}
  - \usepackage{fancyhdr}
  - \usepackage{lastpage}
  - \usepackage{dcolumn}
  - \usepackage{natbib}\bibliographystyle{agsm}
  - \usepackage[nottoc, numbib]{tocbibind}
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
#designed with help of: 
#https://ourcodingclub.github.io/tutorials/rmarkdown-dissertation/

knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)

#install dependencies
if (!requireNamespace("tinytex", quietly=TRUE))
  install.packages("tinytex")

library(tinytex)
```

\allsectionsfont{\centering}
\subsectionfont{\raggedright}
\subsubsectionfont{\raggedright}

\pagenumbering{gobble}

\begin{centering}

\vspace{3cm}
```{r uni_logo, echo=F, out.width="20%"}
knitr::include_graphics("FHCW_logo.png")
knitr::include_graphics("KI_logo.png")
```

\vspace{1cm}

\Large
\doublespacing
{\bf Design and implementation of analysis pipeline for single cell type proteomics data}

\vspace{1 cm}

\normalsize
\singlespacing
By

\vspace{0.5 cm}

\Large

{\bf Lukas Gamp}

\vspace{1.5 cm}

in partial fulfilment of the requirement \\for the degree of MSc with Honours \\in Bioinformatics

\vspace{1.5 cm}

\normalsize
mm yy
\end{centering}

\newpage

\pagenumbering{gobble}

\begin{centering}

{\bf Abstract}

\end{centering}

\spacing{1.5}

(the spacing is set to 1.5)

no more than 250 words for the abstract

- a description of the research question/knowledge gap – what we know and what we don’t know
- how your research has attempted to fill this gap
- a brief description of the methods
- brief results
- key conclusions that put the research into a larger context

\pagenumbering{roman}

\newpage

\centering
\raggedright
\newpage
\tableofcontents

\newpage

```{r acknowledgments, child='acknowledgments.Rmd'}
```

\newpage

```{r intro, child = 'introduction.Rmd'}
```
\pagenumbering{arabic}

\newpage

```{r methods, child = 'methods.Rmd'}
```

\newpage

```{r results, child = 'results.Rmd'}
```

\newpage

```{r discussion, child = 'discussion.Rmd'}
```

\newpage

```{r conclusion, child = 'conclusion.Rmd'}
```

\newpage

```{r biblio, child = 'bibliography.Rmd'}
```

\newpage

```{r appendix, child = 'appendix.Rmd'}
```

### 2023-02-02 markdown/acknowledgments.Rmd
child files added

---
title: ''
output: pdf_document
---

\section*{Acknowledgements}

Thank you for following this tutorial!

I hope you'll find it useful to write a very professional dissertation.

### 2023-02-02 markdown/appendix.Rmd
child files added

---
title: ''
output: pdf_document
---
# Appendix(ces)

## Appendix A: additional tables

Insert content for additional tables here.

\newpage

## Appendix B: additional figures

Insert content for additional figures here.

\newpage

## Appendix C: code

Insert code (if any) used during your dissertation work here.

### 2023-02-02 markdown/conclusion.Rmd
child files added

---
title: ''
output: pdf_document
---

# Conclusion

The conclusion section should specify the key findings of your study, explain their wider significance in the context of the research field and explain how you have filled the knowledge gap that you have identified in the introduction. This is your chance to present to your reader the major take-home messages of your dissertation research. It should be similar in content to the last sentence of your summary abstract. It should not be a repetition of the first paragraph of the discussion. They can be distinguished in their connection to broader issues. The first paragraph of the discussion will tend to focus on the direct scientific implications of your work (i.e. basic science, fundamental knowledge) while the conclusion will tend to focus more on the implications of the results for society, conservation, etc.

### 2023-02-02 markdown/methods.Rmd
child files added

---
title: ''
output: pdf_document
---

# Methods

\vspace{0.5cm}

Write your methods here. In this tutorial you can use this already made file\\to add examples of figures and tables and explore knitr and kableExtra functionalities!

### 2023-02-02 markdown/results.Rmd
child files added

---
title: ''
output: pdf_document
---

# Results

Some more guidlines from the School of Geosciences.

This section should summarise the findings of the research referring to all figures, tables and statistical results (some of which may be placed in appendices).
- include the primary results, ordered logically - it is often useful to follow the same order as presented in the methods.
- alternatively, you may find that ordering the results from the most important to the least important works better for your project.
- data should only be presented in the main text once, either in tables or figures; if presented in figures, data can be tabulated in appendices and referred to at the appropriate point in the main text.

**Often, it is recommended that you write the results section first, so that you can write the methods that are appropriate to describe the results presented. Then you can write the discussion next, then the introduction which includes the relevant literature for the scientific story that you are telling and finally the conclusions and abstract – this approach is called writing backwards.**

### 2023-02-03 markdown/introduction.Rmd
literature recherche

---
title: " "
output: pdf_document
header-includes:
  - \usepackage{float}
  - \usepackage{sectsty}
---

# Introduction

## Mass Spectrometry

### MS-Spectrum

Every peptide is reflected by its` indivual fingerprint in the ms-spectrum. The fingerprint is based on the chemical properties and modifications. In order to identify proteins, fingerprints are matched against a sequence database \citep{Cox2008}.

### Interpretation of the data

Since ms data has a high resoultuion, algorithms are used to convert peaks to interpretable form. These algorithms find local minima of the function to seperate peaks from each other. Peaks include all isotopes of the protein`s containing atoms. MaxQuant \citep{Cox2008} is one of the sofware packages to process the data and provides it for further analysis and statistical testing.

follow sctranscr. analysis pipe

signatures of cell clustures

monocyte data acvailiabe

tmt label --> noise reduction

tools for clustering

	plst, knm,

### 2023-02-03 markdown/methods.Rmd
literature recherche

@@
 
-Write your methods here. In this tutorial you can use this already made file\\to add examples of figures and tables and explore knitr and kableExtra functionalities!
+# Materials and Methods
+
+## Materials
+
+For analysis two types of cells were used. One type is the Jurkat-based cell line (J-lat) with integrated HIV.
+
+The other type of cells are macrophages with a sample size of 72 cells. The analysis is done with two groups. A HIV negative (HIV-) control group and a HIV positive (HIV+) group.
+
+## Data
+
+### Acquisition
+
+Analysis of the data was done with MaxQuant \citep{Cox2008}

### 2023-02-03 markdown/introduction.Rmd
literature update

@@
 ---
-
 # Introduction
@@
 
-Since ms data has a high resoultuion, algorithms are used to convert peaks to interpretable form.
+Since ms data has a high resoultuion, algorithms are used to convert peaks to an interpretable form.
 These algorithms find local minima of the function to seperate peaks from each other.
@@
 MaxQuant \citep{Cox2008} is one of the sofware packages to process the data and provides it for further analysis and statistical testing.
+To find the istopic distribution of a biomolecue, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
+After this procedure a large amount of noise is reduced and a single peak reflects a small biomolecule.
+These biomolecules can now be searched in a database in forward and reverse direction.
+The peptide identification (P-) score indicates the fit of the data to the found sequence in the database according to the length of the peptide and is used to calculate the posterior error probability (FDR).
+The calculation of the false discovery rate is then calculated by taking FDR into contrast.
+After these calculation the peptide peaks are joined to a protein and can be quantified.
 
@@
 	plst, knm,
+
+follow sctranscr. analysis pipe
+
+signatures of cell clustures
+
+monocyte data acvailiabe
+
+tmt label --> noise reduction
+
+tools for clustering
+
+	plst, knm,

### 2023-02-06 markdown/introduction.Rmd
ESI added in introduction

@@
 
+## Proteomics
+
+The proteome is referred to the sum of all proteins of a given sample at a given time.
+In the past several quantitative and qualitative assays were used to enlighten the protein composition of a sample.
+
 ## Mass Spectrometry
+
+Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
+The availability of gene sequences in databases and the ability to match proteins against those sequences with computational methods makes it possible to identify alterations of a sample on a protein level.
+These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylation, methylation or else \citep{Aebersold2003}.
+
+### Instrumentation
+
+#### Ionization
+
+In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase.
+Two techniques are capable of this procedure.
+Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol.
+Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer.
+The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
+Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The carge residue model” (CRM).
+In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
+The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
+In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
+
+Matrix-assisted laser desorption/ionization (MALDI)
 
@@
 
-Every peptide is reflected by its` indivual fingerprint in the ms-spectrum.
+Each peptide is reflected by its` indivual fingerprint in the ms-spectrum.
 The fingerprint is based on the chemical properties and modifications.
 In order to identify proteins, fingerprints are matched against a sequence database \citep{Cox2008}.
+
+### Labeling techniques
 
@@
 	plst, knm,
-
-follow sctranscr. analysis pipe
-
-signatures of cell clustures
-
-monocyte data acvailiabe
-
-tmt label --> noise reduction
-
-tools for clustering
-
-	plst, knm,

### 2023-02-06 markdown/introduction.Rmd
SCoPE added to introduction

@@
 
+An early approach of qualitative analysis of the cellular proteome involved labeling with fluorescent antibodies and imaging.
+The major disadvantage of this technique was the limitation to only stain a few proteins per cell.
+For quantification procedures such as single-cell Western blots, immunoassays or CyTOF have been used.
+Other disadvantages are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
+
+One of those techniques involved RNA-sequencing.
+Since RNA involves also non-coding RNA, the amount of RNA is often not proportional to the content of proteins in a cell.
+So the proteinaceous content of a cell was only predicted and quantitative analysis was not possible.
+
 ## Mass Spectrometry
@@
 
+Mass to charge ratio (m/z)
+
 ### Instrumentation
+
+#### Cell Isolation
+
+#### Lysis
+
+#### Digestion
+
+#### Liquid chromatography
+
+In order to separate proteins according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
 
@@
 The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
-Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The carge residue model” (CRM).
+Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
 In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
@@
 Matrix-assisted laser desorption/ionization (MALDI)
+
+#### Coupled mass-spectrometry (MS/MS)
+
+In order to enhance sequence identification, two MS devices are built in series.
+In the first run (MS1) the m/z is determined and the molecules are passed to the next device.
+Upon passing the molecules are fragmented into smaller ions and analyzed by the second MS.
+The fragmentation highly depends on the chemical bonds found in the molecule.
+The majority of these breaks occur on the peptide bond of the protein, although this is not guaranteed for all bonds and so it can happen that certain peptide ions have a low abundance \citep{Budnik2018}.
+These low abundant peptides will not be detected, hence the problem needs to be faced with another approach .
+A solution for this problem is molecular barcoding with labeling mentioned in the chapter labeling.
+
+#### Labeling techniques
+
+##### Isobaric labeling
+
+###### Tandem mass tag (TMT)
+
+Tandem mass tag (TMT) reagents enable to differentiate multiple samples analyzing in one MS run. 
+The samples are labeled individually and pooled afterwards, this procedure is called multiplexing.
+All TMTs have the same mass and can be found at the same peak in all samples, these peaks are called reporter ions (RI).
+The identification of these RI leads to an enrichment and identification of low abundance peptide ions which is common especially in single-cell techniques.
+With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
+The disadvantage of isobaric labeling is, that the co-fragmentation signals can be observed in the spectrogram and the data needs to be normalized in order to remove unwanted contribution \citep{Marx2019, Budnik2018}.
 
@@
 
-### Labeling techniques
-
 ### Interpretation of the data
@@
 MaxQuant \citep{Cox2008} is one of the sofware packages to process the data and provides it for further analysis and statistical testing.
-To find the istopic distribution of a biomolecue, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
+To find the istopic distribution of a biomolecule, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
 After this procedure a large amount of noise is reduced and a single peak reflects a small biomolecule.
@@
 After these calculation the peptide peaks are joined to a protein and can be quantified.
+
+#### Principle component analysis (PCA)
+
+Protein levels can be projected to their principle components (PC) and clustered to their specific cell type. So cell types are distinguished by their proteinaceous composition.
+
+#### Testing for differential expression
+
+According to the experiment and sample groups, individual statistical methods should be used.
 

### 2023-02-07 markdown/introduction.Rmd
labeling techniques startedxy

@@
 
+For differential analysis proteins need to be labeled to compare mass to charge intensities.
+Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
+Physicochemical properties of the proteins can change the ionization efficiency and detectability of the target.
+However, when comparing the same analyte between multiple runs of labeled proteins, differences in the mass spectrum reflect the abundance of those.
+Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the protein.
+
+#### Metabolic labeling
+
+Feeding cells with aminoacids containing heavy isotopes, is the method of choice in order to label peptides at the earliest possible level.
+This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants.
+Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
+Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
+This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
+
 ##### Isobaric labeling

### 2023-02-15 markdown/introduction.Rmd
data acquisition completed downstream analysis started

@@
 ---
+# Introduction
+
 # Introduction
@@
 
-For differential analysis proteins need to be labeled to compare mass to charge intensities.
+For differential analysis proteins need to be labeled to compare mass to charge intensities in order to quantify observed peptides.
 Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
@@
 The samples are labeled individually and pooled afterwards, this procedure is called multiplexing.
-All TMTs have the same mass and can be found at the same peak in all samples, these peaks are called reporter ions (RI).
+TMTs have the same charge and differ only by their isotopic masses, the peaks found for each sample are called reporter ions (RI).
+Each RI and sample is interpreted as one channel in downstream analysis.
 The identification of these RI leads to an enrichment and identification of low abundance peptide ions which is common especially in single-cell techniques.
@@
 The disadvantage of isobaric labeling is, that the co-fragmentation signals can be observed in the spectrogram and the data needs to be normalized in order to remove unwanted contribution \citep{Marx2019, Budnik2018}.
+Furthermore TMTs have an isotopic distribution according to the distribution found in nature.
+This can be corrected during data-acquisition as a defined spread in other channels.
 
@@
 Each peptide is reflected by its` indivual fingerprint in the ms-spectrum.
-The fingerprint is based on the chemical properties and modifications.
-In order to identify proteins, fingerprints are matched against a sequence database \citep{Cox2008}.
+The fingerprint is based on the chemical properties and modifications of aminoacids.
+These aminoacids can be calculated through their m/z ration and after that interpreted as an aminoacid sequence.
+Due to fragmentation of the protein only peptide sequences are visible in the spectrum.
+In order to identify proteins, peptides are matched against a sequence database \citep{Cox2008}.
+Sequence Databases are simple .fasta files, which can be downloaded on the uniprot webpage (www.uniprot.org).
 
@@
 
-Since ms data has a high resoultuion, algorithms are used to convert peaks to an interpretable form.
-These algorithms find local minima of the function to seperate peaks from each other.
-Peaks include all isotopes of the protein`s containing atoms.
-MaxQuant \citep{Cox2008} is one of the sofware packages to process the data and provides it for further analysis and statistical testing.
-To find the istopic distribution of a biomolecule, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
-After this procedure a large amount of noise is reduced and a single peak reflects a small biomolecule.
-These biomolecules can now be searched in a database in forward and reverse direction.
+#### Data acquisition
+
+Since ms data has a high resolution, algorithms are used to convert the raw signal to an interpretable form. 
+MaxQuant  is one of many sofware packages to process the data and provides it for further analysis and statistical testing.
+Other software solutions are Protein Discoverer Thermo Fisher or even packages for R.
+In this publication we will mainly focus on the data-acquisition with MaxQuant \citep{Cox2008}.
+
+##### Three-dimensional peak detection
+
+The three dimensions of the data are: m/z ratio, intensity and retention time.
+The algorithm finds local minima of the function in order to seperate peaks from each other.
+The centroid of the peak is detected by fitting a so called gaussian peak shape fitting.
+This can be interpreted as finding the peaks of each m/z spectrum as a function of time..
+The centroid of the peak refers to an isotope.
+
+##### Deisotoping
+
+To decrypt the istopic distribution of a biomolecule, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
+Isotoping is the term of such procedure and it is enabled with graph theory.
+After this procedure the amount of data points are reduced by a tenfold and a single peak reflects a small biomolecule.
+
+##### Label detection
+
+The next step in data-acquisition is the detection of labels for quantification.
+Isotopic pairs of the label (e.g.
+N13, N14, N15) contained in the tag or aminoacid can be identified by convoluting the two measured isotope patterns with the theoretical isotope patterns.
+With a least-square method the best fit is found iteratively and the channel/sample can be identified.
+
+##### Improving peptide mass accuracy
+
+The intensity-weighted average of the ms peak centroids (as described in the 3D peak identification) refers to the mass of the peptide.
+Corrections highly depend on the analyzer, MaxQuant uses for an orbitrap typed analyzer a correction value of 1ppm.
+Autocorrelation between centroids is compensated by only using well-identified peptides.
+As published the mass precision within a ms experiment ranges around 10e-7.
+
+##### Peptide search
+
+Biomolecules can now be searched in a database in forward and reverse direction.
 The peptide identification (P-) score indicates the fit of the data to the found sequence in the database according to the length of the peptide and is used to calculate the posterior error probability (FDR).
 The calculation of the false discovery rate is then calculated by taking FDR into contrast.
-After these calculation the peptide peaks are joined to a protein and can be quantified.
 
-#### Principle component analysis (PCA)
+##### Protein assembly
+
+After these calculation the identified peptides can be aggregated according to it`s respective protein and  quantified.
+The mentioned metrics indicating the performance of the peptide search can be used in downstream analysis.
+A so called razor peptide indicates the group with the highest number of identified peptides.
+Quantification is enabled by taking only unique peptides into contrast.
+Posterior error probabilities, which refers to the chance that the found peptide is a random event, are multiplied and only distinct sequences with the highst-scoring are used.
+
+#### Downstream Analysis
+
+Further analysis is done with R and respective packages such as bioconductor.
+
+##### Reading the data
+
+##### Principle component analysis (PCA)
 

### 2023-02-15 scope2.R

@@ -12,7 +12,10 @@ library("dplyr")
 
 
 # read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/combined/txt/evidence.txt")
+mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/mq-run_150223/combined/txt/evidence.txt")
+
+akos_data <- read.csv("/home/lukas/Downloads/20230202_HF2_08_Ujjwal_Monocytes_TMT12_proteins.csv")
+
 
 
 # create annotation file
@@ -31,14 +34,19 @@ generate_var = function(prefix, length) {
   paste(prefix, sprintf('%03d', seq_len(length)), sep = '_')
 }
 
-sampleAnnotation$SampleType <- c(generate_var("example", nrow(sampleAnnotation)))
+samplesA <- c(generate_var("A", 6))
+samplesB <- c(generate_var("B", 6))
+samples <- c(samplesA, samplesB)
+
+
+sampleAnnotation$SampleType <- samples
 
 
 scp <- readSCP(featureData = mqScpData,
                colData = sampleAnnotation,
                channelCol = "Channel",
                batchCol = "Raw.file",
-               suffix = paste0("_TMT", 1:6),
+               suffix = paste0("_TMT", 1:12),
                removeEmptyCols = TRUE)
 
 # check data

### 2023-02-16 markdown/introduction.Rmd
quality control started

@@
 ---
-# Introduction
-
 # Introduction
@@
 
-#### Downstream Analysis
+#### Data processing
 
-Further analysis is done with R and respective packages such as bioconductor.
+Further analysis can be done with R and respective packages such as bioconductor. Since there is no state of the art established, analysis varies upon experimental design.
 
 ##### Reading the data
+
+After processing MaxQuant creates a directory containing all results as .txt file.
+The evidence.txt file include all peptide sequence matches (PSM) with their respective proteins and statistical parameters.
+
+Example fo basic parameters and derivations include:
+
+* Peptide sequence
+* Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)
+	+ Mass
+* Retention time
+* Precursor Ion Fragment
+	+ source of the detected ion also referred as mother ion
+* Fraction of total spectrum 
+* Base peak fraction
+* Reporter intensity (RI)
+	+ corrected RI
+* Posterior error probability (PEP)
+
+#### Object oriented programming
+
+In order to streamline the analysis of multiple experiments, object oriented programming can be applied.
+The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
+The structure enables sub setting for further analysis.
+
+#### Filtering
+
+Peptides having a low abundance will be set to zero by the analysis.
+The value zero would imply that the sample does not contain the respective peptide, therefore replacing 0 with not applicable (NA) is crucial for understanding and interpretation of MS-data.
+
+Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.
+
+The author of the SCoPE2 \citep{Petelski2021} pipeline recommends calculating the sample to carrier ration (SCR).
+This can be done by dviding the ion intensity of the sample by the reporter ion intensity of the carrier channel.
+However in our experiment no carrier and reference channel is used.
+
+The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
+When applying multiple statistical testing (e.g.
+t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed.
+Corrections in statistics are an approach to compensate for the multiplicity of testing.
+There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR.
+In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score.
+The peptides pectrum matching score is defined as -10log10(p).
+Whereas the p-value is defined that the hit is done by chance.
+The calculation of the score is highly dependent on the data acquisition method used.
+MaxQuant uses Andromeda , an integrated search engine.
+Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora.
+As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage.
+However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.
+
+Further filter steps can include correction between multiple runs.
+This kind of process need the addition of a reference channel.
+However when observing a single run, these steps are not crucial for the upcoming analysis.
+
+#### Downstream Analysis
 

### 2023-02-16 scope2.R

@@ -1,3 +1,8 @@
+
+# workflow as per 
+# https://uclouvain-cbio.github.io/SCP.replication/articles/SCoPE2.html
+# and https://bioconductor.org/packages/release/bioc/vignettes/scp/inst/doc/scp.html
+
 # install scp package
 
 #BiocManager::install("scp")
@@ -13,7 +18,6 @@ library("dplyr")
 
 # read in MS result table
 mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/mq-run_150223/combined/txt/evidence.txt")
-
 akos_data <- read.csv("/home/lukas/Downloads/20230202_HF2_08_Ujjwal_Monocytes_TMT12_proteins.csv")
 
 
@@ -22,6 +26,7 @@ akos_data <- read.csv("/home/lukas/Downloads/20230202_HF2_08_Ujjwal_Monocytes_TM
 # this varies upon experimental design
 quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData), value = T)
 
+
 # create sample file
 # this needs to be done by the researcher
 sampleAnnotation <- as.data.frame(quantCols)
@@ -29,19 +34,14 @@ colnames(sampleAnnotation) <-c("Channel")
 sampleAnnotation$Raw.file = unique(mqScpData$Raw.file)
 
 
-# for example generated --> fix accordingly
-generate_var = function(prefix, length) {
-  paste(prefix, sprintf('%03d', seq_len(length)), sep = '_')
-}
-
-samplesA <- c(generate_var("A", 6))
-samplesB <- c(generate_var("B", 6))
+samplesA <- c(replicate(6, "Monocytes-A"))
+samplesB <- c(replicate(6, "Monocytes-B"))
 samples <- c(samplesA, samplesB)
 
 
 sampleAnnotation$SampleType <- samples
 
-
+# create QFeature object
 scp <- readSCP(featureData = mqScpData,
                colData = sampleAnnotation,
                channelCol = "Channel",
@@ -49,160 +49,99 @@ scp <- readSCP(featureData = mqScpData,
                suffix = paste0("_TMT", 1:12),
                removeEmptyCols = TRUE)
 
-# check data
-head(colData(scp))
 
-#plot rawfile
-plot(scp)
+# plot runs (not needed - sample unfractionized)
+# only one run
+# plot(scp)
 
-# workflow as per 
[…]

### 2023-03-08 scope2.R
volcano hover established

@@ -400,8 +400,9 @@ col_index1 <- which(colnames(design) == user_input1)
 col_index2 <- which(colnames(design) == user_input2)
 
 contrast <- paste(user_input1,"-",user_input2)
+contrast = "AvsB = Monocytes_A - Monocytes_B"
 
-cont_matrix <- makeContrasts(AvsB = contrast, levels=design)
+cont_matrix <- makeContrasts(contrast, levels=design)
 
 # Fit the expression matrix to a linear model
 fit <- lmFit(exp_matrix, design)

### 2023-03-08 scpR-app/app.R

@@ -24,7 +24,7 @@ ui <- fluidPage(
       # read in of the data
       fileInput("evidence_file", "Upload evidence.txt from MaxQuant", accept = c("text")),
       fileInput("sample_annotation_file", "Upload sample annotation file", accept = c("text")),
-      textInput("label_suffix", "Input the label suffix (e.g. _TMT"),
+      textInput("label_suffix", "Input the label suffix (e.g. _TMT)"),
       
       # cutoff for PIF
       numericInput("PIF_cutoff", "Input cutoff value for parental ion fraction. PSMs larger than the value remain", 0.1, min = 0, max=1, step = 0.1),
@@ -56,11 +56,13 @@ ui <- fluidPage(
                   tabPanel("Feature wise output", selectInput("selectedProtein", "Choose protein for observation", 
                                                               choices=c(rowData(scp)[["proteins"]][,1])),
                            plotOutput("feature_subset")),
+                  
                   tabPanel("Statistics",
-                           actionButton("run_statistics", "Press to run statistics"), 
-                           selectInput("comp1", "Select comparison 1",choices=c(unique(scp$SampleType))),
-                           selectInput("comp2", "Select comparison 2", choices=c(unique(scp$SampleType))),
-                           plotOutput("volcano"),
+                           actionButton("run_statistics", "Press to run statistics"),
+                           textOutput("model_design"),
+                           textInput("user_contrast", "Define your contrast model (w/o spaces)"),
+                           plotOutput("volcano", hover = hoverOpts(id ="plot_hover")),
+                           verbatimTextOutput("hover_info"),
                            tableOutput("protein_table"),
                            textOutput("summary")
                            )
@@ -261,12 +263,10 @@ server <- function(input, output) {
       # assign the column names
       colnames(design) <- unique(scp_0$SampleType)
       
-      req(input$comp1)
-      req(input$comp2)
-      
-      contrast = paste(input$comp1,"-",input$comp2)
-      
-      cont_matrix <- makeContrasts(AvsB = contrast, levels=design)
+      req(input$user_contrast)
+      contrast_0 <- strsplit(input$user_contrast, " ")
+      # calculates the mean difference between groups
+      cont_matrix <- makeContrasts(contrasts=contrast_0, levels=design)
       
[…]

### 2023-03-08 scpR-app/app.R
statistic update

@@ -61,10 +61,10 @@ ui <- fluidPage(
                            actionButton("run_statistics", "Press to run statistics"),
                            textOutput("model_design"),
                            textInput("user_contrast", "Define your contrast model (w/o spaces)"),
+                           tableOutput("summary"),
                            plotOutput("volcano", hover = hoverOpts(id ="plot_hover")),
                            verbatimTextOutput("hover_info"),
-                           tableOutput("protein_table"),
-                           textOutput("summary")
+                           tableOutput("protein_table")
                            )
       )
     )
@@ -448,10 +448,10 @@ server <- function(input, output) {
       topTable(stat_result(), number = 10, adjust = "BH")
     }, rownames = T)
     
-    output$summary <- renderText({
-      result <- decideTests(stat_result())
-      summary(result)
-    })
+    output$summary <- renderTable({
+      req(stat_result())
+      summary(decideTests(stat_result()))
+    }, colnames = F)
     
     output$model_design <- renderText({
       unique(scp()$SampleType)

### 2023-03-08 markdown/Thesis.tex
Statistical testing in thesis started

--- 2023-02-16 markdown/introduction.Rmd
@@
----
-title: " "
-output: pdf_document
-header-includes:
-  - \usepackage{float}
-  - \usepackage{sectsty}
----
-# Introduction
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  11pt,
+]{article}
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
+2cm]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+\usepackage{float}
+\usepackage{sectsty}
+\usepackage{paralist}
+\usepackage{setspace}\spacing{1.5}
+\usepackage{fancyhdr}
+\usepackage{lastpage}
+\usepackage{dcolumn}
+\usepackage{natbib}\bibliographystyle{agsm}
+\usepackage[nottoc, numbib]{tocbibind}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
-## Proteomics
+\author{}
+\date{\vspace{-2.5em}}
 
-The proteome is referred to the sum of all proteins of a given sample at a given time.
-In the past several quantitative and qualitative assays were used to enlighten the protein composition of a sample.
+\begin{document}
 
-An early approach of qualitative analysis of the cellular proteome involved labeling with fluorescent antibodies and imaging.
-The major disadvantage of this technique was the limitation to only stain a few proteins per cell.
-For quantification procedures such as single-cell Western blots, immunoassays or CyTOF have been used.
-Other disadvantages are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
+\allsectionsfont{\centering}
+\subsectionfont{\raggedright}
+\subsubsectionfont{\raggedright}
 
-One of those techniques involved RNA-sequencing.
-Since RNA involves also non-coding RNA, the amount of RNA is often not proportional to the content of proteins in a cell.
-So the proteinaceous content of a cell was only predicted and quantitative analysis was not possible.
+\pagenumbering{gobble}
 
-## Mass Spectrometry
+\begin{centering}
 
-Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
-The availability of gene sequences in databases and the ability to match proteins against those sequences with computational methods makes it possible to identify alterations of a sample on a protein level.
-These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylation, methylation or else \citep{Aebersold2003}.
+\vspace{3cm}
+
+\includegraphics[width=0.2\linewidth]{FHCW_logo} 
+\includegraphics[width=0.2\linewidth]{KI_logo}
+
+\vspace{1cm}
+
+\Large
+\doublespacing
+{\bf Design and implementation of analysis pipeline for single cell type proteomics data}
+
+\vspace{1 cm}
+
+\normalsize
+\singlespacing
+By
+
+\vspace{0.5 cm}
+
+\Large
+
+{\bf Lukas Gamp}
+
+\vspace{1.5 cm}
+
+in partial fulfilment of the requirement \\for the degree of MSc \\in Bioinformatics
+
+\vspace{1.5 cm}
+
+\normalsize
+mm yy
+\end{centering}
+
+\newpage
+
+\pagenumbering{gobble}
+
+\begin{centering}
+
+{\bf Abstract}
+
+\end{centering}
+
+\spacing{1.5}
+
+(the spacing is set to 1.5)
+
+no more than 250 words for the abstract
+
+\begin{itemize}
+\tightlist
+\item
+  a description of the research question/knowledge gap -- what we know
+  and what we don't know
+\item
+  how your research has attempted to fill this gap
+\item
+  a brief description of the methods
+\item
+  brief results
+\item
+  key conclusions that put the research into a larger context
+\end{itemize}
+
+\pagenumbering{roman}
+
+\newpage
+
+\centering
+\raggedright
+\newpage
+\tableofcontents
+
+\newpage
+
+\section*{Acknowledgements}
+
+Thank you for following this tutorial!
+
+I hope you'll find it useful to write a very professional dissertation.
+
+\newpage
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+\hypertarget{proteomics}{%
+\subsection{Proteomics}\label{proteomics}}
+
+The proteome is referred to the sum of all proteins of a given sample at
+a given time. In the past several quantitative and qualitative assays
+were used to enlighten the protein composition of a sample.
+
+An early approach of qualitative analysis of the cellular proteome
+involved labeling with fluorescent antibodies and imaging. The major
+disadvantage of this technique was the limitation to only stain a few
+proteins per cell. For quantification procedures such as single-cell
+Western blots, immunoassays or CyTOF have been used. Other disadvantages
+are the ability to permeate cells, accessibility and binding of the
+epitope and the creation of specific antibodies for a given protein
+\citep{Budnik2018}.
+
+One of those techniques involved RNA-sequencing. Since RNA involves also
+non-coding RNA, the amount of RNA is often not proportional to the
+content of proteins in a cell. So the proteinaceous content of a cell
+was only predicted and quantitative analysis was not possible.
+
+\hypertarget{mass-spectrometry}{%
+\subsection{Mass Spectrometry}\label{mass-spectrometry}}
+
+Mass spectrometry enables qualitative and quantitative analysis of the
+entire repertoire of a biological sample. The availability of gene
+sequences in databases and the ability to match proteins against those
+sequences with computational methods makes it possible to identify
+alterations of a sample on a protein level. These alterations can rely
+on the sequence level or could be to post-translational modifications
+(PTMs) such as phosphorylation, methylation or else
+\citep{Aebersold2003}.
 
@@
 
-### Instrumentation
+\hypertarget{instrumentation}{%
+\subsubsection{Instrumentation}\label{instrumentation}}
 
-#### Cell Isolation
+\hypertarget{cell-isolation}{%
+\paragraph{Cell Isolation}\label{cell-isolation}}
 
-#### Lysis
+\hypertarget{lysis}{%
+\paragraph{Lysis}\label{lysis}}
 
-#### Digestion
+\hypertarget{digestion}{%
+\paragraph{Digestion}\label{digestion}}
 
-#### Liquid chromatography
+\hypertarget{liquid-chromatography}{%
+\paragraph{Liquid chromatography}\label{liquid-chromatography}}
 
-In order to separate proteins according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
+In order to separate proteins according to their chemical properties,
+size or species a liquid chromatography (LC) is recommended before
+ionization.
 
-#### Ionization
+\hypertarget{ionization}{%
+\paragraph{Ionization}\label{ionization}}
 
-In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase.
-Two techniques are capable of this procedure.
-Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol.
-Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer.
-The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
-Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
-In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
-The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
-In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
+In order to analyze a biological sample consisting of proteins in
+solution the liquid needs to be vaporized into gas phase. Two techniques
+are capable of this procedure. Electrospray ionization (ESI) pushes the
+analyte through a capillary and applies an electric current to the
+liquid, vaporizing the sample to a charged aerosol. Biomolecules are
+fragmented according to their chemical properties and can be further
+handled in the mass spectrometer. The fragmented biomolecules are now in
+charged droplets separated by their charge on the surface, splitting
+further into smaller droplets until they become a gas phase ion. Two
+physical models describe the process from gas phase to ion called ``The
+ion evaporation model'' (IEM) and ``The charge residue model'' (CRM). In
+the ion evaporation model (by Iribarne and Thomoson) the droplets shrink
+by evaporation until ions are expelled. The model had its limitation by
+explaining same evaporation rate constant among ions with different
+chemical properties. In the charge residue model the assumption of one
+molecule per droplet leads to an ionization rate constant, which is
+independent of the ion itself and relies solely on the generation of the
+droplet and the efficiency of the solvent \citep{Wilm2011}.
 
@@
 
-#### Coupled mass-spectrometry (MS/MS)
+\hypertarget{coupled-mass-spectrometry-msms}{%
+\paragraph{Coupled mass-spectrometry
+(MS/MS)}\label{coupled-mass-spectrometry-msms}}
 
-In order to enhance sequence identification, two MS devices are built in series.
-In the first run (MS1) the m/z is determined and the molecules are passed to the next device.
-Upon passing the molecules are fragmented into smaller ions and analyzed by the second MS.
-The fragmentation highly depends on the chemical bonds found in the molecule.
-The majority of these breaks occur on the peptide bond of the protein, although this is not guaranteed for all bonds and so it can happen that certain peptide ions have a low abundance \citep{Budnik2018}.
-These low abundant peptides will not be detected, hence the problem needs to be faced with another approach .
-A solution for this problem is molecular barcoding with labeling mentioned in the chapter labeling.
+In order to enhance sequence identification, two MS devices are built in
+series. In the first run (MS1) the m/z is determined and the molecules
+are passed to the next device. Upon passing the molecules are fragmented
+into smaller ions and analyzed by the second MS. The fragmentation
+highly depends on the chemical bonds found in the molecule. The majority
+of these breaks occur on the peptide bond of the protein, although this
+is not guaranteed for all bonds and so it can happen that certain
+peptide ions have a low abundance \citep{Budnik2018}. These low abundant
+peptides will not be detected, hence the problem needs to be faced with
+another approach . A solution for this problem is molecular barcoding
+with labeling mentioned in the chapter labeling.
 
-#### Labeling techniques
+\hypertarget{labeling-techniques}{%
+\paragraph{Labeling techniques}\label{labeling-techniques}}
 
-For differential analysis proteins need to be labeled to compare mass to charge intensities in order to quantify observed peptides.
-Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
-Physicochemical properties of the proteins can change the ionization efficiency and detectability of the target.
-However, when comparing the same analyte between multiple runs of labeled proteins, differences in the mass spectrum reflect the abundance of those.
-Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the protein.
+For differential analysis proteins need to be labeled to compare mass to
+charge intensities in order to quantify observed peptides. Since mass
+spectrometry is not a quantitative technique by itself, the peak height
+or area does not reflect the abundance of a peptide. Physicochemical
+properties of the proteins can change the ionization efficiency and
+detectability of the target. However, when comparing the same analyte
+between multiple runs of labeled proteins, differences in the mass
+spectrum reflect the abundance of those. Labels should be chosen to
+change solely the mass of the sample and to not affect folding or other
+inherent properties of the protein.
 
-#### Metabolic labeling
+\hypertarget{metabolic-labeling}{%
+\paragraph{Metabolic labeling}\label{metabolic-labeling}}
 
-Feeding cells with aminoacids containing heavy isotopes, is the method of choice in order to label peptides at the earliest possible level.
-This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants.
-Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
-Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
-This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
+Feeding cells with aminoacids containing heavy isotopes, is the method
+of choice in order to label peptides at the earliest possible level.
+This atoms can be heavy nitrogen in aminoacids or salts in fertilizer
+for plants. Mass shifts are proportional to the isotopes incorporated
+during biomass production and are visible after proteolytic cleavage.
+Stable isotope labeling in cell culture (SILAC) was presented in the
+early 2000s. This method used heavy aminoacid enriched media to feed
+cells, in order to quantitatively analyze expression profiles.
 
-##### Isobaric labeling
+\hypertarget{isobaric-labeling}{%
+\subparagraph{Isobaric labeling}\label{isobaric-labeling}}
 
-###### Tandem mass tag (TMT)
+Tandem mass tag (TMT)
 
-Tandem mass tag (TMT) reagents enable to differentiate multiple samples analyzing in one MS run. 
-The samples are labeled individually and pooled afterwards, this procedure is called multiplexing.
-TMTs have the same charge and differ only by their isotopic masses, the peaks found for each sample are called reporter ions (RI).
-Each RI and sample is interpreted as one channel in downstream analysis.
-The identification of these RI leads to an enrichment and identification of low abundance peptide ions which is common especially in single-cell techniques.
-With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
-The disadvantage of isobaric labeling is, that the co-fragmentation signals can be observed in the spectrogram and the data needs to be normalized in order to remove unwanted contribution \citep{Marx2019, Budnik2018}.
-Furthermore TMTs have an isotopic distribution according to the distribution found in nature.
-This can be corrected during data-acquisition as a defined spread in other channels.
+Tandem mass tag (TMT) reagents enable to differentiate multiple samples
+analyzing in one MS run. The samples are labeled individually and pooled
+afterwards, this procedure is called multiplexing. TMTs have the same
+charge and differ only by their isotopic masses, the peaks found for
+each sample are called reporter ions (RI). Each RI and sample is
+interpreted as one channel in downstream analysis. The identification of
+these RI leads to an enrichment and identification of low abundance
+peptide ions which is common especially in single-cell techniques. With
+this technique it is possible to quantify proteins and differ low
+abundant proteins from background noise. The disadvantage of isobaric
+labeling is, that the co-fragmentation signals can be observed in the
+spectrogram and the data needs to be normalized in order to remove
+unwanted contribution \citep{Marx2019, Budnik2018}. Furthermore TMTs
+have an isotopic distribution according to the distribution found in
+nature. This can be corrected during data-acquisition as a defined
+spread in other channels.
 
-### MS-Spectrum
+\hypertarget{ms-spectrum}{%
+\subsubsection{MS-Spectrum}\label{ms-spectrum}}
 
-Each peptide is reflected by its` indivual fingerprint in the ms-spectrum.
-The fingerprint is based on the chemical properties and modifications of aminoacids.
-These aminoacids can be calculated through their m/z ration and after that interpreted as an aminoacid sequence.
-Due to fragmentation of the protein only peptide sequences are visible in the spectrum.
-In order to identify proteins, peptides are matched against a sequence database \citep{Cox2008}.
-Sequence Databases are simple .fasta files, which can be downloaded on the uniprot webpage (www.uniprot.org).
+Each peptide is reflected by its` indivual fingerprint in the
+ms-spectrum. The fingerprint is based on the chemical properties and
+modifications of aminoacids. These aminoacids can be calculated through
+their m/z ration and after that interpreted as an aminoacid sequence.
+Due to fragmentation of the protein only peptide sequences are visible
+in the spectrum. In order to identify proteins, peptides are matched
+against a sequence database \citep{Cox2008}. Sequence Databases are
+simple .fasta files, which can be downloaded on the uniprot webpage
+(www.uniprot.org).
 
-### Interpretation of the data
+\hypertarget{interpretation-of-the-data}{%
+\subsubsection{Interpretation of the
+data}\label{interpretation-of-the-data}}
 
-#### Data acquisition
+\hypertarget{data-acquisition}{%
+\paragraph{Data acquisition}\label{data-acquisition}}
 
-Since ms data has a high resolution, algorithms are used to convert the raw signal to an interpretable form. 
-MaxQuant  is one of many sofware packages to process the data and provides it for further analysis and statistical testing.
-Other software solutions are Protein Discoverer Thermo Fisher or even packages for R.
-In this publication we will mainly focus on the data-acquisition with MaxQuant \citep{Cox2008}.
+Since ms data has a high resolution, algorithms are used to convert the
+raw signal to an interpretable form. MaxQuant is one of many sofware
+packages to process the data and provides it for further analysis and
+statistical testing. Other software solutions are Protein Discoverer
+Thermo Fisher or even packages for R. In this publication we will mainly
+focus on the data-acquisition with MaxQuant \citep{Cox2008}.
 
-##### Three-dimensional peak detection
+\hypertarget{three-dimensional-peak-detection}{%
+\subparagraph{Three-dimensional peak
+detection}\label{three-dimensional-peak-detection}}
 
-The three dimensions of the data are: m/z ratio, intensity and retention time.
-The algorithm finds local minima of the function in order to seperate peaks from each other.
-The centroid of the peak is detected by fitting a so called gaussian peak shape fitting.
-This can be interpreted as finding the peaks of each m/z spectrum as a function of time..
-The centroid of the peak refers to an isotope.
+The three dimensions of the data are: m/z ratio, intensity and retention
+time. The algorithm finds local minima of the function in order to
+seperate peaks from each other. The centroid of the peak is detected by
+fitting a so called gaussian peak shape fitting. This can be interpreted
+as finding the peaks of each m/z spectrum as a function of time.. The
+centroid of the peak refers to an isotope.
 
-##### Deisotoping
+\hypertarget{deisotoping}{%
+\subparagraph{Deisotoping}\label{deisotoping}}
 
-To decrypt the istopic distribution of a biomolecule, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
-Isotoping is the term of such procedure and it is enabled with graph theory.
-After this procedure the amount of data points are reduced by a tenfold and a single peak reflects a small biomolecule.
+To decrypt the istopic distribution of a biomolecule, MaxQuant creates a
+vertex of every single peak and connects them with their possible
+isotopic counterparts by finding the proportion of mass of an average
+aminoacid to its` respective isotope (averagine \citep{Senko1995}).
+Isotoping is the term of such procedure and it is enabled with graph
+theory. After this procedure the amount of data points are reduced by a
+tenfold and a single peak reflects a small biomolecule.
 
-##### Label detection
+\hypertarget{label-detection}{%
+\subparagraph{Label detection}\label{label-detection}}
 
-The next step in data-acquisition is the detection of labels for quantification.
-Isotopic pairs of the label (e.g.
-N13, N14, N15) contained in the tag or aminoacid can be identified by convoluting the two measured isotope patterns with the theoretical isotope patterns.
-With a least-square method the best fit is found iteratively and the channel/sample can be identified.
+The next step in data-acquisition is the detection of labels for
+quantification. Isotopic pairs of the label (e.g.~N13, N14, N15)
+contained in the tag or aminoacid can be identified by convoluting the
+two measured isotope patterns with the theoretical isotope patterns.
+With a least-square method the best fit is found iteratively and the
+channel/sample can be identified.
 
-##### Improving peptide mass accuracy
+\hypertarget{improving-peptide-mass-accuracy}{%
+\subparagraph{Improving peptide mass
+accuracy}\label{improving-peptide-mass-accuracy}}
 
-The intensity-weighted average of the ms peak centroids (as described in the 3D peak identification) refers to the mass of the peptide.
-Corrections highly depend on the analyzer, MaxQuant uses for an orbitrap typed analyzer a correction value of 1ppm.
-Autocorrelation between centroids is compensated by only using well-identified peptides.
-As published the mass precision within a ms experiment ranges around 10e-7.
+The intensity-weighted average of the ms peak centroids (as described in
+the 3D peak identification) refers to the mass of the peptide.
+Corrections highly depend on the analyzer, MaxQuant uses for an orbitrap
+typed analyzer a correction value of 1ppm. Autocorrelation between
+centroids is compensated by only using well-identified peptides. As
+published the mass precision within a ms experiment ranges around 10e-7.
 
-##### Peptide search
+\hypertarget{peptide-search}{%
+\subparagraph{Peptide search}\label{peptide-search}}
 
-Biomolecules can now be searched in a database in forward and reverse direction.
-The peptide identification (P-) score indicates the fit of the data to the found sequence in the database according to the length of the peptide and is used to calculate the posterior error probability (FDR).
-The calculation of the false discovery rate is then calculated by taking FDR into contrast.
+Biomolecules can now be searched in a database in forward and reverse
+direction. The peptide identification (P-) score indicates the fit of
+the data to the found sequence in the database according to the length
+of the peptide and is used to calculate the posterior error probability
+(FDR). The calculation of the false discovery rate is then calculated by
+taking FDR into contrast.
 
-##### Protein assembly
+\hypertarget{protein-assembly}{%
+\subparagraph{Protein assembly}\label{protein-assembly}}
 
-After these calculation the identified peptides can be aggregated according to it`s respective protein and  quantified.
-The mentioned metrics indicating the performance of the peptide search can be used in downstream analysis.
-A so called razor peptide indicates the group with the highest number of identified peptides.
-Quantification is enabled by taking only unique peptides into contrast.
-Posterior error probabilities, which refers to the chance that the found peptide is a random event, are multiplied and only distinct sequences with the highst-scoring are used.
+After these calculation the identified peptides can be aggregated
+according to it`s respective protein and quantified. The mentioned
+metrics indicating the performance of the peptide search can be used in
+downstream analysis. A so called razor peptide indicates the group with
+the highest number of identified peptides. Quantification is enabled by
+taking only unique peptides into contrast. Posterior error
+probabilities, which refers to the chance that the found peptide is a
+random event, are multiplied and only distinct sequences with the
+highst-scoring are used.
 
-#### Data processing
+\hypertarget{data-processing}{%
+\paragraph{Data processing}\label{data-processing}}
 
-Further analysis can be done with R and respective packages such as bioconductor. Since there is no state of the art established, analysis varies upon experimental design.
+Further analysis can be done with R and respective packages such as
+bioconductor. Since there is no state of the art established, analysis
+varies upon experimental design.
 
-##### Reading the data
+\hypertarget{reading-the-data}{%
+\subparagraph{Reading the data}\label{reading-the-data}}
 
-After processing MaxQuant creates a directory containing all results as .txt file.
-The evidence.txt file include all peptide sequence matches (PSM) with their respective proteins and statistical parameters.
+After processing MaxQuant creates a directory containing all results as
+.txt file. The evidence.txt file include all peptide sequence matches
+(PSM) with their respective proteins and statistical parameters.
 
@@
 
-* Peptide sequence
-* Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)
-	+ Mass
-* Retention time
-* Precursor Ion Fragment
-	+ source of the detected ion also referred as mother ion
-* Fraction of total spectrum 
-* Base peak fraction
-* Reporter intensity (RI)
-	+ corrected RI
-* Posterior error probability (PEP)
+\begin{itemize}
+\tightlist
+\item
+  Peptide sequence
+\item
+  Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)
 
-#### Object oriented programming
+  \begin{itemize}
+  \tightlist
+  \item
+    Mass
+  \end{itemize}
+\item
+  Retention time
+\item
+  Precursor Ion Fragment
 
-In order to streamline the analysis of multiple experiments, object oriented programming can be applied.
-The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
-The structure enables sub setting for further analysis.
+  \begin{itemize}
+  \tightlist
+  \item
+    source of the detected ion also referred as mother ion
+  \end{itemize}
+\item
+  Fraction of total spectrum
+\item
+  Base peak fraction
+\item
+  Reporter intensity (RI)
 
-#### Filtering
+  \begin{itemize}
+  \tightlist
+  \item
+    corrected RI
+  \end{itemize}
+\item
+  Posterior error probability (PEP)
+\end{itemize}
 
-Peptides having a low abundance will be set to zero by the analysis.
-The value zero would imply that the sample does not contain the respective peptide, therefore replacing 0 with not applicable (NA) is crucial for understanding and interpretation of MS-data.
+\hypertarget{object-oriented-programming}{%
+\paragraph{Object oriented
+programming}\label{object-oriented-programming}}
 
-Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.
+In order to streamline the analysis of multiple experiments, object
+oriented programming can be applied. The approach in R is to create a so
+called Q-feature object, which contains all variables and metadata in a
+hierarchical structure. The structure enables sub setting for further
+analysis.
 
-The author of the SCoPE2 \citep{Petelski2021} pipeline recommends calculating the sample to carrier ration (SCR).
-This can be done by dviding the ion intensity of the sample by the reporter ion intensity of the carrier channel.
-However in our experiment no carrier and reference channel is used.
+\hypertarget{filtering}{%
+\paragraph{Filtering}\label{filtering}}
 
-The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
-When applying multiple statistical testing (e.g.
-t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed.
-Corrections in statistics are an approach to compensate for the multiplicity of testing.
-There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR.
-In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score.
-The peptides pectrum matching score is defined as -10log10(p).
-Whereas the p-value is defined that the hit is done by chance.
-The calculation of the score is highly dependent on the data acquisition method used.
-MaxQuant uses Andromeda , an integrated search engine.
-Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora.
-As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage.
-However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.
+Peptides having a low abundance will be set to zero by the analysis. The
+value zero would imply that the sample does not contain the respective
+peptide, therefore replacing 0 with not applicable (NA) is crucial for
+understanding and interpretation of MS-data.
 
-Further filter steps can include correction between multiple runs.
-This kind of process need the addition of a reference channel.
-However when observing a single run, these steps are not crucial for the upcoming analysis.
+Peptide sequences matching to the reverse protein sequences (=decoy
+database) are considered as possible contaminants. These matches can be
+excluded from further analysis.
 
-#### Downstream Analysis
+The author of the SCoPE2 \citep{Petelski2021} pipeline recommends
+calculating the sample to carrier ration (SCR). This can be done by
+dviding the ion intensity of the sample by the reporter ion intensity of
+the carrier channel. However in our experiment no carrier and reference
+channel is used.
 
-##### Principle component analysis (PCA)
+The next step for quality control is the exclusion of samples with a
+high false discovery rate (FDR). When applying multiple statistical
+testing (e.g.~t-Test) the obtained p-values can be considered as biased,
+because the probability to observe a significant will iteratively
+increase with each test performed. Corrections in statistics are an
+approach to compensate for the multiplicity of testing. There are many
+ways to do this compensation like the Bonferroni method or
+Benjamini-Hochberg`s FDR. In Mass spectrometry the common ``way to go''
+is calculating a false discovery rate, by dividing false PSMs (=hit of
+the decoy database) through the total number of PSMs above the
+peptide-spectrum matching score. The peptides spectrum matching score is
+defined as -10log10(p). Whereas the p-value is defined that the hit is
+done by chance. The calculation of the score is highly dependent on the
+data acquisition method used. MaxQuant uses Andromeda , an integrated
+search engine. Proteome Discoverer from Thermo Fisher utilizes different
+engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot
+and Andromeda showed similar performance when comparing FDR values as a
+function of coverage. However the observed performance can be lower when
+dealing with a decreased coverage \citep{Cox2011}.The threshold for
+accepting an FDR of an individual PSM is described as q-value.
 
-Protein levels can be projected to their principle components (PC) and clustered to their specific cell type. So cell types are distinguished by their proteinaceous composition.
+Further filter steps can include correction between multiple runs. This
+kind of process need the addition of a reference channel. However when
+observing a single run, these steps are not crucial for the upcoming
+analysis.
 
-#### Testing for differential expression
+\hypertarget{downstream-analysis}{%
+\paragraph{Downstream Analysis}\label{downstream-analysis}}
 
-According to the experiment and sample groups, individual statistical methods should be used.
+\hypertarget{principle-component-analysis-pca}{%
+\subparagraph{Principle component analysis
+(PCA)}\label{principle-component-analysis-pca}}
 
-follow sctranscr. analysis pipe
+Protein levels can be projected to their principle components (PC) and
+clustered to their specific cell type. So cell types are distinguished
+by their proteinaceous composition.
 
-signatures of cell clustures
+\hypertarget{testing-for-differential-expression}{%
+\paragraph{Testing for differential
+expression}\label{testing-for-differential-expression}}
 
-monocyte data acvailiabe
+In order to test for differential expression the package limma from R
+bioconductor was used. \citep{Phipson2016}. The package uses a linear
+model approach to define a fold expression between sample groups. Before
+starting the analysis two matrices need to be computed. The design
+matrix identifies samples according to the sample type and defines the
+experimental design. In order to create the matrix automatically a
+simple algorithm is used within the programmed backend logic. The
+expression matrix contains intensities for each identified protein will
+be obtained at the end of the pipeline. A function call with the
+arguments design and expression matrix calculates the contrast matrix.
+The next step is creating a linear model between groups by log-ratios of
+their expression values.
 
-tmt label --> noise reduction
+\pagenumbering{arabic}
 
-tools for clustering
+\newpage
 
-	plst, knm,
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\vspace{0.5cm}
+
+\hypertarget{materials-and-methods}{%
+\section{Materials and Methods}\label{materials-and-methods}}
+
+\hypertarget{materials}{%
+\subsection{Materials}\label{materials}}
+
+For analysis two types of cells were used. One type is the Jurkat-based
+cell line (J-lat) with integrated HIV.
+
+The other type of cells are macrophages with a sample size of 72 cells.
+The analysis is done with two groups. A HIV negative (HIV-) control
+group and a HIV positive (HIV+) group.
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+\hypertarget{acquisition}{%
+\subsubsection{Acquisition}\label{acquisition}}
+
+Analysis of the data was done with MaxQuant \citep{Cox2008}
+
+\newpage
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+Some more guidlines from the School of Geosciences.
+
+This section should summarise the findings of the research referring to
+all figures, tables and statistical results (some of which may be placed
+in appendices). - include the primary results, ordered logically - it is
+often useful to follow the same order as presented in the methods. -
+alternatively, you may find that ordering the results from the most
+important to the least important works better for your project. - data
+should only be presented in the main text once, either in tables or
+figures; if presented in figures, data can be tabulated in appendices
+and referred to at the appropriate point in the main text.
+
+\textbf{Often, it is recommended that you write the results section
+first, so that you can write the methods that are appropriate to
+describe the results presented. Then you can write the discussion next,
+then the introduction which includes the relevant literature for the
+scientific story that you are telling and finally the conclusions and
+abstract -- this approach is called writing backwards.}
+
+\newpage
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+the purpose of the discussion is to summarise your major findings and
+place them in the context of the current state of knowledge in the
+literature. When you discuss your own work and that of others, back up
+your statements with evidence and citations. - The first part of the
+discussion should contain a summary of your major findings (usually 2 --
+4 points) and a brief summary of the implications of your findings.
+Ideally, it should make reference to whether you found support for your
+hypotheses or answered your questions that were placed at the end of the
+introduction. - The following paragraphs will then usually describe each
+of these findings in greater detail, making reference to previous
+studies. - Often the discussion will include one or a few paragraphs
+describing the limitations of your study and the potential for future
+research. - Subheadings within the discussion can be useful for
+orienting the reader to the major themes that are addressed.
+
+\newpage
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+The conclusion section should specify the key findings of your study,
+explain their wider significance in the context of the research field
+and explain how you have filled the knowledge gap that you have
+identified in the introduction. This is your chance to present to your
+reader the major take-home messages of your dissertation research. It
+should be similar in content to the last sentence of your summary
+abstract. It should not be a repetition of the first paragraph of the
+discussion. They can be distinguished in their connection to broader
+issues. The first paragraph of the discussion will tend to focus on the
+direct scientific implications of your work (i.e.~basic science,
+fundamental knowledge) while the conclusion will tend to focus more on
+the implications of the results for society, conservation, etc.
+
+\newpage
+
+\bibliography{bibliography}
+
+\newpage
+
+\hypertarget{appendixces}{%
+\section{Appendix(ces)}\label{appendixces}}
+
+\hypertarget{appendix-a-additional-tables}{%
+\subsection{Appendix A: additional
+tables}\label{appendix-a-additional-tables}}
+
+Insert content for additional tables here.
+
+\newpage
+
+\hypertarget{appendix-b-additional-figures}{%
+\subsection{Appendix B: additional
+figures}\label{appendix-b-additional-figures}}
+
+Insert content for additional figures here.
+
+\newpage
+
+\hypertarget{appendix-c-code}{%
+\subsection{Appendix C: code}\label{appendix-c-code}}
+
+Insert code (if any) used during your dissertation work here.
+
+\end{document}

### 2023-03-08 markdown/introduction.Rmd
Statistical testing in thesis started

@@
 In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score.
-The peptides pectrum matching score is defined as -10log10(p).
+The peptides spectrum matching score is defined as -10log10(p).
 Whereas the p-value is defined that the hit is done by chance.
@@
 
-According to the experiment and sample groups, individual statistical methods should be used.
-
-follow sctranscr. analysis pipe
-
-signatures of cell clustures
-
-monocyte data acvailiabe
-
-tmt label --> noise reduction
-
-tools for clustering
-
-	plst, knm,
+In order to test for differential expression the package limma from R bioconductor was used.
+\citep{Phipson2016}.
+The package uses a linear model approach to define a fold expression between sample groups.
+Before starting the analysis two matrices need to be computed.
+The design matrix identifies samples according to the sample type and defines the experimental design.
+In order to create the matrix automatically a simple algorithm is used within the programmed backend logic.
+The expression matrix contains intensities for each identified protein will be obtained at the end of the pipeline.
+A function call with the arguments design and expression matrix calculates the contrast matrix.
+The next step is creating a linear model between groups by log-ratios of their expression values.

### 2023-03-09 markdown/introduction.Rmd
thesis changed (intro --> methods) statistical testing tried out paired t-test UMAP in interface included

--- 2023-02-15 markdown/introduction.Rmd
@@
 Mass to charge ratio (m/z)
-
-### Instrumentation
-
-#### Cell Isolation
-
-#### Lysis
-
-#### Digestion
-
-#### Liquid chromatography
-
-In order to separate proteins according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
-
-#### Ionization
-
-In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase.
-Two techniques are capable of this procedure.
-Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol.
-Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer.
-The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
-Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
-In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
-The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
-In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
-
-Matrix-assisted laser desorption/ionization (MALDI)
-
-#### Coupled mass-spectrometry (MS/MS)
-
-In order to enhance sequence identification, two MS devices are built in series.
-In the first run (MS1) the m/z is determined and the molecules are passed to the next device.
-Upon passing the molecules are fragmented into smaller ions and analyzed by the second MS.
-The fragmentation highly depends on the chemical bonds found in the molecule.
-The majority of these breaks occur on the peptide bond of the protein, although this is not guaranteed for all bonds and so it can happen that certain peptide ions have a low abundance \citep{Budnik2018}.
-These low abundant peptides will not be detected, hence the problem needs to be faced with another approach .
-A solution for this problem is molecular barcoding with labeling mentioned in the chapter labeling.
-
-#### Labeling techniques
-
-For differential analysis proteins need to be labeled to compare mass to charge intensities in order to quantify observed peptides.
-Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
-Physicochemical properties of the proteins can change the ionization efficiency and detectability of the target.
-However, when comparing the same analyte between multiple runs of labeled proteins, differences in the mass spectrum reflect the abundance of those.
-Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the protein.
-
-#### Metabolic labeling
-
-Feeding cells with aminoacids containing heavy isotopes, is the method of choice in order to label peptides at the earliest possible level.
-This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants.
-Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
-Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
-This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
-
-##### Isobaric labeling
-
-###### Tandem mass tag (TMT)
-
-Tandem mass tag (TMT) reagents enable to differentiate multiple samples analyzing in one MS run. 
-The samples are labeled individually and pooled afterwards, this procedure is called multiplexing.
-TMTs have the same charge and differ only by their isotopic masses, the peaks found for each sample are called reporter ions (RI).
-Each RI and sample is interpreted as one channel in downstream analysis.
-The identification of these RI leads to an enrichment and identification of low abundance peptide ions which is common especially in single-cell techniques.
-With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
-The disadvantage of isobaric labeling is, that the co-fragmentation signals can be observed in the spectrogram and the data needs to be normalized in order to remove unwanted contribution \citep{Marx2019, Budnik2018}.
-Furthermore TMTs have an isotopic distribution according to the distribution found in nature.
-This can be corrected during data-acquisition as a defined spread in other channels.
-
-### MS-Spectrum
-
-Each peptide is reflected by its` indivual fingerprint in the ms-spectrum.
-The fingerprint is based on the chemical properties and modifications of aminoacids.
-These aminoacids can be calculated through their m/z ration and after that interpreted as an aminoacid sequence.
-Due to fragmentation of the protein only peptide sequences are visible in the spectrum.
-In order to identify proteins, peptides are matched against a sequence database \citep{Cox2008}.
-Sequence Databases are simple .fasta files, which can be downloaded on the uniprot webpage (www.uniprot.org).
-
-### Interpretation of the data
-
-#### Data acquisition
-
-Since ms data has a high resolution, algorithms are used to convert the raw signal to an interpretable form. 
-MaxQuant  is one of many sofware packages to process the data and provides it for further analysis and statistical testing.
-Other software solutions are Protein Discoverer Thermo Fisher or even packages for R.
-In this publication we will mainly focus on the data-acquisition with MaxQuant \citep{Cox2008}.
-
-##### Three-dimensional peak detection
-
-The three dimensions of the data are: m/z ratio, intensity and retention time.
-The algorithm finds local minima of the function in order to seperate peaks from each other.
-The centroid of the peak is detected by fitting a so called gaussian peak shape fitting.
-This can be interpreted as finding the peaks of each m/z spectrum as a function of time..
-The centroid of the peak refers to an isotope.
-
-##### Deisotoping
-
-To decrypt the istopic distribution of a biomolecule, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
-Isotoping is the term of such procedure and it is enabled with graph theory.
-After this procedure the amount of data points are reduced by a tenfold and a single peak reflects a small biomolecule.
-
-##### Label detection
-
-The next step in data-acquisition is the detection of labels for quantification.
-Isotopic pairs of the label (e.g.
-N13, N14, N15) contained in the tag or aminoacid can be identified by convoluting the two measured isotope patterns with the theoretical isotope patterns.
-With a least-square method the best fit is found iteratively and the channel/sample can be identified.
-
-##### Improving peptide mass accuracy
-
-The intensity-weighted average of the ms peak centroids (as described in the 3D peak identification) refers to the mass of the peptide.
-Corrections highly depend on the analyzer, MaxQuant uses for an orbitrap typed analyzer a correction value of 1ppm.
-Autocorrelation between centroids is compensated by only using well-identified peptides.
-As published the mass precision within a ms experiment ranges around 10e-7.
-
-##### Peptide search
-
-Biomolecules can now be searched in a database in forward and reverse direction.
-The peptide identification (P-) score indicates the fit of the data to the found sequence in the database according to the length of the peptide and is used to calculate the posterior error probability (FDR).
-The calculation of the false discovery rate is then calculated by taking FDR into contrast.
-
-##### Protein assembly
-
-After these calculation the identified peptides can be aggregated according to it`s respective protein and  quantified.
-The mentioned metrics indicating the performance of the peptide search can be used in downstream analysis.
-A so called razor peptide indicates the group with the highest number of identified peptides.
-Quantification is enabled by taking only unique peptides into contrast.
-Posterior error probabilities, which refers to the chance that the found peptide is a random event, are multiplied and only distinct sequences with the highst-scoring are used.
-
-#### Downstream Analysis
-
-Further analysis is done with R and respective packages such as bioconductor.
-
-##### Reading the data
-
-##### Principle component analysis (PCA)
-
-Protein levels can be projected to their principle components (PC) and clustered to their specific cell type. So cell types are distinguished by their proteinaceous composition.
-
-#### Testing for differential expression
-
-According to the experiment and sample groups, individual statistical methods should be used.
-
-follow sctranscr. analysis pipe
-
-signatures of cell clustures
-
-monocyte data acvailiabe
-
-tmt label --> noise reduction
-
-tools for clustering
-
-	plst, knm,

### 2023-03-09 markdown/methods.Rmd
thesis changed (intro --> methods) statistical testing tried out paired t-test UMAP in interface included

@@
 ---
-
-# Methods
-
-\vspace{0.5cm}
 
@@
 
-Analysis of the data was done with MaxQuant \citep{Cox2008}
+Acquisition of the data was done with MaxQuant \citep{Cox2008} software package.
+
+### Instrumentation
+
+#### Cell Isolation
+
+#### Lysis
+
+#### Digestion
+
+#### Liquid chromatography
+
+In order to separate proteins according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
+
+#### Ionization
+
+In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase.
+Two techniques are capable of this procedure.
+Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol.
+Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer.
+The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
+Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
+In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
+The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
+In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
+
+Matrix-assisted laser desorption/ionization (MALDI)
+
+#### Coupled mass-spectrometry (MS/MS)
+
+In order to enhance sequence identification, two MS devices are built in series.
+In the first run (MS1) the m/z is determined and the molecules are passed to the next device.
+Upon passing the molecules are fragmented into smaller ions and analyzed by the second MS.
+The fragmentation highly depends on the chemical bonds found in the molecule.
+The majority of these breaks occur on the peptide bond of the protein, although this is not guaranteed for all bonds and so it can happen that certain peptide ions have a low abundance \citep{Budnik2018}.
+These low abundant peptides will not be detected, hence the problem needs to be faced with another approach .
+A solution for this problem is molecular barcoding with labeling mentioned in the chapter labeling.
+
+#### Labeling techniques
+
+For differential analysis proteins need to be labeled to compare mass to charge intensities in order to quantify observed peptides.
+Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
+Physicochemical properties of the proteins can change the ionization efficiency and detectability of the target.
+However, when comparing the same analyte between multiple runs of labeled proteins, differences in the mass spectrum reflect the abundance of those.
+Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the protein.
+
+#### Metabolic labeling
+
+Feeding cells with aminoacids containing heavy isotopes, is the method of choice in order to label peptides at the earliest possible level.
+This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants.
+Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
+Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
+This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
+
+##### Isobaric labeling
+
+###### Tandem mass tag (TMT)
+
+Tandem mass tag (TMT) reagents enable to differentiate multiple samples analyzing in one MS run. 
+The samples are labeled individually and pooled afterwards, this procedure is called multiplexing.
+TMTs have the same charge and differ only by their isotopic masses, the peaks found for each sample are called reporter ions (RI).
+Each RI and sample is interpreted as one channel in downstream analysis.
+The identification of these RI leads to an enrichment and identification of low abundance peptide ions which is common especially in single-cell techniques.
+With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
+The disadvantage of isobaric labeling is, that the co-fragmentation signals can be observed in the spectrogram and the data needs to be normalized in order to remove unwanted contribution \citep{Marx2019, Budnik2018}.
+Furthermore TMTs have an isotopic distribution according to the distribution found in nature.
+This can be corrected during data-acquisition as a defined spread in other channels.
+
+### MS-Spectrum
+
+Each peptide is reflected by its` indivual fingerprint in the ms-spectrum.
+The fingerprint is based on the chemical properties and modifications of aminoacids.
+These aminoacids can be calculated through their m/z ration and after that interpreted as an aminoacid sequence.
+Due to fragmentation of the protein only peptide sequences are visible in the spectrum.
+In order to identify proteins, peptides are matched against a sequence database \citep{Cox2008}.
+Sequence Databases are simple .fasta files, which can be downloaded on the uniprot webpage (www.uniprot.org).
+
+### Interpretation of the data
+
+#### Data acquisition
+
+Since ms data has a high resolution, algorithms are used to convert the raw signal to an interpretable form. 
+MaxQuant  is one of many sofware packages to process the data and provides it for further analysis and statistical testing.
+Other software solutions are Protein Discoverer Thermo Fisher or even packages for R.
+In this publication we will mainly focus on the data-acquisition with MaxQuant \citep{Cox2008}.
+
+##### Three-dimensional peak detection
+
+The three dimensions of the data are: m/z ratio, intensity and retention time.
+The algorithm finds local minima of the function in order to seperate peaks from each other.
+The centroid of the peak is detected by fitting a so called gaussian peak shape fitting.
+This can be interpreted as finding the peaks of each m/z spectrum as a function of time..
+The centroid of the peak refers to an isotope.
+
+##### Deisotoping
+
+To decrypt the istopic distribution of a biomolecule, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
+Isotoping is the term of such procedure and it is enabled with graph theory.
+After this procedure the amount of data points are reduced by a tenfold and a single peak reflects a small biomolecule.
+
+##### Label detection
+
+The next step in data-acquisition is the detection of labels for quantification.
+Isotopic pairs of the label (e.g.
+N13, N14, N15) contained in the tag or aminoacid can be identified by convoluting the two measured isotope patterns with the theoretical isotope patterns.
+With a least-square method the best fit is found iteratively and the channel/sample can be identified.
+
+##### Improving peptide mass accuracy
+
+The intensity-weighted average of the ms peak centroids (as described in the 3D peak identification) refers to the mass of the peptide.
+Corrections highly depend on the analyzer, MaxQuant uses for an orbitrap typed analyzer a correction value of 1ppm.
+Autocorrelation between centroids is compensated by only using well-identified peptides.
+As published the mass precision within a ms experiment ranges around 10e-7.
+
+##### Peptide search
+
+Biomolecules can now be searched in a database in forward and reverse direction.
+The peptide identification (P-) score indicates the fit of the data to the found sequence in the database according to the length of the peptide and is used to calculate the posterior error probability (FDR).
+The calculation of the false discovery rate is then calculated by taking FDR into contrast.
+
+##### Protein assembly
+
+After these calculation the identified peptides can be aggregated according to it`s respective protein and  quantified.
+The mentioned metrics indicating the performance of the peptide search can be used in downstream analysis.
+A so called razor peptide indicates the group with the highest number of identified peptides.
+Quantification is enabled by taking only unique peptides into contrast.
+Posterior error probabilities, which refers to the chance that the found peptide is a random event, are multiplied and only distinct sequences with the highst-scoring are used.
+
+#### Data processing
+
+Further analysis can be done with R and respective packages such as bioconductor. Since there is no state of the art established, analysis varies upon experimental design.
+
+##### Reading the data
+
+After processing MaxQuant creates a directory containing all results as .txt file.
+The evidence.txt file include all peptide sequence matches (PSM) with their respective proteins and statistical parameters.
+
+Example fo basic parameters and derivations include:
+
+* Peptide sequence
+* Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)
+	+ Mass
+* Retention time
+* Precursor Ion Fragment
+	+ source of the detected ion also referred as mother ion
+* Fraction of total spectrum 
+* Base peak fraction
+* Reporter intensity (RI)
+	+ corrected RI
+* Posterior error probability (PEP)
+
+#### Object oriented programming
+
+In order to streamline the analysis of multiple experiments, object oriented programming can be applied.
+The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
+The structure enables sub setting for further analysis.
+
+#### Filtering
+
+Peptides having a low abundance will be set to zero by the analysis.
+The value zero would imply that the sample does not contain the respective peptide, therefore replacing 0 with not applicable (NA) is crucial for understanding and interpretation of MS-data.
+
+Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.
+
+The author of the SCoPE2 \citep{Petelski2021} pipeline recommends calculating the sample to carrier ration (SCR).
+This can be done by dviding the ion intensity of the sample by the reporter ion intensity of the carrier channel.
+However in our experiment no carrier and reference channel is used.
+
+The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
+When applying multiple statistical testing (e.g.
+t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed.
+Corrections in statistics are an approach to compensate for the multiplicity of testing.
+There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR.
+In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score.
+The peptides spectrum matching score is defined as -10log10(p).
+Whereas the p-value is defined that the hit is done by chance.
+The calculation of the score is highly dependent on the data acquisition method used.
+MaxQuant uses Andromeda , an integrated search engine.
+Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora.
+As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage.
+However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.
+
+Further filter steps can include correction between multiple runs.
+This kind of process need the addition of a reference channel.
+However when observing a single run, these steps are not crucial for the upcoming analysis.
+
+#### Downstream Analysis
+
+##### Principle component analysis (PCA)
+
+Protein levels can be projected to their principle components (PC) and clustered to their specific cell type. So cell types are distinguished by their proteinaceous composition.
+
+#### Testing for differential expression
+
+In order to test for differential expression the package limma from R bioconductor was used.
+\citep{Phipson2016}.
+The package uses a linear model approach to define a fold expression between sample groups.
+Before starting the analysis two matrices need to be computed.
+The design matrix identifies samples according to the sample type and defines the experimental design.
+In order to create the matrix automatically a simple algorithm is used within the programmed backend logic.
+The expression matrix contains intensities for each identified protein will be obtained at the end of the pipeline.
+A function call with the arguments design and expression matrix calculates the contrast matrix.
+The next step is creating a linear model between groups by log-ratios of their expression values.

### 2023-03-09 scope2.R

@@ -26,17 +26,20 @@ quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData), value = T)
 
 # create sample file
 # this needs to be done by the researcher
-sampleAnnotation <- as.data.frame(quantCols)
-colnames(sampleAnnotation) <-c("Channel")
-sampleAnnotation$Raw.file = unique(mqScpData$Raw.file)
+# sampleAnnotation <- as.data.frame(quantCols)
+# colnames(sampleAnnotation) <-c("Channel")
+# sampleAnnotation$Raw.file = unique(mqScpData$Raw.file)
+# 
+# 
+# samplesA <- c(replicate(6, "Monocytes_A"))
+# samplesB <- c(replicate(6, "Monocytes_B"))
+# samples <- c(samplesA, samplesB)
+# 
+# 
+# sampleAnnotation$SampleType <- samples
 
+sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/mq-run_150223/combined/txt/sampleAnnotation.txt")
 
-samplesA <- c(replicate(6, "Monocytes_A"))
-samplesB <- c(replicate(6, "Monocytes_B"))
-samples <- c(samplesA, samplesB)
-
-
-sampleAnnotation$SampleType <- samples
 
 # create QFeature object
 scp <- readSCP(featureData = mqScpData,
@@ -388,21 +391,20 @@ num_var <- sapply(factor_var, function(x) {
 # Replace original variable with numeric version
 factor_var <- num_var
 
+
 # Create a design matrix
-design <- model.matrix(~ 0+factor(factor_var))
+design <- model.matrix(~ 0+factor(factor_var)+
+                         factor(c("C", "C", "C", "C", "C", "C", "T", "T", "T", "T", "T", "T")))
 # assign the column names
-colnames(design) <- unique(scp$SampleType)
+colnames(design) <- paste(c(as.character(unique(scp$SampleType)), "Treatment"))
+# 
+# contrast = "Monocytes_2 - Monocytes_1,
+# Monocytes_3 - Monocytes_2,
+# Monocytes_4 -Monocytes_5,
+# Monocytes_5 - Monocytes_6"
 
-user_input1 <- "Monocytes_A"
-user_input2 <- "Monocytes_B"
-
-col_index1 <- which(colnames(design) == user_input1)
-col_index2 <- which(colnames(design) == user_input2)
-
[…]

### 2023-03-09 scpR-app/app.R

@@ -52,7 +52,7 @@ ui <- fluidPage(
                   tabPanel("Summary Barplot", plotOutput("summary_bar")),
                   tabPanel("Reporter Ion Intensity", plotOutput("RI_intensity")),
                   tabPanel("Covariance across razor peptides", plotOutput("CV_median")),
-                  tabPanel("Principle Component Analysis", plotOutput("PCA")),
+                  tabPanel("Dimensionality reduction", plotOutput("PCA"), plotOutput("UMAP")),
                   tabPanel("Feature wise output", selectInput("selectedProtein", "Choose protein for observation", 
                                                               choices=c(rowData(scp)[["proteins"]][,1])),
                            plotOutput("feature_subset")),
@@ -217,13 +217,24 @@ server <- function(input, output) {
       
       
       
-      incProgress(16/17, detail=paste("running PCA"))
+      incProgress(16/17, detail=paste("running dimensionality reduction"))
+     
       scp_0[["proteins_dim_red"]] <- runPCA(scp_0[["proteins_dim_red"]],
                                        ncomponents = 5,
                                        ntop = Inf,
                                        scale = TRUE,
                                        exprs_values = 1,
                                        name = "PCA")
+      
+      scp_0[["proteins_dim_red"]] <- runUMAP(scp_0[["proteins_dim_red"]],
+                                         ncomponents = 2,
+                                         ntop = Inf,
+                                         scale = TRUE,
+                                         exprs_values = 1,
+                                         n_neighbors = 3,
+                                         dimred = "PCA",
+                                         name = "UMAP")
+      
     
       
       incProgress(17/17, detail=paste("analysis finish"))
@@ -393,7 +404,14 @@ server <- function(input, output) {
                      dimred = "PCA",
                      colour_by = "SampleType",
                      point_alpha = 1)
-      
+    })
+    
+    output$UMAP <- renderPlot({
+      scp_0 <- scp()
+      plotReducedDim(scp_0[["proteins_dim_red"]],
+                     dimred = "UMAP",
+                     colour_by = "SampleType",
+                     point_alpha = 1)
     })
     
     output$feature_subset <- renderPlot({

### 2023-04-12 scope2.R
commit before major changes
to do:
refactor exp_matrix_0 <- exp_matrix() in Statistic module
debug if pipeline especially norm module is calculating correctly
try to refactor scp object --> if no normalization or transformation is applied --> overview should not show the changes otherwise missleading
remove debug line in statistic module
MASS package in shiny not working --> fix

@@ -156,8 +156,6 @@ scp <- aggregateFeaturesOverAssays(scp,
 #                  name = "Peptides")
 
 # filter by median intensity
-# calculation of median not applicable if only one run observed
-
 file_name <- sampleAnnotation$Raw.file[1]
 peptides <- paste("peptides_", as.character(file_name), sep = "")
 
@@ -359,7 +357,7 @@ index_B <- st_indeces[choice_B]
 MAplot(assay(scp[["proteins_transf"]][,index_A[[1]]]), assay(scp[["proteins_transf"]][,index_B[[1]]]))
 
 protein_matrix <- assay(scp[["proteins_transf"]])
-protein_matrix <- CONSTANd(protein_matrix)
+#protein_matrix <- CONSTANd(protein_matrix)
 
 sce <- getWithColData(scp, "proteins")
 
@@ -371,7 +369,8 @@ scp <- addAssayLinkOneToOne(scp,
                             from = "proteins",
                             to = "proteins_norm")
 
-assay(scp[["proteins_norm"]]) <- protein_matrix$normalized_data
+assay(scp[["proteins_norm"]]) <- assay(scp[["proteins_transf"]])
+#assay(scp[["proteins_norm"]]) <- protein_matrix$normalized_data
 
 # missing value imputation
 
@@ -508,79 +507,84 @@ hist(exp_matrix[[sample_types[1]]])
 # create design matrix 
 # return the count of individual colnames (Types of experiment)
 # Sample data
-factor_var <- factor(colnames(exp_matrix))
+# factor_var <- factor(colnames(exp_matrix))
+# 
+# factorize_var <- function(i_vector) {
+#   fact_vect <- factor(i_vector)
+#   levels <- unique(fact_vect)
+#   num_values <- seq_along(levels)
+#   lookup_table <- data.frame(fact_vect = levels, num_var = num_values)
+#   
+#   num_var <- sapply(fact_vect, function(x) {
+#     lookup_table$num_var[lookup_table$fact_vect == x]
+#   })
+#   
+#   fact_vect <- factor(num_var)
+#   return(fact_vect)
+# }
+# 
+# factors_sample_type <- factorize_var(colnames(exp_matrix))
+# 
+# user_input <- c(1,2,3,4,5,6,1,2,3,4,5,6)
+# user_chosen_name <- "patient"
+# 
+# factors_patient <- factorize_var(user_input)
+# 
[…]

### 2023-04-12 scpR-app/app.R

@@ -116,8 +116,11 @@ server <- function(input, output, session) {
       incProgress(2/17, detail = paste("replacing zeros with NA"))
       scp_0 <- zeroIsNA(scp_0, 1:length(rowDataNames(scp_0)))
       
+      # filter PSM
+      # filter out potential contaminants
+      # filter out matches to decoy database
+      # keep PSMs with high PIF (parental ion fraction)
       req(input$PIF_cutoff)
-      # apply first filter
       incProgress(3/17, detail = paste("filter contaminants and PIF"))
       scp_0 <- filterFeatures(scp_0,
                               ~ Reverse != "+" &
@@ -159,7 +162,6 @@ server <- function(input, output, session) {
       scp_0$MedianRI <- medians
       
       # Filter based on the median CV -> remove covariant peptides over multiple proteins
-      
       req(input$nObs_pep_razrpr)
       incProgress(9/17, detail=paste("calculate covariance per cell"))
       scp_0 <- medianCVperCell(scp_0,
@@ -177,13 +179,13 @@ server <- function(input, output, session) {
       incProgress(12/17, detail=paste("remove peptides with high missing rate"))
       req(input$pNA)
       scp_0 <- filterNA(scp_0,
-                        i = peptides, #_norm",
+                        i = peptides,
                         pNA = input$pNA)
       
       
       incProgress(13/17, detail=paste("aggregate peptides to proteins"))
       scp_0 <- aggregateFeatures(scp_0,
-                                 i = peptides, #_log",
+                                 i = peptides,
                                  name = "proteins",
                                  fcol = "Leading.razor.protein",
                                  fun = matrixStats::colMedians, na.rm = TRUE)
@@ -435,6 +437,7 @@ server <- function(input, output, session) {
         req(input$selectedComp_stat)
         scp_0 <- scp_0[, scp_0$SampleType %in%  input$selectedComp_stat]
         exp_matrix_0 <- assay(scp_0[["proteins_dim_red"]])
+
         design <- model.matrix(~0+factor(scp_0$SampleType))
         colnames(design) <- unique(scp_0$SampleType)
         
@@ -451,9 +454,10 @@ server <- function(input, output, session) {
 
         scp_0 <- scp_0[, scp_0$SampleType %in%  input$selectedComp_stat]
         exp_matrix_0 <- assay(scp_0[["proteins_dim_red"]])
+        
         fetched_factor <- colData(scp_0)[input$col_factors]
[…]

### 2023-04-12 scpR-app/app.R
refactor exp_matrix_0 <- exp_matrix() in Statistic module --> done

@@ -120,12 +120,14 @@ server <- function(input, output, session) {
       # filter out potential contaminants
       # filter out matches to decoy database
       # keep PSMs with high PIF (parental ion fraction)
-      req(input$PIF_cutoff)
+
       incProgress(3/17, detail = paste("filter contaminants and PIF"))
+      req(input$PIF_cutoff)
+      PIF_cutoff <- input$PIF_cutoff
       scp_0 <- filterFeatures(scp_0,
                               ~ Reverse != "+" &
                                 Potential.contaminant != "+" &
-                                !is.na(PIF) & PIF > input$PIF_cutoff)
+                                !is.na(PIF) & PIF > PIF_cutoff)
       
       # compute qvalues_PSMs to filter out by FDR
       incProgress(4/17, detail=paste("calculate q-value for PSMs"))
@@ -141,9 +143,11 @@ server <- function(input, output, session) {
                           groupBy = "Leading.razor.protein",
                           rowDataName = "qvalue_proteins")
       
-      req(input$qvalue_cutoff)
+      
       incProgress(6/17, detail=paste("filter according to q-value"))
-      scp_0 <- filterFeatures(scp_0, ~ qvalue_proteins < input$qvalue_cutoff)
+      req(input$qvalue_cutoff)
+      qvalue_cutoff <- input$qvalue_cutoff
+      scp_0 <- filterFeatures(scp_0, ~ qvalue_proteins < qvalue_cutoff)
       
       # aggregate PSMS to peptides
       incProgress(7/17, detail=paste("aggregating features"))
@@ -162,8 +166,8 @@ server <- function(input, output, session) {
       scp_0$MedianRI <- medians
       
       # Filter based on the median CV -> remove covariant peptides over multiple proteins
-      req(input$nObs_pep_razrpr)
       incProgress(9/17, detail=paste("calculate covariance per cell"))
+      req(input$nObs_pep_razrpr)
       scp_0 <- medianCVperCell(scp_0,
                                i = length(rowDataNames(scp_0)),
                                groupBy = "Leading.razor.protein",
@@ -412,7 +416,7 @@ server <- function(input, output, session) {
   # expression matrix is used by plot functions
   exp_matrix <- reactive({
     scp_0 <- scp()
-    exp_matrix_0 <- data.frame(assay(scp_0[["proteins_dim_red"]]))
[…]

### 2023-04-13 scpR-app/app.R
coef implemented
fix volcano

@@ -64,7 +64,6 @@ ui <- fluidPage(
                   tabPanel("Statistics",
                            actionButton("run_statistics", "Press to run statistics"),
                            actionButton("qqplot", "check dependencies for linear model"),
-                           textOutput("model_vars"),
                            selectInput("model_design", "Choose your study design",
                                        choices = c("All pairwise comparison",
                                                    "Differential Expression with defined Contrasts",
@@ -72,6 +71,7 @@ ui <- fluidPage(
                            uiOutput("sample_select"),
                            uiOutput("add_factor"),
                            plotOutput("venn_diagram"),
+                           selectInput("chosen_coef", "Select your coefficient of interest", ""),
                            plotOutput("volcano", hover = hoverOpts(id ="plot_hover")),
                            verbatimTextOutput("hover_info"),
                            tableOutput("protein_table")
@@ -120,10 +120,9 @@ server <- function(input, output, session) {
       # filter out potential contaminants
       # filter out matches to decoy database
       # keep PSMs with high PIF (parental ion fraction)
-
       incProgress(3/17, detail = paste("filter contaminants and PIF"))
       req(input$PIF_cutoff)
-      PIF_cutoff <- input$PIF_cutoff
+      PIF_cutoff<-input$PIF_cutoff
       scp_0 <- filterFeatures(scp_0,
                               ~ Reverse != "+" &
                                 Potential.contaminant != "+" &
@@ -143,8 +142,7 @@ server <- function(input, output, session) {
                           groupBy = "Leading.razor.protein",
                           rowDataName = "qvalue_proteins")
       
-      
-      incProgress(6/17, detail=paste("filter according to q-value"))
+      incProgress(6/17, detail=paste("filter according to q-value"))      
       req(input$qvalue_cutoff)
       qvalue_cutoff <- input$qvalue_cutoff
       scp_0 <- filterFeatures(scp_0, ~ qvalue_proteins < qvalue_cutoff)
@@ -168,23 +166,26 @@ server <- function(input, output, session) {
       # Filter based on the median CV -> remove covariant peptides over multiple proteins
[…]

### 2023-04-13 scpR-app/app.R
coef implemented volcano fixed

@@ -804,17 +804,28 @@ server <- function(input, output, session) {
   ## Results 
   # volcanoplot in statistic tab
   output$volcano <- renderPlot({
+    req(input$chosen_coef)
+    req(stat_result())
     volcanoplot(stat_result(), cex = 0.5, coef = input$chosen_coef) 
+  })
+
+  # reactive element for toptable
+  protein_table <- reactive({
+    req(stat_result())
+    req(input$chosen_coef)
+    data.frame(topTable(stat_result(), number = Inf, adjust = "BH", coef = input$chosen_coef))
   })
   
   # reactive element for hover function of volcano plot
   displayed_text <- reactive({
     req(input$plot_hover)
     hover <- input$plot_hover
-    dist <- sqrt((hover$x - stat_result()$coef)^2 + (hover$y - -log10(stat_result()$p.value))^2)
+    req(protein_table())
+    
+    dist <- sqrt((hover$x - protein_table()$logFC)^2 + (hover$y - -log10(protein_table()$P.Value))^2)
     
     if(min(dist) < 0.3) {
-      rownames(stat_result())[which.min(dist)]
+      rownames(protein_table())[which.min(dist)]
     } else {
       NULL
     }
@@ -830,7 +841,8 @@ server <- function(input, output, session) {
   
   # show significant proteins in the table
   output$protein_table <- renderTable({
-    topTable(stat_result(), number = 10, adjust = "BH", coef = input$chosen_coef)
+    req(protein_table())
+    protein_table()
   }, rownames = T)
   
   # venn diagram for significant proteins

### 2023-04-13 scope2.R
box cox fixed

@@ -238,16 +238,69 @@ scp <- aggregateFeatures(scp,
 #                     name = "proteins_transf")
 
 # box cox transf
+boxcox_1 <- function(object, ...) UseMethod("boxcox_1")
 
+boxcox_1.formula <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- object$model
+    m <- length(lambda)
+    object <- stats::lm(object, y = TRUE, qr = TRUE)
+    result <- NextMethod()
+    result
+  }
+
+boxcox_1.lm <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- object$model
+    m <- length(lambda)
+    if(is.null(object$y) || is.null(object$qr))
+      object <- update(object, y = TRUE, qr = TRUE)
+    result <- NextMethod()
+    result
+  }
+
+boxcox_1.default <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- as.matrix(object$model)
+    if(is.null(y <- object$y) || is.null(xqr <- object$qr))
+      stop(gettextf("%s does not have both 'qr' and 'y' components",
+                    sQuote(deparse(substitute(object)))), domain = NA)
+    if(any(y <= 0))
+      stop("response variable must be positive")
+    n <- length(y)
+    ## scale y[]  {for accuracy in  y^la - 1 }:
+    y <- y / exp(mean(log(y)))
+    logy <- log(y) # now  ydot = exp(mean(log(y))) == 1
+    xl <- loglik <- as.vector(lambda)
+    m <- length(xl)
+    for(i in 1L:m) {
+      if(abs(la <- xl[i]) > eps)
+        yt <- (y^la - 1)/la
+      else yt <- logy * (1 + (la * logy)/2 *
+                           (1 + (la * logy)/3 * (1 + (la * logy)/4)))
+      loglik[i] <- - n/2 * log(sum(qr.resid(xqr, yt)^2))
+    }
+    list(x = xl, y = loglik)
+  }
 
 
 protein_matrix <- assay(scp[["proteins"]])
 
-b <- boxcox(lm(protein_matrix ~ 1), plotit=F)
+prot_lm <- lm(protein_matrix ~ 1)
+
+b <- boxcox_1(prot_lm)
+
+b
+
+
+
 
 # Exact lambda
[…]

### 2023-04-13 scpR-app/app.R

@@ -7,7 +7,7 @@ library("reshape2")
 library("scater")
 library("limma")
 library("CONSTANd")
-library("MASS")
+library("stats")
 
 reactlog::reactlog_enable()
 
@@ -80,6 +80,55 @@ ui <- fluidPage(
     )
   )
 )
+
+# box cox transf
+boxcox_1 <- function(object, ...) UseMethod("boxcox_1")
+
+boxcox_1.formula <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- object$model
+    m <- length(lambda)
+    object <- stats::lm(object, y = TRUE, qr = TRUE)
+    result <- NextMethod()
+    result
+  }
+
+boxcox_1.lm <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- object$model
+    m <- length(lambda)
+    if(is.null(object$y) || is.null(object$qr))
+      object <- update(object, y = TRUE, qr = TRUE)
+    result <- NextMethod()
+    result
+  }
+
+boxcox_1.default <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- as.matrix(object$model)
+    if(is.null(y <- object$y) || is.null(xqr <- object$qr))
+      stop(gettextf("%s does not have both 'qr' and 'y' components",
+                    sQuote(deparse(substitute(object)))), domain = NA)
+    if(any(y <= 0))
+      stop("response variable must be positive")
+    n <- length(y)
+    ## scale y[]  {for accuracy in  y^la - 1 }:
+    y <- y / exp(mean(log(y)))
+    logy <- log(y) # now  ydot = exp(mean(log(y))) == 1
+    xl <- loglik <- as.vector(lambda)
+    m <- length(xl)
+    for(i in 1L:m) {
+      if(abs(la <- xl[i]) > eps)
+        yt <- (y^la - 1)/la
+      else yt <- logy * (1 + (la * logy)/2 *
+                           (1 + (la * logy)/3 * (1 + (la * logy)/4)))
+      loglik[i] <- - n/2 * log(sum(qr.resid(xqr, yt)^2))
+    }
+    list(x = xl, y = loglik)
+  }
 
 
 # Define server logic required to draw a histogram
@@ -227,11 +276,10 @@ server <- function(input, output, session) {
       else if (input$transform_base == "BoxCox") {
         
[…]

### 2023-04-13 scpR-app/app.R
minor bugfix
to do:
implement slider for table

@@ -296,6 +296,7 @@ server <- function(input, output, session) {
         }
         if (round(lambda, digits = 0) == 0 || lambda < 0.25 && lambda > - 0.25 ) {
           protein_matrix <- log10(protein_matrix)
+          transform_base_bc <- "log10"
           print("log10(protein_matrix)")
         }
         if (round(lambda, digits = 1) == 0.5 || lambda > 0.25 && lambda < 0.75) {
@@ -337,7 +338,7 @@ server <- function(input, output, session) {
       
       incProgress(15/17, detail=paste("normalizing proteins"))
       req(input$norm_method)
-      if (input$norm_method == "SCoPE2" && input$transform_base == "log2" | input$transform_base == "log10") {
+      if (input$norm_method == "SCoPE2" && input$transform_base == "log2" | input$transform_base == "log10" | !is.na(transform_base_bc == "log10")) {
         # center cols with median
         scp_0 <- sweep(scp_0, i = "proteins_transf",
                        MARGIN = 2,

### 2023-04-13 markdown/methods.Rmd
thesis: flowchart in methods started not knitting correctly --> fix to do

@@
 ---
-
 # Materials and Methods
@@
 
-## Data
+### Cell Isolation
 
-### Acquisition
+### Lysis
 
-Acquisition of the data was done with MaxQuant \citep{Cox2008} software package.
+### Digestion
 
-### Instrumentation
-
-#### Cell Isolation
-
-#### Lysis
-
-#### Digestion
-
-#### Liquid chromatography
-
-In order to separate proteins according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
-
-#### Ionization
-
-In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase.
-Two techniques are capable of this procedure.
-Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol.
-Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer.
-The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
-Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
-In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
-The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
-In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
-
-Matrix-assisted laser desorption/ionization (MALDI)
-
-#### Coupled mass-spectrometry (MS/MS)
-
-In order to enhance sequence identification, two MS devices are built in series.
-In the first run (MS1) the m/z is determined and the molecules are passed to the next device.
-Upon passing the molecules are fragmented into smaller ions and analyzed by the second MS.
-The fragmentation highly depends on the chemical bonds found in the molecule.
-The majority of these breaks occur on the peptide bond of the protein, although this is not guaranteed for all bonds and so it can happen that certain peptide ions have a low abundance \citep{Budnik2018}.
-These low abundant peptides will not be detected, hence the problem needs to be faced with another approach .
-A solution for this problem is molecular barcoding with labeling mentioned in the chapter labeling.
-
-#### Labeling techniques
+### Labeling techniques
 
@@
 
-##### Isobaric labeling
+#### Isobaric labeling
 
-###### Tandem mass tag (TMT)
+##### Tandem mass tag (TMT)
 
@@
 This can be corrected during data-acquisition as a defined spread in other channels.
+
+### Instrumentation
+
+#### Liquid chromatography
+
+In order to separate proteins according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
+
+#### Mass Spectrometry
+
+##### Ionization
+
+In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase.
+Two techniques are capable of this procedure.
+Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol.
+Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer.
+The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
+Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
+In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
+The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
+In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
+
+Matrix-assisted laser desorption/ionization (MALDI)
+
+##### MS.1
+
+##### Coupled mass-spectrometry (MS/MS) & MS.2
+
+In order to enhance sequence identification, two MS devices are built in series.
+In the first run (MS1) the m/z is determined and the molecules are passed to the next device.
+Upon passing the molecules are fragmented into smaller ions and analyzed by the second MS.
+The fragmentation highly depends on the chemical bonds found in the molecule.
+The majority of these breaks occur on the peptide bond of the protein, although this is not guaranteed for all bonds and so it can happen that certain peptide ions have a low abundance \citep{Budnik2018}.
+These low abundant peptides will not be detected, hence the problem needs to be faced with another approach .
+A solution for this problem is molecular barcoding with labeling mentioned in the chapter labeling.
+
+## Data
+```{r Dataflow,  echo=FALSE}
+library(DiagrammeR)
+library(webshot)
+
+DiagrammeR::grViz("digraph {
+	graph [layout  =  dot,
+	       rankdir  = TB]
+
+node [shape = rectangle]
+rawData [label = 'msData.raw']
+node [shape = oval]
+MQ [label = 'MaxQuant']
+node [shape = invhouse]
+MQ_fd [label = 'feature detection']
+node [shape = cylinder]
+MQ_ds [label = 'Database search']
+node [shape = invhouse]
+MQ_fdr [label = 'FDR calculation']
+MQ_quant [label = 'Protein Quantification']
+node [shape = rectangle]
+MQoutput [label = 'evidence.txt']
+sampleAnnotate [label = 'sample_annotation.txt']
+node [shape = oval]
+ProteomicsWorkbench [label = 'Proteomics Workbench']
+
+rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
+sampleAnnotate -> ProteomicsWorkbench
+}",
+height = 500)
+```
+
+### Acquisition
+
+Acquisition of the data was done with MaxQuant \citep{Cox2008} software package.
 

### 2023-04-13 raw_analysis.R

@@ -1,0 +1,242 @@
+
+library("scp")
+library("ggplot2")
+library("magrittr")
+library("dplyr")
+library("CONSTANd")
+library("MASS")
+
+
+# read in MS result table
+mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/mq-run_150223/combined/txt/evidence.txt")
+
+
+sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/mq-run_150223/combined/txt/sampleAnnotation_two_groups.txt")
+
+# create QFeature object
+scp <- readSCP(featureData = mqScpData,
+               colData = sampleAnnotation,
+               channelCol = "Channel",
+               batchCol = "Raw.file",
+               suffix = paste0("_TMT", 1:12),
+               removeEmptyCols = TRUE)
+
+# clean missing data
+scp <- zeroIsNA(scp, i=1:length(rowDataNames(scp)))
+
+# filter PSM
+# filter out potential contaminants
+# filter out matches to decoy database
+# keep PSMs with high PIF (parental ion fraction)
+scp <- filterFeatures(scp,
+                      ~ Reverse != "+" &
+                        Potential.contaminant != "+" &
+                        !is.na(PIF) & PIF > 0.10)
+
+
+# Plot number of peptide sequence matches
+nPSMs <- dims(scp)[1, ]
+#nPSMs <- dims(scp)[1, 1]
+#nPeps <- dims(scp)[1, 2]
+
+ggplot(data.frame(nPSMs)) +
+  aes(x = nPSMs) +
+  geom_histogram(binwidth = 50) +
+  geom_vline(xintercept = 500)
+
+# filter runs with lack of features
+keepAssay <- dims(scp)[1, ] > 150
+scp <- scp[, , keepAssay]
+
+
+# compute qvalues_PSMs to filter out by FDR
+
+scp <- pep2qvalue(scp,
+                  i = 1:length(rowDataNames(scp)),
+                  PEP = "PEP", # by reference the dart_PEP value is used
+                  rowDataName = "qvalue_PSMs")
+
+scp <- pep2qvalue(scp,
+                  i = 1:length(rowDataNames(scp)),
+                  PEP = "PEP",
+                  groupBy = "Leading.razor.protein",
+                  rowDataName = "qvalue_proteins")
+
+
+scp <- filterFeatures(scp, ~ qvalue_proteins < 0.01)
+
+
+# aggregate PSMs to peptides
[…]

### 2023-04-13 scpR-app/app.R

@@ -892,7 +892,7 @@ server <- function(input, output, session) {
   output$protein_table <- renderTable({
     req(protein_table())
     protein_table()
-  }, rownames = T)
+  }, rownames = T, striped = T)
   
   # venn diagram for significant proteins
   output$venn_diagram <- renderPlot({

### 2023-04-18 scope2.R
bug fixing of pipeline finished
MA-plot works now

@@ -420,10 +420,16 @@ assay(scp[["proteins_transf"]]) <- protein_matrix
 
 # make MA plot first
 MAplot <- function(x,y,use.order=FALSE, R=NULL, cex=1.6, showavg=TRUE) {
-  # make an MA plot of y vs. x that shows the rolling average,
-  x[is.na(x)] <- 0
-  y[is.na(y)] <- 0
+  # catch unequal size of matrices
+  if (dim(x)[2] != dim(y)[2]) {
+    if (dim(x)[2] > dim(y)[2]) {
+      x <- x[, 1:dim(y)[2]]
+    } else if (dim(y)[2] > dim(x)[2]) {
+      y <- y[, 1:dim(x)[2]]
+    }
+  }
   
+  # make an MA plot of y vs. x that shows the rolling average,
   M <- log2(y/x)
   xlab = 'A'
   if (!is.null(R)) {r <- R; xlab = "A (re-scaled)"} else r <- 1
@@ -461,7 +467,7 @@ choice_B <- user_choice_vector[[1]][2]
 index_A <- st_indeces[choice_A]
 index_B <- st_indeces[choice_B]
 
-#MAplot(assay(scp[["proteins_transf"]][,index_A[[1]]]), assay(scp[["proteins_transf"]][,index_B[[1]]]))
+MAplot(assay(scp[["proteins_transf"]][,index_A[[1]]]), assay(scp[["proteins_transf"]][,index_B[[1]]]))
 
 protein_matrix <- assay(scp[["proteins_transf"]])
 #protein_matrix <- CONSTANd(protein_matrix)
@@ -501,11 +507,13 @@ library(sva)
 
 if (length(peptide_file) > 1) {
   scp <- impute(scp,
-                  i = "proteins_norm",
-                  name = "proteins_imptd",
-                  method = "knn",
-                  k = 3, rowmax = 1, colmax= 1,
-                  maxp = Inf, rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
+                i = "proteins_norm",
+                name = "proteins_imptd",
+                method = "knn",
+                k = 3, rowmax = 1, colmax= 1,
+                maxp = Inf, rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
+} 
+  
   
   sce <- getWithColData(scp, "proteins_imptd")
   
@@ -782,5 +790,6 @@ results <- decideTests(fit)
[…]

### 2023-04-18 scpR-app/app.R

@@ -51,9 +51,9 @@ ui <- fluidPage(
       selectInput("transform_base", "Choose method for protein data transformation", choices = c("log2", "log10", "sqrt", "quadratic", "BoxCox", "None")),
       # normalization method 
       selectInput("norm_method", "Choose method for protein data normalization", choices = c("SCoPE2", "None", "CONSTANd")),
-      switchInput(inputId = "multiple_batches", onLabel = "Multiple batches", offLabel = "Single batch", value = F),
-      conditionalPanel(condition = "input.multiple_batches"#, 
-                       #switchInput(inputId = "missing_v_impt", onLabel = "Missing value imputation", offLabel = "Missing = NaN", value = F),
+      switchInput(inputId = "opts_multiple_batches", onLabel = "Advanced", offLabel = "Default", value = F, label="Options for multiple batches"),
+      conditionalPanel(condition = "input.opts_multiple_batches", 
+                       selectInput(inputId = "missing_v", "Choose method for missing value handling", choices=c("KNN", "drop rows", "replace with 0", "replace with mean", "replace with median"))
                        #switchInput(inputId = "batch_correction", onLabel = "Batch correction", offLabel = "No batch correction")
                        )
       
@@ -449,14 +449,65 @@ server <- function(input, output, session) {
       }
       
       
-      if (length(peptide_file) > 1 && input$multiple_batches) {
+      if (length(peptide_file) > 1) {
         incProgress(16/17, detail=paste("running missing value imputation"))
-        scp_0 <- impute(scp_0,
-                      i = "proteins_norm",
-                      name = "proteins_imptd",
-                      method = "knn",
-                      k = 3, rowmax = 1, colmax= 1,
-                      maxp = Inf, rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
+        if (input$missing_v == "KNN") {
+          scp_0 <- impute(scp_0,
+                          i = "proteins_norm",
+                          name = "proteins_imptd",
+                          method = "knn",
+                          k = 3, rowmax = 1, colmax= 1,
+                          maxp = Inf, rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
+        } else if (input$missing_v == "drop rows") {
[…]

### 2023-04-18 markdown/methods.Rmd
methods further ellobarted

@@
 Quantification is enabled by taking only unique peptides into contrast.
-Posterior error probabilities, which refers to the chance that the found peptide is a random event, are multiplied and only distinct sequences with the highst-scoring are used.
+Posterior error probabilities, which refers to the chance that the found peptide is a random event, are multiplied and only distinct sequences with the highest-scoring are used.
 
@@
 After processing MaxQuant creates a directory containing all results as .txt file.
-The evidence.txt file include all peptide sequence matches (PSM) with their respective proteins and statistical parameters.
+The evidence.txt file include all peptide to spectrum matches (PSM) with their respective proteins and statistical parameters.
 
@@
 
-#### Object oriented programming
+##### Object oriented programming
 
@@
 The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
-The structure enables sub setting for further analysis.
+The structure enables sub setting for further analysis \citep{Vanderaa2021}.
 
-#### Filtering
+##### Zero values
+Peptides with low abundance are often set to zero during analysis.
+However, assigning a value of zero may incorrectly suggest that the sample does not contain the respective peptide.
+Given that it is highly unlikely for a biological cell of a comparable type and function to not contain a particular protein, replacing the zero value with "not applicable" (NA) is crucial for understanding and interpreting MS data.
 
-Peptides having a low abundance will be set to zero by the analysis.
-The value zero would imply that the sample does not contain the respective peptide, therefore replacing 0 with not applicable (NA) is crucial for understanding and interpretation of MS-data.
-
+##### Exclude reverse matches/contaminants
 Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.
 
-The author of the SCoPE2 \citep{Petelski2021} pipeline recommends calculating the sample to carrier ration (SCR).
-This can be done by dviding the ion intensity of the sample by the reporter ion intensity of the carrier channel.
-However in our experiment no carrier and reference channel is used.
+##### Filter according to precursor ion fraction (PIF)
+During mass spectrometry, the ions detected in MS1 are further fragmented through collision during multiple MS runs.
+The resulting product ions are derived from precursor ions (also known as mother ions or parental ions).
+Contaminant peptides can co-migrate in this process and can be distinguished by the lower fraction of their respective precursor ions \citep{Tannous2020}.
+These peptides need to be filtered out during the analysis pipeline.
+A cutoff value, referenced in the SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface, but it can be adjusted according to the needs of the biologist.
 
+##### Filter by q-value
 The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
@@
 However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.
+
+##### Peptide spectrum match (PSM) aggregation to peptides 
+In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function.
+In the context of processing from peptides to spectrum matches, the desired column is the peptide sequence.
+To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.
+
+##### Join assays when observing multiple comparable batches at once
+Sample size is often a limiting factor in hypothesis testing.
+A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to an early end of analyses.
+To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis.
+
+##### Calculate reporter ion intensity (RI) and filter according to median RI
+Columns which do not meet the desired intensity can be filtered by a threshold set on the RI.
+The median RI can also be used to check if an entire channel has a lower detection level.
+This can be due two reasons.
+One is the expression level of the given proteins in a cell.
+Meaning, that the expression of the observed cell type is simply lower than the other type.
+Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes.
+
+##### Calculate and filter according to median coefficient of variation (CV) per cell/channel
+Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data.
+The coefficient of variation of a peptide is considered as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins).
+Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.
+
+##### Remove peptides with high missing rate
+Although missing value imputation can be performed during the analysis of multiple batches, peptides with missing detections across channels can be problematic for quantification.
+The proteomic composition of a biological sample is similar between replicates and even across groups.
+However, the threshold of missingness (described as a fraction of the row) can be set in the user interface and adjusted to enable different experimental designs.
+
+##### Aggregation of peptides to proteins 
+Similar to the already explained previous aggregation step the peptides will be further processed into their respective proteins after the quality control on the peptide level is performed.
 

### 2023-04-18 scope2.R

@@ -533,20 +533,6 @@ if (length(peptide_file) > 1) {
                                 from = "proteins_imptd",
                                 to = "proteins_dim_red")
   
-  
-} else {
-  sce <- getWithColData(scp, "proteins_norm")
-  
-  scp <- addAssay(scp,
-                    y = sce,
-                    name = "proteins_dim_red")
-  
-  scp <- addAssayLinkOneToOne(scp,
-                                from = "proteins_norm",
-                                to = "proteins_dim_red")  
-}
-
-
 
 # show missing values again
 scp[["proteins_imptd"]] %>%
@@ -563,7 +549,7 @@ scp[["proteins_imptd"]] %>%
 
 library(scater)
 
-scp[["proteins_final"]] <- runPCA(scp[["proteins_final"]],
+scp[["proteins_dim_red"]] <- runPCA(scp[["proteins_dim_red"]],
                                    ncomponents = 5,
                                    ntop = Inf,
                                    scale = TRUE,
@@ -571,13 +557,13 @@ scp[["proteins_final"]] <- runPCA(scp[["proteins_final"]],
                                    name = "PCA")
 
 
-plotReducedDim(scp[["proteins_final"]],
+plotReducedDim(scp[["proteins_dim_red"]],
                dimred = "PCA",
                colour_by = "SampleType",
                point_alpha = 1)
 
 # UMAP
-scp[["proteins_final"]] <- runUMAP(scp[["proteins_final"]],
+scp[["proteins_dim_red"]] <- runUMAP(scp[["proteins_dim_red"]],
                                    ncomponents = 2,
                                    ntop = Inf,
                                    scale = TRUE,
@@ -586,7 +572,7 @@ scp[["proteins_final"]] <- runUMAP(scp[["proteins_final"]],
                                    dimred = "PCA",
                                    name = "UMAP")
 
-plotReducedDim(scp[["proteins_final"]],
+plotReducedDim(scp[["proteins_dim_red"]],
                dimred = "UMAP",
                colour_by = "SampleType",
                point_alpha = 1)

### 2023-04-18 scpR-app/app.R

@@ -53,8 +53,8 @@ ui <- fluidPage(
       selectInput("norm_method", "Choose method for protein data normalization", choices = c("SCoPE2", "None", "CONSTANd")),
       switchInput(inputId = "opts_multiple_batches", onLabel = "Advanced", offLabel = "Default", value = F, label="Options for multiple batches"),
       conditionalPanel(condition = "input.opts_multiple_batches", 
-                       selectInput(inputId = "missing_v", "Choose method for missing value handling", choices=c("KNN", "drop rows", "replace with 0", "replace with mean", "replace with median"))
-                       #switchInput(inputId = "batch_correction", onLabel = "Batch correction", offLabel = "No batch correction")
+                       selectInput(inputId = "missing_v", "Choose method for missing value handling", choices=c("KNN", "drop rows", "replace with 0", "replace with mean", "replace with median")),
+                       selectInput(inputId = "batch_c", "Choose method for batch correction", choices=c("ComBat", "none"))
                        )
       
     ),
@@ -510,24 +510,25 @@ server <- function(input, output, session) {
         }
         
         incProgress(16/17, detail=paste("running batch correction"))
-        
         sce <- getWithColData(scp_0, "proteins_imptd")
+        if (input$batch_c == "ComBat") {
+          batch <- colData(sce)$Raw.file
+          model <- model.matrix(~0 + SampleType, data = colData(sce))
+          
+          assay(sce) <- ComBat(dat = assay(sce),
+                               batch = batch)#,
+                               #mod = model)
+        } else if (input$batch_c == "none") {
+          scp_0 <- addAssay(scp_0,
+                            y = sce,
+                            name = "proteins_dim_red")
+          
+          scp_0 <- addAssayLinkOneToOne(scp_0,
+                                        from = "proteins_imptd",
+                                        to = "proteins_dim_red")          
+        }
         
-        batch <- colData(sce)$Raw.file
-        model <- model.matrix(~0 + SampleType, data = colData(sce))
-        
-        assay(sce) <- ComBat(dat = assay(sce),
-                             batch = batch)#,
-                             #mod = model)
-        
-        
-        scp_0 <- addAssay(scp_0,
-                        y = sce,
-                        name = "proteins_dim_red")
-        
-        scp_0 <- addAssayLinkOneToOne(scp_0,
-                                    from = "proteins_imptd",
-                                    to = "proteins_dim_red")
+
         
         
       } else {

### 2023-04-19 scope2.R
update for multiple selection in dim red plot
fix null bug

@@ -11,12 +11,17 @@
 # read in SCP data
 
 library("scp")
+library("SingleCellExperiment")
 library("ggplot2")
 library("magrittr")
 library("dplyr")
+library("reshape2")
+library("scater")
+library("limma")
 library("CONSTANd")
-library("MASS")
-
+library("stats")
+library("impute")
+library("sva")
 
 # read in MS result table
 mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")

### 2023-04-19 scpR-app/app.R

@@ -3,6 +3,7 @@ library("shinyWidgets")
 library("shinydashboard")
 
 library("scp")
+library("SingleCellExperiment")
 library("ggplot2")
 library("magrittr")
 library("dplyr")
@@ -65,9 +66,18 @@ ui <- fluidPage(
       tabsetPanel(type = "tabs",
                   tabPanel("Overview", plotOutput("overview_plot")),
                   tabPanel("Summary Barplot", plotOutput("summary_bar")),
-                  tabPanel("Reporter Ion Intensity", plotOutput("RI_intensity")),
-                  tabPanel("Covariance across razor peptides", plotOutput("CV_median")),
-                  tabPanel("Dimensionality reduction", plotOutput("PCA"), plotOutput("UMAP")),
+                  tabPanel("Reporter Ion Intensity", 
+                           selectInput("color_variable_ri", "select variable to indicate", ""),
+                           plotOutput("RI_intensity")),
+                  tabPanel("Covariance across razor peptides",
+                           selectInput("color_variable_cv", "select variable to indicate", ""),
+                           plotOutput("CV_median")),
+                  tabPanel("Dimensionality reduction",
+                           selectInput("color_variable_dim_red", "select variable to color", ""),
+                           selectInput("shape_variable_dim_red", "select variable to shape", ""),
+                           selectInput("size_variable_dim_red", "select variable to size", ""),                           
+                           plotOutput("PCA"), 
+                           plotOutput("UMAP")),
                   tabPanel("Feature wise output", 
                            selectInput("selectedProtein", "Choose protein for observation", ""),
                            plotOutput("feature_subset")),
@@ -513,12 +523,34 @@ server <- function(input, output, session) {
         sce <- getWithColData(scp_0, "proteins_imptd")
         if (input$batch_c == "ComBat") {
           batch <- colData(sce)$Raw.file
+          # can be used to aim batch correction for desired result
           model <- model.matrix(~0 + SampleType, data = colData(sce))
           
           assay(sce) <- ComBat(dat = assay(sce),
                                batch = batch)#,
                                #mod = model)
+          
+          scp_0 <- addAssay(scp_0,
+                            y = sce,
+                            name = "proteins_batchC")
+          
+          scp_0 <- addAssayLinkOneToOne(scp_0,
[…]

### 2023-04-24 scpR-app/app.R
sample type = "Pool" will be excluded

@@ -189,6 +189,13 @@ server <- function(input, output, session) {
                        suffix = paste0(input$label_suffix, 1:length(unique(meta_data_0$Channel))),
                        removeEmptyCols = TRUE)
       
+      # skip pool sample
+      if ("Pool" %in% meta_data_0$SampleType) {
+        scp_0 <- scp_0[, scp_0$SampleType !=  "Pool"]
+        print("Pool")
+      }
+      print(scp_0$SampleType)
+      
       # # change zeros to NA, apply first filter
       incProgress(2/17, detail = paste("replacing zeros with NA"))
       scp_0 <- zeroIsNA(scp_0, 1:length(rowDataNames(scp_0)))

### 2023-04-24 markdown/methods.Rmd
methods: transformation started > boxcox explained

@@
 
+##### Transformation of protein expression data
+Data transformation applies a function to each value of a matrix or array, so that: 
+$$
+y_i = f(x_i)
+$$
+Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill dependencies for statistical testing.
+In the shiny application various procedures are implemented and can be further expanded upon request from the user.
+The macrophage analysis done by Specht et.
+Al mentioned in their SCoPE2 publication citep\{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and the remove skewness in the dataset.
+This transformation was used as a reference when comparing methods.
+However using the logarithm to the base 10 can be easy interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method which is also implemented in the application.
+The boxcox transformation developed 1964 by Box and Cox citep\{Sakia1992} applies a recursion on the data by determining the parameter lambda.
+Depending on the size of lambda a certain function will be applied on each value of the expression set in order to introduce normality.
+As already mentioned transformations applied by the BoxCox method range from log10 to taking the square root of each value
+
 Further filter steps can include correction between multiple runs.

### 2023-04-24 scpR-app/app.R

@@ -192,9 +192,7 @@ server <- function(input, output, session) {
       # skip pool sample
       if ("Pool" %in% meta_data_0$SampleType) {
         scp_0 <- scp_0[, scp_0$SampleType !=  "Pool"]
-        print("Pool")
       }
-      print(scp_0$SampleType)
       
       # # change zeros to NA, apply first filter
       incProgress(2/17, detail = paste("replacing zeros with NA"))

### 2023-04-25 scope2.R
started experimenting with time series DE
minor commit before system update

@@ -55,6 +55,12 @@ scp <- readSCP(featureData = mqScpData,
                suffix = paste0("_TMT", 1:length(unique(sampleAnnotation$Channel))),
                removeEmptyCols = TRUE)
 
+# skip pool sample
+if ("Pool" %in% sampleAnnotation$SampleType) {
+  scp <- scp[, scp$SampleType !=  "Pool"]
+}
+
+
 
 # plot runs (not needed - sample unfractionized)
 # only one run
@@ -367,8 +373,8 @@ if (round(lambda, digits = 0) == -1 || lambda < -0.75 && lambda > -1.5) {
   print("1/protein_matrix")
 }
 if (round(lambda, digits = 1) == -0.5 || lambda < -0.25 && lambda > -0.75) {
-  protein_matrix <- 1/protein_matrix**2
-  print("1/protein_matrix**2")
+  protein_matrix <- 1/(protein_matrix**1/2)
+  print("1/(protein_matrix**1/2)")
 }
 if (round(lambda, digits = 0) == 0 || lambda < 0.25 && lambda > - 0.25 ) {
   protein_matrix <- log10(protein_matrix)
@@ -475,7 +481,7 @@ index_B <- st_indeces[choice_B]
 MAplot(assay(scp[["proteins_transf"]][,index_A[[1]]]), assay(scp[["proteins_transf"]][,index_B[[1]]]))
 
 protein_matrix <- assay(scp[["proteins_transf"]])
-#protein_matrix <- CONSTANd(protein_matrix)
+protein_matrix <- CONSTANd(protein_matrix)
 
 sce <- getWithColData(scp, "proteins")
 
@@ -487,8 +493,8 @@ scp <- addAssayLinkOneToOne(scp,
                             from = "proteins_transf",
                             to = "proteins_norm")
 
-assay(scp[["proteins_norm"]]) <- assay(scp[["proteins_transf"]])
-#assay(scp[["proteins_norm"]]) <- protein_matrix$normalized_data
+#assay(scp[["proteins_norm"]]) <- assay(scp[["proteins_transf"]])
+assay(scp[["proteins_norm"]]) <- protein_matrix$normalized_data
 
 
 # missing value imputation
@@ -585,27 +591,27 @@ plotReducedDim(scp[["proteins_dim_red"]],
 
 
 
-## Get the features
-subsetByFeature(scp, "Q9ULV4") %>%
-  ## Format the `QFeatures` to a long format table
[…]

### 2023-04-25 scpR-app/app.R

@@ -324,6 +324,8 @@ server <- function(input, output, session) {
       
       incProgress(14/17, detail=paste("transforming protein data"))
       req(input$transform_base)
+      transform_base_bc <- "NULL"
+      
       if (input$transform_base == "log2") {
         scp_0 <- logTransform(scp_0,
                               base = 2,
@@ -369,8 +371,8 @@ server <- function(input, output, session) {
           print("1/protein_matrix")
         }
         if (round(lambda, digits = 1) == -0.5 || lambda < -0.25 && lambda > -0.75) {
-          protein_matrix <- 1/protein_matrix**2
-          print("1/protein_matrix**2")
+          protein_matrix <- 1/(protein_matrix**1/2)
+          print("1/(protein_matrix**1/2)")
         }
         if (round(lambda, digits = 0) == 0 || lambda < 0.25 && lambda > - 0.25 ) {
           protein_matrix <- log10(protein_matrix)
@@ -416,7 +418,7 @@ server <- function(input, output, session) {
       
       incProgress(15/17, detail=paste("normalizing proteins"))
       req(input$norm_method)
-      if (input$norm_method == "SCoPE2" && input$transform_base == "log2" | input$transform_base == "log10" ) {
+      if (input$norm_method == "SCoPE2" && input$transform_base == "log2" | input$transform_base == "log10" | transform_base_bc == "log10") {
         # center cols with median
         scp_0 <- sweep(scp_0, i = "proteins_transf",
                        MARGIN = 2,

### 2023-04-26 markdown/methods.Rmd
first draft flowchart for analysis pipeline

@@
 ## Data
-```{r Dataflow,  echo=FALSE}
-library(DiagrammeR)
-library(webshot)
-
-DiagrammeR::grViz("digraph {
-	graph [layout  =  dot,
-	       rankdir  = TB]
-
-node [shape = rectangle]
-rawData [label = 'msData.raw']
-node [shape = oval]
-MQ [label = 'MaxQuant']
-node [shape = invhouse]
-MQ_fd [label = 'feature detection']
-node [shape = cylinder]
-MQ_ds [label = 'Database search']
-node [shape = invhouse]
-MQ_fdr [label = 'FDR calculation']
-MQ_quant [label = 'Protein Quantification']
-node [shape = rectangle]
-MQoutput [label = 'evidence.txt']
-sampleAnnotate [label = 'sample_annotation.txt']
-node [shape = oval]
-ProteomicsWorkbench [label = 'Proteomics Workbench']
-
-rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
-sampleAnnotate -> ProteomicsWorkbench
-}",
-height = 500)
+```{r Dataflow, eval=FALSE, include=FALSE}
+# library(DiagrammeR)
+# library(webshot)
+# 
+# DiagrammeR::grViz("digraph {
+# 	graph [layout  =  dot,
+# 	       rankdir  = TB]
+# 
+# node [shape = rectangle]
+# rawData [label = 'msData.raw']
+# node [shape = oval]
+# MQ [label = 'MaxQuant']
+# node [shape = invhouse]
+# MQ_fd [label = 'feature detection']
+# node [shape = cylinder]
+# MQ_ds [label = 'Database search']
+# node [shape = invhouse]
+# MQ_fdr [label = 'FDR calculation']
+# MQ_quant [label = 'Protein Quantification']
+# node [shape = rectangle]
+# MQoutput [label = 'evidence.txt']
+# sampleAnnotate [label = 'sample_annotation.txt']
+# node [shape = oval]
+# ProteomicsWorkbench [label = 'Proteomics Workbench']
+# 
+# rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
+# sampleAnnotate -> ProteomicsWorkbench
+# }",
+# height = 500)
 ```
@@
 
-### MS-Spectrum
+#### MS-Spectrum
 
@@
 
-### Interpretation of the data
-
-#### Data acquisition
-
 Since ms data has a high resolution, algorithms are used to convert the raw signal to an interpretable form. 
@@
 
-##### Three-dimensional peak detection
+#### Three-dimensional peak detection
 
@@
 
-##### Deisotoping
+#### Deisotoping
 
@@
 
-##### Label detection
+#### Label detection
 
@@
 
-##### Improving peptide mass accuracy
+#### Improving peptide mass accuracy
 
@@
 
-##### Peptide search
+#### Peptide search
 
@@
 
-##### Protein assembly
+#### Protein assembly
 
@@
 
-#### Data processing
+### Data processing
 
-Further analysis can be done with R and respective packages such as bioconductor. Since there is no state of the art established, analysis varies upon experimental design.
+Further analysis is done with R and respective packages such as bioconductor.
+Since there is no state of the art established, analysis varies upon experimental design.
+The workflow of the analysis will be processed in a so called pipeline streamlining the data through steps where individual results can be observed in visualizations and indivdual calculations will be adapted according to user demands and experimental properties.
 
-##### Reading the data
+```{r data_processing_pipeline_flowchart, echo=FALSE, fig.height=10}
+library(Gmisc, quietly = TRUE)
+library(glue)
+library(htmlTable)
+library(grid)
+library(magrittr)
+
+opts_box_setting <- gpar(fill="lightyellow", col="darkblue")
+opts_txt_setting <- gpar(col="darkblue", cex=1)
+
+# input files
+evidence_file <- boxGrob({glue("evidence file",
+                           .sep = "\n")})
+sample_annotation_file <- boxGrob({glue("sample annotation",
+                           .sep = "\n")})
+
+read_data <- boxGrob({glue("read data into q-feature object",
+                          .sep = "\n")})
+
+quality_control_1 <- boxGrob({glue("quality control 1",
+                          .sep = "\n")})
+qc1_content <- boxGrob("filter out pool samples\n
+                             replace zeros with NA\n
+                             filter out contaminants\n
+                             filter by q-value cutoff\n",
+                       txt_gp = opts_box_setting,
+                       box_gp = opts_box_setting,
+                       just="left"
+                       )
+
+aggregate_psms <-  boxGrob({glue("aggregate peptide spectrum matches to
+                                 peptides\n
+                                 join multiple batches",
+                          .sep = "\n")})
+
+quality_control_2 <- boxGrob({glue("quality control 2",
+                          .sep = "\n")})
+qc2_content <- boxGrob("filter by reporter ion intensity cutoff\n
+                             filter by pepitde covariance to razor protein\n
+                             remove peptides by missing rate\n",
+                       txt_gp = opts_box_setting,
+                       box_gp = opts_box_setting,
+                       just="left"
+                       )
+
+aggregate_peps <- boxGrob({glue("aggregate peptides to proteins",
+                          .sep = "\n")})
+
+transform_data <- boxGrob({glue("transform data",
+                          .sep = "\n")})
+transform_opts <- boxGrob("options:\n
+                                log2\n
+                                log10\n
+                                sqrt\n
+                                quadratic\n
+                                BoxCox\n
+                                None",
+                       txt_gp = opts_box_setting,
+                       box_gp = opts_box_setting,
+                       just="left"
+                       )
+
+normalize_data <- boxGrob({glue("normalize data",
+                          .sep = "\n")})
+normalize_opts <- boxGrob("options:\n
+                                SCoPE2\n
+                                CONSTANd\n
+                                None",
+                       txt_gp = opts_box_setting,
+                       box_gp = opts_box_setting,
+                       just="left"
+                       )
+
+missing_value_handling <- boxGrob({glue("missing value handling",
+                          .sep = "\n")})
+missing_value_opts <- boxGrob("options:\n
+                                    KNN\n
+                                    drop rows\n
+                                    replace with mean\n
+                                    replace with median\n
+                                    replace with 0",
+                       txt_gp = opts_box_setting,
+                       box_gp = opts_box_setting,
+                       just="left"
+                       )
+
+batch_correction <- boxGrob({glue("batch correction",
+                          .sep = "\n")})
+batch_correction_opts <- boxGrob("options:\n
+                                       ComBat",
+                       txt_gp = opts_box_setting,
+                       box_gp = opts_box_setting,
+                       just="left"
+                       )
+
+dim_red <- boxGrob({glue("dimensionality reduction",
+                          .sep = "\n")})
+dim_red_actions <- boxGrob("procedures:\n
+                                 PCA\n
+                                 UMAP",
+                       txt_gp = opts_box_setting,
+                       box_gp = opts_box_setting,
+                       just="left"
+                       )
+
+stat_module <- boxGrob({glue("statistics module",
+                          .sep = "\n")})
+
+grid.newpage()
+vert <- {spreadVertical(files = evidence_file,
+                       read_data = read_data,
+                       quality_control_1 = quality_control_1,
+                       aggregate_psms,
+                       quality_control_2 = quality_control_2,
+                       aggregate_peps,
+                       transform_data = transform_data,
+                       normalize_data = normalize_data,
+                       missing_value_handling = missing_value_handling,
+                       batch_correction = batch_correction,
+                       dim_red = dim_red,
+                       stat_module
+)}
+input <- alignVertical(reference = vert$files,
+                      evidence_file, sample_annotation_file) %>%
+  spreadHorizontal()
+vert$files <- NULL
+
+# right side
+qc1_content <- moveBox(qc1_content,
+                    x = .8,
+                    y = coords(vert$quality_control_1)$top)
+
+transform_opts <- moveBox(transform_opts,
+                    x = .9,
+                    y = coords(vert$transform_data)$top)
+
+missing_value_opts <- moveBox(missing_value_opts,
+                    x = .7,
+                    y = coords(vert$missing_value_handling)$top)
+
+dim_red_actions <- moveBox(dim_red_actions,
+                    x = .9,
+                    y = coords(vert$dim_red)$top)
+
+# left side
+qc2_content <- moveBox(qc2_content,
+                    x = .2,
+                    y = coords(vert$quality_control_2)$top)
+
+normalize_opts <- moveBox(normalize_opts,
+                    x = .2,
+                    y = coords(vert$normalize_data)$top)
+
+batch_correction_opts <- moveBox(batch_correction_opts,
+                    x = .2,
+                    y = coords(vert$batch_correction)$top)
+
+for (i in 1:(length(vert) - 1)) {
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+    print
+}
+
+# connect boxes
+connectGrob(input[[1]], vert$read_data, type = "N")
+connectGrob(input[[2]], vert$read_data, type = "N")
+
+connectGrob(vert$quality_control_1, qc1_content, type = "horizontal", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$quality_control_2, qc2_content, type = "horizontal", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$transform_data, transform_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$normalize_data, normalize_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$missing_value_handling, missing_value_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$batch_correction, batch_correction_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$dim_red, dim_red_actions, type = "horizontal", arrow_obj = arrow(angle=0))
+
+# Print boxes
+input
+vert
+qc1_content
+qc2_content
+transform_opts
+normalize_opts
+missing_value_opts
+batch_correction_opts
+dim_red_actions
+```
+
+#### Reading the data
 
@@
 
-##### Object oriented programming
+#### Object oriented programming
 
@@
 
-##### Zero values
+#### Zero values
 Peptides with low abundance are often set to zero during analysis.
@@
 
-##### Exclude reverse matches/contaminants
+#### Exclude reverse matches/contaminants
 Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.
 
-##### Filter according to precursor ion fraction (PIF)
+#### Filter according to precursor ion fraction (PIF)
 During mass spectrometry, the ions detected in MS1 are further fragmented through collision during multiple MS runs.
@@
 
-##### Filter by q-value
+#### Filter by q-value
 The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
@@
 
-##### Peptide spectrum match (PSM) aggregation to peptides 
+#### Peptide spectrum match (PSM) aggregation to peptides 
 In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function.
@@
 
-##### Join assays when observing multiple comparable batches at once
+#### Join assays when observing multiple comparable batches at once
 Sample size is often a limiting factor in hypothesis testing.
@@
 
-##### Calculate reporter ion intensity (RI) and filter according to median RI
+#### Calculate reporter ion intensity (RI) and filter according to median RI
 Columns which do not meet the desired intensity can be filtered by a threshold set on the RI.
@@
 
-##### Calculate and filter according to median coefficient of variation (CV) per cell/channel
+#### Calculate and filter according to median coefficient of variation (CV) per cell/channel
 Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data.
@@
 
-##### Remove peptides with high missing rate
+#### Remove peptides with high missing rate
 Although missing value imputation can be performed during the analysis of multiple batches, peptides with missing detections across channels can be problematic for quantification.
@@
 
-##### Aggregation of peptides to proteins 
+#### Aggregation of peptides to proteins 
 Similar to the already explained previous aggregation step the peptides will be further processed into their respective proteins after the quality control on the peptide level is performed.
 
-##### Transformation of protein expression data
+#### Transformation of protein expression data
 Data transformation applies a function to each value of a matrix or array, so that: 

### 2023-04-26 markdown/methods.tex
first draft flowchart for analysis pipeline

--- 2023-03-08 markdown/Thesis.tex
@@
 \documentclass[
-  11pt,
 ]{article}
@@
 \usepackage{xcolor}
-\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
-2cm]{geometry}
+\usepackage[margin=1in]{geometry}
 \usepackage{graphicx}
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage{float}
-\usepackage{sectsty}
-\usepackage{paralist}
-\usepackage{setspace}\spacing{1.5}
-\usepackage{fancyhdr}
-\usepackage{lastpage}
-\usepackage{dcolumn}
-\usepackage{natbib}\bibliographystyle{agsm}
-\usepackage[nottoc, numbib]{tocbibind}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \ifLuaTeX
@@
 
-\allsectionsfont{\centering}
-\subsectionfont{\raggedright}
-\subsubsectionfont{\raggedright}
+\hypertarget{materials-and-methods}{%
+\section{Materials and Methods}\label{materials-and-methods}}
 
-\pagenumbering{gobble}
+\hypertarget{materials}{%
+\subsection{Materials}\label{materials}}
 
-\begin{centering}
+For analysis two types of cells were used. One type is the Jurkat-based
+cell line (J-lat) with integrated HIV.
 
-\vspace{3cm}
-
-\includegraphics[width=0.2\linewidth]{FHCW_logo} 
-\includegraphics[width=0.2\linewidth]{KI_logo}
-
-\vspace{1cm}
-
-\Large
-\doublespacing
-{\bf Design and implementation of analysis pipeline for single cell type proteomics data}
-
-\vspace{1 cm}
-
-\normalsize
-\singlespacing
-By
-
-\vspace{0.5 cm}
-
-\Large
-
-{\bf Lukas Gamp}
-
-\vspace{1.5 cm}
-
-in partial fulfilment of the requirement \\for the degree of MSc \\in Bioinformatics
-
-\vspace{1.5 cm}
-
-\normalsize
-mm yy
-\end{centering}
-
-\newpage
-
-\pagenumbering{gobble}
-
-\begin{centering}
-
-{\bf Abstract}
-
-\end{centering}
-
-\spacing{1.5}
-
-(the spacing is set to 1.5)
-
-no more than 250 words for the abstract
-
-\begin{itemize}
-\tightlist
-\item
-  a description of the research question/knowledge gap -- what we know
-  and what we don't know
-\item
-  how your research has attempted to fill this gap
-\item
-  a brief description of the methods
-\item
-  brief results
-\item
-  key conclusions that put the research into a larger context
-\end{itemize}
-
-\pagenumbering{roman}
-
-\newpage
-
-\centering
-\raggedright
-\newpage
-\tableofcontents
-
-\newpage
-
-\section*{Acknowledgements}
-
-Thank you for following this tutorial!
-
-I hope you'll find it useful to write a very professional dissertation.
-
-\newpage
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-\hypertarget{proteomics}{%
-\subsection{Proteomics}\label{proteomics}}
-
-The proteome is referred to the sum of all proteins of a given sample at
-a given time. In the past several quantitative and qualitative assays
-were used to enlighten the protein composition of a sample.
-
-An early approach of qualitative analysis of the cellular proteome
-involved labeling with fluorescent antibodies and imaging. The major
-disadvantage of this technique was the limitation to only stain a few
-proteins per cell. For quantification procedures such as single-cell
-Western blots, immunoassays or CyTOF have been used. Other disadvantages
-are the ability to permeate cells, accessibility and binding of the
-epitope and the creation of specific antibodies for a given protein
-\citep{Budnik2018}.
-
-One of those techniques involved RNA-sequencing. Since RNA involves also
-non-coding RNA, the amount of RNA is often not proportional to the
-content of proteins in a cell. So the proteinaceous content of a cell
-was only predicted and quantitative analysis was not possible.
-
-\hypertarget{mass-spectrometry}{%
-\subsection{Mass Spectrometry}\label{mass-spectrometry}}
-
-Mass spectrometry enables qualitative and quantitative analysis of the
-entire repertoire of a biological sample. The availability of gene
-sequences in databases and the ability to match proteins against those
-sequences with computational methods makes it possible to identify
-alterations of a sample on a protein level. These alterations can rely
-on the sequence level or could be to post-translational modifications
-(PTMs) such as phosphorylation, methylation or else
-\citep{Aebersold2003}.
-
-Mass to charge ratio (m/z)
-
-\hypertarget{instrumentation}{%
-\subsubsection{Instrumentation}\label{instrumentation}}
+The other type of cells are macrophages with a sample size of 72 cells.
+The analysis is done with two groups. A HIV negative (HIV-) control
+group and a HIV positive (HIV+) group.
 
 \hypertarget{cell-isolation}{%
-\paragraph{Cell Isolation}\label{cell-isolation}}
+\subsubsection{Cell Isolation}\label{cell-isolation}}
 
 \hypertarget{lysis}{%
-\paragraph{Lysis}\label{lysis}}
+\subsubsection{Lysis}\label{lysis}}
 
 \hypertarget{digestion}{%
-\paragraph{Digestion}\label{digestion}}
-
-\hypertarget{liquid-chromatography}{%
-\paragraph{Liquid chromatography}\label{liquid-chromatography}}
-
-In order to separate proteins according to their chemical properties,
-size or species a liquid chromatography (LC) is recommended before
-ionization.
-
-\hypertarget{ionization}{%
-\paragraph{Ionization}\label{ionization}}
-
-In order to analyze a biological sample consisting of proteins in
-solution the liquid needs to be vaporized into gas phase. Two techniques
-are capable of this procedure. Electrospray ionization (ESI) pushes the
-analyte through a capillary and applies an electric current to the
-liquid, vaporizing the sample to a charged aerosol. Biomolecules are
-fragmented according to their chemical properties and can be further
-handled in the mass spectrometer. The fragmented biomolecules are now in
-charged droplets separated by their charge on the surface, splitting
-further into smaller droplets until they become a gas phase ion. Two
-physical models describe the process from gas phase to ion called ``The
-ion evaporation model'' (IEM) and ``The charge residue model'' (CRM). In
-the ion evaporation model (by Iribarne and Thomoson) the droplets shrink
-by evaporation until ions are expelled. The model had its limitation by
-explaining same evaporation rate constant among ions with different
-chemical properties. In the charge residue model the assumption of one
-molecule per droplet leads to an ionization rate constant, which is
-independent of the ion itself and relies solely on the generation of the
-droplet and the efficiency of the solvent \citep{Wilm2011}.
-
-Matrix-assisted laser desorption/ionization (MALDI)
-
-\hypertarget{coupled-mass-spectrometry-msms}{%
-\paragraph{Coupled mass-spectrometry
-(MS/MS)}\label{coupled-mass-spectrometry-msms}}
-
-In order to enhance sequence identification, two MS devices are built in
-series. In the first run (MS1) the m/z is determined and the molecules
-are passed to the next device. Upon passing the molecules are fragmented
-into smaller ions and analyzed by the second MS. The fragmentation
-highly depends on the chemical bonds found in the molecule. The majority
-of these breaks occur on the peptide bond of the protein, although this
-is not guaranteed for all bonds and so it can happen that certain
-peptide ions have a low abundance \citep{Budnik2018}. These low abundant
-peptides will not be detected, hence the problem needs to be faced with
-another approach . A solution for this problem is molecular barcoding
-with labeling mentioned in the chapter labeling.
+\subsubsection{Digestion}\label{digestion}}
 
 \hypertarget{labeling-techniques}{%
-\paragraph{Labeling techniques}\label{labeling-techniques}}
+\subsubsection{Labeling techniques}\label{labeling-techniques}}
 
@@
 \hypertarget{isobaric-labeling}{%
-\subparagraph{Isobaric labeling}\label{isobaric-labeling}}
+\paragraph{Isobaric labeling}\label{isobaric-labeling}}
 
-Tandem mass tag (TMT)
+\hypertarget{tandem-mass-tag-tmt}{%
+\subparagraph{Tandem mass tag (TMT)}\label{tandem-mass-tag-tmt}}
 
@@
 
+\hypertarget{instrumentation}{%
+\subsubsection{Instrumentation}\label{instrumentation}}
+
+\hypertarget{liquid-chromatography}{%
+\paragraph{Liquid chromatography}\label{liquid-chromatography}}
+
+In order to separate proteins according to their chemical properties,
+size or species a liquid chromatography (LC) is recommended before
+ionization.
+
+\hypertarget{mass-spectrometry}{%
+\paragraph{Mass Spectrometry}\label{mass-spectrometry}}
+
+\hypertarget{ionization}{%
+\subparagraph{Ionization}\label{ionization}}
+
+In order to analyze a biological sample consisting of proteins in
+solution the liquid needs to be vaporized into gas phase. Two techniques
+are capable of this procedure. Electrospray ionization (ESI) pushes the
+analyte through a capillary and applies an electric current to the
+liquid, vaporizing the sample to a charged aerosol. Biomolecules are
+fragmented according to their chemical properties and can be further
+handled in the mass spectrometer. The fragmented biomolecules are now in
+charged droplets separated by their charge on the surface, splitting
+further into smaller droplets until they become a gas phase ion. Two
+physical models describe the process from gas phase to ion called ``The
+ion evaporation model'' (IEM) and ``The charge residue model'' (CRM). In
+the ion evaporation model (by Iribarne and Thomoson) the droplets shrink
+by evaporation until ions are expelled. The model had its limitation by
+explaining same evaporation rate constant among ions with different
+chemical properties. In the charge residue model the assumption of one
+molecule per droplet leads to an ionization rate constant, which is
+independent of the ion itself and relies solely on the generation of the
+droplet and the efficiency of the solvent \citep{Wilm2011}.
+
+Matrix-assisted laser desorption/ionization (MALDI)
+
+\hypertarget{ms.1}{%
+\subparagraph{MS.1}\label{ms.1}}
+
+\hypertarget{coupled-mass-spectrometry-msms-ms.2}{%
+\subparagraph{Coupled mass-spectrometry (MS/MS) \&
+MS.2}\label{coupled-mass-spectrometry-msms-ms.2}}
+
+In order to enhance sequence identification, two MS devices are built in
+series. In the first run (MS1) the m/z is determined and the molecules
+are passed to the next device. Upon passing the molecules are fragmented
+into smaller ions and analyzed by the second MS. The fragmentation
+highly depends on the chemical bonds found in the molecule. The majority
+of these breaks occur on the peptide bond of the protein, although this
+is not guaranteed for all bonds and so it can happen that certain
+peptide ions have a low abundance \citep{Budnik2018}. These low abundant
+peptides will not be detected, hence the problem needs to be faced with
+another approach . A solution for this problem is molecular barcoding
+with labeling mentioned in the chapter labeling.
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+\hypertarget{acquisition}{%
+\subsubsection{Acquisition}\label{acquisition}}
+
+Acquisition of the data was done with MaxQuant \citep{Cox2008} software
+package.
+
 \hypertarget{ms-spectrum}{%
-\subsubsection{MS-Spectrum}\label{ms-spectrum}}
+\paragraph{MS-Spectrum}\label{ms-spectrum}}
 
@@
 
-\hypertarget{interpretation-of-the-data}{%
-\subsubsection{Interpretation of the
-data}\label{interpretation-of-the-data}}
-
-\hypertarget{data-acquisition}{%
-\paragraph{Data acquisition}\label{data-acquisition}}
-
 Since ms data has a high resolution, algorithms are used to convert the
@@
 \hypertarget{three-dimensional-peak-detection}{%
-\subparagraph{Three-dimensional peak
+\paragraph{Three-dimensional peak
 detection}\label{three-dimensional-peak-detection}}
@@
 \hypertarget{deisotoping}{%
-\subparagraph{Deisotoping}\label{deisotoping}}
+\paragraph{Deisotoping}\label{deisotoping}}
 
@@
 \hypertarget{label-detection}{%
-\subparagraph{Label detection}\label{label-detection}}
+\paragraph{Label detection}\label{label-detection}}
 
@@
 \hypertarget{improving-peptide-mass-accuracy}{%
-\subparagraph{Improving peptide mass
+\paragraph{Improving peptide mass
 accuracy}\label{improving-peptide-mass-accuracy}}
@@
 \hypertarget{peptide-search}{%
-\subparagraph{Peptide search}\label{peptide-search}}
+\paragraph{Peptide search}\label{peptide-search}}
 
@@
 \hypertarget{protein-assembly}{%
-\subparagraph{Protein assembly}\label{protein-assembly}}
+\paragraph{Protein assembly}\label{protein-assembly}}
 
@@
 random event, are multiplied and only distinct sequences with the
-highst-scoring are used.
+highest-scoring are used.
 
 \hypertarget{data-processing}{%
-\paragraph{Data processing}\label{data-processing}}
+\subsubsection{Data processing}\label{data-processing}}
 
-Further analysis can be done with R and respective packages such as
+Further analysis is done with R and respective packages such as
 bioconductor. Since there is no state of the art established, analysis
-varies upon experimental design.
+varies upon experimental design. The workflow of the analysis will be
+processed in a so called pipeline streamlining the data through steps
+where individual results can be observed in visualizations and indivdual
+calculations will be adapted according to user demands and experimental
+properties.
+
+\includegraphics{methods_files/figure-latex/data_processing_pipeline_flowchart-1.pdf}
 
 \hypertarget{reading-the-data}{%
-\subparagraph{Reading the data}\label{reading-the-data}}
+\paragraph{Reading the data}\label{reading-the-data}}
 
 After processing MaxQuant creates a directory containing all results as
-.txt file. The evidence.txt file include all peptide sequence matches
+.txt file. The evidence.txt file include all peptide to spectrum matches
 (PSM) with their respective proteins and statistical parameters.
@@
 hierarchical structure. The structure enables sub setting for further
-analysis.
+analysis \citep{Vanderaa2021}.
 
-\hypertarget{filtering}{%
-\paragraph{Filtering}\label{filtering}}
+\hypertarget{zero-values}{%
+\paragraph{Zero values}\label{zero-values}}
 
-Peptides having a low abundance will be set to zero by the analysis. The
-value zero would imply that the sample does not contain the respective
-peptide, therefore replacing 0 with not applicable (NA) is crucial for
-understanding and interpretation of MS-data.
+Peptides with low abundance are often set to zero during analysis.
+However, assigning a value of zero may incorrectly suggest that the
+sample does not contain the respective peptide. Given that it is highly
+unlikely for a biological cell of a comparable type and function to not
+contain a particular protein, replacing the zero value with ``not
+applicable'' (NA) is crucial for understanding and interpreting MS data.
+
+\hypertarget{exclude-reverse-matchescontaminants}{%
+\paragraph{Exclude reverse
+matches/contaminants}\label{exclude-reverse-matchescontaminants}}
 
@@
 
-The author of the SCoPE2 \citep{Petelski2021} pipeline recommends
-calculating the sample to carrier ration (SCR). This can be done by
-dviding the ion intensity of the sample by the reporter ion intensity of
-the carrier channel. However in our experiment no carrier and reference
-channel is used.
+\hypertarget{filter-according-to-precursor-ion-fraction-pif}{%
+\paragraph{Filter according to precursor ion fraction
+(PIF)}\label{filter-according-to-precursor-ion-fraction-pif}}
+
+During mass spectrometry, the ions detected in MS1 are further
+fragmented through collision during multiple MS runs. The resulting
+product ions are derived from precursor ions (also known as mother ions
+or parental ions). Contaminant peptides can co-migrate in this process
+and can be distinguished by the lower fraction of their respective
+precursor ions \citep{Tannous2020}. These peptides need to be filtered
+out during the analysis pipeline. A cutoff value, referenced in the
+SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface,
+but it can be adjusted according to the needs of the biologist.
+
+\hypertarget{filter-by-q-value}{%
+\paragraph{Filter by q-value}\label{filter-by-q-value}}
 
@@
 accepting an FDR of an individual PSM is described as q-value.
+
+\hypertarget{peptide-spectrum-match-psm-aggregation-to-peptides}{%
+\paragraph{Peptide spectrum match (PSM) aggregation to
+peptides}\label{peptide-spectrum-match-psm-aggregation-to-peptides}}
+
+In data science, aggregation refers to a row-wise operation that merges
+data based on a particular column using a specific function. In the
+context of processing from peptides to spectrum matches, the desired
+column is the peptide sequence. To account for different distributions
+across multiple assays, the median of the channel is used as the
+function to aggregate multiple matches into one.
+
+\hypertarget{join-assays-when-observing-multiple-comparable-batches-at-once}{%
+\paragraph{Join assays when observing multiple comparable batches at
+once}\label{join-assays-when-observing-multiple-comparable-batches-at-once}}
+
+Sample size is often a limiting factor in hypothesis testing. A strict
+quality control and the fact that TMT reagents are only available up to
+18-plex can reduce the number of observed samples below the critical
+threshold, leading to an early end of analyses. To overcome this
+limitation, the provided software is capable of processing multiple runs
+simultaneously, allowing for testing of multiple batches and increasing
+the number of samples that can be included in the analysis.
+
+\hypertarget{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}{%
+\paragraph{Calculate reporter ion intensity (RI) and filter according to
+median
+RI}\label{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}}
+
+Columns which do not meet the desired intensity can be filtered by a
+threshold set on the RI. The median RI can also be used to check if an
+entire channel has a lower detection level. This can be due two reasons.
+One is the expression level of the given proteins in a cell. Meaning,
+that the expression of the observed cell type is simply lower than the
+other type. Another one could be a spillage of TMT detection in other
+channels due incorrect or missing correction of the TMT isotopes.
+
+\hypertarget{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}{%
+\paragraph{Calculate and filter according to median coefficient of
+variation (CV) per
+cell/channel}\label{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}}
+
+Depending having a bulk sample or single-cell sample choosing a minimum
+of observed peptides and a cutoff value for the CV, changes the level of
+confidence in the peptide data. The coefficient of variation of a
+peptide is considered as the ratio of the standard deviation to the mean
+and describes the relationship of the observed peptide signal over
+multiple proteins (=razor proteins). Peptides having a high coefficient
+of variation over many razor proteins are considered as noise and need
+to be filtered out before statistical analysis.
+
+\hypertarget{remove-peptides-with-high-missing-rate}{%
+\paragraph{Remove peptides with high missing
+rate}\label{remove-peptides-with-high-missing-rate}}
+
+Although missing value imputation can be performed during the analysis
+of multiple batches, peptides with missing detections across channels
+can be problematic for quantification. The proteomic composition of a
+biological sample is similar between replicates and even across groups.
+However, the threshold of missingness (described as a fraction of the
+row) can be set in the user interface and adjusted to enable different
+experimental designs.
+
+\hypertarget{aggregation-of-peptides-to-proteins}{%
+\paragraph{Aggregation of peptides to
+proteins}\label{aggregation-of-peptides-to-proteins}}
+
+Similar to the already explained previous aggregation step the peptides
+will be further processed into their respective proteins after the
+quality control on the peptide level is performed.
+
+\hypertarget{transformation-of-protein-expression-data}{%
+\paragraph{Transformation of protein expression
+data}\label{transformation-of-protein-expression-data}}
+
+Data transformation applies a function to each value of a matrix or
+array, so that: \[
+y_i = f(x_i)
+\] Depending on the distribution of the values in the observed
+expression set, different transformations can be applied to fulfill
+dependencies for statistical testing. In the shiny application various
+procedures are implemented and can be further expanded upon request from
+the user. The macrophage analysis done by Specht et. Al mentioned in
+their SCoPE2 publication citep\{Specht2021\} uses the logarithm to the
+base 2 to spread a compacted distribution and the remove skewness in the
+dataset. This transformation was used as a reference when comparing
+methods. However using the logarithm to the base 10 can be easy
+interpretable way of defining expression data and will also be
+facilitated by other transformation methods such as the boxcox method
+which is also implemented in the application. The boxcox transformation
+developed 1964 by Box and Cox citep\{Sakia1992\} applies a recursion on
+the data by determining the parameter lambda. Depending on the size of
+lambda a certain function will be applied on each value of the
+expression set in order to introduce normality. As already mentioned
+transformations applied by the BoxCox method range from log10 to taking
+the square root of each value
 
@@
 
-\pagenumbering{arabic}
-
-\newpage
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-\vspace{0.5cm}
-
-\hypertarget{materials-and-methods}{%
-\section{Materials and Methods}\label{materials-and-methods}}
-
-\hypertarget{materials}{%
-\subsection{Materials}\label{materials}}
-
-For analysis two types of cells were used. One type is the Jurkat-based
-cell line (J-lat) with integrated HIV.
-
-The other type of cells are macrophages with a sample size of 72 cells.
-The analysis is done with two groups. A HIV negative (HIV-) control
-group and a HIV positive (HIV+) group.
-
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-\hypertarget{acquisition}{%
-\subsubsection{Acquisition}\label{acquisition}}
-
-Analysis of the data was done with MaxQuant \citep{Cox2008}
-
-\newpage
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-Some more guidlines from the School of Geosciences.
-
-This section should summarise the findings of the research referring to
-all figures, tables and statistical results (some of which may be placed
-in appendices). - include the primary results, ordered logically - it is
-often useful to follow the same order as presented in the methods. -
-alternatively, you may find that ordering the results from the most
-important to the least important works better for your project. - data
-should only be presented in the main text once, either in tables or
-figures; if presented in figures, data can be tabulated in appendices
-and referred to at the appropriate point in the main text.
-
-\textbf{Often, it is recommended that you write the results section
-first, so that you can write the methods that are appropriate to
-describe the results presented. Then you can write the discussion next,
-then the introduction which includes the relevant literature for the
-scientific story that you are telling and finally the conclusions and
-abstract -- this approach is called writing backwards.}
-
-\newpage
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-the purpose of the discussion is to summarise your major findings and
-place them in the context of the current state of knowledge in the
-literature. When you discuss your own work and that of others, back up
-your statements with evidence and citations. - The first part of the
-discussion should contain a summary of your major findings (usually 2 --
-4 points) and a brief summary of the implications of your findings.
-Ideally, it should make reference to whether you found support for your
-hypotheses or answered your questions that were placed at the end of the
-introduction. - The following paragraphs will then usually describe each
-of these findings in greater detail, making reference to previous
-studies. - Often the discussion will include one or a few paragraphs
-describing the limitations of your study and the potential for future
-research. - Subheadings within the discussion can be useful for
-orienting the reader to the major themes that are addressed.
-
-\newpage
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-The conclusion section should specify the key findings of your study,
-explain their wider significance in the context of the research field
-and explain how you have filled the knowledge gap that you have
-identified in the introduction. This is your chance to present to your
-reader the major take-home messages of your dissertation research. It
-should be similar in content to the last sentence of your summary
-abstract. It should not be a repetition of the first paragraph of the
-discussion. They can be distinguished in their connection to broader
-issues. The first paragraph of the discussion will tend to focus on the
-direct scientific implications of your work (i.e.~basic science,
-fundamental knowledge) while the conclusion will tend to focus more on
-the implications of the results for society, conservation, etc.
-
-\newpage
-
-\bibliography{bibliography}
-
-\newpage
-
-\hypertarget{appendixces}{%
-\section{Appendix(ces)}\label{appendixces}}
-
-\hypertarget{appendix-a-additional-tables}{%
-\subsection{Appendix A: additional
-tables}\label{appendix-a-additional-tables}}
-
-Insert content for additional tables here.
-
-\newpage
-
-\hypertarget{appendix-b-additional-figures}{%
-\subsection{Appendix B: additional
-figures}\label{appendix-b-additional-figures}}
-
-Insert content for additional figures here.
-
-\newpage
-
-\hypertarget{appendix-c-code}{%
-\subsection{Appendix C: code}\label{appendix-c-code}}
-
-Insert code (if any) used during your dissertation work here.
-
 \end{document}

### 2023-04-26 markdown/methods.Rmd
flowchart updated

@@
 
-```{r data_processing_pipeline_flowchart, echo=FALSE, fig.height=10}
+```{r data_processing_pipeline_flowchart, echo=FALSE, fig.height=17, fig.width=13}
 library(Gmisc, quietly = TRUE)
@@
                           .sep = "\n")})
-qc1_content <- boxGrob("filter out pool samples\n
-                             replace zeros with NA\n
-                             filter out contaminants\n
-                             filter by q-value cutoff\n",
+qc1_content <- boxGrob(
+                      "filter out pool samples\nreplace zeros with NA\nfilter out contaminants\nfilter by q-value cutoff",
                        txt_gp = opts_box_setting,
                        box_gp = opts_box_setting,
-                       just="left"
+                       just="center",
                        )
 
-aggregate_psms <-  boxGrob({glue("aggregate peptide spectrum matches to
-                                 peptides\n
+aggregate_psms <-  boxGrob({glue("aggregate peptide spectrum matches to peptides\n
                                  join multiple batches",
@@
                           .sep = "\n")})
-qc2_content <- boxGrob("filter by reporter ion intensity cutoff\n
-                             filter by pepitde covariance to razor protein\n
-                             remove peptides by missing rate\n",
+qc2_content <- boxGrob(
+                      "filter by reporter ion intensity cutoff\nfilter by pepitde covariance to razor protein\nremove peptides by missing rate",
                        txt_gp = opts_box_setting,
                        box_gp = opts_box_setting,
-                       just="left"
+                       just="center"
                        )
@@
                           .sep = "\n")})
-transform_opts <- boxGrob("options:\n
-                                log2\n
-                                log10\n
-                                sqrt\n
-                                quadratic\n
-                                BoxCox\n
-                                None",
+transform_opts <- boxGrob(
+                      "options:\nlog2\nlog10\nsqrt\nquadratic\nBoxCox\nNone",
                        txt_gp = opts_box_setting,
                        box_gp = opts_box_setting,
-                       just="left"
+                       just="center"
                        )
@@
                           .sep = "\n")})
-normalize_opts <- boxGrob("options:\n
-                                SCoPE2\n
-                                CONSTANd\n
-                                None",
+normalize_opts <- boxGrob(
+                      "options:\nSCoPE2\nCONSTANd\nNone",
                        txt_gp = opts_box_setting,
                        box_gp = opts_box_setting,
-                       just="left"
+                       just="center"
                        )
@@
                           .sep = "\n")})
-missing_value_opts <- boxGrob("options:\n
-                                    KNN\n
-                                    drop rows\n
-                                    replace with mean\n
-                                    replace with median\n
-                                    replace with 0",
+missing_value_opts <- boxGrob(
+                      "options:\nKNN\ndrop rows\nreplace with mean\nreplace with median\nreplace with 0",
                        txt_gp = opts_box_setting,
                        box_gp = opts_box_setting,
-                       just="left"
+                       just="center"
                        )
@@
                           .sep = "\n")})
-batch_correction_opts <- boxGrob("options:\n
-                                       ComBat",
+batch_correction_opts <- boxGrob(
+                      "options:\nComBat",
                        txt_gp = opts_box_setting,
                        box_gp = opts_box_setting,
-                       just="left"
+                       just="center"
                        )
@@
                           .sep = "\n")})
-dim_red_actions <- boxGrob("procedures:\n
-                                 PCA\n
-                                 UMAP",
+dim_red_actions <- boxGrob(
+                      "procedures:\nPCA\nUMAP",
                        txt_gp = opts_box_setting,
                        box_gp = opts_box_setting,
-                       just="left"
+                       just="center"
                        )

### 2023-04-26 markdown/methods.Rmd
flowchart updated

@@
 opts_txt_setting <- gpar(col="darkblue", cex=1)
+refinement_arrow_setting <- arrow(angle=30, ends="both", type = "closed")
 
+plot_box_setting <- gpar(fill="#DDCC77")
 # input files
@@
 
-quality_control_2 <- boxGrob({glue("quality control 2",
-                          .sep = "\n")})
+quality_control_2 <- boxGrob("quality control 2",
+                             box_gp=plot_box_setting)
 qc2_content <- boxGrob(
@@
 
-transform_data <- boxGrob({glue("transform data",
-                          .sep = "\n")})
+transform_data <- boxGrob("transform data",
+                          box_gp=plot_box_setting)
 transform_opts <- boxGrob(
@@
 
-normalize_data <- boxGrob({glue("normalize data",
-                          .sep = "\n")})
+normalize_data <- boxGrob("normalize data",
+                          box_gp=plot_box_setting)
 normalize_opts <- boxGrob(
@@
 
-missing_value_handling <- boxGrob({glue("missing value handling",
-                          .sep = "\n")})
+missing_value_handling <- boxGrob("missing value handling",
+                                  box_gp=plot_box_setting)
 missing_value_opts <- boxGrob(
@@
 
-batch_correction <- boxGrob({glue("batch correction",
-                          .sep = "\n")})
+batch_correction <- boxGrob("batch correction",
+                            box_gp=plot_box_setting)
 batch_correction_opts <- boxGrob(
@@
 
-dim_red <- boxGrob({glue("dimensionality reduction",
-                          .sep = "\n")})
+dim_red <- boxGrob("dimensionality reduction",
+                   box_gp=plot_box_setting)
 dim_red_actions <- boxGrob(
@@
 
-connectGrob(vert$quality_control_2, qc2_content, type = "horizontal", arrow_obj = arrow(angle=0))
+connectGrob(vert$quality_control_2, qc2_content, type = "horizontal", arrow_obj = refinement_arrow_setting)
 
-connectGrob(vert$transform_data, transform_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+connectGrob(vert$transform_data, transform_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
 
-connectGrob(vert$normalize_data, normalize_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+connectGrob(vert$normalize_data, normalize_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
 
-connectGrob(vert$missing_value_handling, missing_value_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+connectGrob(vert$missing_value_handling, missing_value_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
 
-connectGrob(vert$batch_correction, batch_correction_opts, type = "horizontal", arrow_obj = arrow(angle=0))
+connectGrob(vert$batch_correction, batch_correction_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
 

### 2023-04-26 markdown/methods.Rmd
flowchart updated

@@
 
+data_box_setting <- gpar(fill="#56B4E9")
+
 opts_box_setting <- gpar(fill="lightyellow", col="darkblue")
@@
 # input files
-evidence_file <- boxGrob({glue("evidence file",
-                           .sep = "\n")})
-sample_annotation_file <- boxGrob({glue("sample annotation",
-                           .sep = "\n")})
+evidence_file <- boxGrob("evidence file",
+                         box_gp = data_box_setting)
+sample_annotation_file <- boxGrob("sample annotation",
+                                  box_gp=data_box_setting)
 

### 2023-04-26 markdown/methods.Rmd
flowchart update

@@
                     x = .8,
-                    y = coords(vert$quality_control_1)$top)
+                    y = coords(vert$quality_control_1)$y)
 
@@
                     x = .9,
-                    y = coords(vert$transform_data)$top)
+                    y = coords(vert$transform_data)$y)
 
@@
                     x = .7,
-                    y = coords(vert$missing_value_handling)$top)
+                    y = coords(vert$missing_value_handling)$y)
 
@@
                     x = .9,
-                    y = coords(vert$dim_red)$top)
+                    y = coords(vert$dim_red)$y)
 
@@
                     x = .2,
-                    y = coords(vert$quality_control_2)$top)
+                    y = coords(vert$quality_control_2)$y)
 
@@
                     x = .2,
-                    y = coords(vert$normalize_data)$top)
+                    y = coords(vert$normalize_data)$y)
 
@@
                     x = .2,
-                    y = coords(vert$batch_correction)$top)
+                    y = coords(vert$batch_correction)$y)
 

### 2023-04-26 markdown/methods.Rmd
methods --> transformation further ellaborated

@@
 Similar to the already explained previous aggregation step the peptides will be further processed into their respective proteins after the quality control on the peptide level is performed.
+Finally an expression matrix for every protein and their respective channel column is returned.
+Each channel refers to a sample and reflects contains the intensities of each protein found in the biological specimen.
 
@@
 $$
-Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill dependencies for statistical testing.
-In the shiny application various procedures are implemented and can be further expanded upon request from the user.
-The macrophage analysis done by Specht et.
-Al mentioned in their SCoPE2 publication citep\{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and the remove skewness in the dataset.
+Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill assumptions for statistical testing.
+In the shiny application, various procedures are implemented and can be further expanded upon request from the user.
+The macrophage analysis done by Specht et al.
+mentioned in their SCoPE2 publication \citep{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and remove skewness in the dataset.
 This transformation was used as a reference when comparing methods.
-However using the logarithm to the base 10 can be easy interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method which is also implemented in the application.
-The boxcox transformation developed 1964 by Box and Cox citep\{Sakia1992} applies a recursion on the data by determining the parameter lambda.
-Depending on the size of lambda a certain function will be applied on each value of the expression set in order to introduce normality.
-As already mentioned transformations applied by the BoxCox method range from log10 to taking the square root of each value
+However, using the logarithm to the base 10 may be an easily interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
+When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
+Practically impossible, but in data-driven science, occasionally negative expression values may be present, and they can be converted into positive values by taking each one to the power of two.
+Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qqplots, and MA-plots.
+For statistically inexperienced users, the boxcox transformation can help with this decision.
+Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a recursion on the data to determine the statistical parameter lambda.
+
+Depending on the size of lambda, a certain function will be automatically applied to each value of the expression set in order to introduce normality.
+In terms of usability, the boxcox method is the most efficient transformation method, taking the decision of which calculation to apply away from the user.
 

### 2023-04-26 scpR-app/app.R

@@ -290,7 +290,7 @@ server <- function(input, output, session) {
       MedCV_thresh <- input$MedCV_thresh
       scp_0 <- scp_0[, !is.na(scp_0$MedianCV) & scp_0$MedianCV < MedCV_thresh, ]
 
-      incProgress(12/17, detail=paste("remove peptides with high missing rate"))
+      incProgress(12/17, detail=paste("remove peptides by missing rate"))
       req(input$pNA)
       pNA <- input$pNA
       
@@ -1113,7 +1113,6 @@ server <- function(input, output, session) {
   # hover output below volcano chart
   output$hover_info <- renderPrint({
     req(displayed_text())
-    
     cat("UniProt-ID\n")
     displayed_text()
   })

### 2023-04-28 markdown/methods.Rmd
flowchart updated

@@
 ## Data
-```{r Dataflow, eval=FALSE, include=FALSE}
-# library(DiagrammeR)
-# library(webshot)
-# 
-# DiagrammeR::grViz("digraph {
-# 	graph [layout  =  dot,
-# 	       rankdir  = TB]
-# 
-# node [shape = rectangle]
-# rawData [label = 'msData.raw']
-# node [shape = oval]
-# MQ [label = 'MaxQuant']
-# node [shape = invhouse]
-# MQ_fd [label = 'feature detection']
-# node [shape = cylinder]
-# MQ_ds [label = 'Database search']
-# node [shape = invhouse]
-# MQ_fdr [label = 'FDR calculation']
-# MQ_quant [label = 'Protein Quantification']
-# node [shape = rectangle]
-# MQoutput [label = 'evidence.txt']
-# sampleAnnotate [label = 'sample_annotation.txt']
-# node [shape = oval]
-# ProteomicsWorkbench [label = 'Proteomics Workbench']
-# 
-# rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
-# sampleAnnotate -> ProteomicsWorkbench
-# }",
-# height = 500)
+```{r Dataflow, echo=FALSE}
+library(DiagrammeR)
+library(webshot)
+
+DiagrammeR::grViz("digraph {
+	graph [layout  =  dot,
+	       rankdir  = TB]
+
+node [shape = rectangle]
+rawData [label = 'msData.raw']
+node [shape = oval]
+MQ [label = 'MaxQuant']
+node [shape = invhouse]
+MQ_fd [label = 'feature detection']
+node [shape = cylinder]
+MQ_ds [label = 'Database search']
+node [shape = invhouse]
+MQ_fdr [label = 'FDR calculation']
+MQ_quant [label = 'Protein Quantification']
+node [shape = rectangle]
+MQoutput [label = 'evidence.txt']
+sampleAnnotate [label = 'sample_annotation.txt']
+node [shape = oval]
+ProteomicsWorkbench [label = 'Proteomics Workbench']
+
+rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
+sampleAnnotate -> ProteomicsWorkbench
+}",
+height = 500)
 ```
@@
 
+default_txt_setting <- gpar(cex=1.9)
+default_box_setting <- gpar(fill="white")
+
 data_box_setting <- gpar(fill="#56B4E9")
@@
 opts_box_setting <- gpar(fill="lightyellow", col="darkblue")
-opts_txt_setting <- gpar(col="darkblue", cex=1)
+opts_txt_setting <- gpar(col="darkblue", cex=1.5)
 refinement_arrow_setting <- arrow(angle=30, ends="both", type = "closed")
@@
 evidence_file <- boxGrob("evidence file",
-                         box_gp = data_box_setting)
+                         box_gp = data_box_setting,
+                         txt_gp = default_txt_setting)
 sample_annotation_file <- boxGrob("sample annotation",
-                                  box_gp=data_box_setting)
+                                  box_gp = data_box_setting,
+                                  txt_gp = default_txt_setting)
 
-read_data <- boxGrob({glue("read data into q-feature object",
-                          .sep = "\n")})
+read_data <- boxGrob(glue("read data into q-feature object",.sep = "\n"),
+                     box_gp = default_box_setting,
+                     txt_gp = default_txt_setting)
 
-quality_control_1 <- boxGrob({glue("quality control 1",
-                          .sep = "\n")})
-qc1_content <- boxGrob(
-                      "filter out pool samples\nreplace zeros with NA\nfilter out contaminants\nfilter by q-value cutoff",
-                       txt_gp = opts_box_setting,
+quality_control_1 <- boxGrob(glue("quality control 1",.sep = "\n"),
+                             box_gp = default_box_setting,
+                             txt_gp = default_txt_setting)
+
+qc1_content <- boxGrob("filter out pool samples\nreplace zeros with NA\nfilter out contaminants\nfilter by q-value cutoff",
                       box_gp = opts_box_setting,
-                       just="center",
-                       )
+                      txt_gp = opts_txt_setting,
+                      just="center")
 
-aggregate_psms <-  boxGrob({glue("aggregate peptide spectrum matches to peptides\n
-                                 join multiple batches",
-                          .sep = "\n")})
+aggregate_psms <-  boxGrob(glue("aggregate peptide spectrum matches to peptides\n
+                                 join multiple batches", .sep = "\n"),
+                           box_gp = default_box_setting,
+                           txt_gp = default_txt_setting)
 
 quality_control_2 <- boxGrob("quality control 2",
-                             box_gp=plot_box_setting)
-qc2_content <- boxGrob(
-                      "filter by reporter ion intensity cutoff\nfilter by pepitde covariance to razor protein\nremove peptides by missing rate",
-                       txt_gp = opts_box_setting,
+                             box_gp = plot_box_setting,
+                             txt_gp = default_txt_setting)
+
+qc2_content <- boxGrob("filter by:\nreporter ion intensity cutoff\npepitde covariance to razor proteins\npeptide missing rate",
                        box_gp = opts_box_setting,
+                       txt_gp = opts_txt_setting,
+                       just="center")
+
+aggregate_peps <- boxGrob(glue("aggregate peptides to proteins", .sep = "\n"),
+                          box_gp = default_box_setting,
+                          txt_gp = default_txt_setting)
+
+transform_data <- boxGrob("transform data",
+                          box_gp = plot_box_setting,
+                          txt_gp = default_txt_setting)
+
+transform_opts <- boxGrob("options:\nlog2\nlog10\nsqrt\nquadratic\nBoxCox\nNone",
+                          box_gp = opts_box_setting,
+                          txt_gp = opts_txt_setting,
+                          just="center")
+
+normalize_data <- boxGrob("normalize data",
+                          box_gp = plot_box_setting,
+                          txt_gp = default_txt_setting)
+
+normalize_opts <- boxGrob("options:\nSCoPE2\nCONSTANd\nNone",
+                          box_gp = opts_box_setting,
+                          txt_gp = opts_txt_setting,
+                          just="center")
+
+missing_value_handling <- boxGrob("missing value handling",
+                                  box_gp = plot_box_setting,
+                                  txt_gp = default_txt_setting)
+
+missing_value_opts <- boxGrob("options:\nKNN\ndrop rows\nreplace with mean\nreplace with median\nreplace with 0",
+                              box_gp = opts_box_setting,
+                              txt_gp = opts_txt_setting,
+                              just="center")
+
+batch_correction <- boxGrob("batch correction",
+                            box_gp=plot_box_setting,
+                            txt_gp = default_txt_setting)
+
+batch_correction_opts <- boxGrob("options:\nComBat",
+                                 box_gp = opts_box_setting,
+                                 txt_gp = opts_txt_setting,
+                                 just="center")
+
+dim_red <- boxGrob("dimensionality reduction",
+                   box_gp = plot_box_setting,
+                   txt_gp = default_txt_setting)
+
+dim_red_actions <- boxGrob("procedures:\nPCA\nUMAP",
+                       box_gp = opts_box_setting,
+                       txt_gp = opts_txt_setting,
                        just="center"
@@
 
-aggregate_peps <- boxGrob({glue("aggregate peptides to proteins",
-                          .sep = "\n")})
-
-transform_data <- boxGrob("transform data",
-                          box_gp=plot_box_setting)
-transform_opts <- boxGrob(
-                      "options:\nlog2\nlog10\nsqrt\nquadratic\nBoxCox\nNone",
-                       txt_gp = opts_box_setting,
-                       box_gp = opts_box_setting,
-                       just="center"
-                       )
-
-normalize_data <- boxGrob("normalize data",
-                          box_gp=plot_box_setting)
-normalize_opts <- boxGrob(
-                      "options:\nSCoPE2\nCONSTANd\nNone",
-                       txt_gp = opts_box_setting,
-                       box_gp = opts_box_setting,
-                       just="center"
-                       )
-
-missing_value_handling <- boxGrob("missing value handling",
-                                  box_gp=plot_box_setting)
-missing_value_opts <- boxGrob(
-                      "options:\nKNN\ndrop rows\nreplace with mean\nreplace with median\nreplace with 0",
-                       txt_gp = opts_box_setting,
-                       box_gp = opts_box_setting,
-                       just="center"
-                       )
-
-batch_correction <- boxGrob("batch correction",
-                            box_gp=plot_box_setting)
-batch_correction_opts <- boxGrob(
-                      "options:\nComBat",
-                       txt_gp = opts_box_setting,
-                       box_gp = opts_box_setting,
-                       just="center"
-                       )
-
-dim_red <- boxGrob("dimensionality reduction",
-                   box_gp=plot_box_setting)
-dim_red_actions <- boxGrob(
-                      "procedures:\nPCA\nUMAP",
-                       txt_gp = opts_box_setting,
-                       box_gp = opts_box_setting,
-                       just="center"
-                       )
-
-stat_module <- boxGrob({glue("statistics module",
-                          .sep = "\n")})
+stat_module <- boxGrob(glue("statistics module",.sep = "\n"),
+                       box_gp = default_box_setting,
+                       txt_gp = default_txt_setting)
 
@@
 missing_value_opts <- moveBox(missing_value_opts,
-                    x = .7,
+                    x = .85,
                     y = coords(vert$missing_value_handling)$y)
@@
 qc2_content <- moveBox(qc2_content,
-                    x = .2,
+                    x = .17,
                     y = coords(vert$quality_control_2)$y)
@@
 connectGrob(vert$dim_red, dim_red_actions, type = "horizontal", arrow_obj = arrow(angle=0))
+
+# Print boxes
+input
+vert
+qc1_content
+qc2_content
+transform_opts
+normalize_opts
+missing_value_opts
+batch_correction_opts
+dim_red_actions
+```
+
+```{r data_processing_pipeline_flowchart2, echo=FALSE, fig.height=9, fig.width=32}
+library(Gmisc, quietly = TRUE)
+library(glue)
+library(htmlTable)
+library(grid)
+library(magrittr)
+
+default_txt_setting <- gpar(cex=1.5)
+default_box_setting <- gpar(fill="white")
+
+data_box_setting <- gpar(fill="#56B4E9")
+
+opts_box_setting <- gpar(fill="lightyellow", col="darkblue")
+opts_txt_setting <- gpar(col="darkblue", cex=1)
+refinement_arrow_setting <- arrow(angle=30, ends="both", type = "closed")
+
+plot_box_setting <- gpar(fill="#DDCC77")
+# input files
+evidence_file <- boxGrob("evidence file",
+                         box_gp = data_box_setting,
+                         txt_gp = default_txt_setting)
+sample_annotation_file <- boxGrob("sample annotation",
+                                  box_gp = data_box_setting,
+                                  txt_gp = default_txt_setting)
+
+read_data <- boxGrob(glue("read data into q-feature object",.sep = "\n"),
+                     box_gp = default_box_setting,
+                     txt_gp = default_txt_setting)
+
+quality_control_1 <- boxGrob(glue("quality control 1",.sep = "\n"),
+                             box_gp = default_box_setting,
+                             txt_gp = default_txt_setting)
+
+qc1_content <- boxGrob("filter out pool samples\nreplace zeros with NA\nfilter out contaminants\nfilter by q-value cutoff",
+                      box_gp = opts_box_setting,
+                      txt_gp = opts_txt_setting,
+                      just="center")
+
+aggregate_psms <-  boxGrob(glue("aggregate peptide spectrum matches to peptides\n
+                                 join multiple batches", .sep = "\n"),
+                           box_gp = default_box_setting,
+                           txt_gp = default_txt_setting)
+
+quality_control_2 <- boxGrob("quality control 2",
+                             box_gp = plot_box_setting,
+                             txt_gp = default_txt_setting)
+
+qc2_content <- boxGrob("filter by:\nreporter ion intensity cutoff\npepitde covariance to razor proteins\npeptide missing rate",
+                       box_gp = opts_box_setting,
+                       txt_gp = opts_txt_setting,
+                       just="center")
+
+aggregate_peps <- boxGrob(glue("aggregate peptides to proteins", .sep = "\n"),
+                          box_gp = default_box_setting,
+                          txt_gp = default_txt_setting)
+
+transform_data <- boxGrob("transform data",
+                          box_gp = plot_box_setting,
+                          txt_gp = default_txt_setting)
+
+transform_opts <- boxGrob("options:\nlog2\nlog10\nsqrt\nquadratic\nBoxCox\nNone",
+                          box_gp = opts_box_setting,
+                          txt_gp = opts_txt_setting,
+                          just="center")
+
+normalize_data <- boxGrob("normalize data",
+                          box_gp = plot_box_setting,
+                          txt_gp = default_txt_setting)
+
+normalize_opts <- boxGrob("options:\nSCoPE2\nCONSTANd\nNone",
+                          box_gp = opts_box_setting,
+                          txt_gp = opts_txt_setting,
+                          just="center")
+
+missing_value_handling <- boxGrob("missing value handling",
+                                  box_gp = plot_box_setting,
+                                  txt_gp = default_txt_setting)
+
+missing_value_opts <- boxGrob("options:\nKNN\ndrop rows\nreplace with mean\nreplace with median\nreplace with 0",
+                              box_gp = opts_box_setting,
+                              txt_gp = opts_txt_setting,
+                              just="center")
+
+batch_correction <- boxGrob("batch correction",
+                            box_gp=plot_box_setting,
+                            txt_gp = default_txt_setting)
+
+batch_correction_opts <- boxGrob("options:\nComBat",
+                                 box_gp = opts_box_setting,
+                                 txt_gp = opts_txt_setting,
+                                 just="center")
+
+dim_red <- boxGrob("dimensionality reduction",
+                   box_gp = plot_box_setting,
+                   txt_gp = default_txt_setting)
+
+dim_red_actions <- boxGrob("procedures:\nPCA\nUMAP",
+                       box_gp = opts_box_setting,
+                       txt_gp = opts_txt_setting,
+                       just="center"
+                       )
+
+stat_module <- boxGrob(glue("statistics module",.sep = "\n"),
+                       box_gp = default_box_setting,
+                       txt_gp = default_txt_setting)
+
+grid.newpage()
+
+vert <- {spreadHorizontal(files = evidence_file,
+                       read_data = read_data,
+                       quality_control_1 = quality_control_1,
+                       aggregate_psms,
+                       quality_control_2 = quality_control_2,
+                       aggregate_peps,
+                       transform_data = transform_data,
+                       normalize_data = normalize_data,
+                       missing_value_handling = missing_value_handling,
+                       batch_correction = batch_correction,
+                       dim_red = dim_red,
+                       stat_module
+)}
+
+input <- alignHorizontal(reference = vert$files,
+                      evidence_file, sample_annotation_file) %>%
+  spreadVertical()
+vert$files <- NULL
+
+# connect and move the boxes
+for (i in 1:(length(vert) - 1)) {
+    connectGrob(vert[[i]], vert[[i + 1]], type = "horizontal") %>%
+    print}
+
+# right side
+qc1_content <- moveBox(qc1_content,
+                    x = coords(vert$quality_control_1)$x,
+                    y = 0.8)
+
+transform_opts <- moveBox(transform_opts,
+                    y = .9,
+                    x = coords(vert$transform_data)$x)
+
+missing_value_opts <- moveBox(missing_value_opts,
+                    y = .7,
+                    x = coords(vert$missing_value_handling)$x)
+
+dim_red_actions <- moveBox(dim_red_actions,
+                    y = .9,
+                    x = coords(vert$dim_red)$x)
+
+# left side
+qc2_content <- moveBox(qc2_content,
+                    y = .2,
+                    x = coords(vert$quality_control_2)$x)
+
+normalize_opts <- moveBox(normalize_opts,
+                    y = .2,
+                    x = coords(vert$normalize_data)$x)
+
+batch_correction_opts <- moveBox(batch_correction_opts,
+                    y = .2,
+                    x = coords(vert$batch_correction)$x)
+
+# connect boxes
+connectGrob(input[[1]], vert$read_data, type = "vertical")
+connectGrob(input[[2]], vert$read_data, type = "vertical")
+
+connectGrob(vert$quality_control_1, qc1_content, type = "vertical", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$quality_control_2, qc2_content, type = "vertical", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$transform_data, transform_opts, type = "vertical", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$normalize_data, normalize_opts, type = "vertical", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$missing_value_handling, missing_value_opts, type = "vertical", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$batch_correction, batch_correction_opts, type = "vertical", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$dim_red, dim_red_actions, type = "vertical", arrow_obj = arrow(angle=0))
 

### 2023-04-28 scope2.R

@@ -22,30 +22,29 @@ library("CONSTANd")
 library("stats")
 library("impute")
 library("sva")
+library("tibble")
 
 # read in MS result table
 mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")
-
-# create annotation file
-# this varies upon experimental design
-#quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData), value = T)
-
-
-# create sample file
-# this needs to be done by the researcher
-# sampleAnnotation <- as.data.frame(quantCols)
-# colnames(sampleAnnotation) <-c("Channel")
-# sampleAnnotation$Raw.file = unique(mqScpData$Raw.file)
-# 
-# 
-# samplesA <- c(replicate(6, "Monocytes_A"))
-# samplesB <- c(replicate(6, "Monocytes_B"))
-# samples <- c(samplesA, samplesB)
-# 
-# 
-# sampleAnnotation$SampleType <- samples
+mqScpData2 <- read.delim("/home/lukas/Downloads/Raw_data.txt")
 
 sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/sampleAnnotation_tabdel.txt")
+sampleAnnotation2 = read.delim("/home/lukas/Downloads/Design.txt")
+
+
+# define file handling for not mq generated data
+
+#metadata is used as a reference --> change different metadata file accordingly
+if (!"Sequence" %in% colnames(mqScpData2)) {
+  sampleAnnotation2 <- 
+    sampleAnnotation2 %>% rename(
+    SampleType = Group,
+    Raw.File = Batch
+  )
+  sampleAnnotation2 <- sampleAnnotation2[order(sampleAnnotation2$Raw.File),]
+  sampleAnnotation2$Channel <- sampleAnnotation2
+}
+
 
 # create QFeature object
 scp <- readSCP(featureData = mqScpData,
@@ -54,6 +53,30 @@ scp <- readSCP(featureData = mqScpData,
                batchCol = "Raw.file",
                suffix = paste0("_TMT", 1:length(unique(sampleAnnotation$Channel))),
                removeEmptyCols = TRUE)
+
+
+if (!"Sequence" %in% colnames(mqScpData2)) {
+  mqScpData2 <- 
+    mqScpData2 %>% rename(
+    Protein = ID
+  )
+  
+  mqScpData2 %>% 
[…]

### 2023-05-02 scope2.R
replace with 0 from batchC deleted

@@ -26,10 +26,11 @@ library("tibble")
 
 # read in MS result table
 mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")
-mqScpData2 <- read.delim("/home/lukas/Downloads/Raw_data.txt")
+mqScpData2 <- read.csv("/home/lukas/Downloads/Raw_data.csv")
+mqScpData3 <- readxl::read_excel("/home/lukas/Downloads/20230207_L1_UN_Monocytes_TMT12_frac.xlsx")
 
 sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/sampleAnnotation_tabdel.txt")
-sampleAnnotation2 = read.delim("/home/lukas/Downloads/Design.txt")
+sampleAnnotation2 = read.csv("/home/lukas/Downloads/Design.csv")
 
 
 # define file handling for not mq generated data
@@ -39,10 +40,8 @@ if (!"Sequence" %in% colnames(mqScpData2)) {
   sampleAnnotation2 <- 
     sampleAnnotation2 %>% rename(
     SampleType = Group,
-    Raw.File = Batch
-  )
-  sampleAnnotation2 <- sampleAnnotation2[order(sampleAnnotation2$Raw.File),]
-  sampleAnnotation2$Channel <- sampleAnnotation2
+    Raw.file = Batch) %>%
+    arrange(Raw.file)
 }
 
 
@@ -60,21 +59,18 @@ if (!"Sequence" %in% colnames(mqScpData2)) {
     mqScpData2 %>% rename(
     Protein = ID
   )
-  
-  mqScpData2 %>% 
-    rename_with(~deframe(sampleAnnotation2)[.x], .cols = sampleAnnotation2$Channel) %>% 
-    select(Protein, any_of(sampleAnnotation2$Channel))
-  
-  
-  scp2 <- readSCP(featureData = mqScpData2,
-                  colData = sampleAnnotation2,
-                  channelCol = "Channel",
-                  batchCol = "Raw.file",
-                  removeEmptyCols = TRUE,
-                  verbose = T)
+}  
 
-  
-}
+
+quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData), value = TRUE)
+
+scp2 <- readSCP(featureData = mqScpData2,
+                 colData = sampleAnnotation2,
+                 channelCol = "Channel",
[…]

### 2023-05-02 scpR-app/app.R

@@ -54,7 +54,7 @@ ui <- fluidPage(
       selectInput("norm_method", "Choose method for protein data normalization", choices = c("SCoPE2", "None", "CONSTANd")),
       switchInput(inputId = "opts_multiple_batches", onLabel = "Advanced", offLabel = "Default", value = F, label="Options for multiple batches"),
       conditionalPanel(condition = "input.opts_multiple_batches", 
-                       selectInput(inputId = "missing_v", "Choose method for missing value handling", choices=c("KNN", "drop rows", "replace with 0", "replace with mean", "replace with median")),
+                       selectInput(inputId = "missing_v", "Choose method for missing value handling", choices=c("KNN", "drop rows", "replace with mean", "replace with median")),
                        selectInput(inputId = "batch_c", "Choose method for batch correction", choices=c("ComBat", "none"))
                        )
       
@@ -494,19 +494,6 @@ server <- function(input, output, session) {
                                         to = "proteins_imptd")
           
           scp_0 <- filterNA(scp_0, pNA = 0, "proteins_imptd")
-        } else if (input$missing_v == "replace with 0") {
-          sce <- getWithColData(scp_0, "proteins_norm")
-          
-          scp_0 <- addAssay(scp_0,
-                          y = sce,
-                          name = "proteins_imptd")
-          
-          scp_0 <- addAssayLinkOneToOne(scp_0,
-                                      from = "proteins_norm",
-                                      to = "proteins_imptd")
-          
-          assay(scp_0[["proteins_imptd"]]) <- replace(assay(scp_0[["proteins_imptd"]]), is.na(assay(scp_0[["proteins_imptd"]])), 0)
-          
         } else if (input$missing_v == "replace with mean") {
           sce <- getWithColData(scp_0, "proteins_norm")
           

### 2023-05-02 scpR-app/app.R
refactoring
sampletype exclusion

@@ -35,6 +35,8 @@ ui <- fluidPage(
       fileInput("sample_annotation_file", "Upload sample annotation file", accept = c("text")),
       textInput("label_suffix", "Input the label suffix (e.g. _TMT)"),
       
+      selectInput("selectedSampleType_to_exclude", "Select SampleType to exclude", "", multiple=T),
+      
       # cutoff for PIF
       numericInput("PIF_cutoff", "Input cutoff value for parental ion fraction. PSMs larger than the value remain", 0.1, min = 0, max=1, step = 0.1),
       
@@ -109,7 +111,7 @@ ui <- fluidPage(
   )
 )
 
-# box cox transf
+# box cox transformati0n --> alteration of basic MASS package was needed to run it
 boxcox_1 <- function(object, ...) UseMethod("boxcox_1")
 
 boxcox_1.formula <-
@@ -162,517 +164,7 @@ boxcox_1.default <-
 # Define server logic required to draw a histogram
 server <- function(input, output, session) {
   options(shiny.maxRequestSize=30*1024^2)
-  
-  ## Data structures and objects
-  # sample_annotation needs to be accessible for plots
-  meta_data <- reactive({
-    req(input$sample_annotation_file)
-    meta_data_0 <- read.delim(input$sample_annotation_file$datapath)
-    return(meta_data_0)
-  })
-  
-  # result of analysis pipleline
-  scp <- eventReactive(input$update_button, {
-    withProgress(message= "running analysis", value=0, {
-      req(input$evidence_file)
-      req(meta_data)
-      req(input$label_suffix)
-      
-      evidence_data <- read.delim(input$evidence_file$datapath)
-      meta_data_0 <- meta_data()
-      
-      incProgress(1/17, detail = paste("read Data"))
-      scp_0 <- readSCP(featureData = evidence_data,
-                       colData = meta_data_0,
-                       channelCol = "Channel",
-                       batchCol = "Raw.file",
-                       suffix = paste0(input$label_suffix, 1:length(unique(meta_data_0$Channel))),
-                       removeEmptyCols = TRUE)
-      
-      # skip pool sample
-      if ("Pool" %in% meta_data_0$SampleType) {
-        scp_0 <- scp_0[, scp_0$SampleType !=  "Pool"]
-      }
-      
-      # # change zeros to NA, apply first filter
-      incProgress(2/17, detail = paste("replacing zeros with NA"))
[…]

### 2023-05-02 scpR-app/app.R
color the volcano started

@@ -734,8 +734,6 @@ server <- function(input, output, session) {
     return(list)
   })
   
-  
-  
   # observer for sample type exclusion
   observe({
     req(sample_types())

### 2023-05-02 markdown/Thesis.tex
normalization started to ellaborate

@@
 
-in partial fulfilment of the requirement \\for the degree of MSc \\in Bioinformatics
+in partial fullfilment of the requirement \\for the degree of MSc \\in Bioinformatics
 
@@
 
-\hypertarget{instrumentation}{%
-\subsubsection{Instrumentation}\label{instrumentation}}
+\pagenumbering{arabic}
+
+\newpage
+
+\hypertarget{materials-and-methods}{%
+\section{Materials and Methods}\label{materials-and-methods}}
+
+\hypertarget{materials}{%
+\subsection{Materials}\label{materials}}
+
+For analysis two types of cells were used. One type is the Jurkat-based
+cell line (J-lat) with integrated HIV.
+
+The other type of cells are macrophages with a sample size of 72 cells.
+The analysis is done with two groups. A HIV negative (HIV-) control
+group and a HIV positive (HIV+) group.
 
 \hypertarget{cell-isolation}{%
-\paragraph{Cell Isolation}\label{cell-isolation}}
+\subsubsection{Cell Isolation}\label{cell-isolation}}
 
 \hypertarget{lysis}{%
-\paragraph{Lysis}\label{lysis}}
+\subsubsection{Lysis}\label{lysis}}
 
 \hypertarget{digestion}{%
-\paragraph{Digestion}\label{digestion}}
-
-\hypertarget{liquid-chromatography}{%
-\paragraph{Liquid chromatography}\label{liquid-chromatography}}
-
-In order to separate proteins according to their chemical properties,
-size or species a liquid chromatography (LC) is recommended before
-ionization.
-
-\hypertarget{ionization}{%
-\paragraph{Ionization}\label{ionization}}
-
-In order to analyze a biological sample consisting of proteins in
-solution the liquid needs to be vaporized into gas phase. Two techniques
-are capable of this procedure. Electrospray ionization (ESI) pushes the
-analyte through a capillary and applies an electric current to the
-liquid, vaporizing the sample to a charged aerosol. Biomolecules are
-fragmented according to their chemical properties and can be further
-handled in the mass spectrometer. The fragmented biomolecules are now in
-charged droplets separated by their charge on the surface, splitting
-further into smaller droplets until they become a gas phase ion. Two
-physical models describe the process from gas phase to ion called ``The
-ion evaporation model'' (IEM) and ``The charge residue model'' (CRM). In
-the ion evaporation model (by Iribarne and Thomoson) the droplets shrink
-by evaporation until ions are expelled. The model had its limitation by
-explaining same evaporation rate constant among ions with different
-chemical properties. In the charge residue model the assumption of one
-molecule per droplet leads to an ionization rate constant, which is
-independent of the ion itself and relies solely on the generation of the
-droplet and the efficiency of the solvent \citep{Wilm2011}.
-
-Matrix-assisted laser desorption/ionization (MALDI)
-
-\hypertarget{coupled-mass-spectrometry-msms}{%
-\paragraph{Coupled mass-spectrometry
-(MS/MS)}\label{coupled-mass-spectrometry-msms}}
-
-In order to enhance sequence identification, two MS devices are built in
-series. In the first run (MS1) the m/z is determined and the molecules
-are passed to the next device. Upon passing the molecules are fragmented
-into smaller ions and analyzed by the second MS. The fragmentation
-highly depends on the chemical bonds found in the molecule. The majority
-of these breaks occur on the peptide bond of the protein, although this
-is not guaranteed for all bonds and so it can happen that certain
-peptide ions have a low abundance \citep{Budnik2018}. These low abundant
-peptides will not be detected, hence the problem needs to be faced with
-another approach . A solution for this problem is molecular barcoding
-with labeling mentioned in the chapter labeling.
+\subsubsection{Digestion}\label{digestion}}
 
 \hypertarget{labeling-techniques}{%
-\paragraph{Labeling techniques}\label{labeling-techniques}}
+\subsubsection{Labeling techniques}\label{labeling-techniques}}
 
@@
 
+\hypertarget{instrumentation}{%
+\subsubsection{Instrumentation}\label{instrumentation}}
+
+\hypertarget{liquid-chromatography}{%
+\paragraph{Liquid chromatography}\label{liquid-chromatography}}
+
+In order to separate proteins according to their chemical properties,
+size or species a liquid chromatography (LC) is recommended before
+ionization.
+
+\hypertarget{mass-spectrometry-1}{%
+\paragraph{Mass Spectrometry}\label{mass-spectrometry-1}}
+
+\hypertarget{ionization}{%
+\subparagraph{Ionization}\label{ionization}}
+
+In order to analyze a biological sample consisting of proteins in
+solution the liquid needs to be vaporized into gas phase. Two techniques
+are capable of this procedure. Electrospray ionization (ESI) pushes the
+analyte through a capillary and applies an electric current to the
+liquid, vaporizing the sample to a charged aerosol. Biomolecules are
+fragmented according to their chemical properties and can be further
+handled in the mass spectrometer. The fragmented biomolecules are now in
+charged droplets separated by their charge on the surface, splitting
+further into smaller droplets until they become a gas phase ion. Two
+physical models describe the process from gas phase to ion called ``The
+ion evaporation model'' (IEM) and ``The charge residue model'' (CRM). In
+the ion evaporation model (by Iribarne and Thomoson) the droplets shrink
+by evaporation until ions are expelled. The model had its limitation by
+explaining same evaporation rate constant among ions with different
+chemical properties. In the charge residue model the assumption of one
+molecule per droplet leads to an ionization rate constant, which is
+independent of the ion itself and relies solely on the generation of the
+droplet and the efficiency of the solvent \citep{Wilm2011}.
+
+Matrix-assisted laser desorption/ionization (MALDI)
+
+\hypertarget{ms.1}{%
+\subparagraph{MS.1}\label{ms.1}}
+
+\hypertarget{coupled-mass-spectrometry-msms-ms.2}{%
+\subparagraph{Coupled mass-spectrometry (MS/MS) \&
+MS.2}\label{coupled-mass-spectrometry-msms-ms.2}}
+
+In order to enhance sequence identification, two MS devices are built in
+series. In the first run (MS1) the m/z is determined and the molecules
+are passed to the next device. Upon passing the molecules are fragmented
+into smaller ions and analyzed by the second MS. The fragmentation
+highly depends on the chemical bonds found in the molecule. The majority
+of these breaks occur on the peptide bond of the protein, although this
+is not guaranteed for all bonds and so it can happen that certain
+peptide ions have a low abundance \citep{Budnik2018}. These low abundant
+peptides will not be detected, hence the problem needs to be faced with
+another approach . A solution for this problem is molecular barcoding
+with labeling mentioned in the chapter labeling.
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+\hypertarget{acquisition}{%
+\subsubsection{Acquisition}\label{acquisition}}
+
+Acquisition of the data was done with MaxQuant \citep{Cox2008} software
+package.
+
 \hypertarget{ms-spectrum}{%
-\subsubsection{MS-Spectrum}\label{ms-spectrum}}
+\paragraph{MS-Spectrum}\label{ms-spectrum}}
 
@@
 random event, are multiplied and only distinct sequences with the
-highst-scoring are used.
+highest-scoring are used.
 
 \hypertarget{data-processing}{%
-\paragraph{Data processing}\label{data-processing}}
+\subsubsection{Data processing}\label{data-processing}}
 
-Further analysis can be done with R and respective packages such as
+Further analysis is done with R and respective packages such as
 bioconductor. Since there is no state of the art established, analysis
-varies upon experimental design.
+varies upon experimental design. The workflow of the analysis will be
+processed in a so called pipeline streamlining the data through steps
+where individual results can be observed in visualizations and indivdual
+calculations will be adapted according to user demands and experimental
+properties.
+
+\includegraphics{Thesis_files/figure-latex/data_processing_pipeline_flowchart_vertical-1.pdf}
 
 \hypertarget{reading-the-data}{%
-\subparagraph{Reading the data}\label{reading-the-data}}
+\paragraph{Reading the data}\label{reading-the-data}}
 
 After processing MaxQuant creates a directory containing all results as
-.txt file. The evidence.txt file include all peptide sequence matches
+.txt file. The evidence.txt file include all peptide to spectrum matches
 (PSM) with their respective proteins and statistical parameters.
@@
 accepting an FDR of an individual PSM is described as q-value.
+
+\hypertarget{peptide-spectrum-match-psm-aggregation-to-peptides}{%
+\paragraph{Peptide spectrum match (PSM) aggregation to
+peptides}\label{peptide-spectrum-match-psm-aggregation-to-peptides}}
+
+In data science, aggregation refers to a row-wise operation that merges
+data based on a particular column using a specific function. In the
+context of processing from peptides to spectrum matches, the desired
+column is the peptide sequence. To account for different distributions
+across multiple assays, the median of the channel is used as the
+function to aggregate multiple matches into one.
+
+\hypertarget{join-assays-when-observing-multiple-comparable-batches-at-once}{%
+\paragraph{Join assays when observing multiple comparable batches at
+once}\label{join-assays-when-observing-multiple-comparable-batches-at-once}}
+
+Sample size is often a limiting factor in hypothesis testing. A strict
+quality control and the fact that TMT reagents are only available up to
+18-plex can reduce the number of observed samples below the critical
+threshold, leading to an early end of analyses. To overcome this
+limitation, the provided software is capable of processing multiple runs
+simultaneously, allowing for testing of multiple batches and increasing
+the number of samples that can be included in the analysis.
+
+\hypertarget{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}{%
+\paragraph{Calculate reporter ion intensity (RI) and filter according to
+median
+RI}\label{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}}
+
+Columns which do not meet the desired intensity can be filtered by a
+threshold set on the RI. The median RI can also be used to check if an
+entire channel has a lower detection level. This can be due two reasons.
+One is the expression level of the given proteins in a cell. Meaning,
+that the expression of the observed cell type is simply lower than the
+other type. Another one could be a spillage of TMT detection in other
+channels due incorrect or missing correction of the TMT isotopes.
+
+\hypertarget{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}{%
+\paragraph{Calculate and filter according to median coefficient of
+variation (CV) per
+cell/channel}\label{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}}
+
+Depending having a bulk sample or single-cell sample choosing a minimum
+of observed peptides and a cutoff value for the CV, changes the level of
+confidence in the peptide data. The coefficient of variation of a
+peptide is considered as the ratio of the standard deviation to the mean
+and describes the relationship of the observed peptide signal over
+multiple proteins (=razor proteins). Peptides having a high coefficient
+of variation over many razor proteins are considered as noise and need
+to be filtered out before statistical analysis.
+
+\hypertarget{remove-peptides-with-high-missing-rate}{%
+\paragraph{Remove peptides with high missing
+rate}\label{remove-peptides-with-high-missing-rate}}
+
+Although missing value imputation can be performed during the analysis
+of multiple batches, peptides with missing detections across channels
+can be problematic for quantification. The proteomic composition of a
+biological sample is similar between replicates and even across groups.
+However, the threshold of missingness (described as a fraction of the
+row) can be set in the user interface and adjusted to enable different
+experimental designs.
+
+\hypertarget{aggregation-of-peptides-to-proteins}{%
+\paragraph{Aggregation of peptides to
+proteins}\label{aggregation-of-peptides-to-proteins}}
+
+Similar to the already explained previous aggregation step the peptides
+will be further processed into their respective proteins after the
+quality control on the peptide level is performed. Finally an expression
+matrix for every protein and their respective channel column is
+returned. Each channel refers to a sample and reflects contains the
+intensities of each protein found in the biological specimen.
+
+\hypertarget{transformation-of-protein-expression-data}{%
+\paragraph{Transformation of protein expression
+data}\label{transformation-of-protein-expression-data}}
+
+Data transformation applies a function to each value of a matrix or
+array, so that: \[
+y_i = f(x_i)
+\] Depending on the distribution of the values in the observed
+expression set, different transformations can be applied to fulfill
+assumptions for statistical testing. In the shiny application, various
+procedures are implemented and can be further expanded upon request from
+the user. The macrophage analysis done by Specht et al.~mentioned in
+their SCoPE2 publication \citep{Specht2021} uses the logarithm to the
+base 2 to spread a compacted distribution and remove skewness in the
+dataset. This transformation was used as a reference when comparing
+methods. However, using the logarithm to the base 10 may be an easily
+interpretable way of defining expression data and will also be
+facilitated by other transformation methods such as the boxcox method,
+which is also implemented in the application. When observing a wide
+distribution of small and large values of the expression set in a
+histogram, a square root transformation can help increase the
+variability of smaller values and decrease the variability of larger
+values. Practically impossible, but in data-driven science, occasionally
+negative expression values may be present, and they can be converted
+into positive values by taking each one to the power of two. Depending
+on the distribution, the user has to decide which transformation to
+consider and verify that statistical assumptions are met after
+application with visualizations such as histograms, qqplots, and
+MA-plots. For statistically inexperienced users, the boxcox
+transformation can help with this decision. Developed in 1964 by Box and
+Cox \citep{Sakia1992}, this method applies a linear model against 1 on
+the data to determine the statistical parameter lambda by the maximum
+likelihood method. The log-likelihood method takes the 95\% confidence
+interval for the parameter lambda and the final lambda is chosen as the
+value with the highest log-likelihood value.
+
+\begin{tabular}{l|r}
+\hline
+& lambda & Transformation\$\\
+\hline
+-2 & $\frac{1}{x^2}$
+\hline
+-1 & $\frac{1}{x}$
+\hline
+-0.5 & $\frac{1}{\sqrt{x}}$
+\hline
+0 & $\log{x}$
+\hline
+0.5 & $\sqrt{x}$
+\hline
+1 & $x$
+\hline
+2 & $x^2$
+\end{tabular}
+
+Depending on the size of lambda, a certain function will be
+automatically applied to each value of the expression set in order to
+introduce normality. In terms of usability, the boxcox method is the
+most efficient transformation method, taking the decision of which
+calculation to apply away from the user.
+
+\hypertarget{normalization}{%
+\paragraph{Normalization}\label{normalization}}
+
+The term normalization is ambiguous in data-science and will be
+explained briefly in this chapter. Starting with converting the data to
+the Z-distribution (Gaussian result), where 0 reflects the mean and 1
+refers to the standard deviation, until vectorization where every
+feature (=protein) can be explained as a vector which points in a
+specific direction in the unit sphere, normalization can have different
+meanings. Normalization is considered as part of the scaling procedure
+and ensures that every feature contributes equally to our statistical
+model and hinders large values from biasing the model in a particular
+direction. Although this could also be harmful for the modeling
+procedure, since the impact of important features on the dataset will
+also be reduced. Depending on the algorithms later used in the data
+processing, the method of choice for normalization can change the
+results significantly. The K-means clustering algorithm for instance
+relies on the distances between distinct data points by minimizing
+variance of the squared euclidean distances. This method will be applied
+in the missing value imputation which will be the next step applied to
+our data. Hence we want to make sure that the get the maximum accuracy
+for our imputed values. The first option to choose in the user interface
+is the column-wise median and row-wise mean normalization. This method
+was chosen by Specht et al in the SCoPE2 publication 2021
+/citep\{Specht2021\} and worked as a reference during the development
+process to benchmark other methods. When dividing each value of an
+expression set by the median of the particular column, the procedure is
+considered as a column-wise median normalization. Hence the
+mean-normalization works the same way on the column. In the analysis
+pipeline both procedures were applied on the dataset. However when the
+logarithm as the transformation method was chosen in advance the
+pipeline automatically switches from a division to a subtraction, since:
+\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\)
 
@@
 their expression values.
-
-\pagenumbering{arabic}
-
-\newpage
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-\vspace{0.5cm}
-
-\hypertarget{materials-and-methods}{%
-\section{Materials and Methods}\label{materials-and-methods}}
-
-\hypertarget{materials}{%
-\subsection{Materials}\label{materials}}
-
-For analysis two types of cells were used. One type is the Jurkat-based
-cell line (J-lat) with integrated HIV.
-
-The other type of cells are macrophages with a sample size of 72 cells.
-The analysis is done with two groups. A HIV negative (HIV-) control
-group and a HIV positive (HIV+) group.
-
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-\hypertarget{acquisition}{%
-\subsubsection{Acquisition}\label{acquisition}}
-
-Analysis of the data was done with MaxQuant \citep{Cox2008}
 

### 2023-05-02 markdown/methods.Rmd
normalization started to ellaborate

@@
 ## Data
-```{r Dataflow, echo=FALSE}
-library(DiagrammeR)
-library(webshot)
-
-DiagrammeR::grViz("digraph {
-	graph [layout  =  dot,
-	       rankdir  = TB]
-
-node [shape = rectangle]
-rawData [label = 'msData.raw']
-node [shape = oval]
-MQ [label = 'MaxQuant']
-node [shape = invhouse]
-MQ_fd [label = 'feature detection']
-node [shape = cylinder]
-MQ_ds [label = 'Database search']
-node [shape = invhouse]
-MQ_fdr [label = 'FDR calculation']
-MQ_quant [label = 'Protein Quantification']
-node [shape = rectangle]
-MQoutput [label = 'evidence.txt']
-sampleAnnotate [label = 'sample_annotation.txt']
-node [shape = oval]
-ProteomicsWorkbench [label = 'Proteomics Workbench']
-
-rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
-sampleAnnotate -> ProteomicsWorkbench
-}",
-height = 500)
+```{r Dataflow, eval=FALSE, include=FALSE}
+# library(DiagrammeR)
+# library(webshot)
+# 
+# DiagrammeR::grViz("digraph {
+# 	graph [layout  =  dot,
+# 	       rankdir  = TB]
+# 
+# node [shape = rectangle]
+# rawData [label = 'msData.raw']
+# node [shape = oval]
+# MQ [label = 'MaxQuant']
+# node [shape = invhouse]
+# MQ_fd [label = 'feature detection']
+# node [shape = cylinder]
+# MQ_ds [label = 'Database search']
+# node [shape = invhouse]
+# MQ_fdr [label = 'FDR calculation']
+# MQ_quant [label = 'Protein Quantification']
+# node [shape = rectangle]
+# MQoutput [label = 'evidence.txt']
+# sampleAnnotate [label = 'sample_annotation.txt']
+# node [shape = oval]
+# ProteomicsWorkbench [label = 'Proteomics Workbench']
+# 
+# rawData -> MQ -> MQ_fd -> MQ_ds -> MQ_fdr -> MQ_quant -> MQoutput -> ProteomicsWorkbench
+# sampleAnnotate -> ProteomicsWorkbench
+# }",
+# height = 500)
 ```
@@
 
-```{r data_processing_pipeline_flowchart, echo=FALSE, fig.height=17, fig.width=13}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=17, fig.width=13}
 library(Gmisc, quietly = TRUE)
@@
 
-```{r data_processing_pipeline_flowchart2, echo=FALSE, fig.height=9, fig.width=32}
+```{r data_processing_pipeline_flowchart_horizontal, eval=FALSE, fig.height=9, fig.width=32, include=FALSE}
 library(Gmisc, quietly = TRUE)
@@
 For statistically inexperienced users, the boxcox transformation can help with this decision.
-Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a recursion on the data to determine the statistical parameter lambda.
+Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a linear model against 1 on the data to determine the statistical parameter lambda by the maximum likelihood method.
+The log-likelihood method takes the 95% confidence interval for the parameter lambda and the final lambda is chosen as the value with the highest log-likelihood value.
+```{r lambda_table, echo=FALSE}
+# 
+# table <- tibble(lambda=c(-2,-1,-0.5,0,0.5,1,2), Transformation=c("$\frac{1}{x^2}$", "$\frac{1}{x}$", "$\frac{1}{\sqrt{x}}$", "$\log{x}$", "$\sqrt{x}$", "$x$", "$x^2$"))
+
+# knitr::kable(cbind(lambda=c(-2,-1,-0.5,0,0.5,1,2), Transformation=c("\$\\frac{1}{x^2}$", "\$\\frac{1}{x}$", "\$\\frac{1}{\\sqrt{x}}$", "\$\\log{x}$", "\$\\sqrt{x}$", "\$x$", "\$x^2$")), escape = F)
+```
+
+\begin{tabular}{l|r}
+\hline
+& lambda & Transformation\$\\
+\hline
+-2 & $\frac{1}{x^2}$
+\hline
+-1 & $\frac{1}{x}$
+\hline
+-0.5 & $\frac{1}{\sqrt{x}}$
+\hline
+0 & $\log{x}$
+\hline
+0.5 & $\sqrt{x}$
+\hline
+1 & $x$
+\hline
+2 & $x^2$
+\end{tabular}
 
@@
 In terms of usability, the boxcox method is the most efficient transformation method, taking the decision of which calculation to apply away from the user.
+
+#### Normalization
+The term normalization is ambiguous in data-science and will be explained briefly in this chapter.
+Starting with converting the data to the Z-distribution (Gaussian result), where 0 reflects the mean and 1 refers to the standard deviation, until vectorization where every feature (=protein) can be explained as a vector which points in a specific direction in the unit sphere, normalization can have different meanings.
+Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to our statistical model and hinders large values from biasing the model in a particular direction.
+Although this could also be harmful for the modeling procedure, since the impact of important features on the dataset will also be reduced.
+Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly.
+The K-means clustering algorithm for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances.
+This method will be applied in the missing value imputation which will be the next step applied to our data.
+Hence we want to make sure that the get the maximum accuracy for our imputed values.
+The first option to choose in the user interface is the column-wise median and row-wise mean normalization.
+This method was chosen by Specht et al in the SCoPE2 publication 2021 /citep{Specht2021} and worked as a reference during the development process to benchmark other methods.
+When dividing each value of an expression set by the median of the particular column, the procedure is considered as a column-wise median normalization.
+Hence the mean-normalization works the same way on the column.
+In the analysis pipeline both procedures were applied on the dataset.
+However when the logarithm as the transformation method was chosen in advance the pipeline automatically switches from a division to a subtraction, since:
+$\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v$
 

### 2023-05-02 scope2.R

@@ -26,23 +26,23 @@ library("tibble")
 
 # read in MS result table
 mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")
-mqScpData2 <- read.csv("/home/lukas/Downloads/Raw_data.csv")
-mqScpData3 <- readxl::read_excel("/home/lukas/Downloads/20230207_L1_UN_Monocytes_TMT12_frac.xlsx")
+# mqScpData2 <- read.csv("/home/lukas/Downloads/Raw_data.csv")
+# mqScpData3 <- readxl::read_excel("/home/lukas/Downloads/20230207_L1_UN_Monocytes_TMT12_frac.xlsx")
 
 sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/sampleAnnotation_tabdel.txt")
-sampleAnnotation2 = read.csv("/home/lukas/Downloads/Design.csv")
+# sampleAnnotation2 = read.csv("/home/lukas/Downloads/Design.csv")
 
 
 # define file handling for not mq generated data
 
 #metadata is used as a reference --> change different metadata file accordingly
-if (!"Sequence" %in% colnames(mqScpData2)) {
-  sampleAnnotation2 <- 
-    sampleAnnotation2 %>% rename(
-    SampleType = Group,
-    Raw.file = Batch) %>%
-    arrange(Raw.file)
-}
+# if (!"Sequence" %in% colnames(mqScpData2)) {
+#   sampleAnnotation2 <- 
+#     sampleAnnotation2 %>% rename(
+#     SampleType = Group,
+#     Raw.file = Batch) %>%
+#     arrange(Raw.file)
+# }
 
 
 # create QFeature object
@@ -54,23 +54,23 @@ scp <- readSCP(featureData = mqScpData,
                removeEmptyCols = TRUE)
 
 
-if (!"Sequence" %in% colnames(mqScpData2)) {
-  mqScpData2 <- 
-    mqScpData2 %>% rename(
-    Protein = ID
-  )
-}  
-
-
-quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData), value = TRUE)
-
-scp2 <- readSCP(featureData = mqScpData2,
-                 colData = sampleAnnotation2,
-                 channelCol = "Channel",
-                 batchCol = "Raw.file",
-                 suffix=paste0("_TMT", 1:length(unique(sampleAnnotation2$Channel))),
[…]

### 2023-05-03 scope2.R
fancy volcano finished

@@ -813,15 +813,32 @@ genetable <- topTable(fit, number = Inf, adjust = "BH")
 
 volcanoplot(fit)
 
-library(EnhancedVolcano)
-EnhancedVolcano(genetable,
-                lab = rownames(genetable),
-                x = 'logFC',
-                y = 'P.Value',
-                pCutoff = 0.05,
-                FCcutoff = 0.05,
-                ylim = c(-log10(max(genetable$P.Value)), -log10(min(genetable$P.Value))),
-                xlim = c(min(genetable$logFC), max(genetable$logFC))
-                )
+
+library(plotly)
+
+# load data.frame
+toptable <- as.data.frame(genetable)
+toptable <- cbind(rownames(toptable), data.frame(toptable, row.names=NULL))
+colnames(toptable)[1] <- "protein"
+
+# significance cutoffs
+FCcutoff <- 0.05
+pCutoff <- 0.05
+
+# define factor levels for significance
+toptable$Significance <- "not significant"
+toptable$Significance[(abs(toptable$logFC) > FCcutoff)] <- "fold change"
+toptable$Significance[(toptable$P.Value < pCutoff)] <- "p-value"
+toptable$Significance[(toptable$P.Value < pCutoff) & (abs(toptable$logFC) > 
+                                                     FCcutoff)] <- "fold change & p-value"
+toptable$Significance <- factor(toptable$Sig, levels = c("not significant", "fold change", 
+                                                "p-value", "fold change & p-value"))
+
+p <- ggplot(toptable, aes(x = logFC, y = -log10(P.Value), text=protein)) +
+  geom_point(aes(color = Significance), alpha = 1/2, shape = 19, size = 1.5, na.rm = TRUE) +
+  theme(legend.position = "top")
+  
+ggplotly(p, tooltip = "text")
+
 
 

### 2023-05-03 scpR-app/app.R

@@ -14,7 +14,7 @@ library("CONSTANd")
 library("stats")
 library("impute")
 library("sva")
-
+library("plotly")
 
 reactlog::reactlog_enable()
 
@@ -101,9 +101,16 @@ ui <- fluidPage(
                            uiOutput("sample_select"),
                            uiOutput("add_factor"),
                            plotOutput("venn_diagram"),
-                           selectInput("chosen_coef", "Select your coefficient of interest", ""),
-                           plotOutput("volcano", hover = hoverOpts(id ="plot_hover")),
-                           verbatimTextOutput("hover_info"),
+                           fluidRow(width=12,
+                                    column(width=4,
+                                           selectInput("chosen_coef", "Select your coefficient of interest", "")),
+                                    column(width = 4,
+                                           numericInput("p_value_cutoff", "choose p-value cutoff", value = 0.05, min = 0, step = 0.01)),
+                                    column(width=4,
+                                           numericInput("fold_change_cutoff", "choose fold change cutoff", value = 0.05, min = 0, step = 0.01))
+                           ),
+                           
+                           plotlyOutput("volcano"),
                            tableOutput("protein_table")
                   )
       )
@@ -1156,39 +1163,38 @@ server <- function(input, output, session) {
   })
   
   # volcanoplot in statistic tab
-  output$volcano <- renderPlot({
-    req(input$chosen_coef)
-    req(stat_result())
-    volcanoplot(stat_result(), cex = 0.5, coef = input$chosen_coef) 
-  })
+  output$volcano <- renderPlotly({
+    # load data.frame
+    toptable <- protein_table()
+    toptable <- cbind(rownames(toptable), data.frame(toptable, row.names=NULL))
+    colnames(toptable)[1] <- "protein"
+    
+    # significance cutoffs
+    FCcutoff <- input$fold_change_cutoff
+    pCutoff <- input$p_value_cutoff
+    
+    # define factor levels for significance
+    toptable$Significance <- "not significant"
+    toptable$Significance[(abs(toptable$logFC) > FCcutoff)] <- "fold change"
+    toptable$Significance[(toptable$P.Value < pCutoff)] <- "p-value"
+    toptable$Significance[(toptable$P.Value < pCutoff) & (abs(toptable$logFC) > 
+                                                            FCcutoff)] <- "fold change & p-value"
[…]

### 2023-05-03 scpR-app/app.R
significance table improved

@@ -1,6 +1,7 @@
 library("shiny")
 library("shinyWidgets")
 library("shinydashboard")
+library("DT")
 
 library("scp")
 library("SingleCellExperiment")
@@ -94,12 +95,18 @@ ui <- fluidPage(
                   tabPanel("Statistics",
                            actionButton("run_statistics", "Press to run statistics"),
                            actionButton("qqplot", "check dependencies for linear model"),
-                           selectInput("model_design", "Choose your study design",
-                                       choices = c("All pairwise comparison",
-                                                   "Differential Expression with defined Contrasts",
-                                                   "Multi factor additivity")),
-                           uiOutput("sample_select"),
-                           uiOutput("add_factor"),
+                           fluidRow(width=12,
+                                    column(width=4,
+                                           selectInput("model_design", "choose your study design",
+                                                       choices = c("All pairwise comparison",
+                                                                   "Differential Expression with defined Contrasts",
+                                                                   "Multi factor additivity"))),
+                                    column(width = 4,
+                                           uiOutput("sample_select")),
+                                    column(width=4,
+                                           uiOutput("add_factor"))
+                           ),
+                           selectInput("p_value_correction", "select method for p-value correction", choices = c("none", "BH", "fdr", "BY", "holm")),
                            plotOutput("venn_diagram"),
                            fluidRow(width=12,
                                     column(width=4,
@@ -109,9 +116,8 @@ ui <- fluidPage(
                                     column(width=4,
                                            numericInput("fold_change_cutoff", "choose fold change cutoff", value = 0.05, min = 0, step = 0.01))
                            ),
-                           
                            plotlyOutput("volcano"),
-                           tableOutput("protein_table")
+                           DT::dataTableOutput("protein_table")
                   )
       )
     )
@@ -1144,13 +1150,13 @@ server <- function(input, output, session) {
   # additional ui for statistic module
   output$sample_select <- renderUI({
     req(input$model_design == "Differential Expression with defined Contrasts" | input$model_design == "Multi factor additivity")
-    selectInput("selectedComp_stat", "choose your samples of interest", "", multiple = T)
[…]

### 2023-05-03 markdown/methods.Rmd
methods: normalization finished. missing value imputation started

@@
 The log-likelihood method takes the 95% confidence interval for the parameter lambda and the final lambda is chosen as the value with the highest log-likelihood value.
-```{r lambda_table, echo=FALSE}
-# 
-# table <- tibble(lambda=c(-2,-1,-0.5,0,0.5,1,2), Transformation=c("$\frac{1}{x^2}$", "$\frac{1}{x}$", "$\frac{1}{\sqrt{x}}$", "$\log{x}$", "$\sqrt{x}$", "$x$", "$x^2$"))
 
-# knitr::kable(cbind(lambda=c(-2,-1,-0.5,0,0.5,1,2), Transformation=c("\$\\frac{1}{x^2}$", "\$\\frac{1}{x}$", "\$\\frac{1}{\\sqrt{x}}$", "\$\\log{x}$", "\$\\sqrt{x}$", "\$x$", "\$x^2$")), escape = F)
-```
-
-\begin{tabular}{l|r}
-\hline
-& lambda & Transformation\$\\
-\hline
--2 & $\frac{1}{x^2}$
-\hline
--1 & $\frac{1}{x}$
-\hline
--0.5 & $\frac{1}{\sqrt{x}}$
-\hline
-0 & $\log{x}$
-\hline
-0.5 & $\sqrt{x}$
-\hline
-1 & $x$
-\hline
+\begin{center}
+\begin{tabular}{l r}
+$\lambda$ & Transformation \\
+-2 & $\frac{1}{x^2}$ \\
+-1 & $\frac{1}{x}$ \\
+-0.5 & $\frac{1}{\sqrt{x}}$ \\
+0 & $\log{x}$ \\
+0.5 & $\sqrt{x}$ \\
+1 & $x$ \\
 2 & $x^2$
 \end{tabular}
+\end{center}
 
@@
 The term normalization is ambiguous in data-science and will be explained briefly in this chapter.
-Starting with converting the data to the Z-distribution (Gaussian result), where 0 reflects the mean and 1 refers to the standard deviation, until vectorization where every feature (=protein) can be explained as a vector which points in a specific direction in the unit sphere, normalization can have different meanings.
+Starting with converting the data to the Z-distribution (Gaussian result), where 0 represents the mean and 1 represents the standard deviation, in vectorization every feature (or protein) is represented as a vector that points in a specific direction in the unit sphere, normalization can have different meanings.
 Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to our statistical model and hinders large values from biasing the model in a particular direction.
-Although this could also be harmful for the modeling procedure, since the impact of important features on the dataset will also be reduced.
+However, this could also have negative consequences for the modeling procedure, as it may reduce the impact of important features on the dataset.
 Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly.
-The K-means clustering algorithm for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances.
+The K-means clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances.
 This method will be applied in the missing value imputation which will be the next step applied to our data.
-Hence we want to make sure that the get the maximum accuracy for our imputed values.
-The first option to choose in the user interface is the column-wise median and row-wise mean normalization.
-This method was chosen by Specht et al in the SCoPE2 publication 2021 /citep{Specht2021} and worked as a reference during the development process to benchmark other methods.
+Therefore, we want to ensure that we obtain the most accurate imputed values possible.
+The first normalization method available in the user interface is column-wise median and row-wise mean normalization.
+This method was chosen by Specht et al in the SCoPE2 publication 2021 \citep{Specht2021} and worked as a reference during the development process to benchmark other methods.
 When dividing each value of an expression set by the median of the particular column, the procedure is considered as a column-wise median normalization.
-Hence the mean-normalization works the same way on the column.
+Although the mean-normalization works the same way but takes the mean as the second variable.
 In the analysis pipeline both procedures were applied on the dataset.
-However when the logarithm as the transformation method was chosen in advance the pipeline automatically switches from a division to a subtraction, since:
-$\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v$
+However when the logarithm as the transformation method was chosen as transformation method, the pipeline automatically switches from a division to a subtraction, since:
+$\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v$.
+After applying this normalization method, each value can be interpreted as a distance or fraction of the mean or median of the corresponding column or row, enabling fair comparisons between features and samples.
+The second method for normalization included is the CONSTANd method, which was proven suitable for relative quantification in the field of proteomics \citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq data.
+CONSTANd uses a technique called matrix raking, which employs the RAS algorithm.
+The expression set reflects the nonnegative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $B$ which equals to $diag(x) * A *diag(y)$ .
+Whereas $x \in \mathbb{R}^{m}$ and $y \in \mathbb{R}^{n}$.
+The solution to this problem involves finding the row sum of $B=u_i$ and the column sum of $B=v_j$, where $i$ and $j$ denote the row and column indices of the matrix, respectively \citep{Bacharach1965}.
+In other words, the matrix will be alternatively manipulated on rows and columns until the mean of both equals 1.
+In order to apply the method and yield true results, one has to consider 3 major assumptions between sample types.
+This assumption can be observed in the MA-plot which indicates the differences of two samples.
+If we have the intensities of two samples, $R$ and $G$, we can plot them on a graph with the x-axis as $M=\log_2(\frac{R}{G})$ and the y-axis as $A=\frac{1}{2}\log_2(RG)$.
+A comparison between all sample types of identically processed sets is needed before calculation and the assumptions must be fulfilled.
 
-Further filter steps can include correction between multiple runs.
-This kind of process need the addition of a reference channel.
-However when observing a single run, these steps are not crucial for the upcoming analysis.
+\begin{itemize}
+\item{\textbf{1}} \textbf{The majority of proteins are not differentially expressed}
+\\Density of the dots decreases when looking from the middle of the cloud towards the edges.
+
+\item{\textbf{2}} \textbf{Up- and downregulation is balanced around the mean expression}
+\\The scatterplot of the data points is symmetrical around the mean on the horizontal axis.
+
+\item{\textbf{3}} \textbf{Systematic bias correlates with the magnitude of expression}
+\\The symmetry axis is approximately horizontal.
+
+\end{itemize}
+
+If the MA-plot does not visualize any of these violations the CONSTANd method can be applied on the dataset and the biological meaning of the sample types can be assessed.
+The method is utilized by the bioconductor package handler.
+However if the dataset already consists of balanced features and samples or the user wants to benchmark the methods applied, normalization can be skipped also.
+
+#### Missing value handling
+When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches.
+From a biological perspective, it is highly unlikely that a cell completely lacks a single protein, as it would imply a complete loss of function for that protein.
+Therefore, an intensity value of 0 is close to impossible for most proteins.
+In order to handle this dark space in the expression set, several methods can be chosen in the application.
+A variety of function employs replacing the missing values with the mean or median of the rest of the matrix.
+This procedure introduces no bias in the data, although differential expression for the sample carrying the missing feature can not be expected as well.
+Even a more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
+Another common practice is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing.
+$K$ denotes for the count of neighboring values of a gene and can be selected upon preference.
 

### 2023-05-03 markdown/methods.tex
methods: normalization finished. missing value imputation started

@@
 
-\includegraphics{methods_files/figure-latex/data_processing_pipeline_flowchart-1.pdf}
+\includegraphics{methods_files/figure-latex/data_processing_pipeline_flowchart_vertical-1.pdf}
 
@@
 will be further processed into their respective proteins after the
-quality control on the peptide level is performed.
+quality control on the peptide level is performed. Finally an expression
+matrix for every protein and their respective channel column is
+returned. Each channel refers to a sample and reflects contains the
+intensities of each protein found in the biological specimen.
 
@@
 expression set, different transformations can be applied to fulfill
-dependencies for statistical testing. In the shiny application various
+assumptions for statistical testing. In the shiny application, various
 procedures are implemented and can be further expanded upon request from
-the user. The macrophage analysis done by Specht et. Al mentioned in
-their SCoPE2 publication citep\{Specht2021\} uses the logarithm to the
-base 2 to spread a compacted distribution and the remove skewness in the
+the user. The macrophage analysis done by Specht et al.~mentioned in
+their SCoPE2 publication \citep{Specht2021} uses the logarithm to the
+base 2 to spread a compacted distribution and remove skewness in the
 dataset. This transformation was used as a reference when comparing
-methods. However using the logarithm to the base 10 can be easy
+methods. However, using the logarithm to the base 10 may be an easily
 interpretable way of defining expression data and will also be
-facilitated by other transformation methods such as the boxcox method
-which is also implemented in the application. The boxcox transformation
-developed 1964 by Box and Cox citep\{Sakia1992\} applies a recursion on
-the data by determining the parameter lambda. Depending on the size of
-lambda a certain function will be applied on each value of the
-expression set in order to introduce normality. As already mentioned
-transformations applied by the BoxCox method range from log10 to taking
-the square root of each value
+facilitated by other transformation methods such as the boxcox method,
+which is also implemented in the application. When observing a wide
+distribution of small and large values of the expression set in a
+histogram, a square root transformation can help increase the
+variability of smaller values and decrease the variability of larger
+values. Practically impossible, but in data-driven science, occasionally
+negative expression values may be present, and they can be converted
+into positive values by taking each one to the power of two. Depending
+on the distribution, the user has to decide which transformation to
+consider and verify that statistical assumptions are met after
+application with visualizations such as histograms, qqplots, and
+MA-plots. For statistically inexperienced users, the boxcox
+transformation can help with this decision. Developed in 1964 by Box and
+Cox \citep{Sakia1992}, this method applies a linear model against 1 on
+the data to determine the statistical parameter lambda by the maximum
+likelihood method. The log-likelihood method takes the 95\% confidence
+interval for the parameter lambda and the final lambda is chosen as the
+value with the highest log-likelihood value.
+
+\begin{center}
+\begin{tabular}{l r}
+lambda & Transformation \\
+-2 & $\frac{1}{x^2}$ \\
+-1 & $\frac{1}{x}$ \\
+-0.5 & $\frac{1}{\sqrt{x}}$ \\
+0 & $\log{x}$ \\
+0.5 & $\sqrt{x}$ \\
+1 & $x$ \\
+2 & $x^2$
+\end{tabular}
+\end{center}
+
+Depending on the size of lambda, a certain function will be
+automatically applied to each value of the expression set in order to
+introduce normality. In terms of usability, the boxcox method is the
+most efficient transformation method, taking the decision of which
+calculation to apply away from the user.
+
+\hypertarget{normalization}{%
+\paragraph{Normalization}\label{normalization}}
+
+The term normalization is ambiguous in data-science and will be
+explained briefly in this chapter. Starting with converting the data to
+the Z-distribution (Gaussian result), where 0 reflects the mean and 1
+refers to the standard deviation, until vectorization where every
+feature (=protein) can be explained as a vector which points in a
+specific direction in the unit sphere, normalization can have different
+meanings. Normalization is considered as part of the scaling procedure
+and ensures that every feature contributes equally to our statistical
+model and hinders large values from biasing the model in a particular
+direction. Although this could also be harmful for the modeling
+procedure, since the impact of important features on the dataset will
+also be reduced. Depending on the algorithms later used in the data
+processing, the method of choice for normalization can change the
+results significantly. The K-means clustering algorithm for instance
+relies on the distances between distinct data points by minimizing
+variance of the squared euclidean distances. This method will be applied
+in the missing value imputation which will be the next step applied to
+our data. Hence we want to make sure that the get the maximum accuracy
+for our imputed values. The first option to choose in the user interface
+is the column-wise median and row-wise mean normalization. This method
+was chosen by Specht et al in the SCoPE2 publication 2021
+/citep\{Specht2021\} and worked as a reference during the development
+process to benchmark other methods. When dividing each value of an
+expression set by the median of the particular column, the procedure is
+considered as a column-wise median normalization. Hence the
+mean-normalization works the same way on the column. In the analysis
+pipeline both procedures were applied on the dataset. However when the
+logarithm as the transformation method was chosen in advance the
+pipeline automatically switches from a division to a subtraction, since:
+\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\)
 

### 2023-05-08 scope2_on_PD.R
experimenting with PD file

@@ -16,153 +16,174 @@ library("sva")
 library("tibble")
 
 # read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/20230202_HF2_08_Ujjwal_Monocytes_TMT12_PSMs.txt")
+evidence_data <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/20230202_HF2_08_Ujjwal_Monocytes_TMT12_PSMs.txt")
 
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/sampleAnnotate_tabdel.txt")
+meta_data_0 = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/sampleAnnotate_tabdel.txt")
 
-# attempting to calculate PIF
-#The calculation involves dividing the intensity of the selected precursor peak by the total intensity of the isolation window
-#and is performed before and after the MS/MS scan of interest and interpolated at the recorded time of the MS/MS acquisition
+# define cutoff values for example
 
-# exact same values? 
+PIF_cutoff <- 0.9
+qvalue_cutoff <- 0.05
+nObs_pep_razrpr <- 5
+MedCV_thresh <- 0.65
+pNA <- 0.99
+
+if ("Spectrum.File" %in% colnames(evidence_data) && ("Spectrum.File" %in% colnames(meta_data_0))) {
+  
+  # parse colnames to maxquant style
+  colnames(evidence_data)[colnames(evidence_data) == "Spectrum.File"] <- "Raw.file"
+  colnames(meta_data_0)[colnames(meta_data_0) == "Spectrum.File"] <- "Raw.file"
+  
+  if ("Percolator.PEP" %in% colnames(evidence_data)) {
+    colnames(evidence_data)[colnames(evidence_data) == "Percolator.PEP"] <- "PEP"
+  }
+  if ("Master.Protein.Accessions" %in% colnames(evidence_data)) {
+    colnames(evidence_data)[colnames(evidence_data) == "Master.Protein.Accessions"] <- "Leading.razor.protein"
+  }
+  if ("Annotated.Sequence" %in% colnames(evidence_data)) {
+    colnames(evidence_data)[colnames(evidence_data) == "Annotated.Sequence"] <- "Modified.sequence"
+  }
+  
[…]

### 2023-05-08 scope2.R
pipeline until missing v impute for evidence finished

@@ -50,7 +50,6 @@ scp <- readSCP(featureData = mqScpData,
                colData = sampleAnnotation,
                channelCol = "Channel",
                batchCol = "Raw.file",
-               suffix = paste0("_TMT", 1:length(unique(sampleAnnotation$Channel))),
                removeEmptyCols = TRUE)
 
 
@@ -610,27 +609,30 @@ plotReducedDim(scp[["proteins_dim_red"]],
 
 
 
-# ## Get the features
-# subsetByFeature(scp, "Q9ULV4") %>%
-#   ## Format the `QFeatures` to a long format table
-#   longFormat(colvars = c("Raw.file", "SampleType", "Channel")) %>%
-#   data.frame %>%
-#   ## This is used to preserve ordering of the samples and assays in ggplot2
-#   mutate(assay = factor(assay, levels = names(scp)),
-#          Channel = sub("Reporter.intensity.", "Label", Channel),
-#          Channel = factor(Channel, levels = unique(Channel))) %>%
-#   ## Start plotting
-#   ggplot(aes(x = Channel, y = value, group = rowname, col = SampleType)) +
-#   geom_point() +
-#   ## Plot every assay in a separate facet
-#   facet_wrap(facets = vars(assay), scales = "free_y", ncol = 3) +
-#   ## Annotate plot
-#   xlab("Channels") +
-#   ylab("Intensity (arbitrary units)") +
-#   ## Improve plot aspect
-#   theme(axis.text.x = element_text(angle = 90),
-#         strip.text = element_text(hjust = 0),
-#         legend.position = "bottom")
+## Get the features
+subsetByFeature(scp, "Q92954") %>%
+  ## Format the `QFeatures` to a long format table
+  longFormat(colvars = c("Raw.file", "SampleType", "Channel")) %>%
+  data.frame %>%
+  ## This is used to preserve ordering of the samples and assays in ggplot2
+  mutate(assay = factor(assay, levels = names(scp)),
+         Channel = sub("Reporter.intensity.", "", Channel),
+  ) %>%
+  mutate(Channel = as.numeric(Channel)) %>%
+  arrange(Channel) %>%
+  mutate(Channel = factor(Channel, levels = unique(Channel))) %>%
+  ## Start plotting
+  ggplot(aes(x = Channel, y = value, group = rowname, col = SampleType)) +
+  geom_point() +
+  ## Plot every assay in a separate facet
[…]

### 2023-05-08 scpR-app/app.R

@@ -30,34 +30,37 @@ ui <- fluidPage(
       # the button to rule them all
       actionButton("update_button", "Press to run/update"),
       actionButton("help", "Help"),
+      
       # read in of the data
+      materialSwitch(inputId = "file_level", value = F, label="Handle preprocessed expression set", status = "danger"),
       fileInput("evidence_file", "Upload evidence.txt from MaxQuant", accept = c("text")),
       fileInput("sample_annotation_file", "Upload sample annotation file", accept = c("text")),
-      textInput("label_suffix", "Input the label suffix (e.g. _TMT)"),
       
-      selectInput("selectedSampleType_to_exclude", "Select SampleType to exclude", "", multiple=T),
-      
-      # cutoff for PIF
-      numericInput("PIF_cutoff", "Input cutoff value for parental ion fraction. PSMs larger than the value remain", 0.1, min = 0, max=1, step = 0.1),
-      
-      numericInput("qvalue_cutoff", "Input cutoff value for q-value. PSMs with q-value smaller than the value remain", 0.01, min = 0, max=1, step = 0.0001),
-      
-      # minimal observation of a peptide within their corresponding razor protein
-      numericInput("nObs_pep_razrpr", "Input minimal observation of a peptide within their corresponding razor peptide", 5, min = 1, max=100, step = 1),
-      
-      # max covariance accepted
-      numericInput("MedCV_thresh", "Input maximal Covariance accepted", 0.65 , min = 0, max=1, step = 0.01),
-      
-      # remove peptides with missing data
-      numericInput("pNA", "Input percentage threshold for peptide data", 0.99 , min = 0, max=1, step = 0.01),
+      conditionalPanel(condition = "!input.file_level", 
+                       selectInput("selectedSampleType_to_exclude", "Select SampleType to exclude", "", multiple=T),
+                       # cutoff for PIF
+                       numericInput("PIF_cutoff", "Input cutoff value for parental ion fraction. PSMs larger than the value remain", 0.1, min = 0, max=1, step = 0.1),
+                       
+                       numericInput("qvalue_cutoff", "Input cutoff value for q-value. PSMs with q-value smaller than the value remain", 0.01, min = 0, max=1, step = 0.0001),
+                       
+                       # minimal observation of a peptide within their corresponding razor protein
[…]

### 2023-05-08 scpR-app/app.R
minor commit before literature recherche

@@ -415,7 +415,6 @@ server <- function(input, output, session) {
       
       incProgress(14/17, detail=paste("transforming protein data"))
       req(input$transform_base)
-
       if (input$transform_base == "log2") {
         if (input$file_level == FALSE) {
           scp_0 <- logTransform(scp_0,
@@ -651,10 +650,15 @@ server <- function(input, output, session) {
         }
       }
       
-      
       if (length(peptide_file) > 1) {
+        if (input$file_level == FALSE) {
+          
+        }
         incProgress(16/17, detail=paste("running missing value imputation"))
         if (input$missing_v == "KNN") {
+          
+          
+        }
           scp_0 <- impute(scp_0,
                           i = "proteins_norm",
                           name = "proteins_imptd",

### 2023-05-08 scpR-app/app.R
minor bug fix

@@ -651,14 +651,8 @@ server <- function(input, output, session) {
       }
       
       if (length(peptide_file) > 1) {
-        if (input$file_level == FALSE) {
-          
-        }
         incProgress(16/17, detail=paste("running missing value imputation"))
         if (input$missing_v == "KNN") {
-          
-          
-        }
           scp_0 <- impute(scp_0,
                           i = "proteins_norm",
                           name = "proteins_imptd",

### 2023-05-08 scope2_BACKUP_18436.R
files deleted

@@ -1,816 +1,0 @@
-
-# workflow as per 
-# https://uclouvain-cbio.github.io/SCP.replication/articles/SCoPE2.html
-# and https://bioconductor.org/packages/release/bioc/vignettes/scp/inst/doc/scp.html
-
-# install scp package
-
-#BiocManager::install("scp")
-#BiocManager::install("scpdata")
-
-# read in SCP data
-
-library("scp")
-library("SingleCellExperiment")
-library("ggplot2")
-library("magrittr")
-library("dplyr")
-library("reshape2")
-library("scater")
-library("limma")
-library("CONSTANd")
-library("stats")
-library("impute")
-library("sva")
-library("tibble")
-
-# read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/20230202_HF2_08_Ujjwal_Monocytes_TMT12_PSMs.txt")
-
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/sampleAnnotate_tabdel.txt")
-
-
-# create QFeature object
-scp <- readSCP(featureData = mqScpData,
-               colData = sampleAnnotation,
-               channelCol = "Channel",
-<<<<<<< HEAD
-               batchCol = "Raw.file",
-=======
-               batchCol = "Spectrum.File",
->>>>>>> master
-               removeEmptyCols = TRUE)
-
-
-
-# skip pool sample
-if ("Pool" %in% sampleAnnotation$SampleType) {
-  scp <- scp[, scp$SampleType !=  "Pool"]
-}
-
-
-
-# plot runs (not needed - sample unfractionized)
-# only one run
-# plot(scp)
-
-# clean missing data
-scp <- zeroIsNA(scp, i=1:length(rowDataNames(scp)))
-
-# filter PSM
-# filter out potential contaminants
-# filter out matches to decoy database
-# keep PSMs with high PIF (parental ion fraction)
-scp <- filterFeatures(scp,
-                      ~ Reverse != "+" &
-                        Potential.contaminant != "+" &
-                        !is.na(PIF) & PIF > 0.10)
-
-
-nPSMs <- dims(scp)[1, ]
-
-ggplot(data.frame(nPSMs)) +
-  aes(x = nPSMs) +
-  geom_histogram(binwidth = 50) +
-  geom_vline(xintercept = 500)
-
-
[…]

### 2023-05-08 scope2_BASE_18436.R

@@ -1,844 +1,0 @@
-
-# workflow as per 
-# https://uclouvain-cbio.github.io/SCP.replication/articles/SCoPE2.html
-# and https://bioconductor.org/packages/release/bioc/vignettes/scp/inst/doc/scp.html
-
-# install scp package
-
-#BiocManager::install("scp")
-#BiocManager::install("scpdata")
-
-# read in SCP data
-
-library("scp")
-library("SingleCellExperiment")
-library("ggplot2")
-library("magrittr")
-library("dplyr")
-library("reshape2")
-library("scater")
-library("limma")
-library("CONSTANd")
-library("stats")
-library("impute")
-library("sva")
-library("tibble")
-
-# read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")
-# mqScpData2 <- read.csv("/home/lukas/Downloads/Raw_data.csv")
-# mqScpData3 <- readxl::read_excel("/home/lukas/Downloads/20230207_L1_UN_Monocytes_TMT12_frac.xlsx")
-
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/sampleAnnotation_tabdel.txt")
-# sampleAnnotation2 = read.csv("/home/lukas/Downloads/Design.csv")
-
-
-# define file handling for not mq generated data
-
-#metadata is used as a reference --> change different metadata file accordingly
-# if (!"Sequence" %in% colnames(mqScpData2)) {
-#   sampleAnnotation2 <- 
-#     sampleAnnotation2 %>% rename(
-#     SampleType = Group,
-#     Raw.file = Batch) %>%
-#     arrange(Raw.file)
-# }
-
-
-# create QFeature object
-scp <- readSCP(featureData = mqScpData,
-               colData = sampleAnnotation,
-               channelCol = "Channel",
-               batchCol = "Raw.file",
-               suffix = paste0("_TMT", 1:length(unique(sampleAnnotation$Channel))),
-               removeEmptyCols = TRUE)
-
-
-# if (!"Sequence" %in% colnames(mqScpData2)) {
-#   mqScpData2 <- 
-#     mqScpData2 %>% rename(
-#     Protein = ID
-#   )
-# }  
-# 
-# 
-# quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData), value = TRUE)
-# 
[…]

### 2023-05-08 scope2_LOCAL_18436.R

@@ -1,846 +1,0 @@
-
-# workflow as per 
-# https://uclouvain-cbio.github.io/SCP.replication/articles/SCoPE2.html
-# and https://bioconductor.org/packages/release/bioc/vignettes/scp/inst/doc/scp.html
-
-# install scp package
-
-#BiocManager::install("scp")
-#BiocManager::install("scpdata")
-
-# read in SCP data
-
-library("scp")
-library("SingleCellExperiment")
-library("ggplot2")
-library("magrittr")
-library("dplyr")
-library("reshape2")
-library("scater")
-library("limma")
-library("CONSTANd")
-library("stats")
-library("impute")
-library("sva")
-library("tibble")
-
-# read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")
-# mqScpData2 <- read.csv("/home/lukas/Downloads/Raw_data.csv")
-# mqScpData3 <- readxl::read_excel("/home/lukas/Downloads/20230207_L1_UN_Monocytes_TMT12_frac.xlsx")
-
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/sampleAnnotation_tabdel.txt")
-# sampleAnnotation2 = read.csv("/home/lukas/Downloads/Design.csv")
-
-
-# define file handling for not mq generated data
-
-#metadata is used as a reference --> change different metadata file accordingly
-# if (!"Sequence" %in% colnames(mqScpData2)) {
-#   sampleAnnotation2 <- 
-#     sampleAnnotation2 %>% rename(
-#     SampleType = Group,
-#     Raw.file = Batch) %>%
-#     arrange(Raw.file)
-# }
-
-
-# create QFeature object
-scp <- readSCP(featureData = mqScpData,
-               colData = sampleAnnotation,
-               channelCol = "Channel",
-               batchCol = "Raw.file",
-               removeEmptyCols = TRUE)
-
-
-# if (!"Sequence" %in% colnames(mqScpData2)) {
-#   mqScpData2 <- 
-#     mqScpData2 %>% rename(
-#     Protein = ID
-#   )
-# }  
-# 
-# 
-# quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData), value = TRUE)
-# 
-# scp2 <- readSCP(featureData = mqScpData2,
-#                  colData = sampleAnnotation2,
[…]

### 2023-05-08 scope2_REMOTE_18436.R

@@ -1,809 +1,0 @@
-
-# workflow as per 
-# https://uclouvain-cbio.github.io/SCP.replication/articles/SCoPE2.html
-# and https://bioconductor.org/packages/release/bioc/vignettes/scp/inst/doc/scp.html
-
-# install scp package
-
-#BiocManager::install("scp")
-#BiocManager::install("scpdata")
-
-# read in SCP data
-
-library("scp")
-library("SingleCellExperiment")
-library("ggplot2")
-library("magrittr")
-library("dplyr")
-library("reshape2")
-library("scater")
-library("limma")
-library("CONSTANd")
-library("stats")
-library("impute")
-library("sva")
-library("tibble")
-
-# read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/20230202_HF2_08_Ujjwal_Monocytes_TMT12_PSMs.txt")
-
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/sampleAnnotate_tabdel.txt")
-
-
-# create QFeature object
-scp <- readSCP(featureData = mqScpData,
-               colData = sampleAnnotation,
-               channelCol = "Channel",
-               batchCol = "Spectrum.File",
-               removeEmptyCols = TRUE)
-
-
-
-# skip pool sample
-if ("Pool" %in% sampleAnnotation$SampleType) {
-  scp <- scp[, scp$SampleType !=  "Pool"]
-}
-
-
-
-# plot runs (not needed - sample unfractionized)
-# only one run
-# plot(scp)
-
-# clean missing data
-scp <- zeroIsNA(scp, i=1:length(rowDataNames(scp)))
-
-# filter PSM
-# filter out potential contaminants
-# filter out matches to decoy database
-# keep PSMs with high PIF (parental ion fraction)
-scp <- filterFeatures(scp,
-                      ~ Reverse != "+" &
-                        Potential.contaminant != "+" &
-                        !is.na(PIF) & PIF > 0.10)
-
-
-nPSMs <- dims(scp)[1, ]
-
-ggplot(data.frame(nPSMs)) +
-  aes(x = nPSMs) +
-  geom_histogram(binwidth = 50) +
-  geom_vline(xintercept = 500)
-
-
-# filter runs with too few features
-# however only one run is done --> changes nothing
[…]

### 2023-05-08 markdown/methods.Rmd
missing v handling finished. batch correction started

@@
 A variety of function employs replacing the missing values with the mean or median of the rest of the matrix.
-This procedure introduces no bias in the data, although differential expression for the sample carrying the missing feature can not be expected as well.
+This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well.
 Even a more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
-Another common practice is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing.
+A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing.
 $K$ denotes for the count of neighboring values of a gene and can be selected upon preference.
+The euclidean distance between two points is defined as $d = \sqrt{a^2 * b^2}$, the $K$ nearest neighbors average value will be assigned to the intensity of the missing feature.
+Since the average is a statistical metric which can be biased easily by skewed distributions, normalization is an important task before performing missing value imputation.
+KNN was benchmarked  for the regression of protein abundance intensities and showed an Area under the Curve (AUC) above 0.7 when testing on a human proteome test set \citep{Lan2013}.
+
+#### Batch correction 
+Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data.
+This variance would overshadow the biological differences across sample types.
+Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter.
+To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used.
+The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined
 

### 2023-05-09 scpR-app/app.R
expression array until missing v finished

@@ -410,11 +410,15 @@ server <- function(input, output, session) {
         }        
         
         
+      } else {
+        rownames(evidence_data) <- evidence_data$ID
+        evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
       } 
-
       
       incProgress(14/17, detail=paste("transforming protein data"))
       req(input$transform_base)
+      transform_base_bc <- "None"
+      
       if (input$transform_base == "log2") {
         if (input$file_level == FALSE) {
           scp_0 <- logTransform(scp_0,
@@ -509,9 +513,10 @@ server <- function(input, output, session) {
           assay(scp_0[["proteins_transf"]]) <- protein_matrix
         } else {
           protein_matrix <- evidence_data
-          b <- boxcox_1(stats::lm(evidence_data ~ 1))
+          b <- boxcox_1(stats::lm(unlist(protein_matrix) ~ 1))
           # Exact lambda
           lambda <- b$x[which.max(b$y)]
+          print("lambda")
           print(lambda)
           
           if (round(lambda, digits = 0) == -2 || lambda < -1.5) {
@@ -562,9 +567,12 @@ server <- function(input, output, session) {
         }
       }
       
+      print("transformed data")
+      print(evidence_data)
+      
       incProgress(15/17, detail=paste("normalizing proteins"))
       req(input$norm_method)
-      if (input$norm_method == "SCoPE2" && input$transform_base == "log2" | input$transform_base == "log10") {
+      if (input$norm_method == "SCoPE2" && input$transform_base == "log2" | input$transform_base == "log10" | transform_base_bc == "log10") {
         if (input$file_level == FALSE) {
           # center cols with median
           scp_0 <- sweep(scp_0, i = "proteins_transf",
@@ -584,12 +592,14 @@ server <- function(input, output, session) {
           
         } else {
           # #normalize colwise
-          evidence_data <- sweep(evidence_data[,-1], 2, colMedians(evidence_data[,-1]), FUN="-")
+          protein_matrix <- as.matrix(evidence_data)
+          protein_matrix <- sweep(protein_matrix, 2, colMedians(protein_matrix), FUN="-")
           # #normalize rowwise
-          evidence_data <- sweep(evidence_data[,-1], 1, rowMeans(evidence_data[,-1]), FUN="-")
+          protein_matrix <- sweep(protein_matrix, 1, rowMeans(protein_matrix), FUN="-")
+          evidence_data <- data.frame(protein_matrix)
         }
         
[…]

### 2023-05-09 diff_exp_limma.R
expr set file input until batch c finish

@@ -1,24 +1,31 @@
 library(limma)
 library(dplyr)
 
-exp_matrix <- read.delim("/home/lukas/Downloads/Raw_data.txt")
+evidence_data <- read.delim("/home/lukas/Downloads/Raw_data.txt")
 
-design_matrix <- read.delim("/home/lukas/Downloads/Design.txt")
+meta_data_0 <- read.delim("/home/lukas/Downloads/Design.txt")
+
+rownames(evidence_data) <- evidence_data$ID
+evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
 # log transform expression values
-exp_matrix[,-1] <- log10(exp_matrix[,-1]) 
+evidence_data <- log10(evidence_data) 
 
-# #normalize colwise
-exp_matrix_norm <- sweep(exp_matrix[,-1], 2, colSums(exp_matrix[,-1]), FUN="-")
+protein_matrix <- as.matrix(evidence_data)
+protein_matrix <- sweep(protein_matrix, 2, colMedians(protein_matrix), FUN="/")
 # #normalize rowwise
-exp_matrix_norm <- sweep(exp_matrix[,-1], 1, rowSums(exp_matrix[,-1]), FUN="-")
- 
+protein_matrix <- sweep(protein_matrix, 1, rowMeans(protein_matrix), FUN="/")
+evidence_data <- data.frame(protein_matrix)
+
+
+evidence_data <- replace(evidence_data, is.na(evidence_data), median(unlist(evidence_data), na.rm = TRUE))
+
 library('corrr')
 library(ggcorrplot)
 library("FactoMineR")
 library(factoextra)
 
-corr_matrix <- cor(exp_matrix_norm[,-1])
+corr_matrix <- cor(evidence_data_norm[,-1])
 ggcorrplot(corr_matrix)
 
 # in this case a correlation within the batches can be clearly observed 
@@ -43,34 +50,34 @@ fviz_pca_ind(data.pca, col.ind = "cos2",
 )
 
 # batch effects obtained --> use combat to correct
-batch <- design_matrix$Batch
+batch <- meta_data_0$Batch
 
 library(sva)
 
-combat_edata1 = ComBat(dat=exp_matrix_norm, batch=batch)
+combat_edata1 = ComBat(dat=evidence_data_norm, batch=batch)
 
 # or use limma
-y2 <- removeBatchEffect(exp_matrix_norm, batch)
[…]

### 2023-05-09 scpR-app/app.R

@@ -567,9 +567,6 @@ server <- function(input, output, session) {
         }
       }
       
-      print("transformed data")
-      print(evidence_data)
-      
       incProgress(15/17, detail=paste("normalizing proteins"))
       req(input$norm_method)
       if (input$norm_method == "SCoPE2" && input$transform_base == "log2" | input$transform_base == "log10" | transform_base_bc == "log10") {
@@ -662,8 +659,6 @@ server <- function(input, output, session) {
         }
       }
       
-      print("normalized data")
-      print(evidence_data)
       
       if (input$file_level == TRUE) {
         peptide_file <- unique(meta_data_0$Batch)
@@ -679,11 +674,18 @@ server <- function(input, output, session) {
                             name = "proteins_imptd",
                             method = "knn",
                             k = 3, rowmax = 1, colmax= 1,
-                            maxp = Inf, rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
+                            maxp = Inf,
+                            rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
           } else {
+            evidence_data[evidence_data == 0] <- NA
             protein_matrix <- as.matrix(evidence_data)
-            protein_matrix <- impute.knn(protein_matrix, k=5, rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
-            evidence_data <- data.frame(protein_matrix)
+            protein_matrix <- impute.knn(protein_matrix,
+                                         k=5,
+                                         rowmax = 1,
+                                         colmax = 1,
+                                         maxp = Inf,
+                                         rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
+            evidence_data <- data.frame(protein_matrix$data)
           }
         } else if (input$missing_v == "drop rows") {
           if (input$file_level == FALSE) {
@@ -699,41 +701,51 @@ server <- function(input, output, session) {
             
             scp_0 <- filterNA(scp_0, pNA = 0, "proteins_imptd")
           } else {
-            evidence_data %>% drop_na(evidence_data)
+            evidence_data[evidence_data == 0] <- NA
+            evidence_data <- na.omit(evidence_data)
           }
 
         } else if (input$missing_v == "replace with mean") {
-          sce <- getWithColData(scp_0, "proteins_norm")
-          
-          scp_0 <- addAssay(scp_0,
[…]

### 2023-05-09 diff_exp_limma.R
until pca finished

@@ -8,6 +8,28 @@ meta_data_0 <- read.delim("/home/lukas/Downloads/Design.txt")
 rownames(evidence_data) <- evidence_data$ID
 evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
+selectedSampleType_to_exclude <- c("Pool")
+evidence_data <- evidence_data[, !(meta_data_0$Group %in%  selectedSampleType_to_exclude)]      
+meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
+
+
+barplot(log10(dim(evidence_data)[1]), main = "log Count of rows")
+
+evidence_data_medians <- data.frame(median_intensity = colMedians(as.matrix(evidence_data)))
+
+evidence_data_medians %>%
+  ggplot() +
+  aes(x = median_intensity, 
+      y = meta_data_0$Group,
+      fill = meta_data_0$Group) +
+  geom_boxplot() +
+  scale_x_log10() +
+  labs(fill=as.character(meta_data_0$Group), y=as.character(meta_data_0$Group))
+
+
+
+
+
 # log transform expression values
 evidence_data <- log10(evidence_data) 
 
@@ -21,11 +43,11 @@ evidence_data <- data.frame(protein_matrix)
 evidence_data <- replace(evidence_data, is.na(evidence_data), median(unlist(evidence_data), na.rm = TRUE))
 
 library('corrr')
-library(ggcorrplot)
+library("ggcorrplot")
 library("FactoMineR")
-library(factoextra)
+library("factoextra")
 
-corr_matrix <- cor(evidence_data_norm[,-1])
+corr_matrix <- cor(evidence_data)
 ggcorrplot(corr_matrix)
 
 # in this case a correlation within the batches can be clearly observed 
@@ -51,6 +73,18 @@ fviz_pca_ind(data.pca, col.ind = "cos2",
 
 # batch effects obtained --> use combat to correct
 batch <- meta_data_0$Batch
+
+library("scater")
+dimred_pca <- calculatePCA(evidence_data,                                          ncomponents = 5,
+                           ntop = Inf,
+                           scale = TRUE)
+
+dimred_pca <- data.frame(dimred_pca)
+
+ggplot(dimred_pca, aes(x=PC1, y=PC2)) +
+  geom_jitter()
+
+
 
 library(sva)
 

### 2023-05-09 scpR-app/app.R

@@ -255,6 +255,7 @@ server <- function(input, output, session) {
   meta_data <- reactive({
     req(input$sample_annotation_file)
     meta_data_0 <- read.delim(input$sample_annotation_file$datapath)
+    meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
     return(meta_data_0)
   })
   
@@ -413,6 +414,7 @@ server <- function(input, output, session) {
       } else {
         rownames(evidence_data) <- evidence_data$ID
         evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
+        evidence_data <- evidence_data[, !(meta_data_0$Group %in%  input$selectedSampleType_to_exclude)]
       } 
       
       incProgress(14/17, detail=paste("transforming protein data"))
@@ -739,54 +741,55 @@ server <- function(input, output, session) {
             evidence_data <- replace(evidence_data, is.na(evidence_data), median(unlist(evidence_data), na.rm = TRUE))
           }
         }
-      
-        print("imputed_data")
-        print(evidence_data)
-        
+
         incProgress(16/17, detail=paste("running batch correction"))
         if (input$batch_c == "ComBat") {
-          sce <- getWithColData(scp_0, "proteins_imptd")
-          batch <- colData(sce)$Raw.file
-          # can be used to aim batch correction for desired result
-          model <- model.matrix(~0 + SampleType, data = colData(sce))
-          
-          assay(sce) <- ComBat(dat = assay(sce),
-                               batch = batch)#,
-          #mod = model)
-          
-          scp_0 <- addAssay(scp_0,
-                            y = sce,
-                            name = "proteins_batchC")
-          
-          scp_0 <- addAssayLinkOneToOne(scp_0,
-                                        from = "proteins_imptd",
-                                        to = "proteins_batchC")
-          
-          sce <- getWithColData(scp_0, "proteins_batchC")
-          
-          scp_0 <- addAssay(scp_0,
-                            y = sce,
-                            name = "proteins_dim_red")
-          
-          scp_0 <- addAssayLinkOneToOne(scp_0,
-                                        from = "proteins_batchC",
-                                        to = "proteins_dim_red") 
-          
+          if (input$file_level == FALSE) {
+            sce <- getWithColData(scp_0, "proteins_imptd")
+            batch <- colData(sce)$Raw.file
[…]

### 2023-05-09 scpR-app/app.R
minor commit after bugfix

@@ -36,9 +36,9 @@ ui <- fluidPage(
       materialSwitch(inputId = "file_level", value = F, label="Handle preprocessed expression set", status = "danger"),
       fileInput("evidence_file", "Upload evidence.txt from MaxQuant", accept = c("text")),
       fileInput("sample_annotation_file", "Upload sample annotation file", accept = c("text")),
-      
+      selectInput("selectedSampleType_to_exclude", "Select SampleType to exclude", "", multiple=T),
+            
       conditionalPanel(condition = "!input.file_level", 
-                       selectInput("selectedSampleType_to_exclude", "Select SampleType to exclude", "", multiple=T),
                        # cutoff for PIF
                        numericInput("PIF_cutoff", "Input cutoff value for parental ion fraction. PSMs larger than the value remain", 0.1, min = 0, max=1, step = 0.1),
                        
@@ -255,7 +255,6 @@ server <- function(input, output, session) {
   meta_data <- reactive({
     req(input$sample_annotation_file)
     meta_data_0 <- read.delim(input$sample_annotation_file$datapath)
-    meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
     return(meta_data_0)
   })
   
@@ -414,7 +413,7 @@ server <- function(input, output, session) {
       } else {
         rownames(evidence_data) <- evidence_data$ID
         evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
-        evidence_data <- evidence_data[, !(meta_data_0$Group %in%  input$selectedSampleType_to_exclude)]
+        evidence_data <- evidence_data[!(meta_data_0$Group %in%  input$selectedSampleType_to_exclude), ]
       } 
       
       incProgress(14/17, detail=paste("transforming protein data"))

### 2023-05-09 markdown/Thesis.tex
pca finished. to do umap

@@
 
-\begin{tabular}{l|r}
-\hline
-& lambda & Transformation\$\\
-\hline
--2 & $\frac{1}{x^2}$
-\hline
--1 & $\frac{1}{x}$
-\hline
--0.5 & $\frac{1}{\sqrt{x}}$
-\hline
-0 & $\log{x}$
-\hline
-0.5 & $\sqrt{x}$
-\hline
-1 & $x$
-\hline
+\begin{center}
+\begin{tabular}{l r}
+$\lambda$ & Transformation \\
+-2 & $\frac{1}{x^2}$ \\
+-1 & $\frac{1}{x}$ \\
+-0.5 & $\frac{1}{\sqrt{x}}$ \\
+0 & $\log{x}$ \\
+0.5 & $\sqrt{x}$ \\
+1 & $x$ \\
 2 & $x^2$
 \end{tabular}
+\end{center}
 
@@
 explained briefly in this chapter. Starting with converting the data to
-the Z-distribution (Gaussian result), where 0 reflects the mean and 1
-refers to the standard deviation, until vectorization where every
-feature (=protein) can be explained as a vector which points in a
-specific direction in the unit sphere, normalization can have different
-meanings. Normalization is considered as part of the scaling procedure
-and ensures that every feature contributes equally to our statistical
-model and hinders large values from biasing the model in a particular
-direction. Although this could also be harmful for the modeling
-procedure, since the impact of important features on the dataset will
-also be reduced. Depending on the algorithms later used in the data
-processing, the method of choice for normalization can change the
-results significantly. The K-means clustering algorithm for instance
-relies on the distances between distinct data points by minimizing
-variance of the squared euclidean distances. This method will be applied
-in the missing value imputation which will be the next step applied to
-our data. Hence we want to make sure that the get the maximum accuracy
-for our imputed values. The first option to choose in the user interface
-is the column-wise median and row-wise mean normalization. This method
-was chosen by Specht et al in the SCoPE2 publication 2021
-/citep\{Specht2021\} and worked as a reference during the development
+the Z-distribution (Gaussian result), where 0 represents the mean and 1
+represents the standard deviation, in vectorization every feature (or
+protein) is represented as a vector that points in a specific direction
+in the unit sphere, normalization can have different meanings.
+Normalization is considered as part of the scaling procedure and ensures
+that every feature contributes equally to our statistical model and
+hinders large values from biasing the model in a particular direction.
+However, this could also have negative consequences for the modeling
+procedure, as it may reduce the impact of important features on the
+dataset. Depending on the algorithms later used in the data processing,
+the method of choice for normalization can change the results
+significantly. The K-means clustering algorithm, for instance relies on
+the distances between distinct data points by minimizing variance of the
+squared euclidean distances. This method will be applied in the missing
+value imputation which will be the next step applied to our data.
+Therefore, we want to ensure that we obtain the most accurate imputed
+values possible. The first normalization method available in the user
+interface is column-wise median and row-wise mean normalization. This
+method was chosen by Specht et al in the SCoPE2 publication 2021
+\citep{Specht2021} and worked as a reference during the development
 process to benchmark other methods. When dividing each value of an
 expression set by the median of the particular column, the procedure is
-considered as a column-wise median normalization. Hence the
-mean-normalization works the same way on the column. In the analysis
-pipeline both procedures were applied on the dataset. However when the
-logarithm as the transformation method was chosen in advance the
-pipeline automatically switches from a division to a subtraction, since:
-\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\)
+considered as a column-wise median normalization. Although the
+mean-normalization works the same way but takes the mean as the second
+variable. In the analysis pipeline both procedures were applied on the
+dataset. However when the logarithm as the transformation method was
+chosen as transformation method, the pipeline automatically switches
+from a division to a subtraction, since:
+\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\). After applying
+this normalization method, each value can be interpreted as a distance
+or fraction of the mean or median of the corresponding column or row,
+enabling fair comparisons between features and samples. The second
+method for normalization included is the CONSTANd method, which was
+proven suitable for relative quantification in the field of proteomics
+\citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq
+data. CONSTANd uses a technique called matrix raking, which employs the
+RAS algorithm. The expression set reflects the nonnegative real
+\((m, n)\) matrix \(A\) were the bi proportional constrained matrix
+problem will be solved by finding the \((m, n)\) matrix \(B\) which
+equals to \(diag(x) * A *diag(y)\) . Whereas \(x \in \mathbb{R}^{m}\)
+and \(y \in \mathbb{R}^{n}\). The solution to this problem involves
+finding the row sum of \(B=u_i\) and the column sum of \(B=v_j\), where
+\(i\) and \(j\) denote the row and column indices of the matrix,
+respectively \citep{Bacharach1965}. In other words, the matrix will be
+alternatively manipulated on rows and columns until the mean of both
+equals 1. In order to apply the method and yield true results, one has
+to consider 3 major assumptions between sample types. This assumption
+can be observed in the MA-plot which indicates the differences of two
+samples. If we have the intensities of two samples, \(R\) and \(G\), we
+can plot them on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\)
+and the y-axis as \(A=\frac{1}{2}\log_2(RG)\). A comparison between all
+sample types of identically processed sets is needed before calculation
+and the assumptions must be fulfilled.
 
-Further filter steps can include correction between multiple runs. This
-kind of process need the addition of a reference channel. However when
-observing a single run, these steps are not crucial for the upcoming
-analysis.
+\begin{itemize}
+\item{\textbf{1}} \textbf{The majority of proteins are not differentially expressed}
+\\Density of the dots decreases when looking from the middle of the cloud towards the edges.
+
+\item{\textbf{2}} \textbf{Up- and downregulation is balanced around the mean expression}
+\\The scatterplot of the data points is symmetrical around the mean on the horizontal axis.
+
+\item{\textbf{3}} \textbf{Systematic bias correlates with the magnitude of expression}
+\\The symmetry axis is approximately horizontal.
+
+\end{itemize}
+
+If the MA-plot does not visualize any of these violations the CONSTANd
+method can be applied on the dataset and the biological meaning of the
+sample types can be assessed. The method is utilized by the bioconductor
+package handler. However if the dataset already consists of balanced
+features and samples or the user wants to benchmark the methods applied,
+normalization can be skipped also.
+
+\hypertarget{missing-value-handling}{%
+\paragraph{Missing value handling}\label{missing-value-handling}}
+
+When analyzing multiple batches simultaneously, it is possible to
+encounter situations where a particular protein is not detected in one
+of the batches. From a biological perspective, it is highly unlikely
+that a cell completely lacks a single protein, as it would imply a
+complete loss of function for that protein. Therefore, an intensity
+value of 0 is close to impossible for most proteins. In order to handle
+this dark space in the expression set, several methods can be chosen in
+the application. A variety of function employs replacing the missing
+values with the mean or median of the rest of the matrix. This procedure
+introduces no bias in the data if the missingness is low, although
+differential expression for the sample carrying the missing feature can
+not be expected as well. Even a more conservative approach is dropping
+the rows for the feature with the missing value in one of the observed
+samples. A common practice in proteomics is utilizing the K-nearest
+neighbor algorithm (=KNN), which imputes the missing values by a
+euclidean metrics of the neighboring values in the columns where the
+feature is not missing. \(K\) denotes for the count of neighboring
+values of a gene and can be selected upon preference. The euclidean
+distance between two points is defined as \(d = \sqrt{a^2 * b^2}\), the
+\(K\) nearest neighbors average value will be assigned to the intensity
+of the missing feature. Since the average is a statistical metric which
+can be biased easily by skewed distributions, normalization is an
+important task before performing missing value imputation. KNN was
+benchmarked for the regression of protein abundance intensities and
+showed an Area under the Curve (AUC) above 0.7 when testing on a human
+proteome test set \citep{Lan2013}.
+
+\hypertarget{batch-correction}{%
+\paragraph{Batch correction}\label{batch-correction}}
+
+Since mass spectrometry experiments are highly biased by several
+factors, such as sample processing, the technician, manufacturer and
+production of the device (chromatography column, mass-spectrometer), the
+reagents etc., working in multiple batches introduces further variance
+in the protein data. This variance would overshadow the biological
+differences across sample types. Before attempting to remove batch
+effects it is recommended to elucidate the variance with a
+dimensionality reduction technique which will be explained in the next
+chapter. To remove the technical factor also called non-biological
+experimental variation the R package sva \citep{Leek2012} with the
+ComBat function \citep{Johnson2007} was used. The expression of a gene
+\(g\) for the batch \(m_i\) and sample \(j\) is defined as
+\(Y_i_j_g = a_g + X\beta_g + \gamma_i_g + \delta_i_g\epsilon_i_j_g\),
+whereas \(\gamma\) and \(\delta\epsilon\) denote for the systematic
+error caused by the batch. In order to remove the systematic error the
+bayesian framework shrinks the effect by pooling information over the
+features \(g\). Ideally the effects \(a\) and \(\beta\) remain as
+natural as they are and should reflect the true biological difference
+between experimental obtained samples or groups. ComBat standardizes the
+data by a Z-transformation before attempting to remove the batch effect.
+By the method of moments the two batch effect parameters \(\gamma\) and
+\(\delta\epsilon\) were estimated and can be subtracted from each value
+for the particular feature.
 
@@
 
+After cleaning the data and crystallizing from impurities such as noise
+and unwanted effects, the biological differences between sample types
+can be obtained. Proteomic data consists of multi dimensions (=features)
+which are overall hard to understand by a glance. An approach to observe
+the differences without a closer sight of exact effect on the
+proteinaceous ensemble, but still keep a high level of information can
+be dimensionality reduction.
+
 \hypertarget{principle-component-analysis-pca}{%
 \subparagraph{Principle component analysis
-(PCA)}\label{principle-component-analysis-pca}}
+(=PCA)}\label{principle-component-analysis-pca}}
 
-Protein levels can be projected to their principle components (PC) and
-clustered to their specific cell type. So cell types are distinguished
-by their proteinaceous composition.
+The most common approach in data science for any kind of high
+dimensional data is the PCA. Every principle component can be understood
+as a vector pointing in a specific direction derived as the eigenvector
+from the covariance matrix of the expression set. The covariance matrix
+indicates for the co linearity between the variance of all elements
+compared. After calculating the covariance matrix the eigenvectors for
+the covariance matrix will be obtained by solving the quadratic
+equations to obtain the eigenvalues \(\lambda\). Since
+\(A v = \lambda v\) the eigenvector \(v\) can be obtained by solving the
+equation for each element of the vector. The eigenvectors \(v\) never
+changes is direction and explains the covariance within the dataset and
+so also the mathematical differences between the samples. The
+eigenvectors are sorted according to their eigenvalues in decreasing
+order and therefore decreasing explained variance. For graphical
+visualizations the principle components are plotted against each other.
+The first two components, which explain the majority of variance in the
+dateset highlight the difference between samples. After the cleaning
+procedures any additional effect caused by technical or biological
+replication should be removed, however it could be found be highlighting
+according to the particular factor. This makes the PCA to a useful tool
+before starting any statistical analysis by checking for involuntarily
+introduced bias.
 

### 2023-05-10 markdown/Thesis.tex
until feature wise output genex table input finished

@@
 \(g\) for the batch \(m_i\) and sample \(j\) is defined as
-\(Y_i_j_g = a_g + X\beta_g + \gamma_i_g + \delta_i_g\epsilon_i_j_g\),
+\(Y_i_j_g = a_g + X \beta_g + \gamma_i_g + \delta_i_g \epsilon_i_j_g\),
 whereas \(\gamma\) and \(\delta\epsilon\) denote for the systematic

### 2023-05-10 markdown/methods.Rmd
until feature wise output genex table input finished

@@
 To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used.
-The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined
+The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_i_j_g = a_g + X \beta_g + \gamma_i_g + \delta_i_g \epsilon_i_j_g$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch.
+In order to remove the systematic error the bayesian framework shrinks the effect by pooling information over the features $g$.
+Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups.
+ComBat standardizes the data by a Z-transformation before attempting to remove the batch effect.
+By the method of moments the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature.
 
 #### Downstream Analysis
+After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained.
+Proteomic data consists of multi dimensions (=features) which are overall hard to understand by a glance.
+An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keep a high level of information can be dimensionality reduction.
 
-##### Principle component analysis (PCA)
-
-Protein levels can be projected to their principle components (PC) and clustered to their specific cell type. So cell types are distinguished by their proteinaceous composition.
+##### Principle component analysis (=PCA)
+The most common approach in data science for any kind of high dimensional data is the PCA.
+Every principle component can be understood as a vector pointing in a specific direction derived as the eigenvector from the covariance matrix of the expression set.
+The covariance matrix indicates for the co linearity between the variance of all elements compared.
+After calculating the covariance matrix the eigenvectors for the covariance matrix will be obtained by solving the quadratic equations to obtain the eigenvalues $\lambda$.
+Since $A v = \lambda v$ the eigenvector $v$ can be obtained by solving the equation for each element of the vector.
+The eigenvectors $v$ never changes is direction and explains the covariance within the dataset and so also the mathematical differences between the samples.
+The eigenvectors are sorted according to their eigenvalues in decreasing order and therefore decreasing explained variance.
+For graphical visualizations the principle components are plotted against each other.
+The first two components, which explain the majority of variance in the dateset highlight the difference between samples.
+After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor.
+This makes the PCA to a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.
 

### 2023-05-10 diff_exp_limma.R

@@ -1,20 +1,27 @@
 library(limma)
 library(dplyr)
-
+library(ggplot2)
+library(matrixStats)
+# load files
 evidence_data <- read.delim("/home/lukas/Downloads/Raw_data.txt")
 
 meta_data_0 <- read.delim("/home/lukas/Downloads/Design.txt")
 
+
+# handle id column
 rownames(evidence_data) <- evidence_data$ID
 evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
+# handle exclusion dependencies
 selectedSampleType_to_exclude <- c("Pool")
 evidence_data <- evidence_data[, !(meta_data_0$Group %in%  selectedSampleType_to_exclude)]      
 meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
 
 
+# plot basic counts
 barplot(log10(dim(evidence_data)[1]), main = "log Count of rows")
 
+# plot median intensities 
 evidence_data_medians <- data.frame(median_intensity = colMedians(as.matrix(evidence_data)))
 
 evidence_data_medians %>%
@@ -26,9 +33,41 @@ evidence_data_medians %>%
   scale_x_log10() +
   labs(fill=as.character(meta_data_0$Group), y=as.character(meta_data_0$Group))
 
+# plot covariance
+heatmap(cor(t(evidence_data)))
+
+# feature wise output across channels
+selectedProtein <- "P26640"
+to_plot <- data.frame(t(evidence_data[selectedProtein, ]))
 
 
+ggplot(to_plot, aes(x = rownames(to_plot), y = get(selectedProtein))) + 
+  geom_col(fill = meta_data_0$Group) +
+  labs(y=as.character(selectedProtein), x="")
 
+# plot featurewise intensities for each channel
+subsetByFeature(scp_0, input$selectedProtein) %>%
+  ## Format the `QFeatures` to a long format table
+  longFormat(colvars = c("Raw.file", "SampleType", "Channel")) %>%
+  data.frame %>%
+  ## This is used to preserve ordering of the samples and assays in ggplot2
+  mutate(assay = factor(assay, levels = names(scp_0)),
+         Channel = sub("Reporter.intensity.", "", Channel)) %>%
+  mutate(Channel = as.numeric(Channel)) %>%
+  arrange(Channel) %>%
+  mutate(Channel = factor(Channel, levels = unique(Channel))) %>%
+  ## Start plotting
[…]

### 2023-05-10 scope2.R

@@ -25,9 +25,9 @@ library("sva")
 library("tibble")
 
 # read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/20230202_HF2_08_Ujjwal_Monocytes_TMT12_PSMs.txt")
+mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")
 
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Monocytes/gerhard_pd/sampleAnnotate_tabdel.txt")
+sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/sampleAnnotation_tabdel.txt")
 
 
 # create QFeature object
@@ -807,6 +807,8 @@ p <- ggplot(toptable, aes(x = logFC, y = -log10(P.Value), text=protein)) +
   theme(legend.position = "top")
   
 ggplotly(p, tooltip = "text")
+
+heatmap(cor(t(assay(scp_0[["proteins_dim_red"]]))))
 
 
 

### 2023-05-10 scpR-app/app.R

@@ -34,8 +34,8 @@ ui <- fluidPage(
       
       # read in of the data
       materialSwitch(inputId = "file_level", value = F, label="Handle preprocessed expression set", status = "danger"),
-      fileInput("evidence_file", "Upload evidence.txt from MaxQuant", accept = c("text")),
-      fileInput("sample_annotation_file", "Upload sample annotation file", accept = c("text")),
+      fileInput("evidence_file", "Upload tabdel (.txt) PSM file", accept = c("text")),
+      fileInput("sample_annotation_file", "Upload tabdel (.txt) sample annotation file", accept = c("text")),
       selectInput("selectedSampleType_to_exclude", "Select SampleType to exclude", "", multiple=T),
             
       conditionalPanel(condition = "!input.file_level", 
@@ -75,9 +75,11 @@ ui <- fluidPage(
                   tabPanel("Reporter Ion Intensity", 
                            selectInput("color_variable_ri", "select variable to indicate", ""),
                            plotOutput("RI_intensity")),
-                  tabPanel("Covariance across razor peptides",
-                           selectInput("color_variable_cv", "select variable to indicate", ""),
-                           plotOutput("CV_median")),
+                  tabPanel("Covariance and correlation",
+                           conditionalPanel(condition = "!input.file_level",
+                                            selectInput("color_variable_cv", "select variable to indicate", ""),
+                                            plotOutput("CV_median")),
+                           plotOutput("corr_matrix"), width="100%"),
                   tabPanel("Dimensionality reduction",
                            fluidRow(width=12,
                                     column(width=4,
@@ -412,8 +414,9 @@ server <- function(input, output, session) {
         
       } else {
         rownames(evidence_data) <- evidence_data$ID
-        evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
-        evidence_data <- evidence_data[!(meta_data_0$Group %in%  input$selectedSampleType_to_exclude), ]
+        evidence_data <- evidence_data[, !(names(evidence_data) %in% c("ID"))]
+        evidence_data <- evidence_data[, !(meta_data_0$Group %in%  input$selectedSampleType_to_exclude)]
+        meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  input$selectedSampleType_to_exclude), ]
       } 
       
[…]

### 2023-05-10 markdown/methods.Rmd
umap halfway explained

@@
 To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used.
-The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_i_j_g = a_g + X \beta_g + \gamma_i_g + \delta_i_g \epsilon_i_j_g$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch.
+The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch.
 In order to remove the systematic error the bayesian framework shrinks the effect by pooling information over the features $g$.
@@
 
-#### Downstream Analysis
+#### Dimensionality reduction
 After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained.
@@
 An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keep a high level of information can be dimensionality reduction.
+After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor.
+This makes the dimensionality reduction a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.
 
-##### Principle component analysis (=PCA)
+##### Principle component analysis (PCA)
 The most common approach in data science for any kind of high dimensional data is the PCA.
+For the pipeline the bioconductor package scater \citep{McCarthy2017} was used.
 Every principle component can be understood as a vector pointing in a specific direction derived as the eigenvector from the covariance matrix of the expression set.
@@
 The first two components, which explain the majority of variance in the dateset highlight the difference between samples.
-After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor.
-This makes the PCA to a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.
+
+##### Uniform manifold approximation & projection (UMAP)
+UMAP \citep{McInnes2018} is a dimensionality reduction technique that was developed in recent years and is based on Riemann geometry.
+It shares similarities to the  t-distributed stochastic neighbor embedding (t-SNE) in terms of graphical visualization.
+Riemann geometry is a mathematical framework that deals with three dimensional functions using volumes, areas and vectors, which are also concept of topological data analysis.
+Before performing UMAP, three assumptions need to be considered:
+
+\begin{enumerate}
+\item There exists a manifold on which the data would be uniformly distributed.
+$\rightarrow$ This assumption implies that the probability for each value in the biological dataset is the same.
+
+\item The underlying manifold of interest is locally connected.
+$\rightarrow$ In other words, similar samples will have a similar proteinaceous ensemble.
+
+\item Preserving the topological structure of this manifold is the primary
+goal.
+$\rightarrow$ This means that the homeomorphic structure (the property of being able to transform one shape into another without tearing or gluing) is locally preserved, and local values in the dataset can be compared with each other.
+\end{enumerate}
+\citep{McInnes2018}
+
+The UMAP algorithm begins by constructing a graph with a defined neighborhood parameter, typically denoted as $k$, which can be selected computationally.
+UMAP employs the k-nearest neighbor descent algorithm (NN-descent).
+Similar to the k-nearest neighbor algorithm (KNN), used in the missing value imputation part of this project, NN-descent recursively iterates to find the smallest distance to the respective neighbors within the defined neighborhood size $k$.
+Increasing the value of $k$ gradually strengthens the clustering of the data, but at a certain point, fine structure may be lost, and only rough estimates of the underlying principle can be observed.
+The mathematical graph in UMAP is defined by nodes connected by edges, with each node internally connected to at least one other node.
+Edges are distinguished from each other not only by their connection but also by their importance, which is often referred to as weight.
+
+The application utilizes the bioconductor package scater for performing UMAP \citep{McCarthy2017}.
 

### 2023-05-10 diff_exp_limma.R

@@ -146,6 +146,7 @@ ggplot(dimred_umap, aes(x=UMAP1,
                         size = meta_data_0$body_weight)) +
   geom_point(alpha=3/4)
 
+plot_ly(x=dimred_umap$UMAP1, y=dimred_umap$UMAP2, z=dimred_umap$UMAP3, type="scatter3d", mode="markers", color=meta_data_0$Batch)
 
 
 library(sva)

### 2023-05-11 diff_exp_limma.R
3d rep of PCA finished

@@ -19,7 +19,7 @@ meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclu
 
 
 # plot basic counts
-barplot(log10(dim(evidence_data)[1]), main = "log Count of rows")
+#barplot(log10(dim(evidence_data)[1]), main = "log Count of rows")
 
 # plot median intensities 
 evidence_data_medians <- data.frame(median_intensity = colMedians(as.matrix(evidence_data)))
@@ -146,7 +146,12 @@ ggplot(dimred_umap, aes(x=UMAP1,
                         size = meta_data_0$body_weight)) +
   geom_point(alpha=3/4)
 
-plot_ly(x=dimred_umap$UMAP1, y=dimred_umap$UMAP2, z=dimred_umap$UMAP3, type="scatter3d", mode="markers", color=meta_data_0$Batch)
+plot_ly(x=dimred_umap$UMAP1,
+        y=dimred_umap$UMAP2, 
+        z=dimred_umap$UMAP3, 
+        type="scatter3d", 
+        mode="markers",
+        color=meta_data_0$Batch) 
 
 
 library(sva)

### 2023-05-11 scope2.R

@@ -559,7 +559,7 @@ plotReducedDim(scp[["proteins_dim_red"]],
 
 # UMAP
 scp[["proteins_dim_red"]] <- runUMAP(scp[["proteins_dim_red"]],
-                                   ncomponents = 2,
+                                   ncomponents = 3,
                                    ntop = Inf,
                                    scale = TRUE,
                                    exprs_values = 1,
@@ -808,7 +808,19 @@ p <- ggplot(toptable, aes(x = logFC, y = -log10(P.Value), text=protein)) +
   
 ggplotly(p, tooltip = "text")
 
+
+
+
 heatmap(cor(t(assay(scp_0[["proteins_dim_red"]]))))
 
+dimred <- data.frame(reducedDim(scp_0[["proteins_dim_red"]], "PCA"))
+plot_ly(x=dimred$PC1,
+        y=dimred$PC2,
+        z=dimred$PC3,
+        type="scatter3d",
+        mode="markers",
+        color=scp$Raw.file)
 
+selection <- "Channel"
 
+scp_0[["proteins_dim_red"]][[selection]]

### 2023-05-11 scpR-app/app.R

@@ -2,6 +2,7 @@ library("shiny")
 library("shinyWidgets")
 library("shinydashboard")
 library("DT")
+library("spsComps")
 
 library("scp")
 library("SingleCellExperiment")
@@ -70,7 +71,9 @@ ui <- fluidPage(
     mainPanel(
       #create tabs
       tabsetPanel(type = "tabs",
-                  tabPanel("Overview", plotOutput("overview_plot")),
+                  tabPanel("Overview",
+                           plotOutput("overview_plot"),
+                           verbatimTextOutput("debug")),
                   tabPanel("Summary Barplot", plotOutput("summary_bar")),
                   tabPanel("Reporter Ion Intensity", 
                            selectInput("color_variable_ri", "select variable to indicate", ""),
@@ -81,17 +84,29 @@ ui <- fluidPage(
                                             plotOutput("CV_median")),
                            plotOutput("corr_matrix"), width="100%"),
                   tabPanel("Dimensionality reduction",
+                           materialSwitch(inputId = "third_dim", value = F, label="Want to enter the 3rd dimension?", status = "danger"),
                            fluidRow(width=12,
                                     column(width=4,
                                            selectInput("color_variable_dim_red", "select variable to color", "")),
-                                    column(width = 4,
-                                           selectInput("shape_variable_dim_red", "select variable to shape", "")),
-                                    column(width=4,
-                                           selectInput("size_variable_dim_red", "select variable to size", ""))
+                                    conditionalPanel(condition = "!input.third_dim",
+                                                     column(width = 4,
+                                                            selectInput("shape_variable_dim_red", "select variable to shape", "")),
+                                                     column(width=4,
+                                                            selectInput("size_variable_dim_red", "select variable to size", ""))
+                                    )
                                     ),
-                           fluidPage(
-                           plotOutput("PCA"), 
-                           plotOutput("UMAP"))
+                           conditionalPanel(condition = "!input.third_dim",
+                                            fluidPage(
+                                              plotOutput("PCA"), 
+                                              plotOutput("UMAP")
+                                              )
[…]

### 2023-05-15 diff_exp_limma.R
analysis for expression table finished

@@ -12,6 +12,9 @@ meta_data_0 <- read.delim("/home/lukas/Downloads/Design.txt")
 rownames(evidence_data) <- evidence_data$ID
 evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
+
+
+
 # handle exclusion dependencies
 selectedSampleType_to_exclude <- c("Pool")
       
@@ -62,6 +65,65 @@ evidence_data <- data.frame(protein_matrix)
 
 
 evidence_data <- replace(evidence_data, is.na(evidence_data), median(unlist(evidence_data), na.rm = TRUE))
+
+
+# test MA plot and constand
+
+MAplot <- function(x,y,use.order=FALSE, R=NULL, cex=1.6, showavg=TRUE) {
+  # catch unequal size of matrices
+  if (dim(x)[2] != dim(y)[2]) {
+    if (dim(x)[2] > dim(y)[2]) {
+      x <- x[, 1:dim(y)[2]]
+    } else if (dim(y)[2] > dim(x)[2]) {
+      y <- y[, 1:dim(x)[2]]
+    }
+  }
+  
+  # make an MA plot of y vs. x that shows the rolling average,
+  M <- log2(y/x)
+  xlab = 'A'
+  if (!is.null(R)) {r <- R; xlab = "A (re-scaled)"} else r <- 1
+  A <- (log2(y/r)+log2(x/r))/2
+  if (use.order) {
+    orig.order <- order(A)
+    A <- orig.order
+    M <- M[orig.order]
+    xlab = "original rank of feature magnitude within IPS"
+  }
+  # select only finite values
+  use <- is.finite(M)
+  A <- A[use]
+  M <- M[use]
+  # plot
+  print(var(M))
+  plot(A, M, xlab=xlab, cex.lab=cex, cex.axis=cex)
+  # rolling average
+  if (showavg) { lines(lowess(M~A), col='red', lwd=5) }
+}
+
+exp_matrix_0 <- as.matrix(evidence_data)
+colnames(exp_matrix_0) <- meta_data_0$Group
+
+user_choice <- "EC-HC"
+# split user choice of comp back to sample types
+user_choice_vector <- strsplit(user_choice, split = "-")
+# and assign them to a variable
+choice_A <- user_choice_vector[[1]][1]
+choice_B <- user_choice_vector[[1]][2]
+
[…]

### 2023-05-15 scpR-app/app.R

@@ -879,9 +879,9 @@ server <- function(input, output, session) {
       colnames(exp_matrix_0) <- scp_0$SampleType  
     } else {
       req(meta_data())
-      exp_matrix_0 <- scp()
+      exp_matrix_0 <- data.frame(scp())
       meta_data_0 <- meta_data()
-      colnames(exp_matrix_0) <- meta_data_0$group
+      colnames(exp_matrix_0) <- meta_data_0$Group
     }
     return(exp_matrix_0)
   })
@@ -898,7 +898,7 @@ server <- function(input, output, session) {
   })
   
   # reactive element for the protein list --> update if the scp object changes
-  protein_list <- reactive({
+  protein_list <- eventReactive(input$update_button, {
     req(scp())
       if (input$file_level == FALSE) {
         req(scp()[["proteins"]])
@@ -938,7 +938,7 @@ server <- function(input, output, session) {
     return(list)
   })
 
-  columns <- reactive({
+  columns <- eventReactive(input$update_button, {
     if (input$file_level == FALSE) {
       req(scp())
       scp_0 <- scp()
@@ -956,7 +956,7 @@ server <- function(input, output, session) {
     updateSelectInput(session, "selectedSampleType_to_exclude", choices = sample_types())
   })
   
-  sample_types <- reactive({
+  sample_types <- eventReactive(input$update_button, {
     if (input$file_level == FALSE) {
       req(scp())
       scp_0 <- scp()
@@ -1360,23 +1360,7 @@ server <- function(input, output, session) {
   # plot MA plot for the constand method
   output$MAplot <- renderPlot({
     req(input$selectedComp)
-    req(scp())
-    
-    scp_0 <- scp() 
-    scp_0 <- filterNA(scp_0, pNA = 0, "proteins_transf")
-    
     user_choice <- input$selectedComp
-    # split user choice of comp back to sample types
-    user_choice_vector <- strsplit(user_choice, split = "-")
-    # and assign them to a variable
-    choice_A <- user_choice_vector[[1]][1]
-    choice_B <- user_choice_vector[[1]][2]
-    
-    #find the row indeces of corresponding to the individual sample types
-    st_indeces <- split(seq_along(scp_0$SampleType), scp_0$SampleType)
-    index_A <- st_indeces[choice_A]
-    index_B <- st_indeces[choice_B]
-    
     
[…]

### 2023-05-15 markdown/methods.Rmd
UMAP explained

@@
 Edges are distinguished from each other not only by their connection but also by their importance, which is often referred to as weight.
-
+Whereas the weight function is calculated as $w((x_i, x_{i_{j}})) = \exp(\frac{-max(0,d(x_{i},x_{i_{j}}-\rho_{i})}{\sigma_{i}}$ and determine the parameter $\rho$ for the weight function that: $\rho_i=\min \left\{ d(x_i,x_{i_{j}}) \,|\, 1 \leq j \leq k, d(x_i, x_{i_{j}}) > 0 \right\}$.
+Furthermore $\sigma$ will be set to meet the following conditions: $\sum_{j=1}^{k} \exp ( \frac{ -\max(0, d(x_{i}, x_{i_{j}}) - \rho_i)} { \sigma_i } ) = \log_2(k)$ and acts as a normalisation factor.
+The obtained graph $\overline{G}$ is called the fuzzy simplicial set which is obtained by approximating the geodesic distance (= locally length-minimizing curve) of the data points, leading to a topological approximation.
+The second phase of the algorithm adjusts the layout of the weighted graph to a representative view, but preserves the characteristics and shows the underlying topological principle of the data.
+In order to obtain a visual representation of the graph, the algorithm applies repulsive forces $\frac{2b}{(\epsilon + ||y_i-y_j||_{2}^{2})(1+a||y_i-y_j||_{2}^{2b})}(1-w((x_i,x_j)))(y_i-y_j)$ among vertices and attractive forces
+$\frac{ -2ab || y_i -y_j||_{2}^{2(b-1)} }{1+||y_i-y_j||_{2}^{2}} w((x_i,x_j))(y_i-y_j)$ along edges.
+Where $a$ and $b$ are denoted as hyperparameters.
+By iterating through possible conformations until converging towards a local minimum with decreasing repulsive and attractive forces, the UMAP reaches it`s aim to reveal the underlying properties of the data.
 The application utilizes the bioconductor package scater for performing UMAP \citep{McCarthy2017}.
@@
 The next step is creating a linear model between groups by log-ratios of their expression values.
+The general aim of UMAP is transforming the data to a toplogical representation of the initial dataset.

### 2023-05-15 diff_exp_limma.R

@@ -25,34 +25,34 @@ evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data
 # plot basic counts
 #barplot(log10(dim(evidence_data)[1]), main = "log Count of rows")
 
-# plot median intensities 
-evidence_data_medians <- data.frame(median_intensity = colMedians(as.matrix(evidence_data)))
-
-evidence_data_medians %>%
-  ggplot() +
-  aes(x = median_intensity, 
-      y = meta_data_0$Group,
-      fill = meta_data_0$Group) +
-  geom_boxplot() +
-  scale_x_log10() +
-  labs(fill=as.character(meta_data_0$Group), y=as.character(meta_data_0$Group))
+# # plot median intensities 
+# evidence_data_medians <- data.frame(median_intensity = colMedians(as.matrix(evidence_data)))
+# 
+# evidence_data_medians %>%
+#   ggplot() +
+#   aes(x = median_intensity, 
+#       y = meta_data_0$Group,
+#       fill = meta_data_0$Group) +
+#   geom_boxplot() +
+#   scale_x_log10() +
+#   labs(fill=as.character(meta_data_0$Group), y=as.character(meta_data_0$Group))
 
 # plot covariance
 #heatmap(cor(t(evidence_data)))
 
 # feature wise output across channels
-selectedProtein <- "P26640"
-to_plot <- data.frame(t(evidence_data[selectedProtein, ]))
-to_plot$names <- rownames(to_plot)
-
-library(reshape2)
-
-d = melt(to_plot, id.vars = "names")
-
-ggplot(data = d,
-       mapping = aes(x = names, y = value, fill=meta_data_0$Batch)) + 
-  geom_col(position = position_dodge()) +
-  labs(y=paste("intensity of", selectedProtein))
+# selectedProtein <- "P26640"
+# to_plot <- data.frame(t(evidence_data[selectedProtein, ]))
+# to_plot$names <- rownames(to_plot)
+# 
+# library(reshape2)
+# 
+# d = melt(to_plot, id.vars = "names")
+# 
+# ggplot(data = d,
+#        mapping = aes(x = names, y = value, fill=meta_data_0$Batch)) + 
+#   geom_col(position = position_dodge()) +
+#   labs(y=paste("intensity of", selectedProtein))
 
 # log transform expression values
 evidence_data <- log10(evidence_data) 
[…]

### 2023-05-16 diff_exp_limma.R
basic enrichment finished

@@ -3,9 +3,9 @@ library(dplyr)
 library(ggplot2)
 library(matrixStats)
 # load files
-evidence_data <- read.delim("/home/lukas/Downloads/Raw_data.txt")
+evidence_data <- read.delim("/home/lukas/Desktop/MS-Data/example/expression_matrix/Raw_data.txt")
 
-meta_data_0 <- read.delim("/home/lukas/Downloads/Design.txt")
+meta_data_0 <- read.delim("/home/lukas/Desktop/MS-Data/example/expression_matrix/Design.txt")
 
 
 # handle id column
@@ -58,9 +58,9 @@ evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data
 evidence_data <- log10(evidence_data) 
 
 protein_matrix <- as.matrix(evidence_data)
-protein_matrix <- sweep(protein_matrix, 2, colMedians(protein_matrix), FUN="/")
+protein_matrix <- sweep(protein_matrix, 2, colMedians(protein_matrix), FUN="-")
 # #normalize rowwise
-protein_matrix <- sweep(protein_matrix, 1, rowMeans(protein_matrix), FUN="/")
+protein_matrix <- sweep(protein_matrix, 1, rowMeans(protein_matrix), FUN="-")
 evidence_data <- data.frame(protein_matrix)
 
 
@@ -213,15 +213,11 @@ boxplot(as.data.frame(y2),main="Batch corrected with Limma")
 boxplot(as.data.frame(combat_edata1),main="Batch corrected with Combat")
 
 # Perform differential expression analysis between each of the two groups
+evidence_data_de <- combat_edata1
 
-# drop pool columns
-evidence_data_de <- data.frame(combat_edata1) %>% dplyr::select(-starts_with('Control'))
-evidence_data_de <- as.matrix(evidence_data_de)
-rownames(evidence_data_de) <- (evidence_data$ID)
 
 # create design matrix consisting of group variables
-design_samples <- subset(meta_data_0, Group != "Pool")
-design <- model.matrix(~0+factor(design_samples$Group))
+design <- model.matrix(~0+factor(meta_data_0$Group))
 colnames(design) <- c("EC", "HC", "VP")
 
 # Fit the expression matrix to a linear model
[…]

### 2023-05-16 scpR-app/app.R

@@ -141,6 +141,16 @@ ui <- fluidPage(
                            ),
                            plotlyOutput("volcano"),
                            DT::dataTableOutput("protein_table")
+                  ),
+                  tabPanel("Gene set enrichment",
+                           fluidRow(width=12,
+                                    column(width=4,
+                                           selectInput("p_value_correction_gsea", "select method for p-value correction", choices = c("none", "BH", "fdr", "BY", "holm"))),
+                                    column(width = 4,
+                                           numericInput("p_value_cutoff_gsea", "choose p-value cutoff", value = 0.05, min = 0, step = 0.01)),
+                                    column(width=4,
+                                           numericInput("fold_change_cutoff_gsea", "choose fold change cutoff", value = 0.05, min = 0, step = 0.01)),
+                           plotOutput("gsea"))
                   )
       )
     )
@@ -1719,6 +1729,48 @@ server <- function(input, output, session) {
     req(stat_result())
     vennDiagram(decideTests(stat_result(), p.value=input$p_value_cutoff, adjust.method=input$p_value_correction))
   })
+  
+  
+  # gene set enrichment analysis 
+  
+  result_kegg <- reactive({
+    req(protein_table())
+    tt <- protein_table()
+    
+    # filter the data according to user selection
+    req(input$p_value_cutoff_gsea)
+    req(input$fold_change_cutoff_gsea)
+    p_cut <- input$p_value_cutoff_gsea
+    fc_cut <- input$fold_change_cutoff_gsea
+    
+    mask <- tt$adj.P.Val < p_cut & 
+      abs(tt$logFC) > fc_cut
+    
+    deGenes <- rownames(tt)[mask]
+
+    req(input$p_value_correction_gsea)
+    p_correct <- input$p_value_correction_gsea
+    
+    ans.kegg <- enrichKEGG(
+      deGenes,
+      organism = "hsa",
+      keyType = "uniprot",
+      pvalueCutoff = p_cut,
+      pAdjustMethod = p_correct,
+      minGSSize = 10,
+      maxGSSize = 500,
+      qvalueCutoff = 0.2,
+      use_internal_data = FALSE
+    )
+  })
+  
+  output$gsea<- renderPlot({
+    req(result_kegg())
+    ans.kegg <- result_kegg()
+    
+    dotplot(ans.kegg, showCategory=20) + ggtitle("KEGG")
+  })
+  
 }
 
 

### 2023-05-16 markdown/methods.Rmd
limma linear model finished
to do:
explain and understand eBayes function in detail

@@
 
-#### Testing for differential expression
+#### Statistics
+After visualizing the biological tendency in the dataset, the next step is hypothesis testing.
+Several experimental designs can be tested against the null hypothesis $h_0$, which states that there is no significant difference between the sample types.
+Before proceeding with any statistical testing, it is important to observe the data using either a quantile-quantile norm plot (qq-norm plot) or a histogram.
+The quantile-quantile plot (qq-plot) serves a vital role in exploratory statistics by visualizing whether two datasets follow the same underlying distributions.
+In the case of statistical preparation for parametric tests, the first distribution is the dataset of interest, while the second dataset is the normal distribution.
+If the dataset follows an exact normal distribution, the values will align with the red line in the chart.
+However, values at the ends of the distribution may deviate from the normality line, indicating differentially expressed genes.
+These genes can be observed both in the histogram and the qq-norm plot.
 
-In order to test for differential expression the package limma from R bioconductor was used.
-\citep{Phipson2016}.
-The package uses a linear model approach to define a fold expression between sample groups.
-Before starting the analysis two matrices need to be computed.
-The design matrix identifies samples according to the sample type and defines the experimental design.
-In order to create the matrix automatically a simple algorithm is used within the programmed backend logic.
-The expression matrix contains intensities for each identified protein will be obtained at the end of the pipeline.
-A function call with the arguments design and expression matrix calculates the contrast matrix.
-The next step is creating a linear model between groups by log-ratios of their expression values.
-The general aim of UMAP is transforming the data to a toplogical representation of the initial dataset.
+Parametric statistical tests rely on the assumption of normality in the data, which needs to be achieved before building a model.
+To achieve normality, it is recommended to try various combinations of transformation and normalization methods.
+By observing the results in the plots, one can determine which combination of data processing methods satisfies the conditions required for the tests.
+
+##### Differential expression
+
+To test for differential expression, the analysis utilizes the "limma" package from R Bioconductor \citep{Phipson2016}.
+This package employs a linear model approach to determine the fold expression between sample groups.
+Before initiating the analysis, two matrices need to be computed.
+The design matrix identifies samples based on their sample type and defines the experimental design.
+An algorithm within the program's backend logic is utilized to automatically generate this matrix according to the user's selection in the interface.
+The expression matrix contains intensities for each identified protein and is obtained at the end of the pipeline.
+By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors.
+
+The subsequent step involves creating a linear model between groups using log-ratios of their expression values.
+In a two-way design, the expression of the gene $y$ can be explained as $Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \epsilon$, where $\beta_0$ represents the intercept, which can be interpreted as the mean expression of the gene.
+$\beta_1$ represents difference in mean of the treatment (or condition) on the discrete or continuos variable $X_{1}$, and $\epsilon$ acts as an error term.
+The interface allows for additive models with multiple factors, allowing the user to add factors according to various scenarios.
+For example, in a two-factor model, it would look like: $Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + \epsilon$.
+The principle remains the same for models with multiple factors.
+When visualizing the linear model, the x-axis would represent the gene for the sample types or other factors, whereas the y-axis would account for the expression value.
+By drawing a line through the cloud of data points the model is constructed by minimizing the quadratic distance to every data-point, called the least-square method or also ordinary least-square method.
+With the regression line the above mentioned equation can be constructed.
+
+Next the statistical module of the program performs an empirical Bayes model utilizing the eBayes \citep{Smyth2004} function of the Limma package.

### 2023-05-16 diff_exp_limma.R

@@ -302,13 +302,10 @@ library("edgeR")
 # top table of significant genes is the starting point of the analysis
 tt <- top_genes_2
 
-mask <- tt$adj.P.Val < 0.05 &
+mask <- tt$adj.P.Val < 1 &
   abs(tt$logFC) > 0.0005
 deGenes <- rownames(tt)[mask]
 head(deGenes)
-
-geneUniverse <- rownames(tt)
-length(geneUniverse)
 
 library(clusterProfiler)
 
@@ -335,4 +332,16 @@ p1
 p2 <- dotplot(ans.kegg, showCategory=20) + ggtitle("KEGG")
 p2
 
-cowplot::plot_grid(p1, p3, p5, ncol=2, labels=LETTERS[1:3])
+cowplot::plot_grid(p1, p2, ncol=2, labels=LETTERS[1:2])
+
+plot_ly(x=tab.kegg$Description, y=tab.kegg$Count, type = "bar")
+
+
+# plot_ly(x=dimred_umap$UMAP1,
+#         y=dimred_umap$UMAP2, 
+#         z=dimred_umap$UMAP3, 
+#         type="scatter3d", 
+#         mode="markers",
+#         color=meta_data_0$Batch) 
+
+

### 2023-05-17 diff_exp_limma.R
made dot plot for gene enrichment fancy with hover
to do: check x axis scaling, network analysis

@@ -323,7 +323,6 @@ ans.kegg <- enrichKEGG(
 
 tab.kegg <- as.data.frame(ans.kegg)
 tab.kegg<- subset(tab.kegg, Count>5)
-tab.kegg[1:5, 1:6]
 
 library(enrichplot)
 p1 <- graphics::barplot(ans.kegg, showCategory=10)
@@ -334,14 +333,7 @@ p2
 
 cowplot::plot_grid(p1, p2, ncol=2, labels=LETTERS[1:2])
 
-plot_ly(x=tab.kegg$Description, y=tab.kegg$Count, type = "bar")
-
-
-# plot_ly(x=dimred_umap$UMAP1,
-#         y=dimred_umap$UMAP2, 
-#         z=dimred_umap$UMAP3, 
-#         type="scatter3d", 
-#         mode="markers",
-#         color=meta_data_0$Batch) 
-
+library(plotly)
+p3 <- plot_ly(data=tab.kegg, x=~GeneRatio, y=~Description, type = "scatter", color= ~p.adjust, size=~Count, text=~geneID)
+p3
 

### 2023-05-17 scpR-app/app.R

@@ -143,14 +143,19 @@ ui <- fluidPage(
                            DT::dataTableOutput("protein_table")
                   ),
                   tabPanel("Gene set enrichment",
+                           br(),
                            fluidRow(width=12,
                                     column(width=4,
-                                           selectInput("p_value_correction_gsea", "select method for p-value correction", choices = c("none", "BH", "fdr", "BY", "holm"))),
+                                           textOutput("p_correct")),
                                     column(width = 4,
-                                           numericInput("p_value_cutoff_gsea", "choose p-value cutoff", value = 0.05, min = 0, step = 0.01)),
+                                           textOutput("p_cutoff")),
                                     column(width=4,
-                                           numericInput("fold_change_cutoff_gsea", "choose fold change cutoff", value = 0.05, min = 0, step = 0.01)),
-                           plotOutput("gsea"))
+                                           textOutput("fc_cutoff")),
+                                    br(),
+                                    br(),
+                                    br(),
+                                    br(),
+                           plotlyOutput("gsea"))
                   )
       )
     )
@@ -1738,18 +1743,18 @@ server <- function(input, output, session) {
     tt <- protein_table()
     
     # filter the data according to user selection
-    req(input$p_value_cutoff_gsea)
-    req(input$fold_change_cutoff_gsea)
-    p_cut <- input$p_value_cutoff_gsea
-    fc_cut <- input$fold_change_cutoff_gsea
+    req(input$p_value_cutoff)
+    req(input$fold_change_cutoff)
+    p_cut <- input$p_value_cutoff
+    fc_cut <- input$fold_change_cutoff
     
     mask <- tt$adj.P.Val < p_cut & 
       abs(tt$logFC) > fc_cut
     
     deGenes <- rownames(tt)[mask]
 
-    req(input$p_value_correction_gsea)
-    p_correct <- input$p_value_correction_gsea
+    req(input$p_value_correction)
+    p_correct <- input$p_value_correction
     
     ans.kegg <- enrichKEGG(
       deGenes,
@@ -1764,11 +1769,37 @@ server <- function(input, output, session) {
     )
   })
   
-  output$gsea<- renderPlot({
+  output$p_correct <- renderText({
+    req(input$p_value_correction)
+    paste(
[…]

### 2023-05-22 markdown/methods.Rmd
try catch for combat implemented
next --> implement for exp_set

@@
 ComBat standardizes the data by a Z-transformation before attempting to remove the batch effect.
-By the method of moments the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature.
+By the method of moments (= the mean and the variance) the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature driving statistical moments at a comparable level between batches.
+Furthermore the ComBat function allows the user to pass a desired effect (the actual biological variance) to be crystallized out from the non-biological variance.
+This is done by a constructing a design matrix which is also passed to the bayesian framework and can be interpreted as the apriori information about our dataset.
 

### 2023-05-22 diff_exp_limma.R

@@ -55,7 +55,7 @@ evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data
 #   labs(y=paste("intensity of", selectedProtein))
 
 # log transform expression values
-evidence_data <- log10(evidence_data) 
+evidence_data <- log2(evidence_data) 
 
 protein_matrix <- as.matrix(evidence_data)
 protein_matrix <- sweep(protein_matrix, 2, colMedians(protein_matrix), FUN="-")
@@ -198,11 +198,18 @@ fviz_pca_ind(data.pca, col.ind = "cos2",
 
 library(sva)
 
+evidence_data <- data.frame(evidence_data)
 # batch effects obtained --> use combat to correct
 batch <- meta_data_0$Batch
+group <- meta_data_0$Group
 
+model <- model.matrix(~0 + group, data = evidence_data)
 
-combat_edata1 = ComBat(dat=evidence_data, batch=batch)
+qr(model)$rank < ncol(model)
+n.batch <- nlevels(as.factor(batch))
+ncol(model)
+
+combat_edata1 = ComBat(dat=evidence_data, batch=batch, mod = model)
 
 # or use limma
 y2 <- removeBatchEffect(evidence_data, batch)
@@ -221,7 +228,7 @@ design <- model.matrix(~0+factor(meta_data_0$Group))
 colnames(design) <- c("EC", "HC", "VP")
 
 # Fit the expression matrix to a linear model
-fit <- lmFit(evidence_data_de, design)
+fit <- lmFit(evidence_data, design)
 
 # EC vs HC
 cont_matrix_1 <- makeContrasts(ECvsHC = EC-HC,levels=design)
@@ -300,14 +307,12 @@ library("edgeR")
 
 
 # top table of significant genes is the starting point of the analysis
-tt <- top_genes_2
+tt <- top_genes_1
 
 mask <- tt$adj.P.Val < 1 &
   abs(tt$logFC) > 0.0005
 deGenes <- rownames(tt)[mask]
 head(deGenes)
-
-library(clusterProfiler)
 
 ans.kegg <- enrichKEGG(
   deGenes,
@@ -322,7 +327,13 @@ ans.kegg <- enrichKEGG(
 )
 
 tab.kegg <- as.data.frame(ans.kegg)
-tab.kegg<- subset(tab.kegg, Count>5)
+#tab.kegg<- subset(tab.kegg, Count>5)
[…]

### 2023-05-22 scope2.R

@@ -513,11 +513,34 @@ if (length(peptide_file) > 1) {
   sce <- getWithColData(scp, "proteins_imptd")
   
   batch <- colData(sce)$Raw.file
-  model <- model.matrix(~0 + SampleType, data = colData(sce))
+  model <- model.matrix(~SampleType, data = colData(sce))
+  
+  (qr(model)$rank < ncol(model))
   
   assay(sce) <- ComBat(dat = assay(sce),
-                       batch = batch)#,
-                       #mod = model)
+                       batch = batch,
+                       mod = model)
+
+
+  Combat_batchC <- function(i_exp_matrix, i_batch, i_model) {
+    out <- tryCatch(
+      {
+        i_exp_matrix <- ComBat(dat = i_exp_matrix,
+                               batch = i_batch,
+                               mod = i_model)
+        print("batch corrected and optimized")
+      } ,
+      error = function(cond) {
+        i_exp_matrix <- ComBat(dat = i_exp_matrix,
+                               batch = i_batch)
+        print("confounder detected! just batch corrected")
+      }
+    )
+    return(out)
+  }
+  
+  assay(sce) <- Combat_batchC(assay(sce), batch, model) 
+  
   
   
   scp <- addAssay(scp,
@@ -554,8 +577,9 @@ scp[["proteins_dim_red"]] <- runPCA(scp[["proteins_dim_red"]],
 
 plotReducedDim(scp[["proteins_dim_red"]],
                dimred = "PCA",
-               colour_by = "SampleType",
-               point_alpha = 1)
+               colour_by = "Channel",
+               point_alpha = 1,
+               )
 
 # UMAP
 scp[["proteins_dim_red"]] <- runUMAP(scp[["proteins_dim_red"]],

### 2023-05-22 scpR-app/app.R

@@ -150,12 +150,11 @@ ui <- fluidPage(
                                     column(width = 4,
                                            textOutput("p_cutoff")),
                                     column(width=4,
-                                           textOutput("fc_cutoff")),
-                                    br(),
-                                    br(),
-                                    br(),
-                                    br(),
-                           plotlyOutput("gsea"))
+                                           textOutput("fc_cutoff"))
+                                    ),
+                           br(),
+                           materialSwitch("design_plot_gsea", label = "change Plot design"),
+                           plotlyOutput("gsea")
                   )
       )
     )
@@ -795,14 +794,33 @@ server <- function(input, output, session) {
           if (input$batch_c == "ComBat") {
             if (input$file_level == FALSE) {
               sce <- getWithColData(scp_0, "proteins_imptd")
+              
               batch <- colData(sce)$Raw.file
               # can be used to aim batch correction for desired result
-              model <- model.matrix(~0 + SampleType, data = colData(sce))
+              model <- model.matrix(~SampleType, data = colData(sce))
               
-              assay(sce) <- ComBat(dat = assay(sce),
-                                   batch = batch)#,
-              #mod = model)
               
+              Combat_batchC <- function(i_exp_matrix, i_batch, i_model) {
+                i_exp_matrix <- tryCatch(
+                  {
+                    i_exp_matrix <- ComBat(dat = i_exp_matrix,
+                                         batch = i_batch,
+                                         mod = i_model)
+                    errMsg("batch corrected and optimized")
+                    return(i_exp_matrix)
+                  },
+                  error = function(cond) {
+                    i_exp_matrix <- ComBat(dat = i_exp_matrix,
+                                           batch = i_batch)
+                    errMsg("confounder detected! just batch corrected")
+                    return(i_exp_matrix)
+                  }
+                )
+
+              }
+              
+              assay(sce) <- Combat_batchC(assay(sce), batch, model) 
+
               scp_0 <- addAssay(scp_0,
                                 y = sce,
                                 name = "proteins_batchC")
[…]

### 2023-05-23 diff_exp_limma.R
batch correction improved
user will be notified if confounding effect between sampletype and batch is detected.
leading the wetlab scientist to improved experimental design.

to do:
mention in thesis

@@ -16,12 +16,12 @@ evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
 
 # handle exclusion dependencies
-selectedSampleType_to_exclude <- c("Pool")
-      
-meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
+# selectedSampleType_to_exclude <- c("Pool")
+#       
+# meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
+# 
+# evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data))]
 
-
-evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data))]
 # plot basic counts
 #barplot(log10(dim(evidence_data)[1]), main = "log Count of rows")
 
@@ -125,34 +125,34 @@ evidence_data <- data.frame(protein_matrix)
 
 
 
-library('corrr')
-library("ggcorrplot")
-library("FactoMineR")
-library("factoextra")
-
-corr_matrix <- cor(evidence_data)
-ggcorrplot(corr_matrix)
-
-# in this case a correlation within the batches can be clearly observed 
-
-# apply pca on covariance matrix
-data.pca <- princomp(corr_matrix)
-summary(data.pca)
-# scree plot for explained variance by principle component
-# in this case the first compononents explain the majority of variance 
-fviz_eig(data.pca)
-
-# in the biplot the batch effect can be further evaluated.
-# samples within batches show the same vectors
-fviz_pca_var(data.pca, col.var = "black")
-
-
-
-# get results of pca
-fviz_pca_ind(data.pca, col.ind = "cos2", 
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
-             repel = TRUE # Avoid text overlapping (slow if many points)
-)
+# library('corrr')
+# library("ggcorrplot")
+# library("FactoMineR")
+# library("factoextra")
+# 
+# corr_matrix <- cor(evidence_data)
+# ggcorrplot(corr_matrix)
+# 
+# # in this case a correlation within the batches can be clearly observed 
+# 
+# # apply pca on covariance matrix
+# data.pca <- princomp(corr_matrix)
+# summary(data.pca)
+# # scree plot for explained variance by principle component
+# # in this case the first compononents explain the majority of variance 
[…]

### 2023-05-23 scpR-app/app.R

@@ -17,6 +17,7 @@ library("stats")
 library("impute")
 library("sva")
 library("plotly")
+library("clusterProfiler")
 
 reactlog::reactlog_enable()
 
@@ -72,8 +73,8 @@ ui <- fluidPage(
       #create tabs
       tabsetPanel(type = "tabs",
                   tabPanel("Overview",
-                           plotOutput("overview_plot"),
-                           verbatimTextOutput("debug")),
+                           verbatimTextOutput("debug"),
+                           plotOutput("overview_plot")),
                   tabPanel("Summary Barplot", plotOutput("summary_bar")),
                   tabPanel("Reporter Ion Intensity", 
                            selectInput("color_variable_ri", "select variable to indicate", ""),
@@ -211,6 +212,7 @@ boxcox_1.default <-
   }
 
 
+
 # Define server logic required to draw a histogram
 server <- function(input, output, session) {
   options(shiny.maxRequestSize=30*1024^2)
@@ -218,10 +220,34 @@ server <- function(input, output, session) {
   session$onSessionEnded(function() {
     stopApp()
   })
+
+  Combat_batchC <- function(i_exp_matrix, i_batch, i_model) {
+    i_exp_matrix <- tryCatch(
+      {
+        i_exp_matrix <- ComBat(dat = i_exp_matrix,
+                               batch = i_batch,
+                               mod = i_model)
+        return(i_exp_matrix)
+      },
+      error = function(cond) {
+        i_exp_matrix <- ComBat(dat = i_exp_matrix,
+                               batch = i_batch)
+        errMsg("confounder detected! Batch and eventually your desired effect corrected. reconsider your experimental design")
+        return(i_exp_matrix)
+      }
+    )
+  }
   
+  
+    
   #basic error handling
   errMsg <- reactiveVal()
-  output$debug <- renderPrint(errMsg())
+  output$debug <- renderPrint({
+    req(errMsg())
+    if (!is.null(errMsg())) {
+      errMsg()
+    }
+    })
   
   # Help Button and content of the help menu
   observeEvent(input$help,{
@@ -308,6 +334,7 @@ server <- function(input, output, session) {
   # result of analysis pipleline
   scp <- eventReactive(input$update_button, {
     tryCatch({
+      errMsg(NULL)
       withProgress(message= "running analysis", value=0, {
         req(input$evidence_file)
         req(meta_data)
@@ -794,30 +821,8 @@ server <- function(input, output, session) {
[…]

### 2023-05-24 markdown/methods.Rmd
normalization validation with mutual information included
to do:
include in thesis

@@
 #### Batch correction 
-Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data.
-This variance would overshadow the biological differences across sample types.
-Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter.
-To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used.
-The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch.
-In order to remove the systematic error the bayesian framework shrinks the effect by pooling information over the features $g$.
+Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data.
+This variance would overshadow the biological differences across sample types when performing dimensionality reduction.
+A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of the developed pipeline.
+It appears convenient to remove the batch effect before performing any statistical analysis, which is in turn prone to errors especially if the sample groups are not split equally between batches.
+In this case group differences influence the batch effect reducing statistical power after the correction., the both effects are so called interdependent.
+The opposite could happen if the experimental design is heavily unbalanced leading to group differences induced by batch correction.
+Finally it is recommended to include the batch factor should in the linear model as a co factor and statistically quantify it. \citep{Nygaard2016}
+However it could be of interest to observe biological relevance of the factor of interest in the dimensionality reduction visualization, so a batch correction option was included to meet this need.
+To address the non-biological experimental variations, also known as technical factors, the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. 
+The expression of a gene $g$ for the batch $m_i$ and sample $j$ can be defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, where $\gamma$ and $\delta\epsilon$ represent the systematic error caused by the batch.
+The Bayesian framework is employed to remove this systematic error by shrinking the effect and pooling information over the features $g$.
 Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups.
-ComBat standardizes the data by a Z-transformation before attempting to remove the batch effect.
+Before applying batch correction, the data is standardized using Z-transformation.
 By the method of moments (= the mean and the variance) the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature driving statistical moments at a comparable level between batches.
-Furthermore the ComBat function allows the user to pass a desired effect (the actual biological variance) to be crystallized out from the non-biological variance.
-This is done by a constructing a design matrix which is also passed to the bayesian framework and can be interpreted as the apriori information about our dataset.
+The ComBat function allows the user to specify a desired effect, which represents the actual biological variance to be preserved while removing non-biological variance.
+This is achieved by constructing a design matrix, which is provided as prior information about the dataset within the Bayesian framework.
+However, in cases of poor experimental design, the factors contributing to unwanted and desired variation may be confounded.
+In the worst case, if only one sample type per run is observed, it becomes challenging for the program to distinguish between batch-induced variance and biological variance.
+In such cases, the user is advised to adjust the experimental layout for future experiments.
+The detection algorithm for confounding effects is facilitated by QR matrix decomposition. The matrix $A=QR$ is decomposed into an orthogonal matrix $Q$ and an upper triangular matrix $R$. 
+By comparing the rank of matrix $A$ with the number of biologically relevant factors, it is possible to determine the presence of confounding effects in the dataset.
+The rank is defined as the maximum number of linearly independent columns.
+The same approach is used in the limma package \citep{Phipson2016} to identify non-solvable coefficients in experimental designs.
 
@@
 ##### Differential expression 
-
 To test for differential expression, the analysis utilizes the "limma" package from R Bioconductor \citep{Phipson2016}.
@@
 Next the statistical module of the program performs an empirical Bayes model utilizing the eBayes \citep{Smyth2004} function of the Limma package.
+The eBayes function utilizes a moderated t-statistics following t-distributed values.
+The advantage of this method over the posterior odds is the reduction of hyperparameters to estimate used for the modelling process, also called shrinkage.
+In this context, the hyperparameters to estimate are the coefficient $\beta_{gj}$ for gene $g$ and sample $i$, as well as the variance $\sigma^2_g$ across genes $g$.
+In the Bayesian approach, probabilities are updated after obtaining new data, referred to as conditional probabilities.
+For estimation, it is assumed that $\frac{1}{\sigma^2_g} \sim \frac{1}{d_{0}s^2_{0}} \chi^2_{d_{0}}$, representing the prior information of the model.
+The probability that $\beta_{gj} \neq 0$ is denoted as $p_j$, where $p_j$ represents the expected proportion of truly differentially expressed genes.
+The expected distribution of log-fold changes follows the distribution $\beta_{gj} | \sigma^2_g , \beta_{gj} \neq 0 \sim N(0, v_{0j}\sigma^2_g)$.
+The posterior mean of $\sigma^2_g$ given $s^2_g$ is calculated as $s^{-2}_g = E(\sigma^2_g | s^2_g) = \frac{d{0}s^2_{0} + d_{g}s^2_{g}}{d_{0} + d_{g}}$.
+This is where the shrinkage occurs, as the posteriors affect the prior values based on their respective sizes and degrees of freedom.
+The moderated t-statistic is defined as $\tilde{t}_{gj} = \frac{\hat{\beta}_{gj}}{\hat{s}_{g}\sqrt{v_{gj}}}$.
+Once the values of $\tilde{t}$ and $s^{2}$ are calculated, the posterior odds are computed, providing the odds that a particular gene is differentially expressed.
+The odds for gene $g$ being differentially expressed are denoted as $O_{gj} = \frac{p(\beta_{gj} \neq 0 | \tilde{t}_{gj}, s^2{g})}{p(\beta_{gj} = 0 | \tilde{t}_{gj}, s^2{g})}$, where the numerator represents the probability that the gene is differentially expressed, and the denominator represents the probability that the gene is not differentially expressed \citep{Smyth2004}.
+
+When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis.
+This is done by the multifactor option in the user interface by selecting the batch as a co factor and inclusion in the design matrix.

### 2023-05-24 diff_exp_limma.R

@@ -16,11 +16,11 @@ evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
 
 # handle exclusion dependencies
-# selectedSampleType_to_exclude <- c("Pool")
-#       
-# meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
-# 
-# evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data))]
+selectedSampleType_to_exclude <- c("Pool")
+
+meta_data_0 <- meta_data_0[!(meta_data_0$Group %in%  selectedSampleType_to_exclude), ]
+
+evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data))]
 
 # plot basic counts
 #barplot(log10(dim(evidence_data)[1]), main = "log Count of rows")
@@ -56,12 +56,83 @@ evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
 # log transform expression values
 evidence_data <- log2(evidence_data) 
+evidence_data_transf <- evidence_data
+
 
 protein_matrix <- as.matrix(evidence_data)
 protein_matrix <- sweep(protein_matrix, 2, colMedians(protein_matrix), FUN="-")
 # #normalize rowwise
 protein_matrix <- sweep(protein_matrix, 1, rowMeans(protein_matrix), FUN="-")
 evidence_data <- data.frame(protein_matrix)
+
+
+# validate the normalization by entropy estimation with k-nearest neighbor
+library("rmi")
+
+
+
+# split the data according to the column for selection of group
+group_select <- "HC"
+
+evidence_data_group_transf <- evidence_data_transf[ , which(meta_data_0$Group == group_select)]
+evidence_data_group_transf <- data.frame(rowMeans(evidence_data_group_transf))
+colnames(evidence_data_group_transf) <- paste(group_select, "_raw")
+
+
+evidence_data_group_norm <- evidence_data[ , which(meta_data_0$Group == group_select)]
+evidence_data_group_norm <- data.frame(rowMeans(evidence_data_group_norm))
+colnames(evidence_data_group_norm) <- paste(group_select, "_norm")
+
[…]

### 2023-05-24 scope2.R

@@ -515,13 +515,6 @@ if (length(peptide_file) > 1) {
   batch <- colData(sce)$Raw.file
   model <- model.matrix(~SampleType, data = colData(sce))
   
-  (qr(model)$rank < ncol(model))
-  
-  assay(sce) <- ComBat(dat = assay(sce),
-                       batch = batch,
-                       mod = model)
-
-
   Combat_batchC <- function(i_exp_matrix, i_batch, i_model) {
     out <- tryCatch(
       {
@@ -536,7 +529,7 @@ if (length(peptide_file) > 1) {
         print("confounder detected! just batch corrected")
       }
     )
-    return(out)
+    return(i_exp_matrix)
   }
   
   assay(sce) <- Combat_batchC(assay(sce), batch, model) 
@@ -845,6 +838,27 @@ plot_ly(x=dimred$PC1,
         mode="markers",
         color=scp$Raw.file)
 
-selection <- "Channel"
 
-scp_0[["proteins_dim_red"]][[selection]]
+
+## validate normalization
+
+# select data according to the procedure undertaken
+data_final <- assay(scp[["proteins_dim_red"]])
+data_raw <- assay(scp[["proteins"]])
+
+# select data according to groups to observe mutual information within
+data_final <- data_final[, which(scp$SampleType == "PreOp")]
+data_raw <- data_raw[, which(scp$SampleType == "PreOp")]
+            
+# discretize by equal frequencies
+data_final <- discretize(data_final)
+data_raw <- discretize(data_raw)
+
+library("infotheo")
+
+# mutual information is returned in nats
+mi_final <- mutinformation(data_final)
+mi_raw <- mutinformation(data_raw)
+
+gain <- mi_final-mi_raw
+boxplot(stack(gain)$value)

### 2023-05-24 scpR-app/app.R

@@ -18,6 +18,8 @@ library("impute")
 library("sva")
 library("plotly")
 library("clusterProfiler")
+library("infotheo")
+
 
 reactlog::reactlog_enable()
 
@@ -63,7 +65,7 @@ ui <- fluidPage(
       switchInput(inputId = "opts_multiple_batches", onLabel = "Advanced", offLabel = "Default", value = F, label="Options for multiple batches"),
       conditionalPanel(condition = "input.opts_multiple_batches", 
                        selectInput(inputId = "missing_v", "Choose method for missing value handling", choices=c("KNN", "drop rows", "replace with mean", "replace with median")),
-                       selectInput(inputId = "batch_c", "Choose method for batch correction", choices=c("ComBat", "None"))
+                       selectInput(inputId = "batch_c", "Choose method for batch correction", choices=c("None", "ComBat"))
                        )
       
     ),
@@ -76,6 +78,10 @@ ui <- fluidPage(
                            verbatimTextOutput("debug"),
                            plotOutput("overview_plot")),
                   tabPanel("Summary Barplot", plotOutput("summary_bar")),
+                  tabPanel("Normalization validation",
+                           selectInput("selectedSampleType_normval", "Select SampleType to validate", ""),
+                           plotOutput("norm_val_plot")
+                           ),
                   tabPanel("Reporter Ion Intensity", 
                            selectInput("color_variable_ri", "select variable to indicate", ""),
                            plotOutput("RI_intensity")),
@@ -489,6 +495,7 @@ server <- function(input, output, session) {
           rownames(evidence_data) <- evidence_data$ID
           evidence_data <- evidence_data[, !(names(evidence_data) %in% c("ID"))]
           evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data))]
+          evidence_data[evidence_data == 0] <- NA
         } 
         
 
@@ -497,6 +504,7 @@ server <- function(input, output, session) {
         transform_base_bc <- "None"
         
         if (input$transform_base == "log2") {
+          print("log2 transformation")
           if (input$file_level == FALSE) {
             scp_0 <- logTransform(scp_0,
                                   base = 2,
@@ -507,6 +515,7 @@ server <- function(input, output, session) {
           }
         }
         else if (input$transform_base == "log10") {
+          print("log10 transformation")
           if (input$file_level == FALSE) {
             scp_0 <- logTransform(scp_0,
                                   base = 10,
[…]

### 2023-05-24 markdown/methods.Rmd
explained information theory in concept of mutual information applied to the dataset validating normalizatin methods

@@
 
+#### Validation of the methods applied to the dataset
+In the context of the validation of methods applied to a dataset, the mutual information calculation is used to assess the impact of computations on the dataset and the information it carries.
+Specifically, the mutual information is calculated within each sample type to determine if the computations performed affect the dependence between samples of the same type.
+Mutual information, denoted as $I(X;Y)$, is a measure of the dependence or information shared between two variables, $X$ and $Y$, in this case, two samples of the same sample type.
+It is calculated using the equation $I(X;Y) = H(X,Y) - H(X|Y) - H(Y|X)$, where $H$ represents the entropy.
+Entropy, as a concept in information theory, quantifies the level of disorder or uncertainty associated with a random event.
+Events that are highly likely to occur are considered less informative than events that are more unexpected or less likely.
+In a venn diagramm $I(X;Y)$ (the mutual information) is the intersection between the two conditional Entropys $H(X|Y)$ and $H(Y|X)$.
+By comparing the mutual information of a particular sample type before performing any calculations with the mutual information calculated after the computations, it is possible to assess whether the applied methods have significantly altered the dependence or information content within the sample type.
+The goal is to ensure that the computations do not completely decrease the information contributed by each biological sample type to the experiment.
+
+The R package "infotheo" is utilized to perform the mutual information calculations, which is a commonly used package for information-theoretic analysis in R.
+Mutual information is returned as natural unit of information (=nat), a measurement proportional to the to the Shannon entropy ($1nat = \frac{1}{ln2} shannons$).
+Before calculating the pairwise mutual information of the expression matrix, values were discretized into bins with the discretize function of above package.
+After that all pairwise mutual information within the expression set before and after the pipeline is obtained.
+The pairwise differences can be observed as a boxplot for a particular sample type indicating the change in dependence within this sample type.
+
 #### Statistics

### 2023-05-25 scope2.R
normalization validation finished

@@ -841,6 +841,7 @@ plot_ly(x=dimred$PC1,
 
 
 ## validate normalization
+library("infotheo")
 
 # select data according to the procedure undertaken
 data_final <- assay(scp[["proteins_dim_red"]])
@@ -854,7 +855,6 @@ data_raw <- data_raw[, which(scp$SampleType == "PreOp")]
 data_final <- discretize(data_final)
 data_raw <- discretize(data_raw)
 
-library("infotheo")
 
 # mutual information is returned in nats
 mi_final <- mutinformation(data_final)
@@ -862,3 +862,88 @@ mi_raw <- mutinformation(data_raw)
 
 gain <- mi_final-mi_raw
 boxplot(stack(gain)$value)
+
+mi_final_stack <- data.frame(stack(mi_final))
+mi_raw_stack <- data.frame(stack(mi_raw))
+
+mi_total <- cbind(mi_final_stack$value, mi_raw_stack$value)
+colnames(mi_total) <- c("MI_final", "MI_raw")
+
+to_plot <- melt(mi_total)
+
+ggplot(to_plot, aes(x=Var2, y=value)) +
+  geom_boxplot(aes(fill=Var2)) +
+  ggtitle(paste("Change in mutual information after transformation:", 
+                "transform_base_val()", 
+                "and norm_method:", "norm_method_val()")) +
+  xlab(paste("input$selectedSampleType_normval")) +
+  ylab("Change in natural unit of information") + 
+  guides(fill=guide_legend(title=""))
+
+
+# inter group comparision of information content before and after the processes 
+
+data_final <- assay(scp[["proteins_dim_red"]])
+data_raw <- assay(scp[["proteins"]])
+
+user_choice <- "PreOP-S_24h"
+# split user choice of comp back to sample types
+user_choice_vector <- strsplit(user_choice, split = "-")
+# and assign them to a variable
+choice_A <- user_choice_vector[[1]][1]
+choice_B <- user_choice_vector[[1]][2]
+
+
+data_final_group1 <- discretize(data_final[, which(scp$SampleType == choice_A)])
+data_raw_group1 <- discretize(data_raw[, which(scp$SampleType == choice_A)])
+
+data_final_group2 <- discretize(data_final[, which(scp$SampleType == choice_B)])
+data_raw_group2 <- discretize(data_raw[, which(scp$SampleType == choice_B)])
+
+colnames(data_final_group1) <- paste( colnames(data_final_group1), "grp1")
[…]

### 2023-05-25 scpR-app/app.R

@@ -80,7 +80,9 @@ ui <- fluidPage(
                   tabPanel("Summary Barplot", plotOutput("summary_bar")),
                   tabPanel("Normalization validation",
                            selectInput("selectedSampleType_normval", "Select SampleType to validate", ""),
-                           plotOutput("norm_val_plot")
+                           plotOutput("norm_val_plot_intra_group"),
+                           selectInput("selectedComp_normval", "Choose comparison for observation", ""),
+                           plotOutput("norm_val_plot_inter_group")
                            ),
                   tabPanel("Reporter Ion Intensity", 
                            selectInput("color_variable_ri", "select variable to indicate", ""),
@@ -337,18 +339,39 @@ server <- function(input, output, session) {
     return(meta_data_0)
   })
   
+  
+  # read in expression set seperately for fetching raw data
+  exp_set_raw <- eventReactive(input$update_button, {
+    if (input$file_level == TRUE) {
+      req(input$evidence_file)
+      req(meta_data())
+      
+      meta_data_0 <- meta_data()
+      evidence_data <- read.delim(input$evidence_file$datapath)
+      
+      #handle ID column
+      rownames(evidence_data) <- evidence_data$ID
+      evidence_data <- evidence_data[, !(names(evidence_data) %in% c("ID"))]
+      
+      #handle columns to exclude
+      evidence_data <- evidence_data[ , which(meta_data_0$ID == colnames(evidence_data))]
+      evidence_data[evidence_data == 0] <- NA
+      return(evidence_data)
+    }
+  })
+  
   # result of analysis pipleline
   scp <- eventReactive(input$update_button, {
     tryCatch({
       errMsg(NULL)
       withProgress(message= "running analysis", value=0, {
-        req(input$evidence_file)
-        req(meta_data)
         
-        evidence_data <- read.delim(input$evidence_file$datapath)
+        req(meta_data)
         meta_data_0 <- meta_data()
         
         if (input$file_level == FALSE) {
+          req(input$evidence_file)
+          evidence_data <- read.delim(input$evidence_file$datapath)
           
           incProgress(1/17, detail = paste("read Data"))
           scp_0 <- readSCP(featureData = evidence_data,
@@ -492,10 +515,8 @@ server <- function(input, output, session) {
           
           
         } else {
[…]

### 2023-05-25 scpR-app/app.R
minor bugfix

@@ -756,6 +756,7 @@ server <- function(input, output, session) {
             assay(scp_0[["proteins_norm"]]) <- protein_matrix$normalized_data
           } else {
             evidence_data <- CONSTANd(evidence_data)
+            evidence_data <- evidence_data$normalized_data
           }
         } else if (input$norm_method == "None") {
           print("no normalization applied")

### 2023-05-25 markdown/methods.Rmd
validation ellaborated in the methods

--- 2023-05-24 markdown/methods.Rmd
@@
 
+#### Validation of the methods applied to the dataset
+In the context of the validation of methods applied to a dataset, the mutual information calculation is used to assess the impact of computations on the dataset and the information it carries. 
+Mutual information, denoted as $I(X;Y)$, is a measure of the dependence or information shared between two variables, $X$ and $Y$, in this case sample types.
+It is calculated using the equation $I(X;Y) = H(X,Y) - H(X|Y) - H(Y|X)$, where $H$ represents the entropy, $H(X|Y)$ and $H(Y|X)$ denote for the both conditional probabilities and $H(X,Y)$ for the joint entropy.
+Entropy, as a concept in information theory, quantifies the level of disorder or uncertainty associated with a random event.
+Events that are highly likely to occur are considered less informative than events that are more unexpected or less likely.
+In a Venn diagram $I(X;Y)$ (the mutual information) corresponds to the intersection of the two conditional entropys $H(X|Y)$ and $H(Y|X)$.
+To calculate the mutual information between two variables $X$ and $Y$ we can alternatively express the above equation as $I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})$.
+$D_{KL}$ denotes for the Kullback-Leibler divergence which is a concept of relative entropy between two variables and tells us how much 2 probability distributions differ from each other and $P_{X} \otimes P_{Y}$ represents the product of the marginals.
+The underlying principle of divergence is the integral asymmetry in Bayesian inference, which means that divergence does not satisfy the triangle inequality, and differences are not the same depending on the starting point.
+By comparing the mutual information before performing any calculations with the mutual information calculated after the computations, it is possible to assess whether the applied methods have significantly altered the dependence or information content.
+The goal is to choose a method which preserves the information contributed by each biological sample type to the experiment.
+The R package "infotheo" is utilized to perform the estimates for the entropy of the two variables.
+Mutual information is returned as natural unit of information (=nat), a measurement proportional to the to the Shannon entropy ($1nat = \frac{1}{ln2} shannons$).
+Before calculating the pairwise mutual information of the expression matrix, values were discretized into bins with the discretize function of above package.
+The mutual information is calculated within each sample type, and it can also be calculated between a selection of sample types to determine if the computations performed affect the information content.
+After that all pairwise mutual information within the expression set before and after the pipeline is obtained.
+The pairwise differences can be observed as a boxplot for a particular sample type indicating the change in dependence within and also compared to another sample type.
+
 #### Statistics
@@
 By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors.
+When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis.
+This is done by the multi factor option in the user interface by selecting the batch as a co factor and inclusion in the design matrix.
+The advantage of this procedure is, that we can quantify the biological relevance of working in multiple batches and avoid publishing false positive results.
+The multi factor option uses an additive model as a design matrix, whereas continuous variables (such as size, weight etc.) can also be used to build the statistical model.
 
@@
 In a two-way design, the expression of the gene $y$ can be explained as $Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \epsilon$, where $\beta_0$ represents the intercept, which can be interpreted as the mean expression of the gene.
-$\beta_1$ represents difference in mean of the treatment (or condition) on the discrete or continuos variable $X_{1}$, and $\epsilon$ acts as an error term.
+$\beta_1$ represents difference in mean of the treatment (or condition) on the discrete or continuous variable $X_{1}$, and $\epsilon$ acts as an error term.
 The interface allows for additive models with multiple factors, allowing the user to add factors according to various scenarios.
@@
 
-When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis.
-This is done by the multifactor option in the user interface by selecting the batch as a co factor and inclusion in the design matrix.
+Limma outputs a table with all proteins, their respective logfold change, p-values and adjusted p-values.
+The user can choose different methods for the correction of the p-values, which is necessary when testing the same hypothesis multiple times.
+A common method in omics studies is the false discovery rate (=fdr) by Benjamini Hochberg \citep{Benjamini1995}.
+By ranking all the p-values from smallest to largest the p-values are adjusted sequentially using the formula $p_(i) = \frac{r_i}{m} Q$.
+The rank in the p-value table is denoted as $r_i$   , $m$ reflects the total number of tests and $Q$ represents false discovery rate which is set to 5%.
+The second option to choose is the Benjamini & Yekutieli (=BY) \citep{Benjamini2001}method, which is a further development of the fdr by the same statistician and has a wider range of application based on the dependencies for the test.
+The third and most conservative option is the Holm method \citep{Holm1979} which controls the error sequentally in a family wise manner, which is considered as traditional in regards to the above mentioned experimental wise error rates.
+The above mentioned correction methods are ordered in the way from the most liberal to the most conservative one.
+The selection is supposed to give the user the possibility questioning results on multiple levels.

### 2023-05-30 diff_exp_limma.R
bugfix network plot

@@ -501,9 +501,14 @@ par(mar = c(1, 1, 1, 1))
 networkPlot(gsaRes,class="non")
 
 
+nw <- networkPlot2(gsaRes, class="non", significance = 0.5, shiny = T)
+
+library("visNetwork")
+
+visNetwork(nodes = nw$x$nodes, edges = nw$x$edges)
 
 
-networkPlot2(gsaRes, class="non", significance = 0.5, shiny = T)
+
 
 
 

### 2023-05-30 scpR-app/app.R

@@ -24,6 +24,11 @@ library("AnnotationDbi")
 library("piano")
 library("snowfall")
 library("parallel")
+library("visNetwork")
+
+require(shiny)
+require(visNetwork)
+
 
 
 reactlog::reactlog_enable()
@@ -36,7 +41,7 @@ ui <- fluidPage(
   
   # Sidebar
   sidebarLayout(
-    sidebarPanel(
+    sidebarPanel(id="settings_pane",
       # the button to rule them all
       actionButton("update_button", "Press to run/update"),
       actionButton("help", "Help"),
@@ -47,7 +52,7 @@ ui <- fluidPage(
       fileInput("sample_annotation_file", "Upload tabdel (.txt) sample annotation file", accept = c("text")),
       selectInput("selectedSampleType_to_exclude", "Select SampleType to exclude", "", multiple=T),
       
-      conditionalPanel(condition = "!input.file_level", 
+      conditionalPanel(id="full_control_pane", condition = "!input.file_level", 
                        # cutoff for PIF
                        numericInput("PIF_cutoff", "Input cutoff value for parental ion fraction. PSMs larger than the value remain", 0.1, min = 0, max=1, step = 0.1),
                        
@@ -68,7 +73,7 @@ ui <- fluidPage(
       selectInput("norm_method", "Choose method for protein data normalization", choices = c("SCoPE2", "None", "CONSTANd")),
       # multiple batch handling
       switchInput(inputId = "opts_multiple_batches", onLabel = "Advanced", offLabel = "Default", value = F, label="Options for multiple batches"),
-      conditionalPanel(condition = "input.opts_multiple_batches", 
+      conditionalPanel(id="multiple_batch_pane",condition = "input.opts_multiple_batches", 
                        selectInput(inputId = "missing_v", "Choose method for missing value handling", choices=c("KNN", "drop rows", "replace with mean", "replace with median")),
                        selectInput(inputId = "batch_c", "Choose method for batch correction", choices=c("None", "ComBat"))
       )
@@ -76,60 +81,61 @@ ui <- fluidPage(
     ),
     
     # Main Panel for the output
-    mainPanel(
+    mainPanel(id="output_pane",
       #create tabs
-      tabsetPanel(type = "tabs",
-                  tabPanel("Overview",
+      tabsetPanel(id="outbut_tabset", type = "tabs",
+                  tabPanel("Overview", id="overview_pane",
                            verbatimTextOutput("debug"),
                            plotOutput("overview_plot")),
[…]

### 2023-05-30 diff_exp_limma.R
gene ontology visualisation improved.

@@ -67,6 +67,11 @@ evidence_data <- data.frame(protein_matrix)
 
 
 
+
+
+
+
+
 # 
 # # compare mutual info between batches 
 # batch_select1 <- "Batch_1"

### 2023-05-30 scpR-app/app.R

@@ -26,9 +26,6 @@ library("snowfall")
 library("parallel")
 library("visNetwork")
 
-require(shiny)
-require(visNetwork)
-
 
 
 reactlog::reactlog_enable()
@@ -88,7 +85,7 @@ ui <- fluidPage(
                            verbatimTextOutput("debug"),
                            plotOutput("overview_plot")),
                   tabPanel("Summary Barplot", id="summary_pane",
-                           plotOutput("summary_bar")),
+                           plotOutput("summary_bar", width = "auto")),
                   tabPanel("Normalization validation", id="normval_pane",
                            selectInput("selectedSampleType_normval", "Select SampleType to validate", ""),
                            plotOutput("norm_val_plot_intra_group"),
@@ -102,7 +99,7 @@ ui <- fluidPage(
                            conditionalPanel(condition = "!input.file_level", id="razor_prot_pane",
                                             selectInput("color_variable_cv", "select variable to indicate", ""),
                                             plotOutput("CV_median")),
-                           plotOutput("corr_matrix"), width="100%"),
+                           plotOutput("corr_matrix")),
                   tabPanel("Dimensionality reduction", id="dimred_pane",
                            materialSwitch(inputId = "third_dim", value = F, label="3d", status = "danger"),
                            fluidRow(width=12,
@@ -174,7 +171,7 @@ ui <- fluidPage(
                            ),
                            br(),
                            materialSwitch("design_plot_gsea", label = "change Plot design"),
-                           plotlyOutput("gsea")
+                           plotlyOutput("gsea", height="auto", width = "auto")
                   ),
                   tabPanel("Protein set enrichment Ontology based", id="psa_ont_pane",
                            fluidRow(width=12,
@@ -202,10 +199,10 @@ ui <- fluidPage(
                            ),
                            br(),
                            conditionalPanel("!input.go_plot_switch", id="network_pane",
-                                            visNetworkOutput("go_network", width = "100%", height = "200%")
+                                            visNetworkOutput("go_network", width = "1200px", height = "1300px")
                                             ),
                            conditionalPanel("input.go_plot_switch", id="heatmap_pane",
-                                            plotOutput("go_heatmap", width = "100%", height = "200%")
+                                            plotOutput("go_heatmap", width = "1200px", height = "1300px")
                                             )
                   )
       )
[…]

### 2023-05-30 markdown/methods.Rmd
changed graph design

@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=17, fig.width=13}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=13}
 library(Gmisc, quietly = TRUE)
@@
 
-batch_correction <- boxGrob("batch correction",
+batch_correction <- boxGrob("batch correction & dimensionality reduction",
                             box_gp=plot_box_setting,
@@
 
-batch_correction_opts <- boxGrob("options:\nComBat",
+batch_correction_opts <- boxGrob("ComBat\nPCA\nUMAP",
                                  box_gp = opts_box_setting,
@@
 
-stat_module <- boxGrob(glue("statistics module",.sep = "\n"),
+stat_module <- boxGrob(glue("differential expression\npathway enrichment\nontology enrichment",.sep = "\n"),
                        box_gp = default_box_setting,
@@
                        missing_value_handling = missing_value_handling,
-                       batch_correction = batch_correction,
-                       dim_red = dim_red,
-                       stat_module
+#                       batch_correction = batch_correction,
+#                       dim_red = dim_red,
+                       out = stat_module
 )}
@@
 vert$files <- NULL
+
+output <- alignVertical(reference = vert$out,
+                      stat_module, batch_correction) %>%
+  spreadHorizontal()
+vert$out <- NULL
 
@@
 
-missing_value_opts <- moveBox(missing_value_opts,
-                    x = .85,
+# dim_red <- moveBox(dim_red,
+#                    x = coords(output[[2]])$x,
+#                    y = 0)
+
+batch_correction_opts <- moveBox(batch_correction_opts,
+                    x = .8,
                     y = coords(vert$missing_value_handling)$y)
 
-dim_red_actions <- moveBox(dim_red_actions,
-                    x = .9,
-                    y = coords(vert$dim_red)$y)
+# dim_red_actions <- moveBox(dim_red_actions,
+#                     x = (coords(dim_red)$x -.5),
+#                     y = coords(dim_red)$x)
 
@@
 
-batch_correction_opts <- moveBox(batch_correction_opts,
+missing_value_opts <- moveBox(missing_value_opts,
                     x = .2,
-                    y = coords(vert$batch_correction)$y)
+                    y = coords(vert$missing_value_handling)$y)
 
@@
 
-connectGrob(vert$batch_correction, batch_correction_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
+connectGrob(vert$missing_value_handling, output[[1]], type = "N")
+connectGrob(vert$missing_value_handling, output[[2]], type = "N")
 
-connectGrob(vert$dim_red, dim_red_actions, type = "horizontal", arrow_obj = arrow(angle=0))
+connectGrob(output[[2]], batch_correction_opts, type = "N", arrow_obj = arrow(angle=0), subelmnt = "r")
+
+# connectGrob(vert$dim_red, dim_red_actions, type = "horizontal", arrow_obj = arrow(angle=0))
 
@@
 batch_correction_opts
-dim_red_actions
+output
+# dim_red
+# dim_red_actions
 ```
@@
 #### Statistics
+##### Differential expression
 After visualizing the biological tendency in the dataset, the next step is hypothesis testing.
@@
 The selection is supposed to give the user the possibility questioning results on multiple levels.
+
+##### Pathway based protein enrichment analysis
+
+##### Ontology based enrichment

### 2023-05-30 scpR-app/app.R

@@ -199,10 +199,10 @@ ui <- fluidPage(
                            ),
                            br(),
                            conditionalPanel("!input.go_plot_switch", id="network_pane",
-                                            visNetworkOutput("go_network", width = "1200px", height = "1300px")
+                                            visNetworkOutput("go_network", width = "1200px", height = "1200px")
                                             ),
                            conditionalPanel("input.go_plot_switch", id="heatmap_pane",
-                                            plotOutput("go_heatmap", width = "1200px", height = "1300px")
+                                            plotOutput("go_heatmap", width = "1200px", height = "1200px")
                                             )
                   )
       )

### 2023-05-31 diff_exp_limma.R
bugfix of knn imputation

@@ -13,8 +13,6 @@ rownames(evidence_data) <- evidence_data$ID
 evidence_data <- evidence_data[ , !(names(evidence_data) %in% c("ID"))]
 
 
-
-
 # handle exclusion dependencies
 selectedSampleType_to_exclude <- c("Pool")
 
@@ -66,11 +64,22 @@ protein_matrix <- sweep(protein_matrix, 1, rowMeans(protein_matrix), FUN="-")
 evidence_data <- data.frame(protein_matrix)
 
 
+protein_matrix <- as.data.frame(normalize.quantiles(as.matrix(evidence_data)))
+rownames(protein_matrix) <- rownames(evidence_data)
+colnames(protein_matrix) <- colnames(evidence_data)
+evidence_data <- protein_matrix
+
+
+boxplot(evidence_data)
+boxplot(evidence_data_transf)
 
 
 
+summary(evidence_data)
+summary(evidence_data_transf)
 
-
+hist(evidence_data)
+hist(evidence_data_transf)
 
 # 
 # # compare mutual info between batches 

### 2023-05-31 scope2.R

@@ -413,72 +413,90 @@ assay(scp[["proteins_transf"]]) <- protein_matrix
  
  # CONSTANd normalization method
 
-# make MA plot first
-MAplot <- function(x,y,use.order=FALSE, R=NULL, cex=1.6, showavg=TRUE) {
-  # catch unequal size of matrices
-  if (dim(x)[2] != dim(y)[2]) {
-    if (dim(x)[2] > dim(y)[2]) {
-      x <- x[, 1:dim(y)[2]]
-    } else if (dim(y)[2] > dim(x)[2]) {
-      y <- y[, 1:dim(x)[2]]
-    }
-  }
-  
-  # make an MA plot of y vs. x that shows the rolling average,
-  M <- log2(y/x)
-  xlab = 'A'
-  if (!is.null(R)) {r <- R; xlab = "A (re-scaled)"} else r <- 1
-  A <- (log2(y/r)+log2(x/r))/2
-  if (use.order) {
-    orig.order <- order(A)
-    A <- orig.order
-    M <- M[orig.order]
-    xlab = "original rank of feature magnitude within IPS"
-  }
-  # select only finite values
-  use <- is.finite(M)
-  A <- A[use]
-  M <- M[use]
-  # plot
-  print(var(M))
-  plot(A, M, xlab=xlab, cex.lab=cex, cex.axis=cex)
-  # rolling average
-  if (showavg) { lines(lowess(M~A), col='red', lwd=5) }
-}
-
-MAplot(assay(scp[["proteins_transf"]])[,1:6], assay(scp[["proteins_transf"]])[,7:12])
-
-#find all pairwise indeces
-st_indeces <- split(seq_along(scp$SampleType), scp$SampleType)
-
-comp_list <- apply(combn(unique(scp$SampleType),2),2,paste, collapse="-")
-
-index_combis <- apply(combn(st_indeces,2),2,paste)
-
-user_choice <- comp_list[1]
-user_choice_vector <- strsplit(user_choice, split = "-")
-choice_A <- user_choice_vector[[1]][1]
-choice_B <- user_choice_vector[[1]][2]
-index_A <- st_indeces[choice_A]
-index_B <- st_indeces[choice_B]
-
-MAplot(assay(scp[["proteins_transf"]][,index_A[[1]]]), assay(scp[["proteins_transf"]][,index_B[[1]]]))
+# # make MA plot first
[…]

### 2023-05-31 scpR-app/app.R

@@ -67,7 +67,7 @@ ui <- fluidPage(
       # choose log transform base
       selectInput("transform_base", "Choose method for protein data transformation", choices = c("log2", "log10", "sqrt", "quadratic", "BoxCox", "None")),
       # normalization method 
-      selectInput("norm_method", "Choose method for protein data normalization", choices = c("SCoPE2", "None", "CONSTANd")),
+      selectInput("norm_method", "Choose method for protein data normalization", choices = c("col-Median,row_Mean", "None", "CONSTANd", "Quantile")),
       # multiple batch handling
       switchInput(inputId = "opts_multiple_batches", onLabel = "Advanced", offLabel = "Default", value = F, label="Options for multiple batches"),
       conditionalPanel(id="multiple_batch_pane",condition = "input.opts_multiple_batches", 
@@ -306,7 +306,7 @@ server <- function(input, output, session) {
                           title = "Proteomics Workbench",
                           HTML("Created by Lukas Gamp.<br> <br>
       Workflow as described in the publication: <br>
-      Multiplexed single-cell proteomics using SCoPE2 <br>
+      Multiplexed single-cell proteomics using col-Median,row_Mean <br>
              10.1038/s41596-021-00616-z <br>
              <br>
              <br>
@@ -405,8 +405,8 @@ server <- function(input, output, session) {
   
   # result of analysis pipleline
   scp <- eventReactive(input$update_button, {
-    tryCatch({
-      errMsg(NULL)
+    # tryCatch({
+    #   errMsg(NULL)
       withProgress(message= "running analysis", value=0, {
         
         req(meta_data)
@@ -723,8 +723,8 @@ server <- function(input, output, session) {
         
         incProgress(15/17, detail=paste("normalizing proteins"))
         req(input$norm_method)
-        if (input$norm_method == "SCoPE2" && (input$transform_base == "log2" | input$transform_base == "log10" | transform_base_bc == "log10")) {
-          print("SCoPE2 normalization on log transformed values")
+        if (input$norm_method == "col-Median,row_Mean" && (input$transform_base == "log2" | input$transform_base == "log10" | transform_base_bc == "log10")) {
+          print("col-Median,row_Mean normalization on log transformed values")
           if (input$file_level == FALSE) {
             # center cols with median
             scp_0 <- sweep(scp_0, i = "proteins_transf",
@@ -751,8 +751,8 @@ server <- function(input, output, session) {
             evidence_data <- data.frame(protein_matrix)
           }
           
[…]

### 2023-06-01 diff_exp_limma.R
background proteins for enrichment implemented user can now choose between custom set or detected proteins for background

@@ -64,24 +64,7 @@ protein_matrix <- sweep(protein_matrix, 1, rowMeans(protein_matrix), FUN="-")
 evidence_data <- data.frame(protein_matrix)
 
 
-protein_matrix <- as.data.frame(normalize.quantiles(as.matrix(evidence_data)))
-rownames(protein_matrix) <- rownames(evidence_data)
-colnames(protein_matrix) <- colnames(evidence_data)
-evidence_data <- protein_matrix
 
-
-boxplot(evidence_data)
-boxplot(evidence_data_transf)
-
-
-
-summary(evidence_data)
-summary(evidence_data_transf)
-
-hist(evidence_data)
-hist(evidence_data_transf)
-
-# 
 # # compare mutual info between batches 
 # batch_select1 <- "Batch_1"
 # batch_select2 <- "Batch_2"
@@ -372,6 +355,19 @@ library("edgeR")
 # top table of significant genes is the starting point of the analysis
 tt <- top_genes_1
 
+background_genes <- data.frame(rownames(tt))
+
+library(tidyverse)
+write_delim(
+  background_genes,
+  "/home/lukas/Downloads/background_genes.txt",
+  delim = " ",
+)
+
+background <- read.delim("/home/lukas/Downloads/background_genes.txt") 
+background <- as.vector(background[1])
+background <- background[[1]]
+
 mask <- tt$adj.P.Val < 1 &
   abs(tt$logFC) > 0.0005
 deGenes <- rownames(tt)[mask]
@@ -386,7 +382,8 @@ ans.kegg <- enrichKEGG(
   minGSSize = 10,
   maxGSSize = 500,
   qvalueCutoff = 0.2,
-  use_internal_data = FALSE
+  use_internal_data = FALSE,
+  universe = background
 )
 
 tab.kegg <- as.data.frame(ans.kegg)

### 2023-06-01 scpR-app/app.R

@@ -160,6 +160,7 @@ ui <- fluidPage(
                            DT::dataTableOutput("protein_table")
                   ),
                   tabPanel("Protein set enrichment Pathway based", id="psa_pathw_pane",
+                           actionButton("run_kegg", "Press to run pathway enrichment"),
                            br(),
                            fluidRow(width=12,
                                     column(width=4,
@@ -170,7 +171,15 @@ ui <- fluidPage(
                                            textOutput("fc_cutoff_1"))
                            ),
                            br(),
-                           materialSwitch("design_plot_gsea", label = "change Plot design"),
+                           fluidRow(width=12,
+                                    column(width=4,
+                                           materialSwitch("design_plot_gsea", label = "change Plot design")),
+                                    column(width = 4,
+                                           materialSwitch("panel_as_universe", label = "use detected proteins as background")),
+                                    column(width=4,
+                                           fileInput("custom_universe", "Upload tabdel (.txt) uniprot_id background (w.col header)", accept = c("text")))
+                           ),
+                           
                            plotlyOutput("gsea", height="auto", width = "auto")
                   ),
                   tabPanel("Protein set enrichment Ontology based", id="psa_ont_pane",
@@ -2098,7 +2107,7 @@ server <- function(input, output, session) {
   
   # gene set enrichment analysis 
   
-  result_kegg <- reactive({
+  result_kegg <- eventReactive(input$run_kegg, {
     withProgress(message = "running pathway analysis", value = 0, {
       incProgress(1/3, detail = paste("reading data"))
       req(protein_table())
@@ -2115,6 +2124,20 @@ server <- function(input, output, session) {
       
       deGenes <- rownames(tt)[mask]
       
+      if (input$panel_as_universe) {
+        background <- rownames(tt)
+      } else {
+        background <- NULL
+      }
+      
+      if (!is.null(input$custom_universe) & !(input$panel_as_universe)) {
+        background <- read.delim(input$custom_universe$datapath)
+        background <- as.vector(background[1])
+        background <- background[[1]]
+      } else {
+        background <- background
+      }
+      
       req(input$p_value_correction)
       p_correct <- input$p_value_correction
       
@@ -2128,7 +2151,8 @@ server <- function(input, output, session) {
         minGSSize = 10,
         maxGSSize = Inf,
         qvalueCutoff = 0.2,
-        use_internal_data = FALSE
+        use_internal_data = FALSE,
+        universe = background
       )
       incProgress(3/3, detail = paste("success"))
     })

### 2023-06-01 markdown/methods.Rmd
bugfix of knn imputation
enrichment analysis ellaborated in thesis

@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=13}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=16}
 library(Gmisc, quietly = TRUE)
@@
 
-normalize_opts <- boxGrob("options:\nSCoPE2\nCONSTANd\nNone",
+normalize_opts <- boxGrob("options:\ncol-Median,row_Mean\nCONSTANd\nNone",
                           box_gp = opts_box_setting,
@@
 
+#### Protein set enrichment analysis
+After differential expression the upcoming question could be, in which context the up or down regulated proteins are.
+Enriched proteins can be understood as an overabundance within a specific set.
+This set can be a pathway or a selection of common denominators in biological contexts, such as cell types or oncological factors.
+The statistical testing to identify the overabundant proteins of a specific set is similar to a $\chi^2$ test developed by Karl Pearson \citep{Pearson1900} or the exact Fisher test \citep{Sprent2011} for tables with 4 fields in small sample sizes.
+Within the list are all truly differentially expressed genes, an annotation is the identification of a protein in the desired biological context.
+We expect that the values populate the fields in the table as follows:
+
+\begin{center}
+\begin{tabular}{|c|| c| c|| c|} 
+ \hline
+   & in list & not in list & totals \\ [0.5ex] 
+ \hline\hline
+ with annotation & (A+B)(A+C)/N  & (A+B)(B+D)/N & A+B \\ 
+ \hline
+ without annotation & (C+D)(A+C)/N & (C+D)(B+D)/N & C+D \\
+ \hline
+  & A+C & B+D & N \\
+ \hline
+\end{tabular}
+\end{center}
+
+Under the null hypothesis $H_{0}$ states that the distribution and population of the table is random and there is no clear tendency with significant evidence.
+Against that the alternative hypothesis $H_{A}$ says that there is an underlying tendency the table is populated.
+This leads to our test statistics, the observed values (=$O$) in the table are significantly different to the expected (=$E=\frac{(A+B \lor C+D)}{N}$ ones.
+One approach is approximating $\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}$ and deriving the p-value by the area under the density curve of the $\chi^2$ distribution $Q = \sum_{i=1}^{k} Z_{i}^{2}$.
+In the distribution $k$ is defined by the degrees of freedom (=df), what can be obtained by $k=(columns-1)*(rows-1)=df$ of our table.
+Another approach to obtain an exact p-value would be Fisher`s exact test, which states that the margins are distributed according to the hypergeometric distributions.
+The test statistic leads directly to the p-value with no need for approximation:
+$p=1-\sum_{i=0}^{A-1}\frac{\binom{B}{i}\binom{D-B}{A+C-i}}{\binom{D}{A+C}}$
+Here D denotes either for the background distribution which can be the total number of proteins obtained in our experiment, all proteins with annotation or a custom selected background.
+These three options are implemented in the user interface.
+After obtaining the p-values for each protein set, the subsequent critical step involves the correction for multiple testing, considering the inherent risk of false positive results.
+Multiple testing correction methods adjust the significance thresholds to account for the increased probability of detecting false positives when performing multiple statistical tests.
+In this protein set enrichment analysis, the user is provided with a selection of established correction methods, akin to those commonly utilized in the field of differential expression analysis, ensuring robust and reliable interpretation of the results.
+
 ##### Pathway based protein enrichment analysis
+The pathway-based protein enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}.
+The analysis utilizes the “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways.
+Subsequently, the Fisher's exact test is applied to identify significantly enriched pathways.
+To focus on truly differentially expressed proteins, users have the flexibility to specify p-value and fold change cutoffs.
+Furthermore, the method for p-value correction can be selected based on user preference.
+Once the enriched protein pathways are determined, they can be visually represented using bar or dotplots, providing a clear and intuitive representation of the data.
 
-##### Ontology based enrichment
+##### Ontology based protein enrichment analysis

### 2023-06-07 diff_exp_limma.R
ontology tab refined

@@ -398,10 +398,7 @@ visNetwork::visIgraph(graph_info, physics = T, smooth = T)
 
 
 tab.kegg <- as.data.frame(ans.kegg)
-#tab.kegg<- subset(tab.kegg, Count>5)
-# 
-# tab.kegg$GeneRatio[1] <- "10/53"
-# tab.kegg$Count[1] <- 10
+
 
 tab.kegg$denominator <- as.numeric(gsub("^\\d+/(\\d+)$", "\\1", tab.kegg$GeneRatio))
 tab.kegg$decimal_gene_ratio <- tab.kegg$Count / tab.kegg$denominator
@@ -536,8 +533,58 @@ library("visNetwork")
 visNetwork(nodes = nw$x$nodes, edges = nw$x$edges)
 
 
-GSAheatmap(gsaRes_2)
+GSAheatmap(gsaRes)
 
+
+
+
+
+
+
+
+gsa_results <- GSAsummaryTable(gsaRes = gsaRes)
+gsa_results$name_wo_suff <- gsub("_UP*", "", gsa_results$Name)
+gsa_results$name_wo_suff <- gsub("_DN*", "", gsa_results$name_wo_suff)
+
+gsa_results_total_count <- data.frame(aggregate(gsa_results$`Genes (tot)`, list(gsa_results$name_wo_suff), sum))
+colnames(gsa_results_total_count) <- c("Description", "count")
+
+gsa_results_total_pvals <- data.frame(aggregate(gsa_results$`p adj (non-dir.)`, list(gsa_results$name_wo_suff), mean))
+gsa_results_total_pvals <- data.frame(aggregate(gsa_results$`p (non-dir.)`, list(gsa_results$name_wo_suff), mean))
+
+colnames(gsa_results_total_pvals) <- c("Description", "p.adjust")
+
+gsa_results_total <- cbind(gsa_results_total_count, gsa_results_total_pvals)
+
+gsa_results_total$ratio <- paste(as.character(gsa_results_total$count), "/" ,as.character(length(gsa_results_total$Description)))
+
+gsa_results_total$decimal_ratio <- gsa_results_total$count / length(gsa_results_total$Description)
+
+gsa_results_total <- gsa_results_total[, !duplicated(colnames(gsa_results_total))]
+
+
+mask <- gsa_results_total$p.adjust < 0.5
+gsa_results_total <- gsa_results_total[1:100, ]
+
+
+library(plotly)
+
[…]

### 2023-06-07 scpR-app/app.R

@@ -208,12 +208,12 @@ ui <- fluidPage(
                                     column(width = 3,
                                            br(),
                                            br(),
-                                           selectInput("gene_set_stat", label="select stat method", choices = c("fisher", "stouffer","reporter", "tailStrength", "wilcoxon", "mean", "median", "sum", "maxmean","gsea", "fgsea", "page"))),
+                                           selectInput("gene_set_stat", label="select stat method", choices = c("fisher", "stouffer","tailStrength"))),
                                     column(width = 3,
                                            br(),
                                            br(),
                                            br(),
-                                           switchInput("go_plot_switch", label="switch between visualizations", onLabel="Heatmap", offLabel="Network", onStatus="primary", offStatus = "info")),
+                                           switchInput("go_plot_switch", label="switch between visualizations", onLabel="Dotplot", offLabel="Network", onStatus="primary", offStatus = "info")),
                                     column(width=3,
                                            br(),
                                            br(),
@@ -238,8 +238,8 @@ ui <- fluidPage(
                                                              plotOutput("go_network_2", width = "1300px", height = "800px")
                                                              )
                                             ),
-                           conditionalPanel("input.go_plot_switch", id="heatmap_pane",
-                                            plotOutput("go_heatmap", width = "1800px", height = "800px")
+                           conditionalPanel("input.go_plot_switch", id="dotplot_pane",
+                                            plotlyOutput("go_dotplot", width = "1300px")
                                             ), 
                   )
       )
@@ -1950,7 +1950,6 @@ server <- function(input, output, session) {
           
           design <- model.matrix(~0+ . , data=design_frame)
           colnames(design)[1:length(unique(design_frame$Group))] <- sub("Group", "", colnames(design)[1:length(unique(design_frame$Group))])
-          print(design)
           
           fit <- lmFit(exp_matrix_0, design)
  
@@ -2375,27 +2374,71 @@ server <- function(input, output, session) {
   
   
   
-  output$go_heatmap <- renderPlot({
-    withProgress(message = "Plotting heatmap", value=0, 
+  output$go_dotplot <- renderPlotly({
+    withProgress(message = "Plotting dotplot", value=0, 
                  {
                    incProgress(1/3, detail = paste("reading data"))
+                   
[…]

### 2023-06-07 scpR-app/app.R
changed defaults

@@ -22,8 +22,8 @@ library("infotheo")
 library("org.Hs.eg.db")
 library("AnnotationDbi")
 library("piano")
-#library("snowfall")
-#library("parallel")
+library("snowfall")
+library("parallel")
 library("visNetwork")
 
 
@@ -201,20 +201,16 @@ ui <- fluidPage(
                            actionButton("go_ontology", label = "Run ontology", class = "btn-success"),
                            br(),
                            fluidRow(width=12,
-                                    column(width=3,
+                                    column(width=4,
                                            br(),
                                            br(),
                                            fileInput("geneset_collection", "visit gsea-msigdb.org for your collection", accept = c("text"))),
-                                    column(width = 3,
-                                           br(),
-                                           br(),
-                                           selectInput("gene_set_stat", label="select stat method", choices = c("fisher", "stouffer","tailStrength"))),
-                                    column(width = 3,
+                                    column(width = 4,
                                            br(),
                                            br(),
                                            br(),
                                            switchInput("go_plot_switch", label="switch between visualizations", onLabel="Dotplot", offLabel="Network", onStatus="primary", offStatus = "info")),
-                                    column(width=3,
+                                    column(width=4,
                                            br(),
                                            br(),
                                            br(),
@@ -2316,18 +2312,15 @@ server <- function(input, output, session) {
       req(input$p_value_correction)
       p_correct <- input$p_value_correction
       
-#      cores <- detectCores()
-      
-      req(input$gene_set_stat)
-      stat_meth <- input$gene_set_stat
-      
-      gsaRes <- runGSA(geneLevelStats = pVals,
-                       directions = logFCs,
-                       gsc=myGSC,
-                       adjMethod = p_correct,
-                       geneSetStat = as.character(stat_meth)#,
-#                       ncpus = cores
-                       )
+     cores <- detectCores()
+
+     gsaRes <- runGSA(geneLevelStats = pVals,
+                      directions = logFCs,
+                      gsc=myGSC,
+                      adjMethod = p_correct,
+                      geneSetStat = "fisher",
+                      ncpus = cores
+                      )
       incProgress(4/4, detail=paste("success"))
       
     })

### 2023-06-07 markdown/methods.Rmd
methods finished

@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=16}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=18}
 library(Gmisc, quietly = TRUE)
@@
 After differential expression the upcoming question could be, in which context the up or down regulated proteins are.
-Enriched proteins can be understood as an overabundance within a specific set.
+Enriched proteins can be understood as an over-representation (ORA = over-representation analysis) within a specific set.
 This set can be a pathway or a selection of common denominators in biological contexts, such as cell types or oncological factors.
@@
 Here D denotes either for the background distribution which can be the total number of proteins obtained in our experiment, all proteins with annotation or a custom selected background.
-These three options are implemented in the user interface.
+The background indicates all proteins which could be positive tested for in our observed sample.
+The above mentioned three options are implemented in the user interface.
 After obtaining the p-values for each protein set, the subsequent critical step involves the correction for multiple testing, considering the inherent risk of false positive results.
@@
 In this protein set enrichment analysis, the user is provided with a selection of established correction methods, akin to those commonly utilized in the field of differential expression analysis, ensuring robust and reliable interpretation of the results.
+Once the enriched protein pathways are determined, they can be visually represented using bar or dotplots, and a network of the found pathways.
+In the network ontologies are depicted as nodes, whereas edges show the strength of the interaction by thickness corresponding to the number of associated proteins.
 
@@
 The analysis utilizes the “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways.
-Subsequently, the Fisher's exact test is applied to identify significantly enriched pathways.
-To focus on truly differentially expressed proteins, users have the flexibility to specify p-value and fold change cutoffs.
-Furthermore, the method for p-value correction can be selected based on user preference.
-Once the enriched protein pathways are determined, they can be visually represented using bar or dotplots, providing a clear and intuitive representation of the data.
 
-##### Ontology based protein enrichment analysis
+##### Custom ontology based protein enrichment analysis
+For custom ontologies the protein enrichment analysis is done using the R bioconductor package piano \citep{Vaeremo2013}.
+In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
+For statistical testing the above explained Fisher exact test is used.

### 2023-06-07 scpR-app/app.R

@@ -184,10 +184,10 @@ ui <- fluidPage(
                                            textOutput("fc_cutoff_1"))
                            ),
                            conditionalPanel("!input.pw_plot_switch", id="network_pane",
-                                            conditionalPanel("!input.design_plot_gsea",
+                                            conditionalPanel("input.design_plot_gsea",
                                                              visNetworkOutput("gsea_network_1", width = "1300px", height="800px")
                                                              ),
-                                            conditionalPanel("input.design_plot_gsea",
+                                            conditionalPanel("!input.design_plot_gsea",
                                                              plotOutput("gsea_network_2", width = "1300px", height="800px")
                                                              )
                            ),
@@ -227,15 +227,15 @@ ui <- fluidPage(
                            ),
                            br(),
                            conditionalPanel("!input.go_plot_switch", id="network_pane",
-                                            conditionalPanel("!input.design_plot_go",
+                                            conditionalPanel("input.design_plot_go",
                                                              visNetworkOutput("go_network_1", width = "1300px", height = "800px")
                                                              ),
-                                            conditionalPanel("input.design_plot_go",
+                                            conditionalPanel("!input.design_plot_go",
                                                              plotOutput("go_network_2", width = "1300px", height = "800px")
                                                              )
                                             ),
                            conditionalPanel("input.go_plot_switch", id="dotplot_pane",
-                                            plotlyOutput("go_dotplot", width = "1300px")
+                                            plotlyOutput("go_dotplot", width = "1300px", height = "800px")
                                             ), 
                   )
       )
@@ -2248,7 +2248,7 @@ server <- function(input, output, session) {
   
   
   output$gsea_network_1 <- renderVisNetwork({
-    if (!input$design_plot_gsea & !input$pw_plot_switch) {
+    if (input$design_plot_gsea & !input$pw_plot_switch) {
       withProgress(message = "Plotting network", value=0,{
         incProgress(1/3, detail=paste("reading data"))
         req(gsea_network_data())
@@ -2330,7 +2330,7 @@ server <- function(input, output, session) {
   
   
   output$go_network_1 <- renderVisNetwork({
-    if (!input$design_plot_go & !input$go_plot_switch) {
+    if (input$design_plot_go & !input$go_plot_switch) {
       withProgress(message = "Plotting network", value=0,{
         incProgress(1/3, detail=paste("reading data"))
         req(result_piano())

### 2023-06-07 markdown/introduction.Rmd
introduction further ellaborated

@@
 
+## From the Central Dogma of Molecular Biology to Omics and Systems Biology: Exploring Interactions and Quantifying Biological Relevance
+Before understanding the underlying principle biological systems undergo and the initial exploration of molecular biology, the processes of live were roughly inferred and not well understood.
+The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins.
+This framework consists of two major processes, which involve transcription, converting DNA into RNA, and translation, converting RNA into functional proteins.
+In between those processes regulation takes base on a molecular level, by altering structure with chemical adaptations.
+The regulation is undertaken to provide biological organisms with an highly efficient output of their  genetic products called proteins.
+Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb, 2017}.
+Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as omics, to comprehensively analyze biomolecules in diverse contexts.
+Omics research has emerged as an approach for comprehensively studying biomolecules and their interactions within biological systems.
+Omics encompasses various fields, including genomics, transcriptomics, proteomics, metabolomics, and others, each focusing on a specific class of biomolecules.
+By analyzing multiple omics layers, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks with computational methods.
+Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks.
+Imagining biological systems as complex, multi-layer networks, the ensemble of proteins, known as the proteome, plays a critical role in this interactive structure with many functions such as structural integrity, catalyzing chemical reaction and regulation of cellular funcitons.
+
 ## Proteomics
+The central dogma of molecular biology offers insights not only at the level of individual terms but also provides a hierarchical understanding of biology.
+Hierarchical organization is the underlying principle in all topics of biology.
+Proteins exhibit hierarchical structures that contribute to their functionality.
+At the primary structure level, proteins are composed of smaller building blocks called amino acids.
+Combining them results in the protein secondary structure such as alpha helix and beta sheet.
+These secondary structures further assemble to create higher-order tertiary structures, which represent specific domains within the protein.
+Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
+By studying the proteome, which encompasses all the proteins present in a biological system at a specific moment, we can gain a comprehensive understanding of the hierarchical organization and functional dynamics of biological processes.
 
-The proteome is referred to the sum of all proteins of a given sample at a given time.
-In the past several quantitative and qualitative assays were used to enlighten the protein composition of a sample.
+### Bulk proteomics
+Higher organisms consist of highly differentiated cells in an arrangement called tissue.
+The term tissue includes skin, muscle or blood and is represented by an ensemble of living cells.
+Taking a sample of any particular tissue will include all types of cells.
+
+In tissues cells have specific functions and not all proteins are equally expressed in all cells.
+Taking samples can lead to an effect called averaging over the entire cellular ensemble.
+Cell sorting techniques like fluorescence activated cell sorting, magnetic bead isolation and liquid chromatography can compensate for this effect and help for studying single cell lines or types.
+
+### Single cell type
+
+### Single cell proteomics
+
+Single-cell proteomics (=SCP) provides a snapshot of the underlying processes in a cell.
+These proteins can be influenced by various environmental factors and the understanding of abundance or differential expression is crucial to understand the principles of cellular behavior.
+SCP gives insight in a qualitative but also a quantitative way and can elucidate responses of drugs, infectious diseases, and the development of an organism.
+In the last 10 years SCP became an indispensable tool for global research communities (Minakshi et al., 2019).
 
@@
 
-## Mass Spectrometry
+## Effectors of the proteome
+
+### External effectors
+
+When pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
+These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications and more (Jean Beltran et al., 2017).
+
+## Methods for Quantifications
+
+A holistic approach to analyze the expression can be sequencing of RNA (RNAseq) or mass-spectrometry (MS).
+RNAseq refers to the task of examining the transcriptome of bulk tissue sample  with deep-sequencing technologies.
+RNA in a sample includes mRNAs, non-coding RNAs, and small RNAs.
+One of the main disadvantages is, that the mapping relies on the knowledge of the DNA sequence of the observed organism (reference genome) (Wang et al., 2009).
+Another disadvantage is, that the amount of protein translated from the determined mRNA can only be estimated (Gygi et al., 1999).
+However, today these predictions can be done with computational methods such as LASSO and the accuracy of the estimate highly depends on selection and validation of this method (Magnusson et al., 2022).
+
+When observing the protein ensemble of a cellular environment with a top-down approach  like DNAseq or RNAseq, the characteristics (e.g.
+post-translational modifications or charge) and local concentration of the proteins remain hidden.
+The cause of this phenomenon is that only a fraction of the DNA is transcribed into mRNA and not all mRNA is further translated into protein.
+Furthermore multiple proteins can be derived from a single DNA-sequence with alternative splicing and different proteins coming from non related sequences acting like building blocks can form protein complexes and/or Protein-RNA complexes.
+Cells are highly efficient and do not waste energy by producing obsolete proteins, moreover the degradation of proteins is an important energy recovery mechanism.
+A bottom-up strategy can elucidate these properties of the protein ensemble at a given time leading to proteomics.
+Early methods in molecular biology were western blots with the disadvantage of having a-priori knowledge of the analyzed protein and quantitative protein assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay measuring the entire protein concentration.
+When studying impact of an external signal or stressor to a cell the abundance and   properties of a protein or protein-ensemble can enlighten the underlying pathway leading to a biological response.
+Since some proteins do not even function before post-translational modifications happen, a knowledge about these can be crucial for further studies.
+
+### Mass Spectrometry
 
@@
 
-Mass to charge ratio (m/z)
+Mass spectrometers measure the mass to charge ratio (m/z) of a particle from a fragmented larger molecule.
+This is achieved by a physical procedure done with a device which is made of three major components: the ion source, an analyzer, and a detector.
+The ion sources charges molecules and accelerate them through a magnetic field.
+The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal (Parker et al., 2010).
+MS provides great sensitivity for traces, hence getting an average concentration for all biomolecules in a sample.
+To identify protein abundance from a single type of cell, cell sorting techniques needed to be improved and applied to MS (Minakshi et al., 2019).
+Single cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
+Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics (Ye et al., 2022).
+
+MS provides great sensitivity for traces, hence getting an average concentration for all biomolecules in a sample.
+To identify protein abundance from a single type of cell, cell sorting techniques needed to be improved and applied to MS (Minakshi et al., 2019).
+Single cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
+Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics (Ye et al., 2022).

### 2023-06-08 markdown/introduction.Rmd
introduction bulk proteomics finished, single cell type started

@@
 
-## From the Central Dogma of Molecular Biology to Omics and Systems Biology: Exploring Interactions and Quantifying Biological Relevance
+## From the Central Dogma of Molecular Biology to Omics and Systems Biology
 Before understanding the underlying principle biological systems undergo and the initial exploration of molecular biology, the processes of live were roughly inferred and not well understood.
@@
 ### Bulk proteomics
-Higher organisms consist of highly differentiated cells in an arrangement called tissue.
-The term tissue includes skin, muscle or blood and is represented by an ensemble of living cells.
-Taking a sample of any particular tissue will include all types of cells.
-
-In tissues cells have specific functions and not all proteins are equally expressed in all cells.
-Taking samples can lead to an effect called averaging over the entire cellular ensemble.
-Cell sorting techniques like fluorescence activated cell sorting, magnetic bead isolation and liquid chromatography can compensate for this effect and help for studying single cell lines or types.
+Higher organisms are composed of specialized cells organized into tissues, such as skin, muscle, and blood.
+Each tissue consists of cells with specific functions, resulting in variations in protein expression.
+Bulk proteomics is a technique used to analyze the protein composition of a tissue sample, which contains all types of cells present in that particular tissue.
+This approach finds valuable applications for instance in oncology, where it can provide insights into the protein profile of a tumor.
+By studying oncogenic biomarkers and expression patterns, this knowledge can contribute to the development of screening methods and the customization of treatments based on observed expression patterns.
+Additionally, bulk proteomics can aid in identifying the progression of tumor development \citep{Kwon2021}.
+Bulk analysis expression profiles provide scientists with an average protein abundance across all cells in a sample, offering a comprehensive view of protein expression.
+Additionally, compared to single-cell methods, bulk samples typically contain a greater number of proteins.
+In fact the expanded library can be used to validate the performance of single-cell type and single-cell proteomics methods and bioinformatics pipeline \citep{Schoof2021}.
+However, in this type of analysis, the precise properties of individual cell types can only be inferred and remain obscured.
 
 ### Single cell type
+Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
+To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed \citep{Liou2015}.
+These techniques rely on the detection of surface proteins on cells and the specific binding of antibodies to these proteins.
+Antibodies, typically derived from the immune system of other species, bind to antigens in a lock-and-key manner.
+Antibodies are widely used in molecular biology and find applications in cell isolation.
+Depending on the desired application, the antibodies can be labeled with fluorescent markers (FACS), magnetic beads (MACS), or biotin (BACS).
+These methods have varying technical requirements, with flow cytometers being the most complex and costly.
+In MACS, magnetic beads are used to isolate cells from a solution by utilizing a magnet.
+However, mechanical forces involved in MACS can potentially damage the cells.
+BACS, a method developed in 2015, employs microbubbles to carry the antibodies for cellular isolation.
+It is important to note that FACS is considered the gold standard for validating the effectiveness of MACS and BACS, as it was used in the validation of these two methods.\citep{Liou2015}
 

### 2023-06-08 markdown/introduction.Rmd
scp done

@@
 The regulation is undertaken to provide biological organisms with an highly efficient output of their  genetic products called proteins.
-Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb, 2017}.
+Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
 Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as omics, to comprehensively analyze biomolecules in diverse contexts.
@@
 
-### Single cell type
+### Single-cell type
 Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
@@
 BACS, a method developed in 2015, employs microbubbles to carry the antibodies for cellular isolation.
-It is important to note that FACS is considered the gold standard for validating the effectiveness of MACS and BACS, as it was used in the validation of these two methods.\citep{Liou2015}
+The method showed improvements compared to MACS in regards of viability after sorting.
+It is important to note that FACS is considered the gold standard for validating the effectiveness of MACS and BACS due to its high purity and yield.
+FACS was used in the validation of these two methods \citep{Liou2015, Sutermaster2019}.
+Single-cell type proteomics reflects the protein ensemble of a specific cell type at a particular time, providing a focused perspective on the studied field compared to bulk methods \citep{Maes2020}.
 
-### Single cell proteomics
-
-Single-cell proteomics (=SCP) provides a snapshot of the underlying processes in a cell.
-These proteins can be influenced by various environmental factors and the understanding of abundance or differential expression is crucial to understand the principles of cellular behavior.
-SCP gives insight in a qualitative but also a quantitative way and can elucidate responses of drugs, infectious diseases, and the development of an organism.
-In the last 10 years SCP became an indispensable tool for global research communities (Minakshi et al., 2019).
+### Single-cell proteomics
+Cell sorting techniques primarily rely on extracellular proteins to identify and separate different cell types.
+However, the intracellular content of cells can still exhibit variability in experiments.
+Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the underlying principles of cell type behavior, such as differentiation.
+The immune system, for instance, relies on specialized cells for specific functions, and the acquisition of these functions occurs through a process called differentiation.
+Studying differentiation in multiple cells can be challenging due to the inherent heterogeneity during the initiation of differentiation.
+By analyzing one cell at a time, stimulating and observing it throughout the differentiation process, researchers can capture the entire process at a single-cell level.
+However, sample preparation and computational analysis in SCP can be tedious, and statistical testing can be challenging due to low cell counts.
+In 2021, Specht et al.
+developed the SCoPE2 sample preparation and analysis pipeline, focusing on macrophage-monocyte differentiation experiments \citep{Specht2021}.
+This pipeline, which reduced labor time, served as a reference model for subsequent projects.
+SCP provides insights qualitatively and quantitatively, enabling the study of drug responses, infectious diseases, and organism development.
+Over the past decade, single cell techniques became indispensable tools for global research communities \citep{Minakshi2019}.
 
@@
 To identify protein abundance from a single type of cell, cell sorting techniques needed to be improved and applied to MS (Minakshi et al., 2019).
-Single cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
+Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
 Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics (Ye et al., 2022).

### 2023-06-13 markdown/introduction.Rmd
ms explanation moved from methods to intro

@@
 The regulation is undertaken to provide biological organisms with an highly efficient output of their  genetic products called proteins.
-Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
+Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb, 2017}.
 Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as omics, to comprehensively analyze biomolecules in diverse contexts.
@@
 Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
+For instance pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
+These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications and more \citep{Beltran2017}.
 By studying the proteome, which encompasses all the proteins present in a biological system at a specific moment, we can gain a comprehensive understanding of the hierarchical organization and functional dynamics of biological processes.
@@
 
-### Single-cell type
+### Single-cell type proteomics
 Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
@@
 However, the intracellular content of cells can still exhibit variability in experiments.
-Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the underlying principles of cell type behavior, such as differentiation.
+Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the  principles of cell type behavior, such as differentiation.
 The immune system, for instance, relies on specialized cells for specific functions, and the acquisition of these functions occurs through a process called differentiation.
@@
 
+## Capturing the proteome
 An early approach of qualitative analysis of the cellular proteome involved labeling with fluorescent antibodies and imaging.
-The major disadvantage of this technique was the limitation to only stain a few proteins per cell.
-For quantification procedures such as single-cell Western blots, immunoassays or CyTOF have been used.
+The major disadvantage of this technique was the limitation to only stain a few proteins per cell and the requirement of a-priori knowledge of the protein to analyze.
+For quantification procedures such as Western blots, immunoassays or cytometry by time of flight (CyTOF) have been used.
 Other disadvantages are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
-
-One of those techniques involved RNA-sequencing.
-Since RNA involves also non-coding RNA, the amount of RNA is often not proportional to the content of proteins in a cell.
-So the proteinaceous content of a cell was only predicted and quantitative analysis was not possible.
-
-## Effectors of the proteome
-
-### External effectors
-
-When pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
-These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications and more (Jean Beltran et al., 2017).
-
-## Methods for Quantifications
-
-A holistic approach to analyze the expression can be sequencing of RNA (RNAseq) or mass-spectrometry (MS).
+Early quantitative methods in molecular biology were assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay measuring the entire protein concentration.
+These methods are still in use to obtain total concentration for sample preparation.
+Another quantitative technique involved RNA-sequencing.
 RNAseq refers to the task of examining the transcriptome of bulk tissue sample  with deep-sequencing technologies.
 RNA in a sample includes mRNAs, non-coding RNAs, and small RNAs.
-One of the main disadvantages is, that the mapping relies on the knowledge of the DNA sequence of the observed organism (reference genome) (Wang et al., 2009).
-Another disadvantage is, that the amount of protein translated from the determined mRNA can only be estimated (Gygi et al., 1999).
-However, today these predictions can be done with computational methods such as LASSO and the accuracy of the estimate highly depends on selection and validation of this method (Magnusson et al., 2022).
-
+One of the main disadvantages is, that the mapping relies on the knowledge of the DNA sequence of the observed organism (reference genome) \citep{Wang2009}.
+Another disadvantage is, that the amount of protein translated from the determined mRNA can only be estimated \citep{Gygi1999}.
+However, today these predictions can be done with computational methods such as LASSO and the accuracy of the estimate highly depends on selection and validation of this method \citep{Magnusson2022}.
 When observing the protein ensemble of a cellular environment with a top-down approach  like DNAseq or RNAseq, the characteristics (e.g.
@@
 A bottom-up strategy can elucidate these properties of the protein ensemble at a given time leading to proteomics.
-Early methods in molecular biology were western blots with the disadvantage of having a-priori knowledge of the analyzed protein and quantitative protein assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay measuring the entire protein concentration.
 When studying impact of an external signal or stressor to a cell the abundance and   properties of a protein or protein-ensemble can enlighten the underlying pathway leading to a biological response.
-Since some proteins do not even function before post-translational modifications happen, a knowledge about these can be crucial for further studies.
+Since some proteins do not even function before post-translational modifications (PTMs) happen, a knowledge about these can be crucial for further studies.
+Mass spectrometry (MS) fusions the advantages of  the above mentioned techniques, because it combines the specificity of qualitative assays which are even able to sense (PTMs) with the quantitative accuracy of  protein concentration assays.
+Furthermore MS does not require any a-priori knowledge of the protein ensemble of the specimen, making it the method of choice for untargeted approaches.
 
 ### Mass Spectrometry
-
 Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
-The availability of gene sequences in databases and the ability to match proteins against those sequences with computational methods makes it possible to identify alterations of a sample on a protein level.
-These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylation, methylation or else \citep{Aebersold2003}.
-
 Mass spectrometers measure the mass to charge ratio (m/z) of a particle from a fragmented larger molecule.
@@
 The ion sources charges molecules and accelerate them through a magnetic field.
-The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal (Parker et al., 2010).
-MS provides great sensitivity for traces, hence getting an average concentration for all biomolecules in a sample.
-To identify protein abundance from a single type of cell, cell sorting techniques needed to be improved and applied to MS (Minakshi et al., 2019).
+The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal \citep{Parker2010}.
+However before acquiring any data, sample preparation and pre-processing is required.
+After sample preparation the first step involves tryptic digestion.
+It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the protease enzyme trypsin.
+This process is known as proteolysis.
+The enyme cleaves specifically at the caroboxyl side of arginine and lysis residues.
+Afterwards a charge remains on the residues making them detectable for the mass spectrometer.
+It is proven that proteins which are digested to the limit, meaning no further proteolysis can be induced, provide the best results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.
+
+To observe multiple samples in one run, labeling is needed to identify each sample within a batch. 
+Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
+Physicochemical properties of the peptides can change the ionization efficiency and detectability of the target.
+However, when comparing the same analyte between multiple runs of labeled peptides, differences in the reporter ion intensity reflect the abundance of those.
+Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the peptide.
+The two techniques for labeling peptides are metabolic labeling and isobaric labling.
+In metabolic labeling the cells were fed with aminoacids containing heavy isotopes, it is the method of choice in order to label peptides at the earliest possible level.
+This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants.
+Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
+Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
+This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
+The limitation of this method is, that it can only be applied to a human cell culture (in vitro) or to model organisms (in vivo).
+Isobaric labeling solely applies to in vitro techniques, with TMT and iTRAQ.
+The data analyzed by the developed pipeline primarily utilized TMT tags.
+Tandem mass tag (TMT) reagents enable differentiation of multiple samples in a single mass spectrometry (MS) run.
+Each sample is individually labeled with TMT reagents and then pooled together, this procedure is called multiplexing.
+TMTs have the same charge and differ only by their isotopic masses, the peaks observed for each sample are called reporter ions (RI).
+Each RI and sample is interpreted as one channel in downstream analysis.
+This technique allows for the identification, enrichment, and quantification of low-abundance peptide ions, which is particularly valuable in single-cell techniques.
+With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
+However, a drawback of isobaric labeling is the presence of co-fragmentation signals in the spectrogram, which requires data normalization to remove unwanted contributions \citep{Marx2019, Budnik2018}.
+Isotopic distribution in TMTs, reflecting the natural distribution, can be corrected during data acquisition by considering a defined spread in other channels.
 Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
-Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics (Ye et al., 2022).
+Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics \citep{Ye2022}.
 
-MS provides great sensitivity for traces, hence getting an average concentration for all biomolecules in a sample.
-To identify protein abundance from a single type of cell, cell sorting techniques needed to be improved and applied to MS (Minakshi et al., 2019).
-Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
-Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics (Ye et al., 2022).
+Before ionization peptides need to be further separated according to their chemical properties, size or species by liquid chromatography (LC), because the number of peptide processed simultaneously by the MS device is limited.
+In the case of mass spectrometry, liquid chromatography especially high performance LC (HPLC) is the technique of choice and replaces the instrumentation-wise simpler technique gel electrophoresis.
+In HPLC the molecules eluate through a narrow column with high pressures (50-350 bar), compared to LC where the separation is gravity dependent.
+Similar to gel electrophoresis molecules can be separated upon different properties depending on the type of the chromatography column.
+In a typical proteomics experiment peptides are separated with reversed phase chromatography (RPLC).
+The principle behind RPLC is a column with a nonpolar stationary phase which interacts with the nonpolar residues of the peptides.
+Peptides were fractionated according to their hydrophobicity.
+
+To analyze a biological sample consisting of peptides in solution the liquid needs to be vaporized into gas phase.
+Two techniques are capable of this procedure.
+Electrospray ionization (ESI) pushes the analyte through a capillary while applying an electric current to the liquid, vaporizing the sample to a  charged aerosol.
+Peptides are further fragmented according to their chemical properties and can be further handled in the mass spectrometer.
+The fragmented peptides are now in charged droplets separated based on their surface charge, splitting further into smaller droplets until they reach the gas phase as ions.
+Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
+In the ion evaporation model the droplets shrink by evaporation until ions are expelled \citep{Iribarne1976}.
+The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
+In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
+
+Subsequently after the ionization peptides are accelerated through the magnetic field generated by the mass analyzer.
+Six general types of mass analyzers are availaible on the market, there are Quadropole, time of flight (TOF), magnetic sector, electrostatic sector, quadropole ion trap and ion cyclotron resonance -mass analyzer.
+The function of the mass analyzer is to separate ions based on their mass to charge ratio (m/z).
+
+Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their (m/z) consecutively.
+Ion trapping enables the technique of coupled mass spectrometers, where ions were further fragmented by collision induced dissociation and analyzed subsequently.
+The detector transforms the impact energy of the ion into an electric signal, called spectrum which can be processed computationally.
+
+Given the high resolution of MS data, algorithms are employed to convert the raw signal into an interpretable form.
+Software packages like MaxQuant are commonly used to process the data, providing it for further analysis and statistical testing.
+Other software solutions include Protein Discoverer by Thermo Fisher and various R packages.
+
+The data obtained from mass spectrometry has three dimensions: m/z ratio, intensity, and retention time.
+To separate peaks from each other, algorithms are used to identify local minima in the data.
+The centroid of each peak is determined by fitting a Gaussian peak shape, which can be interpreted as locating the peaks of each m/z spectrum as a function of retention time.
+The centroid of a peak corresponds to an isotope.
+
+In order to decipher the isotopic distribution of a biomolecule, MaxQuant employs a vertex-based approach.
+This process, referred to as de-isotoping, utilizes graph theory.
+It creates a vertex for each individual peak and connects them with their potential isotopic counterparts.
+This is achieved by determining the mass proportion of an average amino acid to its respective isotope using a concept known as averagine \citep{Senko1995}.
+By applying this procedure, the number of data points is reduced by a factor of ten, and each peak represents a small biomolecule.
+
+The subsequent step in data acquisition involves the detection of labels for quantification purposes.
+Isotopic pairs of the label (such as N13, N14, N15) present in the tag or aminoacid are identified by convolving the two measured isotope patterns with the theoretical isotope patterns.
+Through an iterative process using a least-squares method, the best fit is determined, enabling the identification of the specific channel or sample.
+
+The intensity-weighted average of the MS peak centroids, as described as 3D peak identification, corresponds to the mass of the peptide.
+The accuracy of mass measurements relies on the specific analyzer used.
+For an Orbitrap-type analyzer (which was used during this project), MaxQuant applies a correction value of 1 ppm (parts per million).
+Autocorrelation between centroids is accounted for by considering only well-identified peptides.
+Published data indicates that the mass precision within an MS experiment typically ranges around 10^-7.
+
+Each peptide is represented by its individual fingerprint in the MS spectrum, which is based on the sequence and modifications of amino acids.
+The m/z ratio of the amino acids can be calculated, and the resulting peaks are interpreted as an amino acid sequence.
+However, only peptide fragments are visible in the spectrum due to peptide fragmentation.
+To identify proteins, peptides are matched against a sequence database \citep{Cox2008}.
+Sequence databases are typically stored as .fasta files and can be downloaded from the UniProt website (www.uniprot.org).
+The availability of gene sequences in databases and the computational matching of peptides against those sequences enable the identification of alterations in a sample at the protein level.
+These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylation, methylation \citep{Aebersold2003}.
+The peptide identification (P-) score reflects the goodness of fit between the data and the identified sequence in the database, taking into account the length of the peptide.
+This score is used to calculate the posterior error probability, which is then utilized to estimate the false discovery rate (FDR).
+The FDR calculation provides a measure of the rate at which false identifications are expected in the dataset.
+
+After these calculations, the identified peptides can be grouped and aggregated according to their respective proteins for quantification.
+Various metrics that indicate the performance of the peptide search can be utilized in downstream analysis.
+Within the identified peptide groups, a "razor protein" refers to the protein with the highest number of ambiguously identified peptides.
+Quantification is performed by considering only unique peptides for comparison.
+Posterior error probabilities, which represent the likelihood of the identified peptide being a random event, are multiplied together, and only distinct sequences with the highest scores are considered for further analysis.

### 2023-06-13 markdown/introduction.tex
ms explanation moved from methods to intro

--- 2023-06-08 markdown/introduction.Rmd
@@
----
-title: " "
-output: pdf_document
-header-includes:
-  - \usepackage{float}
-  - \usepackage{sectsty}
----
-# Introduction
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\usepackage{float}
+\usepackage{sectsty}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
-## From the Central Dogma of Molecular Biology to Omics and Systems Biology
-Before understanding the underlying principle biological systems undergo and the initial exploration of molecular biology, the processes of live were roughly inferred and not well understood.
-The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins.
-This framework consists of two major processes, which involve transcription, converting DNA into RNA, and translation, converting RNA into functional proteins.
-In between those processes regulation takes base on a molecular level, by altering structure with chemical adaptations.
-The regulation is undertaken to provide biological organisms with an highly efficient output of their  genetic products called proteins.
-Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
-Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as omics, to comprehensively analyze biomolecules in diverse contexts.
-Omics research has emerged as an approach for comprehensively studying biomolecules and their interactions within biological systems.
-Omics encompasses various fields, including genomics, transcriptomics, proteomics, metabolomics, and others, each focusing on a specific class of biomolecules.
-By analyzing multiple omics layers, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks with computational methods.
-Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks.
-Imagining biological systems as complex, multi-layer networks, the ensemble of proteins, known as the proteome, plays a critical role in this interactive structure with many functions such as structural integrity, catalyzing chemical reaction and regulation of cellular funcitons.
+\author{}
+\date{\vspace{-2.5em}}
 
-## Proteomics
-The central dogma of molecular biology offers insights not only at the level of individual terms but also provides a hierarchical understanding of biology.
-Hierarchical organization is the underlying principle in all topics of biology.
-Proteins exhibit hierarchical structures that contribute to their functionality.
-At the primary structure level, proteins are composed of smaller building blocks called amino acids.
-Combining them results in the protein secondary structure such as alpha helix and beta sheet.
-These secondary structures further assemble to create higher-order tertiary structures, which represent specific domains within the protein.
-Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
-By studying the proteome, which encompasses all the proteins present in a biological system at a specific moment, we can gain a comprehensive understanding of the hierarchical organization and functional dynamics of biological processes.
+\begin{document}
 
-### Bulk proteomics
-Higher organisms are composed of specialized cells organized into tissues, such as skin, muscle, and blood.
-Each tissue consists of cells with specific functions, resulting in variations in protein expression.
-Bulk proteomics is a technique used to analyze the protein composition of a tissue sample, which contains all types of cells present in that particular tissue.
-This approach finds valuable applications for instance in oncology, where it can provide insights into the protein profile of a tumor.
-By studying oncogenic biomarkers and expression patterns, this knowledge can contribute to the development of screening methods and the customization of treatments based on observed expression patterns.
-Additionally, bulk proteomics can aid in identifying the progression of tumor development \citep{Kwon2021}.
-Bulk analysis expression profiles provide scientists with an average protein abundance across all cells in a sample, offering a comprehensive view of protein expression.
-Additionally, compared to single-cell methods, bulk samples typically contain a greater number of proteins.
-In fact the expanded library can be used to validate the performance of single-cell type and single-cell proteomics methods and bioinformatics pipeline \citep{Schoof2021}.
-However, in this type of analysis, the precise properties of individual cell types can only be inferred and remain obscured.
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-### Single-cell type
-Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
-To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed \citep{Liou2015}.
-These techniques rely on the detection of surface proteins on cells and the specific binding of antibodies to these proteins.
-Antibodies, typically derived from the immune system of other species, bind to antigens in a lock-and-key manner.
-Antibodies are widely used in molecular biology and find applications in cell isolation.
-Depending on the desired application, the antibodies can be labeled with fluorescent markers (FACS), magnetic beads (MACS), or biotin (BACS).
-These methods have varying technical requirements, with flow cytometers being the most complex and costly.
-In MACS, magnetic beads are used to isolate cells from a solution by utilizing a magnet.
-However, mechanical forces involved in MACS can potentially damage the cells.
-BACS, a method developed in 2015, employs microbubbles to carry the antibodies for cellular isolation.
-The method showed improvements compared to MACS in regards of viability after sorting.
-It is important to note that FACS is considered the gold standard for validating the effectiveness of MACS and BACS due to its high purity and yield.
-FACS was used in the validation of these two methods \citep{Liou2015, Sutermaster2019}.
-Single-cell type proteomics reflects the protein ensemble of a specific cell type at a particular time, providing a focused perspective on the studied field compared to bulk methods \citep{Maes2020}.
+\hypertarget{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}{%
+\subsection{From the Central Dogma of Molecular Biology to Omics and
+Systems
+Biology}\label{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}}
 
-### Single-cell proteomics
-Cell sorting techniques primarily rely on extracellular proteins to identify and separate different cell types.
-However, the intracellular content of cells can still exhibit variability in experiments.
-Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the underlying principles of cell type behavior, such as differentiation.
-The immune system, for instance, relies on specialized cells for specific functions, and the acquisition of these functions occurs through a process called differentiation.
-Studying differentiation in multiple cells can be challenging due to the inherent heterogeneity during the initiation of differentiation.
-By analyzing one cell at a time, stimulating and observing it throughout the differentiation process, researchers can capture the entire process at a single-cell level.
-However, sample preparation and computational analysis in SCP can be tedious, and statistical testing can be challenging due to low cell counts.
-In 2021, Specht et al.
-developed the SCoPE2 sample preparation and analysis pipeline, focusing on macrophage-monocyte differentiation experiments \citep{Specht2021}.
-This pipeline, which reduced labor time, served as a reference model for subsequent projects.
-SCP provides insights qualitatively and quantitatively, enabling the study of drug responses, infectious diseases, and organism development.
-Over the past decade, single cell techniques became indispensable tools for global research communities \citep{Minakshi2019}.
+Before understanding the underlying principle biological systems undergo
+and the initial exploration of molecular biology, the processes of live
+were roughly inferred and not well understood. The central dogma of
+molecular biology serves as the cornerstone of biological processes,
+providing a framework for understanding how genetic information is
+converted into functional proteins. This framework consists of two major
+processes, which involve transcription, converting DNA into RNA, and
+translation, converting RNA into functional proteins. In between those
+processes regulation takes base on a molecular level, by altering
+structure with chemical adaptations. The regulation is undertaken to
+provide biological organisms with an highly efficient output of their
+genetic products called proteins. Crick's formulation of the central
+dogma has guided research in molecular biology for over 60 years
+\citep{Cobb2017}. Over the past six decades, researchers have explored
+various aspects of this principle, collectively referred to as omics, to
+comprehensively analyze biomolecules in diverse contexts. Omics research
+has emerged as an approach for comprehensively studying biomolecules and
+their interactions within biological systems. Omics encompasses various
+fields, including genomics, transcriptomics, proteomics, metabolomics,
+and others, each focusing on a specific class of biomolecules. By
+analyzing multiple omics layers, researchers can gain a holistic view of
+biological processes, uncover regulatory mechanisms, and identify key
+players in complex biological networks with computational methods.
+Systems biology takes a quantitative approach to investigate the
+interactions, dynamics, and emergent properties of biomolecular
+networks. Imagining biological systems as complex, multi-layer networks,
+the ensemble of proteins, known as the proteome, plays a critical role
+in this interactive structure with many functions such as structural
+integrity, catalyzing chemical reaction and regulation of cellular
+funcitons.
 
-An early approach of qualitative analysis of the cellular proteome involved labeling with fluorescent antibodies and imaging.
-The major disadvantage of this technique was the limitation to only stain a few proteins per cell.
-For quantification procedures such as single-cell Western blots, immunoassays or CyTOF have been used.
-Other disadvantages are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
+\hypertarget{proteomics}{%
+\subsection{Proteomics}\label{proteomics}}
 
-One of those techniques involved RNA-sequencing.
-Since RNA involves also non-coding RNA, the amount of RNA is often not proportional to the content of proteins in a cell.
-So the proteinaceous content of a cell was only predicted and quantitative analysis was not possible.
+The central dogma of molecular biology offers insights not only at the
+level of individual terms but also provides a hierarchical understanding
+of biology. Hierarchical organization is the underlying principle in all
+topics of biology. Proteins exhibit hierarchical structures that
+contribute to their functionality. At the primary structure level,
+proteins are composed of smaller building blocks called amino acids.
+Combining them results in the protein secondary structure such as alpha
+helix and beta sheet. These secondary structures further assemble to
+create higher-order tertiary structures, which represent specific
+domains within the protein. Finally, the quaternary structure describes
+the functional state of the protein as a whole at a given time. By
+studying the proteome, which encompasses all the proteins present in a
+biological system at a specific moment, we can gain a comprehensive
+understanding of the hierarchical organization and functional dynamics
+of biological processes.
 
-## Effectors of the proteome
+\hypertarget{bulk-proteomics}{%
+\subsubsection{Bulk proteomics}\label{bulk-proteomics}}
 
-### External effectors
+Higher organisms are composed of specialized cells organized into
+tissues, such as skin, muscle, and blood. Each tissue consists of cells
+with specific functions, resulting in variations in protein expression.
+Bulk proteomics is a technique used to analyze the protein composition
+of a tissue sample, which contains all types of cells present in that
+particular tissue. This approach finds valuable applications for
+instance in oncology, where it can provide insights into the protein
+profile of a tumor. By studying oncogenic biomarkers and expression
+patterns, this knowledge can contribute to the development of screening
+methods and the customization of treatments based on observed expression
+patterns. Additionally, bulk proteomics can aid in identifying the
+progression of tumor development \citep{Kwon2021}. Bulk analysis
+expression profiles provide scientists with an average protein abundance
+across all cells in a sample, offering a comprehensive view of protein
+expression. Additionally, compared to single-cell methods, bulk samples
+typically contain a greater number of proteins. In fact the expanded
+library can be used to validate the performance of single-cell type and
+single-cell proteomics methods and bioinformatics pipeline
+\citep{Schoof2021}. However, in this type of analysis, the precise
+properties of individual cell types can only be inferred and remain
+obscured.
 
-When pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
-These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications and more (Jean Beltran et al., 2017).
+\hypertarget{single-cell-type}{%
+\subsubsection{Single-cell type}\label{single-cell-type}}
 
-## Methods for Quantifications
+Taking tissue samples can lead to an averaging effect across the entire
+cellular ensemble, making it difficult to discern specific cell types.
+To overcome this limitation, cell sorting techniques such as
+fluorescence-activated cell sorting (FACS), magnetic-activated cell
+sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed
+\citep{Liou2015}. These techniques rely on the detection of surface
+proteins on cells and the specific binding of antibodies to these
+proteins. Antibodies, typically derived from the immune system of other
+species, bind to antigens in a lock-and-key manner. Antibodies are
+widely used in molecular biology and find applications in cell
+isolation. Depending on the desired application, the antibodies can be
+labeled with fluorescent markers (FACS), magnetic beads (MACS), or
+biotin (BACS). These methods have varying technical requirements, with
+flow cytometers being the most complex and costly. In MACS, magnetic
+beads are used to isolate cells from a solution by utilizing a magnet.
+However, mechanical forces involved in MACS can potentially damage the
+cells. BACS, a method developed in 2015, employs microbubbles to carry
+the antibodies for cellular isolation. The method showed improvements
+compared to MACS in regards of viability after sorting. It is important
+to note that FACS is considered the gold standard for validating the
+effectiveness of MACS and BACS due to its high purity and yield. FACS
+was used in the validation of these two methods
+\citep{Liou2015, Sutermaster2019}. Single-cell type proteomics reflects
+the protein ensemble of a specific cell type at a particular time,
+providing a focused perspective on the studied field compared to bulk
+methods \citep{Maes2020}.
 
-A holistic approach to analyze the expression can be sequencing of RNA (RNAseq) or mass-spectrometry (MS).
-RNAseq refers to the task of examining the transcriptome of bulk tissue sample  with deep-sequencing technologies.
-RNA in a sample includes mRNAs, non-coding RNAs, and small RNAs.
-One of the main disadvantages is, that the mapping relies on the knowledge of the DNA sequence of the observed organism (reference genome) (Wang et al., 2009).
-Another disadvantage is, that the amount of protein translated from the determined mRNA can only be estimated (Gygi et al., 1999).
-However, today these predictions can be done with computational methods such as LASSO and the accuracy of the estimate highly depends on selection and validation of this method (Magnusson et al., 2022).
+\hypertarget{single-cell-proteomics}{%
+\subsubsection{Single-cell proteomics}\label{single-cell-proteomics}}
 
-When observing the protein ensemble of a cellular environment with a top-down approach  like DNAseq or RNAseq, the characteristics (e.g.
-post-translational modifications or charge) and local concentration of the proteins remain hidden.
-The cause of this phenomenon is that only a fraction of the DNA is transcribed into mRNA and not all mRNA is further translated into protein.
-Furthermore multiple proteins can be derived from a single DNA-sequence with alternative splicing and different proteins coming from non related sequences acting like building blocks can form protein complexes and/or Protein-RNA complexes.
-Cells are highly efficient and do not waste energy by producing obsolete proteins, moreover the degradation of proteins is an important energy recovery mechanism.
-A bottom-up strategy can elucidate these properties of the protein ensemble at a given time leading to proteomics.
-Early methods in molecular biology were western blots with the disadvantage of having a-priori knowledge of the analyzed protein and quantitative protein assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay measuring the entire protein concentration.
-When studying impact of an external signal or stressor to a cell the abundance and   properties of a protein or protein-ensemble can enlighten the underlying pathway leading to a biological response.
-Since some proteins do not even function before post-translational modifications happen, a knowledge about these can be crucial for further studies.
+Cell sorting techniques primarily rely on extracellular proteins to
+identify and separate different cell types. However, the intracellular
+content of cells can still exhibit variability in experiments.
+Single-cell proteomics (SCP) addresses this limitation by providing a
+snapshot of the underlying processes within an individual cell,
+eliminating this variability and allowing for the elucidation of the
+underlying principles of cell type behavior, such as differentiation.
+The immune system, for instance, relies on specialized cells for
+specific functions, and the acquisition of these functions occurs
+through a process called differentiation. Studying differentiation in
+multiple cells can be challenging due to the inherent heterogeneity
+during the initiation of differentiation. By analyzing one cell at a
+time, stimulating and observing it throughout the differentiation
+process, researchers can capture the entire process at a single-cell
+level. However, sample preparation and computational analysis in SCP can
+be tedious, and statistical testing can be challenging due to low cell
+counts. In 2021, Specht et al.~developed the SCoPE2 sample preparation
+and analysis pipeline, focusing on macrophage-monocyte differentiation
+experiments \citep{Specht2021}. This pipeline, which reduced labor time,
+served as a reference model for subsequent projects. SCP provides
+insights qualitatively and quantitatively, enabling the study of drug
+responses, infectious diseases, and organism development. Over the past
+decade, single cell techniques became indispensable tools for global
+research communities \citep{Minakshi et al., 2019}.
 
-### Mass Spectrometry
+An early approach of qualitative analysis of the cellular proteome
+involved labeling with fluorescent antibodies and imaging. The major
+disadvantage of this technique was the limitation to only stain a few
+proteins per cell. For quantification procedures such as single-cell
+Western blots, immunoassays or CyTOF have been used. Other disadvantages
+are the ability to permeate cells, accessibility and binding of the
+epitope and the creation of specific antibodies for a given protein
+\citep{Budnik2018}.
 
-Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
-The availability of gene sequences in databases and the ability to match proteins against those sequences with computational methods makes it possible to identify alterations of a sample on a protein level.
-These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylation, methylation or else \citep{Aebersold2003}.
+One of those techniques involved RNA-sequencing. Since RNA involves also
+non-coding RNA, the amount of RNA is often not proportional to the
+content of proteins in a cell. So the proteinaceous content of a cell
+was only predicted and quantitative analysis was not possible.
 
-Mass spectrometers measure the mass to charge ratio (m/z) of a particle from a fragmented larger molecule.
-This is achieved by a physical procedure done with a device which is made of three major components: the ion source, an analyzer, and a detector.
-The ion sources charges molecules and accelerate them through a magnetic field.
-The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal (Parker et al., 2010).
-MS provides great sensitivity for traces, hence getting an average concentration for all biomolecules in a sample.
-To identify protein abundance from a single type of cell, cell sorting techniques needed to be improved and applied to MS (Minakshi et al., 2019).
-Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
-Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics (Ye et al., 2022).
+\hypertarget{effectors-of-the-proteome}{%
+\subsection{Effectors of the proteome}\label{effectors-of-the-proteome}}
 
-MS provides great sensitivity for traces, hence getting an average concentration for all biomolecules in a sample.
-To identify protein abundance from a single type of cell, cell sorting techniques needed to be improved and applied to MS (Minakshi et al., 2019).
-Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
-Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics (Ye et al., 2022).
+\hypertarget{external-effectors}{%
+\subsubsection{External effectors}\label{external-effectors}}
+
+When pathogens interact with cells of an organism a cascade of
+biochemical reactions takes place. These reactions influence the outcome
+of the cellular proteome in regards of localization, abundance,
+post-translational modifications and more (Jean Beltran et al., 2017).
+
+\hypertarget{methods-for-quantifications}{%
+\subsection{Methods for
+Quantifications}\label{methods-for-quantifications}}
+
+A holistic approach to analyze the expression can be sequencing of RNA
+(RNAseq) or mass-spectrometry (MS). RNAseq refers to the task of
+examining the transcriptome of bulk tissue sample with deep-sequencing
+technologies. RNA in a sample includes mRNAs, non-coding RNAs, and small
+RNAs. One of the main disadvantages is, that the mapping relies on the
+knowledge of the DNA sequence of the observed organism (reference
+genome) (Wang et al., 2009). Another disadvantage is, that the amount of
+protein translated from the determined mRNA can only be estimated (Gygi
+et al., 1999). However, today these predictions can be done with
+computational methods such as LASSO and the accuracy of the estimate
+highly depends on selection and validation of this method (Magnusson et
+al., 2022).
+
+When observing the protein ensemble of a cellular environment with a
+top-down approach like DNAseq or RNAseq, the characteristics
+(e.g.~post-translational modifications or charge) and local
+concentration of the proteins remain hidden. The cause of this
+phenomenon is that only a fraction of the DNA is transcribed into mRNA
+and not all mRNA is further translated into protein. Furthermore
+multiple proteins can be derived from a single DNA-sequence with
+alternative splicing and different proteins coming from non related
+sequences acting like building blocks can form protein complexes and/or
+Protein-RNA complexes. Cells are highly efficient and do not waste
+energy by producing obsolete proteins, moreover the degradation of
+proteins is an important energy recovery mechanism. A bottom-up strategy
+can elucidate these properties of the protein ensemble at a given time
+leading to proteomics. Early methods in molecular biology were western
+blots with the disadvantage of having a-priori knowledge of the analyzed
+protein and quantitative protein assays like the UV-vis-, Bradford-
+and/or bicinchoninic-acid-assay measuring the entire protein
+concentration. When studying impact of an external signal or stressor to
+a cell the abundance and properties of a protein or protein-ensemble can
+enlighten the underlying pathway leading to a biological response. Since
+some proteins do not even function before post-translational
+modifications happen, a knowledge about these can be crucial for further
+studies.
+
+\hypertarget{mass-spectrometry}{%
+\subsubsection{Mass Spectrometry}\label{mass-spectrometry}}
+
+Mass spectrometry enables qualitative and quantitative analysis of the
+entire repertoire of a biological sample. The availability of gene
+sequences in databases and the ability to match proteins against those
+sequences with computational methods makes it possible to identify
+alterations of a sample on a protein level. These alterations can rely
+on the sequence level or could be to post-translational modifications
+(PTMs) such as phosphorylation, methylation or else
+\citep{Aebersold2003}.
+
+Mass spectrometers measure the mass to charge ratio (m/z) of a particle
+from a fragmented larger molecule. This is achieved by a physical
+procedure done with a device which is made of three major components:
+the ion source, an analyzer, and a detector. The ion sources charges
+molecules and accelerate them through a magnetic field. The analyzer
+separates particles according to their mass to charge ratio and the
+detector senses charged particles and amplifies their signal (Parker et
+al., 2010). MS provides great sensitivity for traces, hence getting an
+average concentration for all biomolecules in a sample. To identify
+protein abundance from a single type of cell, cell sorting techniques
+needed to be improved and applied to MS (Minakshi et al., 2019).
+Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric
+labeling with a tandem mass tag (TMT) to enhance signal intensity of a
+protein species. Loading of a single-cell channel (SCC) with the right
+amount can be balanced to the protein concentration of a single cell,
+leading to single-cell proteomics (Ye et al., 2022).
+
+MS provides great sensitivity for traces, hence getting an average
+concentration for all biomolecules in a sample. To identify protein
+abundance from a single type of cell, cell sorting techniques needed to
+be improved and applied to MS (Minakshi et al., 2019). Single-cell
+proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with
+a tandem mass tag (TMT) to enhance signal intensity of a protein
+species. Loading of a single-cell channel (SCC) with the right amount
+can be balanced to the protein concentration of a single cell, leading
+to single-cell proteomics (Ye et al., 2022).
+
+\end{document}

### 2023-06-13 markdown/methods.Rmd
ms explanation moved from methods to intro

@@
 
-### Labeling techniques
-
-For differential analysis proteins need to be labeled to compare mass to charge intensities in order to quantify observed peptides.
-Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
-Physicochemical properties of the proteins can change the ionization efficiency and detectability of the target.
-However, when comparing the same analyte between multiple runs of labeled proteins, differences in the mass spectrum reflect the abundance of those.
-Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the protein.
-
 #### Metabolic labeling
-
-Feeding cells with aminoacids containing heavy isotopes, is the method of choice in order to label peptides at the earliest possible level.
-This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants.
-Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
-Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
-This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
 
@@
 
-Tandem mass tag (TMT) reagents enable to differentiate multiple samples analyzing in one MS run. 
-The samples are labeled individually and pooled afterwards, this procedure is called multiplexing.
-TMTs have the same charge and differ only by their isotopic masses, the peaks found for each sample are called reporter ions (RI).
-Each RI and sample is interpreted as one channel in downstream analysis.
-The identification of these RI leads to an enrichment and identification of low abundance peptide ions which is common especially in single-cell techniques.
-With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
-The disadvantage of isobaric labeling is, that the co-fragmentation signals can be observed in the spectrogram and the data needs to be normalized in order to remove unwanted contribution \citep{Marx2019, Budnik2018}.
-Furthermore TMTs have an isotopic distribution according to the distribution found in nature.
-This can be corrected during data-acquisition as a defined spread in other channels.
-
 ### Instrumentation
@@
 
-In order to separate proteins according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
+In order to separate peptides according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
+For the reference data used for developing the pipeline the following intrumentation was used.
+EASY-Spray™ HPLC column, ES802A, 2 $\mu$m × 75 $\mu$m ID × 250 mm (ThermoFisher Scientific)
+Acclaim PepMap trap column, 2 $\mu$m × 75 $\mu$m ID × 2 cm (ThermoFisher Scientific)
 
@@
 ##### Ionization
+For the reference data ESI ionization was used.
 
-In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase.
-Two techniques are capable of this procedure.
-Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol.
-Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer.
-The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion.
-Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
-In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled.
-The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
-In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
-
-Matrix-assisted laser desorption/ionization (MALDI)
-
-##### MS.1
-
-##### Coupled mass-spectrometry (MS/MS) & MS.2
-
-In order to enhance sequence identification, two MS devices are built in series.
-In the first run (MS1) the m/z is determined and the molecules are passed to the next device.
-Upon passing the molecules are fragmented into smaller ions and analyzed by the second MS.
-The fragmentation highly depends on the chemical bonds found in the molecule.
-The majority of these breaks occur on the peptide bond of the protein, although this is not guaranteed for all bonds and so it can happen that certain peptide ions have a low abundance \citep{Budnik2018}.
-These low abundant peptides will not be detected, hence the problem needs to be faced with another approach .
-A solution for this problem is molecular barcoding with labeling mentioned in the chapter labeling.
+##### MS/MS
+For the reference data used for developing the pipeline an Orbitrap™ Fusion™ Lumos™ tribrid mass spectrometer (ThermoFisher Scientific) was used.
+The instrumentation uses two subsequent runs in the analyzer (=Orbitrap) to increase mass accuracy.
 
@@
 
-### Acquisition
-
-Acquisition of the data was done with MaxQuant \citep{Cox2008} software package.
+### Data-acquisition and raw-file analysis
+For the data-acquisition thermo fisher acquisition software was used.
+Raw-file analysis of the data was done with MaxQuant \citep{Cox2008} software package.
 
 #### MS-Spectrum
-
-Each peptide is reflected by its` indivual fingerprint in the ms-spectrum.
-The fingerprint is based on the chemical properties and modifications of aminoacids.
-These aminoacids can be calculated through their m/z ration and after that interpreted as an aminoacid sequence.
-Due to fragmentation of the protein only peptide sequences are visible in the spectrum.
-In order to identify proteins, peptides are matched against a sequence database \citep{Cox2008}.
-Sequence Databases are simple .fasta files, which can be downloaded on the uniprot webpage (www.uniprot.org).
-
-Since ms data has a high resolution, algorithms are used to convert the raw signal to an interpretable form. 
-MaxQuant  is one of many sofware packages to process the data and provides it for further analysis and statistical testing.
-Other software solutions are Protein Discoverer Thermo Fisher or even packages for R.
  In this publication we will mainly focus on the data-acquisition with MaxQuant \citep{Cox2008}.
@@
 
-The three dimensions of the data are: m/z ratio, intensity and retention time.
-The algorithm finds local minima of the function in order to seperate peaks from each other.
-The centroid of the peak is detected by fitting a so called gaussian peak shape fitting.
-This can be interpreted as finding the peaks of each m/z spectrum as a function of time..
-The centroid of the peak refers to an isotope.
-
 #### Deisotoping
-
-To decrypt the istopic distribution of a biomolecule, MaxQuant creates a vertex of every single peak and connects them with their possible isotopic counterparts by finding the proportion of mass of an average aminoacid to its` respective isotope (averagine \citep{Senko1995}).
-Isotoping is the term of such procedure and it is enabled with graph theory.
-After this procedure the amount of data points are reduced by a tenfold and a single peak reflects a small biomolecule.
 
@@
 
-The next step in data-acquisition is the detection of labels for quantification.
-Isotopic pairs of the label (e.g.
-N13, N14, N15) contained in the tag or aminoacid can be identified by convoluting the two measured isotope patterns with the theoretical isotope patterns.
-With a least-square method the best fit is found iteratively and the channel/sample can be identified.
-
 #### Improving peptide mass accuracy
-
-The intensity-weighted average of the ms peak centroids (as described in the 3D peak identification) refers to the mass of the peptide.
-Corrections highly depend on the analyzer, MaxQuant uses for an orbitrap typed analyzer a correction value of 1ppm.
-Autocorrelation between centroids is compensated by only using well-identified peptides.
-As published the mass precision within a ms experiment ranges around 10e-7.
 
@@
 
-Biomolecules can now be searched in a database in forward and reverse direction.
-The peptide identification (P-) score indicates the fit of the data to the found sequence in the database according to the length of the peptide and is used to calculate the posterior error probability (FDR).
-The calculation of the false discovery rate is then calculated by taking FDR into contrast.
-
 #### Protein assembly
 
-After these calculation the identified peptides can be aggregated according to it`s respective protein and  quantified.
-The mentioned metrics indicating the performance of the peptide search can be used in downstream analysis.
-A so called razor peptide indicates the group with the highest number of identified peptides.
-Quantification is enabled by taking only unique peptides into contrast.
-Posterior error probabilities, which refers to the chance that the found peptide is a random event, are multiplied and only distinct sequences with the highest-scoring are used.
-
 ### Data processing
-
 Further analysis is done with R and respective packages such as bioconductor.
@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=18}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=16}
 library(Gmisc, quietly = TRUE)

### 2023-06-13 markdown/appendix.Rmd
discussion and conclusion started

@@
 ---
-# Appendix(ces)
+# Appendix
 
-## Appendix A: additional tables
+## Cells sorting protocol
+\begin{enumerate}
+\item 96 single cells from each donor were sorted into a single plate. Each well contained a single cell. As such, 12 different plates containing single cells from 12 different donors were prepared. 
+\item Carrier plates were also prepared. In the carrier plates, 100 cells from one donor were sorted into each well. 24 such wells (3 columns) were prepared for each donor.
+\end{enumerate}
 
-Insert content for additional tables here.
+### Digestion protocol
+\begin{enumerate}
+\item Prepare 25 ng/ $\mu$L trypsin in 100 mmol/L Triethylammonium bicarbonate buffer (TEAB).
+\item Add 25 ng/$\mu$L trypsin: 2 $\mu$L to the carrier samples, 1 $\mu$L to the single cell samples.
+\item Seal the plates.
+\item Centrifuge (2 min at 120 G).
+\item Incubate overnight at 37 °C.
+\end{enumerate}
 
-\newpage
+### Labeling protocol
+\begin{enumerate}
+\item Let the TMTpro labels come to room temperature so that all the condensation on the tube is gone.
+Dissolve the TMTpro reagents in anhydrous acetonitril (ACN) to obtain a concentration of 10 $\mu$g/$\mu$L.
+\item Add 1 $\mu$L of TMTpro reagents according to the tables. For one set of plates, you will need 16 $\mu$L of each label. It is recommended to aliquot a slightly larger volume, around 20 $\mu$L.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Prepare a 5% hydroxylamine solution in water.
+\item Add 1 $\mu$L of the 5% hydroxylamine solution to each well.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Cover the plates with cap strips.
+\end{enumerate}
 
-## Appendix B: additional figures
+### Pooling protocol
+\begin{enumerate}
+\item Wash a 10 $\mu$L glass microsyringe five times with ACN.
+\item Use the microsyringe to combine TMTpro-labeled samples into 36 LC autosampler vials with built-in glass inserts.
+\item Combine six carrier wells from each sample and add the carrier proteome to the 36 TMTpro sets.
+\item Combine the first set by taking well A1 from the first single-cell plate, well A2 from the second plate, well A3 from the third plate, and so on.
+For the next set, take well A1 from the second plate, well A2 from the third plate, and so on.
+Combine 12 sets following this scheme.
+Combine the next 24 sets by following the same scheme for rows B and C.
+\item Vacuum centrifuge the combined sets to dryness.
+\item Dissolve the dried samples in 7 $\mu$L of buffer A and transfer them to LC-MS/MS for further analysis.
+\end{enumerate}
 
-Insert content for additional figures here.
+### R-version info
 
-\newpage
+\begin{verbatim}
+R version 4.3.0 (2023-04-21)
+Platform: x86_64-pc-linux-gnu (64-bit)
+Running under: Ubuntu 22.04.2 LTS
 
-## Appendix C: code
-
-Insert code (if any) used during your dissertation work here.
+ [1] visNetwork_2.1.2            snowfall_1.84-6.2          
+ [3] snow_0.4-4                  piano_2.16.0               
+ [5] org.Hs.eg.db_3.17.0         AnnotationDbi_1.62.1       
+ [7] infotheo_1.2.0.1            clusterProfiler_4.8.1      
+ [9] plotly_4.10.1               sva_3.48.0                 
+[11] BiocParallel_1.34.2         genefilter_1.82.1          
+[13] mgcv_1.8-42                 nlme_3.1-162               
+[15] impute_1.74.1               CONSTANd_1.8.0             
+[17] limma_3.56.1                scater_1.28.0              
+[19] scuttle_1.10.1              reshape2_1.4.4             
+[21] dplyr_1.1.2                 magrittr_2.0.3             
+[23] ggplot2_3.4.2               SingleCellExperiment_1.22.0
+[25] scp_1.10.0                  QFeatures_1.10.0           
+[27] MultiAssayExperiment_1.26.0 SummarizedExperiment_1.30.1
+[29] Biobase_2.60.0              GenomicRanges_1.52.0       
+[31] GenomeInfoDb_1.36.0         IRanges_2.34.0             
+[33] S4Vectors_0.38.1            BiocGenerics_0.46.0        
+[35] MatrixGenerics_1.12.0       matrixStats_0.63.0         
+[37] spsComps_0.3.2.1            DT_0.28                    
+[39] shinydashboard_0.7.2        shinyWidgets_0.7.6         
+[41] shiny_1.7.4                
+\end{verbatim}

### 2023-06-13 markdown/conclusion.Rmd
discussion and conclusion started

@@
 
-The conclusion section should specify the key findings of your study, explain their wider significance in the context of the research field and explain how you have filled the knowledge gap that you have identified in the introduction.
-This is your chance to present to your reader the major take-home messages of your dissertation research.
-It should be similar in content to the last sentence of your summary abstract.
-It should not be a repetition of the first paragraph of the discussion.
-They can be distinguished in their connection to broader issues.
-The first paragraph of the discussion will tend to focus on the direct scientific implications of your work (i.e.
-basic science, fundamental knowledge) while the conclusion will tend to focus more on the implications of the results for society, conservation, etc.
+Through this project, we have demonstrated that computational analysis can be made accessible to a wide range of biologists by providing an interactive environment.
+It is important to recognize that there cannot be a one-size-fits-all approach for proteomics datasets.
+Therefore, our application offers easy-to-interpret visualizations that allow users to assess the patterns and trends in their observed datasets.
+The default options provided in the application are based on successful analysis pipelines such as ScoPE2 \citep{Specht2021}, serving as a starting point for customization at multiple levels.
+
+The developed pipeline and user interface serve as valuable tools for individual researchers and research communities, enabling them to analyze and visualize their data in meetings, even on non-performant computers.
+However, our future perspective involves deploying the application on a server, making it accessible to research groups around the world.
+This would enhance collaboration and facilitate the sharing of insights and findings across different geographical locations.
+By providing a centralized platform for proteomics analysis, we aim to foster scientific advancements and enable researchers to make significant contributions to their respective fields.
+Furthermore by having a central hub for proteomic data analysis and the use of artificial intelligence the biological meaning could be brought to a wider context.
+Since proteomic data could be saved without violating data protection laws, algorithms could be employed to learn from large datasets.

### 2023-06-13 markdown/discussion.Rmd
discussion and conclusion started

---
title: ''
output: pdf_document
---

# Discussion

The developed application and the proteomics experiment pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset. By utilizing an entropy-based visualization approach, the conservation of information can be validated, allowing users to select appropriate methods and adjust thresholds, cutoffs, and techniques accordingly. With the selection of factors to indicate over multiple plots, biases can be identified on various levels. Additionally, the analysis can be examined on an individual protein basis, enabling the identification of suspicious results. The statistics module provides users with a simple tool to assess and visualize the results, offering clarity to dense scatter plots through hover functions. Enrichment mapping links statistically significant proteins to either the KEGG pathway database or individual research datasets, depending on user selection. These final results place the data in a biological context, which can be explored interactively in graphical visualizations such as network.

Information theory and the concept of mutual information provide a framework for understanding the conservation of entropy in a dataset. Traditionally, the validation of data transformation and normalization methods has been limited to observing distributions through histograms or Quantile-Quantile plots. However, the developed approach goes beyond these conventional methods, providing a deeper context for the variation in the data. By allowing users to choose the method that preserves the maximum amount of information, the approach enables researchers to extract the most valuable insights from their experiments.

Linear programming scripts are widely utilized for data analysis, and R-shiny provides an efficient platform for implementing these scripts in applications. This approach not only benefits scientists who are not proficient in programming, but it is also convenient for programmers. The developed user interface offers a visually appealing design, and the interactive menus enhance the user experience by eliminating the need to search for specific lines of code on a black screen. Modifying parameters of functions becomes effortless, and data can be visualized without rerunning the entire script. R-shiny executes functions on demand, which means that graphics, variables, and data frames are not recalculated if their dependencies remain unchanged. This feature makes the application both cost-effective and time-efficient.

The analysis pathway is visualized through multiple components, providing users with a comprehensive view of the data. The overview plot, generated using the Q-feature object-oriented programming in the scp R library \citep{Vanderaa2021}, serves as a visual representation of the analysis process. This plot allows users to track the progression of the analysis and identify any missing files or samples. MS2 reporter ion intensities can be visualized as medians in form of a boxplot, allowing users to assess the distribution of intensities across different factors such as sample type, batch, or channel. This visualization helps identify potential biases or variations in the dataset.The feature-wise output further enhances the analysis pathway by enabling users to select individual proteins and explore their associated peptide spectrum, peptide and protein values over the course of the analysis. This feature provides a detailed view of the analysis results, allowing users to detect any missing or erroneous data points, particularly in channels with low or high abundance. By examining these individual protein profiles, users can evaluate the quality of the wet lab work and identify any possible errors or inconsistencies.
The statistics module within the application provides users with a user-friendly interface for hypothesis testing, even without prior expertise in statistics. By leveraging the powerful R-package limma \citep{Ritchie2015}, the module offers robust statistical analysis methods, empowering users to perform tests and obtain results to validate their hypotheses.One key feature of the statistics module is its ability to handle multiple factors and therefore correct for batch effects. Users can extend the statistical analysis across various factors, allowing them to quantify the impact of different variables and identify differentially expressed proteins. To enhance the user experience, the visualization of statistical results has been improved, resulting in visually appealing and interactive displays. These visualizations enable users to explore and interpret the results effortlessly, aiding in the comprehension of the statistical findings. By providing intuitive and engaging visual representations, the module facilitates the understanding and communication of complex statistical analyses.

The enrichment analysis module within the application addresses the challenge of selecting protein sets for mapping by incorporating a database-fetching functionality. This feature streamlines the enrichment analysis process and eliminates the need for users to manually select sets to map proteins against. The first variant of the enrichment analysis module utilizes the KEGG database, which offers a comprehensive collection of pathways and their interactions. By accessing this database, users can gain valuable insights into the underlying pathways and their relationships. This automated mapping against the KEGG database provides a convenient and efficient way to perform enrichment analysis. The second variant of the module caters to sophisticated users who may have specific protein sets in mind based on their research question. This flexibility allows users to select and incorporate their own protein sets into the analysis, enabling research in multiple contexts and providing a comprehensive insight into the dataset. Both variants of the enrichment analysis module present the results in interactive plots, offering users a dynamic and immersive experience with the data. Users can explore the enriched pathways and interact with the visualizations, gaining a deeper understanding of the relationships and connections within the dataset. Additionally, if desired, users can generate static plots and download them for use in presentations or other offline contexts.

One future research question could utilize artificial intelligence by predicting cellular behaviour, cell types, or even diseases. Running the application on a server or high-performance cluster would be advantageous to address this question and employ simple machine learning tools such as decision trees, random forests, or support vector machines. The application of artificial intelligence on biological data could reduce the hands-on time by identifying unclear results caused by experimental design, unclean preparation, or incorrect computation.
Another aspect, not limited to research topics, would be to focus on enhancing usability. While a development time of 6 months for a pipeline and user interface is sufficient to meet the analytical needs, there is still room for improvement in terms of user-friendliness. It is crucial to make the application more intuitive and robust to prevent crashes or errors caused by user mistakes. Ensuring a user-friendly experience is particularly important for commercial software, as robustness and ease of use are essential factors for success. However, since all software should be open source and developers should have the freedom to satisfy their curiosity, we can overlook this fact.

### 2023-06-13 markdown/introduction.Rmd
discussion and conclusion started

@@
 Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the  principles of cell type behavior, such as differentiation.
-The immune system, for instance, relies on specialized cells for specific functions, and the acquisition of these functions occurs through a process called differentiation.
-Studying differentiation in multiple cells can be challenging due to the inherent heterogeneity during the initiation of differentiation.
-By analyzing one cell at a time, stimulating and observing it throughout the differentiation process, researchers can capture the entire process at a single-cell level.
-However, sample preparation and computational analysis in SCP can be tedious, and statistical testing can be challenging due to low cell counts.
+Differentiation, a critical process in various biological systems such as the immune system, involves the acquisition of specialized functions by specific cell types.
+Investigating differentiation in multiple cells can be challenging due to the inherent heterogeneity during the initiation of this process.
+However, by analyzing one cell at a time and observing it throughout the differentiation process, researchers can capture the complete trajectory of differentiation at a single-cell level.
+This approach provides valuable insights into the dynamic changes occurring within cells during differentiation.
 In 2021, Specht et al.
-developed the SCoPE2 sample preparation and analysis pipeline, focusing on macrophage-monocyte differentiation experiments \citep{Specht2021}.
-This pipeline, which reduced labor time, served as a reference model for subsequent projects.
-SCP provides insights qualitatively and quantitatively, enabling the study of drug responses, infectious diseases, and organism development.
-Over the past decade, single cell techniques became indispensable tools for global research communities \citep{Minakshi2019}.
+developed the SCoPE2 sample preparation and analysis pipeline, performing macrophage-monocyte differentiation experiments \citep{Specht2021}.
+This pipeline streamlined the sample preparation process, reducing labor time and providing a reference model for subsequent SCP projects.
+SCP has proven to be a powerful tool for studying a wide range of biological phenomena, including drug responses, infectious diseases, and organism development.
+Over the past decade, single-cell techniques, including SCP, have become indispensable tools for the global research community \citep{Minakshi2019}.
 
@@
 
+### Data processing
 Given the high resolution of MS data, algorithms are employed to convert the raw signal into an interpretable form.
-Software packages like MaxQuant are commonly used to process the data, providing it for further analysis and statistical testing.
-Other software solutions include Protein Discoverer by Thermo Fisher and various R packages.
+Software packages like MaxQuant \citep{Cox2008} are commonly used to process the data, providing it for further analysis and statistical testing.
+Other software solutions include Protein Discoverer by Thermo Fisher.
 
@@
 However, only peptide fragments are visible in the spectrum due to peptide fragmentation.
-To identify proteins, peptides are matched against a sequence database \citep{Cox2008}.
+To identify proteins, peptides are matched against a sequence database.
 Sequence databases are typically stored as .fasta files and can be downloaded from the UniProt website (www.uniprot.org).
@@
 The FDR calculation provides a measure of the rate at which false identifications are expected in the dataset.
-
-After these calculations, the identified peptides can be grouped and aggregated according to their respective proteins for quantification.
-Various metrics that indicate the performance of the peptide search can be utilized in downstream analysis.
 Within the identified peptide groups, a "razor protein" refers to the protein with the highest number of ambiguously identified peptides.
@@
 Posterior error probabilities, which represent the likelihood of the identified peptide being a random event, are multiplied together, and only distinct sequences with the highest scores are considered for further analysis.
+Various metrics that indicate the performance of the peptide search can be utilized in downstream analysis.
+
+### Downstream analysis
+With advancements in mass spectrometry technology and computational methods, proteomic analysis has emerged as a powerful tool for investigating complex protein samples.
+However, analyzing proteomic data poses challenges in data processing, statistical analysis, and interpretation.
+This thesis aims to address these challenges by exploring computational methods for downstream analysis of proteomic data.
+The methods employed in this study will be thoroughly explained in the methods section, with a specific focus on developing an interactive environment equipped with a graphical user interface.
+This user-friendly interface will enable researchers, including those with limited computational expertise, to navigate and comprehend the data effectively.
+
+The analysis pipeline comprises a series of steps to streamline the data.
+It begins with quality control measures at the peptide spectrum match level, involving the removal of contaminants and irrelevant channels.
+Significance levels of peptide spectrum matches serve as confidence thresholds for match selection.
+The data is then aggregated at the peptide level, followed by additional filtering procedures based on metrics such as reporter ion intensity, peptide covariance to razor proteins, and missing rates for peptides.
+
+To derive meaningful biological insights from the dataset, statistical testing plays a crucial role.
+Selecting appropriate statistical tests and ensuring proper data transformation and normalization are essential steps.
+Many statistical methods rely on data distribution and require careful consideration to preserve the biological interpretation of the results.
+This thesis employs various transformation and normalization techniques, aiming to maintain data integrity and avoid biased interpretations.
+Additionally, an entropy-based visualization approach will provide users with a valuable tool for validating the computations and selected thresholds for the filtering steps.
+
+Handling missing values poses further challenges in the pipeline development since a missing signal for a peptide results in a missing protein, which is highly unlikely in a biological context.
+Several options for missing value handling will be included and thoroughly explained in the methods section.
+
+Dimensionality reduction and visualization techniques are widely used to cluster data.
+In this project, multiple approaches will be utilized, and users will have the ability to visually and interactively observe the data.
+
+It is important to acknowledge that downstream analysis in proteomics lacks a standardized approach, and the selection of computational methods depends on the specific dataset and research objectives.
+Furthermore, statistical analysis and the interpretation of biological meaning often present misconceptions that need to be addressed.
+For instance, batch correction, a commonly practiced method, has been overestimated in numerous studies, leading to misleading results \citep{Nygaard2016}.
+This thesis aims to contribute to the field by providing researchers with a comprehensive understanding of computational methods in proteomics analysis and their application in an interactive environment.

### 2023-06-13 markdown/methods.Rmd
discussion and conclusion started

@@
 
-## Materials
+## Wetlab - Materials and Methods
+For testing and development of the pipeline the following materials and methods were performed. For the sample preparation protocol see appendix.
 
-For analysis two types of cells were used. One type is the Jurkat-based cell line (J-lat) with integrated HIV.
-
-The other type of cells are macrophages with a sample size of 72 cells. The analysis is done with two groups. A HIV negative (HIV-) control group and a HIV positive (HIV+) group.
-
-### Cell Isolation
-
-### Lysis
+### Celltypes
+Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome. Cell sorting was performed with FACSAria Fusion (BD Biosciences).
 
 ### Digestion
+Protein digestion was performed using trypsin (Promega, Madison, WA) buffered in Triethylammonium bicarbonate buffer (TEAB).
+Reagents were handled with MANTIS automatic dispenser (Formulatrix, Bedford, MA).
 
-#### Metabolic labeling
+### Labeling
+Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.
 
-#### Isobaric labeling
+### Pooling
+Carrier proteome was added to the 36 TMTpro sets before pooling sample types for each of the 36 MS-runs.
 
-##### Tandem mass tag (TMT)
-
-### Instrumentation
-
-#### Liquid chromatography
-
+### Liquid chromatography
 In order to separate peptides according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
@@
 
-#### Mass Spectrometry
+### Mass Spectrometry
 
-##### Ionization
+### Ionization
 For the reference data ESI ionization was used.
 
-##### MS/MS
+### Mass spectrometer
 For the reference data used for developing the pipeline an Orbitrap™ Fusion™ Lumos™ tribrid mass spectrometer (ThermoFisher Scientific) was used.
@@
 
-## Data
+## Computational methods
 ```{r Dataflow, eval=FALSE, include=FALSE}
@@
 ### Data-acquisition and raw-file analysis
-For the data-acquisition thermo fisher acquisition software was used.
-Raw-file analysis of the data was done with MaxQuant \citep{Cox2008} software package.
-
-#### MS-Spectrum
- In this publication we will mainly focus on the data-acquisition with MaxQuant \citep{Cox2008}.
-
-#### Three-dimensional peak detection
-
-#### Deisotoping
-
-#### Label detection
-
-#### Improving peptide mass accuracy
-
-#### Peptide search
-
-#### Protein assembly
+For the data-acquisition Thermo Fisher acquisition software was used.
+Raw-file analysis of the data was done with  MaxQuantCmd 2.2.0.0 \citep{Cox2008}.
+The default settings were used. TMTpro labels were corrected according to the isotopical distribution of the Lot-number provided by Thermo Fisher.
 
 ### Data processing
-Further analysis is done with R and respective packages such as bioconductor.
-Since there is no state of the art established, analysis varies upon experimental design.
-The workflow of the analysis will be processed in a so called pipeline streamlining the data through steps where individual results can be observed in visualizations and indivdual calculations will be adapted according to user demands and experimental properties.
+Further analysis is done with R and respective packages such as bioconductor. Please find version info in the appendix.
 
@@
 
-In order to streamline the analysis of multiple experiments, object oriented programming can be applied.
+In order to streamline the analysis of multiple experiments, object oriented programming was applied.
 The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
@@
 #### Exclude reverse matches/contaminants
-Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches can be excluded from further analysis.
+Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches were be excluded from further analysis.
 
@@
 The resulting product ions are derived from precursor ions (also known as mother ions or parental ions).
-Contaminant peptides can co-migrate in this process and can be distinguished by the lower fraction of their respective precursor ions \citep{Tannous2020}.
+Contaminant peptides can co-migrate in this process and were distinguished by the lower fraction of their respective precursor ions \citep{Tannous2020}.
 These peptides need to be filtered out during the analysis pipeline.

### 2023-06-14 scope2.R
single cell data set handling implemented

@@ -25,9 +25,30 @@ library("sva")
 library("tibble")
 
 # read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/evidence.txt")
+mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/evidence.txt")
 
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Apr12/combined/txt/sampleAnnotation_tabdel.txt")
+sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/sampleAnnotation.txt")
+
+
+
+ 
+# # create sample annotation
+# Raw.files <- unique(mqScpData$Raw.file)
+# 
+# library(stringr)
+# Raw.files <- Raw.files[str_order(str_extract_all(Raw.files, "\\d+"), numeric = TRUE)]
+# sampleAnnotation <- data.frame(Raw.file=rep(Raw.files, each=12))
+# 
+# channels <- colnames(mqScpData[ , grepl( "Reporter.intensity.corrected" , names(mqScpData) ) ])
+# sampleAnnotation$Channel <- rep(channels, 12)
+# 
+# sampletypes <- c(rep("HC", 4), rep("HIV_MetS", 4), rep("HIV_noMetS", 4))
+# sampleAnnotation$SampleType <- rep(sampletypes, 12)
+# 
+# outfile <- "/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/sampleAnnotation.txt"
+# 
+# library("readr")
+# write_delim(sampleAnnotation, outfile, delim = "\t")
 
 
 # create QFeature object
@@ -630,16 +651,17 @@ plotReducedDim(scp[["proteins_dim_red"]],
                point_alpha = 1)
 
 
+channelstring <- gsub("[0-9]{1,2}$","", scp$Channel[1]) 
 
 
 ## Get the features
-subsetByFeature(scp, "Q92954") %>%
+subsetByFeature(scp, "E9PAV3") %>%
   ## Format the `QFeatures` to a long format table
   longFormat(colvars = c("Raw.file", "SampleType", "Channel")) %>%
   data.frame %>%
   ## This is used to preserve ordering of the samples and assays in ggplot2
   mutate(assay = factor(assay, levels = names(scp)),
-         Channel = sub("Reporter.intensity.", "", Channel),
[…]

### 2023-06-14 scpR-app/app.R

@@ -407,6 +407,7 @@ server <- function(input, output, session) {
   meta_data <- eventReactive(input$update_button, {
     req(input$sample_annotation_file)
     meta_data_0 <- read.delim(input$sample_annotation_file$datapath)
+
     if (input$file_level == FALSE) {
       meta_data_0 <- meta_data_0[!(meta_data_0$SampleType %in%  input$selectedSampleType_to_exclude), ]
     } else {
@@ -1507,6 +1508,9 @@ server <- function(input, output, session) {
   output$feature_subset <- renderPlot({
     scp_0 <- scp()
     
+    
+    channelstring <- gsub("[0-9]{1,2}$","", scp_0$Channel[1]) 
+    
     if (input$file_level == FALSE) {
       subsetByFeature(scp_0, input$selectedProtein) %>%
         ## Format the `QFeatures` to a long format table
@@ -1514,7 +1518,7 @@ server <- function(input, output, session) {
         data.frame %>%
         ## This is used to preserve ordering of the samples and assays in ggplot2
         mutate(assay = factor(assay, levels = names(scp_0)),
-               Channel = sub("Reporter.intensity.", "", Channel)) %>%
+               Channel = sub(channelstring, "", Channel)) %>%
         mutate(Channel = as.numeric(Channel)) %>%
         arrange(Channel) %>%
         mutate(Channel = factor(Channel, levels = unique(Channel))) %>%

### 2023-06-14 markdown/results.Rmd
results started

@@
 
-Some more guidlines from the School of Geosciences.
+After runring MaxQuant mode with MS2 search for TMTpro reporter ions with a runtime of 6 hours, the data was analyzed using the developed shiny application. 
+Using the application starts with uploading the tabdelimited evidence and sample annotation file. The application is ready to run with default values. Fine tuning will be done subsequently.
 
-This section should summarise the findings of the research referring to all figures, tables and statistical results (some of which may be placed in appendices).
-- include the primary results, ordered logically - it is often useful to follow the same order as presented in the methods.
-- alternatively, you may find that ordering the results from the most important to the least important works better for your project.
-- data should only be presented in the main text once, either in tables or figures; if presented in figures, data can be tabulated in appendices and referred to at the appropriate point in the main text.
+```{r ui_summary, echo=F, out.width="100%", fig.align='center', fig.cap="Summary showing the process of the analysis pipeline"}
+knitr::include_graphics("screenshots/summary.png")
+```
 
-**Often, it is recommended that you write the results section first, so that you can write the methods that are appropriate to describe the results presented.
-Then you can write the discussion next, then the introduction which includes the relevant literature for the scientific story that you are telling and finally the conclusions and abstract – this approach is called writing backwards.**
+On the left side of the user interface the analysis pipeline settings ca be changed.
+With the update button the user can apply the changes made and observe alterations in the dataset directly.
+The summary visualization looks cluttered since 36 files were processed, in smaller datasets the nodes of the network graph can be distuingished clearly.
+
+By running the application with SCoPE2 \citep{Specht2021} reference cutoffs the following results were visualized.
+```{r ui_numbers_default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
+knitr::include_graphics("screenshots/numbers_default.png")
+```
+
+The default settings yield a total number of unique proteins below $\approx 10^{3}$ aggregated from $10^{4.5} \approx$ peptide spectrum matches also aggregated over $10^{3.5}\approx$ peptides.
+The aggregation is set fixed by median.
+
+```{r ui_mutual_info_default, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with default setting"}
+knitr::include_graphics("screenshots/mutual_info_default.png")
+```
+
+The mutual information (MI) within the sample type of the healthy control group shows a reduction to the lower edge of the IQR after the transformation and normalization.
+When comparing the MI with the sample type HIV with metabolic syndrome, the reduction is within the inter quantile range (IQR = 25 - 75 quantile) of the raw data MI.
+
+```{r ui_RI_channel_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on channel.
+The selected factor will be highlighted and indicated in the legend."}
+knitr::include_graphics("screenshots/RI_channel_default.png")
+```
+
+When comparing the median reporter ion intensities over channels show no systematic bias.
+Considering the 3 sample types occupy in every run the same channels, it is important to look at this visualization.
+Since the IQR overlap over the majority of samples, the change in expression with the factor channel can be considered as non-significant.
+However the channel will be later on tested for significance in the dimensionality redution plot and statistic module.
+
+```{r ui_RI_file_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on file"}
+knitr::include_graphics("screenshots/RI_file_default.png")
+```
+
+Comparing median reporter ion intensities with the focus on the batch (also Raw.file), it is obvious that the changes in signal are significant and need to be considered in further analysis.
+For the dimensionality reduction the user can choose in the sidepane the batch correction option ComBat \citep{Leek2012}.
+However this option is not chosen as default, since the conservative method includes the batch factor in the statistic module.
+
+```{r ui_RI_st_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on sampletype"}
+knitr::include_graphics("screenshots/RI_st_default.png")
+```
+
+Observing the Median RIs with a focus on sample types, IQRs partly overlap so the binding efficiency of the TMTpro reagent is not affected by the sample type.
+
+```{r ui_cv_st_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on sampletype"}
+knitr::include_graphics("screenshots/cv_st_default.png")
+```
+
+```{r ui_mutual_info_finetune, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with adjusted settings"}
+knitr::include_graphics("screenshots/mutual_info_finetune.png")
+```

### 2023-06-14 scpR-app/app.R

@@ -34,7 +34,9 @@ reactlog::reactlog_enable()
 ui <- fluidPage(
   # useShinyjs(),
   # Application title
-  titlePanel("Proteomics Workbench"),
+  titlePanel(fluidRow(width=15,
+                      column(h1("Proteomics Workbench"), h5("by Lukas Gamp"), offset = 0.5, width=5))
+    ),
   
   # Sidebar
   sidebarLayout(
@@ -83,9 +85,9 @@ ui <- fluidPage(
       tabsetPanel(id="outbut_tabset", type = "tabs",
                   tabPanel("Overview", id="overview_pane",
                            verbatimTextOutput("debug"),
-                           plotOutput("overview_plot")),
+                           plotOutput("overview_plot", width = "1500px", height = "1000px")),
                   tabPanel("Summary Barplot", id="summary_pane",
-                           plotOutput("summary_bar", width = "auto")),
+                           plotOutput("summary_bar", width = "1300px", height = "1000px")),
                   tabPanel("Normalization validation", id="normval_pane",
                            selectInput("selectedSampleType_normval", "Select SampleType to validate", ""),
                            plotOutput("norm_val_plot_intra_group"),
@@ -94,12 +96,24 @@ ui <- fluidPage(
                   ),
                   tabPanel("Reporter Ion Intensity", id="ri_pane",
                            selectInput("color_variable_ri", "select variable to indicate", ""),
-                           plotOutput("RI_intensity")),
+                           plotOutput("RI_intensity", width = "1300px", height = "800px")),
                   tabPanel("Covariance and correlation", id="cov_cor_pane",
-                           conditionalPanel(condition = "!input.file_level", id="razor_prot_pane",
-                                            selectInput("color_variable_cv", "select variable to indicate", ""),
-                                            plotOutput("CV_median")),
-                           plotOutput("corr_matrix")),
+                           fluidRow(width=10,
+                                    conditionalPanel(condition = "!input.file_level", id="razor_prot_pane",
+                                                     column(width = 5, conditionalPanel(condition = "input.cv_plot_switch",
+                                                                                        selectInput("color_variable_cv", "select variable to indicate", ""))
+                                                                             ),
+
+                                                     column(width = 5, switchInput("cv_plot_switch", label="switch between visualizations", onLabel="Cov boxlot", offLabel="Cor heatmap", onStatus="primary", offStatus = "info"))
+                                                     )
+                                    ),
[…]

### 2023-06-15 markdown/Thesis.Rmd
literature recherche
refactor exp_matrix_0 <- exp_matrix() in Statistic module --> done
thesis: flowchart in methods started not knitting correctly --> fix to do
discussion and conclusion started
results until stats finished

@@
   - \usepackage[nottoc, numbib]{tocbibind}
+
 bibliography: bibliography.bib
@@
 library(tinytex)
+
+#install dependencies
+if (!requireNamespace("DiagrammeR", quietly=TRUE))
+  install.packages("DiagrammeR")
+
+library(DiagrammeR)
 ```
@@
 
-in partial fulfilment of the requirement \\for the degree of MSc with Honours \\in Bioinformatics
+in partial fullfilment of the requirement \\for the degree of MSc \\in Bioinformatics
 
@@
 ```
+\pagenumbering{arabic}
 
@@
 ```
-\pagenumbering{arabic}
 

### 2023-06-15 markdown/discussion.Rmd
results until stats finished

@@
 
+## Intro
 The developed application and the proteomics experiment pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
@@
 
+## Mutual Information
 Information theory and the concept of mutual information provide a framework for understanding the conservation of entropy in a dataset.
@@
 
+## Interactive pipeline
 Linear programming scripts are widely utilized for data analysis, and R-shiny provides an efficient platform for implementing these scripts in applications.
@@
 This feature makes the application both cost-effective and time-efficient.
-
 The analysis pathway is visualized through multiple components, providing users with a comprehensive view of the data.
@@
 This plot allows users to track the progression of the analysis and identify any missing files or samples.
+
+## Reporter ion intensity
 MS2 reporter ion intensities can be visualized as medians in form of a boxplot, allowing users to assess the distribution of intensities across different factors such as sample type, batch, or channel.
-This visualization helps identify potential biases or variations in the dataset.The feature-wise output further enhances the analysis pathway by enabling users to select individual proteins and explore their associated peptide spectrum, peptide and protein values over the course of the analysis.
+This visualization helps identify potential biases or variations in the dataset.
+
+## median coefficient of variation
+The reference values of the SCoPE2 publication \citep{Specht2021} could be tested and confirmed.
+5 as the number of observation for leading razor peptides was suitable to identify the cutoff of 0.65 by observing whiskers of the boxplots in order to reduce noisy peptides in the dataset.
+
+## Feature wise plot
+The feature-wise output further enhances the analysis pathway by enabling users to select individual proteins and explore their associated peptide spectrum, peptide and protein values over the course of the analysis.
 This feature provides a detailed view of the analysis results, allowing users to detect any missing or erroneous data points, particularly in channels with low or high abundance.
 By examining these individual protein profiles, users can evaluate the quality of the wet lab work and identify any possible errors or inconsistencies.
+
+## Statistics module
 The statistics module within the application provides users with a user-friendly interface for hypothesis testing, even without prior expertise in statistics.
@@
 
+All-Round Manipulation of the Actin Cytoskeleton by HIV \citep{Turville2018} example for the most differentially expressed gene. --> elaborate biological meaning.
+P68133 (Actin, alpha skeletal muscle)
+
+SAMHD1-Deficient CD14+ Cells from Individuals with Aicardi-Goutières Syndrome Are Highly Susceptible to HIV-1 Infection \citep{Berger2011} example for the most differentially expressed gene.
+--> elaborate biological meaning.
+P31153 (S-adenosylmethionine synthase isoform type-2)
+
+## Enrichment 
 The enrichment analysis module within the application addresses the challenge of selecting protein sets for mapping by incorporating a database-fetching functionality.
@@
 
+## Future perspectives
+
+Clustering for stratification of cells.
+
 One future research question could utilize artificial intelligence by predicting cellular behaviour, cell types, or even diseases.

### 2023-06-15 markdown/introduction.Rmd
discussion and conclusion started
results until stats finished

@@
 Significance levels of peptide spectrum matches serve as confidence thresholds for match selection.
-The data is then aggregated at the peptide level, followed by additional filtering procedures based on metrics such as reporter ion intensity, peptide covariance to razor proteins, and missing rates for peptides.
+The data is then aggregated at the peptide level, followed by additional filtering procedures based on metrics such as reporter ion intensity, peptide coefficient of variation over razor proteins, and missing rates for peptides.
 

### 2023-06-15 markdown/results.Rmd
results until stats finished

--- 2023-02-02 markdown/results.Rmd
@@
 
-Some more guidlines from the School of Geosciences.
+To begin using the application, the first step is to upload the tab-delimited evidence file and the sample annotation file.
+Once these files are uploaded, the application is ready to run with default values.
+The default settings are preconfigured for an initial analysis, and any further adjustments or fine-tuning can be made in subsequent steps of the analysis process.
 
-This section should summarise the findings of the research referring to all figures, tables and statistical results (some of which may be placed in appendices).
-- include the primary results, ordered logically - it is often useful to follow the same order as presented in the methods.
-- alternatively, you may find that ordering the results from the most important to the least important works better for your project.
-- data should only be presented in the main text once, either in tables or figures; if presented in figures, data can be tabulated in appendices and referred to at the appropriate point in the main text.
+```{r ui_summary, echo=F, out.width="100%", fig.align='center', fig.cap="Summary showing the process of the analysis pipeline"}
+knitr::include_graphics("screenshots/overview.png")
+```
 
-**Often, it is recommended that you write the results section first, so that you can write the methods that are appropriate to describe the results presented.
-Then you can write the discussion next, then the introduction which includes the relevant literature for the scientific story that you are telling and finally the conclusions and abstract – this approach is called writing backwards.**
+On the left side of the user interface, users can find the analysis pipeline settings that can be adjusted.
+Once the desired changes are made, they can be applied by clicking the update button.
+This allows users to observe the direct impact of the applied changes on the dataset.
+In the summary visualization, it may appear cluttered due to the processing of 36 files. In smaller datasets, the nodes of the network graph can be more easily distinguished and understood.
+When running the application with SCoPE2 \citep{Specht2021} reference cutoffs, the following results were visualized.
+
+\newpage
+```{r ui_numbers_default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
+knitr::include_graphics("screenshots/numbers_default.png")
+```
+
+The default settings of the analysis pipeline resulted in a total number of unique proteins below approximately $10^3$, which were aggregated from around $10^{3.5}$ peptides.
+Before these peptides were aggregated over approximately $10^{4.5}$ peptide spectrum matches.
+The aggregation method used is fixed by median.
+
+\newpage
+```{r ui_mutual_info_default, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with default setting"}
+knitr::include_graphics("screenshots/mutual_info_default.png")
+```
+
+After applying logarithmic transformation to the base 2 and performing column median row mean normalization, the mutual information (MI) within the healthy control (HC) group's sample type exhibits a reduction towards the lower edge of the interquartile range (IQR).
+
+When comparing the MI between the HC group and the sample type corresponding to individuals with both HIV and metabolic syndrome, the IQR is further reduced and shifted towards the lower end compared to the MI values observed in the raw data.
+
+\newpage
+```{r ui_mutual_info_quant_norm, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with quantile normalization"}
+knitr::include_graphics("screenshots/mutual_info_quant_norm.png")
+```
+
+After performing a logarithmic transformation to the base 2 and quantile normalization, the mutual information (MI) within the healthy control (HC) group's sample type exhibits a minor reduction in the interquartile range (IQR).
+However, it should be noted that there is an increase in the number of outliers compared to the raw data.
+
+When comparing the MI IQR of the HC group with the sample type corresponding to individuals with both HIV and metabolic syndrome, the reduction in MI falls within the interquartile range of the raw data MI.
+Additionally, the median remains close to unaffected, indicating that the overall central tendency of the MI values is relatively preserved.
+
+\newpage
+```{r ui_RI_channel_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on channel.
+The selected factor will be highlighted and indicated in the legend."}
+knitr::include_graphics("screenshots/RI_channel_default.png")
+```
+
+When comparing the median reporter ion intensities across channels, no systematic bias is observed.
+However, it is worth noting that channel 7 exhibits a wide range of values spanning an order of magnitude.
+This comparison is crucial since the three sample types consistently occupy the same channels in every run, making the visualization an important aspect of the analysis.
+Despite the wide range of values in channel 7, there is still overlap in the interquartile ranges (IQR) across the majority of samples.
+
+\newpage
+```{r ui_RI_file_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on file"}
+knitr::include_graphics("screenshots/RI_file_default.png")
+```
+
+When comparing the median reporter ion intensities with a focus on the batch (=Raw.file), it becomes evident that there are significant changes in signal that need to be taken into account in further analysis.
+After applying quality control measures, several values have been lost due to cutoffs in parameters such as parental ion fraction (PIF), q-value, coefficient of variation, and missingness of peptide data.
+
+For dimensionality reduction, the user has the option to choose the batch correction method ComBat \citep{Leek2012} in the side panel.
+However, this option is not set as the default choice.
+Instead, the conservative approach is followed, where the batch factor is included in the statistical module to account for any potential batch effects that may influence the results.
+
+\newpage
+```{r ui_RI_st_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on sampletype"}
+knitr::include_graphics("screenshots/RI_st_default.png")
+```
+
+The sample type does not appear to have an impact on the median reporter ion intensities, as suggested by the overlapping interquartile ranges (IQRs).
+Nevertheless, the HIV group with metabolic syndrome displays a broader range of values spanning tenfold in arbitrary units.
+
+\newpage
+```{r ui_cv_st_default, echo=F, out.width="100%", fig.align='center', fig.cap="Median covariance of razor proteins with selection on sampletype"}
+knitr::include_graphics("screenshots/cv_st_default.png")
+```
+
+A cutoff value of 0.65 for the coefficient of variation (CV) of razor proteins as shown in the SCoPE2 publication \citep{Specht2021} demonstrates a satisfactory fit to the data.
+The minimum number of observations determines the threshold for considering a razor protein ambiguous, based on the number of matching peptides.
+In this case, the reference publication set the threshold at 5 observations.
+
+The interquartile ranges (IQRs) remained unchanged by this cutoff.
+The sample type did not influence the CV of razor proteins.
+Boxplots illustrate that the values are distributed similarly, as indicated by the overlapping IQRs.
+
+\newpage
+```{r ui_cv_rawfile_default, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file"}
+knitr::include_graphics("screenshots/cv_rawfile_default.png")
+```
+
+The plot illustrates that certain batches (=raw files) exhibit a higher number of distinct matches compared to others, as indicated by a thin line depicting low density.
+Most interquartile ranges (IQRs) overlap, suggesting similarity among the majority of batches.
+However, specific batches display a lower IQR in median CV across razor protein peptides, indicating distinct matches.
+
+\newpage
+```{r ui_cv_rawfile_nobs2, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins calculated for 2  observations per cell, with selection on Raw.file"}
+knitr::include_graphics("screenshots/cv_rawfile_nobs2.png")
+```
+By setting the number of observations to 2, the distributions become more equalized.
+Additionally, batches that initially exhibited a decreased coefficient of variation (CV) tend to inflate towards the threshold.
+
+\newpage
+```{r ui_cv_rawfile_nobs10, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins calculated for 10 observations per cell, with selection on Raw.file"}
+knitr::include_graphics("screenshots/cv_rawfile_nobs10.png")
+```
+
+When the minimum number of observations for ambiguous peptides is set to 10, the distribution of medians across channels shows a decrease in density.
+This can be attributed to the rare occurrence of 10 unique peptides matching to the same protein, which is not commonly observed in the dataset.
+
+\newpage
+```{r ui_heatmap, echo=F, out.width="100%", fig.align='center', fig.cap="Correlation heatmap"}
+knitr::include_graphics("screenshots/heatmap.png")
+```
+The heatmap reveals the Pearson pairwise correlation coefficients between all proteins.
+The dendrogram indicates that there are two main protein clusters, with one of the clusters showing further sub-clustering, indicating a strong relationship among the proteins in the sub groups.
+However, the other cluster, consisting of individual proteins, shows weaker correlations, leading to a more intricate branching pattern in the dendrogram.
+
+\newpage
+```{r ui_dim_red_default, echo=F, out.width="100%", fig.align='center', fig.cap="Dimensionality reduction analysis depicting sample types (indicated by shape) and raw files (indicated by color).
+Top: principle component analysis (PCA).
+Lower: Uniform Manifold Approximation and Projection (UMAP)"}
+knitr::include_graphics("screenshots/dim_red_default.png")
+```
+
+In both the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP) plot, the second component is plotted against the first component.
+Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch.
+For the k-nearest neighbor descent algorithm (NN-descent) employed in the UMAP, first a default value of k=3 was selected.
+
+\newpage
+```{r ui_dim_red_changed_k, echo=F, out.width="100%", fig.align='center', fig.cap="Dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+Top: Principal Component Analysis (PCA).
+Bottom: Uniform Manifold Approximation and Projection (UMAP).
+For the UMAP visualization, the k value for the NN-descent algorithm was modified to 15."}
+knitr::include_graphics("screenshots/dim_red_changed_k.png")
+```
+
+The second figure of the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP), both displaying the second component against the first.
+Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch information.
+To enhance the resolution and capture more intricate patterns, the k-nearest neighbor descent algorithm (NN-descent) was adjusted, with a value of K set to 15.
+While the resulting plot exhibits improved detail, the underlying biological context or background remains elusive and does not emerge in the visual representation.
+
+\newpage
+```{r ui_dim_red_c3d, echo=F, out.width="100%", fig.align='center', fig.cap="3D visualization of dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+Top: Principal Component Analysis (PCA).
+Bottom: Uniform Manifold Approximation and Projection (UMAP)"}
+knitr::include_graphics("screenshots/dim_red_3d.png")
+```
+
+In the 3D visualization, a subtle clustering pattern can be observed for the first two batches.
+Upon switching to the indicator sample type in the 3D visualization, no clustering can be observed whatsoever.
+This suggests that the variance in the dataset is not attributed to sample type or batch, reinforcing the notion that other factors could contribute to the observed variability.
+
+\newpage
+```{r ui_dim_red_batchC, echo=F, out.width="100%", fig.align='center', fig.cap="Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+upper top: Principal Component Analysis (PCA).
+upper bottom: Uniform Manifold Approximation and Projection (UMAP).
+lower top: 3D Principal Component Analysis (PCA).
+lower bottom: 3D Uniform Manifold Approximation and Projection (UMAP)"}
+knitr::include_graphics("screenshots/dim_red_batchC.png")
+knitr::include_graphics("screenshots/dim_red_batchC_3d.png")
+```
+
+After applying batch correction using ComBat \citep{Leek2012}, neither the 2D nor the 3D visualization demonstrates any noticeable effects related to batches or sample types.
+The previously observed subtle clustering of the first two batches is now eliminated, indicating successful batch correction.
+However, it's important to note that the default pipeline was executed without batch correction for correct statistical testing.
+
+\newpage
+```{r ui_feature, echo=F, out.width="100%", fig.align='center', fig.cap="Feature wise output for the protein P61313 (=60S ribosomal protein L15) over the course of the pipeline."}
+knitr::include_graphics("screenshots/feature.png")
+```
+
+The scatter plots indicate the detected values associated with the protein UniProt ID P61313 (= 60S ribosomal protein L15).
+Starting from the peptide spectrum level and progressing to peptides aggregated for individual raw files, these plots offer a comprehensive view of the protein's peptide abundance across various batches.
+After merging the batches, the peptide appears to be abundant in all channels.
+Consequently, the need for missing value imputation with K-nearest neighbors (KNN) in this dataset becomes less likely, as missing values for the protein are uncommon once the batches are combined on the protein level.
+
+\newpage
+```{r ui_stat_HC_HIV_noMets, echo=F, out.width="100%", fig.align='center', fig.cap="Statistic module indicating the comparison of healthy control (HC) versus HIV without metabolic syndrom (HIV_noMetS).
+Venndiagram indicates significantly up/downregulated proteins.
+Volcano represents log10(adjusted p values) against log10 fold changes (logFC).
+Table shows logFC, average expression, t value, raw p.value, adjusted p value and B statistic.
+P.values were adjusted with false discovery rate (FDR)"}
+knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
+```
+
+By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group, six differentially expressed proteins were identified.
+These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
+A total of 652 proteins have been identified as not significantly upregulated or downregulated in the comparison analysis.
+These proteins did not exhibit statistically significant differences in expression between the compared groups or conditions.
+
+The direction of the contrast model determines the change in log-fold change (logFC) values.
+For instance, in the depicted figure, P68133 (Actin, alpha skeletal muscle) shows an increased expression of 0.25 (25%) in HIV_noMetS macrophages compared to HC macrophages.
+It's important to note that the logFC value changes by multiplication with -1 when the direction of the contrast model is reversed.
+
+To account for potential batch effects, the batch factor was included as a cofactor in the model, ensuring that any observed differential expression is not confounded by batch variation.

### 2023-06-15 scpR-app/app.R

@@ -60,8 +60,8 @@ ui <- fluidPage(
                        # minimal observation of a peptide within their corresponding razor protein
                        numericInput("nObs_pep_razrpr", "Input minimal observation of a peptide within their corresponding razor peptide", 5, min = 1, max=100, step = 1),
                        
-                       # max covariance accepted
-                       numericInput("MedCV_thresh", "Input maximal Covariance accepted", 0.65 , min = 0, max=1, step = 0.01),
+                       # max median coefficient of variation (CV) accepted
+                       numericInput("MedCV_thresh", "Input maximal CV accepted", 0.65 , min = 0, max=1, step = 0.01),
                        
                        # remove peptides with missing data
                        numericInput("pNA", "Input missingness percentage threshold for peptide data", 0.99 , min = 0, max=1, step = 0.01)
@@ -97,7 +97,7 @@ ui <- fluidPage(
                   tabPanel("Reporter Ion Intensity", id="ri_pane",
                            selectInput("color_variable_ri", "select variable to indicate", ""),
                            plotOutput("RI_intensity", width = "1300px", height = "800px")),
-                  tabPanel("Covariance and correlation", id="cov_cor_pane",
+                  tabPanel("median coefficient of variation and correlation", id="cov_cor_pane",
                            fluidRow(width=10,
                                     conditionalPanel(condition = "!input.file_level", id="razor_prot_pane",
                                                      column(width = 5, conditionalPanel(condition = "input.cv_plot_switch",
@@ -379,8 +379,8 @@ server <- function(input, output, session) {
              visualization in RI plot. <br>
              <br>
              <br>
-             5.) Covariance across razor protein calculation and filtering according to n-observations (nobs) <br>
-             Adapt nobs and accepted covariance according to your preference of analysis in order to filter out noisy quantification. Visualization can be found in the CV plot. <br>
+             5.) median coefficient of variation (CV) across razor protein calculation and filtering according to n-observations (nobs) <br>
+             Adapt nobs and accepted median coefficient of variation (CV) according to your preference of analysis in order to filter out noisy quantification. Visualization can be found in the CV plot. <br>
              <br>
              <br>
              6.) Normalization of peptide data <br>
@@ -457,8 +457,8 @@ server <- function(input, output, session) {
   
   # result of analysis pipleline
[…]

### 2023-06-16 markdown/introduction.Rmd
proofreading of intro[1] and methods[umap] done

@@
 ## From the Central Dogma of Molecular Biology to Omics and Systems Biology
-Before understanding the underlying principle biological systems undergo and the initial exploration of molecular biology, the processes of live were roughly inferred and not well understood.
+Before the initial exploration of molecular biology, the processes of life were roughly inferred and not well understood.
 The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins.
-This framework consists of two major processes, which involve transcription, converting DNA into RNA, and translation, converting RNA into functional proteins.
-In between those processes regulation takes base on a molecular level, by altering structure with chemical adaptations.
-The regulation is undertaken to provide biological organisms with an highly efficient output of their  genetic products called proteins.
+This framework consists of two major processes: transcription, which converts DNA into RNA, and translation, which converts RNA into functional proteins.
+Between these processes, molecular-level regulation occurs through chemical adaptations.
+This regulation is crucial for biological organisms to efficiently produce their genetic products, known as proteins.
 Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
-Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as omics, to comprehensively analyze biomolecules in diverse contexts.
-Omics research has emerged as an approach for comprehensively studying biomolecules and their interactions within biological systems.
-Omics encompasses various fields, including genomics, transcriptomics, proteomics, metabolomics, and others, each focusing on a specific class of biomolecules.
-By analyzing multiple omics layers, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks with computational methods.
+Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as "omics," to comprehensively analyze biomolecules in diverse contexts.
+Omics research has emerged as an approach to studying biomolecules and their interactions within biological systems comprehensively.
+Omics encompasses various fields, including genomics, transcriptomics, proteomics, and metabolomics, with each field focusing on a specific class of biomolecules.
+By analyzing multiple layers of omics data, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks using computational methods.
 Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks.
-Imagining biological systems as complex, multi-layer networks, the ensemble of proteins, known as the proteome, plays a critical role in this interactive structure with many functions such as structural integrity, catalyzing chemical reaction and regulation of cellular funcitons.
+Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role in this interactive structure.
+Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}
 
@@
 Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
-For instance pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
+For instance if pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
 These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications and more \citep{Beltran2017}.

### 2023-06-16 markdown/methods.Rmd
proofreading of intro[1] and methods[umap] done

@@
 Edges are distinguished from each other not only by their connection but also by their importance, which is often referred to as weight.
-Whereas the weight function is calculated as $w((x_i, x_{i_{j}})) = \exp(\frac{-max(0,d(x_{i},x_{i_{j}}-\rho_{i})}{\sigma_{i}}$ and determine the parameter $\rho$ for the weight function that: $\rho_i=\min \left\{ d(x_i,x_{i_{j}}) \,|\, 1 \leq j \leq k, d(x_i, x_{i_{j}}) > 0 \right\}$.
-Furthermore $\sigma$ will be set to meet the following conditions: $\sum_{j=1}^{k} \exp ( \frac{ -\max(0, d(x_{i}, x_{i_{j}}) - \rho_i)} { \sigma_i } ) = \log_2(k)$ and acts as a normalisation factor.
+Whereas the weight function is calculated as
+
+\[w((x_i, x_{i_{j}})) = \exp(\frac{-max(0,d(x_{i},x_{i_{j}}-\rho_{i})}{\sigma_{i}})\]
+
+and determine the parameter $\rho$ for the weight function that:
+
+\[\rho_i=\min \left\{ d(x_i,x_{i_{j}}) \,|\, 1 \leq j \leq k, d(x_i, x_{i_{j}}) > 0 \right\}\]
+
+Furthermore $\sigma$ will be set to meet the following conditions: 
+\[\sum_{j=1}^{k} \exp ( \frac{ -\max(0, d(x_{i}, x_{i_{j}}) - \rho_i)} { \sigma_i } ) = \log_2(k)\] 
+and acts as a normalization factor.
 The obtained graph $\overline{G}$ is called the fuzzy simplicial set which is obtained by approximating the geodesic distance (= locally length-minimizing curve) of the data points, leading to a topological approximation.
 The second phase of the algorithm adjusts the layout of the weighted graph to a representative view, but preserves the characteristics and shows the underlying topological principle of the data.
-In order to obtain a visual representation of the graph, the algorithm applies repulsive forces $\frac{2b}{(\epsilon + ||y_i-y_j||_{2}^{2})(1+a||y_i-y_j||_{2}^{2b})}(1-w((x_i,x_j)))(y_i-y_j)$ among vertices and attractive forces
-$\frac{ -2ab || y_i -y_j||_{2}^{2(b-1)} }{1+||y_i-y_j||_{2}^{2}} w((x_i,x_j))(y_i-y_j)$ along edges.
+In order to obtain a visual representation of the graph, the algorithm applies repulsive forces:
+\[\frac{2b}{(\epsilon + ||y_i-y_j||_{2}^{2})(1+a||y_i-y_j||_{2}^{2b})}(1-w((x_i,x_j)))(y_i-y_j)\]
+among vertices and attractive forces
+\[\frac{ -2ab || y_i -y_j||_{2}^{2(b-1)} }{1+||y_i-y_j||_{2}^{2}} w((x_i,x_j))(y_i-y_j)\] 
+along edges.
 Where $a$ and $b$ are denoted as hyperparameters.
@@
 #### Validation of the methods applied to the dataset
-In the context of the validation of methods applied to a dataset, the mutual information calculation is used to assess the impact of computations on the dataset and the information it carries. 
+For the validation of methods applied to the dataset, the mutual information calculation is used to assess the impact of computations and the information it carries. 
 Mutual information, denoted as $I(X;Y)$, is a measure of the dependence or information shared between two variables, $X$ and $Y$, in this case sample types.
@@
 To calculate the mutual information between two variables $X$ and $Y$ we can alternatively express the above equation as $I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})$.
-$D_{KL}$ denotes for the Kullback-Leibler divergence which is a concept of relative entropy between two variables and tells us how much 2 probability distributions differ from each other and $P_{X} \otimes P_{Y}$ represents the product of the marginals.
-The underlying principle of divergence is the integral asymmetry in Bayesian inference, which means that divergence does not satisfy the triangle inequality, and differences are not the same depending on the starting point.
+$D_{KL}$ denotes for the Kullback-Leibler \citep{Kullback1951} divergence which is a concept of relative entropy between two variables and tells us how much 2 probability distributions differ from each other.
+$P_{X} \otimes P_{Y}$ represents the product of the marginals.
+The underlying principle of divergence is the integral asymmetry in Bayesian inference, which means that divergence does not satisfy the triangle inequality.
+Differences are not the same depending on the starting point.
 By comparing the mutual information before performing any calculations with the mutual information calculated after the computations, it is possible to assess whether the applied methods have significantly altered the dependence or information content.

### 2023-06-16 markdown/results.Rmd
proofreading of intro[1] and methods[umap] done

@@
 \newpage
-```{r ui_stat_HC_HIV_noMets, echo=F, out.width="100%", fig.align='center', fig.cap="Statistic module indicating the comparison of healthy control (HC) versus HIV without metabolic syndrom (HIV_noMetS).
-Venndiagram indicates significantly up/downregulated proteins.
-Volcano represents log10(adjusted p values) against log10 fold changes (logFC).
-Table shows logFC, average expression, t value, raw p.value, adjusted p value and B statistic.
-P.values were adjusted with false discovery rate (FDR)"}
+```{r ui_stat_dependency, echo=F, out.width="100%", fig.align='center', fig.cap="Popup window for checking the dependency of the statistic module"}
+knitr::include_graphics("screenshots/stat_dependency.png")
+```
+
+The t-statistics employed in limma are known for their robustness in handling non-normally distributed values.
+Nonetheless, it is recommended for users to assess the distribution of their data based on sample types before proceeding with statistical analysis.
+If the user is not satisfied with the results or identifies any distributional issues, they can consider revisiting the transformation or normalization steps to address any concerns or improve the data distribution prior to statistical analysis.
+
+In the example dataset, the expressi
+on values for all sample types exhibit a unimodal distribution.
+However, when examining the quantile-quantile (qq) plot, heavy tails can be observed for all three groups.
+This indicates deviations from a perfectly normal distribution, suggesting potential outliers in the data.
+
+\newpage
+```{r ui_stat_HC_HIV_noMets, echo=F, out.width="90%", fig.align='center', fig.cap="Statistic module"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group, six differentially expressed proteins were identified.
+The Venndiagram indicates significantly up/downregulated proteins.
+The Volcano represents log10(adjusted p values) against log10 fold changes (logFC).
 These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
@@
 To account for potential batch effects, the batch factor was included as a cofactor in the model, ensuring that any observed differential expression is not confounded by batch variation.
+
+\newpage
+```{r ui_path_enrich_nw, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment. Networkplot indicating the interaction of differentially expressed proteins"}
+knitr::include_graphics("screenshots/path_enrich_nw.png")
+```
+
+The pathway enrichment mapped and tested with the exact Fisher test \citep{Sprent2011} the differentially expressed proteins P46777 (60S ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the ribosome and coronavirus disease (COVID-19) nodes.
+
+\newpage
+```{r ui_path_enrich_bar, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment. Barplot indicating the interaction significance of differentially expressed proteins"}
+knitr::include_graphics("screenshots/path_enrich_bar.png")
+```
+
+The barplot shows the significance of the two nodes.
+The count indicates how many proteins are associated with the node of the network.
+When hovering over the bar the user can read the p-value of the performed Fisher test.
+The background for the statistical test can be chosen as genome wide (default) or as the detected proteins.
+When choosing the detected proteins as a background no protein could be considered as statistically significant.

### 2023-06-16 markdown/results.Rmd
clustering tried --> no significant results obtained

@@
 When comparing the MI between the HC group and the sample type corresponding to individuals with both HIV and metabolic syndrome, the IQR is further reduced and shifted towards the lower end compared to the MI values observed in the raw data.
+
+\newpage
+```{r ui_constand_norm, echo=F, out.width="100%", fig.align='center', fig.cap="Popup showing the dependencies for the CONSTANd normalization"}
+knitr::include_graphics("screenshots/constand_norm_dependency.png")
+```
 
@@
 Starting from the peptide spectrum level and progressing to peptides aggregated for individual raw files, these plots offer a comprehensive view of the protein's peptide abundance across various batches.
-After merging the batches, the peptide appears to be abundant in all channels.
-Consequently, the need for missing value imputation with K-nearest neighbors (KNN) in this dataset becomes less likely, as missing values for the protein are uncommon once the batches are combined on the protein level.
+As indicated 14 batches show no abundance of peptides derived from the protein P61313 at all.
+After merging the batches, the peptides appear to be abundant in all channels.
 
@@
 Nonetheless, it is recommended for users to assess the distribution of their data based on sample types before proceeding with statistical analysis.
-If the user is not satisfied with the results or identifies any distributional issues, they can consider revisiting the transformation or normalization steps to address any concerns or improve the data distribution prior to statistical analysis.
+If the user is not satisfied with the results or identifies any distributional issues, the user can consider revisiting the transformation or normalization steps to address any concerns or improve the data distribution prior to statistical analysis.
 
-In the example dataset, the expressi
-on values for all sample types exhibit a unimodal distribution.
+The values for all sample types exhibit a unimodal distribution.
 However, when examining the quantile-quantile (qq) plot, heavy tails can be observed for all three groups.
@@
 When choosing the detected proteins as a background no protein could be considered as statistically significant.
+
+\newpage
+```{r ui_go_enrich_bar, echo=F, out.width="90%", fig.align='center', fig.cap="Gene ontology enrichment against the complete human proteome.
+Snapshot dotplot indicating the interaction significance of found proteins"}
+knitr::include_graphics("screenshots/go_enrich_dot.png")
+```
+
+The position of the dot shows the number of matches for the particular gene set. Colors indicate 
+for the significance level. In this analysis the entire repetoire of detected proteins was used. Only the gene set collections with the most protein matches were used.

### 2023-06-16 scope2.R

@@ -60,11 +60,11 @@ scp <- readSCP(featureData = mqScpData,
 
 
 
-# skip pool sample
-if ("Pool" %in% sampleAnnotation$SampleType) {
-  scp <- scp[, scp$SampleType !=  "Pool"]
-}
-
+# # skip pool sample
+# if ("Pool" %in% sampleAnnotation$SampleType) {
+#   scp <- scp[, scp$SampleType !=  "Pool"]
+# }
+# 
 
 
 # plot runs (not needed - sample unfractionized)
@@ -534,6 +534,11 @@ longFormat(scp[, , "proteins_norm"]) %>%
   ggplot(aes(x = missingness)) +
   geom_histogram()
 
+missing_values <- data.frame(longFormat(scp[, , "proteins_norm"]) %>%
+  data.frame %>%
+  group_by(colname) %>%
+  summarize(missingness = mean(is.na(value))))
+
 
 # use knn to impute missing values
 library(impute)
@@ -569,31 +574,37 @@ if (length(peptide_file) > 1) {
   assay(scp[["proteins_imptd"]]) <- protein_matrix$data
   } 
 
+# show missing values again
+scp[["proteins_imptd"]] %>%
+  assay %>%
+  is.na %>%
+  mean
 
   
   sce <- getWithColData(scp, "proteins_imptd")
-  
-  batch <- colData(sce)$Raw.file
-  model <- model.matrix(~SampleType, data = colData(sce))
-  
-  Combat_batchC <- function(i_exp_matrix, i_batch, i_model) {
-    out <- tryCatch(
-      {
-        i_exp_matrix <- ComBat(dat = i_exp_matrix,
-                               batch = i_batch,
-                               mod = i_model)
-        print("batch corrected and optimized")
-      } ,
-      error = function(cond) {
-        i_exp_matrix <- ComBat(dat = i_exp_matrix,
-                               batch = i_batch)
-        print("confounder detected! just batch corrected")
-      }
-    )
-    return(i_exp_matrix)
-  }
-  
-  assay(sce) <- Combat_batchC(assay(sce), batch, model) 
+
+# Batch correction   
+  # batch <- colData(sce)$Raw.file
+  # model <- model.matrix(~SampleType, data = colData(sce))
+  # 
+  # Combat_batchC <- function(i_exp_matrix, i_batch, i_model) {
+  #   out <- tryCatch(
+  #     {
+  #       i_exp_matrix <- ComBat(dat = i_exp_matrix,
[…]

### 2023-06-16 scpR-app/app.R

@@ -1887,8 +1887,7 @@ server <- function(input, output, session) {
       scp_0 <- scp()
       exp_matrix_0 <- exp_matrix()
       meta_data_0 <- meta_data()
-      
-      
+
       incProgress(2/3, detail=paste("creating linear model"))
       req(input$model_design)
       # Create a design matrix
@@ -1915,6 +1914,7 @@ server <- function(input, output, session) {
           SampleType <- scp_0$SampleType
           design <- model.matrix(~0+SampleType)
           colnames(design) <- sub("SampleType", "", colnames(design))
+
           
           fit <- lmFit(exp_matrix_0, design)
           
@@ -1928,7 +1928,7 @@ server <- function(input, output, session) {
           Group <- meta_data_0$Group
           design <- model.matrix(~0+Group)
           colnames(design) <- sub("Group", "", colnames(design))
-          
+
           fit <- lmFit(exp_matrix_0, design)
           
           user_contrast <- paste(input$selectedComp_stat, collapse = "-")
@@ -1950,7 +1950,7 @@ server <- function(input, output, session) {
           
           design <- model.matrix(~0+ . , data=design_frame)
           colnames(design)[1:length(unique(design_frame$SampleType))] <- sub("SampleType", "", colnames(design)[1:length(unique(design_frame$SampleType))])
-          
+
           fit <- lmFit(exp_matrix_0, design)
           
           user_contrast <- paste(input$selectedComp_stat, collapse = "-")

### 2023-06-19 markdown/discussion.Rmd
clustering tried --> no significant results obtained
discussion finished

@@
 The statistics module provides users with a simple tool to assess and visualize the results, offering clarity to dense scatter plots through hover functions.
-Enrichment mapping links statistically significant proteins to either the KEGG pathway database or individual research datasets, depending on user selection.
-These final results place the data in a biological context, which can be explored interactively in graphical visualizations such as network.
-
-## Mutual Information
-Information theory and the concept of mutual information provide a framework for understanding the conservation of entropy in a dataset.
-Traditionally, the validation of data transformation and normalization methods has been limited to observing distributions through histograms or Quantile-Quantile plots.
-However, the developed approach goes beyond these conventional methods, providing a deeper context for the variation in the data.
-By allowing users to choose the method that preserves the maximum amount of information, the approach enables researchers to extract the most valuable insights from their experiments.
+Enrichment mapping links statistically significant proteins to either the KEGG pathway database or individual research data sets, depending on user selection.
+These final results place the data in a biological context, which can be explored interactively in graphical visualizations.
 
@@
 
+## Mutual Information
+Information theory and the concept of mutual information provide a framework for understanding the conservation of entropy in a dataset.
+Traditionally, the validation of data transformation and normalization methods has been limited to observing distributions through histograms or Quantile-Quantile plots.
+However, the developed approach goes beyond these conventional methods, providing a deeper context for the variation in the data.
+By allowing users to choose the method that preserves the maximum amount of information, the approach enables researchers to extract the most valuable insights from their experiments.
+Normalization methods play a crucial role in single-cell data analysis, and their selection depends on the nature of the data and the specific analysis objectives.
+The mutual information analysis revealed that quantile normalization exhibited the smallest loss of entropy compared to other normalization methods
+
 ## Reporter ion intensity
-MS2 reporter ion intensities can be visualized as medians in form of a boxplot, allowing users to assess the distribution of intensities across different factors such as sample type, batch, or channel.
-This visualization helps identify potential biases or variations in the dataset.
+MS2 reporter ion intensities can be effectively visualized using a boxplot, where the medians represent the intensity distribution across various factors such as sample type, batch, or channel.
+This visualization aids users in assessing the variation and potential biases in the dataset.
+During the example analysis, the reporter ion intensity plot revealed inconsistencies in the dataset based on signal intensities associated with different batches.
+It was observed that certain runs exhibited a wide spread of signal intensities, which could indicate technical issues or batch-related variations.
+These inconsistencies were partly filtered out during the quality control process to ensure the reliability and accuracy of the data.
 
-## median coefficient of variation
-The reference values of the SCoPE2 publication \citep{Specht2021} could be tested and confirmed.
-5 as the number of observation for leading razor peptides was suitable to identify the cutoff of 0.65 by observing whiskers of the boxplots in order to reduce noisy peptides in the dataset.
+## Median coefficient of variation
+The median coefficient of variation (CV) was calculated for a razor protein in relation to its ambiguous peptides.
+This statistical parameter measures the variability of the razor protein's abundance across different peptide measurements.
+The results were then visualized using a boxplot, allowing users to assess the distribution of CV values based on a selected factor.
+During testing and fine-tuning, the number of observations for ambiguously matched peptides was adjusted to evaluate its impact on the CV.
+By varying the number of observations, researchers could determine the optimal parameter that yielded meaningful CV values.
+In this context, the reference values from the SCoPE2 publication \citep{Specht2021} were employed as benchmarks and subjected to validation.
+The results showed that using a number of observations of 5 for leading razor peptides was suitable for identifying a cutoff value of 0.65.
+This cutoff value was determined by observing the whiskers of the boxplots.
+By setting this threshold, noisy peptides in the dataset could be effectively reduced, enhancing the quality and reliability of the data.
 
-## Feature wise plot
-The feature-wise output further enhances the analysis pathway by enabling users to select individual proteins and explore their associated peptide spectrum, peptide and protein values over the course of the analysis.
-This feature provides a detailed view of the analysis results, allowing users to detect any missing or erroneous data points, particularly in channels with low or high abundance.
-By examining these individual protein profiles, users can evaluate the quality of the wet lab work and identify any possible errors or inconsistencies.
+## Feature-wise plot
+The feature-wise output plays a crucial role in enhancing the analysis pathway by offering users the ability to select specific proteins and explore their associated peptide spectra, peptide values, and protein values throughout the analysis.
+This feature provides a detailed and comprehensive view of the analysis results, enabling users to identify any missing or erroneous data points, particularly in channels with low or high abundance.
+By examining these individual protein profiles, users can assess the quality of the wet lab work and detect any potential errors or inconsistencies.
+In the case of observing the abundance of the protein P61313 (60S ribosomal protein L15) over the entire course of the experiment, it becomes apparent that there are missing values within certain runs and channels.
+Specifically, 14 runs exhibit no signal for this protein.
+This observation underscores the significance of aggregating data at the peptide and protein levels and combining individual runs to construct the final protein expression set.
+The scattering of dots in the plot indicates the inflation of small values and the deflation of large values after transformation.
+The scatter plot of quantile normalized values demonstrates the processing steps taken while preserving outliers and maintaining the biological relevance of the data.
+This processing steps aid in achieving a more balanced distribution of data and avoids the distortion of biological meaning.
 
@@
 The statistics module within the application provides users with a user-friendly interface for hypothesis testing, even without prior expertise in statistics.
-By leveraging the powerful R-package limma \citep{Ritchie2015}, the module offers robust statistical analysis methods, empowering users to perform tests and obtain results to validate their hypotheses.One key feature of the statistics module is its ability to handle multiple factors and therefore correct for batch effects.
+By leveraging the powerful R-package limma \citep{Ritchie2015}, the module offers robust statistical analysis methods, empowering users to perform tests and obtain results to validate their hypotheses.
+One key feature of the statistics module is its ability to handle multiple factors and therefore correct for batch effects.
 Users can extend the statistical analysis across various factors, allowing them to quantify the impact of different variables and identify differentially expressed proteins.
@@
 By providing intuitive and engaging visual representations, the module facilitates the understanding and communication of complex statistical analyses.
-
-All-Round Manipulation of the Actin Cytoskeleton by HIV \citep{Turville2018} example for the most differentially expressed gene. --> elaborate biological meaning.
-P68133 (Actin, alpha skeletal muscle)
-
-SAMHD1-Deficient CD14+ Cells from Individuals with Aicardi-Goutières Syndrome Are Highly Susceptible to HIV-1 Infection \citep{Berger2011} example for the most differentially expressed gene.
---> elaborate biological meaning.
-P31153 (S-adenosylmethionine synthase isoform type-2)
+When comparing the group HIV without metabolic syndrome and the healthy control 6 proteins could be found as differentially expressed after correcting the p-values by false discovery rate (FDR \citep{Benjamini1995}).
+For example, the protein P68133 (Actin, alpha skeletal muscle) exhibits a log fold change (logFC) of -0.25, indicating a significant decrease in actin levels within the HIV group.
+This finding aligns with previous publications suggesting that HIV exploits cellular skeleton proteins and alters their expression levels \citep{Turville2018}.
+By inducing orchestrated structural changes in compartments composed of actin filaments, the virus facilitates its spread and infects other lymphocytes \citep{Rodrigues2022}.
+Another protein of interest, P47914 (60S ribosomal protein L29), demonstrates the impact of viral infection on protein synthesis.
+Differential expression of this protein further highlights the perturbations in cellular processes caused by the virus.
 
@@
 Additionally, if desired, users can generate static plots and download them for use in presentations or other offline contexts.
+As indicated by the statistical findings, the KEGG enrichment analysis identified proteins associated with the ribosome and COVID-19.
+Both protein sets were connected to two ribosomal proteins, namely P47914 (60S ribosomal protein L29) and P46777 (60S ribosomal protein L5).
+These findings align with the fact that HIV and COVID-19, like all known viruses, depend entirely on the host cell's translation machinery for protein synthesis.
+This includes utilizing the host's ribosomes, tRNAs, amino acids, and all necessary factors for protein synthesis initiation, elongation, and termination \citep{Ohlmann2014}.
+Consequently, it is not surprising that these ribosomal proteins were found to be abundant in HIV-infected individuals.
+
+## Limitations
+
+```{r missingness, echo=FALSE, message=FALSE, warning=FALSE,  out.width="90%", fig.align='center', fig.cap="Missingness over channels.
+Y-axis: Height of the histogram bars show number of channels.
+X-axis: fraction of missing values.
+Red line: Mean of number of missing values over all runs"}
+
+library("scp")
+library("SingleCellExperiment")
+library("ggplot2")
+library("magrittr")
+library("dplyr")
+library("reshape2")
+library("scater")
+library("limma")
+library("CONSTANd")
+library("stats")
+library("impute")
+library("sva")
+library("tibble")
+
+# read in MS result table
+mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/evidence.txt")
+
+sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/sampleAnnotation.txt")
+
+# create QFeature object
+scp <- readSCP(featureData = mqScpData,
+               colData = sampleAnnotation,
+               channelCol = "Channel",
+               batchCol = "Raw.file",
+               removeEmptyCols = TRUE)
+
+# clean missing data
+scp <- zeroIsNA(scp, i=1:length(rowDataNames(scp)))
+
+# filter PSM
+# filter out potential contaminants
+# filter out matches to decoy database
+# keep PSMs with high PIF (parental ion fraction)
+scp <- filterFeatures(scp,
+                      ~ Reverse != "+" &
+                        Potential.contaminant != "+" &
+                        !is.na(PIF) & PIF > 0.10)
+
+nPSMs <- dims(scp)[1, ]
+
+# filter runs with too few features
+# however only one run is done --> changes nothing
+keepAssay <- dims(scp)[1, ] > 150
+scp <- scp[, , keepAssay]
+
+scp <- pep2qvalue(scp,
+                  i = names(scp),
+                  PEP = "PEP", # by reference the dart_PEP value is used
+                  rowDataName = "qvalue_PSMs")
+
+scp <- pep2qvalue(scp,
+                  i = names(scp),
+                  PEP = "PEP",
+                  groupBy = "Leading.razor.protein",
+                  rowDataName = "qvalue_proteins")
+
+scp <- filterFeatures(scp, ~ qvalue_proteins < 0.01)
+
+# matrixStats::colMedians, na.rm = TRUE aggregate over median 
+scp <- aggregateFeaturesOverAssays(scp,
+                                  i = names(scp),
+                                  fcol = "Modified.sequence",
+                                  name = paste0("peptides_", names(scp)),
+                                  fun = matrixStats::colMedians, na.rm = TRUE)
+
+# join assays to one
+if (length(names(scp)) > 1) {
+  scp <- joinAssays(scp,
+                    i = ((length(names(scp))/2)+1):length(names(scp)),
+                    name = "peptides")
+}
+
+# filter by median intensity
+file_name <- unique(sampleAnnotation$Raw.file)
+peptide_file <- paste("peptides_", as.character(file_name), sep = "")
+
+if (length(peptide_file) > 1) {
+  medians <- c()
+  for (assay_name in peptide_file) {
+    new_medians <- colMedians(assay(scp[[assay_name]]), na.rm = TRUE)
+    medians <- c(medians, new_medians)
+  }
+} else {
+  medians <- colMedians(assay(scp[[peptide_file]]), na.rm = TRUE)
+}
+scp$MedianRI <- medians
+
+# calculate median CV (coefficient of variation)
+if (length(peptide_file) > 1) {
+  scp <- medianCVperCell(scp,
+                         i = "peptides",
+                         groupBy = "Leading.razor.protein",
+                         nobs = 5, 
+                         norm = "div.median",
+                         na.rm = TRUE,
+                         colDataName = "MedianCV")
+} else {
+  scp <- medianCVperCell(scp,
+                         i = peptide_file,
+                         groupBy = "Leading.razor.protein",
+                         nobs = 5, 
+                         norm = "div.median",
+                         na.rm = TRUE,
+                         colDataName = "MedianCV")
+}
+
+scp <- scp[, !is.na(scp$MedianCV) & scp$MedianCV < 0.65, ]
+
+if (length(peptide_file) > 1) {
+  scp <- filterNA(scp,
+                  i = "peptides",
+                  pNA = 0.99)
+} else {
+  scp <- filterNA(scp,
+                  i = peptide_file,
+                  pNA = 0.99)
+}
+
+# aggregate peptide to protein
+if (length(peptide_file) > 1) {
+  scp <- aggregateFeatures(scp,
+                           i = "peptides",
+                           name = "proteins",
+                           fcol = "Leading.razor.protein",
+                           fun = matrixStats::colMedians, na.rm = TRUE)
+
+} else {
+  scp <- aggregateFeatures(scp,
+                           i = peptide_file,
+                           name = "proteins",
+                           fcol = "Leading.razor.protein",
+                           fun = matrixStats::colMedians, na.rm = TRUE)
+}
+
+# box cox transf
+boxcox_1 <- function(object, ...) UseMethod("boxcox_1")
+
+boxcox_1.formula <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- object$model
+    m <- length(lambda)
+    object <- stats::lm(object, y = TRUE, qr = TRUE)
+    result <- NextMethod()
+    result
+  }
+
+boxcox_1.lm <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- object$model
+    m <- length(lambda)
+    if(is.null(object$y) || is.null(object$qr))
+      object <- update(object, y = TRUE, qr = TRUE)
+    result <- NextMethod()
+    result
+  }
+
+boxcox_1.default <-
+  function(object, lambda = seq(-2, 2, 1/10), eps=1/50)
+  {
+    object$y <- as.matrix(object$model)
+    if(is.null(y <- object$y) || is.null(xqr <- object$qr))
+      stop(gettextf("%s does not have both 'qr' and 'y' components",
+                    sQuote(deparse(substitute(object)))), domain = NA)
+    if(any(y <= 0))
+      stop("response variable must be positive")
+    n <- length(y)
+    ## scale y[]  {for accuracy in  y^la - 1 }:
+    y <- y / exp(mean(log(y)))
+    logy <- log(y) # now  ydot = exp(mean(log(y))) == 1
+    xl <- loglik <- as.vector(lambda)
+    m <- length(xl)
+    for(i in 1L:m) {
+      if(abs(la <- xl[i]) > eps)
+        yt <- (y^la - 1)/la
+      else yt <- logy * (1 + (la * logy)/2 *
+                           (1 + (la * logy)/3 * (1 + (la * logy)/4)))
+      loglik[i] <- - n/2 * log(sum(qr.resid(xqr, yt)^2))
+    }
+    list(x = xl, y = loglik)
+  }
+
+protein_matrix <- assay(scp[["proteins"]])
+
+prot_lm <- lm(protein_matrix ~ 1)
+
+b <- boxcox_1(prot_lm)
+
+# Exact lambda
+lambda <- b$x[which.max(b$y)]
+
+if (round(lambda, digits = 0) == -2 || lambda < -1.5) {
+  protein_matrix <- 1/protein_matrix**2
+  ##print("1/protein_matrix**2")
+}
+if (round(lambda, digits = 0) == -1 || lambda < -0.75 && lambda > -1.5) {
+  protein_matrix <- 1/protein_matrix
+  ##print("1/protein_matrix")
+}
+if (round(lambda, digits = 1) == -0.5 || lambda < -0.25 && lambda > -0.75) {
+  protein_matrix <- 1/(protein_matrix**1/2)
+  ##print("1/(protein_matrix**1/2)")
+}
+if (round(lambda, digits = 0) == 0 || lambda < 0.25 && lambda > - 0.25 ) {
+  protein_matrix <- log10(protein_matrix)
+  ##print("log10(protein_matrix)")
+}
+if (round(lambda, digits = 1) == 0.5 || lambda > 0.25 && lambda < 0.75) {
+  protein_matrix <- protein_matrix**1/2
+  ##print("protein_matrix**1/2")
+}
+if (round(lambda, digits = 0) == 1 || lambda > 0.75 && lamdba < 1.5) {
+  protein_matrix <- protein_matrix
+  ##print("protein_matrix")
+}
+if (round(lambda, digits = 0) == 2 || lambda > 1.5) {
+  protein_matrix <- protein_matrix**2
+  ##print("protein_matrix**2")
+}
+
+sce <- getWithColData(scp, "proteins")
+
+scp <- addAssay(scp,
+                y = sce,
+                name = "proteins_transf")
+
+scp <- addAssayLinkOneToOne(scp,
+                            from = "proteins",
+                            to = "proteins_transf")
+
+assay(scp[["proteins_transf"]]) <- protein_matrix
+
+protein_matrix <- assay(scp[["proteins_transf"]])
+
+sce <- getWithColData(scp, "proteins_transf")
+
+scp <- addAssay(scp,
+                  y = sce,
+                  name = "proteins_norm")
+
+scp <- addAssayLinkOneToOne(scp,
+                              from = "proteins_transf",
+                              to = "proteins_norm")
+
+protein_matrix <- normalizeQuantiles(protein_matrix)
+
+assay(scp[["proteins_norm"]]) <- protein_matrix
+
+# missing value imputation
+# show missing values
+mean_missingness <- scp[["proteins_norm"]] %>%
+  assay %>%
+  is.na %>%
+  mean
+
+nrows_dataset <- nrow(assay(scp[["proteins_norm"]]))
+
+longFormat(scp[, , "proteins_norm"]) %>%
+  data.frame %>%
+  group_by(colname) %>%
+  summarize(missingness = mean(is.na(value))) %>%
+  ggplot(aes(x = missingness)) +
+  geom_histogram() +
+  stat_bin(bins=ncol(assay(scp[["proteins_norm"]]))) +
+  geom_vline(aes(xintercept=mean_missingness, col="red"))+
+  geom_text(aes(label=round(mean_missingness, 2), y=0, x=mean_missingness),
+            vjust=-1, col="red", size=3.5) +
+  guides(fill="none", colour="none") +
+  labs(y="n channels")
+
+```
+
+Unfortunately, a substantial number of individual proteins in the dataset (totaling r nrows_dataset) have missing values.
+On average, this amounts to approximately r round(mean_missingness, 2)*100% of the proteins across all runs and channels.
+This situation leads to the need for missing value imputation using techniques such as k-means, which can result in averaged results.
+Additionally, it is important to note that this imputation process can introduce potential loss of power in subsequent statistical analysis.
+To improve future experiments, it is recommended to increase the number of proteins in the carrier proteome.
+By expanding the repertoire of proteins included in the carrier proteome, it can enhance the coverage and depth of analysis, providing a more comprehensive understanding of the biological system under investigation.
+This can potentially lead to more robust and accurate results in downstream analyses.
 
@@
 
-Clustering for stratification of cells.
+### Clustering for stratification of cells.
 
-One future research question could utilize artificial intelligence by predicting cellular behaviour, cell types, or even diseases.
-Running the application on a server or high-performance cluster would be advantageous to address this question and employ simple machine learning tools such as decision trees, random forests, or support vector machines.
-The application of artificial intelligence on biological data could reduce the hands-on time by identifying unclear results caused by experimental design, unclean preparation, or incorrect computation.
-Another aspect, not limited to research topics, would be to focus on enhancing usability.
-While a development time of 6 months for a pipeline and user interface is sufficient to meet the analytical needs, there is still room for improvement in terms of user-friendliness.
-It is crucial to make the application more intuitive and robust to prevent crashes or errors caused by user mistakes.
-Ensuring a user-friendly experience is particularly important for commercial software, as robustness and ease of use are essential factors for success.
-However, since all software should be open source and developers should have the freedom to satisfy their curiosity, we can overlook this fact.
+The advantage of performing single-cell analysis lies in its ability to provide a fine resolution of cellular differences.
+Unfortunately variation causing effects such as cellular differentiation stages, are mainly unknown when starting single-cell experiments.
+The next steps in the development of the program aim to enable the stratification of single-cell data sets.
+The analysis typically begins by finding clusters within the datasets, identifying differentially expressed proteins, and annotating cell types based on their expression profiles.
+Subsequently, the analysis aims to distinguish sample types based on their expression profiles.
+
+Here I will explain how the analysis could start and elaborate the limitations and challenges for this approach.
+Additional processing steps will be included after the existing pipeline, so the developed program provides a starting point for further analysis.
+
+```{r cluster_analysis_dendrogram, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%", fig.align='center', fig.cap="Dendrogram showing the similarity over sample types.
+Tree was created using Ward`s minimum variance method."}
+
+# use knn to impute missing values
+library(impute)
+library(sva)
+
+if (length(peptide_file) > 1) {
+  protein_matrix <- assay(scp[["proteins_norm"]])
+
+  sce <- getWithColData(scp, "proteins_norm")
+
+  scp <- addAssay(scp,
+                  y = sce,
+                  name = "proteins_imptd")
+
+  scp <- addAssayLinkOneToOne(scp,
+                              from = "proteins_norm",
+                              to = "proteins_imptd")
+
+  protein_matrix <- impute.knn(protein_matrix, 
+                               k=3, 
+                               rowmax = 1, 
+                               colmax = 1, 
+                               maxp = Inf, 
+                               rng.seed = as.numeric(gsub('[^0-9]', '', Sys.Date())))
+
+  assay(scp[["proteins_imptd"]]) <- protein_matrix$data
+  }
+
+  sce <- getWithColData(scp, "proteins_imptd")
+
+  scp <- addAssay(scp,
+                    y = sce,
+                    name = "proteins_dim_red")
+
+  scp <- addAssayLinkOneToOne(scp,
+                                from = "proteins_imptd",
+                                to = "proteins_dim_red")
+
+# perform clustering before the rest
+
+  protein_matrix <- assay(scp[["proteins_dim_red"]])
+  colnames(protein_matrix) <- scp$SampleType
+
+  protein_matrix_t <- t(protein_matrix)
+
+  library("factoextra")
+  library("cluster")
+
+  clust <- agnes(protein_matrix_t, method = "ward")
+  pltree(clust, cex = 0.6, hang = -1, main = "Dendrogram") 
+```
+
+As observed in the dendrogram, it is evident that sample type alone does not account for the variation observed in the data.
+The branching patterns in the dendrogram suggest that other factors might be significant for contributing to the averaging of protein expression levels.
+This can potentially lead to a decrease in the statistical power of tests aimed at detecting differential expression between sample types.
+Ideally, clustering approaches should result in individual clusters that capture distinct sample types, enabling the testing of differential expression within each cluster.
+This stratification of samples within clusters allows for the identification of unique expression patterns associated with specific biological variations.
+The dendrogram, as demonstrated above, is an example of hierarchical clustering, which provides a visual representation of the clustering results.
+In this context, users should have the flexibility to select the desired number of clusters that best explain the underlying biological variation in their dataset.
+This allows for a more targeted and precise analysis of differential expression within each cluster, facilitating the identification of biologically meaningful patterns and insights.
+To begin with the k-means clustering approach, it is crucial to determine the optimal number of clusters before conducting the analysis.
+This can be achieved using various approaches, one of which is the gap statistic method \citep{Tibshirani2001}.
+
+```{r cluster_analysis_gap_stat, echo=FALSE, message=FALSE, warning=FALSE, out.width="90%", fig.align='center', fig.cap="Optimal numbers of clusters calculated with gap statistics method"}
+
+  protein_matrix_t_names <- protein_matrix_t 
+  rownames(protein_matrix_t_names) <- make.names(rownames(protein_matrix_t_names), unique = T)
+
+  gap_stat <- clusGap(protein_matrix_t_names, FUN = hcut, nstart = 25, K.max = 15, B = 50)
+  gap_stat_plot <-fviz_gap_stat(gap_stat)
+
+  plot_data <- gap_stat_plot$data
+
+    gap <- gap_stat$Tab[, "gap"]
+    se <- gap_stat$Tab[, "SE.sim"]
+optimal_clusters_gap <- maxSE(gap, se, method = "firstSEmax", SE.factor = 1)
+
+  gap_stat_plot
+```
+
+The gap statistic compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution.
+The null reference distribution represents a scenario where no clear clustering patterns exist in the data.
+However as seen in the gap statistic a high number of clusters will be difficult to distinguish from each other and explore the biological meaning such as differentiation in the example macrophage dataset.
+Another approach used to determine the optimal number of clusters is the silhouette method.
+
+```{r cluster_analysis_silh, echo=FALSE, message=FALSE, warning=FALSE, out.width="90%", fig.align='center', fig.cap="Optimal number of clusters calculated with silhoutte method."}
+
+  # unsupervised
+  sil_plot <- fviz_nbclust(protein_matrix_t, kmeans, method='silhouette', k.max=15)
+  plot_data <- sil_plot$data
+
+  optimal_clusters_sil <- as.numeric(plot_data[which(plot_data$y == max(plot_data$y)), "clusters"])
+
+  sil_plot
+
+```
+
+The silhouette method calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters. 
+If the overall distances between clusters are small, indicating overlapping or unclear separation, we take the minimum k.
+This suggests that the data points may not be well-separated into distinct clusters.
+As suggested by the silhouette method in this case a lower number of clusters to differentiate cell types is recommended.
+
+```{r cluster_analysis_dim_red, echo=FALSE, message=FALSE, warning=FALSE,  out.width="90%", fig.align='center', fig.cap="PCA plot with cluster indication.
+Shape of the datapoints indicate sample types.
+Clusters are encircled and color coded."}
+
+if (optimal_clusters_sil > optimal_clusters_gap) {
+  optimal_clusters <- optimal_clusters_gap
+} else {
+  optimal_clusters <- optimal_clusters_sil
+}
+
+km.final <- kmeans(protein_matrix_t, optimal_clusters)
+
+scp$cluster <- data.frame(km.final$cluster)
+
+library(scater)
+
+dimred_pca <- data.frame(calculatePCA(assay(scp[["proteins_dim_red"]]),
+                           ncomponents = 5,
+                           ntop = Inf,
+                           scale = TRUE))
+
+dimred_pca$cluster <- scp$cluster$km.final.cluster
+
+ggplot(dimred_pca, aes(x=PC1, 
+                       y=PC2,
+                       shape = scp$SampleType)) +
+  geom_point() +
+  guides(color="none", shape="none") +
+  stat_ellipse(data=dimred_pca, aes(fill=factor(cluster)),
+               geom="polygon", alpha=0.2, level=0.9, type="t", linetype=2, show.legend = F) +
+  coord_fixed()
+
+```
+
+As seen in the principal components analysis, clusters are densely packed, and the plot shows no clear differentiation between the `r optimal_clusters` suggested clusters.
+The dense packing and poor resolution indicates that other factors may be influencing the lack of distinct separation.
+One of the other factors involves the large number of imputed values due the high rate of missing values in the protein expression set.
+K-NN is not able to impute biological differences and therefore averages the particular missing value by its k neighbors.
+
+This suggests the utilization of a different data set for developing the clustering module of the future program.
+By employing a data set with a lower rate of missing values in the protein expression set, the accuracy and reliability of the clustering analysis can be enhanced.
+Having a data set with less missing data will allow for a more precise identification of clusters and differentiation between cell types.
+
+### Machine learning
+One potential target for future research could involve leveraging artificial intelligence techniques to predict cellular behavior, cell types, or even diseases.
+By running the application on a server and creating a database, it would be possible to harness the power of AI algorithms such as decision trees, random forests, or support vector machines.
+Applying artificial intelligence to biological data has the potential to enhance the classification of cell types, similar to the earlier discussed clustering approach.
+However, it is important to note that training and validating these algorithms typically requires supervision and can be a time-consuming task.
+
+Another area of focus, not limited to research topics, could be to enhance the usability of the application.
+While a development time of 5 months for the pipeline and user interface is sufficient to meet the analytical needs, there is always room for improvement in terms of user-friendliness.
+It is essential to consider the needs of all users, including the least experienced or "dumbest assumable user" (DAU).
+To prevent program crashes and errors, it may be necessary to implement try-catch around code blocks, as well as validate-need statements for the validation of user input.
+Additionally, creating a user-friendly environment involves establishing effective channels of communication between users and developers, such as utilizing platforms like GitHub for collaboration and feedback.

### 2023-06-19 markdown/results.Rmd
discussion finished

@@
 
-When comparing the MI between the HC group and the sample type corresponding to individuals with both HIV and metabolic syndrome, the IQR is further reduced and shifted towards the lower end compared to the MI values observed in the raw data.
+MI between the HC group and the sample type corresponding to individuals with both HIV and metabolic syndrome, the IQR is further reduced and shifted towards the lower end compared to the MI values observed in the raw data.
 
@@
 
-When comparing the MI IQR of the HC group with the sample type corresponding to individuals with both HIV and metabolic syndrome, the reduction in MI falls within the interquartile range of the raw data MI.
+Observing MI IQR of the HC group with the sample type corresponding to individuals with both HIV and metabolic syndrome, the reduction in MI falls within the interquartile range of the raw data MI.
 Additionally, the median remains close to unaffected, indicating that the overall central tendency of the MI values is relatively preserved.
@@
 
-When comparing the median reporter ion intensities with a focus on the batch (=Raw.file), it becomes evident that there are significant changes in signal that need to be taken into account in further analysis.
+Comparing the median reporter ion intensities with a focus on the batch (=Raw.file), it becomes evident that there are significant changes in signal that need to be taken into account in further analysis.
 After applying quality control measures, several values have been lost due to cutoffs in parameters such as parental ion fraction (PIF), q-value, coefficient of variation, and missingness of peptide data.
@@
 Upon switching to the indicator sample type in the 3D visualization, no clustering can be observed whatsoever.
-This suggests that the variance in the dataset is not attributed to sample type or batch, reinforcing the notion that other factors could contribute to the observed variability.
+This suggests that the variance in the data set is not attributed to sample type or batch, reinforcing the notion that other factors could contribute to the observed variability.
 
@@
 The position of the dot shows the number of matches for the particular gene set. Colors indicate 
-for the significance level. In this analysis the entire repetoire of detected proteins was used. Only the gene set collections with the most protein matches were used.
+for the significance level. In this analysis the entire repertoire of detected proteins was used. Only the gene set collections with the most protein matches were used.

### 2023-06-19 scope2.R

@@ -575,7 +575,7 @@ if (length(peptide_file) > 1) {
   } 
 
 # show missing values again
-scp[["proteins_imptd"]] %>%
+scp[["proteins_norm"]] %>%
   assay %>%
   is.na %>%
   mean
@@ -642,14 +642,16 @@ scp[["proteins_imptd"]] %>%
   groups <- cutree(final_clust, k=3)
   
   # unsupervised
-  plot <- fviz_nbclust(protein_matrix_t, kmeans, method='silhouette', k.max=100)
-  plot_data <- plot$data
+  sil_plot <- fviz_nbclust(protein_matrix_t, kmeans, method='silhouette', k.max=100)
+  plot_data <- sil_plot$data
   
-  optimal_clusters <- as.numeric(plot_data[which(plot_data$y == max(plot_data$y)), "clusters"])
+  optimal_clusters_sil <- as.numeric(plot_data[which(plot_data$y == max(plot_data$y)), "clusters"])
+
   
-  km.final <- kmeans(protein_matrix_t, optimal_clusters)
   
-  colData(scp)$cluster <- km.final$cluster
+  km.final <- kmeans(protein_matrix_t, optimal_clusters_sil)
+  
+  scp$cluster <- km.final$cluster
   
   
 # dimensionality reduction
@@ -667,7 +669,7 @@ scp[["proteins_dim_red"]] <- runPCA(scp[["proteins_dim_red"]],
 
 plotReducedDim(scp[["proteins_dim_red"]],
                dimred = "PCA",
-               colour_by = "SampleType",
+               colour_by = "MedianRI",
                point_alpha = 1,
                )
 

### 2023-06-21 markdown/Thesis.Rmd
first draft finished

@@
 
-(the spacing is set to 1.5)
-
-no more than 250 words for the abstract
-
-- a description of the research question/knowledge gap – what we know and what we don’t know
-- how your research has attempted to fill this gap
-- a brief description of the methods
-- brief results
-- key conclusions that put the research into a larger context
+With the growing prominence of single-cell techniques across various omics fields, there is a pressing need to develop a standardized pipeline for proteomics data in the realm of systems biology.
+Unlike DNA sequencing and RNA sequencing, proteomics analysis via mass spectrometry incurs high costs in terms of labor and equipment.
+Additionally, commercially available software solutions often come with hefty price tags and limited transparency regarding the underlying methods employed.
+However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility of post-processing using the R programming language.
+Guided by publications such as SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of the proteomics data analysis pipeline was thoroughly tested and validated using both bulk and single-cell mass spectrometry datasets.
+The pipeline implementation leverages the power of an interactive environment built in R Shiny, empowering users to discover the gold standard for their specific dataset.
+Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction.
+The developed application served as a valuable tool in guiding the identification of significantly differentially expressed proteins in HIV cells.
+Moreover, it facilitated the mapping of these proteins to relevant pathways and provided additional biological contexts for a comprehensive understanding of their functional implications.
+The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics.
+Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.
 

### 2023-06-21 markdown/conclusion.Rmd
first draft finished

@@
 
-Through this project, we have demonstrated that computational analysis can be made accessible to a wide range of biologists by providing an interactive environment.
+Through this project, I demonstrated that computational analysis can be made accessible to a wide range of biologists by providing an interactive environment.
 It is important to recognize that there cannot be a one-size-fits-all approach for proteomics datasets.
-Therefore, our application offers easy-to-interpret visualizations that allow users to assess the patterns and trends in their observed datasets.
+Therefore, the application offers easy-to-interpret visualizations that allow users to assess the patterns and trends in their observed datasets.
 The default options provided in the application are based on successful analysis pipelines such as ScoPE2 \citep{Specht2021}, serving as a starting point for customization at multiple levels.
 
+The example dataset provided showcases the inherent challenges in proteomic data analysis.
+It exhibits skewed distributions, batch effects, and a substantial number of missing values.
+Despite these complexities, the developed pipeline and user interface offer biologists a convenient means to address their research questions and validate the computations performed.
+The example use case also sheds light on the fact that previous proteomics publications, such as the ScoPE2 study \citep{Specht2021}, may not always employ the gold standard in data processing.
+This emphasizes the importance of critically evaluating and refining data processing steps through trial and error to achieve optimal results.
+The interactive environment provided by the application allows users to navigate the trial and error process without experiencing frustration.
+This user-friendly interface ensures that researchers can iteratively refine their data processing steps and analyze the results in a seamless manner.
+Furthermore, the scalability of the application was initially unknown until the final submission of the single-cell dataset.
+The programming and testing of the application primarily focused on bulk datasets, typically consisting of up to four runs at once.
+However, it is noteworthy that the application demonstrates remarkable efficiency and performance, even when running large datasets.
+The ability of the application to efficiently handle larger datasets, such as single-cell studies, is a significant advantage.
+It allows researchers to analyze complex and extensive datasets without sacrificing performance or compromising the quality of the results.
+This scalability makes the application a valuable tool for analyzing diverse types of biological data, accommodating the needs of researchers working with both smaller and larger datasets.
+
 The developed pipeline and user interface serve as valuable tools for individual researchers and research communities, enabling them to analyze and visualize their data in meetings, even on non-performant computers.
-However, our future perspective involves deploying the application on a server, making it accessible to research groups around the world.
+However, one of the future perspectives involves deploying the application on a server, making it accessible to research groups around the world.
 This would enhance collaboration and facilitate the sharing of insights and findings across different geographical locations.
-By providing a centralized platform for proteomics analysis, we aim to foster scientific advancements and enable researchers to make significant contributions to their respective fields.
+By providing a centralized platform for proteomics analysis, aiming to foster scientific advancements and enable researchers to make significant contributions to their respective fields.
 Furthermore by having a central hub for proteomic data analysis and the use of artificial intelligence the biological meaning could be brought to a wider context.
 Since proteomic data could be saved without violating data protection laws, algorithms could be employed to learn from large datasets.
+
+An important aspect of learning from the data involves annotating specimens, particularly in the context of single-cell proteomics.
+The clustering approach is a primary focus for future development and analysis in the near term.
+By employing clustering algorithms and stratifying the data, individual cells can be labeled and annotated based on their phenotypes.
+This annotation process involves associating protein sets collections with specific clusters or cell types, providing a more comprehensive understanding of cellular heterogeneity and functional characteristics.
+
+In conclusion, the developed pipeline and user interface provide a user-friendly and efficient approach for proteomic data analysis, enabling biologists to address their research questions and validate computations.
+Future perspectives include deploying the application on a server for global accessibility, leveraging artificial intelligence techniques, and implementing clustering approaches to further enhance annotation and understanding of cellular heterogeneity and functional characteristics.

### 2023-06-21 markdown/methods.Rmd
first draft finished

@@
 
-### Ionization
+#### Ionization
 For the reference data ESI ionization was used.
 
-### Mass spectrometer
+#### Mass spectrometer
 For the reference data used for developing the pipeline an Orbitrap™ Fusion™ Lumos™ tribrid mass spectrometer (ThermoFisher Scientific) was used.
@@
 
-## Computational methods
+## Drylab - Computational methods
 ```{r Dataflow, eval=FALSE, include=FALSE}
@@
 The default settings were used. TMTpro labels were corrected according to the isotopical distribution of the Lot-number provided by Thermo Fisher.
-
-### Data processing
-Further analysis is done with R and respective packages such as bioconductor. Please find version info in the appendix.
 
@@
 
-#### Reading the data
+### Processing the data
+Further analysis is done with R and respective packages such as bioconductor. Please find version info in the appendix.
 
 After processing MaxQuant creates a directory containing all results as .txt file.
-The evidence.txt file include all peptide to spectrum matches (PSM) with their respective proteins and statistical parameters.
+The evidence.txt file includes all peptide to spectrum matches (PSM) with their respective proteins and statistical parameters.
 
@@
 * Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)
-	+ Mass
 * Retention time
 * Precursor Ion Fragment
-	+ source of the detected ion also referred as mother ion
 * Fraction of total spectrum 
@@
 * Reporter intensity (RI)
-	+ corrected RI
 * Posterior error probability (PEP)
 
-#### Object oriented programming
+### Object oriented programming
 
@@
 
-#### Zero values
+### Zero values
 Peptides with low abundance are often set to zero during analysis.
@@
 Given that it is highly unlikely for a biological cell of a comparable type and function to not contain a particular protein, replacing the zero value with "not applicable" (NA) is crucial for understanding and interpreting MS data.
+NA values will be managed in missing value handling later on the protein level.
 
-#### Exclude reverse matches/contaminants
-Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches were be excluded from further analysis.
+### Exclude reverse matches/contaminants
+Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches will be excluded from further analysis.
 
-#### Filter according to precursor ion fraction (PIF)
+### Filter according to precursor ion fraction (PIF)
 During mass spectrometry, the ions detected in MS1 are further fragmented through collision during multiple MS runs.
@@
 These peptides need to be filtered out during the analysis pipeline.
-A cutoff value, referenced in the SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface, but it can be adjusted according to the needs of the biologist.
+A cutoff value, referenced in the SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface, but it can be adjusted.
 
-#### Filter by q-value
+### Filter by q-value
 The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
@@
 In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score.
-The peptides spectrum matching score is defined as -10log10(p).
+The peptides spectrum matching score is defined as $-10 * log(p)$.
 Whereas the p-value is defined that the hit is done by chance.
 The calculation of the score is highly dependent on the data acquisition method used.
-MaxQuant uses Andromeda , an integrated search engine.
+MaxQuant uses Andromeda, an integrated search engine.
 Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora.
@@
 
-#### Peptide spectrum match (PSM) aggregation to peptides 
+### Peptide spectrum match (PSM) aggregation to peptides 
 In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function.
-In the context of processing from peptides to spectrum matches, the desired column is the peptide sequence.
+In the context of processing from PSMs to unique peptides, the desired column is the PSM sequence.
 To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.
 
-#### Join assays when observing multiple comparable batches at once
+### Join assays when observing multiple comparable batches at once
 Sample size is often a limiting factor in hypothesis testing.
@@
 To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis.
+For further processing the runs are joined in one object at the peptide level.
 
-#### Calculate reporter ion intensity (RI) and filter according to median RI
+### Calculate reporter ion intensity (RI) and filter according to median RI
 Columns which do not meet the desired intensity can be filtered by a threshold set on the RI.
@@
 Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes.
+The correction values can be found on the vendor`s webpage (in this case Thermo Fisher) and correction needs to be set up in MaxQuant.
 
-#### Calculate and filter according to median coefficient of variation (CV) per cell/channel
+### Calculate and filter according to median coefficient of variation (CV) per cell/channel
 Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data.
@@
 
-#### Remove peptides with high missing rate
+### Remove peptides with high missing rate
 Although missing value imputation can be performed during the analysis of multiple batches, peptides with missing detections across channels can be problematic for quantification.
 The proteomic composition of a biological sample is similar between replicates and even across groups.
-However, the threshold of missingness (described as a fraction of the row) can be set in the user interface and adjusted to enable different experimental designs.
+However, the threshold of missingness (described as a fraction of the row) can be set in the user interface and adjusted.
 
-#### Aggregation of peptides to proteins 
+### Aggregation of peptides to proteins 
 Similar to the already explained previous aggregation step the peptides will be further processed into their respective proteins after the quality control on the peptide level is performed.
 Finally an expression matrix for every protein and their respective channel column is returned.
-Each channel refers to a sample and reflects contains the intensities of each protein found in the biological specimen.
+Each channel refers to a sample and contains the intensities of each protein found in the biological specimen.
 
-#### Transformation of protein expression data
+### Transformation of protein expression data
 Data transformation applies a function to each value of a matrix or array, so that: 
@@
 When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
-Practically impossible, but in data-driven science, occasionally negative expression values may be present, and they can be converted into positive values by taking each one to the power of two.
-Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qqplots, and MA-plots.
+Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qq-plots, and MA-plots.
 For statistically inexperienced users, the boxcox transformation can help with this decision.
@@
 
-#### Normalization
+### Normalization
 The term normalization is ambiguous in data-science and will be explained briefly in this chapter.
 Starting with converting the data to the Z-distribution (Gaussian result), where 0 represents the mean and 1 represents the standard deviation, in vectorization every feature (or protein) is represented as a vector that points in a specific direction in the unit sphere, normalization can have different meanings.
-Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to our statistical model and hinders large values from biasing the model in a particular direction.
+Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to the statistical model and hinders large values from biasing the model in a particular direction.
 However, this could also have negative consequences for the modeling procedure, as it may reduce the impact of important features on the dataset.
@@
 The K-means clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances.
-This method will be applied in the missing value imputation which will be the next step applied to our data.
-Therefore, we want to ensure that we obtain the most accurate imputed values possible.
+This method will be applied in the missing value imputation, which will be the next step applied to the data.
 The first normalization method available in the user interface is column-wise median and row-wise mean normalization.
@@
 CONSTANd uses a technique called matrix raking, which employs the RAS algorithm.
-The expression set reflects the nonnegative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $B$ which equals to $diag(x) * A *diag(y)$ .
+The expression set reflects the non-negative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $B$ which equals to $diag(x) * A *diag(y)$ .
 Whereas $x \in \mathbb{R}^{m}$ and $y \in \mathbb{R}^{n}$.
@@
 This assumption can be observed in the MA-plot which indicates the differences of two samples.
-If we have the intensities of two samples, $R$ and $G$, we can plot them on a graph with the x-axis as $M=\log_2(\frac{R}{G})$ and the y-axis as $A=\frac{1}{2}\log_2(RG)$.
+Given the intensities of two samples, $R$ and $G$, visualized on a graph with the x-axis as $M=\log_2(\frac{R}{G})$ and the y-axis as $A=\frac{1}{2}\log_2(RG)$.
 A comparison between all sample types of identically processed sets is needed before calculation and the assumptions must be fulfilled.
@@
 The method is utilized by the bioconductor package handler.
-However if the dataset already consists of balanced features and samples or the user wants to benchmark the methods applied, normalization can be skipped also.
+Another method developed in the application is the quantile normalization.
+Quantile normalization is performed over all columns, equalizing statistical parameters such as inter quartile range, median and mean.
+However the application of this method is only suitable if the number of differentially expressed proteins is approximately equal over sample types.
+Otherwise statistical power could be lost by removing fine deviations \citep{Zhao2020}.
+For the quantile normalization method the package limma was utilized.
+Quantile normalization ranks protein expression by magnitude, then averages the expression values occupying the same rank and substitute the inital value with the average.
+However if the dataset already consists of balanced features and samples or the user wants to benchmark the methods applied, normalization can be skipped also, by setting to "none".
 
-#### Missing value handling
+### Missing value handling
 When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches.
@@
 
-#### Batch correction 
+### Batch correction 
 Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data.
@@
 It appears convenient to remove the batch effect before performing any statistical analysis, which is in turn prone to errors especially if the sample groups are not split equally between batches.
-In this case group differences influence the batch effect reducing statistical power after the correction., the both effects are so called interdependent.
+In this case, group differences influence the batch effect reducing statistical power after the correction.
+These both effects are so called interdependent.
 The opposite could happen if the experimental design is heavily unbalanced leading to group differences induced by batch correction.
-Finally it is recommended to include the batch factor should in the linear model as a co factor and statistically quantify it. \citep{Nygaard2016}
+Finally it is recommended to include the batch factor should in the linear model as a co-factor and statistically quantify it \citep{Nygaard2016}.
 However it could be of interest to observe biological relevance of the factor of interest in the dimensionality reduction visualization, so a batch correction option was included to meet this need.
@@
 
-#### Dimensionality reduction
+### Dimensionality reduction
 After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained.
@@
 
-##### Principle component analysis (PCA)
+#### Principle component analysis (PCA)
 The most common approach in data science for any kind of high dimensional data is the PCA.
@@
 
-##### Uniform manifold approximation & projection (UMAP)
+#### Uniform manifold approximation & projection (UMAP)
 UMAP \citep{McInnes2018} is a dimensionality reduction technique that was developed in recent years and is based on Riemann geometry.
@@
 
-#### Validation of the methods applied to the dataset
+### Validation of the methods applied to the dataset
 For the validation of methods applied to the dataset, the mutual information calculation is used to assess the impact of computations and the information it carries. 
@@
 In a Venn diagram $I(X;Y)$ (the mutual information) corresponds to the intersection of the two conditional entropys $H(X|Y)$ and $H(Y|X)$.
-To calculate the mutual information between two variables $X$ and $Y$ we can alternatively express the above equation as $I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})$.
+To calculate the mutual information between two variables $X$ and $Y$ the above equation can be alternatively expressed as $I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})$.
 $D_{KL}$ denotes for the Kullback-Leibler \citep{Kullback1951} divergence which is a concept of relative entropy between two variables and tells us how much 2 probability distributions differ from each other.
@@
 
-#### Statistics
-##### Differential expression
+### Statistics
+#### Differential expression
 After visualizing the biological tendency in the dataset, the next step is hypothesis testing.
 Several experimental designs can be tested against the null hypothesis $h_0$, which states that there is no significant difference between the sample types.
-Before proceeding with any statistical testing, it is important to observe the data using either a quantile-quantile norm plot (qq-norm plot) or a histogram.
-The quantile-quantile plot (qq-plot) serves a vital role in exploratory statistics by visualizing whether two datasets follow the same underlying distributions.
+Before proceeding with any statistical testing, it is important to observe the data using either a quantile-quantile plot (qq-plot) or a histogram.
+The qq-plot serves a vital role in exploratory statistics by visualizing whether two datasets follow the same underlying distributions.
 In the case of statistical preparation for parametric tests, the first distribution is the dataset of interest, while the second dataset is the normal distribution.
@@
 However, values at the ends of the distribution may deviate from the normality line, indicating differentially expressed genes.
-These genes can be observed both in the histogram and the qq-norm plot.
-
+These genes can be observed both in the histogram and the qq-plot.
 Parametric statistical tests rely on the assumption of normality in the data, which needs to be achieved before building a model.
@@
 By observing the results in the plots, one can determine which combination of data processing methods satisfies the conditions required for the tests.
+However the limma method utilized in the statistics module shows great robustness to imperfect normal distributed values \citep{Ritchie2015}.
 
-##### Differential expression 
 To test for differential expression, the analysis utilizes the "limma" package from R Bioconductor \citep{Phipson2016}.
@@
 When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis.
-This is done by the multi factor option in the user interface by selecting the batch as a co factor and inclusion in the design matrix.
-The advantage of this procedure is, that we can quantify the biological relevance of working in multiple batches and avoid publishing false positive results.
+This is done by the multi-factor option in the user interface by selecting the batch as a co-factor and inclusion in the design matrix.
+The advantage of this procedure is, that the biological relevance of working in multiple batches can be quantified and the publishing of false positive results avoided.
 The multi factor option uses an additive model as a design matrix, whereas continuous variables (such as size, weight etc.) can also be used to build the statistical model.
@@
 For estimation, it is assumed that $\frac{1}{\sigma^2_g} \sim \frac{1}{d_{0}s^2_{0}} \chi^2_{d_{0}}$, representing the prior information of the model.
-The probability that $\beta_{gj} \neq 0$ is denoted as $p_j$, where $p_j$ represents the expected proportion of truly differentially expressed genes.
+The probability for $\beta_{gj} \neq 0$ is denoted as $p_j$, where $p_j$ represents the expected proportion of truly differentially expressed genes.
 The expected distribution of log-fold changes follows the distribution $\beta_{gj} | \sigma^2_g , \beta_{gj} \neq 0 \sim N(0, v_{0j}\sigma^2_g)$.
@@
 
-#### Protein set enrichment analysis
+### Protein set enrichment analysis
 After differential expression the upcoming question could be, in which context the up or down regulated proteins are.
@@
 Within the list are all truly differentially expressed genes, an annotation is the identification of a protein in the desired biological context.
-We expect that the values populate the fields in the table as follows:
+The table is expected to be populated by the values as follows:
 
@@
 Against that the alternative hypothesis $H_{A}$ says that there is an underlying tendency the table is populated.
-This leads to our test statistics, the observed values (=$O$) in the table are significantly different to the expected (=$E=\frac{(A+B \lor C+D)}{N}$ ones.
+This leads to the test statistics, the observed values (=$O$) in the table are significantly different to the expected (=$E=\frac{(A+B \lor C+D)}{N}$ ones.
 One approach is approximating $\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}$ and deriving the p-value by the area under the density curve of the $\chi^2$ distribution $Q = \sum_{i=1}^{k} Z_{i}^{2}$.
-In the distribution $k$ is defined by the degrees of freedom (=df), what can be obtained by $k=(columns-1)*(rows-1)=df$ of our table.
+In the distribution $k$ is defined by the degrees of freedom (=df), what can be obtained by $k=(columns-1)*(rows-1)=df$ of the table.
 Another approach to obtain an exact p-value would be Fisher`s exact test, which states that the margins are distributed according to the hypergeometric distributions.
@@
 $p=1-\sum_{i=0}^{A-1}\frac{\binom{B}{i}\binom{D-B}{A+C-i}}{\binom{D}{A+C}}$
-Here D denotes either for the background distribution which can be the total number of proteins obtained in our experiment, all proteins with annotation or a custom selected background.
-The background indicates all proteins which could be positive tested for in our observed sample.
+Here D denotes either for the background distribution which can be the total number of proteins obtained in the experiment, all proteins with annotation or a custom selected background.
+The background indicates all proteins which could be positive tested for in the observed sample.
 The above mentioned three options are implemented in the user interface.
@@
 
-##### Pathway based protein enrichment analysis
+#### Pathway based protein enrichment analysis
 The pathway-based protein enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}.
@@
 
-##### Custom ontology based protein enrichment analysis
+#### Custom ontology based protein enrichment analysis
 For custom ontologies the protein enrichment analysis is done using the R bioconductor package piano \citep{Vaeremo2013}.

### 2023-06-21 markdown/results.Rmd
first draft finished

@@
 
-The position of the dot shows the number of matches for the particular gene set. Colors indicate 
-for the significance level. In this analysis the entire repertoire of detected proteins was used. Only the gene set collections with the most protein matches were used.
+The position of the dot shows the number of matches for the particular protein set. Colors indicate 
+for the significance level. In this analysis the entire repertoire of detected proteins was used. Only the protein set collections with the most protein matches were used.

### 2023-06-21 markdown/acknowledgments.Rmd
proofread

@@
 
-Thank you for following this tutorial!
+In addition to my faithful feline companion, Apollo, who took care of my place in Austria while I was working on this project in Stockholm, I am deeply grateful to all the members of Ujjwal Neogi's research group.
 
-I hope you'll find it useful to write a very professional dissertation.
+Their unwavering guidance and support throughout this project have been truly invaluable, and I am profoundly thankful for their expertise and assistance.

### 2023-06-21 markdown/discussion.Rmd
first draft finished
proofread

@@
 # Discussion
-
-## Intro
 The developed application and the proteomics experiment pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
@@
 The silhouette method calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters. 
-If the overall distances between clusters are small, indicating overlapping or unclear separation, we take the minimum k.
+If the overall distances between clusters are small, indicating overlapping or unclear separation, it is suggested to take the minimum k.
 This suggests that the data points may not be well-separated into distinct clusters.

### 2023-06-21 markdown/introduction.Rmd
proofread

@@
 Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as "omics," to comprehensively analyze biomolecules in diverse contexts.
-Omics research has emerged as an approach to studying biomolecules and their interactions within biological systems comprehensively.
+Omics research has emerged as an approach to study biomolecules and their interactions within biological systems comprehensively.
 Omics encompasses various fields, including genomics, transcriptomics, proteomics, and metabolomics, with each field focusing on a specific class of biomolecules.
@@
 Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
+Proteomics is a comprehensive approach that investigates the complete protein composition of a specimen, aiming to understand the intricate biological network it represents.
 For instance if pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
-These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications and more \citep{Beltran2017}.
-By studying the proteome, which encompasses all the proteins present in a biological system at a specific moment, we can gain a comprehensive understanding of the hierarchical organization and functional dynamics of biological processes.
+These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications \citep{Beltran2017}.
+By studying the proteome, we can gain a comprehensive understanding of the hierarchical organization and functional dynamics of biological processes.
 
@@
 This approach finds valuable applications for instance in oncology, where it can provide insights into the protein profile of a tumor.
-By studying oncogenic biomarkers and expression patterns, this knowledge can contribute to the development of screening methods and the customization of treatments based on observed expression patterns.
+By studying oncogenic biomarkers and expression patterns, this knowledge can contribute to the development of screening methods and the customization of treatments.
 Additionally, bulk proteomics can aid in identifying the progression of tumor development \citep{Kwon2021}.
@@
 To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed \citep{Liou2015}.
-These techniques rely on the detection of surface proteins on cells and the specific binding of antibodies to these proteins.
-Antibodies, typically derived from the immune system of other species, bind to antigens in a lock-and-key manner.
+These techniques primarily rely on the detection of surface proteins on cells and the specific binding of antibodies to these proteins.
+Auto-fluorescence is observed in certain cells, wherein they possess the capability to emit a light signal when stimulated, and this emission can be used to sort them accordingly.
+However auto-fluorescence can lack of accuracy as shown in studies indicating the effect of laboratory gloves on cellular samples \citep{Philpott2017}.
+On the other side antibodies, typically derived from the immune system of other species, bind to antigens in a highly specific lock-and-key manner.
 Antibodies are widely used in molecular biology and find applications in cell isolation.
@@
 Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the  principles of cell type behavior, such as differentiation.
-Differentiation, a critical process in various biological systems such as the immune system, involves the acquisition of specialized functions by specific cell types.
+Differentiation, a critical process in various biological systems such as the immune system, involves the acquisition of specialized functions by specific cell types \citep{Fang2018}.
 Investigating differentiation in multiple cells can be challenging due to the inherent heterogeneity during the initiation of this process.
@@
 The major disadvantage of this technique was the limitation to only stain a few proteins per cell and the requirement of a-priori knowledge of the protein to analyze.
-For quantification procedures such as Western blots, immunoassays or cytometry by time of flight (CyTOF) have been used.
-Other disadvantages are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
-Early quantitative methods in molecular biology were assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay measuring the entire protein concentration.
+For quantification, procedures such as Western blots, immunoassays or cytometry by time of flight (CyTOF) have been used.
+Challenges are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
+Early quantitative methods in molecular biology, measuring the entire protein concentration, were assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay.
 These methods are still in use to obtain total concentration for sample preparation.
@@
 Furthermore multiple proteins can be derived from a single DNA-sequence with alternative splicing and different proteins coming from non related sequences acting like building blocks can form protein complexes and/or Protein-RNA complexes.
-Cells are highly efficient and do not waste energy by producing obsolete proteins, moreover the degradation of proteins is an important energy recovery mechanism.
+Cells are highly efficient and do not waste energy by producing obsolete proteins, moreover the degradation of proteins in the proteasome is an important energy recovery mechanism \citep{TANAKA2009}.
 A bottom-up strategy can elucidate these properties of the protein ensemble at a given time leading to proteomics.
@@
 It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the protease enzyme trypsin.
-This process is known as proteolysis.
-The enyme cleaves specifically at the caroboxyl side of arginine and lysis residues.
+The process known as proteolysis takes place in multiple parts of multi cellular organisms and was firstly detected in the small intestine digesting proteins \citep{Wang2008}.
+The enzyme cleaves specifically at the caroboxyl side of arginine and lysin residues.
 Afterwards a charge remains on the residues making them detectable for the mass spectrometer.
@@
 In metabolic labeling the cells were fed with aminoacids containing heavy isotopes, it is the method of choice in order to label peptides at the earliest possible level.
-This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants.
+This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants \citep{Karen2012}.
 Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
@@
 Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
-Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics \citep{Ye2022}.
+Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell \citep{Ye2022}.
 
@@
 In HPLC the molecules eluate through a narrow column with high pressures (50-350 bar), compared to LC where the separation is gravity dependent.
-Similar to gel electrophoresis molecules can be separated upon different properties depending on the type of the chromatography column.
+Similar to gel electrophoresis molecules can be separated upon different chemical properties, depending on the type of the chromatography column.
 In a typical proteomics experiment peptides are separated with reversed phase chromatography (RPLC).
 The principle behind RPLC is a column with a nonpolar stationary phase which interacts with the nonpolar residues of the peptides.
-Peptides were fractionated according to their hydrophobicity.
+Peptides were fractionated according to their hydrophobicity \citep{Pitt2009}.
 
@@
 Autocorrelation between centroids is accounted for by considering only well-identified peptides.
-Published data indicates that the mass precision within an MS experiment typically ranges around 10^-7.
+Published data indicates that the mass precision within an MS experiment typically ranges around 10^-7 \citep{Wilschefski2019}.
 
@@
 The availability of gene sequences in databases and the computational matching of peptides against those sequences enable the identification of alterations in a sample at the protein level.
-These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylation, methylation \citep{Aebersold2003}.
+These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylations or methylations \citep{Aebersold2003}.
 The peptide identification (P-) score reflects the goodness of fit between the data and the identified sequence in the database, taking into account the length of the peptide.

### 2023-06-21 markdown/methods.Rmd
proofread

@@
 
-### Join assays when observing multiple comparable batches at once
+### Join assays
 Sample size is often a limiting factor in hypothesis testing.
@@
 
-### Calculate reporter ion intensity (RI) and filter according to median RI
+### Calculate median reporter ion intensity (RI) and filter
 Columns which do not meet the desired intensity can be filtered by a threshold set on the RI.
@@
 
-### Calculate and filter according to median coefficient of variation (CV) per cell/channel
+### Calculate median coefficient of variation (CV) and filter
 Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data.

### 2023-06-26 markdown/Thesis.Rmd
anoops improvements implemented

@@
   - \usepackage{float}
+  - \let\origfigure\figure
+  - \let\endorigfigure\endfigure
+  - \renewenvironment{figure}[1][2] {
+      \expandafter\origfigure\expandafter[H]
+    } {
+    \endorigfigure
+    }
   - \usepackage{sectsty}
@@
 {\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
+presenting ProteoScanR & Proteomics Workbench
 
@@
 {\bf Lukas Gamp}
+01630003
 
@@
 
-in partial fullfilment of the requirement \\for the degree of MSc \\in Bioinformatics
+in partial fulfillment of the requirement \\for the degree of MSc \\in Bioinformatics
 
@@
 \normalsize
-mm yy
+July 2023
+
+Main Supervisor: 
+Dr. DI(FH) Gerhard Duernberger
+
+External Supervison:
+Assoc. Prof. Ujjwal Neogi, M.Sc. PhD
+
+Second examiner:
+FH-Prof. Dr. Alexandra Graf
+
 \end{centering}
@@
 Additionally, commercially available software solutions often come with hefty price tags and limited transparency regarding the underlying methods employed.
-However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility of post-processing using the R programming language.
-Guided by publications such as SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of the proteomics data analysis pipeline was thoroughly tested and validated using both bulk and single-cell mass spectrometry datasets.
-The pipeline implementation leverages the power of an interactive environment built in R Shiny, empowering users to discover the gold standard for their specific dataset.
+However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility using the R programming language.
+Introducing ProteoScanR, a state-of-the-art proteomics pipeline integrated into the user-friendly Proteomics Workbench interface. 
+Guided by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of ProteoScanR was thoroughly tested and validated using both bulk and single-cell mass spectrometry data sets.
+The pipeline implementation in the Proteomics Workbench leverages the power of an interactive environment built in R Shiny, empowering users to discover valuable insights for their specific data set.
 Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction.
-The developed application served as a valuable tool in guiding the identification of significantly differentially expressed proteins in HIV cells.
-Moreover, it facilitated the mapping of these proteins to relevant pathways and provided additional biological contexts for a comprehensive understanding of their functional implications.
+ProteoScanR and the Proteomics Workbench served as a valuable tool in guiding the identification of significantly expressed proteins in cells under study.
+Subsequently, the pathway enrichment analysis provided additional biological contexts for a comprehensive understanding of their functional implications.
 The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics.
@@
 ```
+
+\newpage
+\listoffigures

### 2023-06-26 markdown/conclusion.Rmd
anoops improvements implemented

@@
 It exhibits skewed distributions, batch effects, and a substantial number of missing values.
-Despite these complexities, the developed pipeline and user interface offer biologists a convenient means to address their research questions and validate the computations performed.
+Despite these complexities, the ProteoScanR pipeline and Proteomics Workbench interface offer biologists a convenient means to address their research questions and validate the computations performed.
 The example use case also sheds light on the fact that previous proteomics publications, such as the ScoPE2 study \citep{Specht2021}, may not always employ the gold standard in data processing.
@@
 
-The developed pipeline and user interface serve as valuable tools for individual researchers and research communities, enabling them to analyze and visualize their data in meetings, even on non-performant computers.
+ProteoScanR pipeline and Proteomics Workbench user interface serve as valuable tools for individual researchers and research communities, enabling them to analyze and visualize their data in meetings, even on non-performant computers.
 However, one of the future perspectives involves deploying the application on a server, making it accessible to research groups around the world.
@@
 
-In conclusion, the developed pipeline and user interface provide a user-friendly and efficient approach for proteomic data analysis, enabling biologists to address their research questions and validate computations.
+In conclusion, the ProteoScanR pipeline and Proteomics Workbench interface provide a user-friendly and efficient approach for proteomic data analysis, enabling biologists to address their research questions and validate computations.
 Future perspectives include deploying the application on a server for global accessibility, leveraging artificial intelligence techniques, and implementing clustering approaches to further enhance annotation and understanding of cellular heterogeneity and functional characteristics.

### 2023-06-26 markdown/discussion.Rmd
anoops improvements implemented

@@
 # Discussion
-The developed application and the proteomics experiment pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
+The Proteomics Workbench interface and the ProteoScanR pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
 By utilizing an entropy-based visualization approach, the conservation of information can be validated, allowing users to select appropriate methods and adjust thresholds, cutoffs, and techniques accordingly.
@@
 This approach not only benefits scientists who are not proficient in programming, but it is also convenient for programmers.
-The developed user interface offers a visually appealing design, and the interactive menus enhance the user experience by eliminating the need to search for specific lines of code on a black screen.
+The Proteomics Workbench user interface offers a visually appealing design, and the interactive menus enhance the user experience by eliminating the need to search for specific lines of code on a black screen.
 Modifying parameters of functions becomes effortless, and data can be visualized without rerunning the entire script.
@@
 By allowing users to choose the method that preserves the maximum amount of information, the approach enables researchers to extract the most valuable insights from their experiments.
-Normalization methods play a crucial role in single-cell data analysis, and their selection depends on the nature of the data and the specific analysis objectives.
+Normalization methods play a crucial role in data analysis, and their selection depends on the nature of the data and the specific analysis objectives.
 The mutual information analysis revealed that quantile normalization exhibited the smallest loss of entropy compared to other normalization methods
@@
 
-Unfortunately, a substantial number of individual proteins in the dataset (totaling r nrows_dataset) have missing values.
-On average, this amounts to approximately r round(mean_missingness, 2)*100% of the proteins across all runs and channels.
-This situation leads to the need for missing value imputation using techniques such as k-means, which can result in averaged results.
+Unfortunately, a substantial number of individual proteins in the data set (totaling `r nrows_dataset`) have missing values.
+On average, this amounts to approximately `r round(mean_missingness, 2)*100`% of the proteins across all runs and channels.
+This situation leads to the need for missing value imputation using techniques such as k-nearest neighbor (K-NN), which can result in averaged results.
 Additionally, it is important to note that this imputation process can introduce potential loss of power in subsequent statistical analysis.
@@
 Here I will explain how the analysis could start and elaborate the limitations and challenges for this approach.
-Additional processing steps will be included after the existing pipeline, so the developed program provides a starting point for further analysis.
+Additional processing steps will be included after the existing pipeline, so ProteoScanR provides a starting point for further analysis.
 
@@
 While a development time of 5 months for the pipeline and user interface is sufficient to meet the analytical needs, there is always room for improvement in terms of user-friendliness.
-It is essential to consider the needs of all users, including the least experienced or "dumbest assumable user" (DAU).
+It is essential to consider the needs of all users, including the least experienced user.
 To prevent program crashes and errors, it may be necessary to implement try-catch around code blocks, as well as validate-need statements for the validation of user input.

### 2023-06-26 markdown/introduction.Rmd
anoops improvements implemented

@@
 Isobaric labeling solely applies to in vitro techniques, with TMT and iTRAQ.
-The data analyzed by the developed pipeline primarily utilized TMT tags.
+The data analyzed by ProteoScanR primarily utilized TMT tags.
 Tandem mass tag (TMT) reagents enable differentiation of multiple samples in a single mass spectrometry (MS) run.
@@
 Software packages like MaxQuant \citep{Cox2008} are commonly used to process the data, providing it for further analysis and statistical testing.
-Other software solutions include Protein Discoverer by Thermo Fisher.
+Other software solutions include Proteome Discoverer by Thermo Fisher.
 
@@
 To identify proteins, peptides are matched against a sequence database.
-Sequence databases are typically stored as .fasta files and can be downloaded from the UniProt website (www.uniprot.org).
+Sequence databases are typically stored as .fasta files and can be downloaded from the UniProt database (www.uniprot.org).
 The availability of gene sequences in databases and the computational matching of peptides against those sequences enable the identification of alterations in a sample at the protein level.
@@
 However, analyzing proteomic data poses challenges in data processing, statistical analysis, and interpretation.
-This thesis aims to address these challenges by exploring computational methods for downstream analysis of proteomic data.
-The methods employed in this study will be thoroughly explained in the methods section, with a specific focus on developing an interactive environment equipped with a graphical user interface.
+This thesis aims to address these challenges by exploring computational methods for downstream analysis of proteomics data.
+The methods employed in this study will be thoroughly explained in the methods section, with a specific focus on developing an interactive environment equipped with a graphical user interface, called Proteomics Workbench.
 This user-friendly interface will enable researchers, including those with limited computational expertise, to navigate and comprehend the data effectively.
 
-The analysis pipeline comprises a series of steps to streamline the data.
+ProteoScanR comprises a series of steps to streamline the data.
 It begins with quality control measures at the peptide spectrum match level, involving the removal of contaminants and irrelevant channels.

### 2023-06-26 markdown/methods.Rmd
anoops improvements implemented

@@
 ## Wetlab - Materials and Methods
-For testing and development of the pipeline the following materials and methods were performed. For the sample preparation protocol see appendix.
+For testing and development of the ProteoScanR pipeline the following materials and methods were performed. For the sample preparation protocol see appendix.
 
@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=16}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=17, fig.align='center', fig.cap="Flowchart presenting the processing by ProteoScanR"}
 library(Gmisc, quietly = TRUE)
@@
 batch_correction_opts <- moveBox(batch_correction_opts,
-                    x = .8,
+                    x = coords(output[[2]])$x,
                     y = coords(vert$missing_value_handling)$y)
@@
 ### Processing the data
-Further analysis is done with R and respective packages such as bioconductor. Please find version info in the appendix.
+Further analysis is done with R and respective bioconductor packages. Please find version info in the appendix.
 
@@
 This transformation was used as a reference when comparing methods.
-However, using the logarithm to the base 10 may be an easily interpretable way of defining expression data and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
+However, using the logarithm to the base 10 may be an easy way to recalculate expression data in the mind and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
 When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
@@
 Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly.
-The K-means clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances.
+The K-nearest neighbor (K-NN) clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances.
 This method will be applied in the missing value imputation, which will be the next step applied to the data.
@@
 In the analysis pipeline both procedures were applied on the dataset.
-However when the logarithm as the transformation method was chosen as transformation method, the pipeline automatically switches from a division to a subtraction, since:
+However when the logarithm as the transformation method was chosen as transformation method, ProteoScanR automatically switches from a division to a subtraction, since:
 $\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v$.
@@
 This variance would overshadow the biological differences across sample types when performing dimensionality reduction.
-A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of the developed pipeline.
+A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of ProteoScanR.
 It appears convenient to remove the batch effect before performing any statistical analysis, which is in turn prone to errors especially if the sample groups are not split equally between batches.
@@
 The most common approach in data science for any kind of high dimensional data is the PCA.
-For the pipeline the bioconductor package scater \citep{McCarthy2017} was used.
+For ProteoScanR the bioconductor package scater \citep{McCarthy2017} was used.
 Every principle component can be understood as a vector pointing in a specific direction derived as the eigenvector from the covariance matrix of the expression set.
@@
 The mutual information is calculated within each sample type, and it can also be calculated between a selection of sample types to determine if the computations performed affect the information content.
-After that all pairwise mutual information within the expression set before and after the pipeline is obtained.
+After that all pairwise mutual information within the expression set before and after ProteoScanR is obtained.
 The pairwise differences can be observed as a boxplot for a particular sample type indicating the change in dependence within and also compared to another sample type.
@@
 An algorithm within the program's backend logic is utilized to automatically generate this matrix according to the user's selection in the interface.
-The expression matrix contains intensities for each identified protein and is obtained at the end of the pipeline.
+The expression matrix contains intensities for each identified protein and is obtained at the end of ProteoScanR.
 By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors.
@@
 After differential expression the upcoming question could be, in which context the up or down regulated proteins are.
-Enriched proteins can be understood as an over-representation (ORA = over-representation analysis) within a specific set.
+Enriched pathways can be understood as an over-representation (ORA = over-representation analysis) of proteins within a specific set.
 This set can be a pathway or a selection of common denominators in biological contexts, such as cell types or oncological factors.
@@
 
-#### Pathway based protein enrichment analysis
-The pathway-based protein enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}.
-The analysis utilizes the “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways.
+#### Pathway enrichment analysis
+The pathway enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}.
+The analysis utilizes the entire “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways.
 

### 2023-06-26 markdown/results.Rmd
anoops improvements implemented

@@
 \newpage
-```{r ui_stat_HC_HIV_noMets, echo=F, out.width="90%", fig.align='center', fig.cap="Statistic module"}
+```{r ui_stat_HC_HIV_noMets, echo=F, out.width="70%", fig.align='center', fig.cap="Statistic module"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 
+\newpage
+
 By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group, six differentially expressed proteins were identified.
 The Venndiagram indicates significantly up/downregulated proteins.
-The Volcano represents log10(adjusted p values) against log10 fold changes (logFC).
+The Volcano represents -log10(adjusted p values) against log2 fold changes (logFC).
 These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
@@
 \newpage
-```{r ui_path_enrich_nw, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment. Networkplot indicating the interaction of differentially expressed proteins"}
+```{r ui_path_enrich_nw, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment. Networkplot indicating the association of differentially expressed proteins with pathways"}
 knitr::include_graphics("screenshots/path_enrich_nw.png")
@@
 The background for the statistical test can be chosen as genome wide (default) or as the detected proteins.
-When choosing the detected proteins as a background no protein could be considered as statistically significant.
+When choosing the detected proteins as a background no pathway could be considered as statistically significant.
 

### 2023-06-26 scpR-app/app.R

@@ -2108,9 +2108,9 @@ server <- function(input, output, session) {
     
     req(input$p_value_correction)
     if (input$p_value_correction == "none") {
-      p_val_correct <- "p-value"
+      p_val_correct <- "-log10(p-value)"
     } else {
-      p_val_correct <- "adj.p.value"
+      p_val_correct <- "-log10(adj.p.value)"
     }
     
     p <- ggplot(toptable, aes(x = logFC, y = -log10(adj.P.Val), text=protein)) +

### 2023-06-26 markdown/Thesis.Rmd
cite file added

@@
 {\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
-presenting ProteoScanR & Proteomics Workbench
+presenting ProteoScanR \& Proteomics Workbench
 
@@
 Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.
+https://github.com/Lukas67/ProteoScanR
 

### 2023-06-26 markdown/Thesis.tex
cite file added

--- 2023-05-09 markdown/Thesis.tex
@@
 \usepackage{float}
+\let\origfigure\figure
+\let\endorigfigure\endfigure
+\renewenvironment{figure}[1][2] { \expandafter\origfigure\expandafter[H] } { \endorigfigure }
 \usepackage{sectsty}
@@
 
-(the spacing is set to 1.5)
-
-no more than 250 words for the abstract
-
-\begin{itemize}
-\tightlist
-\item
-  a description of the research question/knowledge gap -- what we know
-  and what we don't know
-\item
-  how your research has attempted to fill this gap
-\item
-  a brief description of the methods
-\item
-  brief results
-\item
-  key conclusions that put the research into a larger context
-\end{itemize}
+With the growing prominence of single-cell techniques across various
+omics fields, there is a pressing need to develop a standardized
+pipeline for proteomics data in the realm of systems biology. Unlike DNA
+sequencing and RNA sequencing, proteomics analysis via mass spectrometry
+incurs high costs in terms of labor and equipment. Additionally,
+commercially available software solutions often come with hefty price
+tags and limited transparency regarding the underlying methods employed.
+However, MaxQuant \citep{Cox2008} presents a promising alternative,
+complemented by the flexibility using the R programming language.
+Introducing ProteoScanR, a state-of-the-art proteomics pipeline
+integrated into the user-friendly Proteomics Workbench interface. Guided
+by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the
+development process of ProteoScanR was thoroughly tested and validated
+using both bulk and single-cell mass spectrometry data sets. The
+pipeline implementation in the Proteomics Workbench leverages the power
+of an interactive environment built in R Shiny, empowering users to
+discover valuable insights for their specific data set. Within the
+interactive environment, users have the flexibility to customize cutoffs
+and thresholds for quality control, as well as employ various approaches
+for data transformation, normalization, missing value imputation, and
+batch correction. ProteoScanR and the Proteomics Workbench served as a
+valuable tool in guiding the identification of significantly expressed
+proteins in cells under study. Subsequently, the pathway enrichment
+analysis provided additional biological contexts for a comprehensive
+understanding of their functional implications. The master thesis
+project serves as a foundation for future advancements in the field of
+single-cell proteomics. Moreover, the codebase has been designed with
+robustness and scalability in mind, ensuring ease of maintenance and
+future expansion of the application.
+\url{https://github.com/Lukas67/ProteoScanR}
 
@@
 
-Thank you for following this tutorial!
-
-I hope you'll find it useful to write a very professional dissertation.
+In addition to my faithful feline companion, Apollo, who took care of my
+place in Austria while I was working on this project in Stockholm, I am
+deeply grateful to all the members of Ujjwal Neogi's research group.
+
+Their unwavering guidance and support throughout this project have been
+truly invaluable, and I am profoundly thankful for their expertise and
+assistance. \pagenumbering{arabic}
 
@@
 
+\hypertarget{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}{%
+\subsection{From the Central Dogma of Molecular Biology to Omics and
+Systems
+Biology}\label{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}}
+
+Before the initial exploration of molecular biology, the processes of
+life were roughly inferred and not well understood. The central dogma of
+molecular biology serves as the cornerstone of biological processes,
+providing a framework for understanding how genetic information is
+converted into functional proteins. This framework consists of two major
+processes: transcription, which converts DNA into RNA, and translation,
+which converts RNA into functional proteins. Between these processes,
+molecular-level regulation occurs through chemical adaptations. This
+regulation is crucial for biological organisms to efficiently produce
+their genetic products, known as proteins. Crick's formulation of the
+central dogma has guided research in molecular biology for over 60 years
+\citep{Cobb2017}. Over the past six decades, researchers have explored
+various aspects of this principle, collectively referred to as
+``omics,'' to comprehensively analyze biomolecules in diverse contexts.
+Omics research has emerged as an approach to study biomolecules and
+their interactions within biological systems comprehensively. Omics
+encompasses various fields, including genomics, transcriptomics,
+proteomics, and metabolomics, with each field focusing on a specific
+class of biomolecules. By analyzing multiple layers of omics data,
+researchers can gain a holistic view of biological processes, uncover
+regulatory mechanisms, and identify key players in complex biological
+networks using computational methods. Systems biology takes a
+quantitative approach to investigate the interactions, dynamics, and
+emergent properties of biomolecular networks. Biological systems are
+imagined as complex, multi-layer networks, with the ensemble of
+proteins, known as the proteome, playing a critical role in this
+interactive structure. Proteins have various functions, such as
+providing structural integrity, catalyzing chemical reactions, and
+regulating cellular functions \citep{Karahalil2016}
+
 \hypertarget{proteomics}{%
@@
 
-The proteome is referred to the sum of all proteins of a given sample at
-a given time. In the past several quantitative and qualitative assays
-were used to enlighten the protein composition of a sample.
+The central dogma of molecular biology offers insights not only at the
+level of individual terms but also provides a hierarchical understanding
+of biology. Hierarchical organization is the underlying principle in all
+topics of biology. Proteins exhibit hierarchical structures that
+contribute to their functionality. At the primary structure level,
+proteins are composed of smaller building blocks called amino acids.
+Combining them results in the protein secondary structure such as alpha
+helix and beta sheet. These secondary structures further assemble to
+create higher-order tertiary structures, which represent specific
+domains within the protein. Finally, the quaternary structure describes
+the functional state of the protein as a whole at a given time.
+Proteomics is a comprehensive approach that investigates the complete
+protein composition of a specimen, aiming to understand the intricate
+biological network it represents. For instance if pathogens interact
+with cells of an organism a cascade of biochemical reactions takes
+place. These reactions influence the outcome of the cellular proteome in
+regards of localization, abundance, post-translational modifications
+\citep{Beltran2017}. By studying the proteome, we can gain a
+comprehensive understanding of the hierarchical organization and
+functional dynamics of biological processes.
+
+\hypertarget{bulk-proteomics}{%
+\subsubsection{Bulk proteomics}\label{bulk-proteomics}}
+
+Higher organisms are composed of specialized cells organized into
+tissues, such as skin, muscle, and blood. Each tissue consists of cells
+with specific functions, resulting in variations in protein expression.
+Bulk proteomics is a technique used to analyze the protein composition
+of a tissue sample, which contains all types of cells present in that
+particular tissue. This approach finds valuable applications for
+instance in oncology, where it can provide insights into the protein
+profile of a tumor. By studying oncogenic biomarkers and expression
+patterns, this knowledge can contribute to the development of screening
+methods and the customization of treatments. Additionally, bulk
+proteomics can aid in identifying the progression of tumor development
+\citep{Kwon2021}. Bulk analysis expression profiles provide scientists
+with an average protein abundance across all cells in a sample, offering
+a comprehensive view of protein expression. Additionally, compared to
+single-cell methods, bulk samples typically contain a greater number of
+proteins. In fact the expanded library can be used to validate the
+performance of single-cell type and single-cell proteomics methods and
+bioinformatics pipeline \citep{Schoof2021}. However, in this type of
+analysis, the precise properties of individual cell types can only be
+inferred and remain obscured.
+
+\hypertarget{single-cell-type-proteomics}{%
+\subsubsection{Single-cell type
+proteomics}\label{single-cell-type-proteomics}}
+
+Taking tissue samples can lead to an averaging effect across the entire
+cellular ensemble, making it difficult to discern specific cell types.
+To overcome this limitation, cell sorting techniques such as
+fluorescence-activated cell sorting (FACS), magnetic-activated cell
+sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed
+\citep{Liou2015}. These techniques primarily rely on the detection of
+surface proteins on cells and the specific binding of antibodies to
+these proteins. Auto-fluorescence is observed in certain cells, wherein
+they possess the capability to emit a light signal when stimulated, and
+this emission can be used to sort them accordingly. However
+auto-fluorescence can lack of accuracy as shown in studies indicating
+the effect of laboratory gloves on cellular samples
+\citep{Philpott2017}. On the other side antibodies, typically derived
+from the immune system of other species, bind to antigens in a highly
+specific lock-and-key manner. Antibodies are widely used in molecular
+biology and find applications in cell isolation. Depending on the
+desired application, the antibodies can be labeled with fluorescent
+markers (FACS), magnetic beads (MACS), or biotin (BACS). These methods
+have varying technical requirements, with flow cytometers being the most
+complex and costly. In MACS, magnetic beads are used to isolate cells
+from a solution by utilizing a magnet. However, mechanical forces
+involved in MACS can potentially damage the cells. BACS, a method
+developed in 2015, employs microbubbles to carry the antibodies for
+cellular isolation. The method showed improvements compared to MACS in
+regards of viability after sorting. It is important to note that FACS is
+considered the gold standard for validating the effectiveness of MACS
+and BACS due to its high purity and yield. FACS was used in the
+validation of these two methods \citep{Liou2015, Sutermaster2019}.
+Single-cell type proteomics reflects the protein ensemble of a specific
+cell type at a particular time, providing a focused perspective on the
+studied field compared to bulk methods \citep{Maes2020}.
+
+\hypertarget{single-cell-proteomics}{%
+\subsubsection{Single-cell proteomics}\label{single-cell-proteomics}}
+
+Cell sorting techniques primarily rely on extracellular proteins to
+identify and separate different cell types. However, the intracellular
+content of cells can still exhibit variability in experiments.
+Single-cell proteomics (SCP) addresses this limitation by providing a
+snapshot of the underlying processes within an individual cell,
+eliminating this variability and allowing for the elucidation of the
+principles of cell type behavior, such as differentiation.
+Differentiation, a critical process in various biological systems such
+as the immune system, involves the acquisition of specialized functions
+by specific cell types \citep{Fang2018}. Investigating differentiation
+in multiple cells can be challenging due to the inherent heterogeneity
+during the initiation of this process. However, by analyzing one cell at
+a time and observing it throughout the differentiation process,
+researchers can capture the complete trajectory of differentiation at a
+single-cell level. This approach provides valuable insights into the
+dynamic changes occurring within cells during differentiation. In 2021,
+Specht et al.~developed the SCoPE2 sample preparation and analysis
+pipeline, performing macrophage-monocyte differentiation experiments
+\citep{Specht2021}. This pipeline streamlined the sample preparation
+process, reducing labor time and providing a reference model for
+subsequent SCP projects. SCP has proven to be a powerful tool for
+studying a wide range of biological phenomena, including drug responses,
+infectious diseases, and organism development. Over the past decade,
+single-cell techniques, including SCP, have become indispensable tools
+for the global research community \citep{Minakshi2019}.
+
+\hypertarget{capturing-the-proteome}{%
+\subsection{Capturing the proteome}\label{capturing-the-proteome}}
 
@@
 disadvantage of this technique was the limitation to only stain a few
-proteins per cell. For quantification procedures such as single-cell
-Western blots, immunoassays or CyTOF have been used. Other disadvantages
-are the ability to permeate cells, accessibility and binding of the
-epitope and the creation of specific antibodies for a given protein
-\citep{Budnik2018}.
-
-One of those techniques involved RNA-sequencing. Since RNA involves also
-non-coding RNA, the amount of RNA is often not proportional to the
-content of proteins in a cell. So the proteinaceous content of a cell
-was only predicted and quantitative analysis was not possible.
+proteins per cell and the requirement of a-priori knowledge of the
+protein to analyze. For quantification, procedures such as Western
+blots, immunoassays or cytometry by time of flight (CyTOF) have been
+used. Challenges are the ability to permeate cells, accessibility and
+binding of the epitope and the creation of specific antibodies for a
+given protein \citep{Budnik2018}. Early quantitative methods in
+molecular biology, measuring the entire protein concentration, were
+assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay.
+These methods are still in use to obtain total concentration for sample
+preparation. Another quantitative technique involved RNA-sequencing.
+RNAseq refers to the task of examining the transcriptome of bulk tissue
+sample with deep-sequencing technologies. RNA in a sample includes
+mRNAs, non-coding RNAs, and small RNAs. One of the main disadvantages
+is, that the mapping relies on the knowledge of the DNA sequence of the
+observed organism (reference genome) \citep{Wang2009}. Another
+disadvantage is, that the amount of protein translated from the
+determined mRNA can only be estimated \citep{Gygi1999}. However, today
+these predictions can be done with computational methods such as LASSO
+and the accuracy of the estimate highly depends on selection and
+validation of this method \citep{Magnusson2022}. When observing the
+protein ensemble of a cellular environment with a top-down approach like
+DNAseq or RNAseq, the characteristics (e.g.~post-translational
+modifications or charge) and local concentration of the proteins remain
+hidden. The cause of this phenomenon is that only a fraction of the DNA
+is transcribed into mRNA and not all mRNA is further translated into
+protein. Furthermore multiple proteins can be derived from a single
+DNA-sequence with alternative splicing and different proteins coming
+from non related sequences acting like building blocks can form protein
+complexes and/or Protein-RNA complexes. Cells are highly efficient and
+do not waste energy by producing obsolete proteins, moreover the
+degradation of proteins in the proteasome is an important energy
+recovery mechanism \citep{TANAKA2009}. A bottom-up strategy can
+elucidate these properties of the protein ensemble at a given time
+leading to proteomics. When studying impact of an external signal or
+stressor to a cell the abundance and properties of a protein or
+protein-ensemble can enlighten the underlying pathway leading to a
+biological response. Since some proteins do not even function before
+post-translational modifications (PTMs) happen, a knowledge about these
+can be crucial for further studies. Mass spectrometry (MS) fusions the
+advantages of the above mentioned techniques, because it combines the
+specificity of qualitative assays which are even able to sense (PTMs)
+with the quantitative accuracy of protein concentration assays.
+Furthermore MS does not require any a-priori knowledge of the protein
+ensemble of the specimen, making it the method of choice for untargeted
+approaches.
 
 \hypertarget{mass-spectrometry}{%
-\subsection{Mass Spectrometry}\label{mass-spectrometry}}
+\subsubsection{Mass Spectrometry}\label{mass-spectrometry}}
 
 Mass spectrometry enables qualitative and quantitative analysis of the
-entire repertoire of a biological sample. The availability of gene
-sequences in databases and the ability to match proteins against those
-sequences with computational methods makes it possible to identify
-alterations of a sample on a protein level. These alterations can rely
-on the sequence level or could be to post-translational modifications
-(PTMs) such as phosphorylation, methylation or else
-\citep{Aebersold2003}.
-
-Mass to charge ratio (m/z)
-
-\pagenumbering{arabic}
+entire repertoire of a biological sample. Mass spectrometers measure the
+mass to charge ratio (m/z) of a particle from a fragmented larger
+molecule. This is achieved by a physical procedure done with a device
+which is made of three major components: the ion source, an analyzer,
+and a detector. The ion sources charges molecules and accelerate them
+through a magnetic field. The analyzer separates particles according to
+their mass to charge ratio and the detector senses charged particles and
+amplifies their signal \citep{Parker2010}. However before acquiring any
+data, sample preparation and pre-processing is required. After sample
+preparation the first step involves tryptic digestion. It is a widely
+used technique in shotgun proteomics and involves the enzymatic cleavage
+of proteins into smaller peptides using the protease enzyme trypsin. The
+process known as proteolysis takes place in multiple parts of multi
+cellular organisms and was firstly detected in the small intestine
+digesting proteins \citep{Wang2008}. The enzyme cleaves specifically at
+the caroboxyl side of arginine and lysin residues. Afterwards a charge
+remains on the residues making them detectable for the mass
+spectrometer. It is proven that proteins which are digested to the
+limit, meaning no further proteolysis can be induced, provide the best
+results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.
+
+To observe multiple samples in one run, labeling is needed to identify
+each sample within a batch. Since mass spectrometry is not a
+quantitative technique by itself, the peak height or area does not
+reflect the abundance of a peptide. Physicochemical properties of the
+peptides can change the ionization efficiency and detectability of the
+target. However, when comparing the same analyte between multiple runs
+of labeled peptides, differences in the reporter ion intensity reflect
+the abundance of those. Labels should be chosen to change solely the
+mass of the sample and to not affect folding or other inherent
+properties of the peptide. The two techniques for labeling peptides are
+metabolic labeling and isobaric labling. In metabolic labeling the cells
+were fed with aminoacids containing heavy isotopes, it is the method of
+choice in order to label peptides at the earliest possible level. This
+atoms can be heavy nitrogen in aminoacids or salts in fertilizer for
+plants \citep{Karen2012}. Mass shifts are proportional to the isotopes
+incorporated during biomass production and are visible after proteolytic
+cleavage. Stable isotope labeling in cell culture (SILAC) was presented
+in the early 2000s. This method used heavy aminoacid enriched media to
+feed cells, in order to quantitatively analyze expression profiles. The
+limitation of this method is, that it can only be applied to a human
+cell culture (in vitro) or to model organisms (in vivo). Isobaric
+labeling solely applies to in vitro techniques, with TMT and iTRAQ. The
+data analyzed by ProteoScanR primarily utilized TMT tags. Tandem mass
+tag (TMT) reagents enable differentiation of multiple samples in a
+single mass spectrometry (MS) run. Each sample is individually labeled
+with TMT reagents and then pooled together, this procedure is called
+multiplexing. TMTs have the same charge and differ only by their
+isotopic masses, the peaks observed for each sample are called reporter
+ions (RI). Each RI and sample is interpreted as one channel in
+downstream analysis. This technique allows for the identification,
+enrichment, and quantification of low-abundance peptide ions, which is
+particularly valuable in single-cell techniques. With this technique it
+is possible to quantify proteins and differ low abundant proteins from
+background noise. However, a drawback of isobaric labeling is the
+presence of co-fragmentation signals in the spectrogram, which requires
+data normalization to remove unwanted contributions
+\citep{Marx2019, Budnik2018}. Isotopic distribution in TMTs, reflecting
+the natural distribution, can be corrected during data acquisition by
+considering a defined spread in other channels. Single-cell proteomics
+by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem
+mass tag (TMT) to enhance signal intensity of a protein species. Loading
+of a single-cell channel (SCC) with the right amount can be balanced to
+the protein concentration of a single cell \citep{Ye2022}.
+
+Before ionization peptides need to be further separated according to
+their chemical properties, size or species by liquid chromatography
+(LC), because the number of peptide processed simultaneously by the MS
+device is limited. In the case of mass spectrometry, liquid
+chromatography especially high performance LC (HPLC) is the technique of
+choice and replaces the instrumentation-wise simpler technique gel
+electrophoresis. In HPLC the molecules eluate through a narrow column
+with high pressures (50-350 bar), compared to LC where the separation is
+gravity dependent. Similar to gel electrophoresis molecules can be
+separated upon different chemical properties, depending on the type of
+the chromatography column. In a typical proteomics experiment peptides
+are separated with reversed phase chromatography (RPLC). The principle
+behind RPLC is a column with a nonpolar stationary phase which interacts
+with the nonpolar residues of the peptides. Peptides were fractionated
+according to their hydrophobicity \citep{Pitt2009}.
+
+To analyze a biological sample consisting of peptides in solution the
+liquid needs to be vaporized into gas phase. Two techniques are capable
+of this procedure. Electrospray ionization (ESI) pushes the analyte
+through a capillary while applying an electric current to the liquid,
+vaporizing the sample to a charged aerosol. Peptides are further
+fragmented according to their chemical properties and can be further
+handled in the mass spectrometer. The fragmented peptides are now in
+charged droplets separated based on their surface charge, splitting
+further into smaller droplets until they reach the gas phase as ions.
+Two physical models describe the process from gas phase to ion called
+``The ion evaporation model'' (IEM) and ``The charge residue model''
+(CRM). In the ion evaporation model the droplets shrink by evaporation
+until ions are expelled \citep{Iribarne1976}. The model had its
+limitation by explaining same evaporation rate constant among ions with
+different chemical properties. In the charge residue model the
+assumption of one molecule per droplet leads to an ionization rate
+constant, which is independent of the ion itself and relies solely on
+the generation of the droplet and the efficiency of the solvent
+\citep{Wilm2011}.
+
+Subsequently after the ionization peptides are accelerated through the
+magnetic field generated by the mass analyzer. Six general types of mass
+analyzers are availaible on the market, there are Quadropole, time of
+flight (TOF), magnetic sector, electrostatic sector, quadropole ion trap
+and ion cyclotron resonance -mass analyzer. The function of the mass
+analyzer is to separate ions based on their mass to charge ratio (m/z).
+
+Before the signal of the separated ions can be obtained an optional
+subsequent step involves trapping the ions an electric and/or magnetic
+field in order to detect ions based on their (m/z) consecutively. Ion
+trapping enables the technique of coupled mass spectrometers, where ions
+were further fragmented by collision induced dissociation and analyzed
+subsequently. The detector transforms the impact energy of the ion into
+an electric signal, called spectrum which can be processed
+computationally.
+
+\hypertarget{data-processing}{%
+\subsubsection{Data processing}\label{data-processing}}
+
+Given the high resolution of MS data, algorithms are employed to convert
+the raw signal into an interpretable form. Software packages like
+MaxQuant \citep{Cox2008} are commonly used to process the data,
+providing it for further analysis and statistical testing. Other
+software solutions include Proteome Discoverer by Thermo Fisher.
+
+The data obtained from mass spectrometry has three dimensions: m/z
+ratio, intensity, and retention time. To separate peaks from each other,
+algorithms are used to identify local minima in the data. The centroid
+of each peak is determined by fitting a Gaussian peak shape, which can
+be interpreted as locating the peaks of each m/z spectrum as a function
+of retention time. The centroid of a peak corresponds to an isotope.
+
+In order to decipher the isotopic distribution of a biomolecule,
+MaxQuant employs a vertex-based approach. This process, referred to as
+de-isotoping, utilizes graph theory. It creates a vertex for each
+individual peak and connects them with their potential isotopic
+counterparts. This is achieved by determining the mass proportion of an
+average amino acid to its respective isotope using a concept known as
+averagine \citep{Senko1995}. By applying this procedure, the number of
+data points is reduced by a factor of ten, and each peak represents a
+small biomolecule.
+
+The subsequent step in data acquisition involves the detection of labels
+for quantification purposes. Isotopic pairs of the label (such as N13,
+N14, N15) present in the tag or aminoacid are identified by convolving
+the two measured isotope patterns with the theoretical isotope patterns.
+Through an iterative process using a least-squares method, the best fit
+is determined, enabling the identification of the specific channel or
+sample.
+
+The intensity-weighted average of the MS peak centroids, as described as
+3D peak identification, corresponds to the mass of the peptide. The
+accuracy of mass measurements relies on the specific analyzer used. For
+an Orbitrap-type analyzer (which was used during this project), MaxQuant
+applies a correction value of 1 ppm (parts per million). Autocorrelation
+between centroids is accounted for by considering only well-identified
+peptides. Published data indicates that the mass precision within an MS
+experiment typically ranges around 10\^{}-7 \citep{Wilschefski2019}.
+
+Each peptide is represented by its individual fingerprint in the MS
+spectrum, which is based on the sequence and modifications of amino
+acids. The m/z ratio of the amino acids can be calculated, and the
+resulting peaks are interpreted as an amino acid sequence. However, only
+peptide fragments are visible in the spectrum due to peptide
+fragmentation. To identify proteins, peptides are matched against a
+sequence database. Sequence databases are typically stored as .fasta
+files and can be downloaded from the UniProt database (www.uniprot.org).
+The availability of gene sequences in databases and the computational
+matching of peptides against those sequences enable the identification
+of alterations in a sample at the protein level. These alterations can
+rely on the sequence level or could be to post-translational
+modifications (PTMs) such as phosphorylations or methylations
+\citep{Aebersold2003}. The peptide identification (P-) score reflects
+the goodness of fit between the data and the identified sequence in the
+database, taking into account the length of the peptide. This score is
+used to calculate the posterior error probability, which is then
+utilized to estimate the false discovery rate (FDR). The FDR calculation
+provides a measure of the rate at which false identifications are
+expected in the dataset. Within the identified peptide groups, a ``razor
+protein'' refers to the protein with the highest number of ambiguously
+identified peptides. Quantification is performed by considering only
+unique peptides for comparison. Posterior error probabilities, which
+represent the likelihood of the identified peptide being a random event,
+are multiplied together, and only distinct sequences with the highest
+scores are considered for further analysis. Various metrics that
+indicate the performance of the peptide search can be utilized in
+downstream analysis.
+
+\hypertarget{downstream-analysis}{%
+\subsubsection{Downstream analysis}\label{downstream-analysis}}
+
+With advancements in mass spectrometry technology and computational
+methods, proteomic analysis has emerged as a powerful tool for
+investigating complex protein samples. However, analyzing proteomic data
+poses challenges in data processing, statistical analysis, and
+interpretation. This thesis aims to address these challenges by
+exploring computational methods for downstream analysis of proteomics
+data. The methods employed in this study will be thoroughly explained in
+the methods section, with a specific focus on developing an interactive
+environment equipped with a graphical user interface, called Proteomics
+Workbench. This user-friendly interface will enable researchers,
+including those with limited computational expertise, to navigate and
+comprehend the data effectively.
+
+ProteoScanR comprises a series of steps to streamline the data. It
+begins with quality control measures at the peptide spectrum match
+level, involving the removal of contaminants and irrelevant channels.
+Significance levels of peptide spectrum matches serve as confidence
+thresholds for match selection. The data is then aggregated at the
+peptide level, followed by additional filtering procedures based on
+metrics such as reporter ion intensity, peptide coefficient of variation
+over razor proteins, and missing rates for peptides.
+
+To derive meaningful biological insights from the dataset, statistical
+testing plays a crucial role. Selecting appropriate statistical tests
+and ensuring proper data transformation and normalization are essential
+steps. Many statistical methods rely on data distribution and require
+careful consideration to preserve the biological interpretation of the
+results. This thesis employs various transformation and normalization
+techniques, aiming to maintain data integrity and avoid biased
+interpretations. Additionally, an entropy-based visualization approach
+will provide users with a valuable tool for validating the computations
+and selected thresholds for the filtering steps.
+
+Handling missing values poses further challenges in the pipeline
+development since a missing signal for a peptide results in a missing
+protein, which is highly unlikely in a biological context. Several
+options for missing value handling will be included and thoroughly
+explained in the methods section.
+
+Dimensionality reduction and visualization techniques are widely used to
+cluster data. In this project, multiple approaches will be utilized, and
+users will have the ability to visually and interactively observe the
+data.
+
+It is important to acknowledge that downstream analysis in proteomics
+lacks a standardized approach, and the selection of computational
+methods depends on the specific dataset and research objectives.
+Furthermore, statistical analysis and the interpretation of biological
+meaning often present misconceptions that need to be addressed. For
+instance, batch correction, a commonly practiced method, has been
+overestimated in numerous studies, leading to misleading results
+\citep{Nygaard2016}. This thesis aims to contribute to the field by
+providing researchers with a comprehensive understanding of
+computational methods in proteomics analysis and their application in an
+interactive environment.
 
@@
 
-\hypertarget{materials}{%
-\subsection{Materials}\label{materials}}
-
-For analysis two types of cells were used. One type is the Jurkat-based
-cell line (J-lat) with integrated HIV.
-
-The other type of cells are macrophages with a sample size of 72 cells.
-The analysis is done with two groups. A HIV negative (HIV-) control
-group and a HIV positive (HIV+) group.
-
-\hypertarget{cell-isolation}{%
-\subsubsection{Cell Isolation}\label{cell-isolation}}
-
-\hypertarget{lysis}{%
-\subsubsection{Lysis}\label{lysis}}
+\hypertarget{wetlab---materials-and-methods}{%
+\subsection{Wetlab - Materials and
+Methods}\label{wetlab---materials-and-methods}}
+
+For testing and development of the ProteoScanR pipeline the following
+materials and methods were performed. For the sample preparation
+protocol see appendix.
+
+\hypertarget{celltypes}{%
+\subsubsection{Celltypes}\label{celltypes}}
+
+Macrophages from 12 healthy donors and 12 HIV patients with metabolic
+syndrome and 12 HIV patients without metabolic syndrome. Cell sorting
+was performed with FACSAria Fusion (BD Biosciences).
 
@@
 
-\hypertarget{labeling-techniques}{%
-\subsubsection{Labeling techniques}\label{labeling-techniques}}
-
-For differential analysis proteins need to be labeled to compare mass to
-charge intensities in order to quantify observed peptides. Since mass
-spectrometry is not a quantitative technique by itself, the peak height
-or area does not reflect the abundance of a peptide. Physicochemical
-properties of the proteins can change the ionization efficiency and
-detectability of the target. However, when comparing the same analyte
-between multiple runs of labeled proteins, differences in the mass
-spectrum reflect the abundance of those. Labels should be chosen to
-change solely the mass of the sample and to not affect folding or other
-inherent properties of the protein.
-
-\hypertarget{metabolic-labeling}{%
-\paragraph{Metabolic labeling}\label{metabolic-labeling}}
-
-Feeding cells with aminoacids containing heavy isotopes, is the method
-of choice in order to label peptides at the earliest possible level.
-This atoms can be heavy nitrogen in aminoacids or salts in fertilizer
-for plants. Mass shifts are proportional to the isotopes incorporated
-during biomass production and are visible after proteolytic cleavage.
-Stable isotope labeling in cell culture (SILAC) was presented in the
-early 2000s. This method used heavy aminoacid enriched media to feed
-cells, in order to quantitatively analyze expression profiles.
-
-\hypertarget{isobaric-labeling}{%
-\paragraph{Isobaric labeling}\label{isobaric-labeling}}
-
-\hypertarget{tandem-mass-tag-tmt}{%
-\subparagraph{Tandem mass tag (TMT)}\label{tandem-mass-tag-tmt}}
-
-Tandem mass tag (TMT) reagents enable to differentiate multiple samples
-analyzing in one MS run. The samples are labeled individually and pooled
-afterwards, this procedure is called multiplexing. TMTs have the same
-charge and differ only by their isotopic masses, the peaks found for
-each sample are called reporter ions (RI). Each RI and sample is
-interpreted as one channel in downstream analysis. The identification of
-these RI leads to an enrichment and identification of low abundance
-peptide ions which is common especially in single-cell techniques. With
-this technique it is possible to quantify proteins and differ low
-abundant proteins from background noise. The disadvantage of isobaric
-labeling is, that the co-fragmentation signals can be observed in the
-spectrogram and the data needs to be normalized in order to remove
-unwanted contribution \citep{Marx2019, Budnik2018}. Furthermore TMTs
-have an isotopic distribution according to the distribution found in
-nature. This can be corrected during data-acquisition as a defined
-spread in other channels.
-
-\hypertarget{instrumentation}{%
-\subsubsection{Instrumentation}\label{instrumentation}}
+Protein digestion was performed using trypsin (Promega, Madison, WA)
+buffered in Triethylammonium bicarbonate buffer (TEAB). Reagents were
+handled with MANTIS automatic dispenser (Formulatrix, Bedford, MA).
+
+\hypertarget{labeling}{%
+\subsubsection{Labeling}\label{labeling}}
+
+Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher)
+in anhydrous acetonitril (ACN) solution.
+
+\hypertarget{pooling}{%
+\subsubsection{Pooling}\label{pooling}}
+
+Carrier proteome was added to the 36 TMTpro sets before pooling sample
+types for each of the 36 MS-runs.
 
 \hypertarget{liquid-chromatography}{%
-\paragraph{Liquid chromatography}\label{liquid-chromatography}}
-
-In order to separate proteins according to their chemical properties,
+\subsubsection{Liquid chromatography}\label{liquid-chromatography}}
+
+In order to separate peptides according to their chemical properties,
 size or species a liquid chromatography (LC) is recommended before
-ionization.
+ionization. For the reference data used for developing the pipeline the
+following intrumentation was used. EASY-Spray™ HPLC column, ES802A, 2
+\(\mu\)m × 75 \(\mu\)m ID × 250 mm (ThermoFisher Scientific) Acclaim
+PepMap trap column, 2 \(\mu\)m × 75 \(\mu\)m ID × 2 cm (ThermoFisher
+Scientific)
 
 \hypertarget{mass-spectrometry-1}{%
-\paragraph{Mass Spectrometry}\label{mass-spectrometry-1}}
+\subsubsection{Mass Spectrometry}\label{mass-spectrometry-1}}
 
 \hypertarget{ionization}{%
-\subparagraph{Ionization}\label{ionization}}
-
-In order to analyze a biological sample consisting of proteins in
-solution the liquid needs to be vaporized into gas phase. Two techniques
-are capable of this procedure. Electrospray ionization (ESI) pushes the
-analyte through a capillary and applies an electric current to the
-liquid, vaporizing the sample to a charged aerosol. Biomolecules are
-fragmented according to their chemical properties and can be further
-handled in the mass spectrometer. The fragmented biomolecules are now in
-charged droplets separated by their charge on the surface, splitting
-further into smaller droplets until they become a gas phase ion. Two
-physical models describe the process from gas phase to ion called ``The
-ion evaporation model'' (IEM) and ``The charge residue model'' (CRM). In
-the ion evaporation model (by Iribarne and Thomoson) the droplets shrink
-by evaporation until ions are expelled. The model had its limitation by
-explaining same evaporation rate constant among ions with different
-chemical properties. In the charge residue model the assumption of one
-molecule per droplet leads to an ionization rate constant, which is
-independent of the ion itself and relies solely on the generation of the
-droplet and the efficiency of the solvent \citep{Wilm2011}.
-
-Matrix-assisted laser desorption/ionization (MALDI)
-
-\hypertarget{ms.1}{%
-\subparagraph{MS.1}\label{ms.1}}
-
-\hypertarget{coupled-mass-spectrometry-msms-ms.2}{%
-\subparagraph{Coupled mass-spectrometry (MS/MS) \&
-MS.2}\label{coupled-mass-spectrometry-msms-ms.2}}
-
-In order to enhance sequence identification, two MS devices are built in
-series. In the first run (MS1) the m/z is determined and the molecules
-are passed to the next device. Upon passing the molecules are fragmented
-into smaller ions and analyzed by the second MS. The fragmentation
-highly depends on the chemical bonds found in the molecule. The majority
-of these breaks occur on the peptide bond of the protein, although this
-is not guaranteed for all bonds and so it can happen that certain
-peptide ions have a low abundance \citep{Budnik2018}. These low abundant
-peptides will not be detected, hence the problem needs to be faced with
-another approach . A solution for this problem is molecular barcoding
-with labeling mentioned in the chapter labeling.
-
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-\hypertarget{acquisition}{%
-\subsubsection{Acquisition}\label{acquisition}}
-
-Acquisition of the data was done with MaxQuant \citep{Cox2008} software
-package.
-
-\hypertarget{ms-spectrum}{%
-\paragraph{MS-Spectrum}\label{ms-spectrum}}
-
-Each peptide is reflected by its` indivual fingerprint in the
-ms-spectrum. The fingerprint is based on the chemical properties and
-modifications of aminoacids. These aminoacids can be calculated through
-their m/z ration and after that interpreted as an aminoacid sequence.
-Due to fragmentation of the protein only peptide sequences are visible
-in the spectrum. In order to identify proteins, peptides are matched
-against a sequence database \citep{Cox2008}. Sequence Databases are
-simple .fasta files, which can be downloaded on the uniprot webpage
-(www.uniprot.org).
-
-Since ms data has a high resolution, algorithms are used to convert the
-raw signal to an interpretable form. MaxQuant is one of many sofware
-packages to process the data and provides it for further analysis and
-statistical testing. Other software solutions are Protein Discoverer
-Thermo Fisher or even packages for R. In this publication we will mainly
-focus on the data-acquisition with MaxQuant \citep{Cox2008}.
-
-\hypertarget{three-dimensional-peak-detection}{%
-\paragraph{Three-dimensional peak
-detection}\label{three-dimensional-peak-detection}}
-
-The three dimensions of the data are: m/z ratio, intensity and retention
-time. The algorithm finds local minima of the function in order to
-seperate peaks from each other. The centroid of the peak is detected by
-fitting a so called gaussian peak shape fitting. This can be interpreted
-as finding the peaks of each m/z spectrum as a function of time.. The
-centroid of the peak refers to an isotope.
-
-\hypertarget{deisotoping}{%
-\paragraph{Deisotoping}\label{deisotoping}}
-
-To decrypt the istopic distribution of a biomolecule, MaxQuant creates a
-vertex of every single peak and connects them with their possible
-isotopic counterparts by finding the proportion of mass of an average
-aminoacid to its` respective isotope (averagine \citep{Senko1995}).
-Isotoping is the term of such procedure and it is enabled with graph
-theory. After this procedure the amount of data points are reduced by a
-tenfold and a single peak reflects a small biomolecule.
-
-\hypertarget{label-detection}{%
-\paragraph{Label detection}\label{label-detection}}
-
-The next step in data-acquisition is the detection of labels for
-quantification. Isotopic pairs of the label (e.g.~N13, N14, N15)
-contained in the tag or aminoacid can be identified by convoluting the
-two measured isotope patterns with the theoretical isotope patterns.
-With a least-square method the best fit is found iteratively and the
-channel/sample can be identified.
-
-\hypertarget{improving-peptide-mass-accuracy}{%
-\paragraph{Improving peptide mass
-accuracy}\label{improving-peptide-mass-accuracy}}
-
-The intensity-weighted average of the ms peak centroids (as described in
-the 3D peak identification) refers to the mass of the peptide.
-Corrections highly depend on the analyzer, MaxQuant uses for an orbitrap
-typed analyzer a correction value of 1ppm. Autocorrelation between
-centroids is compensated by only using well-identified peptides. As
-published the mass precision within a ms experiment ranges around 10e-7.
-
-\hypertarget{peptide-search}{%
-\paragraph{Peptide search}\label{peptide-search}}
-
-Biomolecules can now be searched in a database in forward and reverse
-direction. The peptide identification (P-) score indicates the fit of
-the data to the found sequence in the database according to the length
-of the peptide and is used to calculate the posterior error probability
-(FDR). The calculation of the false discovery rate is then calculated by
-taking FDR into contrast.
-
-\hypertarget{protein-assembly}{%
-\paragraph{Protein assembly}\label{protein-assembly}}
-
-After these calculation the identified peptides can be aggregated
-according to it`s respective protein and quantified. The mentioned
-metrics indicating the performance of the peptide search can be used in
-downstream analysis. A so called razor peptide indicates the group with
-the highest number of identified peptides. Quantification is enabled by
-taking only unique peptides into contrast. Posterior error
-probabilities, which refers to the chance that the found peptide is a
-random event, are multiplied and only distinct sequences with the
-highest-scoring are used.
-
-\hypertarget{data-processing}{%
-\subsubsection{Data processing}\label{data-processing}}
-
-Further analysis is done with R and respective packages such as
-bioconductor. Since there is no state of the art established, analysis
-varies upon experimental design. The workflow of the analysis will be
-processed in a so called pipeline streamlining the data through steps
-where individual results can be observed in visualizations and indivdual
-calculations will be adapted according to user demands and experimental
-properties.
-
-\includegraphics{Thesis_files/figure-latex/data_processing_pipeline_flowchart_vertical-1.pdf}
-
-\hypertarget{reading-the-data}{%
-\paragraph{Reading the data}\label{reading-the-data}}
+\paragraph{Ionization}\label{ionization}}
+
+For the reference data ESI ionization was used.
+
+\hypertarget{mass-spectrometer}{%
+\paragraph{Mass spectrometer}\label{mass-spectrometer}}
+
+For the reference data used for developing the pipeline an Orbitrap™
+Fusion™ Lumos™ tribrid mass spectrometer (ThermoFisher Scientific) was
+used. The instrumentation uses two subsequent runs in the analyzer
+(=Orbitrap) to increase mass accuracy.
+
+\hypertarget{drylab---computational-methods}{%
+\subsection{Drylab - Computational
+methods}\label{drylab---computational-methods}}
+
+\hypertarget{hardware}{%
+\subsubsection{Hardware}\label{hardware}}
+
+Lenovo Laptop X360 with: Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz 8 GB
+RAM
+
+\hypertarget{data-acquisition-and-raw-file-analysis}{%
+\subsubsection{Data-acquisition and raw-file
+analysis}\label{data-acquisition-and-raw-file-analysis}}
+
+For the data-acquisition Thermo Fisher acquisition software was used.
+Raw-file analysis of the data was done with MaxQuantCmd 2.2.0.0
+\citep{Cox2008}. The default settings were used. TMTpro labels were
+corrected according to the isotopical distribution of the Lot-number
+provided by Thermo Fisher.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics{Thesis_files/figure-latex/data_processing_pipeline_flowchart_vertical-1}
+
+}
+
+\caption{Flowchart presenting the processing by ProteoScanR}\label{fig:data_processing_pipeline_flowchart_vertical}
+ \endfigure\egroup
+
+\hypertarget{processing-the-data}{%
+\subsubsection{Processing the data}\label{processing-the-data}}
+
+Further analysis is done with R and respective bioconductor packages.
+Please find version info in the appendix.
 
 After processing MaxQuant creates a directory containing all results as
-.txt file. The evidence.txt file include all peptide to spectrum matches
-(PSM) with their respective proteins and statistical parameters.
+.txt file. The evidence.txt file includes all peptide to spectrum
+matches (PSM) with their respective proteins and statistical parameters.
 
@@
   Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Mass
-  \end{itemize}
 \item
@@
   Precursor Ion Fragment
-
-  \begin{itemize}
-  \tightlist
-  \item
-    source of the detected ion also referred as mother ion
-  \end{itemize}
 \item
@@
   Reporter intensity (RI)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    corrected RI
-  \end{itemize}
 \item
@@
 \hypertarget{object-oriented-programming}{%
-\paragraph{Object oriented
+\subsubsection{Object oriented
 programming}\label{object-oriented-programming}}
@@
 In order to streamline the analysis of multiple experiments, object
-oriented programming can be applied. The approach in R is to create a so
+oriented programming was applied. The approach in R is to create a so
 called Q-feature object, which contains all variables and metadata in a
@@
 \hypertarget{zero-values}{%
-\paragraph{Zero values}\label{zero-values}}
+\subsubsection{Zero values}\label{zero-values}}
 
@@
 applicable'' (NA) is crucial for understanding and interpreting MS data.
+NA values will be managed in missing value handling later on the protein
+level.
 
 \hypertarget{exclude-reverse-matchescontaminants}{%
-\paragraph{Exclude reverse
+\subsubsection{Exclude reverse
 matches/contaminants}\label{exclude-reverse-matchescontaminants}}
@@
 Peptide sequences matching to the reverse protein sequences (=decoy
-database) are considered as possible contaminants. These matches can be
+database) are considered as possible contaminants. These matches will be
 excluded from further analysis.
@@
 \hypertarget{filter-according-to-precursor-ion-fraction-pif}{%
-\paragraph{Filter according to precursor ion fraction
+\subsubsection{Filter according to precursor ion fraction
 (PIF)}\label{filter-according-to-precursor-ion-fraction-pif}}
@@
 or parental ions). Contaminant peptides can co-migrate in this process
-and can be distinguished by the lower fraction of their respective
+and were distinguished by the lower fraction of their respective
 precursor ions \citep{Tannous2020}. These peptides need to be filtered
@@
 SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface,
-but it can be adjusted according to the needs of the biologist.
+but it can be adjusted.
 
 \hypertarget{filter-by-q-value}{%
-\paragraph{Filter by q-value}\label{filter-by-q-value}}
+\subsubsection{Filter by q-value}\label{filter-by-q-value}}
 
@@
 peptide-spectrum matching score. The peptides spectrum matching score is
-defined as -10log10(p). Whereas the p-value is defined that the hit is
-done by chance. The calculation of the score is highly dependent on the
-data acquisition method used. MaxQuant uses Andromeda , an integrated
+defined as \(-10 * log(p)\). Whereas the p-value is defined that the hit
+is done by chance. The calculation of the score is highly dependent on
+the data acquisition method used. MaxQuant uses Andromeda, an integrated
 search engine. Proteome Discoverer from Thermo Fisher utilizes different
@@
 \hypertarget{peptide-spectrum-match-psm-aggregation-to-peptides}{%
-\paragraph{Peptide spectrum match (PSM) aggregation to
+\subsubsection{Peptide spectrum match (PSM) aggregation to
 peptides}\label{peptide-spectrum-match-psm-aggregation-to-peptides}}
@@
 data based on a particular column using a specific function. In the
-context of processing from peptides to spectrum matches, the desired
-column is the peptide sequence. To account for different distributions
-across multiple assays, the median of the channel is used as the
-function to aggregate multiple matches into one.
-
-\hypertarget{join-assays-when-observing-multiple-comparable-batches-at-once}{%
-\paragraph{Join assays when observing multiple comparable batches at
-once}\label{join-assays-when-observing-multiple-comparable-batches-at-once}}
+context of processing from PSMs to unique peptides, the desired column
+is the PSM sequence. To account for different distributions across
+multiple assays, the median of the channel is used as the function to
+aggregate multiple matches into one.
+
+\hypertarget{join-assays}{%
+\subsubsection{Join assays}\label{join-assays}}
 
@@
 simultaneously, allowing for testing of multiple batches and increasing
-the number of samples that can be included in the analysis.
-
-\hypertarget{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}{%
-\paragraph{Calculate reporter ion intensity (RI) and filter according to
-median
-RI}\label{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}}
+the number of samples that can be included in the analysis. For further
+processing the runs are joined in one object at the peptide level.
+
+\hypertarget{calculate-median-reporter-ion-intensity-ri-and-filter}{%
+\subsubsection{Calculate median reporter ion intensity (RI) and
+filter}\label{calculate-median-reporter-ion-intensity-ri-and-filter}}
 
@@
 other type. Another one could be a spillage of TMT detection in other
-channels due incorrect or missing correction of the TMT isotopes.
-
-\hypertarget{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}{%
-\paragraph{Calculate and filter according to median coefficient of
-variation (CV) per
-cell/channel}\label{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}}
+channels due incorrect or missing correction of the TMT isotopes. The
+correction values can be found on the vendor`s webpage (in this case
+Thermo Fisher) and correction needs to be set up in MaxQuant.
+
+\hypertarget{calculate-median-coefficient-of-variation-cv-and-filter}{%
+\subsubsection{Calculate median coefficient of variation (CV) and
+filter}\label{calculate-median-coefficient-of-variation-cv-and-filter}}
 
@@
 \hypertarget{remove-peptides-with-high-missing-rate}{%
-\paragraph{Remove peptides with high missing
+\subsubsection{Remove peptides with high missing
 rate}\label{remove-peptides-with-high-missing-rate}}
@@
 However, the threshold of missingness (described as a fraction of the
-row) can be set in the user interface and adjusted to enable different
-experimental designs.
+row) can be set in the user interface and adjusted.
 
 \hypertarget{aggregation-of-peptides-to-proteins}{%
-\paragraph{Aggregation of peptides to
+\subsubsection{Aggregation of peptides to
 proteins}\label{aggregation-of-peptides-to-proteins}}
@@
 matrix for every protein and their respective channel column is
-returned. Each channel refers to a sample and reflects contains the
-intensities of each protein found in the biological specimen.
+returned. Each channel refers to a sample and contains the intensities
+of each protein found in the biological specimen.
 
 \hypertarget{transformation-of-protein-expression-data}{%
-\paragraph{Transformation of protein expression
+\subsubsection{Transformation of protein expression
 data}\label{transformation-of-protein-expression-data}}
@@
 dataset. This transformation was used as a reference when comparing
-methods. However, using the logarithm to the base 10 may be an easily
-interpretable way of defining expression data and will also be
-facilitated by other transformation methods such as the boxcox method,
-which is also implemented in the application. When observing a wide
-distribution of small and large values of the expression set in a
-histogram, a square root transformation can help increase the
-variability of smaller values and decrease the variability of larger
-values. Practically impossible, but in data-driven science, occasionally
-negative expression values may be present, and they can be converted
-into positive values by taking each one to the power of two. Depending
-on the distribution, the user has to decide which transformation to
-consider and verify that statistical assumptions are met after
-application with visualizations such as histograms, qqplots, and
-MA-plots. For statistically inexperienced users, the boxcox
-transformation can help with this decision. Developed in 1964 by Box and
-Cox \citep{Sakia1992}, this method applies a linear model against 1 on
-the data to determine the statistical parameter lambda by the maximum
-likelihood method. The log-likelihood method takes the 95\% confidence
-interval for the parameter lambda and the final lambda is chosen as the
-value with the highest log-likelihood value.
+methods. However, using the logarithm to the base 10 may be an easy way
+to recalculate expression data in the mind and will also be facilitated
+by other transformation methods such as the boxcox method, which is also
+implemented in the application. When observing a wide distribution of
+small and large values of the expression set in a histogram, a square
+root transformation can help increase the variability of smaller values
+and decrease the variability of larger values. Depending on the
+distribution, the user has to decide which transformation to consider
+and verify that statistical assumptions are met after application with
+visualizations such as histograms, qq-plots, and MA-plots. For
+statistically inexperienced users, the boxcox transformation can help
+with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992},
+this method applies a linear model against 1 on the data to determine
+the statistical parameter lambda by the maximum likelihood method. The
+log-likelihood method takes the 95\% confidence interval for the
+parameter lambda and the final lambda is chosen as the value with the
+highest log-likelihood value.
 
@@
 \hypertarget{normalization}{%
-\paragraph{Normalization}\label{normalization}}
+\subsubsection{Normalization}\label{normalization}}
 
@@
 Normalization is considered as part of the scaling procedure and ensures
-that every feature contributes equally to our statistical model and
+that every feature contributes equally to the statistical model and
 hinders large values from biasing the model in a particular direction.
@@
 the method of choice for normalization can change the results
-significantly. The K-means clustering algorithm, for instance relies on
-the distances between distinct data points by minimizing variance of the
-squared euclidean distances. This method will be applied in the missing
-value imputation which will be the next step applied to our data.
-Therefore, we want to ensure that we obtain the most accurate imputed
-values possible. The first normalization method available in the user
-interface is column-wise median and row-wise mean normalization. This
-method was chosen by Specht et al in the SCoPE2 publication 2021
+significantly. The K-nearest neighbor (K-NN) clustering algorithm, for
+instance relies on the distances between distinct data points by
+minimizing variance of the squared euclidean distances. This method will
+be applied in the missing value imputation, which will be the next step
+applied to the data. The first normalization method available in the
+user interface is column-wise median and row-wise mean normalization.
+This method was chosen by Specht et al in the SCoPE2 publication 2021
 \citep{Specht2021} and worked as a reference during the development
@@
 dataset. However when the logarithm as the transformation method was
-chosen as transformation method, the pipeline automatically switches
-from a division to a subtraction, since:
+chosen as transformation method, ProteoScanR automatically switches from
+a division to a subtraction, since:
 \(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\). After applying
@@
 data. CONSTANd uses a technique called matrix raking, which employs the
-RAS algorithm. The expression set reflects the nonnegative real
+RAS algorithm. The expression set reflects the non-negative real
 \((m, n)\) matrix \(A\) were the bi proportional constrained matrix
@@
 can be observed in the MA-plot which indicates the differences of two
-samples. If we have the intensities of two samples, \(R\) and \(G\), we
-can plot them on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\)
-and the y-axis as \(A=\frac{1}{2}\log_2(RG)\). A comparison between all
+samples. Given the intensities of two samples, \(R\) and \(G\),
+visualized on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\) and
+the y-axis as \(A=\frac{1}{2}\log_2(RG)\). A comparison between all
 sample types of identically processed sets is needed before calculation
@@
 sample types can be assessed. The method is utilized by the bioconductor
-package handler. However if the dataset already consists of balanced
-features and samples or the user wants to benchmark the methods applied,
-normalization can be skipped also.
+package handler. Another method developed in the application is the
+quantile normalization. Quantile normalization is performed over all
+columns, equalizing statistical parameters such as inter quartile range,
+median and mean. However the application of this method is only suitable
+if the number of differentially expressed proteins is approximately
+equal over sample types. Otherwise statistical power could be lost by
+removing fine deviations \citep{Zhao2020}. For the quantile
+normalization method the package limma was utilized. Quantile
+normalization ranks protein expression by magnitude, then averages the
+expression values occupying the same rank and substitute the inital
+value with the average. However if the dataset already consists of
+balanced features and samples or the user wants to benchmark the methods
+applied, normalization can be skipped also, by setting to ``none''.
 
 \hypertarget{missing-value-handling}{%
-\paragraph{Missing value handling}\label{missing-value-handling}}
+\subsubsection{Missing value handling}\label{missing-value-handling}}
 
@@
 \hypertarget{batch-correction}{%
-\paragraph{Batch correction}\label{batch-correction}}
-
-Since mass spectrometry experiments are highly biased by several
+\subsubsection{Batch correction}\label{batch-correction}}
+
+Since mass spectrometry experiments are highly influenced by several
 factors, such as sample processing, the technician, manufacturer and
-production of the device (chromatography column, mass-spectrometer), the
-reagents etc., working in multiple batches introduces further variance
-in the protein data. This variance would overshadow the biological
-differences across sample types. Before attempting to remove batch
-effects it is recommended to elucidate the variance with a
-dimensionality reduction technique which will be explained in the next
-chapter. To remove the technical factor also called non-biological
-experimental variation the R package sva \citep{Leek2012} with the
-ComBat function \citep{Johnson2007} was used. The expression of a gene
-\(g\) for the batch \(m_i\) and sample \(j\) is defined as
-\(Y_i_j_g = a_g + X\beta_g + \gamma_i_g + \delta_i_g\epsilon_i_j_g\),
-whereas \(\gamma\) and \(\delta\epsilon\) denote for the systematic
-error caused by the batch. In order to remove the systematic error the
-bayesian framework shrinks the effect by pooling information over the
-features \(g\). Ideally the effects \(a\) and \(\beta\) remain as
+production of the device, the reagents etc., working in multiple batches
+introduces further variance in the protein data. This variance would
+overshadow the biological differences across sample types when
+performing dimensionality reduction. A common misconception in biology
+is, that batches need to be corrected before differential expression
+analysis, which is the final aim of ProteoScanR. It appears convenient
+to remove the batch effect before performing any statistical analysis,
+which is in turn prone to errors especially if the sample groups are not
+split equally between batches. In this case, group differences influence
+the batch effect reducing statistical power after the correction. These
+both effects are so called interdependent. The opposite could happen if
+the experimental design is heavily unbalanced leading to group
+differences induced by batch correction. Finally it is recommended to
+include the batch factor should in the linear model as a co-factor and
+statistically quantify it \citep{Nygaard2016}. However it could be of
+interest to observe biological relevance of the factor of interest in
+the dimensionality reduction visualization, so a batch correction option
+was included to meet this need. To address the non-biological
+experimental variations, also known as technical factors, the R package
+sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was
+used. The expression of a gene \(g\) for the batch \(m_i\) and sample
+\(j\) can be defined as
+\(Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}\),
+where \(\gamma\) and \(\delta\epsilon\) represent the systematic error
+caused by the batch. The Bayesian framework is employed to remove this
+systematic error by shrinking the effect and pooling information over
+the features \(g\). Ideally the effects \(a\) and \(\beta\) remain as
 natural as they are and should reflect the true biological difference
-between experimental obtained samples or groups. ComBat standardizes the
-data by a Z-transformation before attempting to remove the batch effect.
-By the method of moments the two batch effect parameters \(\gamma\) and
-\(\delta\epsilon\) were estimated and can be subtracted from each value
-for the particular feature.
-
-\hypertarget{downstream-analysis}{%
-\paragraph{Downstream Analysis}\label{downstream-analysis}}
+between experimental obtained samples or groups. Before applying batch
+correction, the data is standardized using Z-transformation. By the
+method of moments (= the mean and the variance) the two batch effect
+parameters \(\gamma\) and \(\delta\epsilon\) were estimated and can be
+subtracted from each value for the particular feature driving
+statistical moments at a comparable level between batches. The ComBat
+function allows the user to specify a desired effect, which represents
+the actual biological variance to be preserved while removing
+non-biological variance. This is achieved by constructing a design
+matrix, which is provided as prior information about the dataset within
+the Bayesian framework. However, in cases of poor experimental design,
+the factors contributing to unwanted and desired variation may be
+confounded. In the worst case, if only one sample type per run is
+observed, it becomes challenging for the program to distinguish between
+batch-induced variance and biological variance. In such cases, the user
+is advised to adjust the experimental layout for future experiments. The
+detection algorithm for confounding effects is facilitated by QR matrix
+decomposition. The matrix \(A=QR\) is decomposed into an orthogonal
+matrix \(Q\) and an upper triangular matrix \(R\). By comparing the rank
+of matrix \(A\) with the number of biologically relevant factors, it is
+possible to determine the presence of confounding effects in the
+dataset. The rank is defined as the maximum number of linearly
+independent columns. The same approach is used in the limma package
+\citep{Phipson2016} to identify non-solvable coefficients in
+experimental designs.
+
+\hypertarget{dimensionality-reduction}{%
+\subsubsection{Dimensionality
+reduction}\label{dimensionality-reduction}}
 
@@
 proteinaceous ensemble, but still keep a high level of information can
-be dimensionality reduction.
+be dimensionality reduction. After the cleaning procedures any
+additional effect caused by technical or biological replication should
+be removed, however it could be found be highlighting according to the
+particular factor. This makes the dimensionality reduction a useful tool
+before starting any statistical analysis by checking for involuntarily
+introduced bias.
 
 \hypertarget{principle-component-analysis-pca}{%
-\subparagraph{Principle component analysis
-(=PCA)}\label{principle-component-analysis-pca}}
+\paragraph{Principle component analysis
+(PCA)}\label{principle-component-analysis-pca}}
 
 The most common approach in data science for any kind of high
-dimensional data is the PCA. Every principle component can be understood
-as a vector pointing in a specific direction derived as the eigenvector
-from the covariance matrix of the expression set. The covariance matrix
-indicates for the co linearity between the variance of all elements
-compared. After calculating the covariance matrix the eigenvectors for
-the covariance matrix will be obtained by solving the quadratic
-equations to obtain the eigenvalues \(\lambda\). Since
+dimensional data is the PCA. For ProteoScanR the bioconductor package
+scater \citep{McCarthy2017} was used. Every principle component can be
+understood as a vector pointing in a specific direction derived as the
+eigenvector from the covariance matrix of the expression set. The
+covariance matrix indicates for the co linearity between the variance of
+all elements compared. After calculating the covariance matrix the
+eigenvectors for the covariance matrix will be obtained by solving the
+quadratic equations to obtain the eigenvalues \(\lambda\). Since
 \(A v = \lambda v\) the eigenvector \(v\) can be obtained by solving the
@@
 The first two components, which explain the majority of variance in the
-dateset highlight the difference between samples. After the cleaning
-procedures any additional effect caused by technical or biological
-replication should be removed, however it could be found be highlighting
-according to the particular factor. This makes the PCA to a useful tool
-before starting any statistical analysis by checking for involuntarily
-introduced bias.
-
-\hypertarget{testing-for-differential-expression}{%
-\paragraph{Testing for differential
-expression}\label{testing-for-differential-expression}}
-
-In order to test for differential expression the package limma from R
-bioconductor was used. \citep{Phipson2016}. The package uses a linear
-model approach to define a fold expression between sample groups. Before
-starting the analysis two matrices need to be computed. The design
-matrix identifies samples according to the sample type and defines the
-experimental design. In order to create the matrix automatically a
-simple algorithm is used within the programmed backend logic. The
-expression matrix contains intensities for each identified protein will
-be obtained at the end of the pipeline. A function call with the
-arguments design and expression matrix calculates the contrast matrix.
-The next step is creating a linear model between groups by log-ratios of
-their expression values.
+dateset highlight the difference between samples.
+
+\hypertarget{uniform-manifold-approximation-projection-umap}{%
+\paragraph{Uniform manifold approximation \& projection
+(UMAP)}\label{uniform-manifold-approximation-projection-umap}}
+
+UMAP \citep{McInnes2018} is a dimensionality reduction technique that
+was developed in recent years and is based on Riemann geometry. It
+shares similarities to the t-distributed stochastic neighbor embedding
+(t-SNE) in terms of graphical visualization. Riemann geometry is a
+mathematical framework that deals with three dimensional functions using
+volumes, areas and vectors, which are also concept of topological data
+analysis. Before performing UMAP, three assumptions need to be
+considered:
+
+\begin{enumerate}
+\item There exists a manifold on which the data would be uniformly distributed.
+$\rightarrow$ This assumption implies that the probability for each value in the biological dataset is the same.
+
+\item The underlying manifold of interest is locally connected.
+$\rightarrow$ In other words, similar samples will have a similar proteinaceous ensemble.
+
+\item Preserving the topological structure of this manifold is the primary
+goal.
+$\rightarrow$ This means that the homeomorphic structure (the property of being able to transform one shape into another without tearing or gluing) is locally preserved, and local values in the dataset can be compared with each other.
+\end{enumerate}
+
+\citep{McInnes2018}
+
+The UMAP algorithm begins by constructing a graph with a defined
+neighborhood parameter, typically denoted as \(k\), which can be
+selected computationally. UMAP employs the k-nearest neighbor descent
+algorithm (NN-descent). Similar to the k-nearest neighbor algorithm
+(KNN), used in the missing value imputation part of this project,
+NN-descent recursively iterates to find the smallest distance to the
+respective neighbors within the defined neighborhood size \(k\).
+Increasing the value of \(k\) gradually strengthens the clustering of
+the data, but at a certain point, fine structure may be lost, and only
+rough estimates of the underlying principle can be observed. The
+mathematical graph in UMAP is defined by nodes connected by edges, with
+each node internally connected to at least one other node. Edges are
+distinguished from each other not only by their connection but also by
+their importance, which is often referred to as weight. Whereas the
+weight function is calculated as
+
+\[w((x_i, x_{i_{j}})) = \exp(\frac{-max(0,d(x_{i},x_{i_{j}}-\rho_{i})}{\sigma_{i}})\]
+
+and determine the parameter \(\rho\) for the weight function that:
+
+\[\rho_i=\min \left\{ d(x_i,x_{i_{j}}) \,|\, 1 \leq j \leq k, d(x_i, x_{i_{j}}) > 0 \right\}\]
+
+Furthermore \(\sigma\) will be set to meet the following conditions:
+\[\sum_{j=1}^{k} \exp ( \frac{ -\max(0, d(x_{i}, x_{i_{j}}) - \rho_i)} { \sigma_i } ) = \log_2(k)\]
+and acts as a normalization factor. The obtained graph \(\overline{G}\)
+is called the fuzzy simplicial set which is obtained by approximating
+the geodesic distance (= locally length-minimizing curve) of the data
+points, leading to a topological approximation. The second phase of the
+algorithm adjusts the layout of the weighted graph to a representative
+view, but preserves the characteristics and shows the underlying
+topological principle of the data. In order to obtain a visual
+representation of the graph, the algorithm applies repulsive forces:
+\[\frac{2b}{(\epsilon + ||y_i-y_j||_{2}^{2})(1+a||y_i-y_j||_{2}^{2b})}(1-w((x_i,x_j)))(y_i-y_j)\]
+among vertices and attractive forces
+\[\frac{ -2ab || y_i -y_j||_{2}^{2(b-1)} }{1+||y_i-y_j||_{2}^{2}} w((x_i,x_j))(y_i-y_j)\]
+along edges. Where \(a\) and \(b\) are denoted as hyperparameters. By
+iterating through possible conformations until converging towards a
+local minimum with decreasing repulsive and attractive forces, the UMAP
+reaches it`s aim to reveal the underlying properties of the data. The
+application utilizes the bioconductor package scater for performing UMAP
+\citep{McCarthy2017}.
+
+\hypertarget{validation-of-the-methods-applied-to-the-dataset}{%
+\subsubsection{Validation of the methods applied to the
+dataset}\label{validation-of-the-methods-applied-to-the-dataset}}
+
+For the validation of methods applied to the dataset, the mutual
+information calculation is used to assess the impact of computations and
+the information it carries. Mutual information, denoted as \(I(X;Y)\),
+is a measure of the dependence or information shared between two
+variables, \(X\) and \(Y\), in this case sample types. It is calculated
+using the equation \(I(X;Y) = H(X,Y) - H(X|Y) - H(Y|X)\), where \(H\)
+represents the entropy, \(H(X|Y)\) and \(H(Y|X)\) denote for the both
+conditional probabilities and \(H(X,Y)\) for the joint entropy. Entropy,
+as a concept in information theory, quantifies the level of disorder or
+uncertainty associated with a random event. Events that are highly
+likely to occur are considered less informative than events that are
+more unexpected or less likely. In a Venn diagram \(I(X;Y)\) (the mutual
+information) corresponds to the intersection of the two conditional
+entropys \(H(X|Y)\) and \(H(Y|X)\). To calculate the mutual information
+between two variables \(X\) and \(Y\) the above equation can be
+alternatively expressed as
+\(I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})\). \(D_{KL}\)
+denotes for the Kullback-Leibler \citep{Kullback1951} divergence which
+is a concept of relative entropy between two variables and tells us how
+much 2 probability distributions differ from each other.
+\(P_{X} \otimes P_{Y}\) represents the product of the marginals. The
+underlying principle of divergence is the integral asymmetry in Bayesian
+inference, which means that divergence does not satisfy the triangle
+inequality. Differences are not the same depending on the starting
+point. By comparing the mutual information before performing any
+calculations with the mutual information calculated after the
+computations, it is possible to assess whether the applied methods have
+significantly altered the dependence or information content. The goal is
+to choose a method which preserves the information contributed by each
+biological sample type to the experiment. The R package ``infotheo'' is
+utilized to perform the estimates for the entropy of the two variables.
+Mutual information is returned as natural unit of information (=nat), a
+measurement proportional to the to the Shannon entropy
+(\(1nat = \frac{1}{ln2} shannons\)). Before calculating the pairwise
+mutual information of the expression matrix, values were discretized
+into bins with the discretize function of above package. The mutual
+information is calculated within each sample type, and it can also be
+calculated between a selection of sample types to determine if the
+computations performed affect the information content. After that all
+pairwise mutual information within the expression set before and after
+ProteoScanR is obtained. The pairwise differences can be observed as a
+boxplot for a particular sample type indicating the change in dependence
+within and also compared to another sample type.
+
+\hypertarget{statistics}{%
+\subsubsection{Statistics}\label{statistics}}
+
+\hypertarget{differential-expression}{%
+\paragraph{Differential expression}\label{differential-expression}}
+
+After visualizing the biological tendency in the dataset, the next step
+is hypothesis testing. Several experimental designs can be tested
+against the null hypothesis \(h_0\), which states that there is no
+significant difference between the sample types. Before proceeding with
+any statistical testing, it is important to observe the data using
+either a quantile-quantile plot (qq-plot) or a histogram. The qq-plot
+serves a vital role in exploratory statistics by visualizing whether two
+datasets follow the same underlying distributions. In the case of
+statistical preparation for parametric tests, the first distribution is
+the dataset of interest, while the second dataset is the normal
+distribution. If the dataset follows an exact normal distribution, the
+values will align with the red line in the chart. However, values at the
+ends of the distribution may deviate from the normality line, indicating
+differentially expressed genes. These genes can be observed both in the
+histogram and the qq-plot. Parametric statistical tests rely on the
+assumption of normality in the data, which needs to be achieved before
+building a model. To achieve normality, it is recommended to try various
+combinations of transformation and normalization methods. By observing
+the results in the plots, one can determine which combination of data
+processing methods satisfies the conditions required for the tests.
+However the limma method utilized in the statistics module shows great
+robustness to imperfect normal distributed values \citep{Ritchie2015}.
+
+To test for differential expression, the analysis utilizes the ``limma''
+package from R Bioconductor \citep{Phipson2016}. This package employs a
+linear model approach to determine the fold expression between sample
+groups. Before initiating the analysis, two matrices need to be
+computed. The design matrix identifies samples based on their sample
+type and defines the experimental design. An algorithm within the
+program's backend logic is utilized to automatically generate this
+matrix according to the user's selection in the interface. The
+expression matrix contains intensities for each identified protein and
+is obtained at the end of ProteoScanR. By calling a function with the
+design and expression matrices as arguments, the contrast matrix is
+calculated, enabling the user to decide which comparisons to consider
+and also perform tests on multiple factors. When observing multiple
+batches at once, as already explained in the chapter batch correction
+this effect needs to be taken into consideration when performing
+differential expression analysis. This is done by the multi-factor
+option in the user interface by selecting the batch as a co-factor and
+inclusion in the design matrix. The advantage of this procedure is, that
+the biological relevance of working in multiple batches can be
+quantified and the publishing of false positive results avoided. The
+multi factor option uses an additive model as a design matrix, whereas
+continuous variables (such as size, weight etc.) can also be used to
+build the statistical model.
+
+The subsequent step involves creating a linear model between groups
+using log-ratios of their expression values. In a two-way design, the
+expression of the gene \(y\) can be explained as
+\(Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \epsilon\), where \(\beta_0\)
+represents the intercept, which can be interpreted as the mean
+expression of the gene. \(\beta_1\) represents difference in mean of the
+treatment (or condition) on the discrete or continuous variable
+\(X_{1}\), and \(\epsilon\) acts as an error term. The interface allows
+for additive models with multiple factors, allowing the user to add
+factors according to various scenarios. For example, in a two-factor
+model, it would look like:
+\(Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + \epsilon\).
+The principle remains the same for models with multiple factors. When
+visualizing the linear model, the x-axis would represent the gene for
+the sample types or other factors, whereas the y-axis would account for
+the expression value. By drawing a line through the cloud of data points
+the model is constructed by minimizing the quadratic distance to every
+data-point, called the least-square method or also ordinary least-square
+method. With the regression line the above mentioned equation can be
+constructed.
+
+Next the statistical module of the program performs an empirical Bayes
+model utilizing the eBayes \citep{Smyth2004} function of the Limma
+package. The eBayes function utilizes a moderated t-statistics following
+t-distributed values. The advantage of this method over the posterior
+odds is the reduction of hyperparameters to estimate used for the
+modelling process, also called shrinkage. In this context, the
+hyperparameters to estimate are the coefficient \(\beta_{gj}\) for gene
+\(g\) and sample \(i\), as well as the variance \(\sigma^2_g\) across
+genes \(g\). In the Bayesian approach, probabilities are updated after
+obtaining new data, referred to as conditional probabilities. For
+estimation, it is assumed that
+\(\frac{1}{\sigma^2_g} \sim \frac{1}{d_{0}s^2_{0}} \chi^2_{d_{0}}\),
+representing the prior information of the model. The probability for
+\(\beta_{gj} \neq 0\) is denoted as \(p_j\), where \(p_j\) represents
+the expected proportion of truly differentially expressed genes. The
+expected distribution of log-fold changes follows the distribution
+\(\beta_{gj} | \sigma^2_g , \beta_{gj} \neq 0 \sim N(0, v_{0j}\sigma^2_g)\).
+The posterior mean of \(\sigma^2_g\) given \(s^2_g\) is calculated as
+\(s^{-2}_g = E(\sigma^2_g | s^2_g) = \frac{d{0}s^2_{0} + d_{g}s^2_{g}}{d_{0} + d_{g}}\).
+This is where the shrinkage occurs, as the posteriors affect the prior
+values based on their respective sizes and degrees of freedom. The
+moderated t-statistic is defined as
+\(\tilde{t}_{gj} = \frac{\hat{\beta}_{gj}}{\hat{s}_{g}\sqrt{v_{gj}}}\).
+Once the values of \(\tilde{t}\) and \(s^{2}\) are calculated, the
+posterior odds are computed, providing the odds that a particular gene
+is differentially expressed. The odds for gene \(g\) being
+differentially expressed are denoted as
+\(O_{gj} = \frac{p(\beta_{gj} \neq 0 | \tilde{t}_{gj}, s^2{g})}{p(\beta_{gj} = 0 | \tilde{t}_{gj}, s^2{g})}\),
+where the numerator represents the probability that the gene is
+differentially expressed, and the denominator represents the probability
+that the gene is not differentially expressed \citep{Smyth2004}.
+
+Limma outputs a table with all proteins, their respective logfold
+change, p-values and adjusted p-values. The user can choose different
+methods for the correction of the p-values, which is necessary when
+testing the same hypothesis multiple times. A common method in omics
+studies is the false discovery rate (=fdr) by Benjamini Hochberg
+\citep{Benjamini1995}. By ranking all the p-values from smallest to
+largest the p-values are adjusted sequentially using the formula
+\(p_(i) = \frac{r_i}{m} Q\). The rank in the p-value table is denoted as
+\(r_i\) , \(m\) reflects the total number of tests and \(Q\) represents
+false discovery rate which is set to 5\%. The second option to choose is
+the Benjamini \& Yekutieli (=BY) \citep{Benjamini2001}method, which is a
+further development of the fdr by the same statistician and has a wider
+range of application based on the dependencies for the test. The third
+and most conservative option is the Holm method \citep{Holm1979} which
+controls the error sequentally in a family wise manner, which is
+considered as traditional in regards to the above mentioned experimental
+wise error rates. The above mentioned correction methods are ordered in
+the way from the most liberal to the most conservative one. The
+selection is supposed to give the user the possibility questioning
+results on multiple levels.
+
+\hypertarget{protein-set-enrichment-analysis}{%
+\subsubsection{Protein set enrichment
+analysis}\label{protein-set-enrichment-analysis}}
+
+After differential expression the upcoming question could be, in which
+context the up or down regulated proteins are. Enriched pathways can be
+understood as an over-representation (ORA = over-representation
+analysis) of proteins within a specific set. This set can be a pathway
+or a selection of common denominators in biological contexts, such as
+cell types or oncological factors. The statistical testing to identify
+the overabundant proteins of a specific set is similar to a \(\chi^2\)
+test developed by Karl Pearson \citep{Pearson1900} or the exact Fisher
+test \citep{Sprent2011} for tables with 4 fields in small sample sizes.
+Within the list are all truly differentially expressed genes, an
+annotation is the identification of a protein in the desired biological
+context. The table is expected to be populated by the values as follows:
+
+\begin{center}
+\begin{tabular}{|c|| c| c|| c|} 
+ \hline
+   & in list & not in list & totals \\ [0.5ex] 
+ \hline\hline
+ with annotation & (A+B)(A+C)/N  & (A+B)(B+D)/N & A+B \\ 
+ \hline
+ without annotation & (C+D)(A+C)/N & (C+D)(B+D)/N & C+D \\
+ \hline
+  & A+C & B+D & N \\
+ \hline
+\end{tabular}
+\end{center}
+
+Under the null hypothesis \(H_{0}\) states that the distribution and
+population of the table is random and there is no clear tendency with
+significant evidence. Against that the alternative hypothesis \(H_{A}\)
+says that there is an underlying tendency the table is populated. This
+leads to the test statistics, the observed values (=\(O\)) in the table
+are significantly different to the expected
+(=\(E=\frac{(A+B \lor C+D)}{N}\) ones. One approach is approximating
+\(\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}\) and deriving the
+p-value by the area under the density curve of the \(\chi^2\)
+distribution \(Q = \sum_{i=1}^{k} Z_{i}^{2}\). In the distribution \(k\)
+is defined by the degrees of freedom (=df), what can be obtained by
+\(k=(columns-1)*(rows-1)=df\) of the table. Another approach to obtain
+an exact p-value would be Fisher`s exact test, which states that the
+margins are distributed according to the hypergeometric distributions.
+The test statistic leads directly to the p-value with no need for
+approximation:
+\(p=1-\sum_{i=0}^{A-1}\frac{\binom{B}{i}\binom{D-B}{A+C-i}}{\binom{D}{A+C}}\)
+Here D denotes either for the background distribution which can be the
+total number of proteins obtained in the experiment, all proteins with
+annotation or a custom selected background. The background indicates all
+proteins which could be positive tested for in the observed sample. The
+above mentioned three options are implemented in the user interface.
+After obtaining the p-values for each protein set, the subsequent
+critical step involves the correction for multiple testing, considering
+the inherent risk of false positive results. Multiple testing correction
+methods adjust the significance thresholds to account for the increased
+probability of detecting false positives when performing multiple
+statistical tests. In this protein set enrichment analysis, the user is
+provided with a selection of established correction methods, akin to
+those commonly utilized in the field of differential expression
+analysis, ensuring robust and reliable interpretation of the results.
+Once the enriched protein pathways are determined, they can be visually
+represented using bar or dotplots, and a network of the found pathways.
+In the network ontologies are depicted as nodes, whereas edges show the
+strength of the interaction by thickness corresponding to the number of
+associated proteins.
+
+\hypertarget{pathway-enrichment-analysis}{%
+\paragraph{Pathway enrichment
+analysis}\label{pathway-enrichment-analysis}}
+
+The pathway enrichment analysis is performed using the R Bioconductor
+package clusterProfiler \citep{Wu2021}. The analysis utilizes the entire
+``Kyoto Encyclopedia of Genes and Genomes'' (KEGG) database, which is
+accessed through a function that maps the UniProt IDs to corresponding
+pathways.
+
+\hypertarget{custom-ontology-based-protein-enrichment-analysis}{%
+\paragraph{Custom ontology based protein enrichment
+analysis}\label{custom-ontology-based-protein-enrichment-analysis}}
+
+For custom ontologies the protein enrichment analysis is done using the
+R bioconductor package piano \citep{Vaeremo2013}. In opposite to the
+pathway enrichment, piano maps proteins to a pre-selected gene
+collection set, which can be either downloaded from gsea-msigdb.org or
+custom made for the particular experiment and explains conditions,
+phenotype or other desired properties. For statistical testing the above
+explained Fisher exact test is used.
 
@@
 
-Some more guidlines from the School of Geosciences.
-
-This section should summarise the findings of the research referring to
-all figures, tables and statistical results (some of which may be placed
-in appendices). - include the primary results, ordered logically - it is
-often useful to follow the same order as presented in the methods. -
-alternatively, you may find that ordering the results from the most
-important to the least important works better for your project. - data
-should only be presented in the main text once, either in tables or
-figures; if presented in figures, data can be tabulated in appendices
-and referred to at the appropriate point in the main text.
-
-\textbf{Often, it is recommended that you write the results section
-first, so that you can write the methods that are appropriate to
-describe the results presented. Then you can write the discussion next,
-then the introduction which includes the relevant literature for the
-scientific story that you are telling and finally the conclusions and
-abstract -- this approach is called writing backwards.}
+To begin using the application, the first step is to upload the
+tab-delimited evidence file and the sample annotation file. Once these
+files are uploaded, the application is ready to run with default values.
+The default settings are preconfigured for an initial analysis, and any
+further adjustments or fine-tuning can be made in subsequent steps of
+the analysis process.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/overview}
+
+}
+
+\caption{Summary showing the process of the analysis pipeline}\label{fig:ui_summary}
+ \endfigure\egroup
+
+After a runtime of 2 minutes on the 4 core laptop the data is ready to
+visualize. On the left side of the user interface, users can find the
+analysis pipeline settings that can be adjusted. Once the desired
+changes are made, they can be applied by clicking the update button.
+This allows users to observe the direct impact of the applied changes on
+the dataset. In the summary visualization, it may appear cluttered due
+to the processing of 36 files. In smaller datasets, the nodes of the
+network graph can be more easily distinguished and understood. When
+running the application with SCoPE2 \citep{Specht2021} reference
+cutoffs, the following results were visualized.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/numbers_default}
+
+}
+
+\caption{Barchart showing the number of peptide spectrum matches, peptides and proteins}\label{fig:ui_numbers_default}
+ \endfigure\egroup
+
+The default settings of the analysis pipeline resulted in a total number
+of unique proteins below approximately \(10^3\), which were aggregated
+from around \(10^{3.5}\) peptides. Before these peptides were aggregated
+over approximately \(10^{4.5}\) peptide spectrum matches. The
+aggregation method used is fixed by median.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/mutual_info_default}
+
+}
+
+\caption{Mutual information with default setting}\label{fig:ui_mutual_info_default}
+ \endfigure\egroup
+
+After applying logarithmic transformation to the base 2 and performing
+column median row mean normalization, the mutual information (MI) within
+the healthy control (HC) group's sample type exhibits a reduction
+towards the lower edge of the interquartile range (IQR).
+
+MI between the HC group and the sample type corresponding to individuals
+with both HIV and metabolic syndrome, the IQR is further reduced and
+shifted towards the lower end compared to the MI values observed in the
+raw data.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/constand_norm_dependency}
+
+}
+
+\caption{Popup showing the dependencies for the CONSTANd normalization}\label{fig:ui_constand_norm}
+ \endfigure\egroup
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/mutual_info_quant_norm}
+
+}
+
+\caption{Mutual information with quantile normalization}\label{fig:ui_mutual_info_quant_norm}
+ \endfigure\egroup
+
+After performing a logarithmic transformation to the base 2 and quantile
+normalization, the mutual information (MI) within the healthy control
+(HC) group's sample type exhibits a minor reduction in the interquartile
+range (IQR). However, it should be noted that there is an increase in
+the number of outliers compared to the raw data.
+
+Observing MI IQR of the HC group with the sample type corresponding to
+individuals with both HIV and metabolic syndrome, the reduction in MI
+falls within the interquartile range of the raw data MI. Additionally,
+the median remains close to unaffected, indicating that the overall
+central tendency of the MI values is relatively preserved.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/RI_channel_default}
+
+}
+
+\caption{Reporter ion intensity with selection on channel. The selected factor will be highlighted and indicated in the legend.}\label{fig:ui_RI_channel_default}
+ \endfigure\egroup
+
+When comparing the median reporter ion intensities across channels, no
+systematic bias is observed. However, it is worth noting that channel 7
+exhibits a wide range of values spanning an order of magnitude. This
+comparison is crucial since the three sample types consistently occupy
+the same channels in every run, making the visualization an important
+aspect of the analysis. Despite the wide range of values in channel 7,
+there is still overlap in the interquartile ranges (IQR) across the
+majority of samples.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/RI_file_default}
+
+}
+
+\caption{Reporter ion intensity with selection on file}\label{fig:ui_RI_file_default}
+ \endfigure\egroup
+
+Comparing the median reporter ion intensities with a focus on the batch
+(=Raw.file), it becomes evident that there are significant changes in
+signal that need to be taken into account in further analysis. After
+applying quality control measures, several values have been lost due to
+cutoffs in parameters such as parental ion fraction (PIF), q-value,
+coefficient of variation, and missingness of peptide data.
+
+For dimensionality reduction, the user has the option to choose the
+batch correction method ComBat \citep{Leek2012} in the side panel.
+However, this option is not set as the default choice. Instead, the
+conservative approach is followed, where the batch factor is included in
+the statistical module to account for any potential batch effects that
+may influence the results.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/RI_st_default}
+
+}
+
+\caption{Reporter ion intensity with selection on sampletype}\label{fig:ui_RI_st_default}
+ \endfigure\egroup
+
+The sample type does not appear to have an impact on the median reporter
+ion intensities, as suggested by the overlapping interquartile ranges
+(IQRs). Nevertheless, the HIV group with metabolic syndrome displays a
+broader range of values spanning tenfold in arbitrary units.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/cv_st_default}
+
+}
+
+\caption{Median covariance of razor proteins with selection on sampletype}\label{fig:ui_cv_st_default}
+ \endfigure\egroup
+
+A cutoff value of 0.65 for the coefficient of variation (CV) of razor
+proteins as shown in the SCoPE2 publication \citep{Specht2021}
+demonstrates a satisfactory fit to the data. The minimum number of
+observations determines the threshold for considering a razor protein
+ambiguous, based on the number of matching peptides. In this case, the
+reference publication set the threshold at 5 observations.
+
+The interquartile ranges (IQRs) remained unchanged by this cutoff. The
+sample type did not influence the CV of razor proteins. Boxplots
+illustrate that the values are distributed similarly, as indicated by
+the overlapping IQRs.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/cv_rawfile_default}
+
+}
+
+\caption{Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file}\label{fig:ui_cv_rawfile_default}
+ \endfigure\egroup
+
+The plot illustrates that certain batches (=raw files) exhibit a higher
+number of distinct matches compared to others, as indicated by a thin
+line depicting low density. Most interquartile ranges (IQRs) overlap,
+suggesting similarity among the majority of batches. However, specific
+batches display a lower IQR in median CV across razor protein peptides,
+indicating distinct matches.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/cv_rawfile_nobs2}
+
+}
+
+\caption{Median coefficient of variation (CV) of leading razor proteins calculated for 2  observations per cell, with selection on Raw.file}\label{fig:ui_cv_rawfile_nobs2}
+ \endfigure\egroup
+
+By setting the number of observations to 2, the distributions become
+more equalized. Additionally, batches that initially exhibited a
+decreased coefficient of variation (CV) tend to inflate towards the
+threshold.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/cv_rawfile_nobs10}
+
+}
+
+\caption{Median coefficient of variation (CV) of leading razor proteins calculated for 10 observations per cell, with selection on Raw.file}\label{fig:ui_cv_rawfile_nobs10}
+ \endfigure\egroup
+
+When the minimum number of observations for ambiguous peptides is set to
+10, the distribution of medians across channels shows a decrease in
+density. This can be attributed to the rare occurrence of 10 unique
+peptides matching to the same protein, which is not commonly observed in
+the dataset.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/heatmap}
+
+}
+
+\caption{Correlation heatmap}\label{fig:ui_heatmap}
+ \endfigure\egroup
+
+The heatmap reveals the Pearson pairwise correlation coefficients
+between all proteins. The dendrogram indicates that there are two main
+protein clusters, with one of the clusters showing further
+sub-clustering, indicating a strong relationship among the proteins in
+the sub groups. However, the other cluster, consisting of individual
+proteins, shows weaker correlations, leading to a more intricate
+branching pattern in the dendrogram.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/dim_red_default}
+
+}
+
+\caption{Dimensionality reduction analysis depicting sample types (indicated by shape) and raw files (indicated by color).
+Top: principle component analysis (PCA).
+Lower: Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_default}
+ \endfigure\egroup
+
+In both the Principal Component Analysis (PCA) plot and the Uniform
+Manifold Approximation and Projection (UMAP) plot, the second component
+is plotted against the first component. Surprisingly, neither
+visualization reveals any discernible clustering patterns based on
+sample type or batch. For the k-nearest neighbor descent algorithm
+(NN-descent) employed in the UMAP, first a default value of k=3 was
+selected.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/dim_red_changed_k}
+
+}
+
+\caption{Dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+Top: Principal Component Analysis (PCA).
+Bottom: Uniform Manifold Approximation and Projection (UMAP).
+For the UMAP visualization, the k value for the NN-descent algorithm was modified to 15.}\label{fig:ui_dim_red_changed_k}
+ \endfigure\egroup
+
+The second figure of the Principal Component Analysis (PCA) plot and the
+Uniform Manifold Approximation and Projection (UMAP), both displaying
+the second component against the first. Surprisingly, neither
+visualization reveals any discernible clustering patterns based on
+sample type or batch information. To enhance the resolution and capture
+more intricate patterns, the k-nearest neighbor descent algorithm
+(NN-descent) was adjusted, with a value of K set to 15. While the
+resulting plot exhibits improved detail, the underlying biological
+context or background remains elusive and does not emerge in the visual
+representation.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/dim_red_3d}
+
+}
+
+\caption{3D visualization of dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+Top: Principal Component Analysis (PCA).
+Bottom: Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_c3d}
+ \endfigure\egroup
+
+In the 3D visualization, a subtle clustering pattern can be observed for
+the first two batches. Upon switching to the indicator sample type in
+the 3D visualization, no clustering can be observed whatsoever. This
+suggests that the variance in the data set is not attributed to sample
+type or batch, reinforcing the notion that other factors could
+contribute to the observed variability.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/dim_red_batchC}
+
+}
+
+\caption{Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+upper top: Principal Component Analysis (PCA).
+upper bottom: Uniform Manifold Approximation and Projection (UMAP).
+lower top: 3D Principal Component Analysis (PCA).
+lower bottom: 3D Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_batchC-1}
+ \endfigure\egroup 
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/dim_red_batchC_3d}
+
+}
+
+\caption{Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+upper top: Principal Component Analysis (PCA).
+upper bottom: Uniform Manifold Approximation and Projection (UMAP).
+lower top: 3D Principal Component Analysis (PCA).
+lower bottom: 3D Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_batchC-2}
+ \endfigure\egroup
+
+After applying batch correction using ComBat \citep{Leek2012}, neither
+the 2D nor the 3D visualization demonstrates any noticeable effects
+related to batches or sample types. The previously observed subtle
+clustering of the first two batches is now eliminated, indicating
+successful batch correction. However, it's important to note that the
+default pipeline was executed without batch correction for correct
+statistical testing.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/feature}
+
+}
+
+\caption{Feature wise output for the protein P61313 (=60S ribosomal protein L15) over the course of the pipeline.}\label{fig:ui_feature}
+ \endfigure\egroup
+
+The scatter plots indicate the detected values associated with the
+protein UniProt ID P61313 (= 60S ribosomal protein L15). Starting from
+the peptide spectrum level and progressing to peptides aggregated for
+individual raw files, these plots offer a comprehensive view of the
+protein's peptide abundance across various batches. As indicated 14
+batches show no abundance of peptides derived from the protein P61313 at
+all. After merging the batches, the peptides appear to be abundant in
+all channels.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{screenshots/stat_dependency}
+
+}
+
+\caption{Popup window for checking the dependency of the statistic module}\label{fig:ui_stat_dependency}
+ \endfigure\egroup
+
+The t-statistics employed in limma are known for their robustness in
+handling non-normally distributed values. Nonetheless, it is recommended
+for users to assess the distribution of their data based on sample types
+before proceeding with statistical analysis. If the user is not
+satisfied with the results or identifies any distributional issues, the
+user can consider revisiting the transformation or normalization steps
+to address any concerns or improve the data distribution prior to
+statistical analysis.
+
+The values for all sample types exhibit a unimodal distribution.
+However, when examining the quantile-quantile (qq) plot, heavy tails can
+be observed for all three groups. This indicates deviations from a
+perfectly normal distribution, suggesting potential outliers in the
+data.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.7\linewidth]{screenshots/stat_HC_HIV_noMets}
+
+}
+
+\caption{Statistic module}\label{fig:ui_stat_HC_HIV_noMets}
+ \endfigure\egroup
+
+\newpage
+
+By comparing the log10-transformed and quantile-normalized expression
+values between the HIV group without metabolic syndrome and the healthy
+control (HC) group, six differentially expressed proteins were
+identified. The Venndiagram indicates significantly up/downregulated
+proteins. The Volcano represents -log10(adjusted p values) against log2
+fold changes (logFC). These proteins include P68133 (Actin, alpha
+skeletal muscle), P47914 (60S ribosomal protein L29), P25311
+(Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear
+ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform
+type-2), and P46777 (60S ribosomal protein L5). A total of 652 proteins
+have been identified as not significantly upregulated or downregulated
+in the comparison analysis. These proteins did not exhibit statistically
+significant differences in expression between the compared groups or
+conditions.
+
+The direction of the contrast model determines the change in log-fold
+change (logFC) values. For instance, in the depicted figure, P68133
+(Actin, alpha skeletal muscle) shows an increased expression of 0.25
+(25\%) in HIV\_noMetS macrophages compared to HC macrophages. It's
+important to note that the logFC value changes by multiplication with -1
+when the direction of the contrast model is reversed.
+
+To account for potential batch effects, the batch factor was included as
+a cofactor in the model, ensuring that any observed differential
+expression is not confounded by batch variation.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{screenshots/path_enrich_nw}
+
+}
+
+\caption{Pathway enrichment. Networkplot indicating the association of differentially expressed proteins with pathways}\label{fig:ui_path_enrich_nw}
+ \endfigure\egroup
+
+The pathway enrichment mapped and tested with the exact Fisher test
+\citep{Sprent2011} the differentially expressed proteins P46777 (60S
+ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the
+ribosome and coronavirus disease (COVID-19) nodes.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{screenshots/path_enrich_bar}
+
+}
+
+\caption{Pathway enrichment. Barplot indicating the interaction significance of differentially expressed proteins}\label{fig:ui_path_enrich_bar}
+ \endfigure\egroup
+
+The barplot shows the significance of the two nodes. The count indicates
+how many proteins are associated with the node of the network. When
+hovering over the bar the user can read the p-value of the performed
+Fisher test. The background for the statistical test can be chosen as
+genome wide (default) or as the detected proteins. When choosing the
+detected proteins as a background no pathway could be considered as
+statistically significant.
+
+\newpage
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{screenshots/go_enrich_dot}
+
+}
+
+\caption{Gene ontology enrichment against the complete human proteome. Snapshot dotplot indicating the interaction significance of found proteins}\label{fig:ui_go_enrich_bar}
+ \endfigure\egroup
+
+The position of the dot shows the number of matches for the particular
+protein set. Colors indicate for the significance level. In this
+analysis the entire repertoire of detected proteins was used and fetched
+against the bundle containing all Human gene sets in the Human Molecular
+Signatures Database (MSigDB). After a runtime of 3 minutes 672 of 24667
+gene sets were found significantly over represented. Only the protein
+set collections with the most protein matches were used (snipping tool
+Plotly).
 
@@
 
-the purpose of the discussion is to summarise your major findings and
-place them in the context of the current state of knowledge in the
-literature. When you discuss your own work and that of others, back up
-your statements with evidence and citations. - The first part of the
-discussion should contain a summary of your major findings (usually 2 --
-4 points) and a brief summary of the implications of your findings.
-Ideally, it should make reference to whether you found support for your
-hypotheses or answered your questions that were placed at the end of the
-introduction. - The following paragraphs will then usually describe each
-of these findings in greater detail, making reference to previous
-studies. - Often the discussion will include one or a few paragraphs
-describing the limitations of your study and the potential for future
-research. - Subheadings within the discussion can be useful for
-orienting the reader to the major themes that are addressed.
+The Proteomics Workbench interface and the ProteoScanR pipeline
+demonstrated how interactive engagement with the data not only enhances
+the experience of biologists but also improves the comprehension of the
+underlying significance of a biological dataset. By utilizing an
+entropy-based visualization approach, the conservation of information
+can be validated, allowing users to select appropriate methods and
+adjust thresholds, cutoffs, and techniques accordingly. With the
+selection of factors to indicate over multiple plots, biases can be
+identified on various levels. Additionally, the analysis can be examined
+on an individual protein basis, enabling the identification of
+suspicious results. The statistics module provides users with a simple
+tool to assess and visualize the results, offering clarity to dense
+scatter plots through hover functions. Enrichment mapping links
+statistically significant proteins to either the KEGG pathway database
+or individual research data sets, depending on user selection. These
+final results place the data in a biological context, which can be
+explored interactively in graphical visualizations.
+
+\hypertarget{interactive-pipeline}{%
+\subsection{Interactive pipeline}\label{interactive-pipeline}}
+
+Linear programming scripts are widely utilized for data analysis, and
+R-shiny provides an efficient platform for implementing these scripts in
+applications. This approach not only benefits scientists who are not
+proficient in programming, but it is also convenient for programmers.
+The Proteomics Workbench user interface offers a visually appealing
+design, and the interactive menus enhance the user experience by
+eliminating the need to search for specific lines of code on a black
+screen. Modifying parameters of functions becomes effortless, and data
+can be visualized without rerunning the entire script. R-shiny executes
+functions on demand, which means that graphics, variables, and data
+frames are not recalculated if their dependencies remain unchanged. This
+feature makes the application both cost-effective and time-efficient.
+The analysis pathway is visualized through multiple components,
+providing users with a comprehensive view of the data. The overview
+plot, generated using the Q-feature object-oriented programming in the
+scp R library \citep{Vanderaa2021}, serves as a visual representation of
+the analysis process. This plot allows users to track the progression of
+the analysis and identify any missing files or samples.
+
+\hypertarget{mutual-information}{%
+\subsection{Mutual Information}\label{mutual-information}}
+
+Information theory and the concept of mutual information provide a
+framework for understanding the conservation of entropy in a dataset.
+Traditionally, the validation of data transformation and normalization
+methods has been limited to observing distributions through histograms
+or Quantile-Quantile plots. However, the developed approach goes beyond
+these conventional methods, providing a deeper context for the variation
+in the data. By allowing users to choose the method that preserves the
+maximum amount of information, the approach enables researchers to
+extract the most valuable insights from their experiments. Normalization
+methods play a crucial role in data analysis, and their selection
+depends on the nature of the data and the specific analysis objectives.
+The mutual information analysis revealed that quantile normalization
+exhibited the smallest loss of entropy compared to other normalization
+methods
+
+\hypertarget{reporter-ion-intensity}{%
+\subsection{Reporter ion intensity}\label{reporter-ion-intensity}}
+
+MS2 reporter ion intensities can be effectively visualized using a
+boxplot, where the medians represent the intensity distribution across
+various factors such as sample type, batch, or channel. This
+visualization aids users in assessing the variation and potential biases
+in the dataset. During the example analysis, the reporter ion intensity
+plot revealed inconsistencies in the dataset based on signal intensities
+associated with different batches. It was observed that certain runs
+exhibited a wide spread of signal intensities, which could indicate
+technical issues or batch-related variations. These inconsistencies were
+partly filtered out during the quality control process to ensure the
+reliability and accuracy of the data.
+
+\hypertarget{median-coefficient-of-variation}{%
+\subsection{Median coefficient of
+variation}\label{median-coefficient-of-variation}}
+
+The median coefficient of variation (CV) was calculated for a razor
+protein in relation to its ambiguous peptides. This statistical
+parameter measures the variability of the razor protein's abundance
+across different peptide measurements. The results were then visualized
+using a boxplot, allowing users to assess the distribution of CV values
+based on a selected factor. During testing and fine-tuning, the number
+of observations for ambiguously matched peptides was adjusted to
+evaluate its impact on the CV. By varying the number of observations,
+researchers could determine the optimal parameter that yielded
+meaningful CV values. In this context, the reference values from the
+SCoPE2 publication \citep{Specht2021} were employed as benchmarks and
+subjected to validation. The results showed that using a number of
+observations of 5 for leading razor peptides was suitable for
+identifying a cutoff value of 0.65. This cutoff value was determined by
+observing the whiskers of the boxplots. By setting this threshold, noisy
+peptides in the dataset could be effectively reduced, enhancing the
+quality and reliability of the data.
+
+\hypertarget{feature-wise-plot}{%
+\subsection{Feature-wise plot}\label{feature-wise-plot}}
+
+The feature-wise output plays a crucial role in enhancing the analysis
+pathway by offering users the ability to select specific proteins and
+explore their associated peptide spectra, peptide values, and protein
+values throughout the analysis. This feature provides a detailed and
+comprehensive view of the analysis results, enabling users to identify
+any missing or erroneous data points, particularly in channels with low
+or high abundance. By examining these individual protein profiles, users
+can assess the quality of the wet lab work and detect any potential
+errors or inconsistencies. In the case of observing the abundance of the
+protein P61313 (60S ribosomal protein L15) over the entire course of the
+experiment, it becomes apparent that there are missing values within
+certain runs and channels. Specifically, 14 runs exhibit no signal for
+this protein. This observation underscores the significance of
+aggregating data at the peptide and protein levels and combining
+individual runs to construct the final protein expression set. The
+scattering of dots in the plot indicates the inflation of small values
+and the deflation of large values after transformation. The scatter plot
+of quantile normalized values demonstrates the processing steps taken
+while preserving outliers and maintaining the biological relevance of
+the data. This processing steps aid in achieving a more balanced
+distribution of data and avoids the distortion of biological meaning.
+
+\hypertarget{statistics-module}{%
+\subsection{Statistics module}\label{statistics-module}}
+
+The statistics module within the application provides users with a
+user-friendly interface for hypothesis testing, even without prior
+expertise in statistics. By leveraging the powerful R-package limma
+\citep{Ritchie2015}, the module offers robust statistical analysis
+methods, empowering users to perform tests and obtain results to
+validate their hypotheses. One key feature of the statistics module is
+its ability to handle multiple factors and therefore correct for batch
+effects. Users can extend the statistical analysis across various
+factors, allowing them to quantify the impact of different variables and
+identify differentially expressed proteins. To enhance the user
+experience, the visualization of statistical results has been improved,
+resulting in visually appealing and interactive displays. These
+visualizations enable users to explore and interpret the results
+effortlessly, aiding in the comprehension of the statistical findings.
+By providing intuitive and engaging visual representations, the module
+facilitates the understanding and communication of complex statistical
+analyses. When comparing the group HIV without metabolic syndrome and
+the healthy control 6 proteins could be found as differentially
+expressed after correcting the p-values by false discovery rate (FDR
+\citep{Benjamini1995}). For example, the protein P68133 (Actin, alpha
+skeletal muscle) exhibits a log fold change (logFC) of -0.25, indicating
+a significant decrease in actin levels within the HIV group. This
+finding aligns with previous publications suggesting that HIV exploits
+cellular skeleton proteins and alters their expression levels
+\citep{Turville2018}. By inducing orchestrated structural changes in
+compartments composed of actin filaments, the virus facilitates its
+spread and infects other lymphocytes \citep{Rodrigues2022}. Another
+protein of interest, P47914 (60S ribosomal protein L29), demonstrates
+the impact of viral infection on protein synthesis. Differential
+expression of this protein further highlights the perturbations in
+cellular processes caused by the virus.
+
+\hypertarget{enrichment}{%
+\subsection{Enrichment}\label{enrichment}}
+
+The enrichment analysis module within the application addresses the
+challenge of selecting protein sets for mapping by incorporating a
+database-fetching functionality. This feature streamlines the enrichment
+analysis process and eliminates the need for users to manually select
+sets to map proteins against. The first variant of the enrichment
+analysis module utilizes the KEGG database, which offers a comprehensive
+collection of pathways and their interactions. By accessing this
+database, users can gain valuable insights into the underlying pathways
+and their relationships. This automated mapping against the KEGG
+database provides a convenient and efficient way to perform enrichment
+analysis. The second variant of the module caters to sophisticated users
+who may have specific protein sets in mind based on their research
+question. This flexibility allows users to select and incorporate their
+own protein sets into the analysis, enabling research in multiple
+contexts and providing a comprehensive insight into the dataset. Both
+variants of the enrichment analysis module present the results in
+interactive plots, offering users a dynamic and immersive experience
+with the data. Users can explore the enriched pathways and interact with
+the visualizations, gaining a deeper understanding of the relationships
+and connections within the dataset. Additionally, if desired, users can
+generate static plots and download them for use in presentations or
+other offline contexts. As indicated by the statistical findings, the
+KEGG enrichment analysis identified proteins associated with the
+ribosome and COVID-19. Both protein sets were connected to two ribosomal
+proteins, namely P47914 (60S ribosomal protein L29) and P46777 (60S
+ribosomal protein L5). These findings align with the fact that HIV and
+COVID-19, like all known viruses, depend entirely on the host cell's
+translation machinery for protein synthesis. This includes utilizing the
+host's ribosomes, tRNAs, amino acids, and all necessary factors for
+protein synthesis initiation, elongation, and termination
+\citep{Ohlmann2014}. Consequently, it is not surprising that these
+ribosomal proteins were found to be abundant in HIV-infected
+individuals.
+
+\hypertarget{limitations}{%
+\subsection{Limitations}\label{limitations}}
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/missingness-1}
+
+}
+
+\caption{Missingness over channels.
+Y-axis: Height of the histogram bars show number of channels.
+X-axis: fraction of missing values.
+Red line: Mean of number of missing values over all runs}\label{fig:missingness}
+ \endfigure\egroup
+
+Unfortunately, a substantial number of individual proteins in the data
+set (totaling 658) have missing values. On average, this amounts to
+approximately 53\% of the proteins across all runs and channels. This
+situation leads to the need for missing value imputation using
+techniques such as k-nearest neighbor (K-NN), which can result in
+averaged results. Additionally, it is important to note that this
+imputation process can introduce potential loss of power in subsequent
+statistical analysis. To improve future experiments, it is recommended
+to increase the number of proteins in the carrier proteome. By expanding
+the repertoire of proteins included in the carrier proteome, it can
+enhance the coverage and depth of analysis, providing a more
+comprehensive understanding of the biological system under
+investigation. This can potentially lead to more robust and accurate
+results in downstream analyses.
+
+\hypertarget{future-perspectives}{%
+\subsection{Future perspectives}\label{future-perspectives}}
+
+\hypertarget{clustering-for-stratification-of-cells.}{%
+\subsubsection{Clustering for stratification of
+cells.}\label{clustering-for-stratification-of-cells.}}
+
+The advantage of performing single-cell analysis lies in its ability to
+provide a fine resolution of cellular differences. Unfortunately
+variation causing effects such as cellular differentiation stages, are
+mainly unknown when starting single-cell experiments. The next steps in
+the development of the program aim to enable the stratification of
+single-cell data sets. The analysis typically begins by finding clusters
+within the datasets, identifying differentially expressed proteins, and
+annotating cell types based on their expression profiles. Subsequently,
+the analysis aims to distinguish sample types based on their expression
+profiles.
+
+Here I will explain how the analysis could start and elaborate the
+limitations and challenges for this approach. Additional processing
+steps will be included after the existing pipeline, so ProteoScanR
+provides a starting point for further analysis.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{Thesis_files/figure-latex/cluster_analysis_dendrogram-1}
+
+}
+
+\caption{Dendrogram showing the similarity over sample types. Tree was created using Ward`s minimum variance method.}\label{fig:cluster_analysis_dendrogram}
+ \endfigure\egroup
+
+As observed in the dendrogram, it is evident that sample type alone does
+not account for the variation observed in the data. The branching
+patterns in the dendrogram suggest that other factors might be
+significant for contributing to the averaging of protein expression
+levels. This can potentially lead to a decrease in the statistical power
+of tests aimed at detecting differential expression between sample
+types. Ideally, clustering approaches should result in individual
+clusters that capture distinct sample types, enabling the testing of
+differential expression within each cluster. This stratification of
+samples within clusters allows for the identification of unique
+expression patterns associated with specific biological variations. The
+dendrogram, as demonstrated above, is an example of hierarchical
+clustering, which provides a visual representation of the clustering
+results. In this context, users should have the flexibility to select
+the desired number of clusters that best explain the underlying
+biological variation in their dataset. This allows for a more targeted
+and precise analysis of differential expression within each cluster,
+facilitating the identification of biologically meaningful patterns and
+insights. To begin with the k-means clustering approach, it is crucial
+to determine the optimal number of clusters before conducting the
+analysis. This can be achieved using various approaches, one of which is
+the gap statistic method \citep{Tibshirani2001}.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster_analysis_gap_stat-1}
+
+}
+
+\caption{Optimal numbers of clusters calculated with gap statistics method}\label{fig:cluster_analysis_gap_stat}
+ \endfigure\egroup
+
+The gap statistic compares the total intracluster variation across
+various values of k (number of clusters) to the expected variation under
+a null reference distribution. The null reference distribution
+represents a scenario where no clear clustering patterns exist in the
+data. However as seen in the gap statistic a high number of clusters
+will be difficult to distinguish from each other and explore the
+biological meaning such as differentiation in the example macrophage
+dataset. Another approach used to determine the optimal number of
+clusters is the silhouette method.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster_analysis_silh-1}
+
+}
+
+\caption{Optimal number of clusters calculated with silhoutte method.}\label{fig:cluster_analysis_silh}
+ \endfigure\egroup
+
+The silhouette method calculates silhouette coefficients for each data
+point, which quantify how closely the point resembles its own cluster
+compared to other clusters. If the overall distances between clusters
+are small, indicating overlapping or unclear separation, it is suggested
+to take the minimum k. This suggests that the data points may not be
+well-separated into distinct clusters. As suggested by the silhouette
+method in this case a lower number of clusters to differentiate cell
+types is recommended.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster_analysis_dim_red-1}
+
+}
+
+\caption{PCA plot with cluster indication. Shape of the datapoints indicate sample types. Clusters are encircled and color coded.}\label{fig:cluster_analysis_dim_red}
+ \endfigure\egroup
+
+As seen in the principal components analysis, clusters are densely
+packed, and the plot shows no clear differentiation between the 8
+suggested clusters. The dense packing and poor resolution indicates that
+other factors may be influencing the lack of distinct separation. One of
+the other factors involves the large number of imputed values due the
+high rate of missing values in the protein expression set. K-NN is not
+able to impute biological differences and therefore averages the
+particular missing value by its k neighbors.
+
+This suggests the utilization of a different data set for developing the
+clustering module of the future program. By employing a data set with a
+lower rate of missing values in the protein expression set, the accuracy
+and reliability of the clustering analysis can be enhanced. Having a
+data set with less missing data will allow for a more precise
+identification of clusters and differentiation between cell types.
+
+\hypertarget{machine-learning}{%
+\subsubsection{Machine learning}\label{machine-learning}}
+
+One potential target for future research could involve leveraging
+artificial intelligence techniques to predict cellular behavior, cell
+types, or even diseases. By running the application on a server and
+creating a database, it would be possible to harness the power of AI
+algorithms such as decision trees, random forests, or support vector
+machines. Applying artificial intelligence to biological data has the
+potential to enhance the classification of cell types, similar to the
+earlier discussed clustering approach. However, it is important to note
+that training and validating these algorithms typically requires
+supervision and can be a time-consuming task.
+
+Another area of focus, not limited to research topics, could be to
+enhance the usability of the application. While a development time of 5
+months for the pipeline and user interface is sufficient to meet the
+analytical needs, there is always room for improvement in terms of
+user-friendliness. It is essential to consider the needs of all users,
+including the least experienced user. To prevent program crashes and
+errors, it may be necessary to implement try-catch around code blocks,
+as well as validate-need statements for the validation of user input.
+Additionally, creating a user-friendly environment involves establishing
+effective channels of communication between users and developers, such
+as utilizing platforms like GitHub for collaboration and feedback.
 
@@
 
-The conclusion section should specify the key findings of your study,
-explain their wider significance in the context of the research field
-and explain how you have filled the knowledge gap that you have
-identified in the introduction. This is your chance to present to your
-reader the major take-home messages of your dissertation research. It
-should be similar in content to the last sentence of your summary
-abstract. It should not be a repetition of the first paragraph of the
-discussion. They can be distinguished in their connection to broader
-issues. The first paragraph of the discussion will tend to focus on the
-direct scientific implications of your work (i.e.~basic science,
-fundamental knowledge) while the conclusion will tend to focus more on
-the implications of the results for society, conservation, etc.
+Through this project, I demonstrated that computational analysis can be
+made accessible to a wide range of biologists by providing an
+interactive environment. It is important to recognize that there cannot
+be a one-size-fits-all approach for proteomics datasets. Therefore, the
+application offers easy-to-interpret visualizations that allow users to
+assess the patterns and trends in their observed datasets. The default
+options provided in the application are based on successful analysis
+pipelines such as ScoPE2 \citep{Specht2021}, serving as a starting point
+for customization at multiple levels.
+
+The example dataset provided showcases the inherent challenges in
+proteomic data analysis. It exhibits skewed distributions, batch
+effects, and a substantial number of missing values. Despite these
+complexities, the ProteoScanR pipeline and Proteomics Workbench
+interface offer biologists a convenient means to address their research
+questions and validate the computations performed. The example use case
+also sheds light on the fact that previous proteomics publications, such
+as the ScoPE2 study \citep{Specht2021}, may not always employ the gold
+standard in data processing. This emphasizes the importance of
+critically evaluating and refining data processing steps through trial
+and error to achieve optimal results. The interactive environment
+provided by the application allows users to navigate the trial and error
+process without experiencing frustration. This user-friendly interface
+ensures that researchers can iteratively refine their data processing
+steps and analyze the results in a seamless manner. Furthermore, the
+scalability of the application was initially unknown until the final
+submission of the single-cell dataset. The programming and testing of
+the application primarily focused on bulk datasets, typically consisting
+of up to four runs at once. However, it is noteworthy that the
+application demonstrates remarkable efficiency and performance, even
+when running large datasets. The ability of the application to
+efficiently handle larger datasets, such as single-cell studies, is a
+significant advantage. It allows researchers to analyze complex and
+extensive datasets without sacrificing performance or compromising the
+quality of the results. This scalability makes the application a
+valuable tool for analyzing diverse types of biological data,
+accommodating the needs of researchers working with both smaller and
+larger datasets.
+
+ProteoScanR pipeline and Proteomics Workbench user interface serve as
+valuable tools for individual researchers and research communities,
+enabling them to analyze and visualize their data in meetings, even on
+non-performant computers. However, one of the future perspectives
+involves deploying the application on a server, making it accessible to
+research groups around the world. This would enhance collaboration and
+facilitate the sharing of insights and findings across different
+geographical locations. By providing a centralized platform for
+proteomics analysis, aiming to foster scientific advancements and enable
+researchers to make significant contributions to their respective
+fields. Furthermore by having a central hub for proteomic data analysis
+and the use of artificial intelligence the biological meaning could be
+brought to a wider context. Since proteomic data could be saved without
+violating data protection laws, algorithms could be employed to learn
+from large datasets.
+
+An important aspect of learning from the data involves annotating
+specimens, particularly in the context of single-cell proteomics. The
+clustering approach is a primary focus for future development and
+analysis in the near term. By employing clustering algorithms and
+stratifying the data, individual cells can be labeled and annotated
+based on their phenotypes. This annotation process involves associating
+protein sets collections with specific clusters or cell types, providing
+a more comprehensive understanding of cellular heterogeneity and
+functional characteristics.
+
+In conclusion, the ProteoScanR pipeline and Proteomics Workbench
+interface provide a user-friendly and efficient approach for proteomic
+data analysis, enabling biologists to address their research questions
+and validate computations. Future perspectives include deploying the
+application on a server for global accessibility, leveraging artificial
+intelligence techniques, and implementing clustering approaches to
+further enhance annotation and understanding of cellular heterogeneity
+and functional characteristics.
 
@@
 
-\hypertarget{appendixces}{%
-\section{Appendix(ces)}\label{appendixces}}
-
-\hypertarget{appendix-a-additional-tables}{%
-\subsection{Appendix A: additional
-tables}\label{appendix-a-additional-tables}}
-
-Insert content for additional tables here.
-
-\newpage
-
-\hypertarget{appendix-b-additional-figures}{%
-\subsection{Appendix B: additional
-figures}\label{appendix-b-additional-figures}}
-
-Insert content for additional figures here.
-
-\newpage
-
-\hypertarget{appendix-c-code}{%
-\subsection{Appendix C: code}\label{appendix-c-code}}
-
-Insert code (if any) used during your dissertation work here.
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+\hypertarget{wet-lab}{%
+\subsection{Wet lab}\label{wet-lab}}
+
+\hypertarget{cells-sorting-protocol}{%
+\subsubsection{Cells sorting protocol}\label{cells-sorting-protocol}}
+
+\begin{enumerate}
+\item 96 single cells from each donor were sorted into a single plate. Each well contained a single cell. As such, 12 different plates containing single cells from 12 different donors were prepared. 
+\item Carrier plates were also prepared. In the carrier plates, 100 cells from one donor were sorted into each well. 24 such wells (3 columns) were prepared for each donor.
+\end{enumerate}
+
+\hypertarget{digestion-protocol}{%
+\subsubsection{Digestion protocol}\label{digestion-protocol}}
+
+\begin{enumerate}
+\item Prepare 25 ng/ $\mu$L trypsin in 100 mmol/L Triethylammonium bicarbonate buffer (TEAB).
+\item Add 25 ng/$\mu$L trypsin: 2 $\mu$L to the carrier samples, 1 $\mu$L to the single cell samples.
+\item Seal the plates.
+\item Centrifuge (2 min at 120 G).
+\item Incubate overnight at 37 °C.
+\end{enumerate}
+
+\hypertarget{labeling-protocol}{%
+\subsubsection{Labeling protocol}\label{labeling-protocol}}
+
+\begin{enumerate}
+\item Let the TMTpro labels come to room temperature so that all the condensation on the tube is gone.
+Dissolve the TMTpro reagents in anhydrous acetonitril (ACN) to obtain a concentration of 10 $\mu$g/$\mu$L.
+\item Add 1 $\mu$L of TMTpro reagents according to the tables. For one set of plates, you will need 16 $\mu$L of each label. It is recommended to aliquot a slightly larger volume, around 20 $\mu$L.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Prepare a 5% hydroxylamine solution in water.
+\item Add 1 $\mu$L of the 5% hydroxylamine solution to each well.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Cover the plates with cap strips.
+\end{enumerate}
+
+\hypertarget{pooling-protocol}{%
+\subsubsection{Pooling protocol}\label{pooling-protocol}}
+
+\begin{enumerate}
+\item Wash a 10 $\mu$L glass microsyringe five times with ACN.
+\item Use the microsyringe to combine TMTpro-labeled samples into 36 LC autosampler vials with built-in glass inserts.
+\item Combine six carrier wells from each sample and add the carrier proteome to the 36 TMTpro sets.
+\item Combine the first set by taking well A1 from the first single-cell plate, well A2 from the second plate, well A3 from the third plate, and so on.
+For the next set, take well A1 from the second plate, well A2 from the third plate, and so on.
+Combine 12 sets following this scheme.
+Combine the next 24 sets by following the same scheme for rows B and C.
+\item Vacuum centrifuge the combined sets to dryness.
+\item Dissolve the dried samples in 7 $\mu$L of buffer A and transfer them to LC-MS/MS for further analysis.
+\end{enumerate}
+
+\hypertarget{dry-lab}{%
+\subsection{Dry lab}\label{dry-lab}}
+
+\hypertarget{r-version-info}{%
+\subsubsection{R-version info}\label{r-version-info}}
+
+\begin{verbatim}
+R version 4.3.0 (2023-04-21)
+Platform: x86_64-pc-linux-gnu (64-bit)
+Running under: Ubuntu 22.04.2 LTS
+
+ [1] visNetwork_2.1.2            snowfall_1.84-6.2          
+ [3] snow_0.4-4                  piano_2.16.0               
+ [5] org.Hs.eg.db_3.17.0         AnnotationDbi_1.62.1       
+ [7] infotheo_1.2.0.1            clusterProfiler_4.8.1      
+ [9] plotly_4.10.1               sva_3.48.0                 
+[11] BiocParallel_1.34.2         genefilter_1.82.1          
+[13] mgcv_1.8-42                 nlme_3.1-162               
+[15] impute_1.74.1               CONSTANd_1.8.0             
+[17] limma_3.56.1                scater_1.28.0              
+[19] scuttle_1.10.1              reshape2_1.4.4             
+[21] dplyr_1.1.2                 magrittr_2.0.3             
+[23] ggplot2_3.4.2               SingleCellExperiment_1.22.0
+[25] scp_1.10.0                  QFeatures_1.10.0           
+[27] MultiAssayExperiment_1.26.0 SummarizedExperiment_1.30.1
+[29] Biobase_2.60.0              GenomicRanges_1.52.0       
+[31] GenomeInfoDb_1.36.0         IRanges_2.34.0             
+[33] S4Vectors_0.38.1            BiocGenerics_0.46.0        
+[35] MatrixGenerics_1.12.0       matrixStats_0.63.0         
+[37] spsComps_0.3.2.1            DT_0.28                    
+[39] shinydashboard_0.7.2        shinyWidgets_0.7.6         
+[41] shiny_1.7.4                
+\end{verbatim}
+
+\newpage
+\listoffigures
 

### 2023-06-26 markdown/methods.Rmd
cite file added

@@
 ## Drylab - Computational methods
+
+### Hardware
+Lenovo Laptop X360 with:
+Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz
+8 GB RAM
+
 ```{r Dataflow, eval=FALSE, include=FALSE}

### 2023-06-26 markdown/results.Rmd
cite file added

@@
 
+After a runtime of 2 minutes on the 4 core laptop the data is ready to visualize.
 On the left side of the user interface, users can find the analysis pipeline settings that can be adjusted.
@@
 
-The position of the dot shows the number of matches for the particular protein set. Colors indicate 
-for the significance level. In this analysis the entire repertoire of detected proteins was used. Only the protein set collections with the most protein matches were used.
+The position of the dot shows the number of matches for the particular protein set.
+Colors indicate for the significance level.
+In this analysis the entire repertoire of detected proteins was used and fetched against the bundle containing all Human gene sets in the Human Molecular Signatures Database (MSigDB).
+After a runtime of 3 minutes 672 of 24667 gene sets were found significantly over represented.
+Only the protein set collections with the most protein matches were used (snipping tool Plotly).

### 2023-06-26 markdown/Thesis.Rmd
formating

@@
 {\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
-presenting ProteoScanR \& Proteomics Workbench
+\\ presenting ProteoScanR \& Proteomics Workbench
 
@@
 \singlespacing
-By
 
@@
 {\bf Lukas Gamp}
-01630003
+\\ 01630003
 
@@
 
-in partial fulfillment of the requirement \\for the degree of MSc \\in Bioinformatics
+In partial fulfillment of the requirement \\for the degree of MSc \\in Bioinformatics
 
@@
 
-Main Supervisor: 
-Dr. DI(FH) Gerhard Duernberger
-
-External Supervison:
-Assoc. Prof. Ujjwal Neogi, M.Sc. PhD
-
-Second examiner:
-FH-Prof. Dr. Alexandra Graf
+\doublespacing
+\doublespacing
+Main Supervisor: Dr. DI(FH) Gerhard Duernberger \\
+External Supervison: Assoc. Prof. Ujjwal Neogi, M.Sc. PhD \\
+Second examiner: FH-Prof. Dr. Alexandra Graf \\
 

### 2023-06-26 markdown/appendix.Rmd
results until stats finished
formating

@@
 
-## Cells sorting protocol
+## Wet lab
+### Cells sorting protocol
 \begin{enumerate}
@@
 
+## Dry lab
 ### R-version info
 
-\begin{verbatim}
-R version 4.3.0 (2023-04-21)
-Platform: x86_64-pc-linux-gnu (64-bit)
-Running under: Ubuntu 22.04.2 LTS
+```{r load_libraries, message=FALSE, warning=FALSE, include=FALSE}
+library("shiny")
+library("shinyWidgets")
+library("shinydashboard")
+library("DT")
+library("spsComps")
 
- [1] visNetwork_2.1.2            snowfall_1.84-6.2          
- [3] snow_0.4-4                  piano_2.16.0               
- [5] org.Hs.eg.db_3.17.0         AnnotationDbi_1.62.1       
- [7] infotheo_1.2.0.1            clusterProfiler_4.8.1      
- [9] plotly_4.10.1               sva_3.48.0                 
-[11] BiocParallel_1.34.2         genefilter_1.82.1          
-[13] mgcv_1.8-42                 nlme_3.1-162               
-[15] impute_1.74.1               CONSTANd_1.8.0             
-[17] limma_3.56.1                scater_1.28.0              
-[19] scuttle_1.10.1              reshape2_1.4.4             
-[21] dplyr_1.1.2                 magrittr_2.0.3             
-[23] ggplot2_3.4.2               SingleCellExperiment_1.22.0
-[25] scp_1.10.0                  QFeatures_1.10.0           
-[27] MultiAssayExperiment_1.26.0 SummarizedExperiment_1.30.1
-[29] Biobase_2.60.0              GenomicRanges_1.52.0       
-[31] GenomeInfoDb_1.36.0         IRanges_2.34.0             
-[33] S4Vectors_0.38.1            BiocGenerics_0.46.0        
-[35] MatrixGenerics_1.12.0       matrixStats_0.63.0         
-[37] spsComps_0.3.2.1            DT_0.28                    
-[39] shinydashboard_0.7.2        shinyWidgets_0.7.6         
-[41] shiny_1.7.4                
-\end{verbatim}
+library("scp")
+library("SingleCellExperiment")
+library("ggplot2")
+library("magrittr")
+library("dplyr")
+library("reshape2")
+library("scater")
+library("limma")
+library("CONSTANd")
+library("stats")
+library("impute")
+library("sva")
+library("plotly")
+library("clusterProfiler")
+library("infotheo")
+library("org.Hs.eg.db")
+library("AnnotationDbi")
+library("piano")
+library("snowfall")
+library("parallel")
+library("visNetwork")
+```
+
+```{r session_info, echo=FALSE}
+sessionInfo()
+```

### 2023-06-26 markdown/methods.Rmd
formating

@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=17, fig.align='center', fig.cap="Flowchart presenting the processing by ProteoScanR"}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="90%", fig.align='center', fig.cap="Flowchart presenting the processing by ProteoScanR"}
 library(Gmisc, quietly = TRUE)

### 2023-06-26 markdown/discussion.Rmd
tried to fix clusters

@@
 Here I will explain how the analysis could start and elaborate the limitations and challenges for this approach.
-Additional processing steps will be included after the existing pipeline, so ProteoScanR provides a starting point for further analysis.
+Additional processing steps will be included after the existing pipeline, so the ProteoScanR provides a starting point for further analysis.
 
@@
 
-km.final <- kmeans(protein_matrix_t, optimal_clusters)
+km.final <- kmeans(protein_matrix_t, 2)
 
@@
                geom="polygon", alpha=0.2, level=0.9, type="t", linetype=2, show.legend = F) +
-  coord_fixed()
+  coord_fixed() +
+  labs(title=paste("Principle Components with", 2, "clusters"))
 

### 2023-06-26 markdown/methods.Rmd
tried to fix clusters

--- 2023-06-26 markdown/methods.Rmd
@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=17, fig.align='center', fig.cap="Flowchart presenting the processing by ProteoScanR"}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="80%", fig.align='center', fig.cap="Flowchart presenting the processing by ProteoScanR"}
 library(Gmisc, quietly = TRUE)
@@
 
-```{r data_processing_pipeline_flowchart_horizontal, eval=FALSE, fig.height=9, fig.width=32, include=FALSE}
+```{r data_processing_pipeline_flowchart_horizontal, eval=FALSE, fig.height=14, fig.width=45, include=FALSE}
 library(Gmisc, quietly = TRUE)
@@
 
-default_txt_setting <- gpar(cex=1.5)
+default_txt_setting <- gpar(cex=1.9)
 default_box_setting <- gpar(fill="white")
@@
 
-batch_correction <- boxGrob("batch correction",
+batch_correction <- boxGrob("batch correction &\n dimensionality reduction",
                             box_gp=plot_box_setting,
@@
 grid.newpage()
-
 vert <- {spreadHorizontal(files = evidence_file,
@@
                        missing_value_handling = missing_value_handling,
-                       batch_correction = batch_correction,
-                       dim_red = dim_red,
-                       stat_module
+#                       batch_correction = batch_correction,
+#                       dim_red = dim_red,
+                       out = stat_module
 )}
-
 input <- alignHorizontal(reference = vert$files,
@@
 
-# connect and move the boxes
-for (i in 1:(length(vert) - 1)) {
-    connectGrob(vert[[i]], vert[[i + 1]], type = "horizontal") %>%
-    print}
+output <- alignHorizontal(reference = vert$out,
+                      stat_module, batch_correction) %>%
+  spreadVertical()
+vert$out <- NULL
 
@@
 qc1_content <- moveBox(qc1_content,
-                    x = coords(vert$quality_control_1)$x,
-                    y = 0.8)
+                    y = .8,
+                    x = coords(vert$quality_control_1)$x)
 
@@
 
-missing_value_opts <- moveBox(missing_value_opts,
-                    y = .7,
+# dim_red <- moveBox(dim_red,
+#                    x = coords(output[[2]])$x,
+#                    y = 0)
+
+batch_correction_opts <- moveBox(batch_correction_opts,
+                    y = coords(output[[2]])$y,
                     x = coords(vert$missing_value_handling)$x)
 
-dim_red_actions <- moveBox(dim_red_actions,
-                    y = .9,
-                    x = coords(vert$dim_red)$x)
+# dim_red_actions <- moveBox(dim_red_actions,
+#                     x = (coords(dim_red)$x -.5),
+#                     y = coords(dim_red)$x)
 
@@
 qc2_content <- moveBox(qc2_content,
-                    y = .2,
+                    y = .17,
                     x = coords(vert$quality_control_2)$x)
@@
 
-batch_correction_opts <- moveBox(batch_correction_opts,
+missing_value_opts <- moveBox(missing_value_opts,
                     y = .2,
-                    x = coords(vert$batch_correction)$x)
+                    x = coords(vert$missing_value_handling)$x)
+
+for (i in 1:(length(vert) - 1)) {
+  connectGrob(vert[[i]], vert[[i + 1]], type = "horizontal") %>%
+    print
+}
 
 # connect boxes
-connectGrob(input[[1]], vert$read_data, type = "vertical")
-connectGrob(input[[2]], vert$read_data, type = "vertical")
+connectGrob(input[[1]], vert$read_data, type = "N")
+connectGrob(input[[2]], vert$read_data, type = "N")
 
@@
 
-connectGrob(vert$batch_correction, batch_correction_opts, type = "vertical", arrow_obj = refinement_arrow_setting)
+connectGrob(vert$missing_value_handling, output[[1]], type = "vertical", subelmnt = "right")
+connectGrob(vert$missing_value_handling, output[[2]], type = "vertical", subelmnt = "right")
 
-connectGrob(vert$dim_red, dim_red_actions, type = "vertical", arrow_obj = arrow(angle=0))
+connectGrob(output[[2]], batch_correction_opts, type = "horizontal", arrow_obj = arrow(angle=0), subelmnt = "l")
+
+# connectGrob(vert$dim_red, dim_red_actions, type = "horizontal", arrow_obj = arrow(angle=0))
 
@@
 Enriched pathways can be understood as an over-representation (ORA = over-representation analysis) of proteins within a specific set.
-This set can be a pathway or a selection of common denominators in biological contexts, such as cell types or oncological factors.
+This set, also known as ontology can be a pathway or a selection of common denominators in biological contexts, such as cell types or oncological factors.
 The statistical testing to identify the overabundant proteins of a specific set is similar to a $\chi^2$ test developed by Karl Pearson \citep{Pearson1900} or the exact Fisher test \citep{Sprent2011} for tables with 4 fields in small sample sizes.
-Within the list are all truly differentially expressed genes, an annotation is the identification of a protein in the desired biological context.
+Within the list are all truly differential expressed genes, an annotation is the identification of a protein in the desired biological context.
 The table is expected to be populated by the values as follows:
@@
 Once the enriched protein pathways are determined, they can be visually represented using bar or dotplots, and a network of the found pathways.
-In the network ontologies are depicted as nodes, whereas edges show the strength of the interaction by thickness corresponding to the number of associated proteins.
 
@@
 In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
+The online database offers pre-built gene set collections for previous research questions in multiple fields of human biology.
 For statistical testing the above explained Fisher exact test is used.
+The output of the Fisher test does not show any directionality in regards of protein expression.
+If the user wants to include directionality in the network diagram, the default option of piano can be included in future versions of the program.

### 2023-06-26 markdown/discussion.Rmd
tried to fix clusters

@@
 
-km.final <- kmeans(protein_matrix_t, 2)
+#optimal_clusters <- 4
+
+km.final <- kmeans(protein_matrix_t, optimal_clusters)
 
@@
 
-dimred_pca$cluster <- scp$cluster$km.final.cluster
+dimred_pca$cluster <- factor(scp$cluster$km.final.cluster) 
+dimred_pca$SampleType <- scp$SampleType
 
@@
                        y=PC2,
-                       shape = scp$SampleType)) +
+                       shape = SampleType,
+                       color=cluster)) +
   geom_point() +
   guides(color="none", shape="none") +
-  stat_ellipse(data=dimred_pca, aes(fill=factor(cluster)),
-               geom="polygon", alpha=0.2, level=0.9, type="t", linetype=2, show.legend = F) +
-  coord_fixed() +
-  labs(title=paste("Principle Components with", 2, "clusters"))
+  labs(title=paste("Principle Components with", optimal_clusters, "clusters")) +
+  stat_ellipse(data = dimred_pca, 
+               aes(fill = cluster),
+               geom = "polygon", 
+               alpha = 0.2, 
+               level = 0.9, 
+               type = "norm", 
+               linetype = 2, 
+               show.legend = F,
+               na.rm = T) +
+  coord_fixed()
 
+# ggplot(dimred_pca, aes(x=PC1, 
+#                        y=PC2,
+#                        shape = SampleType
+#                        )) +
+#   geom_mark_ellipse(aes(fill=cluster, color=cluster)) +
+#   geom_point()
 ```

### 2023-06-27 markdown/discussion.Rmd
minor bugfix

@@
 
-#optimal_clusters <- 4
+#optimal_clusters <- 2
 
@@
   geom_point() +
-  guides(color="none", shape="none") +
+#  guides(color="none", shape="none") +
   labs(title=paste("Principle Components with", optimal_clusters, "clusters")) +
-  stat_ellipse(data = dimred_pca, 
-               aes(fill = cluster),
-               geom = "polygon", 
-               alpha = 0.2, 
-               level = 0.9, 
-               type = "norm", 
-               linetype = 2, 
-               show.legend = F,
-               na.rm = T) +
+  # stat_ellipse(data = dimred_pca, 
+  #              aes(fill = cluster),
+  #              geom = "polygon", 
+  #              alpha = 0.2, 
+  #              level = 0.9, 
+  #              type = "norm", 
+  #              linetype = 2, 
+  #              show.legend = F,
+  #              na.rm = T) +
   coord_fixed()
-
-# ggplot(dimred_pca, aes(x=PC1, 
-#                        y=PC2,
-#                        shape = SampleType
-#                        )) +
-#   geom_mark_ellipse(aes(fill=cluster, color=cluster)) +
-#   geom_point()
 ```

### 2023-07-17 markdown/discussion.Rmd
app on desktop pc established

@@
 
+folder <- "/media/lukas/DAB0041DB003FEAF/MS-Data"
+
 # read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/evidence.txt")
+mqScpData <- read.delim("/media/lukas/DAB0041DB003FEAF/MS-Data/Lukas/Jun13/combined/txt/evidence.txt")
 
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/sampleAnnotation.txt")
+sampleAnnotation = read.delim("/media/lukas/DAB0041DB003FEAF/MS-Data/Lukas/Jun13/combined/txt/sampleAnnotation.txt")
 

### 2023-07-18 markdown/appendix.Rmd
headlines added. appendix moved to methods

--- 2023-02-02 markdown/appendix.Rmd
@@
 ---
-# Appendix(ces)
+# Appendix
 
-## Appendix A: additional tables
+## Dry lab
+### R-version info
 
-Insert content for additional tables here.
+```{r load_libraries, message=FALSE, warning=FALSE, include=FALSE}
+library("shiny")
+library("shinyWidgets")
+library("shinydashboard")
+library("DT")
+library("spsComps")
 
-\newpage
+library("scp")
+library("SingleCellExperiment")
+library("ggplot2")
+library("magrittr")
+library("dplyr")
+library("reshape2")
+library("scater")
+library("limma")
+library("CONSTANd")
+library("stats")
+library("impute")
+library("sva")
+library("plotly")
+library("clusterProfiler")
+library("infotheo")
+library("org.Hs.eg.db")
+library("AnnotationDbi")
+library("piano")
+library("snowfall")
+library("parallel")
+library("visNetwork")
+```
 
-## Appendix B: additional figures
-
-Insert content for additional figures here.
-
-\newpage
-
-## Appendix C: code
-
-Insert code (if any) used during your dissertation work here.
+```{r session_info, echo=FALSE}
+sessionInfo()
+```

### 2023-07-18 markdown/discussion.Rmd
headlines added. appendix moved to methods

--- 2023-06-27 markdown/discussion.Rmd
@@
 # read in MS result table
-mqScpData <- read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/evidence.txt")
+mqScpData <- read.delim("/home/lukas/Desktop/MS_data/example/single_cell/evidence.txt")
 
-sampleAnnotation = read.delim("/home/lukas/Desktop/MS-Data/Lukas/Jun13/combined/txt/sampleAnnotation.txt")
+sampleAnnotation = read.delim("/home/lukas/Desktop/MS_data/example/single_cell/sampleAnnotation.txt")
 

### 2023-07-18 markdown/methods.Rmd
minor bugfix
headlines added. appendix moved to methods

@@
 ## Wetlab - Materials and Methods
-For testing and development of the ProteoScanR pipeline the following materials and methods were performed. For the sample preparation protocol see appendix.
+For testing and development of the ProteoScanR pipeline the following materials and methods were performed.
 
@@
 Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome. Cell sorting was performed with FACSAria Fusion (BD Biosciences).
+
+#### Cells sorting protocol
+\begin{enumerate}
+\item 96 single cells from each donor were sorted into a single plate. Each well contained a single cell. As such, 12 different plates containing single cells from 12 different donors were prepared. 
+\item Carrier plates were also prepared. In the carrier plates, 100 cells from one donor were sorted into each well. 24 such wells (3 columns) were prepared for each donor.
+\end{enumerate}
 
@@
 
+#### Digestion protocol
+\begin{enumerate}
+\item Prepare 25 ng/ $\mu$L trypsin in 100 mmol/L Triethylammonium bicarbonate buffer (TEAB).
+\item Add 25 ng/$\mu$L trypsin: 2 $\mu$L to the carrier samples, 1 $\mu$L to the single cell samples.
+\item Seal the plates.
+\item Centrifuge (2 min at 120 G).
+\item Incubate overnight at 37 °C.
+\end{enumerate}
+
 ### Labeling
@@
 
+#### Labeling protocol
+\begin{enumerate}
+\item Let the TMTpro labels come to room temperature so that all the condensation on the tube is gone.
+Dissolve the TMTpro reagents in anhydrous acetonitril (ACN) to obtain a concentration of 10 $\mu$g/$\mu$L.
+\item Add 1 $\mu$L of TMTpro reagents according to the tables. For one set of plates, you will need 16 $\mu$L of each label. It is recommended to aliquot a slightly larger volume, of around 20 $\mu$L.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Prepare a 5% hydroxylamine solution in water.
+\item Add 1 $\mu$L of the 5% hydroxylamine solution to each well.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Cover the plates with cap strips.
+\end{enumerate}
+
 ### Pooling
 Carrier proteome was added to the 36 TMTpro sets before pooling sample types for each of the 36 MS-runs.
+
+#### Pooling protocol
+\begin{enumerate}
+\item Wash a 10 $\mu$L glass microsyringe five times with ACN.
+\item Use the microsyringe to combine TMTpro-labeled samples into 36 LC autosampler vials with built-in glass inserts.
+\item Combine six carrier wells from each sample and add the carrier proteome to the 36 TMTpro sets.
+\item Combine the first set by taking well A1 from the first single-cell plate, well A2 from the second plate, well A3 from the third plate, and so on.
+For the next set, take well A1 from the second plate, well A2 from the third plate, and so on.
+Combine 12 sets following this scheme.
+Combine the next 24 sets by following the same scheme for rows B and C.
+\item Vacuum centrifuge the combined sets to dryness.
+\item Dissolve the dried samples in 7 $\mu$L of buffer A and transfer them to LC-MS/MS for further analysis.
+\end{enumerate}
 
@@
                          txt_gp = default_txt_setting)
-sample_annotation_file <- boxGrob("sample annotation",
+sample_annotation_file <- boxGrob("sample \nannotation",
                                   box_gp = data_box_setting,

### 2023-07-18 markdown/results.Rmd
headlines added. appendix moved to methods

@@
 
-```{r ui_summary, echo=F, out.width="100%", fig.align='center', fig.cap="Summary showing the process of the analysis pipeline"}
+## Pre-processing and quality control
+
+```{r ui_summary, echo=F, out.width="100%", fig.align='center', fig.cap="Summary showing the processing steps of the analysis pipeline"}
 knitr::include_graphics("screenshots/overview.png")
@@
 
+## Downstream analysis
+
 \newpage
@@
 
+## Statistics
+
 \newpage

### 2023-07-18 markdown/Thesis.Rmd
tried to fix clusters
cross references added

@@
 output:
-  pdf_document:
-    number_sections: TRUE
+  bookdown::pdf_document2:
+    toc: false
+#    number_sections: TRUE
 geometry: "left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm"
@@
 Main Supervisor: Dr. DI(FH) Gerhard Duernberger \\
-External Supervison: Assoc. Prof. Ujjwal Neogi, M.Sc. PhD \\
+External Supervisor: Assoc. Prof. Ujjwal Neogi, M.Sc. PhD \\
 Second examiner: FH-Prof. Dr. Alexandra Graf \\

### 2023-07-18 markdown/discussion.Rmd
cross references added

@@
 
-```{r cluster_analysis_dendrogram, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%", fig.align='center', fig.cap="Dendrogram showing the similarity over sample types.
+```{r cluster-analysis-dendrogram, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%", fig.align='center', fig.cap="Dendrogram showing the similarity over sample types.
 Tree was created using Ward`s minimum variance method."}
@@
 
-```{r cluster_analysis_gap_stat, echo=FALSE, message=FALSE, warning=FALSE, out.width="90%", fig.align='center', fig.cap="Optimal numbers of clusters calculated with gap statistics method"}
+```{r cluster-analysis-gap-stat, echo=FALSE, message=FALSE, warning=FALSE, out.width="90%", fig.align='center', fig.cap="Optimal numbers of clusters calculated with gap statistics method"}
 
@@
 
-```{r cluster_analysis_silh, echo=FALSE, message=FALSE, warning=FALSE, out.width="90%", fig.align='center', fig.cap="Optimal number of clusters calculated with silhoutte method."}
+```{r cluster-analysis-silh, echo=FALSE, message=FALSE, warning=FALSE, out.width="90%", fig.align='center', fig.cap="Optimal number of clusters calculated with silhoutte method."}
 
@@
 
-```{r cluster_analysis_dim_red, echo=FALSE, message=FALSE, warning=FALSE,  out.width="90%", fig.align='center', fig.cap="PCA plot with cluster indication.
+```{r cluster-analysis-dim-red, echo=FALSE, message=FALSE, warning=FALSE,  out.width="90%", fig.align='center', fig.cap="PCA plot with cluster indication.
 Shape of the datapoints indicate sample types.

### 2023-07-18 markdown/methods.Rmd
cross references added

@@
 
-```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="80%", fig.align='center', fig.cap="Flowchart presenting the processing by ProteoScanR"}
+```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="80%", fig.align='center', fig.cap= "Flowchart presenting the processing by ProteoScanR"}
 library(Gmisc, quietly = TRUE)

### 2023-07-18 markdown/results.Rmd
cross references added

--- 2023-06-26 markdown/results.Rmd
@@
 
-To begin using the application, the first step is to upload the tab-delimited evidence file and the sample annotation file.
+To begin using the application, the first step is to upload the tab-delimited evidence file and the sample annotation file after processing the data with MaxQuant.
 Once these files are uploaded, the application is ready to run with default values.
@@
 
-```{r ui_summary, echo=F, out.width="100%", fig.align='center', fig.cap="Summary showing the process of the analysis pipeline"}
+## Pre-processing and quality control
+The initial stages of the pipeline are focused on creating a reliable dataset for analysis that is free from noise and bias.
+
+```{r ui-summary, echo=F, out.width="100%", fig.align='center', fig.cap="Summary showing the processing steps of the analysis pipeline"}
 knitr::include_graphics("screenshots/overview.png")
@@
 After a runtime of 2 minutes on the 4 core laptop the data is ready to visualize.
-On the left side of the user interface, users can find the analysis pipeline settings that can be adjusted.
+On the left side of the user interface (see figure \@ref(fig:ui-summary)), users can find the analysis pipeline settings that can be adjusted.
 Once the desired changes are made, they can be applied by clicking the update button.
@@
 \newpage
-```{r ui_numbers_default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
+```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
 knitr::include_graphics("screenshots/numbers_default.png")
 ```
-
-The default settings of the analysis pipeline resulted in a total number of unique proteins below approximately $10^3$, which were aggregated from around $10^{3.5}$ peptides.
-Before these peptides were aggregated over approximately $10^{4.5}$ peptide spectrum matches.
-The aggregation method used is fixed by median.
+In order to get an idea of the magnitude of the data set, the barplot provides an overview (see figure \@ref(fig:ui-numbers-default))
+The default settings of the analysis pipeline resulted in a total number of unique proteins below approximately 1,000, which were aggregated from around 3,000 peptides.
+Before these peptides were aggregated over approximately 32,000 peptide spectrum matches.
+The aggregation for both steps was performed by the column median.
 
 \newpage
-```{r ui_mutual_info_default, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with default setting"}
+```{r ui-mutual-info-default, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with default setting"}
 knitr::include_graphics("screenshots/mutual_info_default.png")
 ```
+One of the first steps of quality control is to assess the loss of information in the observed data set.
+The entropy based analysis gives insight in impact the computations had on the initial data set.
+When exploring the intra and inter groupwise comparison as a boxplot (see figure \@ref(fig:ui-mutual-info-default)), the user can elaborate the loss in mutual information.
 
@@
 
+Therefore, it is advisable to experiment with various normalization techniques.
+
 \newpage
-```{r ui_constand_norm, echo=F, out.width="100%", fig.align='center', fig.cap="Popup showing the dependencies for the CONSTANd normalization"}
+```{r ui-constand-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Popup showing the dependencies for the CONSTANd normalization"}
 knitr::include_graphics("screenshots/constand_norm_dependency.png")
@@
 
+As seen in the dependency popup (see figure \@ref(fig:ui-constand-norm)) for the CONSTANd normalization \citep{VanHoutven2021}, requirements are not fulfilled.
+The MA-plot shows a strong heteroskedasticity indicated by the red trendline.
+
 \newpage
-```{r ui_mutual_info_quant_norm, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with quantile normalization"}
+```{r ui-mutual-info-quant-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with quantile normalization"}
 knitr::include_graphics("screenshots/mutual_info_quant_norm.png")
 ```
+When switching to the quantile normalization the loss of mutual information decreased.
+Therefore the advised normalization technique for this particular dataset would be the quantile normalization (see figure \@ref(fig:ui-mutual-info-quant-norm) compared to figure \@ref(fig:ui-mutual-info-default)).
 
@@
 \newpage
-```{r ui_RI_channel_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on channel.
+```{r ui-RI-channel-default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on channel.
 The selected factor will be highlighted and indicated in the legend."}
@@
 
+In order to check if the TMTpro correction in MaxQuant removed the signal spillage in other channels correctly, it is recommended to compare channel intensity (see figure \@ref(fig:ui-RI-channel-default)).
 When comparing the median reporter ion intensities across channels, no systematic bias is observed.
@@
 \newpage
-```{r ui_RI_file_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on file"}
+```{r ui-RI-file-default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on file"}
 knitr::include_graphics("screenshots/RI_file_default.png")
 ```
+The reporter ion intensity highlights also differences on raw files, which is advisable to observe prior to any statistics or dimensionality reduction.
 
-Comparing the median reporter ion intensities with a focus on the batch (=Raw.file), it becomes evident that there are significant changes in signal that need to be taken into account in further analysis.
+Comparing the median reporter ion intensities with a focus on the batch (=Raw.file), it becomes evident that there are significant changes in signal that need to be taken into account in further analysis (see figure \@ref(fig:ui-RI-file-default)).
 After applying quality control measures, several values have been lost due to cutoffs in parameters such as parental ion fraction (PIF), q-value, coefficient of variation, and missingness of peptide data.
@@
 \newpage
-```{r ui_RI_st_default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on sampletype"}
+```{r ui-RI-st-default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on sampletype"}
 knitr::include_graphics("screenshots/RI_st_default.png")
 ```
+Comparing sample types in regards of reporter ion intensity is recommended to observe any experimental bias such as batch/channel confounding effects which would influence the statistical model afterwards.
 
 The sample type does not appear to have an impact on the median reporter ion intensities, as suggested by the overlapping interquartile ranges (IQRs).
-Nevertheless, the HIV group with metabolic syndrome displays a broader range of values spanning tenfold in arbitrary units.
+Nevertheless, the HIV group with metabolic syndrome displays a broader range of values spanning tenfold in arbitrary units (see figure \@ref(fig:ui-RI-st-default)).
 
 \newpage
-```{r ui_cv_st_default, echo=F, out.width="100%", fig.align='center', fig.cap="Median covariance of razor proteins with selection on sampletype"}
+```{r ui-cv-st-default, echo=F, out.width="100%", fig.align='center', fig.cap="Median covariance of razor proteins with selection on sampletype"}
 knitr::include_graphics("screenshots/cv_st_default.png")
 ```
+The distribution of median coefficient of variation (CV) shows the uniqueness in peptides for their respective razor peptides.
+This metric represents the presence or absence of noisy matches in the data set.
 
@@
 The minimum number of observations determines the threshold for considering a razor protein ambiguous, based on the number of matching peptides.
-In this case, the reference publication set the threshold at 5 observations.
+In this case, the recommended minimum number of observation of 5 (by \citep{Specht2021}) was used.
 
-The interquartile ranges (IQRs) remained unchanged by this cutoff.
+The interquartile ranges (IQRs) remained unchanged by this cutoff (see figure \@ref(fig:ui-cv-st-default)).
 The sample type did not influence the CV of razor proteins.
@@
 \newpage
-```{r ui_cv_rawfile_default, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file"}
+```{r ui-cv-rawfile-default, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file"}
 knitr::include_graphics("screenshots/cv_rawfile_default.png")
 ```
+Razor peptide ambiguity over multiple batches can indicate experimental failures during sample prep or during acquisition.
 
-The plot illustrates that certain batches (=raw files) exhibit a higher number of distinct matches compared to others, as indicated by a thin line depicting low density.
+The plot (see figure \ref@(fig:ui-cv-rawfile-default)) illustrates that certain batches (=raw files) exhibit a higher number of distinct matches compared to others, as indicated by a thin line depicting low density.
 Most interquartile ranges (IQRs) overlap, suggesting similarity among the majority of batches.
@@
 \newpage
-```{r ui_cv_rawfile_nobs2, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins calculated for 2  observations per cell, with selection on Raw.file"}
+```{r ui-cv-rawfile-nobs2, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins calculated for 2  observations per cell, grouped by Raw.file"}
 knitr::include_graphics("screenshots/cv_rawfile_nobs2.png")
 ```
-By setting the number of observations to 2, the distributions become more equalized.
+Experimenting with a lower minimum number of observations for the calculation of the median coefficient of variation, includes also less ambigous peptides.
+
+By setting the number of observations to 2, the distributions become more equalized (see figure \ref@(fig:ui-cv-rawfile-nobs2)).
 Additionally, batches that initially exhibited a decreased coefficient of variation (CV) tend to inflate towards the threshold.
@@
 \newpage
-```{r ui_cv_rawfile_nobs10, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins calculated for 10 observations per cell, with selection on Raw.file"}
+```{r ui-cv-rawfile-nobs10, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins calculated for 10 observations per cell, with selection on Raw.file"}
 knitr::include_graphics("screenshots/cv_rawfile_nobs10.png")
 ```
+In opposite increasing the minimum number of observation takes only highly ambiguous peptides into account and the coefficent of variation shows the interaction of these razor proteins.
 
-When the minimum number of observations for ambiguous peptides is set to 10, the distribution of medians across channels shows a decrease in density.
-This can be attributed to the rare occurrence of 10 unique peptides matching to the same protein, which is not commonly observed in the dataset.
+When the minimum number of observations for ambiguous peptides is set to 10, the distribution of medians across channels shows a decrease in density (see figure \ref@(fig:ui-cv-rawfile-nobs10)).
+This can be attributed to the rare occurrence of 10 unique peptides matching to the same protein, which is rarely observed in the dataset.
 
 \newpage
-```{r ui_heatmap, echo=F, out.width="100%", fig.align='center', fig.cap="Correlation heatmap"}
+## Downstream analysis
+The subsequent analysis reveals the initial biological effects, serving as an indicator of the effectiveness or ineffectiveness of the pre-processing and quality control steps.
+
+```{r ui-heatmap, echo=F, out.width="100%", fig.align='center', fig.cap="Correlation heatmap"}
 knitr::include_graphics("screenshots/heatmap.png")
 ```
-The heatmap reveals the Pearson pairwise correlation coefficients between all proteins.
+The heatmap (see figure \ref@(fig:ui-heatmap)) reveals the Pearson pairwise correlation coefficients between all proteins.
 The dendrogram indicates that there are two main protein clusters, with one of the clusters showing further sub-clustering, indicating a strong relationship among the proteins in the sub groups.
@@
 \newpage
-```{r ui_dim_red_default, echo=F, out.width="100%", fig.align='center', fig.cap="Dimensionality reduction analysis depicting sample types (indicated by shape) and raw files (indicated by color).
+```{r ui-dim-red-default, echo=F, out.width="100%", fig.align='center', fig.cap="Dimensionality reduction analysis depicting sample types (indicated by shape) and raw files (indicated by color).
 Top: principle component analysis (PCA).
@@
 
-In both the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP) plot, the second component is plotted against the first component.
+In both the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP) plot (see figure \ref@(fig:ui-dim-red-default)), the second component is plotted against the first component.
 Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch.
@@
 \newpage
-```{r ui_dim_red_changed_k, echo=F, out.width="100%", fig.align='center', fig.cap="Dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+```{r ui-dim-red-changed-k, echo=F, out.width="100%", fig.align='center', fig.cap="Dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
 Top: Principal Component Analysis (PCA).
@@
 
-The second figure of the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP), both displaying the second component against the first.
+The second figure (\ref@(fig:ui-dim-red-changed-k) of the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP), both displaying the second component against the first.
 Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch information.
@@
 \newpage
-```{r ui_dim_red_c3d, echo=F, out.width="100%", fig.align='center', fig.cap="3D visualization of dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
+```{r ui-dim-red-c3d, echo=F, out.width="100%", fig.align='center', fig.cap="3D visualization of dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
 Top: Principal Component Analysis (PCA).
@@
 
-In the 3D visualization, a subtle clustering pattern can be observed for the first two batches.
+In the 3D visualization (see figure \ref@(fig:ui-dim-red-c3d)), a subtle clustering pattern can be observed for the first two batches.
 Upon switching to the indicator sample type in the 3D visualization, no clustering can be observed whatsoever.
@@
 \newpage
-```{r ui_dim_red_batchC, echo=F, out.width="100%", fig.align='center', fig.cap="Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
-upper top: Principal Component Analysis (PCA).
-upper bottom: Uniform Manifold Approximation and Projection (UMAP).
-lower top: 3D Principal Component Analysis (PCA).
-lower bottom: 3D Uniform Manifold Approximation and Projection (UMAP)"}
+```{r ui-dim-red-batchC, echo=F, out.width="100%", fig.align='center', fig.cap="Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color)."}
 knitr::include_graphics("screenshots/dim_red_batchC.png")
+```
+
+```{r ui-dim-red-batchC-3d, echo=F, out.width="100%", fig.align='center', fig.cap="Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color) in 3D."}
 knitr::include_graphics("screenshots/dim_red_batchC_3d.png")
@@
 
-After applying batch correction using ComBat \citep{Leek2012}, neither the 2D nor the 3D visualization demonstrates any noticeable effects related to batches or sample types.
+After applying batch correction using ComBat \citep{Leek2012}, neither the 2D (figure \ref@(fig:ui-dim-red-batchC)) nor the 3D (figure \ref@(fig:ui-dim-red-batchC-3d)) visualization demonstrates any noticeable effects related to batches or sample types.
 The previously observed subtle clustering of the first two batches is now eliminated, indicating successful batch correction.
@@
 \newpage
-```{r ui_feature, echo=F, out.width="100%", fig.align='center', fig.cap="Feature wise output for the protein P61313 (=60S ribosomal protein L15) over the course of the pipeline."}
+```{r ui-feature, echo=F, out.width="100%", fig.align='center', fig.cap="Feature wise output for the protein P61313 (=60S ribosomal protein L15) over the course of the pipeline."}
 knitr::include_graphics("screenshots/feature.png")
 ```
+The feature-wise output allows for the observation of abundance in raw files or batches, providing a useful tool to assess the effectiveness of pre-processing and/or the work conducted in the wet lab.
 
-The scatter plots indicate the detected values associated with the protein UniProt ID P61313 (= 60S ribosomal protein L15).
+The scatter plots (see figure \ref@(fig:ui-feature)) indicate the detected values associated with the protein UniProt ID P61313 (= 60S ribosomal protein L15).
 Starting from the peptide spectrum level and progressing to peptides aggregated for individual raw files, these plots offer a comprehensive view of the protein's peptide abundance across various batches.
@@
 \newpage
-```{r ui_stat_dependency, echo=F, out.width="100%", fig.align='center', fig.cap="Popup window for checking the dependency of the statistic module"}
+## Statistics
+
+```{r ui-stat-dependency, echo=F, out.width="100%", fig.align='center', fig.cap="Popup window for checking the dependency of the statistic module"}
 knitr::include_graphics("screenshots/stat_dependency.png")
@@
 
-The values for all sample types exhibit a unimodal distribution.
+The values for all sample types exhibit a unimodal distribution (see figure \ref@(fig:ui-stat-dependency)).
 However, when examining the quantile-quantile (qq) plot, heavy tails can be observed for all three groups.
@@
 \newpage
-```{r ui_stat_HC_HIV_noMets, echo=F, out.width="70%", fig.align='center', fig.cap="Statistic module"}
+```{r ui-stat-HC-HIV-noMets, echo=F, out.width="70%", fig.align='center', fig.cap="Statistics module"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 
-By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group, six differentially expressed proteins were identified.
-The Venndiagram indicates significantly up/downregulated proteins.
+By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group (see figure \ref@(fig:ui-stat-HC-HIV-noMets)), six differentially expressed proteins were identified.
+The Venn diagram indicates significantly up/down regulated proteins.
 The Volcano represents -log10(adjusted p values) against log2 fold changes (logFC).
 These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
-A total of 652 proteins have been identified as not significantly upregulated or downregulated in the comparison analysis.
+The remaining 652 proteins have been identified as not significantly up/down regulated in the comparison analysis.
 These proteins did not exhibit statistically significant differences in expression between the compared groups or conditions.
@@
 \newpage
-```{r ui_path_enrich_nw, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment. Networkplot indicating the association of differentially expressed proteins with pathways"}
+```{r ui-path-enrich-nw, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment. Networkplot indicating the association of differentially expressed proteins with pathways"}
 knitr::include_graphics("screenshots/path_enrich_nw.png")
 ```
+The enrichment represents overabundant proteins for specific pathways. Results can be shown as a network or barplot.
 
-The pathway enrichment mapped and tested with the exact Fisher test \citep{Sprent2011} the differentially expressed proteins P46777 (60S ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the ribosome and coronavirus disease (COVID-19) nodes.
+The pathway enrichment mapped and tested with the exact Fisher test \citep{Sprent2011} the differentially expressed proteins P46777 (60S ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the ribosome and coronavirus disease (COVID-19) nodes (see figure \ref@(fig:ui-path-enrich-nw)).
 
 \newpage
-```{r ui_path_enrich_bar, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment. Barplot indicating the interaction significance of differentially expressed proteins"}
+```{r ui-path-enrich-bar, echo=F, out.width="90%", fig.align='center', fig.cap="Pathway enrichment.
+Barplot indicating the number of differentially expressed proteins to a particular pathwaz.
+The color highlights the significance of the mapping, with exact p-values from the Fisher test after false discovery rate correction."}
 knitr::include_graphics("screenshots/path_enrich_bar.png")
@@
 
-The barplot shows the significance of the two nodes.
+The barplot (see figure \ref@(fig:ui-path-enrich-bar)) shows the significance of the two nodes.
 The count indicates how many proteins are associated with the node of the network.
 When hovering over the bar the user can read the p-value of the performed Fisher test.
-The background for the statistical test can be chosen as genome wide (default) or as the detected proteins.
+The background for the statistical test can be chosen as genome wide (default) or as the set of detected proteins.
 When choosing the detected proteins as a background no pathway could be considered as statistically significant.
@@
 \newpage
-```{r ui_go_enrich_bar, echo=F, out.width="90%", fig.align='center', fig.cap="Gene ontology enrichment against the complete human proteome.
+```{r ui-go-enrich-bar, echo=F, out.width="90%", fig.align='center', fig.cap="Gene ontology enrichment against the complete human proteome.
 Snapshot dotplot indicating the interaction significance of found proteins"}
@@
 ```
+Ontology enrichment offers a more comprehensive mapping, providing a broader understanding of the data.
+By allowing users to select a custom protein set collection, they can either narrow down the potential results for a targeted approach or expand their scope for an untargeted analysis.
 
-The position of the dot shows the number of matches for the particular protein set.
+The position of the dot (see figure \ref@(fig:ui-go-enrich-bar)) shows the number of matches for the particular protein set.
 Colors indicate for the significance level.

### 2023-07-18 markdown/Thesis.tex
until batch correction done

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \usepackage{xcolor}
-\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
-2cm]{geometry}
+\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 {\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
-presenting ProteoScanR & Proteomics Workbench
+\\ presenting ProteoScanR \& Proteomics Workbench
 
@@
 
-Main Supervisor: 
-Dr. DI(FH) Gerhard Duernberger
-
-External Supervison:
-Assoc. Prof. Ujjwal Neogi, M.Sc. PhD
-
-Second examiner:
-FH-Prof. Dr. Alexandra Graf
+\doublespacing
+\doublespacing
+Main Supervisor: Dr. DI(FH) Gerhard Duernberger \\
+External Supervisor: Assoc. Prof. Ujjwal Neogi, M.Sc. PhD \\
+Second examiner: FH-Prof. Dr. Alexandra Graf \\
 
@@
 
-With the growing prominence of single-cell techniques across various
-omics fields, there is a pressing need to develop a standardized
-pipeline for proteomics data in the realm of systems biology. Unlike DNA
-sequencing and RNA sequencing, proteomics analysis via mass spectrometry
-incurs high costs in terms of labor and equipment. Additionally,
-commercially available software solutions often come with hefty price
-tags and limited transparency regarding the underlying methods employed.
-However, MaxQuant \citep{Cox2008} presents a promising alternative,
-complemented by the flexibility using the R programming language.
-Introducing ProteoScanR, a state-of-the-art proteomics pipeline
-integrated into the user-friendly Proteomics Workbench interface. Guided
-by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the
-development process of ProteoScanR was thoroughly tested and validated
-using both bulk and single-cell mass spectrometry data sets. The
-pipeline implementation in the Proteomics Workbench leverages the power
-of an interactive environment built in R Shiny, empowering users to
-discover valuable insights for their specific data set. Within the
-interactive environment, users have the flexibility to customize cutoffs
-and thresholds for quality control, as well as employ various approaches
-for data transformation, normalization, missing value imputation, and
-batch correction. ProteoScanR and the Proteomics Workbench served as a
-valuable tool in guiding the identification of significantly expressed
-proteins in cells under study. Subsequently, the pathway enrichment
-analysis provided additional biological contexts for a comprehensive
-understanding of their functional implications. The master thesis
-project serves as a foundation for future advancements in the field of
-single-cell proteomics. Moreover, the codebase has been designed with
-robustness and scalability in mind, ensuring ease of maintenance and
-future expansion of the application.
+With the growing prominence of single-cell techniques across various omics fields, there is a pressing need to develop a standardized pipeline for proteomics data in the realm of systems biology.
+Unlike DNA sequencing and RNA sequencing, proteomics analysis via mass spectrometry incurs high costs in terms of labor and equipment.
+Additionally, commercially available software solutions often come with hefty price tags and limited transparency regarding the underlying methods employed.
+However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility using the R programming language.
+Introducing ProteoScanR, a state-of-the-art proteomics pipeline integrated into the user-friendly Proteomics Workbench interface.
+Guided by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of ProteoScanR was thoroughly tested and validated using both bulk and single-cell mass spectrometry data sets.
+The pipeline implementation in the Proteomics Workbench leverages the power of an interactive environment built in R Shiny, empowering users to discover valuable insights for their specific data set.
+Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction.
+ProteoScanR and the Proteomics Workbench served as a valuable tool in guiding the identification of significantly expressed proteins in cells under study.
+Subsequently, the pathway enrichment analysis provided additional biological contexts for a comprehensive understanding of their functional implications.
+The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics.
+Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.
 \url{https://github.com/Lukas67/ProteoScanR}
@@
 
-In addition to my faithful feline companion, Apollo, who took care of my
-place in Austria while I was working on this project in Stockholm, I am
-deeply grateful to all the members of Ujjwal Neogi's research group.
-
-Their unwavering guidance and support throughout this project have been
-truly invaluable, and I am profoundly thankful for their expertise and
-assistance. \pagenumbering{arabic}
+In addition to my faithful feline companion, Apollo, who took care of my place in Austria while I was working on this project in Stockholm, I am deeply grateful to all the members of Ujjwal Neogi's research group.
+
+Their unwavering guidance and support throughout this project have been truly invaluable, and I am profoundly thankful for their expertise and assistance.
+\pagenumbering{arabic}
 
@@
 \hypertarget{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}{%
-\subsection{From the Central Dogma of Molecular Biology to Omics and
-Systems
-Biology}\label{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}}
-
-Before the initial exploration of molecular biology, the processes of
-life were roughly inferred and not well understood. The central dogma of
-molecular biology serves as the cornerstone of biological processes,
-providing a framework for understanding how genetic information is
-converted into functional proteins. This framework consists of two major
-processes: transcription, which converts DNA into RNA, and translation,
-which converts RNA into functional proteins. Between these processes,
-molecular-level regulation occurs through chemical adaptations. This
-regulation is crucial for biological organisms to efficiently produce
-their genetic products, known as proteins. Crick's formulation of the
-central dogma has guided research in molecular biology for over 60 years
-\citep{Cobb2017}. Over the past six decades, researchers have explored
-various aspects of this principle, collectively referred to as
-``omics,'' to comprehensively analyze biomolecules in diverse contexts.
-Omics research has emerged as an approach to study biomolecules and
-their interactions within biological systems comprehensively. Omics
-encompasses various fields, including genomics, transcriptomics,
-proteomics, and metabolomics, with each field focusing on a specific
-class of biomolecules. By analyzing multiple layers of omics data,
-researchers can gain a holistic view of biological processes, uncover
-regulatory mechanisms, and identify key players in complex biological
-networks using computational methods. Systems biology takes a
-quantitative approach to investigate the interactions, dynamics, and
-emergent properties of biomolecular networks. Biological systems are
-imagined as complex, multi-layer networks, with the ensemble of
-proteins, known as the proteome, playing a critical role in this
-interactive structure. Proteins have various functions, such as
-providing structural integrity, catalyzing chemical reactions, and
-regulating cellular functions \citep{Karahalil2016}
+\subsection{From the Central Dogma of Molecular Biology to Omics and Systems Biology}\label{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}}
+
+Before the initial exploration of molecular biology, the processes of life were roughly inferred and not well understood.
+The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins.
+This framework consists of two major processes: transcription, which converts DNA into RNA, and translation, which converts RNA into functional proteins.
+Between these processes, molecular-level regulation occurs through chemical adaptations.
+This regulation is crucial for biological organisms to efficiently produce their genetic products, known as proteins.
+Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
+Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as ``omics,'' to comprehensively analyze biomolecules in diverse contexts.
+Omics research has emerged as an approach to study biomolecules and their interactions within biological systems comprehensively.
+Omics encompasses various fields, including genomics, transcriptomics, proteomics, and metabolomics, with each field focusing on a specific class of biomolecules.
+By analyzing multiple layers of omics data, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks using computational methods.
+Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks.
+Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role in this interactive structure.
+Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}
 
@@
 
-The central dogma of molecular biology offers insights not only at the
-level of individual terms but also provides a hierarchical understanding
-of biology. Hierarchical organization is the underlying principle in all
-topics of biology. Proteins exhibit hierarchical structures that
-contribute to their functionality. At the primary structure level,
-proteins are composed of smaller building blocks called amino acids.
-Combining them results in the protein secondary structure such as alpha
-helix and beta sheet. These secondary structures further assemble to
-create higher-order tertiary structures, which represent specific
-domains within the protein. Finally, the quaternary structure describes
-the functional state of the protein as a whole at a given time.
-Proteomics is a comprehensive approach that investigates the complete
-protein composition of a specimen, aiming to understand the intricate
-biological network it represents. For instance if pathogens interact
-with cells of an organism a cascade of biochemical reactions takes
-place. These reactions influence the outcome of the cellular proteome in
-regards of localization, abundance, post-translational modifications
-\citep{Beltran2017}. By studying the proteome, we can gain a
-comprehensive understanding of the hierarchical organization and
-functional dynamics of biological processes.
+The central dogma of molecular biology offers insights not only at the level of individual terms but also provides a hierarchical understanding of biology.
+Hierarchical organization is the underlying principle in all topics of biology.
+Proteins exhibit hierarchical structures that contribute to their functionality.
+At the primary structure level, proteins are composed of smaller building blocks called amino acids.
+Combining them results in the protein secondary structure such as alpha helix and beta sheet.
+These secondary structures further assemble to create higher-order tertiary structures, which represent specific domains within the protein.
+Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
+Proteomics is a comprehensive approach that investigates the complete protein composition of a specimen, aiming to understand the intricate biological network it represents.
+For instance if pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
+These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications \citep{Beltran2017}.
+By studying the proteome, we can gain a comprehensive understanding of the hierarchical organization and functional dynamics of biological processes.
 
@@
 
-Higher organisms are composed of specialized cells organized into
-tissues, such as skin, muscle, and blood. Each tissue consists of cells
-with specific functions, resulting in variations in protein expression.
-Bulk proteomics is a technique used to analyze the protein composition
-of a tissue sample, which contains all types of cells present in that
-particular tissue. This approach finds valuable applications for
-instance in oncology, where it can provide insights into the protein
-profile of a tumor. By studying oncogenic biomarkers and expression
-patterns, this knowledge can contribute to the development of screening
-methods and the customization of treatments. Additionally, bulk
-proteomics can aid in identifying the progression of tumor development
-\citep{Kwon2021}. Bulk analysis expression profiles provide scientists
-with an average protein abundance across all cells in a sample, offering
-a comprehensive view of protein expression. Additionally, compared to
-single-cell methods, bulk samples typically contain a greater number of
-proteins. In fact the expanded library can be used to validate the
-performance of single-cell type and single-cell proteomics methods and
-bioinformatics pipeline \citep{Schoof2021}. However, in this type of
-analysis, the precise properties of individual cell types can only be
-inferred and remain obscured.
+Higher organisms are composed of specialized cells organized into tissues, such as skin, muscle, and blood.
+Each tissue consists of cells with specific functions, resulting in variations in protein expression.
+Bulk proteomics is a technique used to analyze the protein composition of a tissue sample, which contains all types of cells present in that particular tissue.
+This approach finds valuable applications for instance in oncology, where it can provide insights into the protein profile of a tumor.
+By studying oncogenic biomarkers and expression patterns, this knowledge can contribute to the development of screening methods and the customization of treatments.
+Additionally, bulk proteomics can aid in identifying the progression of tumor development \citep{Kwon2021}.
+Bulk analysis expression profiles provide scientists with an average protein abundance across all cells in a sample, offering a comprehensive view of protein expression.
+Additionally, compared to single-cell methods, bulk samples typically contain a greater number of proteins.
+In fact the expanded library can be used to validate the performance of single-cell type and single-cell proteomics methods and bioinformatics pipeline \citep{Schoof2021}.
+However, in this type of analysis, the precise properties of individual cell types can only be inferred and remain obscured.
 
 \hypertarget{single-cell-type-proteomics}{%
-\subsubsection{Single-cell type
-proteomics}\label{single-cell-type-proteomics}}
-
-Taking tissue samples can lead to an averaging effect across the entire
-cellular ensemble, making it difficult to discern specific cell types.
-To overcome this limitation, cell sorting techniques such as
-fluorescence-activated cell sorting (FACS), magnetic-activated cell
-sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed
-\citep{Liou2015}. These techniques primarily rely on the detection of
-surface proteins on cells and the specific binding of antibodies to
-these proteins. Auto-fluorescence is observed in certain cells, wherein
-they possess the capability to emit a light signal when stimulated, and
-this emission can be used to sort them accordingly. However
-auto-fluorescence can lack of accuracy as shown in studies indicating
-the effect of laboratory gloves on cellular samples
-\citep{Philpott2017}. On the other side antibodies, typically derived
-from the immune system of other species, bind to antigens in a highly
-specific lock-and-key manner. Antibodies are widely used in molecular
-biology and find applications in cell isolation. Depending on the
-desired application, the antibodies can be labeled with fluorescent
-markers (FACS), magnetic beads (MACS), or biotin (BACS). These methods
-have varying technical requirements, with flow cytometers being the most
-complex and costly. In MACS, magnetic beads are used to isolate cells
-from a solution by utilizing a magnet. However, mechanical forces
-involved in MACS can potentially damage the cells. BACS, a method
-developed in 2015, employs microbubbles to carry the antibodies for
-cellular isolation. The method showed improvements compared to MACS in
-regards of viability after sorting. It is important to note that FACS is
-considered the gold standard for validating the effectiveness of MACS
-and BACS due to its high purity and yield. FACS was used in the
-validation of these two methods \citep{Liou2015, Sutermaster2019}.
-Single-cell type proteomics reflects the protein ensemble of a specific
-cell type at a particular time, providing a focused perspective on the
-studied field compared to bulk methods \citep{Maes2020}.
+\subsubsection{Single-cell type proteomics}\label{single-cell-type-proteomics}}
+
+Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
+To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed \citep{Liou2015}.
+These techniques primarily rely on the detection of surface proteins on cells and the specific binding of antibodies to these proteins.
+Auto-fluorescence is observed in certain cells, wherein they possess the capability to emit a light signal when stimulated, and this emission can be used to sort them accordingly.
+However auto-fluorescence can lack of accuracy as shown in studies indicating the effect of laboratory gloves on cellular samples \citep{Philpott2017}.
+On the other side antibodies, typically derived from the immune system of other species, bind to antigens in a highly specific lock-and-key manner.
+Antibodies are widely used in molecular biology and find applications in cell isolation.
+Depending on the desired application, the antibodies can be labeled with fluorescent markers (FACS), magnetic beads (MACS), or biotin (BACS).
+These methods have varying technical requirements, with flow cytometers being the most complex and costly.
+In MACS, magnetic beads are used to isolate cells from a solution by utilizing a magnet.
+However, mechanical forces involved in MACS can potentially damage the cells.
+BACS, a method developed in 2015, employs microbubbles to carry the antibodies for cellular isolation.
+The method showed improvements compared to MACS in regards of viability after sorting.
+It is important to note that FACS is considered the gold standard for validating the effectiveness of MACS and BACS due to its high purity and yield.
+FACS was used in the validation of these two methods \citep{Liou2015, Sutermaster2019}.
+Single-cell type proteomics reflects the protein ensemble of a specific cell type at a particular time, providing a focused perspective on the studied field compared to bulk methods \citep{Maes2020}.
 
@@
 
-Cell sorting techniques primarily rely on extracellular proteins to
-identify and separate different cell types. However, the intracellular
-content of cells can still exhibit variability in experiments.
-Single-cell proteomics (SCP) addresses this limitation by providing a
-snapshot of the underlying processes within an individual cell,
-eliminating this variability and allowing for the elucidation of the
-principles of cell type behavior, such as differentiation.
-Differentiation, a critical process in various biological systems such
-as the immune system, involves the acquisition of specialized functions
-by specific cell types \citep{Fang2018}. Investigating differentiation
-in multiple cells can be challenging due to the inherent heterogeneity
-during the initiation of this process. However, by analyzing one cell at
-a time and observing it throughout the differentiation process,
-researchers can capture the complete trajectory of differentiation at a
-single-cell level. This approach provides valuable insights into the
-dynamic changes occurring within cells during differentiation. In 2021,
-Specht et al.~developed the SCoPE2 sample preparation and analysis
-pipeline, performing macrophage-monocyte differentiation experiments
-\citep{Specht2021}. This pipeline streamlined the sample preparation
-process, reducing labor time and providing a reference model for
-subsequent SCP projects. SCP has proven to be a powerful tool for
-studying a wide range of biological phenomena, including drug responses,
-infectious diseases, and organism development. Over the past decade,
-single-cell techniques, including SCP, have become indispensable tools
-for the global research community \citep{Minakshi2019}.
+Cell sorting techniques primarily rely on extracellular proteins to identify and separate different cell types.
+However, the intracellular content of cells can still exhibit variability in experiments.
+Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the principles of cell type behavior, such as differentiation.
+Differentiation, a critical process in various biological systems such as the immune system, involves the acquisition of specialized functions by specific cell types \citep{Fang2018}.
+Investigating differentiation in multiple cells can be challenging due to the inherent heterogeneity during the initiation of this process.
+However, by analyzing one cell at a time and observing it throughout the differentiation process, researchers can capture the complete trajectory of differentiation at a single-cell level.
+This approach provides valuable insights into the dynamic changes occurring within cells during differentiation.
+In 2021, Specht et al.~developed the SCoPE2 sample preparation and analysis pipeline, performing macrophage-monocyte differentiation experiments \citep{Specht2021}.
+This pipeline streamlined the sample preparation process, reducing labor time and providing a reference model for subsequent SCP projects.
+SCP has proven to be a powerful tool for studying a wide range of biological phenomena, including drug responses, infectious diseases, and organism development.
+Over the past decade, single-cell techniques, including SCP, have become indispensable tools for the global research community \citep{Minakshi2019}.
 
@@
 
-An early approach of qualitative analysis of the cellular proteome
-involved labeling with fluorescent antibodies and imaging. The major
-disadvantage of this technique was the limitation to only stain a few
-proteins per cell and the requirement of a-priori knowledge of the
-protein to analyze. For quantification, procedures such as Western
-blots, immunoassays or cytometry by time of flight (CyTOF) have been
-used. Challenges are the ability to permeate cells, accessibility and
-binding of the epitope and the creation of specific antibodies for a
-given protein \citep{Budnik2018}. Early quantitative methods in
-molecular biology, measuring the entire protein concentration, were
-assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay.
-These methods are still in use to obtain total concentration for sample
-preparation. Another quantitative technique involved RNA-sequencing.
-RNAseq refers to the task of examining the transcriptome of bulk tissue
-sample with deep-sequencing technologies. RNA in a sample includes
-mRNAs, non-coding RNAs, and small RNAs. One of the main disadvantages
-is, that the mapping relies on the knowledge of the DNA sequence of the
-observed organism (reference genome) \citep{Wang2009}. Another
-disadvantage is, that the amount of protein translated from the
-determined mRNA can only be estimated \citep{Gygi1999}. However, today
-these predictions can be done with computational methods such as LASSO
-and the accuracy of the estimate highly depends on selection and
-validation of this method \citep{Magnusson2022}. When observing the
-protein ensemble of a cellular environment with a top-down approach like
-DNAseq or RNAseq, the characteristics (e.g.~post-translational
-modifications or charge) and local concentration of the proteins remain
-hidden. The cause of this phenomenon is that only a fraction of the DNA
-is transcribed into mRNA and not all mRNA is further translated into
-protein. Furthermore multiple proteins can be derived from a single
-DNA-sequence with alternative splicing and different proteins coming
-from non related sequences acting like building blocks can form protein
-complexes and/or Protein-RNA complexes. Cells are highly efficient and
-do not waste energy by producing obsolete proteins, moreover the
-degradation of proteins in the proteasome is an important energy
-recovery mechanism \citep{TANAKA2009}. A bottom-up strategy can
-elucidate these properties of the protein ensemble at a given time
-leading to proteomics. When studying impact of an external signal or
-stressor to a cell the abundance and properties of a protein or
-protein-ensemble can enlighten the underlying pathway leading to a
-biological response. Since some proteins do not even function before
-post-translational modifications (PTMs) happen, a knowledge about these
-can be crucial for further studies. Mass spectrometry (MS) fusions the
-advantages of the above mentioned techniques, because it combines the
-specificity of qualitative assays which are even able to sense (PTMs)
-with the quantitative accuracy of protein concentration assays.
-Furthermore MS does not require any a-priori knowledge of the protein
-ensemble of the specimen, making it the method of choice for untargeted
-approaches.
+An early approach of qualitative analysis of the cellular proteome involved labeling with fluorescent antibodies and imaging.
+The major disadvantage of this technique was the limitation to only stain a few proteins per cell and the requirement of a-priori knowledge of the protein to analyze.
+For quantification, procedures such as Western blots, immunoassays or cytometry by time of flight (CyTOF) have been used.
+Challenges are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
+Early quantitative methods in molecular biology, measuring the entire protein concentration, were assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay.
+These methods are still in use to obtain total concentration for sample preparation.
+Another quantitative technique involved RNA-sequencing.
+RNAseq refers to the task of examining the transcriptome of bulk tissue sample with deep-sequencing technologies.
+RNA in a sample includes mRNAs, non-coding RNAs, and small RNAs.
+One of the main disadvantages is, that the mapping relies on the knowledge of the DNA sequence of the observed organism (reference genome) \citep{Wang2009}.
+Another disadvantage is, that the amount of protein translated from the determined mRNA can only be estimated \citep{Gygi1999}.
+However, today these predictions can be done with computational methods such as LASSO and the accuracy of the estimate highly depends on selection and validation of this method \citep{Magnusson2022}.
+When observing the protein ensemble of a cellular environment with a top-down approach like DNAseq or RNAseq, the characteristics (e.g.~post-translational modifications or charge) and local concentration of the proteins remain hidden.
+The cause of this phenomenon is that only a fraction of the DNA is transcribed into mRNA and not all mRNA is further translated into protein.
+Furthermore multiple proteins can be derived from a single DNA-sequence with alternative splicing and different proteins coming from non related sequences acting like building blocks can form protein complexes and/or Protein-RNA complexes.
+Cells are highly efficient and do not waste energy by producing obsolete proteins, moreover the degradation of proteins in the proteasome is an important energy recovery mechanism \citep{TANAKA2009}.
+A bottom-up strategy can elucidate these properties of the protein ensemble at a given time leading to proteomics.
+When studying impact of an external signal or stressor to a cell the abundance and properties of a protein or protein-ensemble can enlighten the underlying pathway leading to a biological response.
+Since some proteins do not even function before post-translational modifications (PTMs) happen, a knowledge about these can be crucial for further studies.
+Mass spectrometry (MS) fusions the advantages of the above mentioned techniques, because it combines the specificity of qualitative assays which are even able to sense (PTMs) with the quantitative accuracy of protein concentration assays.
+Furthermore MS does not require any a-priori knowledge of the protein ensemble of the specimen, making it the method of choice for untargeted approaches.
 
@@
 
-Mass spectrometry enables qualitative and quantitative analysis of the
-entire repertoire of a biological sample. Mass spectrometers measure the
-mass to charge ratio (m/z) of a particle from a fragmented larger
-molecule. This is achieved by a physical procedure done with a device
-which is made of three major components: the ion source, an analyzer,
-and a detector. The ion sources charges molecules and accelerate them
-through a magnetic field. The analyzer separates particles according to
-their mass to charge ratio and the detector senses charged particles and
-amplifies their signal \citep{Parker2010}. However before acquiring any
-data, sample preparation and pre-processing is required. After sample
-preparation the first step involves tryptic digestion. It is a widely
-used technique in shotgun proteomics and involves the enzymatic cleavage
-of proteins into smaller peptides using the protease enzyme trypsin. The
-process known as proteolysis takes place in multiple parts of multi
-cellular organisms and was firstly detected in the small intestine
-digesting proteins \citep{Wang2008}. The enzyme cleaves specifically at
-the caroboxyl side of arginine and lysin residues. Afterwards a charge
-remains on the residues making them detectable for the mass
-spectrometer. It is proven that proteins which are digested to the
-limit, meaning no further proteolysis can be induced, provide the best
-results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.
-
-To observe multiple samples in one run, labeling is needed to identify
-each sample within a batch. Since mass spectrometry is not a
-quantitative technique by itself, the peak height or area does not
-reflect the abundance of a peptide. Physicochemical properties of the
-peptides can change the ionization efficiency and detectability of the
-target. However, when comparing the same analyte between multiple runs
-of labeled peptides, differences in the reporter ion intensity reflect
-the abundance of those. Labels should be chosen to change solely the
-mass of the sample and to not affect folding or other inherent
-properties of the peptide. The two techniques for labeling peptides are
-metabolic labeling and isobaric labling. In metabolic labeling the cells
-were fed with aminoacids containing heavy isotopes, it is the method of
-choice in order to label peptides at the earliest possible level. This
-atoms can be heavy nitrogen in aminoacids or salts in fertilizer for
-plants \citep{Karen2012}. Mass shifts are proportional to the isotopes
-incorporated during biomass production and are visible after proteolytic
-cleavage. Stable isotope labeling in cell culture (SILAC) was presented
-in the early 2000s. This method used heavy aminoacid enriched media to
-feed cells, in order to quantitatively analyze expression profiles. The
-limitation of this method is, that it can only be applied to a human
-cell culture (in vitro) or to model organisms (in vivo). Isobaric
-labeling solely applies to in vitro techniques, with TMT and iTRAQ. The
-data analyzed by ProteoScanR primarily utilized TMT tags. Tandem mass
-tag (TMT) reagents enable differentiation of multiple samples in a
-single mass spectrometry (MS) run. Each sample is individually labeled
-with TMT reagents and then pooled together, this procedure is called
-multiplexing. TMTs have the same charge and differ only by their
-isotopic masses, the peaks observed for each sample are called reporter
-ions (RI). Each RI and sample is interpreted as one channel in
-downstream analysis. This technique allows for the identification,
-enrichment, and quantification of low-abundance peptide ions, which is
-particularly valuable in single-cell techniques. With this technique it
-is possible to quantify proteins and differ low abundant proteins from
-background noise. However, a drawback of isobaric labeling is the
-presence of co-fragmentation signals in the spectrogram, which requires
-data normalization to remove unwanted contributions
-\citep{Marx2019, Budnik2018}. Isotopic distribution in TMTs, reflecting
-the natural distribution, can be corrected during data acquisition by
-considering a defined spread in other channels. Single-cell proteomics
-by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem
-mass tag (TMT) to enhance signal intensity of a protein species. Loading
-of a single-cell channel (SCC) with the right amount can be balanced to
-the protein concentration of a single cell \citep{Ye2022}.
-
-Before ionization peptides need to be further separated according to
-their chemical properties, size or species by liquid chromatography
-(LC), because the number of peptide processed simultaneously by the MS
-device is limited. In the case of mass spectrometry, liquid
-chromatography especially high performance LC (HPLC) is the technique of
-choice and replaces the instrumentation-wise simpler technique gel
-electrophoresis. In HPLC the molecules eluate through a narrow column
-with high pressures (50-350 bar), compared to LC where the separation is
-gravity dependent. Similar to gel electrophoresis molecules can be
-separated upon different chemical properties, depending on the type of
-the chromatography column. In a typical proteomics experiment peptides
-are separated with reversed phase chromatography (RPLC). The principle
-behind RPLC is a column with a nonpolar stationary phase which interacts
-with the nonpolar residues of the peptides. Peptides were fractionated
-according to their hydrophobicity \citep{Pitt2009}.
-
-To analyze a biological sample consisting of peptides in solution the
-liquid needs to be vaporized into gas phase. Two techniques are capable
-of this procedure. Electrospray ionization (ESI) pushes the analyte
-through a capillary while applying an electric current to the liquid,
-vaporizing the sample to a charged aerosol. Peptides are further
-fragmented according to their chemical properties and can be further
-handled in the mass spectrometer. The fragmented peptides are now in
-charged droplets separated based on their surface charge, splitting
-further into smaller droplets until they reach the gas phase as ions.
-Two physical models describe the process from gas phase to ion called
-``The ion evaporation model'' (IEM) and ``The charge residue model''
-(CRM). In the ion evaporation model the droplets shrink by evaporation
-until ions are expelled \citep{Iribarne1976}. The model had its
-limitation by explaining same evaporation rate constant among ions with
-different chemical properties. In the charge residue model the
-assumption of one molecule per droplet leads to an ionization rate
-constant, which is independent of the ion itself and relies solely on
-the generation of the droplet and the efficiency of the solvent
-\citep{Wilm2011}.
-
-Subsequently after the ionization peptides are accelerated through the
-magnetic field generated by the mass analyzer. Six general types of mass
-analyzers are availaible on the market, there are Quadropole, time of
-flight (TOF), magnetic sector, electrostatic sector, quadropole ion trap
-and ion cyclotron resonance -mass analyzer. The function of the mass
-analyzer is to separate ions based on their mass to charge ratio (m/z).
-
-Before the signal of the separated ions can be obtained an optional
-subsequent step involves trapping the ions an electric and/or magnetic
-field in order to detect ions based on their (m/z) consecutively. Ion
-trapping enables the technique of coupled mass spectrometers, where ions
-were further fragmented by collision induced dissociation and analyzed
-subsequently. The detector transforms the impact energy of the ion into
-an electric signal, called spectrum which can be processed
-computationally.
+Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
+Mass spectrometers measure the mass to charge ratio (m/z) of a particle from a fragmented larger molecule.
+This is achieved by a physical procedure done with a device which is made of three major components: the ion source, an analyzer, and a detector.
+The ion sources charges molecules and accelerate them through a magnetic field.
+The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal \citep{Parker2010}.
+However before acquiring any data, sample preparation and pre-processing is required.
+After sample preparation the first step involves tryptic digestion.
+It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the protease enzyme trypsin.
+The process known as proteolysis takes place in multiple parts of multi cellular organisms and was firstly detected in the small intestine digesting proteins \citep{Wang2008}.
+The enzyme cleaves specifically at the caroboxyl side of arginine and lysin residues.
+Afterwards a charge remains on the residues making them detectable for the mass spectrometer.
+It is proven that proteins which are digested to the limit, meaning no further proteolysis can be induced, provide the best results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.
+
+To observe multiple samples in one run, labeling is needed to identify each sample within a batch.
+Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
+Physicochemical properties of the peptides can change the ionization efficiency and detectability of the target.
+However, when comparing the same analyte between multiple runs of labeled peptides, differences in the reporter ion intensity reflect the abundance of those.
+Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the peptide.
+The two techniques for labeling peptides are metabolic labeling and isobaric labling.
+In metabolic labeling the cells were fed with aminoacids containing heavy isotopes, it is the method of choice in order to label peptides at the earliest possible level.
+This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants \citep{Karen2012}.
+Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
+Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
+This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
+The limitation of this method is, that it can only be applied to a human cell culture (in vitro) or to model organisms (in vivo).
+Isobaric labeling solely applies to in vitro techniques, with TMT and iTRAQ.
+The data analyzed by ProteoScanR primarily utilized TMT tags.
+Tandem mass tag (TMT) reagents enable differentiation of multiple samples in a single mass spectrometry (MS) run.
+Each sample is individually labeled with TMT reagents and then pooled together, this procedure is called multiplexing.
+TMTs have the same charge and differ only by their isotopic masses, the peaks observed for each sample are called reporter ions (RI).
+Each RI and sample is interpreted as one channel in downstream analysis.
+This technique allows for the identification, enrichment, and quantification of low-abundance peptide ions, which is particularly valuable in single-cell techniques.
+With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
+However, a drawback of isobaric labeling is the presence of co-fragmentation signals in the spectrogram, which requires data normalization to remove unwanted contributions \citep{Marx2019, Budnik2018}.
+Isotopic distribution in TMTs, reflecting the natural distribution, can be corrected during data acquisition by considering a defined spread in other channels.
+Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
+Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell \citep{Ye2022}.
+
+Before ionization peptides need to be further separated according to their chemical properties, size or species by liquid chromatography (LC), because the number of peptide processed simultaneously by the MS device is limited.
+In the case of mass spectrometry, liquid chromatography especially high performance LC (HPLC) is the technique of choice and replaces the instrumentation-wise simpler technique gel electrophoresis.
+In HPLC the molecules eluate through a narrow column with high pressures (50-350 bar), compared to LC where the separation is gravity dependent.
+Similar to gel electrophoresis molecules can be separated upon different chemical properties, depending on the type of the chromatography column.
+In a typical proteomics experiment peptides are separated with reversed phase chromatography (RPLC).
+The principle behind RPLC is a column with a nonpolar stationary phase which interacts with the nonpolar residues of the peptides.
+Peptides were fractionated according to their hydrophobicity \citep{Pitt2009}.
+
+To analyze a biological sample consisting of peptides in solution the liquid needs to be vaporized into gas phase.
+Two techniques are capable of this procedure.
+Electrospray ionization (ESI) pushes the analyte through a capillary while applying an electric current to the liquid, vaporizing the sample to a charged aerosol.
+Peptides are further fragmented according to their chemical properties and can be further handled in the mass spectrometer.
+The fragmented peptides are now in charged droplets separated based on their surface charge, splitting further into smaller droplets until they reach the gas phase as ions.
+Two physical models describe the process from gas phase to ion called ``The ion evaporation model'' (IEM) and ``The charge residue model'' (CRM).
+In the ion evaporation model the droplets shrink by evaporation until ions are expelled \citep{Iribarne1976}.
+The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties.
+In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.
+
+Subsequently after the ionization peptides are accelerated through the magnetic field generated by the mass analyzer.
+Six general types of mass analyzers are availaible on the market, there are Quadropole, time of flight (TOF), magnetic sector, electrostatic sector, quadropole ion trap and ion cyclotron resonance -mass analyzer.
+The function of the mass analyzer is to separate ions based on their mass to charge ratio (m/z).
+
+Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their (m/z) consecutively.
+Ion trapping enables the technique of coupled mass spectrometers, where ions were further fragmented by collision induced dissociation and analyzed subsequently.
+The detector transforms the impact energy of the ion into an electric signal, called spectrum which can be processed computationally.
 
@@
 
-Given the high resolution of MS data, algorithms are employed to convert
-the raw signal into an interpretable form. Software packages like
-MaxQuant \citep{Cox2008} are commonly used to process the data,
-providing it for further analysis and statistical testing. Other
-software solutions include Proteome Discoverer by Thermo Fisher.
-
-The data obtained from mass spectrometry has three dimensions: m/z
-ratio, intensity, and retention time. To separate peaks from each other,
-algorithms are used to identify local minima in the data. The centroid
-of each peak is determined by fitting a Gaussian peak shape, which can
-be interpreted as locating the peaks of each m/z spectrum as a function
-of retention time. The centroid of a peak corresponds to an isotope.
-
-In order to decipher the isotopic distribution of a biomolecule,
-MaxQuant employs a vertex-based approach. This process, referred to as
-de-isotoping, utilizes graph theory. It creates a vertex for each
-individual peak and connects them with their potential isotopic
-counterparts. This is achieved by determining the mass proportion of an
-average amino acid to its respective isotope using a concept known as
-averagine \citep{Senko1995}. By applying this procedure, the number of
-data points is reduced by a factor of ten, and each peak represents a
-small biomolecule.
-
-The subsequent step in data acquisition involves the detection of labels
-for quantification purposes. Isotopic pairs of the label (such as N13,
-N14, N15) present in the tag or aminoacid are identified by convolving
-the two measured isotope patterns with the theoretical isotope patterns.
-Through an iterative process using a least-squares method, the best fit
-is determined, enabling the identification of the specific channel or
-sample.
-
-The intensity-weighted average of the MS peak centroids, as described as
-3D peak identification, corresponds to the mass of the peptide. The
-accuracy of mass measurements relies on the specific analyzer used. For
-an Orbitrap-type analyzer (which was used during this project), MaxQuant
-applies a correction value of 1 ppm (parts per million). Autocorrelation
-between centroids is accounted for by considering only well-identified
-peptides. Published data indicates that the mass precision within an MS
-experiment typically ranges around 10\^{}-7 \citep{Wilschefski2019}.
-
-Each peptide is represented by its individual fingerprint in the MS
-spectrum, which is based on the sequence and modifications of amino
-acids. The m/z ratio of the amino acids can be calculated, and the
-resulting peaks are interpreted as an amino acid sequence. However, only
-peptide fragments are visible in the spectrum due to peptide
-fragmentation. To identify proteins, peptides are matched against a
-sequence database. Sequence databases are typically stored as .fasta
-files and can be downloaded from the UniProt database (www.uniprot.org).
-The availability of gene sequences in databases and the computational
-matching of peptides against those sequences enable the identification
-of alterations in a sample at the protein level. These alterations can
-rely on the sequence level or could be to post-translational
-modifications (PTMs) such as phosphorylations or methylations
-\citep{Aebersold2003}. The peptide identification (P-) score reflects
-the goodness of fit between the data and the identified sequence in the
-database, taking into account the length of the peptide. This score is
-used to calculate the posterior error probability, which is then
-utilized to estimate the false discovery rate (FDR). The FDR calculation
-provides a measure of the rate at which false identifications are
-expected in the dataset. Within the identified peptide groups, a ``razor
-protein'' refers to the protein with the highest number of ambiguously
-identified peptides. Quantification is performed by considering only
-unique peptides for comparison. Posterior error probabilities, which
-represent the likelihood of the identified peptide being a random event,
-are multiplied together, and only distinct sequences with the highest
-scores are considered for further analysis. Various metrics that
-indicate the performance of the peptide search can be utilized in
-downstream analysis.
+Given the high resolution of MS data, algorithms are employed to convert the raw signal into an interpretable form.
+Software packages like MaxQuant \citep{Cox2008} are commonly used to process the data, providing it for further analysis and statistical testing.
+Other software solutions include Proteome Discoverer by Thermo Fisher.
+
+The data obtained from mass spectrometry has three dimensions: m/z ratio, intensity, and retention time.
+To separate peaks from each other, algorithms are used to identify local minima in the data.
+The centroid of each peak is determined by fitting a Gaussian peak shape, which can be interpreted as locating the peaks of each m/z spectrum as a function of retention time.
+The centroid of a peak corresponds to an isotope.
+
+In order to decipher the isotopic distribution of a biomolecule, MaxQuant employs a vertex-based approach.
+This process, referred to as de-isotoping, utilizes graph theory.
+It creates a vertex for each individual peak and connects them with their potential isotopic counterparts.
+This is achieved by determining the mass proportion of an average amino acid to its respective isotope using a concept known as averagine \citep{Senko1995}.
+By applying this procedure, the number of data points is reduced by a factor of ten, and each peak represents a small biomolecule.
+
+The subsequent step in data acquisition involves the detection of labels for quantification purposes.
+Isotopic pairs of the label (such as N13, N14, N15) present in the tag or aminoacid are identified by convolving the two measured isotope patterns with the theoretical isotope patterns.
+Through an iterative process using a least-squares method, the best fit is determined, enabling the identification of the specific channel or sample.
+
+The intensity-weighted average of the MS peak centroids, as described as 3D peak identification, corresponds to the mass of the peptide.
+The accuracy of mass measurements relies on the specific analyzer used.
+For an Orbitrap-type analyzer (which was used during this project), MaxQuant applies a correction value of 1 ppm (parts per million).
+Autocorrelation between centroids is accounted for by considering only well-identified peptides.
+Published data indicates that the mass precision within an MS experiment typically ranges around 10\^{}-7 \citep{Wilschefski2019}.
+
+Each peptide is represented by its individual fingerprint in the MS spectrum, which is based on the sequence and modifications of amino acids.
+The m/z ratio of the amino acids can be calculated, and the resulting peaks are interpreted as an amino acid sequence.
+However, only peptide fragments are visible in the spectrum due to peptide fragmentation.
+To identify proteins, peptides are matched against a sequence database.
+Sequence databases are typically stored as .fasta files and can be downloaded from the UniProt database (www.uniprot.org).
+The availability of gene sequences in databases and the computational matching of peptides against those sequences enable the identification of alterations in a sample at the protein level.
+These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylations or methylations \citep{Aebersold2003}.
+The peptide identification (P-) score reflects the goodness of fit between the data and the identified sequence in the database, taking into account the length of the peptide.
+This score is used to calculate the posterior error probability, which is then utilized to estimate the false discovery rate (FDR).
+The FDR calculation provides a measure of the rate at which false identifications are expected in the dataset.
+Within the identified peptide groups, a ``razor protein'' refers to the protein with the highest number of ambiguously identified peptides.
+Quantification is performed by considering only unique peptides for comparison.
+Posterior error probabilities, which represent the likelihood of the identified peptide being a random event, are multiplied together, and only distinct sequences with the highest scores are considered for further analysis.
+Various metrics that indicate the performance of the peptide search can be utilized in downstream analysis.
 
@@
 
-With advancements in mass spectrometry technology and computational
-methods, proteomic analysis has emerged as a powerful tool for
-investigating complex protein samples. However, analyzing proteomic data
-poses challenges in data processing, statistical analysis, and
-interpretation. This thesis aims to address these challenges by
-exploring computational methods for downstream analysis of proteomics
-data. The methods employed in this study will be thoroughly explained in
-the methods section, with a specific focus on developing an interactive
-environment equipped with a graphical user interface, called Proteomics
-Workbench. This user-friendly interface will enable researchers,
-including those with limited computational expertise, to navigate and
-comprehend the data effectively.
-
-ProteoScanR comprises a series of steps to streamline the data. It
-begins with quality control measures at the peptide spectrum match
-level, involving the removal of contaminants and irrelevant channels.
-Significance levels of peptide spectrum matches serve as confidence
-thresholds for match selection. The data is then aggregated at the
-peptide level, followed by additional filtering procedures based on
-metrics such as reporter ion intensity, peptide coefficient of variation
-over razor proteins, and missing rates for peptides.
-
-To derive meaningful biological insights from the dataset, statistical
-testing plays a crucial role. Selecting appropriate statistical tests
-and ensuring proper data transformation and normalization are essential
-steps. Many statistical methods rely on data distribution and require
-careful consideration to preserve the biological interpretation of the
-results. This thesis employs various transformation and normalization
-techniques, aiming to maintain data integrity and avoid biased
-interpretations. Additionally, an entropy-based visualization approach
-will provide users with a valuable tool for validating the computations
-and selected thresholds for the filtering steps.
-
-Handling missing values poses further challenges in the pipeline
-development since a missing signal for a peptide results in a missing
-protein, which is highly unlikely in a biological context. Several
-options for missing value handling will be included and thoroughly
-explained in the methods section.
-
-Dimensionality reduction and visualization techniques are widely used to
-cluster data. In this project, multiple approaches will be utilized, and
-users will have the ability to visually and interactively observe the
-data.
-
-It is important to acknowledge that downstream analysis in proteomics
-lacks a standardized approach, and the selection of computational
-methods depends on the specific dataset and research objectives.
-Furthermore, statistical analysis and the interpretation of biological
-meaning often present misconceptions that need to be addressed. For
-instance, batch correction, a commonly practiced method, has been
-overestimated in numerous studies, leading to misleading results
-\citep{Nygaard2016}. This thesis aims to contribute to the field by
-providing researchers with a comprehensive understanding of
-computational methods in proteomics analysis and their application in an
-interactive environment.
+With advancements in mass spectrometry technology and computational methods, proteomic analysis has emerged as a powerful tool for investigating complex protein samples.
+However, analyzing proteomic data poses challenges in data processing, statistical analysis, and interpretation.
+This thesis aims to address these challenges by exploring computational methods for downstream analysis of proteomics data.
+The methods employed in this study will be thoroughly explained in the methods section, with a specific focus on developing an interactive environment equipped with a graphical user interface, called Proteomics Workbench.
+This user-friendly interface will enable researchers, including those with limited computational expertise, to navigate and comprehend the data effectively.
+
+ProteoScanR comprises a series of steps to streamline the data.
+It begins with quality control measures at the peptide spectrum match level, involving the removal of contaminants and irrelevant channels.
+Significance levels of peptide spectrum matches serve as confidence thresholds for match selection.
+The data is then aggregated at the peptide level, followed by additional filtering procedures based on metrics such as reporter ion intensity, peptide coefficient of variation over razor proteins, and missing rates for peptides.
+
+To derive meaningful biological insights from the dataset, statistical testing plays a crucial role.
+Selecting appropriate statistical tests and ensuring proper data transformation and normalization are essential steps.
+Many statistical methods rely on data distribution and require careful consideration to preserve the biological interpretation of the results.
+This thesis employs various transformation and normalization techniques, aiming to maintain data integrity and avoid biased interpretations.
+Additionally, an entropy-based visualization approach will provide users with a valuable tool for validating the computations and selected thresholds for the filtering steps.
+
+Handling missing values poses further challenges in the pipeline development since a missing signal for a peptide results in a missing protein, which is highly unlikely in a biological context.
+Several options for missing value handling will be included and thoroughly explained in the methods section.
+
+Dimensionality reduction and visualization techniques are widely used to cluster data.
+In this project, multiple approaches will be utilized, and users will have the ability to visually and interactively observe the data.
+
+It is important to acknowledge that downstream analysis in proteomics lacks a standardized approach, and the selection of computational methods depends on the specific dataset and research objectives.
+Furthermore, statistical analysis and the interpretation of biological meaning often present misconceptions that need to be addressed.
+For instance, batch correction, a commonly practiced method, has been overestimated in numerous studies, leading to misleading results \citep{Nygaard2016}.
+This thesis aims to contribute to the field by providing researchers with a comprehensive understanding of computational methods in proteomics analysis and their application in an interactive environment.
 
@@
 \hypertarget{wetlab---materials-and-methods}{%
-\subsection{Wetlab - Materials and
-Methods}\label{wetlab---materials-and-methods}}
-
-For testing and development of the ProteoScanR pipeline the following
-materials and methods were performed. For the sample preparation
-protocol see appendix.
+\subsection{Wetlab - Materials and Methods}\label{wetlab---materials-and-methods}}
+
+For testing and development of the ProteoScanR pipeline the following materials and methods were applied.
 
@@
 
-Macrophages from 12 healthy donors and 12 HIV patients with metabolic
-syndrome and 12 HIV patients without metabolic syndrome. Cell sorting
-was performed with FACSAria Fusion (BD Biosciences).
+Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome were isolated.
+Cell sorting was performed with FACSAria Fusion (BD Biosciences) by auto fluorescence.
+
+\hypertarget{cells-sorting-protocol}{%
+\paragraph{Cells sorting protocol}\label{cells-sorting-protocol}}
+
+\begin{enumerate}
+\item 96 single cells from each donor were sorted into a single plate. Each well contained a single cell. As such, 12 different plates containing single cells from 12 different donors were prepared. 
+\item Carrier plates were also prepared. In the carrier plates, 100 cells from one donor were sorted into each well. 24 such wells (3 columns) were prepared for each donor.
+\end{enumerate}
 
@@
 
-Protein digestion was performed using trypsin (Promega, Madison, WA)
-buffered in Triethylammonium bicarbonate buffer (TEAB). Reagents were
-handled with MANTIS automatic dispenser (Formulatrix, Bedford, MA).
+Protein digestion was performed using trypsin (Promega, Madison, WA) buffered in Triethylammonium bicarbonate buffer (TEAB).
+Reagents were handled with MANTIS automatic dispenser (Formulatrix, Bedford, MA).
+
+\hypertarget{digestion-protocol}{%
+\paragraph{Digestion protocol}\label{digestion-protocol}}
+
+\begin{enumerate}
+\item Prepare 25 ng/ $\mu$L trypsin in 100 mmol/L Triethylammonium bicarbonate buffer (TEAB).
+\item Add 25 ng/$\mu$L trypsin: 2 $\mu$L to the carrier samples, 1 $\mu$L to the single cell samples.
+\item Seal the plates.
+\item Centrifuge (2 min at 120 G).
+\item Incubate overnight at 37 °C.
+\end{enumerate}
 
@@
 
-Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher)
-in anhydrous acetonitril (ACN) solution.
+Samples of the test data set were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.
+
+\hypertarget{labeling-protocol}{%
+\paragraph{Labeling protocol}\label{labeling-protocol}}
+
+\begin{enumerate}
+\item Let the TMTpro labels come to room temperature so that all the condensation on the tube is gone.
+Dissolve the TMTpro reagents in anhydrous acetonitril (ACN) to obtain a concentration of 10 $\mu$g/$\mu$L.
+\item Add 1 $\mu$L of TMTpro reagents according to the tables. For one set of plates, you will need 16 $\mu$L of each label. It is recommended to aliquot a slightly larger volume, of around 20 $\mu$L.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Prepare a 5% hydroxylamine solution in water.
+\item Add 1 $\mu$L of the 5% hydroxylamine solution to each well.
+\item Seal the plates.
+\item Centrifuge for 2 minutes at 120 G.
+\item Incubate for 1 hour at room temperature.
+\item Cover the plates with cap strips.
+\end{enumerate}
 
@@
 
-Carrier proteome was added to the 36 TMTpro sets before pooling sample
-types for each of the 36 MS-runs.
+Carrier proteome was added to the 36 TMTpro sets before pooling sample types for each of the 36 MS-runs.
+
+\hypertarget{pooling-protocol}{%
+\paragraph{Pooling protocol}\label{pooling-protocol}}
+
+\begin{enumerate}
+\item Wash a 10 $\mu$L glass microsyringe five times with ACN.
+\item Use the microsyringe to combine TMTpro-labeled samples into 36 LC autosampler vials with built-in glass inserts.
+\item Combine six carrier wells from each sample and add the carrier proteome to the 36 TMTpro sets.
+\item Combine the first set by taking well A1 from the first single-cell plate, well A2 from the second plate, well A3 from the third plate, and so on.
+For the next set, take well A1 from the second plate, well A2 from the third plate, and so on.
+Combine 12 sets following this scheme.
+Combine the next 24 sets by following the same scheme for rows B and C.
+\item Vacuum centrifuge the combined sets to dryness.
+\item Dissolve the dried samples in 7 $\mu$L of buffer A and transfer them to LC-MS/MS for further analysis.
+\end{enumerate}
 
@@
 
-In order to separate peptides according to their chemical properties,
-size or species a liquid chromatography (LC) is recommended before
-ionization. For the reference data used for developing the pipeline the
-following intrumentation was used. EASY-Spray™ HPLC column, ES802A, 2
-\(\mu\)m × 75 \(\mu\)m ID × 250 mm (ThermoFisher Scientific) Acclaim
-PepMap trap column, 2 \(\mu\)m × 75 \(\mu\)m ID × 2 cm (ThermoFisher
-Scientific)
+In order to separate peptides according to their chemical properties, size or species a liquid chromatography (LC) is recommended before ionization.
+For the reference data used for developing the pipeline the following intrumentation was used.
+EASY-Spray™ HPLC column, ES802A, 2 \(\mu\)m × 75 \(\mu\)m ID × 250 mm (ThermoFisher Scientific)
+Acclaim PepMap trap column, 2 \(\mu\)m × 75 \(\mu\)m ID × 2 cm (ThermoFisher Scientific)
 
@@
 
-For the reference data used for developing the pipeline an Orbitrap™
-Fusion™ Lumos™ tribrid mass spectrometer (ThermoFisher Scientific) was
-used. The instrumentation uses two subsequent runs in the analyzer
-(=Orbitrap) to increase mass accuracy.
+For the reference data used for developing the pipeline an Orbitrap™ Fusion™ Lumos™ tribrid mass spectrometer (ThermoFisher Scientific) was used.
+The instrumentation uses two subsequent runs in the analyzer (=Orbitrap) to increase mass accuracy.
 
 \hypertarget{drylab---computational-methods}{%
-\subsection{Drylab - Computational
-methods}\label{drylab---computational-methods}}
+\subsection{Drylab - Computational methods}\label{drylab---computational-methods}}
 
@@
 
-Lenovo Laptop X360 with: Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz 8 GB
-RAM
+Data was analysed using commercially available hardware, find details below:
+Lenovo Laptop X360 with:
+Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz
+8 GB RAM
 
 \hypertarget{data-acquisition-and-raw-file-analysis}{%
-\subsubsection{Data-acquisition and raw-file
-analysis}\label{data-acquisition-and-raw-file-analysis}}
+\subsubsection{Data-acquisition and raw-file analysis}\label{data-acquisition-and-raw-file-analysis}}
 
 For the data-acquisition Thermo Fisher acquisition software was used.
-Raw-file analysis of the data was done with MaxQuantCmd 2.2.0.0
-\citep{Cox2008}. The default settings were used. TMTpro labels were
-corrected according to the isotopical distribution of the Lot-number
-provided by Thermo Fisher.
-
-\bgroup  \origfigure[H]
-
-{\centering \includegraphics{Thesis_files/figure-latex/data_processing_pipeline_flowchart_vertical-1}
-
-}
-
-\caption{Flowchart presenting the processing by ProteoScanR}\label{fig:data_processing_pipeline_flowchart_vertical}
+Raw-file analysis of the data was done with MaxQuantCmd 2.2.0.0 \citep{Cox2008}.
+The default settings were used. TMTpro labels were corrected according to the isotopical distribution of the Lot-number provided by Thermo Fisher.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.8\linewidth]{Thesis_files/figure-latex/data-processing-pipeline-flowchart-vertical-1}
+
+}
+
+\caption{Flowchart presenting the processing by ProteoScanR}\label{fig:data-processing-pipeline-flowchart-vertical}
  \endfigure\egroup
@@
 
-Further analysis is done with R and respective bioconductor packages.
-Please find version info in the appendix.
-
-After processing MaxQuant creates a directory containing all results as
-.txt file. The evidence.txt file includes all peptide to spectrum
-matches (PSM) with their respective proteins and statistical parameters.
+Further analysis is done with R and respective bioconductor packages. Please find version info in the appendix.
+
+After processing MaxQuant creates a directory containing all results as .txt file.
+The evidence.txt file includes all peptide to spectrum matches (PSM) with their respective proteins and statistical parameters.
 
@@
 \hypertarget{object-oriented-programming}{%
-\subsubsection{Object oriented
-programming}\label{object-oriented-programming}}
-
-In order to streamline the analysis of multiple experiments, object
-oriented programming was applied. The approach in R is to create a so
-called Q-feature object, which contains all variables and metadata in a
-hierarchical structure. The structure enables sub setting for further
-analysis \citep{Vanderaa2021}.
+\subsubsection{Object oriented programming}\label{object-oriented-programming}}
+
+In order to streamline the analysis of multiple experiments and software development, object oriented programming was applied.
+The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
+The structure enables sub setting for further analysis \citep{Vanderaa2021}.
 
@@
 Peptides with low abundance are often set to zero during analysis.
-However, assigning a value of zero may incorrectly suggest that the
-sample does not contain the respective peptide. Given that it is highly
-unlikely for a biological cell of a comparable type and function to not
-contain a particular protein, replacing the zero value with ``not
-applicable'' (NA) is crucial for understanding and interpreting MS data.
-NA values will be managed in missing value handling later on the protein
-level.
+However, assigning a value of zero may incorrectly suggest that the sample does not contain the respective peptide.
+Given that it is highly unlikely for a biological cell of a comparable type and function to not contain a particular protein, replacing the zero value with ``not applicable'' (NA) is crucial for understanding and interpreting MS data.
+NA values will be managed in missing value handling later on the protein level.
 
 \hypertarget{exclude-reverse-matchescontaminants}{%
-\subsubsection{Exclude reverse
-matches/contaminants}\label{exclude-reverse-matchescontaminants}}
-
-Peptide sequences matching to the reverse protein sequences (=decoy
-database) are considered as possible contaminants. These matches will be
-excluded from further analysis.
+\subsubsection{Exclude reverse matches/contaminants}\label{exclude-reverse-matchescontaminants}}
+
+Peptide sequences matching to the reverse protein sequences (=decoy database) are considered as possible contaminants. These matches will be excluded from further analysis.
 
 \hypertarget{filter-according-to-precursor-ion-fraction-pif}{%
-\subsubsection{Filter according to precursor ion fraction
-(PIF)}\label{filter-according-to-precursor-ion-fraction-pif}}
-
-During mass spectrometry, the ions detected in MS1 are further
-fragmented through collision during multiple MS runs. The resulting
-product ions are derived from precursor ions (also known as mother ions
-or parental ions). Contaminant peptides can co-migrate in this process
-and were distinguished by the lower fraction of their respective
-precursor ions \citep{Tannous2020}. These peptides need to be filtered
-out during the analysis pipeline. A cutoff value, referenced in the
-SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface,
-but it can be adjusted.
+\subsubsection{Filter according to precursor ion fraction (PIF)}\label{filter-according-to-precursor-ion-fraction-pif}}
+
+During mass spectrometry, the ions detected in MS1 are further fragmented through collision during multiple MS runs.
+The resulting product ions are derived from precursor ions (also known as mother ions or parental ions).
+Contaminant peptides can co-migrate in this process and were distinguished by the lower fraction of their respective precursor ions \citep{Tannous2020}.
+These peptides need to be filtered out during the analysis pipeline.
+A cutoff value, referenced in the SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface, but it can be adjusted.
 
@@
 
-The next step for quality control is the exclusion of samples with a
-high false discovery rate (FDR). When applying multiple statistical
-testing (e.g.~t-Test) the obtained p-values can be considered as biased,
-because the probability to observe a significant will iteratively
-increase with each test performed. Corrections in statistics are an
-approach to compensate for the multiplicity of testing. There are many
-ways to do this compensation like the Bonferroni method or
-Benjamini-Hochberg`s FDR. In Mass spectrometry the common ``way to go''
-is calculating a false discovery rate, by dividing false PSMs (=hit of
-the decoy database) through the total number of PSMs above the
-peptide-spectrum matching score. The peptides spectrum matching score is
-defined as \(-10 * log(p)\). Whereas the p-value is defined that the hit
-is done by chance. The calculation of the score is highly dependent on
-the data acquisition method used. MaxQuant uses Andromeda, an integrated
-search engine. Proteome Discoverer from Thermo Fisher utilizes different
-engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot
-and Andromeda showed similar performance when comparing FDR values as a
-function of coverage. However the observed performance can be lower when
-dealing with a decreased coverage \citep{Cox2011}.The threshold for
-accepting an FDR of an individual PSM is described as q-value.
+The next step for quality control is the exclusion of spectra with a high false discovery rate (FDR).
+When applying multiple statistical testing (e.g.~t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant result will iteratively increase with each test performed.
+Corrections in statistics are an approach to compensate for the multiplicity of testing.
+There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR.
+Many search engines face that problem by calculating the false discovery rate, by dividing false PSMs (=hit of the decoy database) with the total number of PSMs above the peptide-spectrum matching score.
+The peptides spectrum matching score is defined as \(-10 * log(p)\).
+Whereas the p-value is defined that the hit is done by chance.
+The calculation of the score is highly dependent on the data acquisition method used.
+MaxQuant uses Andromeda \citep{Cox2011}, an integrated search engine.
+Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot \citep{Brosch2009}.
+As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage.
+However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.
 
 \hypertarget{peptide-spectrum-match-psm-aggregation-to-peptides}{%
-\subsubsection{Peptide spectrum match (PSM) aggregation to
-peptides}\label{peptide-spectrum-match-psm-aggregation-to-peptides}}
-
-In data science, aggregation refers to a row-wise operation that merges
-data based on a particular column using a specific function. In the
-context of processing from PSMs to unique peptides, the desired column
-is the PSM sequence. To account for different distributions across
-multiple assays, the median of the channel is used as the function to
-aggregate multiple matches into one.
+\subsubsection{Peptide spectrum match (PSM) aggregation to peptides}\label{peptide-spectrum-match-psm-aggregation-to-peptides}}
+
+In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function.
+In the context of processing of PSMs to unique peptides, the desired column is the peptide sequence.
+To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.
 
@@
 
-Sample size is often a limiting factor in hypothesis testing. A strict
-quality control and the fact that TMT reagents are only available up to
-18-plex can reduce the number of observed samples below the critical
-threshold, leading to an early end of analyses. To overcome this
-limitation, the provided software is capable of processing multiple runs
-simultaneously, allowing for testing of multiple batches and increasing
-the number of samples that can be included in the analysis. For further
-processing the runs are joined in one object at the peptide level.
+Sample size is often a limiting factor in hypothesis testing.
+A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to diminished statistical power.
+To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis.
+For further processing the runs are joined in one object at the peptide level.
 
 \hypertarget{calculate-median-reporter-ion-intensity-ri-and-filter}{%
-\subsubsection{Calculate median reporter ion intensity (RI) and
-filter}\label{calculate-median-reporter-ion-intensity-ri-and-filter}}
-
-Columns which do not meet the desired intensity can be filtered by a
-threshold set on the RI. The median RI can also be used to check if an
-entire channel has a lower detection level. This can be due two reasons.
-One is the expression level of the given proteins in a cell. Meaning,
-that the expression of the observed cell type is simply lower than the
-other type. Another one could be a spillage of TMT detection in other
-channels due incorrect or missing correction of the TMT isotopes. The
-correction values can be found on the vendor`s webpage (in this case
-Thermo Fisher) and correction needs to be set up in MaxQuant.
+\subsubsection{Calculate median reporter ion intensity (RI) and filter}\label{calculate-median-reporter-ion-intensity-ri-and-filter}}
+
+Columns which do not meet the desired intensity can be filtered by a threshold set on the RI.
+The median RI can also be used to check if an entire channel has a lower detection level.
+This can be due two reasons.
+One is the expression level of the given proteins in a sample.
+Meaning, that the expression of the observed sample type is simply lower than in the other type.
+Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes.
+The correction values can be found on the vendor`s webpage (in this case Thermo Fisher) and correction needs to be set up in MaxQuant.
 
 \hypertarget{calculate-median-coefficient-of-variation-cv-and-filter}{%
-\subsubsection{Calculate median coefficient of variation (CV) and
-filter}\label{calculate-median-coefficient-of-variation-cv-and-filter}}
-
-Depending having a bulk sample or single-cell sample choosing a minimum
-of observed peptides and a cutoff value for the CV, changes the level of
-confidence in the peptide data. The coefficient of variation of a
-peptide is considered as the ratio of the standard deviation to the mean
-and describes the relationship of the observed peptide signal over
-multiple proteins (=razor proteins). Peptides having a high coefficient
-of variation over many razor proteins are considered as noise and need
-to be filtered out before statistical analysis.
+\subsubsection{Calculate median coefficient of variation (CV) and filter}\label{calculate-median-coefficient-of-variation-cv-and-filter}}
+
+Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data.
+The coefficient of variation of a peptide is defined as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins).
+Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.
 
 \hypertarget{remove-peptides-with-high-missing-rate}{%
-\subsubsection{Remove peptides with high missing
-rate}\label{remove-peptides-with-high-missing-rate}}
-
-Although missing value imputation can be performed during the analysis
-of multiple batches, peptides with missing detections across channels
-can be problematic for quantification. The proteomic composition of a
-biological sample is similar between replicates and even across groups.
-However, the threshold of missingness (described as a fraction of the
-row) can be set in the user interface and adjusted.
+\subsubsection{Remove peptides with high missing rate}\label{remove-peptides-with-high-missing-rate}}
+
+Although missing value imputation can be performed during the analysis of multiple batches, peptides with missing detections across channels can be problematic for quantification.
+The proteomic composition of a biological sample is similar between replicates and even across groups.
+However, the threshold of missingness (described as a fraction of the row) can be set in the user interface and adjusted.
 
 \hypertarget{aggregation-of-peptides-to-proteins}{%
-\subsubsection{Aggregation of peptides to
-proteins}\label{aggregation-of-peptides-to-proteins}}
-
-Similar to the already explained previous aggregation step the peptides
-will be further processed into their respective proteins after the
-quality control on the peptide level is performed. Finally an expression
-matrix for every protein and their respective channel column is
-returned. Each channel refers to a sample and contains the intensities
-of each protein found in the biological specimen.
+\subsubsection{Aggregation of peptides to proteins}\label{aggregation-of-peptides-to-proteins}}
+
+Similar to the already explained previous aggregation step the peptides will be further processed into their respective proteins after the quality control on the peptide level is performed.
+Finally an expression matrix for every protein and their respective channel column is returned.
+Each channel refers to a sample and contains the intensities of each protein found in the biological specimen.
 
 \hypertarget{transformation-of-protein-expression-data}{%
-\subsubsection{Transformation of protein expression
-data}\label{transformation-of-protein-expression-data}}
-
-Data transformation applies a function to each value of a matrix or
-array, so that: \[
+\subsubsection{Transformation of protein expression data}\label{transformation-of-protein-expression-data}}
+
+Data transformation applies a function to each value of a matrix or array, so that:
+\[
 y_i = f(x_i)
-\] Depending on the distribution of the values in the observed
-expression set, different transformations can be applied to fulfill
-assumptions for statistical testing. In the shiny application, various
-procedures are implemented and can be further expanded upon request from
-the user. The macrophage analysis done by Specht et al.~mentioned in
-their SCoPE2 publication \citep{Specht2021} uses the logarithm to the
-base 2 to spread a compacted distribution and remove skewness in the
-dataset. This transformation was used as a reference when comparing
-methods. However, using the logarithm to the base 10 may be an easy way
-to recalculate expression data in the mind and will also be facilitated
-by other transformation methods such as the boxcox method, which is also
-implemented in the application. When observing a wide distribution of
-small and large values of the expression set in a histogram, a square
-root transformation can help increase the variability of smaller values
-and decrease the variability of larger values. Depending on the
-distribution, the user has to decide which transformation to consider
-and verify that statistical assumptions are met after application with
-visualizations such as histograms, qq-plots, and MA-plots. For
-statistically inexperienced users, the boxcox transformation can help
-with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992},
-this method applies a linear model against 1 on the data to determine
-the statistical parameter lambda by the maximum likelihood method. The
-log-likelihood method takes the 95\% confidence interval for the
-parameter lambda and the final lambda is chosen as the value with the
-highest log-likelihood value.
+\]
+Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill assumptions for statistical testing.
+In the shiny application, various procedures are implemented and can be further expanded upon request from the user.
+The macrophage analysis done by Specht et al.~mentioned in their SCoPE2 publication \citep{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and remove skewness in the dataset.
+This transformation was used as a reference when comparing methods.
+However, using the logarithm to the base 10 has the same effect and improves the interpretability of the values.
+Furthermore the log10 transformation will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
+When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
+Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qq-plots, and MA-plots.
+For statistically inexperienced users, the boxcox transformation can help with this decision.
+Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a linear model against 1 on the data to determine the statistical parameter lambda by the maximum likelihood method.
+The log-likelihood method takes the 95\% confidence interval for the parameter lambda and the final lambda is chosen as the value with the highest log-likelihood value.
 
@@
 
-Depending on the size of lambda, a certain function will be
-automatically applied to each value of the expression set in order to
-introduce normality. In terms of usability, the boxcox method is the
-most efficient transformation method, taking the decision of which
-calculation to apply away from the user.
+Depending on the size of lambda, a certain function will be automatically applied to each value of the expression set in order to introduce normality.
+In terms of usability, the boxcox method is the most efficient transformation method, taking the decision of which calculation to apply away from the user.
 
@@
 
-The term normalization is ambiguous in data-science and will be
-explained briefly in this chapter. Starting with converting the data to
-the Z-distribution (Gaussian result), where 0 represents the mean and 1
-represents the standard deviation, in vectorization every feature (or
-protein) is represented as a vector that points in a specific direction
-in the unit sphere, normalization can have different meanings.
-Normalization is considered as part of the scaling procedure and ensures
-that every feature contributes equally to the statistical model and
-hinders large values from biasing the model in a particular direction.
-However, this could also have negative consequences for the modeling
-procedure, as it may reduce the impact of important features on the
-dataset. Depending on the algorithms later used in the data processing,
-the method of choice for normalization can change the results
-significantly. The K-nearest neighbor (K-NN) clustering algorithm, for
-instance relies on the distances between distinct data points by
-minimizing variance of the squared euclidean distances. This method will
-be applied in the missing value imputation, which will be the next step
-applied to the data. The first normalization method available in the
-user interface is column-wise median and row-wise mean normalization.
-This method was chosen by Specht et al in the SCoPE2 publication 2021
-\citep{Specht2021} and worked as a reference during the development
-process to benchmark other methods. When dividing each value of an
-expression set by the median of the particular column, the procedure is
-considered as a column-wise median normalization. Although the
-mean-normalization works the same way but takes the mean as the second
-variable. In the analysis pipeline both procedures were applied on the
-dataset. However when the logarithm as the transformation method was
-chosen as transformation method, ProteoScanR automatically switches from
-a division to a subtraction, since:
-\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\). After applying
-this normalization method, each value can be interpreted as a distance
-or fraction of the mean or median of the corresponding column or row,
-enabling fair comparisons between features and samples. The second
-method for normalization included is the CONSTANd method, which was
-proven suitable for relative quantification in the field of proteomics
-\citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq
-data. CONSTANd uses a technique called matrix raking, which employs the
-RAS algorithm. The expression set reflects the non-negative real
-\((m, n)\) matrix \(A\) were the bi proportional constrained matrix
-problem will be solved by finding the \((m, n)\) matrix \(B\) which
-equals to \(diag(x) * A *diag(y)\) . Whereas \(x \in \mathbb{R}^{m}\)
-and \(y \in \mathbb{R}^{n}\). The solution to this problem involves
-finding the row sum of \(B=u_i\) and the column sum of \(B=v_j\), where
-\(i\) and \(j\) denote the row and column indices of the matrix,
-respectively \citep{Bacharach1965}. In other words, the matrix will be
-alternatively manipulated on rows and columns until the mean of both
-equals 1. In order to apply the method and yield true results, one has
-to consider 3 major assumptions between sample types. This assumption
-can be observed in the MA-plot which indicates the differences of two
-samples. Given the intensities of two samples, \(R\) and \(G\),
-visualized on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\) and
-the y-axis as \(A=\frac{1}{2}\log_2(RG)\). A comparison between all
-sample types of identically processed sets is needed before calculation
-and the assumptions must be fulfilled.
+The term normalization is ambiguous in data-science and will be explained briefly in this chapter.
+Starting with converting the data to the Z-distribution (Gaussian result), where 0 represents the mean and 1 represents the standard deviation, in vectorization every feature (or protein) is represented as a vector that points in a specific direction in the unit sphere, normalization can have different meanings.
+Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to the statistical model and hinders large values from biasing the model in a particular direction.
+However, this could also have negative consequences for the modeling procedure, as it may reduce the impact of important features on the dataset.
+Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly.
+The K-nearest neighbor (K-NN) clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances.
+This method will be applied in the missing value imputation, which will be the next step applied to the data.
+The first normalization method available in the user interface is column-wise median and row-wise mean normalization.
+This method was chosen by Specht et al in the SCoPE2 publication 2021 \citep{Specht2021} and worked as a reference during the development process to benchmark other methods.
+When dividing each value of an expression set by the median of the particular column, the procedure is considered as a column-wise median normalization.
+Although the mean-normalization works the same way but takes the mean as the second variable.
+In the analysis pipeline both procedures were applied on the dataset.
+However when the logarithm as the transformation method was chosen as transformation method, ProteoScanR automatically switches from a division to a subtraction, since:
+\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\).
+After applying this normalization method, each value can be interpreted as a distance or fraction of the mean or median of the corresponding column or row, enabling fair comparisons between features and samples.
+The second method for normalization included is the CONSTANd method, which was proven suitable for relative quantification in the field of proteomics \citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq data.
+CONSTANd uses a technique called matrix raking, which employs the RAS (\(K=RAS\)) algorithm.
+The expression set reflects the non-negative real \((m, n)\) matrix \(A\) were the bi proportional constrained matrix problem will be solved by finding the \((m, n)\) matrix \(K\) which equals to \(diag(x) * A *diag(y)\), so that \(K=diag(x) * A *diag(y)= RAS\).
+Whereas \(x \in \mathbb{R}^{m}\), \(y \in \mathbb{R}^{n}\) and \(R\) denotes for the row multiplier and \(S\) for the column multiplier.
+The solution to this problem involves finding the row sum of \(K=u_i\) and the column sum of \(K=v_j\), where \(i\) and \(j\) denote the row and column indices of the matrix, respectively \citep{Bacharach1965}.
+In other words, the matrix will be alternatingly manipulated on rows and columns until the mean of both equals 1.
+In order to apply the method and yield true results, one has to consider 3 major assumptions between sample types.
+These assumption can be observed in the MA-plot which indicates the differences of two samples.
+Given the intensities of two samples, \(R\) and \(G\), visualized on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\) and the y-axis as \(A=\frac{1}{2}\log_2(RG)\).
+A comparison between all sample types of identically processed sets is needed before calculation and the assumptions must be fulfilled.
 
@@
 
-If the MA-plot does not visualize any of these violations the CONSTANd
-method can be applied on the dataset and the biological meaning of the
-sample types can be assessed. The method is utilized by the bioconductor
-package handler. Another method developed in the application is the
-quantile normalization. Quantile normalization is performed over all
-columns, equalizing statistical parameters such as inter quartile range,
-median and mean. However the application of this method is only suitable
-if the number of differentially expressed proteins is approximately
-equal over sample types. Otherwise statistical power could be lost by
-removing fine deviations \citep{Zhao2020}. For the quantile
-normalization method the package limma was utilized. Quantile
-normalization ranks protein expression by magnitude, then averages the
-expression values occupying the same rank and substitute the inital
-value with the average. However if the dataset already consists of
-balanced features and samples or the user wants to benchmark the methods
-applied, normalization can be skipped also, by setting to ``none''.
+If the MA-plot does not visualize any of these violations the CONSTANd method can be applied on the dataset and the biological meaning of the sample types can be assessed.
+The method is utilized by the bioconductor package handler.
+Another method developed in the application is the quantile normalization.
+Quantile normalization is performed over all columns, equalizing statistical parameters such as inter quartile range, median and mean.
+However the application of this method is only suitable if the number of differentially expressed proteins is approximately equal over sample types.
+Otherwise statistical power could be lost by removing fine deviations \citep{Zhao2020}.
+For the quantile normalization method the package limma was utilized.
+Quantile normalization ranks protein expression by magnitude, then averages the expression values occupying the same rank and substitute the inital value with the average.
+However if the dataset already consists of balanced features and samples or the user wants to benchmark the methods applied, normalization can be skipped also, by setting to ``none''.
 
@@
 
-When analyzing multiple batches simultaneously, it is possible to
-encounter situations where a particular protein is not detected in one
-of the batches. From a biological perspective, it is highly unlikely
-that a cell completely lacks a single protein, as it would imply a
-complete loss of function for that protein. Therefore, an intensity
-value of 0 is close to impossible for most proteins. In order to handle
-this dark space in the expression set, several methods can be chosen in
-the application. A variety of function employs replacing the missing
-values with the mean or median of the rest of the matrix. This procedure
-introduces no bias in the data if the missingness is low, although
-differential expression for the sample carrying the missing feature can
-not be expected as well. Even a more conservative approach is dropping
-the rows for the feature with the missing value in one of the observed
-samples. A common practice in proteomics is utilizing the K-nearest
-neighbor algorithm (=KNN), which imputes the missing values by a
-euclidean metrics of the neighboring values in the columns where the
-feature is not missing. \(K\) denotes for the count of neighboring
-values of a gene and can be selected upon preference. The euclidean
-distance between two points is defined as \(d = \sqrt{a^2 * b^2}\), the
-\(K\) nearest neighbors average value will be assigned to the intensity
-of the missing feature. Since the average is a statistical metric which
-can be biased easily by skewed distributions, normalization is an
-important task before performing missing value imputation. KNN was
-benchmarked for the regression of protein abundance intensities and
-showed an Area under the Curve (AUC) above 0.7 when testing on a human
-proteome test set \citep{Lan2013}.
+When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches.
+In order to handle this dark space in the expression set, several methods can be chosen in the application.
+A variety of function employs replacing the missing values with the mean or median of the rest of the matrix (total mean/median).
+This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well.
+An even more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
+A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values according to an euclidean metric of the neighboring values in the columns where the feature is not missing.
+\(K\) denotes for the count of neighboring values of a protein which are considered and can be selected upon preference.
+The euclidean distance between two points is defined as \(d = \sqrt{a^2 * b^2}\), the \(K\) nearest neighbors average value will be assigned to the intensity of the missing feature.
+Since the average is a statistical metric which can be biased easily by skewed distributions, normalization is an important task before performing missing value imputation.
+KNN was benchmarked for the regression of protein abundance intensities and showed an Area under the Curve (AUC) above 0.7 when testing on a human proteome test set \citep{Lan2013}.
 
@@
 
-Since mass spectrometry experiments are highly influenced by several
-factors, such as sample processing, the technician, manufacturer and
-production of the device, the reagents etc., working in multiple batches
-introduces further variance in the protein data. This variance would
-overshadow the biological differences across sample types when
-performing dimensionality reduction. A common misconception in biology
-is, that batches need to be corrected before differential expression
-analysis, which is the final aim of ProteoScanR. It appears convenient
-to remove the batch effect before performing any statistical analysis,
-which is in turn prone to errors especially if the sample groups are not
-split equally between batches. In this case, group differences influence
-the batch effect reducing statistical power after the correction. These
-both effects are so called interdependent. The opposite could happen if
-the experimental design is heavily unbalanced leading to group
-differences induced by batch correction. Finally it is recommended to
-include the batch factor should in the linear model as a co-factor and
-statistically quantify it \citep{Nygaard2016}. However it could be of
-interest to observe biological relevance of the factor of interest in
-the dimensionality reduction visualization, so a batch correction option
-was included to meet this need. To address the non-biological
-experimental variations, also known as technical factors, the R package
-sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was
-used. The expression of a gene \(g\) for the batch \(m_i\) and sample
-\(j\) can be defined as
-\(Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}\),
-where \(\gamma\) and \(\delta\epsilon\) represent the systematic error
-caused by the batch. The Bayesian framework is employed to remove this
-systematic error by shrinking the effect and pooling information over
-the features \(g\). Ideally the effects \(a\) and \(\beta\) remain as
-natural as they are and should reflect the true biological difference
-between experimental obtained samples or groups. Before applying batch
-correction, the data is standardized using Z-transformation. By the
-method of moments (= the mean and the variance) the two batch effect
-parameters \(\gamma\) and \(\delta\epsilon\) were estimated and can be
-subtracted from each value for the particular feature driving
-statistical moments at a comparable level between batches. The ComBat
-function allows the user to specify a desired effect, which represents
-the actual biological variance to be preserved while removing
-non-biological variance. This is achieved by constructing a design
-matrix, which is provided as prior information about the dataset within
-the Bayesian framework. However, in cases of poor experimental design,
-the factors contributing to unwanted and desired variation may be
-confounded. In the worst case, if only one sample type per run is
-observed, it becomes challenging for the program to distinguish between
-batch-induced variance and biological variance. In such cases, the user
-is advised to adjust the experimental layout for future experiments. The
-detection algorithm for confounding effects is facilitated by QR matrix
-decomposition. The matrix \(A=QR\) is decomposed into an orthogonal
-matrix \(Q\) and an upper triangular matrix \(R\). By comparing the rank
-of matrix \(A\) with the number of biologically relevant factors, it is
-possible to determine the presence of confounding effects in the
-dataset. The rank is defined as the maximum number of linearly
-independent columns. The same approach is used in the limma package
-\citep{Phipson2016} to identify non-solvable coefficients in
-experimental designs.
+Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data.
+This variance would overshadow the biological differences across sample types when performing dimensionality reduction.
+A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of ProteoScanR.
+It appears convenient to remove the batch effect before performing any statistical analysis, which is in turn prone to errors especially if the sample groups are not split equally between batches.
+In this case, group differences influence the batch effect reducing statistical power after the correction.
+These both effects are so called interdependent.
+The opposite could happen if the experimental design is heavily unbalanced leading to group differences induced by batch correction.
+Finally it is recommended to include the batch factor should in the linear model as a co-factor and statistically quantify it \citep{Nygaard2016}.
+However it could be of interest to observe biological relevance of the factor of interest in the dimensionality reduction visualization, so a batch correction option was included to meet this need.
+To address the non-biological experimental variations, also known as technical factors, the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used.
+The expression of a gene \(g\) for the batch \(m_i\) and sample \(j\) can be defined as \(Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}\), where \(\gamma\) and \(\delta\epsilon\) represent the systematic error caused by the batch.
+The Bayesian framework is employed to remove this systematic error by shrinking the effect and pooling information over the features \(g\).
+Ideally the effects \(a\) and \(\beta\) remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups.
+Before applying batch correction, the data is standardized using Z-transformation.
+By the method of moments (= the mean and the variance) the two batch effect parameters \(\gamma\) and \(\delta\epsilon\) were estimated and can be subtracted from each value for the particular feature driving statistical moments at a comparable level between batches.
+The ComBat function allows the user to specify a desired effect, which represents the actual biological variance to be preserved while removing non-biological variance.
+This is achieved by constructing a design matrix, which is provided as prior information about the dataset within the Bayesian framework.
+However, in cases of poor experimental design, the factors contributing to unwanted and desired variation may be confounded.
+In the worst case, if only one sample type per run is observed, it becomes challenging for the program to distinguish between batch-induced variance and biological variance.
+In such cases, the user is advised to adjust the experimental layout for future experiments.
+The detection algorithm for confounding effects is facilitated by QR matrix decomposition. The matrix \(A=QR\) is decomposed into an orthogonal matrix \(Q\) and an upper triangular matrix \(R\).
+By comparing the rank of matrix \(A\) with the number of biologically relevant factors, it is possible to determine the presence of confounding effects in the dataset.
+The rank is defined as the maximum number of linearly independent columns.
+The same approach is used in the limma package \citep{Phipson2016} to identify non-solvable coefficients in experimental designs.
 
 \hypertarget{dimensionality-reduction}{%
-\subsubsection{Dimensionality
-reduction}\label{dimensionality-reduction}}
-
-After cleaning the data and crystallizing from impurities such as noise
-and unwanted effects, the biological differences between sample types
-can be obtained. Proteomic data consists of multi dimensions (=features)
-which are overall hard to understand by a glance. An approach to observe
-the differences without a closer sight of exact effect on the
-proteinaceous ensemble, but still keep a high level of information can
-be dimensionality reduction. After the cleaning procedures any
-additional effect caused by technical or biological replication should
-be removed, however it could be found be highlighting according to the
-particular factor. This makes the dimensionality reduction a useful tool
-before starting any statistical analysis by checking for involuntarily
-introduced bias.
+\subsubsection{Dimensionality reduction}\label{dimensionality-reduction}}
+
+After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained.
+Proteomic data consists of multi dimensions (=features) which are overall hard to understand by a glance.
+An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keep a high level of information can be dimensionality reduction.
+After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor.
+This makes the dimensionality reduction a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.
 
 \hypertarget{principle-component-analysis-pca}{%
-\paragraph{Principle component analysis
-(PCA)}\label{principle-component-analysis-pca}}
-
-The most common approach in data science for any kind of high
-dimensional data is the PCA. For ProteoScanR the bioconductor package
-scater \citep{McCarthy2017} was used. Every principle component can be
-understood as a vector pointing in a specific direction derived as the
-eigenvector from the covariance matrix of the expression set. The
-covariance matrix indicates for the co linearity between the variance of
-all elements compared. After calculating the covariance matrix the
-eigenvectors for the covariance matrix will be obtained by solving the
-quadratic equations to obtain the eigenvalues \(\lambda\). Since
-\(A v = \lambda v\) the eigenvector \(v\) can be obtained by solving the
-equation for each element of the vector. The eigenvectors \(v\) never
-changes is direction and explains the covariance within the dataset and
-so also the mathematical differences between the samples. The
-eigenvectors are sorted according to their eigenvalues in decreasing
-order and therefore decreasing explained variance. For graphical
-visualizations the principle components are plotted against each other.
-The first two components, which explain the majority of variance in the
-dateset highlight the difference between samples.
+\paragraph{Principle component analysis (PCA)}\label{principle-component-analysis-pca}}
+
+The most common approach in data science for any kind of high dimensional data is the PCA.
+For ProteoScanR the bioconductor package scater \citep{McCarthy2017} was used.
+Every principle component can be understood as a vector pointing in a specific direction derived as the eigenvector from the covariance matrix of the expression set.
+The covariance matrix indicates for the co linearity between the variance of all elements compared.
+After calculating the covariance matrix the eigenvectors for the covariance matrix will be obtained by solving the quadratic equations to obtain the eigenvalues \(\lambda\).
+Since \(A v = \lambda v\) the eigenvector \(v\) can be obtained by solving the equation for each element of the vector.
+The eigenvectors \(v\) never changes is direction and explains the covariance within the dataset and so also the mathematical differences between the samples.
+The eigenvectors are sorted according to their eigenvalues in decreasing order and therefore decreasing explained variance.
+For graphical visualizations the principle components are plotted against each other.
+The first two components, which explain the majority of variance in the dateset highlight the difference between samples.
 
 \hypertarget{uniform-manifold-approximation-projection-umap}{%
-\paragraph{Uniform manifold approximation \& projection
-(UMAP)}\label{uniform-manifold-approximation-projection-umap}}
-
-UMAP \citep{McInnes2018} is a dimensionality reduction technique that
-was developed in recent years and is based on Riemann geometry. It
-shares similarities to the t-distributed stochastic neighbor embedding
-(t-SNE) in terms of graphical visualization. Riemann geometry is a
-mathematical framework that deals with three dimensional functions using
-volumes, areas and vectors, which are also concept of topological data
-analysis. Before performing UMAP, three assumptions need to be
-considered:
+\paragraph{Uniform manifold approximation \& projection (UMAP)}\label{uniform-manifold-approximation-projection-umap}}
+
+UMAP \citep{McInnes2018} is a dimensionality reduction technique that was developed in recent years and is based on Riemann geometry.
+It shares similarities to the t-distributed stochastic neighbor embedding (t-SNE) in terms of graphical visualization.
+Riemann geometry is a mathematical framework that deals with three dimensional functions using volumes, areas and vectors, which are also concept of topological data analysis.
+Before performing UMAP, three assumptions need to be considered:
 
@@
 
-The UMAP algorithm begins by constructing a graph with a defined
-neighborhood parameter, typically denoted as \(k\), which can be
-selected computationally. UMAP employs the k-nearest neighbor descent
-algorithm (NN-descent). Similar to the k-nearest neighbor algorithm
-(KNN), used in the missing value imputation part of this project,
-NN-descent recursively iterates to find the smallest distance to the
-respective neighbors within the defined neighborhood size \(k\).
-Increasing the value of \(k\) gradually strengthens the clustering of
-the data, but at a certain point, fine structure may be lost, and only
-rough estimates of the underlying principle can be observed. The
-mathematical graph in UMAP is defined by nodes connected by edges, with
-each node internally connected to at least one other node. Edges are
-distinguished from each other not only by their connection but also by
-their importance, which is often referred to as weight. Whereas the
-weight function is calculated as
+The UMAP algorithm begins by constructing a graph with a defined neighborhood parameter, typically denoted as \(k\), which can be selected computationally.
+UMAP employs the k-nearest neighbor descent algorithm (NN-descent).
+Similar to the k-nearest neighbor algorithm (KNN), used in the missing value imputation part of this project, NN-descent recursively iterates to find the smallest distance to the respective neighbors within the defined neighborhood size \(k\).
+Increasing the value of \(k\) gradually strengthens the clustering of the data, but at a certain point, fine structure may be lost, and only rough estimates of the underlying principle can be observed.
+The mathematical graph in UMAP is defined by nodes connected by edges, with each node internally connected to at least one other node.
+Edges are distinguished from each other not only by their connection but also by their importance, which is often referred to as weight.
+Whereas the weight function is calculated as
 
@@
 \[\sum_{j=1}^{k} \exp ( \frac{ -\max(0, d(x_{i}, x_{i_{j}}) - \rho_i)} { \sigma_i } ) = \log_2(k)\]
-and acts as a normalization factor. The obtained graph \(\overline{G}\)
-is called the fuzzy simplicial set which is obtained by approximating
-the geodesic distance (= locally length-minimizing curve) of the data
-points, leading to a topological approximation. The second phase of the
-algorithm adjusts the layout of the weighted graph to a representative
-view, but preserves the characteristics and shows the underlying
-topological principle of the data. In order to obtain a visual
-representation of the graph, the algorithm applies repulsive forces:
+and acts as a normalization factor.
+The obtained graph \(\overline{G}\) is called the fuzzy simplicial set which is obtained by approximating the geodesic distance (= locally length-minimizing curve) of the data points, leading to a topological approximation.
+The second phase of the algorithm adjusts the layout of the weighted graph to a representative view, but preserves the characteristics and shows the underlying topological principle of the data.
+In order to obtain a visual representation of the graph, the algorithm applies repulsive forces:
 \[\frac{2b}{(\epsilon + ||y_i-y_j||_{2}^{2})(1+a||y_i-y_j||_{2}^{2b})}(1-w((x_i,x_j)))(y_i-y_j)\]
@@
 \[\frac{ -2ab || y_i -y_j||_{2}^{2(b-1)} }{1+||y_i-y_j||_{2}^{2}} w((x_i,x_j))(y_i-y_j)\]
-along edges. Where \(a\) and \(b\) are denoted as hyperparameters. By
-iterating through possible conformations until converging towards a
-local minimum with decreasing repulsive and attractive forces, the UMAP
-reaches it`s aim to reveal the underlying properties of the data. The
-application utilizes the bioconductor package scater for performing UMAP
-\citep{McCarthy2017}.
+along edges.
+Where \(a\) and \(b\) are denoted as hyperparameters.
+By iterating through possible conformations until converging towards a local minimum with decreasing repulsive and attractive forces, the UMAP reaches it`s aim to reveal the underlying properties of the data.
+The application utilizes the bioconductor package scater for performing UMAP \citep{McCarthy2017}.
 
 \hypertarget{validation-of-the-methods-applied-to-the-dataset}{%
-\subsubsection{Validation of the methods applied to the
-dataset}\label{validation-of-the-methods-applied-to-the-dataset}}
-
-For the validation of methods applied to the dataset, the mutual
-information calculation is used to assess the impact of computations and
-the information it carries. Mutual information, denoted as \(I(X;Y)\),
-is a measure of the dependence or information shared between two
-variables, \(X\) and \(Y\), in this case sample types. It is calculated
-using the equation \(I(X;Y) = H(X,Y) - H(X|Y) - H(Y|X)\), where \(H\)
-represents the entropy, \(H(X|Y)\) and \(H(Y|X)\) denote for the both
-conditional probabilities and \(H(X,Y)\) for the joint entropy. Entropy,
-as a concept in information theory, quantifies the level of disorder or
-uncertainty associated with a random event. Events that are highly
-likely to occur are considered less informative than events that are
-more unexpected or less likely. In a Venn diagram \(I(X;Y)\) (the mutual
-information) corresponds to the intersection of the two conditional
-entropys \(H(X|Y)\) and \(H(Y|X)\). To calculate the mutual information
-between two variables \(X\) and \(Y\) the above equation can be
-alternatively expressed as
-\(I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})\). \(D_{KL}\)
-denotes for the Kullback-Leibler \citep{Kullback1951} divergence which
-is a concept of relative entropy between two variables and tells us how
-much 2 probability distributions differ from each other.
-\(P_{X} \otimes P_{Y}\) represents the product of the marginals. The
-underlying principle of divergence is the integral asymmetry in Bayesian
-inference, which means that divergence does not satisfy the triangle
-inequality. Differences are not the same depending on the starting
-point. By comparing the mutual information before performing any
-calculations with the mutual information calculated after the
-computations, it is possible to assess whether the applied methods have
-significantly altered the dependence or information content. The goal is
-to choose a method which preserves the information contributed by each
-biological sample type to the experiment. The R package ``infotheo'' is
-utilized to perform the estimates for the entropy of the two variables.
-Mutual information is returned as natural unit of information (=nat), a
-measurement proportional to the to the Shannon entropy
-(\(1nat = \frac{1}{ln2} shannons\)). Before calculating the pairwise
-mutual information of the expression matrix, values were discretized
-into bins with the discretize function of above package. The mutual
-information is calculated within each sample type, and it can also be
-calculated between a selection of sample types to determine if the
-computations performed affect the information content. After that all
-pairwise mutual information within the expression set before and after
-ProteoScanR is obtained. The pairwise differences can be observed as a
-boxplot for a particular sample type indicating the change in dependence
-within and also compared to another sample type.
+\subsubsection{Validation of the methods applied to the dataset}\label{validation-of-the-methods-applied-to-the-dataset}}
+
+For the validation of methods applied to the dataset, the mutual information calculation is used to assess the impact of computations and the information it carries.
+Mutual information, denoted as \(I(X;Y)\), is a measure of the dependence or information shared between two variables, \(X\) and \(Y\), in this case sample types.
+It is calculated using the equation \(I(X;Y) = H(X,Y) - H(X|Y) - H(Y|X)\), where \(H\) represents the entropy, \(H(X|Y)\) and \(H(Y|X)\) denote for the both conditional probabilities and \(H(X,Y)\) for the joint entropy.
+Entropy, as a concept in information theory, quantifies the level of disorder or uncertainty associated with a random event.
+Events that are highly likely to occur are considered less informative than events that are more unexpected or less likely.
+In a Venn diagram \(I(X;Y)\) (the mutual information) corresponds to the intersection of the two conditional entropys \(H(X|Y)\) and \(H(Y|X)\).
+To calculate the mutual information between two variables \(X\) and \(Y\) the above equation can be alternatively expressed as \(I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})\).
+\(D_{KL}\) denotes for the Kullback-Leibler \citep{Kullback1951} divergence which is a concept of relative entropy between two variables and tells us how much 2 probability distributions differ from each other.
+\(P_{X} \otimes P_{Y}\) represents the product of the marginals.
+The underlying principle of divergence is the integral asymmetry in Bayesian inference, which means that divergence does not satisfy the triangle inequality.
+Differences are not the same depending on the starting point.
+By comparing the mutual information before performing any calculations with the mutual information calculated after the computations, it is possible to assess whether the applied methods have significantly altered the dependence or information content.
+The goal is to choose a method which preserves the information contributed by each biological sample type to the experiment.
+The R package ``infotheo'' is utilized to perform the estimates for the entropy of the two variables.
+Mutual information is returned as natural unit of information (=nat), a measurement proportional to the to the Shannon entropy (\(1nat = \frac{1}{ln2} shannons\)).
+Before calculating the pairwise mutual information of the expression matrix, values were discretized into bins with the discretize function of above package.
+The mutual information is calculated within each sample type, and it can also be calculated between a selection of sample types to determine if the computations performed affect the information content.
+After that all pairwise mutual information within the expression set before and after ProteoScanR is obtained.
+The pairwise differences can be observed as a boxplot for a particular sample type indicating the change in dependence within and also compared to another sample type.
 
@@
 
-After visualizing the biological tendency in the dataset, the next step
-is hypothesis testing. Several experimental designs can be tested
-against the null hypothesis \(h_0\), which states that there is no
-significant difference between the sample types. Before proceeding with
-any statistical testing, it is important to observe the data using
-either a quantile-quantile plot (qq-plot) or a histogram. The qq-plot
-serves a vital role in exploratory statistics by visualizing whether two
-datasets follow the same underlying distributions. In the case of
-statistical preparation for parametric tests, the first distribution is
-the dataset of interest, while the second dataset is the normal
-distribution. If the dataset follows an exact normal distribution, the
-values will align with the red line in the chart. However, values at the
-ends of the distribution may deviate from the normality line, indicating
-differentially expressed genes. These genes can be observed both in the
-histogram and the qq-plot. Parametric statistical tests rely on the
-assumption of normality in the data, which needs to be achieved before
-building a model. To achieve normality, it is recommended to try various
-combinations of transformation and normalization methods. By observing
-the results in the plots, one can determine which combination of data
-processing methods satisfies the conditions required for the tests.
-However the limma method utilized in the statistics module shows great
-robustness to imperfect normal distributed values \citep{Ritchie2015}.
-
-To test for differential expression, the analysis utilizes the ``limma''
-package from R Bioconductor \citep{Phipson2016}. This package employs a
-linear model approach to determine the fold expression between sample
-groups. Before initiating the analysis, two matrices need to be
-computed. The design matrix identifies samples based on their sample
-type and defines the experimental design. An algorithm within the
-program's backend logic is utilized to automatically generate this
-matrix according to the user's selection in the interface. The
-expression matrix contains intensities for each identified protein and
-is obtained at the end of ProteoScanR. By calling a function with the
-design and expression matrices as arguments, the contrast matrix is
-calculated, enabling the user to decide which comparisons to consider
-and also perform tests on multiple factors. When observing multiple
-batches at once, as already explained in the chapter batch correction
-this effect needs to be taken into consideration when performing
-differential expression analysis. This is done by the multi-factor
-option in the user interface by selecting the batch as a co-factor and
-inclusion in the design matrix. The advantage of this procedure is, that
-the biological relevance of working in multiple batches can be
-quantified and the publishing of false positive results avoided. The
-multi factor option uses an additive model as a design matrix, whereas
-continuous variables (such as size, weight etc.) can also be used to
-build the statistical model.
-
-The subsequent step involves creating a linear model between groups
-using log-ratios of their expression values. In a two-way design, the
-expression of the gene \(y\) can be explained as
-\(Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \epsilon\), where \(\beta_0\)
-represents the intercept, which can be interpreted as the mean
-expression of the gene. \(\beta_1\) represents difference in mean of the
-treatment (or condition) on the discrete or continuous variable
-\(X_{1}\), and \(\epsilon\) acts as an error term. The interface allows
-for additive models with multiple factors, allowing the user to add
-factors according to various scenarios. For example, in a two-factor
-model, it would look like:
-\(Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + \epsilon\).
-The principle remains the same for models with multiple factors. When
-visualizing the linear model, the x-axis would represent the gene for
-the sample types or other factors, whereas the y-axis would account for
-the expression value. By drawing a line through the cloud of data points
-the model is constructed by minimizing the quadratic distance to every
-data-point, called the least-square method or also ordinary least-square
-method. With the regression line the above mentioned equation can be
-constructed.
-
-Next the statistical module of the program performs an empirical Bayes
-model utilizing the eBayes \citep{Smyth2004} function of the Limma
-package. The eBayes function utilizes a moderated t-statistics following
-t-distributed values. The advantage of this method over the posterior
-odds is the reduction of hyperparameters to estimate used for the
-modelling process, also called shrinkage. In this context, the
-hyperparameters to estimate are the coefficient \(\beta_{gj}\) for gene
-\(g\) and sample \(i\), as well as the variance \(\sigma^2_g\) across
-genes \(g\). In the Bayesian approach, probabilities are updated after
-obtaining new data, referred to as conditional probabilities. For
-estimation, it is assumed that
-\(\frac{1}{\sigma^2_g} \sim \frac{1}{d_{0}s^2_{0}} \chi^2_{d_{0}}\),
-representing the prior information of the model. The probability for
-\(\beta_{gj} \neq 0\) is denoted as \(p_j\), where \(p_j\) represents
-the expected proportion of truly differentially expressed genes. The
-expected distribution of log-fold changes follows the distribution
-\(\beta_{gj} | \sigma^2_g , \beta_{gj} \neq 0 \sim N(0, v_{0j}\sigma^2_g)\).
-The posterior mean of \(\sigma^2_g\) given \(s^2_g\) is calculated as
-\(s^{-2}_g = E(\sigma^2_g | s^2_g) = \frac{d{0}s^2_{0} + d_{g}s^2_{g}}{d_{0} + d_{g}}\).
-This is where the shrinkage occurs, as the posteriors affect the prior
-values based on their respective sizes and degrees of freedom. The
-moderated t-statistic is defined as
-\(\tilde{t}_{gj} = \frac{\hat{\beta}_{gj}}{\hat{s}_{g}\sqrt{v_{gj}}}\).
-Once the values of \(\tilde{t}\) and \(s^{2}\) are calculated, the
-posterior odds are computed, providing the odds that a particular gene
-is differentially expressed. The odds for gene \(g\) being
-differentially expressed are denoted as
-\(O_{gj} = \frac{p(\beta_{gj} \neq 0 | \tilde{t}_{gj}, s^2{g})}{p(\beta_{gj} = 0 | \tilde{t}_{gj}, s^2{g})}\),
-where the numerator represents the probability that the gene is
-differentially expressed, and the denominator represents the probability
-that the gene is not differentially expressed \citep{Smyth2004}.
-
-Limma outputs a table with all proteins, their respective logfold
-change, p-values and adjusted p-values. The user can choose different
-methods for the correction of the p-values, which is necessary when
-testing the same hypothesis multiple times. A common method in omics
-studies is the false discovery rate (=fdr) by Benjamini Hochberg
-\citep{Benjamini1995}. By ranking all the p-values from smallest to
-largest the p-values are adjusted sequentially using the formula
-\(p_(i) = \frac{r_i}{m} Q\). The rank in the p-value table is denoted as
-\(r_i\) , \(m\) reflects the total number of tests and \(Q\) represents
-false discovery rate which is set to 5\%. The second option to choose is
-the Benjamini \& Yekutieli (=BY) \citep{Benjamini2001}method, which is a
-further development of the fdr by the same statistician and has a wider
-range of application based on the dependencies for the test. The third
-and most conservative option is the Holm method \citep{Holm1979} which
-controls the error sequentally in a family wise manner, which is
-considered as traditional in regards to the above mentioned experimental
-wise error rates. The above mentioned correction methods are ordered in
-the way from the most liberal to the most conservative one. The
-selection is supposed to give the user the possibility questioning
-results on multiple levels.
+After visualizing the biological tendency in the dataset, the next step is hypothesis testing.
+Several experimental designs can be tested against the null hypothesis \(h_0\), which states that there is no significant difference between the sample types.
+Before proceeding with any statistical testing, it is important to observe the data using either a quantile-quantile plot (qq-plot) or a histogram.
+The qq-plot serves a vital role in exploratory statistics by visualizing whether two datasets follow the same underlying distributions.
+In the case of statistical preparation for parametric tests, the first distribution is the dataset of interest, while the second dataset is the normal distribution.
+If the dataset follows an exact normal distribution, the values will align with the red line in the chart.
+However, values at the ends of the distribution may deviate from the normality line, indicating differentially expressed genes.
+These genes can be observed both in the histogram and the qq-plot.
+Parametric statistical tests rely on the assumption of normality in the data, which needs to be achieved before building a model.
+To achieve normality, it is recommended to try various combinations of transformation and normalization methods.
+By observing the results in the plots, one can determine which combination of data processing methods satisfies the conditions required for the tests.
+However the limma method utilized in the statistics module shows great robustness to imperfect normal distributed values \citep{Ritchie2015}.
+
+To test for differential expression, the analysis utilizes the ``limma'' package from R Bioconductor \citep{Phipson2016}.
+This package employs a linear model approach to determine the fold expression between sample groups.
+Before initiating the analysis, two matrices need to be computed.
+The design matrix identifies samples based on their sample type and defines the experimental design.
+An algorithm within the program's backend logic is utilized to automatically generate this matrix according to the user's selection in the interface.
+The expression matrix contains intensities for each identified protein and is obtained at the end of ProteoScanR.
+By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors.
+When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis.
+This is done by the multi-factor option in the user interface by selecting the batch as a co-factor and inclusion in the design matrix.
+The advantage of this procedure is, that the biological relevance of working in multiple batches can be quantified and the publishing of false positive results avoided.
+The multi factor option uses an additive model as a design matrix, whereas continuous variables (such as size, weight etc.) can also be used to build the statistical model.
+
+The subsequent step involves creating a linear model between groups using log-ratios of their expression values.
+In a two-way design, the expression of the gene \(y\) can be explained as \(Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \epsilon\), where \(\beta_0\) represents the intercept, which can be interpreted as the mean expression of the gene.
+\(\beta_1\) represents difference in mean of the treatment (or condition) on the discrete or continuous variable \(X_{1}\), and \(\epsilon\) acts as an error term.
+The interface allows for additive models with multiple factors, allowing the user to add factors according to various scenarios.
+For example, in a two-factor model, it would look like: \(Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + \epsilon\).
+The principle remains the same for models with multiple factors.
+When visualizing the linear model, the x-axis would represent the gene for the sample types or other factors, whereas the y-axis would account for the expression value.
+By drawing a line through the cloud of data points the model is constructed by minimizing the quadratic distance to every data-point, called the least-square method or also ordinary least-square method.
+With the regression line the above mentioned equation can be constructed.
+
+Next the statistical module of the program performs an empirical Bayes model utilizing the eBayes \citep{Smyth2004} function of the Limma package.
+The eBayes function utilizes a moderated t-statistics following t-distributed values.
+The advantage of this method over the posterior odds is the reduction of hyperparameters to estimate used for the modelling process, also called shrinkage.
+In this context, the hyperparameters to estimate are the coefficient \(\beta_{gj}\) for gene \(g\) and sample \(i\), as well as the variance \(\sigma^2_g\) across genes \(g\).
+In the Bayesian approach, probabilities are updated after obtaining new data, referred to as conditional probabilities.
+For estimation, it is assumed that \(\frac{1}{\sigma^2_g} \sim \frac{1}{d_{0}s^2_{0}} \chi^2_{d_{0}}\), representing the prior information of the model.
+The probability for \(\beta_{gj} \neq 0\) is denoted as \(p_j\), where \(p_j\) represents the expected proportion of truly differentially expressed genes.
+The expected distribution of log-fold changes follows the distribution \(\beta_{gj} | \sigma^2_g , \beta_{gj} \neq 0 \sim N(0, v_{0j}\sigma^2_g)\).
+The posterior mean of \(\sigma^2_g\) given \(s^2_g\) is calculated as \(s^{-2}_g = E(\sigma^2_g | s^2_g) = \frac{d{0}s^2_{0} + d_{g}s^2_{g}}{d_{0} + d_{g}}\).
+This is where the shrinkage occurs, as the posteriors affect the prior values based on their respective sizes and degrees of freedom.
+The moderated t-statistic is defined as \(\tilde{t}_{gj} = \frac{\hat{\beta}_{gj}}{\hat{s}_{g}\sqrt{v_{gj}}}\).
+Once the values of \(\tilde{t}\) and \(s^{2}\) are calculated, the posterior odds are computed, providing the odds that a particular gene is differentially expressed.
+The odds for gene \(g\) being differentially expressed are denoted as \(O_{gj} = \frac{p(\beta_{gj} \neq 0 | \tilde{t}_{gj}, s^2{g})}{p(\beta_{gj} = 0 | \tilde{t}_{gj}, s^2{g})}\), where the numerator represents the probability that the gene is differentially expressed, and the denominator represents the probability that the gene is not differentially expressed \citep{Smyth2004}.
+
+Limma outputs a table with all proteins, their respective logfold change, p-values and adjusted p-values.
+The user can choose different methods for the correction of the p-values, which is necessary when testing the same hypothesis multiple times.
+A common method in omics studies is the false discovery rate (=fdr) by Benjamini Hochberg \citep{Benjamini1995}.
+By ranking all the p-values from smallest to largest the p-values are adjusted sequentially using the formula \(p_(i) = \frac{r_i}{m} Q\).
+The rank in the p-value table is denoted as \(r_i\) , \(m\) reflects the total number of tests and \(Q\) represents false discovery rate which is set to 5\%.
+The second option to choose is the Benjamini \& Yekutieli (=BY) \citep{Benjamini2001}method, which is a further development of the fdr by the same statistician and has a wider range of application based on the dependencies for the test.
+The third and most conservative option is the Holm method \citep{Holm1979} which controls the error sequentally in a family wise manner, which is considered as traditional in regards to the above mentioned experimental wise error rates.
+The above mentioned correction methods are ordered in the way from the most liberal to the most conservative one.
+The selection is supposed to give the user the possibility questioning results on multiple levels.
 
 \hypertarget{protein-set-enrichment-analysis}{%
-\subsubsection{Protein set enrichment
-analysis}\label{protein-set-enrichment-analysis}}
-
-After differential expression the upcoming question could be, in which
-context the up or down regulated proteins are. Enriched pathways can be
-understood as an over-representation (ORA = over-representation
-analysis) of proteins within a specific set. This set can be a pathway
-or a selection of common denominators in biological contexts, such as
-cell types or oncological factors. The statistical testing to identify
-the overabundant proteins of a specific set is similar to a \(\chi^2\)
-test developed by Karl Pearson \citep{Pearson1900} or the exact Fisher
-test \citep{Sprent2011} for tables with 4 fields in small sample sizes.
-Within the list are all truly differentially expressed genes, an
-annotation is the identification of a protein in the desired biological
-context. The table is expected to be populated by the values as follows:
+\subsubsection{Protein set enrichment analysis}\label{protein-set-enrichment-analysis}}
+
+After differential expression the upcoming question could be, in which context the up or down regulated proteins are.
+Enriched pathways can be understood as an over-representation (ORA = over-representation analysis) of proteins within a specific set.
+This set, also known as ontology can be a pathway or a selection of common denominators in biological contexts, such as cell types or oncological factors.
+The statistical testing to identify the overabundant proteins of a specific set is similar to a \(\chi^2\) test developed by Karl Pearson \citep{Pearson1900} or the exact Fisher test \citep{Sprent2011} for tables with 4 fields in small sample sizes.
+Within the list are all truly differential expressed genes, an annotation is the identification of a protein in the desired biological context.
+The table is expected to be populated by the values as follows:
 
@@
 
-Under the null hypothesis \(H_{0}\) states that the distribution and
-population of the table is random and there is no clear tendency with
-significant evidence. Against that the alternative hypothesis \(H_{A}\)
-says that there is an underlying tendency the table is populated. This
-leads to the test statistics, the observed values (=\(O\)) in the table
-are significantly different to the expected
-(=\(E=\frac{(A+B \lor C+D)}{N}\) ones. One approach is approximating
-\(\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}\) and deriving the
-p-value by the area under the density curve of the \(\chi^2\)
-distribution \(Q = \sum_{i=1}^{k} Z_{i}^{2}\). In the distribution \(k\)
-is defined by the degrees of freedom (=df), what can be obtained by
-\(k=(columns-1)*(rows-1)=df\) of the table. Another approach to obtain
-an exact p-value would be Fisher`s exact test, which states that the
-margins are distributed according to the hypergeometric distributions.
-The test statistic leads directly to the p-value with no need for
-approximation:
+Under the null hypothesis \(H_{0}\) states that the distribution and population of the table is random and there is no clear tendency with significant evidence.
+Against that the alternative hypothesis \(H_{A}\) says that there is an underlying tendency the table is populated.
+This leads to the test statistics, the observed values (=\(O\)) in the table are significantly different to the expected (=\(E=\frac{(A+B \lor C+D)}{N}\) ones.
+One approach is approximating \(\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}\) and deriving the p-value by the area under the density curve of the \(\chi^2\) distribution \(Q = \sum_{i=1}^{k} Z_{i}^{2}\).
+In the distribution \(k\) is defined by the degrees of freedom (=df), what can be obtained by \(k=(columns-1)*(rows-1)=df\) of the table.
+Another approach to obtain an exact p-value would be Fisher`s exact test, which states that the margins are distributed according to the hypergeometric distributions.
+The test statistic leads directly to the p-value with no need for approximation:
 \(p=1-\sum_{i=0}^{A-1}\frac{\binom{B}{i}\binom{D-B}{A+C-i}}{\binom{D}{A+C}}\)
-Here D denotes either for the background distribution which can be the
-total number of proteins obtained in the experiment, all proteins with
-annotation or a custom selected background. The background indicates all
-proteins which could be positive tested for in the observed sample. The
-above mentioned three options are implemented in the user interface.
-After obtaining the p-values for each protein set, the subsequent
-critical step involves the correction for multiple testing, considering
-the inherent risk of false positive results. Multiple testing correction
-methods adjust the significance thresholds to account for the increased
-probability of detecting false positives when performing multiple
-statistical tests. In this protein set enrichment analysis, the user is
-provided with a selection of established correction methods, akin to
-those commonly utilized in the field of differential expression
-analysis, ensuring robust and reliable interpretation of the results.
-Once the enriched protein pathways are determined, they can be visually
-represented using bar or dotplots, and a network of the found pathways.
-In the network ontologies are depicted as nodes, whereas edges show the
-strength of the interaction by thickness corresponding to the number of
-associated proteins.
+Here D denotes either for the background distribution which can be the total number of proteins obtained in the experiment, all proteins with annotation or a custom selected background.
+The background indicates all proteins which could be positive tested for in the observed sample.
+The above mentioned three options are implemented in the user interface.
+After obtaining the p-values for each protein set, the subsequent critical step involves the correction for multiple testing, considering the inherent risk of false positive results.
+Multiple testing correction methods adjust the significance thresholds to account for the increased probability of detecting false positives when performing multiple statistical tests.
+In this protein set enrichment analysis, the user is provided with a selection of established correction methods, akin to those commonly utilized in the field of differential expression analysis, ensuring robust and reliable interpretation of the results.
+Once the enriched protein pathways are determined, they can be visually represented using bar or dotplots, and a network of the found pathways.
 
 \hypertarget{pathway-enrichment-analysis}{%
-\paragraph{Pathway enrichment
-analysis}\label{pathway-enrichment-analysis}}
-
-The pathway enrichment analysis is performed using the R Bioconductor
-package clusterProfiler \citep{Wu2021}. The analysis utilizes the entire
-``Kyoto Encyclopedia of Genes and Genomes'' (KEGG) database, which is
-accessed through a function that maps the UniProt IDs to corresponding
-pathways.
+\paragraph{Pathway enrichment analysis}\label{pathway-enrichment-analysis}}
+
+The pathway enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}.
+The analysis utilizes the entire ``Kyoto Encyclopedia of Genes and Genomes'' (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways.
 
 \hypertarget{custom-ontology-based-protein-enrichment-analysis}{%
-\paragraph{Custom ontology based protein enrichment
-analysis}\label{custom-ontology-based-protein-enrichment-analysis}}
-
-For custom ontologies the protein enrichment analysis is done using the
-R bioconductor package piano \citep{Vaeremo2013}. In opposite to the
-pathway enrichment, piano maps proteins to a pre-selected gene
-collection set, which can be either downloaded from gsea-msigdb.org or
-custom made for the particular experiment and explains conditions,
-phenotype or other desired properties. For statistical testing the above
-explained Fisher exact test is used.
+\paragraph{Custom ontology based protein enrichment analysis}\label{custom-ontology-based-protein-enrichment-analysis}}
+
+For custom ontologies the protein enrichment analysis is done using the R bioconductor package piano \citep{Vaeremo2013}.
+In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
+The online database offers pre-built gene set collections for previous research questions in multiple fields of human biology.
+For statistical testing the above explained Fisher exact test is used.
+The output of the Fisher test does not show any directionality in regards of protein expression.
+If the user wants to include directionality in the network diagram, the default option of piano can be included in future versions of the program.
 
@@
 
-To begin using the application, the first step is to upload the
-tab-delimited evidence file and the sample annotation file. Once these
-files are uploaded, the application is ready to run with default values.
-The default settings are preconfigured for an initial analysis, and any
-further adjustments or fine-tuning can be made in subsequent steps of
-the analysis process.
+To begin using the application, the first step is to upload the tab-delimited evidence file and the sample annotation file after processing the data with MaxQuant.
+Once these files are uploaded, the application is ready to run with default values.
+The default settings are preconfigured for an initial analysis, and any further adjustments or fine-tuning can be made in subsequent steps of the analysis process.
+
+\hypertarget{pre-processing-and-quality-control}{%
+\subsection{Pre-processing and quality control}\label{pre-processing-and-quality-control}}
+
+The initial stages of the pipeline are focused on creating a reliable dataset for analysis that is free from noise and bias.
 
@@
 
-\caption{Summary showing the process of the analysis pipeline}\label{fig:ui_summary}
- \endfigure\egroup
-
-After a runtime of 2 minutes on the 4 core laptop the data is ready to
-visualize. On the left side of the user interface, users can find the
-analysis pipeline settings that can be adjusted. Once the desired
-changes are made, they can be applied by clicking the update button.
-This allows users to observe the direct impact of the applied changes on
-the dataset. In the summary visualization, it may appear cluttered due
-to the processing of 36 files. In smaller datasets, the nodes of the
-network graph can be more easily distinguished and understood. When
-running the application with SCoPE2 \citep{Specht2021} reference
-cutoffs, the following results were visualized.
+\caption{Summary showing the processing steps of the analysis pipeline}\label{fig:ui-summary}
+ \endfigure\egroup
+
+After a runtime of 2 minutes on the 4 core laptop the data is ready to visualize.
+On the left side of the user interface (see figure \ref{fig:ui-summary}), users can find the analysis pipeline settings that can be adjusted.
+Once the desired changes are made, they can be applied by clicking the update button.
+This allows users to observe the direct impact of the applied changes on the dataset.
+In the summary visualization, it may appear cluttered due to the processing of 36 files. In smaller data sets, the nodes of the network graph can be more easily distinguished and understood.
+When running the application with SCoPE2 \citep{Specht2021} reference cutoffs, the following results were visualized.
 
@@
 
-\caption{Barchart showing the number of peptide spectrum matches, peptides and proteins}\label{fig:ui_numbers_default}
- \endfigure\egroup
-
-The default settings of the analysis pipeline resulted in a total number
-of unique proteins below approximately \(10^3\), which were aggregated
-from around \(10^{3.5}\) peptides. Before these peptides were aggregated
-over approximately \(10^{4.5}\) peptide spectrum matches. The
-aggregation method used is fixed by median.
+\caption{Barchart showing the number of peptide spectrum matches, peptides and proteins}\label{fig:ui-numbers-default}
+ \endfigure\egroup
+
+In order to get an idea of the magnitude of the data set, the barplot provides an overview (see figure \ref{fig:ui-numbers-default}).
+The default settings of the analysis pipeline resulted in a total number of unique proteins below approximately 1,000, which were aggregated from around 3,000 peptides.
+Before these peptides were aggregated over approximately 32,000 peptide spectrum matches.
+The aggregation for both steps was performed by the column median.
 
@@
 
-\caption{Mutual information with default setting}\label{fig:ui_mutual_info_default}
- \endfigure\egroup
-
-After applying logarithmic transformation to the base 2 and performing
-column median row mean normalization, the mutual information (MI) within
-the healthy control (HC) group's sample type exhibits a reduction
-towards the lower edge of the interquartile range (IQR).
-
-MI between the HC group and the sample type corresponding to individuals
-with both HIV and metabolic syndrome, the IQR is further reduced and
-shifted towards the lower end compared to the MI values observed in the
-raw data.
+\caption{Mutual information with default setting}\label{fig:ui-mutual-info-default}
+ \endfigure\egroup
+
+One of the first steps of quality control is to assess the loss of information in the observed data set.
+The entropy based analysis gives insight in impact the computations had on the initial data set.
+When exploring the intra and inter groupwise comparison as a boxplot (see figure \ref{fig:ui-mutual-info-default}), the user can elaborate the loss in mutual information.
+
+After applying logarithmic transformation to the base 2 and performing column median row mean normalization, the mutual information (MI) within the healthy control (HC) group's sample type exhibits a reduction towards the lower edge of the interquartile range (IQR).
+
+MI between the HC group and the sample type corresponding to individuals with both HIV and metabolic syndrome, the IQR is further reduced and shifted towards the lower end compared to the MI values observed in the raw data.
+
+Therefore, it is advisable to experiment with various normalization techniques.
 
@@
 
-\caption{Popup showing the dependencies for the CONSTANd normalization}\label{fig:ui_constand_norm}
- \endfigure\egroup
+\caption{Popup showing the dependencies for the CONSTANd normalization}\label{fig:ui-constand-norm}
+ \endfigure\egroup
+
+Choosing the CONSTANd option for normalization a popup window appears to test for assumptions on the dataset.
+
+As seen in the dependency popup (see figure \ref{fig:ui-constand-norm}) for the CONSTANd normalization \citep{VanHoutven2021}, requirements are not fulfilled.
+The MA-plot shows a strong heteroskedasticity indicated by the red trendline.
 
@@
 
-\caption{Mutual information with quantile normalization}\label{fig:ui_mutual_info_quant_norm}
- \endfigure\egroup
-
-After performing a logarithmic transformation to the base 2 and quantile
-normalization, the mutual information (MI) within the healthy control
-(HC) group's sample type exhibits a minor reduction in the interquartile
-range (IQR). However, it should be noted that there is an increase in
-the number of outliers compared to the raw data.
-
-Observing MI IQR of the HC group with the sample type corresponding to
-individuals with both HIV and metabolic syndrome, the reduction in MI
-falls within the interquartile range of the raw data MI. Additionally,
-the median remains close to unaffected, indicating that the overall
-central tendency of the MI values is relatively preserved.
+\caption{Mutual information with quantile normalization}\label{fig:ui-mutual-info-quant-norm}
+ \endfigure\egroup
+
+When switching to the quantile normalization the loss of mutual information decreased.
+Therefore the advised normalization technique for this particular dataset would be the quantile normalization (see figure \ref{fig:ui-mutual-info-quant-norm} compared to figure \ref{fig:ui-mutual-info-default}).
+
+After performing a logarithmic transformation to the base 2 and quantile normalization, the mutual information (MI) within the healthy control (HC) group's sample type exhibits a minor reduction in the interquartile range (IQR).
+However, it should be noted that there is an increase in the number of outliers compared to the raw data.
+
+Observing MI IQR of the HC group with the sample type corresponding to individuals with both HIV and metabolic syndrome, the reduction in MI falls within the interquartile range of the raw data MI.
+Additionally, the median remains close to unaffected, indicating that the overall central tendency of the MI values is relatively preserved.
 
@@
 
-\caption{Reporter ion intensity with selection on channel. The selected factor will be highlighted and indicated in the legend.}\label{fig:ui_RI_channel_default}
- \endfigure\egroup
-
-When comparing the median reporter ion intensities across channels, no
-systematic bias is observed. However, it is worth noting that channel 7
-exhibits a wide range of values spanning an order of magnitude. This
-comparison is crucial since the three sample types consistently occupy
-the same channels in every run, making the visualization an important
-aspect of the analysis. Despite the wide range of values in channel 7,
-there is still overlap in the interquartile ranges (IQR) across the
-majority of samples.
+\caption{Reporter ion intensity with selection on channel. The selected factor will be highlighted and indicated in the legend.}\label{fig:ui-RI-channel-default}
+ \endfigure\egroup
+
+In order to check if the TMTpro correction in MaxQuant removed the signal spillage in other channels correctly, it is recommended to compare channel intensity (see figure \ref{fig:ui-RI-channel-default}).
+When comparing the median reporter ion intensities across channels, no systematic bias is observed.
+However, it is worth noting that channel 7 exhibits a wide range of values spanning an order of magnitude.
+This comparison is crucial since the three sample types consistently occupy the same channels in every run, making the visualization an important aspect of the analysis.
+Despite the wide range of values in channel 7, there is still overlap in the interquartile ranges (IQR) across the majority of samples.
 
@@
 
-\caption{Reporter ion intensity with selection on file}\label{fig:ui_RI_file_default}
- \endfigure\egroup
-
-Comparing the median reporter ion intensities with a focus on the batch
-(=Raw.file), it becomes evident that there are significant changes in
-signal that need to be taken into account in further analysis. After
-applying quality control measures, several values have been lost due to
-cutoffs in parameters such as parental ion fraction (PIF), q-value,
-coefficient of variation, and missingness of peptide data.
-
-For dimensionality reduction, the user has the option to choose the
-batch correction method ComBat \citep{Leek2012} in the side panel.
-However, this option is not set as the default choice. Instead, the
-conservative approach is followed, where the batch factor is included in
-the statistical module to account for any potential batch effects that
-may influence the results.
+\caption{Reporter ion intensity with selection on file}\label{fig:ui-RI-file-default}
+ \endfigure\egroup
+
+The reporter ion intensity highlights also differences on raw files, which is advisable to observe prior to any statistics or dimensionality reduction.
+
+Comparing the median reporter ion intensities with a focus on the batch (=Raw.file), it becomes evident that there are significant changes in signal that need to be taken into account in further analysis (see figure \ref{fig:ui-RI-file-default}).
+After applying quality control measures, several values have been lost due to cutoffs in parameters such as parental ion fraction (PIF), q-value, coefficient of variation, and missingness of peptide data.
+
+For dimensionality reduction, the user has the option to choose the batch correction method ComBat \citep{Leek2012} in the side panel.
+However, this option is not set as the default choice.
+Instead, the conservative approach is followed, where the batch factor is included in the statistical module to account for any potential batch effects that may influence the results.
 
@@
 
-\caption{Reporter ion intensity with selection on sampletype}\label{fig:ui_RI_st_default}
- \endfigure\egroup
-
-The sample type does not appear to have an impact on the median reporter
-ion intensities, as suggested by the overlapping interquartile ranges
-(IQRs). Nevertheless, the HIV group with metabolic syndrome displays a
-broader range of values spanning tenfold in arbitrary units.
+\caption{Reporter ion intensity with selection on sampletype}\label{fig:ui-RI-st-default}
+ \endfigure\egroup
+
+Comparing sample types in regards of reporter ion intensity is recommended to observe any experimental bias such as batch/channel confounding effects which would negatively impact the statistical model afterwards.
+
+The sample type does not appear to have an impact on the median reporter ion intensities, as suggested by the overlapping interquartile ranges (IQRs).
+Nevertheless, the HIV group with metabolic syndrome displays a broader range of values spanning tenfold in arbitrary units (see figure \ref{fig:ui-RI-st-default}).
 
@@
 
-\caption{Median covariance of razor proteins with selection on sampletype}\label{fig:ui_cv_st_default}
- \endfigure\egroup
-
-A cutoff value of 0.65 for the coefficient of variation (CV) of razor
-proteins as shown in the SCoPE2 publication \citep{Specht2021}
-demonstrates a satisfactory fit to the data. The minimum number of
-observations determines the threshold for considering a razor protein
-ambiguous, based on the number of matching peptides. In this case, the
-reference publication set the threshold at 5 observations.
-
-The interquartile ranges (IQRs) remained unchanged by this cutoff. The
-sample type did not influence the CV of razor proteins. Boxplots
-illustrate that the values are distributed similarly, as indicated by
-the overlapping IQRs.
+\caption{Median covariance of razor proteins with selection on sampletype}\label{fig:ui-cv-st-default}
+ \endfigure\egroup
+
+The distribution of median coefficient of variation (CV) shows the uniqueness in peptides for their respective razor protein.
+This metric represents the presence or absence of noisy matches in the data set.
+
+A cutoff value of 0.65 for the coefficient of variation (CV) of razor proteins as shown in the SCoPE2 publication \citep{Specht2021} demonstrates a satisfactory fit to the data.
+The minimum number of observations determines the threshold for considering a razor protein ambiguous, based on the number of matching peptides.
+In this case, the recommended minimum number of observation of 5 (by \citep{Specht2021}) was used.
+
+The interquartile ranges (IQRs) remained unchanged by this cutoff (see figure \ref{fig:ui-cv-st-default}).
+The sample type did not influence the CV of razor proteins.
+Boxplots illustrate that the values are distributed similarly, as indicated by the overlapping IQRs.
 
@@
 
-\caption{Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file}\label{fig:ui_cv_rawfile_default}
- \endfigure\egroup
-
-The plot illustrates that certain batches (=raw files) exhibit a higher
-number of distinct matches compared to others, as indicated by a thin
-line depicting low density. Most interquartile ranges (IQRs) overlap,
-suggesting similarity among the majority of batches. However, specific
-batches display a lower IQR in median CV across razor protein peptides,
-indicating distinct matches.
+\caption{Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file}\label{fig:ui-cv-rawfile-default}
+ \endfigure\egroup
+
+Razor peptide ambiguity over multiple batches can indicate experimental failures during sample prep (especially digestion) or during acquisition.
+
+The plot (see figure \ref@(fig:ui-cv-rawfile-default)) illustrates that certain batches (=raw files) exhibit a higher number of distinct matches compared to others, as indicated by a thin line depicting low density.
+Most interquartile ranges (IQRs) overlap, suggesting similarity among the majority of batches.
+However, specific batches display a lower IQR in median CV across razor protein peptides, indicating distinct matches.
 
@@
 
-\caption{Median coefficient of variation (CV) of leading razor proteins calculated for 2  observations per cell, with selection on Raw.file}\label{fig:ui_cv_rawfile_nobs2}
- \endfigure\egroup
-
-By setting the number of observations to 2, the distributions become
-more equalized. Additionally, batches that initially exhibited a
-decreased coefficient of variation (CV) tend to inflate towards the
-threshold.
+\caption{Median coefficient of variation (CV) of leading razor proteins calculated for 2  observations per cell, grouped by Raw.file}\label{fig:ui-cv-rawfile-nobs2}
+ \endfigure\egroup
+
+Experimenting with a lower minimum number of observations for the calculation of the median coefficient of variation, includes also less ambigous peptides.
+
+By setting the number of observations to 2, the distributions become more equalized (see figure \ref@(fig:ui-cv-rawfile-nobs2)).
+Additionally, batches that initially exhibited a decreased coefficient of variation (CV) tend to inflate towards the threshold.
 
@@
 
-\caption{Median coefficient of variation (CV) of leading razor proteins calculated for 10 observations per cell, with selection on Raw.file}\label{fig:ui_cv_rawfile_nobs10}
- \endfigure\egroup
-
-When the minimum number of observations for ambiguous peptides is set to
-10, the distribution of medians across channels shows a decrease in
-density. This can be attributed to the rare occurrence of 10 unique
-peptides matching to the same protein, which is not commonly observed in
-the dataset.
-
-\newpage
+\caption{Median coefficient of variation (CV) of leading razor proteins calculated for 10 observations per cell, with selection on Raw.file}\label{fig:ui-cv-rawfile-nobs10}
+ \endfigure\egroup
+
+In opposite increasing the minimum number of observation takes only highly ambiguous peptides into account and the coefficent of variation shows the interaction of these razor proteins.
+
+When the minimum number of observations for ambiguous peptides is set to 10, the distribution of medians across channels shows a decrease in density (see figure \ref@(fig:ui-cv-rawfile-nobs10)).
+This can be attributed to the rare occurrence of 10 unique peptides matching to the same protein, which is rarely observed in the dataset.
+
+\newpage
+
+\hypertarget{downstream-analysis-1}{%
+\subsection{Downstream analysis}\label{downstream-analysis-1}}
+
+The subsequent analysis reveals the initial biological effects, serving as an indicator of the effectiveness or ineffectiveness of the pre-processing and quality control steps.
+
 \bgroup  \origfigure[H]
@@
 
-\caption{Correlation heatmap}\label{fig:ui_heatmap}
- \endfigure\egroup
-
-The heatmap reveals the Pearson pairwise correlation coefficients
-between all proteins. The dendrogram indicates that there are two main
-protein clusters, with one of the clusters showing further
-sub-clustering, indicating a strong relationship among the proteins in
-the sub groups. However, the other cluster, consisting of individual
-proteins, shows weaker correlations, leading to a more intricate
-branching pattern in the dendrogram.
+\caption{Correlation heatmap}\label{fig:ui-heatmap}
+ \endfigure\egroup
+
+The heatmap (see figure \ref@(fig:ui-heatmap)) reveals the Pearson pairwise correlation coefficients between all proteins.
+The dendrogram indicates that there are two main protein clusters, with one of the clusters showing further sub-clustering, indicating a strong relationship among the proteins in the sub groups.
+However, the other cluster, consisting of individual proteins, shows weaker correlations, leading to a more intricate branching pattern in the dendrogram.
 
@@
 Top: principle component analysis (PCA).
-Lower: Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_default}
- \endfigure\egroup
-
-In both the Principal Component Analysis (PCA) plot and the Uniform
-Manifold Approximation and Projection (UMAP) plot, the second component
-is plotted against the first component. Surprisingly, neither
-visualization reveals any discernible clustering patterns based on
-sample type or batch. For the k-nearest neighbor descent algorithm
-(NN-descent) employed in the UMAP, first a default value of k=3 was
-selected.
+Lower: Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui-dim-red-default}
+ \endfigure\egroup
+
+In both the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP) plot (see figure \ref@(fig:ui-dim-red-default)), the second component is plotted against the first component.
+Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch.
+For the k-nearest neighbor descent algorithm (NN-descent) employed in the UMAP, first a default value of k=3 was selected.
 
@@
 Bottom: Uniform Manifold Approximation and Projection (UMAP).
-For the UMAP visualization, the k value for the NN-descent algorithm was modified to 15.}\label{fig:ui_dim_red_changed_k}
- \endfigure\egroup
-
-The second figure of the Principal Component Analysis (PCA) plot and the
-Uniform Manifold Approximation and Projection (UMAP), both displaying
-the second component against the first. Surprisingly, neither
-visualization reveals any discernible clustering patterns based on
-sample type or batch information. To enhance the resolution and capture
-more intricate patterns, the k-nearest neighbor descent algorithm
-(NN-descent) was adjusted, with a value of K set to 15. While the
-resulting plot exhibits improved detail, the underlying biological
-context or background remains elusive and does not emerge in the visual
-representation.
+For the UMAP visualization, the k value for the NN-descent algorithm was modified to 15.}\label{fig:ui-dim-red-changed-k}
+ \endfigure\egroup
+
+The second figure (\ref@(fig:ui-dim-red-changed-k) of the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP), both displaying the second component against the first.
+Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch information.
+To enhance the resolution and capture more intricate patterns, the k-nearest neighbor descent algorithm (NN-descent) was adjusted, with a value of K set to 15.
+While the resulting plot exhibits improved detail, the underlying biological context or background remains elusive and does not emerge in the visual representation.
 
@@
 Top: Principal Component Analysis (PCA).
-Bottom: Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_c3d}
- \endfigure\egroup
-
-In the 3D visualization, a subtle clustering pattern can be observed for
-the first two batches. Upon switching to the indicator sample type in
-the 3D visualization, no clustering can be observed whatsoever. This
-suggests that the variance in the data set is not attributed to sample
-type or batch, reinforcing the notion that other factors could
-contribute to the observed variability.
+Bottom: Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui-dim-red-c3d}
+ \endfigure\egroup
+
+In the 3D visualization (see figure \ref@(fig:ui-dim-red-c3d)), a subtle clustering pattern can be observed for the first two batches.
+Upon switching to the indicator sample type in the 3D visualization, no clustering can be observed whatsoever.
+This suggests that the variance in the data set is not attributed to sample type or batch, reinforcing the notion that other factors could contribute to the observed variability.
 
@@
 
-\caption{Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
-upper top: Principal Component Analysis (PCA).
-upper bottom: Uniform Manifold Approximation and Projection (UMAP).
-lower top: 3D Principal Component Analysis (PCA).
-lower bottom: 3D Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_batchC-1}
- \endfigure\egroup 
+\caption{Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).}\label{fig:ui-dim-red-batchC}
+ \endfigure\egroup
+
 \bgroup  \origfigure[H]
@@
 
-\caption{Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color).
-upper top: Principal Component Analysis (PCA).
-upper bottom: Uniform Manifold Approximation and Projection (UMAP).
-lower top: 3D Principal Component Analysis (PCA).
-lower bottom: 3D Uniform Manifold Approximation and Projection (UMAP)}\label{fig:ui_dim_red_batchC-2}
- \endfigure\egroup
-
-After applying batch correction using ComBat \citep{Leek2012}, neither
-the 2D nor the 3D visualization demonstrates any noticeable effects
-related to batches or sample types. The previously observed subtle
-clustering of the first two batches is now eliminated, indicating
-successful batch correction. However, it's important to note that the
-default pipeline was executed without batch correction for correct
-statistical testing.
+\caption{Batch corrected dimensionality reduction analysis showcasing sample types (represented by shape) and raw files (represented by color) in 3D.}\label{fig:ui-dim-red-batchC-3d}
+ \endfigure\egroup
+
+After applying batch correction using ComBat \citep{Leek2012}, neither the 2D (figure \ref@(fig:ui-dim-red-batchC)) nor the 3D (figure \ref@(fig:ui-dim-red-batchC-3d)) visualization demonstrates any noticeable effects related to batches or sample types.
+The previously observed subtle clustering of the first two batches is now eliminated, indicating successful batch correction.
+However, it's important to note that the default pipeline was executed without batch correction for correct statistical testing.
 
@@
 
-\caption{Feature wise output for the protein P61313 (=60S ribosomal protein L15) over the course of the pipeline.}\label{fig:ui_feature}
- \endfigure\egroup
-
-The scatter plots indicate the detected values associated with the
-protein UniProt ID P61313 (= 60S ribosomal protein L15). Starting from
-the peptide spectrum level and progressing to peptides aggregated for
-individual raw files, these plots offer a comprehensive view of the
-protein's peptide abundance across various batches. As indicated 14
-batches show no abundance of peptides derived from the protein P61313 at
-all. After merging the batches, the peptides appear to be abundant in
-all channels.
-
-\newpage
+\caption{Feature wise output for the protein P61313 (=60S ribosomal protein L15) over the course of the pipeline.}\label{fig:ui-feature}
+ \endfigure\egroup
+
+The feature-wise output allows for the observation of abundance in raw files or batches, providing a useful tool to assess the effectiveness of pre-processing and/or the work conducted in the wet lab.
+
+The scatter plots (see figure \ref@(fig:ui-feature)) indicate the detected values associated with the protein UniProt ID P61313 (= 60S ribosomal protein L15).
+Starting from the peptide spectrum level and progressing to peptides aggregated for individual raw files, these plots offer a comprehensive view of the protein's peptide abundance across various batches.
+As indicated 14 batches show no abundance of peptides derived from the protein P61313 at all.
+After merging the batches, the peptides appear to be abundant in all channels.
+
+\newpage
+
+\hypertarget{statistics-1}{%
+\subsection{Statistics}\label{statistics-1}}
+
 \bgroup  \origfigure[H]
@@
 
-\caption{Popup window for checking the dependency of the statistic module}\label{fig:ui_stat_dependency}
- \endfigure\egroup
-
-The t-statistics employed in limma are known for their robustness in
-handling non-normally distributed values. Nonetheless, it is recommended
-for users to assess the distribution of their data based on sample types
-before proceeding with statistical analysis. If the user is not
-satisfied with the results or identifies any distributional issues, the
-user can consider revisiting the transformation or normalization steps
-to address any concerns or improve the data distribution prior to
-statistical analysis.
-
-The values for all sample types exhibit a unimodal distribution.
-However, when examining the quantile-quantile (qq) plot, heavy tails can
-be observed for all three groups. This indicates deviations from a
-perfectly normal distribution, suggesting potential outliers in the
-data.
+\caption{Popup window for checking the dependency of the statistic module}\label{fig:ui-stat-dependency}
+ \endfigure\egroup
+
+The t-statistics employed in limma are known for their robustness in handling non-normally distributed values.
+Nonetheless, it is recommended for users to assess the distribution of their data based on sample types before proceeding with statistical analysis.
+If the user is not satisfied with the results or identifies any distributional issues, the user can consider revisiting the transformation or normalization steps to address any concerns or improve the data distribution prior to statistical analysis.
+
+The values for all sample types exhibit a unimodal distribution (see figure \ref@(fig:ui-stat-dependency)).
+However, when examining the quantile-quantile (qq) plot, heavy tails can be observed for all three groups.
+This indicates deviations from a perfectly normal distribution, suggesting potential outliers in the data.
 
@@
 
-\caption{Statistic module}\label{fig:ui_stat_HC_HIV_noMets}
- \endfigure\egroup
-
-\newpage
-
-By comparing the log10-transformed and quantile-normalized expression
-values between the HIV group without metabolic syndrome and the healthy
-control (HC) group, six differentially expressed proteins were
-identified. The Venndiagram indicates significantly up/downregulated
-proteins. The Volcano represents -log10(adjusted p values) against log2
-fold changes (logFC). These proteins include P68133 (Actin, alpha
-skeletal muscle), P47914 (60S ribosomal protein L29), P25311
-(Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear
-ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform
-type-2), and P46777 (60S ribosomal protein L5). A total of 652 proteins
-have been identified as not significantly upregulated or downregulated
-in the comparison analysis. These proteins did not exhibit statistically
-significant differences in expression between the compared groups or
-conditions.
-
-The direction of the contrast model determines the change in log-fold
-change (logFC) values. For instance, in the depicted figure, P68133
-(Actin, alpha skeletal muscle) shows an increased expression of 0.25
-(25\%) in HIV\_noMetS macrophages compared to HC macrophages. It's
-important to note that the logFC value changes by multiplication with -1
-when the direction of the contrast model is reversed.
-
-To account for potential batch effects, the batch factor was included as
-a cofactor in the model, ensuring that any observed differential
-expression is not confounded by batch variation.
+\caption{Statistics module}\label{fig:ui-stat-HC-HIV-noMets}
+ \endfigure\egroup
+
+\newpage
+
+By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group (see figure \ref@(fig:ui-stat-HC-HIV-noMets)), six differentially expressed proteins were identified.
+The Venn diagram indicates significantly up/down regulated proteins.
+The Volcano represents -log10(adjusted p values) against log2 fold changes (logFC).
+These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
+The remaining 652 proteins have been identified as not significantly up/down regulated in the comparison analysis.
+These proteins did not exhibit statistically significant differences in expression between the compared groups or conditions.
+
+The direction of the contrast model determines the change in log-fold change (logFC) values.
+For instance, in the depicted figure, P68133 (Actin, alpha skeletal muscle) shows an increased expression of 0.25 (25\%) in HIV\_noMetS macrophages compared to HC macrophages.
+It's important to note that the logFC value changes by multiplication with -1 when the direction of the contrast model is reversed.
+
+To account for potential batch effects, the batch factor was included as a cofactor in the model, ensuring that any observed differential expression is not confounded by batch variation.
 
@@
 
-\caption{Pathway enrichment. Networkplot indicating the association of differentially expressed proteins with pathways}\label{fig:ui_path_enrich_nw}
- \endfigure\egroup
-
-The pathway enrichment mapped and tested with the exact Fisher test
-\citep{Sprent2011} the differentially expressed proteins P46777 (60S
-ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the
-ribosome and coronavirus disease (COVID-19) nodes.
+\caption{Pathway enrichment. Networkplot indicating the association of differentially expressed proteins with pathways}\label{fig:ui-path-enrich-nw}
+ \endfigure\egroup
+
+The enrichment represents overabundant proteins for specific pathways. Results can be shown as a network or barplot.
+
+The pathway enrichment mapped and tested with the exact Fisher test \citep{Sprent2011} the differentially expressed proteins P46777 (60S ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the ribosome and coronavirus disease (COVID-19) nodes (see figure \ref@(fig:ui-path-enrich-nw)).
 
@@
 
-\caption{Pathway enrichment. Barplot indicating the interaction significance of differentially expressed proteins}\label{fig:ui_path_enrich_bar}
- \endfigure\egroup
-
-The barplot shows the significance of the two nodes. The count indicates
-how many proteins are associated with the node of the network. When
-hovering over the bar the user can read the p-value of the performed
-Fisher test. The background for the statistical test can be chosen as
-genome wide (default) or as the detected proteins. When choosing the
-detected proteins as a background no pathway could be considered as
-statistically significant.
+\caption{Pathway enrichment.
+Barplot indicating the number of differentially expressed proteins to a particular pathwaz.
+The color highlights the significance of the mapping, with exact p-values from the Fisher test after false discovery rate correction.}\label{fig:ui-path-enrich-bar}
+ \endfigure\egroup
+
+The barplot (see figure \ref@(fig:ui-path-enrich-bar)) shows the significance of the two nodes.
+The count indicates how many proteins are associated with the node of the network.
+When hovering over the bar the user can read the p-value of the performed Fisher test.
+The background for the statistical test can be chosen as genome wide (default) or as the set of detected proteins.
+When choosing the detected proteins as a background no pathway could be considered as statistically significant.
 
@@
 
-\caption{Gene ontology enrichment against the complete human proteome. Snapshot dotplot indicating the interaction significance of found proteins}\label{fig:ui_go_enrich_bar}
- \endfigure\egroup
-
-The position of the dot shows the number of matches for the particular
-protein set. Colors indicate for the significance level. In this
-analysis the entire repertoire of detected proteins was used and fetched
-against the bundle containing all Human gene sets in the Human Molecular
-Signatures Database (MSigDB). After a runtime of 3 minutes 672 of 24667
-gene sets were found significantly over represented. Only the protein
-set collections with the most protein matches were used (snipping tool
-Plotly).
+\caption{Gene ontology enrichment against the complete human proteome. Snapshot dotplot indicating the interaction significance of found proteins}\label{fig:ui-go-enrich-bar}
+ \endfigure\egroup
+
+Ontology enrichment offers a more comprehensive mapping, providing a broader understanding of the data.
+By allowing users to select a custom protein set collection, they can either narrow down the potential results for a targeted approach or expand their scope for an untargeted analysis.
+
+The position of the dot (see figure \ref@(fig:ui-go-enrich-bar)) shows the number of matches for the particular protein set.
+Colors indicate for the significance level.
+In this analysis the entire repertoire of detected proteins was used and fetched against the bundle containing all Human gene sets in the Human Molecular Signatures Database (MSigDB).
+After a runtime of 3 minutes 672 of 24667 gene sets were found significantly over represented.
+Only the protein set collections with the most protein matches were used (snipping tool Plotly).
 
@@
 
-The Proteomics Workbench interface and the ProteoScanR pipeline
-demonstrated how interactive engagement with the data not only enhances
-the experience of biologists but also improves the comprehension of the
-underlying significance of a biological dataset. By utilizing an
-entropy-based visualization approach, the conservation of information
-can be validated, allowing users to select appropriate methods and
-adjust thresholds, cutoffs, and techniques accordingly. With the
-selection of factors to indicate over multiple plots, biases can be
-identified on various levels. Additionally, the analysis can be examined
-on an individual protein basis, enabling the identification of
-suspicious results. The statistics module provides users with a simple
-tool to assess and visualize the results, offering clarity to dense
-scatter plots through hover functions. Enrichment mapping links
-statistically significant proteins to either the KEGG pathway database
-or individual research data sets, depending on user selection. These
-final results place the data in a biological context, which can be
-explored interactively in graphical visualizations.
+The Proteomics Workbench interface and the ProteoScanR pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
+By utilizing an entropy-based visualization approach, the conservation of information can be validated, allowing users to select appropriate methods and adjust thresholds, cutoffs, and techniques accordingly.
+With the selection of factors to indicate over multiple plots, biases can be identified on various levels.
+Additionally, the analysis can be examined on an individual protein basis, enabling the identification of suspicious results and closer validation of significant candidates.
+The statistics module provides users with a simple tool to assess and visualize the results, offering clarity to dense scatter plots through hover functions.
+Enrichment mapping links statistically significant proteins to either the KEGG pathway database or individual research data sets, depending on user selection.
+These final results place the data in a biological context, which can be explored interactively in graphical visualizations.
 
@@
 
-Linear programming scripts are widely utilized for data analysis, and
-R-shiny provides an efficient platform for implementing these scripts in
-applications. This approach not only benefits scientists who are not
-proficient in programming, but it is also convenient for programmers.
-The Proteomics Workbench user interface offers a visually appealing
-design, and the interactive menus enhance the user experience by
-eliminating the need to search for specific lines of code on a black
-screen. Modifying parameters of functions becomes effortless, and data
-can be visualized without rerunning the entire script. R-shiny executes
-functions on demand, which means that graphics, variables, and data
-frames are not recalculated if their dependencies remain unchanged. This
-feature makes the application both cost-effective and time-efficient.
-The analysis pathway is visualized through multiple components,
-providing users with a comprehensive view of the data. The overview
-plot, generated using the Q-feature object-oriented programming in the
-scp R library \citep{Vanderaa2021}, serves as a visual representation of
-the analysis process. This plot allows users to track the progression of
-the analysis and identify any missing files or samples.
+Linear programming scripts are widely utilized for data analysis, and R-shiny provides an efficient platform for implementing these scripts in applications.
+This approach not only benefits scientists who are not proficient in programming, but it is also convenient for programmers.
+The Proteomics Workbench user interface offers a visually appealing design, and the interactive menus enhance the user experience by eliminating the need to search for specific lines of code on a black screen.
+Modifying parameters of functions becomes effortless, and data can be visualized without rerunning the entire script.
+R-shiny executes functions on demand, which means that graphics, variables, and data frames are not recalculated if their dependencies remain unchanged.
+This feature makes the application both cost-effective and time-efficient.
+The analysis pathway is visualized through multiple components, providing users with a comprehensive view of the data.
+The overview plot, generated using the Q-feature object-oriented programming in the scp R library \citep{Vanderaa2021}, serves as a visual representation of the analysis process.
+This plot allows users to track the progression of the analysis and identify any missing files or samples.
 
@@
 
-Information theory and the concept of mutual information provide a
-framework for understanding the conservation of entropy in a dataset.
-Traditionally, the validation of data transformation and normalization
-methods has been limited to observing distributions through histograms
-or Quantile-Quantile plots. However, the developed approach goes beyond
-these conventional methods, providing a deeper context for the variation
-in the data. By allowing users to choose the method that preserves the
-maximum amount of information, the approach enables researchers to
-extract the most valuable insights from their experiments. Normalization
-methods play a crucial role in data analysis, and their selection
-depends on the nature of the data and the specific analysis objectives.
-The mutual information analysis revealed that quantile normalization
-exhibited the smallest loss of entropy compared to other normalization
-methods
+Information theory and the concept of mutual information provide a framework for understanding the conservation of entropy in a dataset.
+Traditionally, the validation of data transformation and normalization methods has been limited to observing distributions through histograms or Quantile-Quantile plots.
+However, the developed approach goes beyond these conventional methods, providing a deeper context for the variation in the data.
+By allowing users to choose the method that preserves the maximum amount of information, the approach enables researchers to extract the most valuable insights from their experiments.
+Normalization methods play a crucial role in data analysis, and their selection depends on the nature of the data and the specific analysis objectives.
+The mutual information analysis revealed that quantile normalization exhibited the smallest loss of entropy compared to other normalization methods
 
@@
 
-MS2 reporter ion intensities can be effectively visualized using a
-boxplot, where the medians represent the intensity distribution across
-various factors such as sample type, batch, or channel. This
-visualization aids users in assessing the variation and potential biases
-in the dataset. During the example analysis, the reporter ion intensity
-plot revealed inconsistencies in the dataset based on signal intensities
-associated with different batches. It was observed that certain runs
-exhibited a wide spread of signal intensities, which could indicate
-technical issues or batch-related variations. These inconsistencies were
-partly filtered out during the quality control process to ensure the
-reliability and accuracy of the data.
+MS2 reporter ion intensities can be effectively visualized using a boxplot, where the medians represent the intensity distribution across various factors such as sample type, batch, or channel.
+This visualization aids users in assessing the variation and potential biases in the dataset.
+During the example analysis, the reporter ion intensity plot revealed inconsistencies in the dataset based on signal intensities associated with different batches.
+It was observed that certain runs exhibited a wide spread of signal intensities, which could indicate technical issues or batch-related variations.
+These inconsistencies were partly filtered out during the quality control process to ensure the reliability and accuracy of the data.
 
 \hypertarget{median-coefficient-of-variation}{%
-\subsection{Median coefficient of
-variation}\label{median-coefficient-of-variation}}
-
-The median coefficient of variation (CV) was calculated for a razor
-protein in relation to its ambiguous peptides. This statistical
-parameter measures the variability of the razor protein's abundance
-across different peptide measurements. The results were then visualized
-using a boxplot, allowing users to assess the distribution of CV values
-based on a selected factor. During testing and fine-tuning, the number
-of observations for ambiguously matched peptides was adjusted to
-evaluate its impact on the CV. By varying the number of observations,
-researchers could determine the optimal parameter that yielded
-meaningful CV values. In this context, the reference values from the
-SCoPE2 publication \citep{Specht2021} were employed as benchmarks and
-subjected to validation. The results showed that using a number of
-observations of 5 for leading razor peptides was suitable for
-identifying a cutoff value of 0.65. This cutoff value was determined by
-observing the whiskers of the boxplots. By setting this threshold, noisy
-peptides in the dataset could be effectively reduced, enhancing the
-quality and reliability of the data.
+\subsection{Median coefficient of variation}\label{median-coefficient-of-variation}}
+
+The median coefficient of variation (CV) was calculated for a razor protein in relation to its ambiguous peptides.
+This statistical parameter measures the variability of the razor protein's abundance across different peptide measurements.
+The results were then visualized using a boxplot, allowing users to assess the distribution of CV values based on a selected factor.
+During testing and fine-tuning, the number of observations for ambiguously matched peptides was adjusted to evaluate its impact on the CV.
+By varying the number of observations, researchers could determine the optimal parameter that yielded meaningful CV values.
+In this context, the reference values from the SCoPE2 publication \citep{Specht2021} were employed as benchmarks and subjected to validation.
+The results showed that using a number of observations of 5 for leading razor peptides was suitable for identifying a cutoff value of 0.65.
+This cutoff value was determined by observing the whiskers of the boxplots.
+By setting this threshold, noisy peptides in the dataset could be effectively reduced, enhancing the quality and reliability of the data.
 
@@
 
-The feature-wise output plays a crucial role in enhancing the analysis
-pathway by offering users the ability to select specific proteins and
-explore their associated peptide spectra, peptide values, and protein
-values throughout the analysis. This feature provides a detailed and
-comprehensive view of the analysis results, enabling users to identify
-any missing or erroneous data points, particularly in channels with low
-or high abundance. By examining these individual protein profiles, users
-can assess the quality of the wet lab work and detect any potential
-errors or inconsistencies. In the case of observing the abundance of the
-protein P61313 (60S ribosomal protein L15) over the entire course of the
-experiment, it becomes apparent that there are missing values within
-certain runs and channels. Specifically, 14 runs exhibit no signal for
-this protein. This observation underscores the significance of
-aggregating data at the peptide and protein levels and combining
-individual runs to construct the final protein expression set. The
-scattering of dots in the plot indicates the inflation of small values
-and the deflation of large values after transformation. The scatter plot
-of quantile normalized values demonstrates the processing steps taken
-while preserving outliers and maintaining the biological relevance of
-the data. This processing steps aid in achieving a more balanced
-distribution of data and avoids the distortion of biological meaning.
+The feature-wise output plays a crucial role in enhancing the analysis workflow by offering users the ability to select specific proteins and explore their associated peptide spectra, peptide values, and protein values throughout the analysis.
+This feature provides a detailed and comprehensive view of the analysis results, enabling users to identify any missing or erroneous data points, particularly in channels with low or high abundance.
+By examining these individual protein profiles, users can assess the quality of the wet lab work and detect any potential errors or inconsistencies.
+In the case of observing the abundance of the protein P61313 (60S ribosomal protein L15) over the entire course of the experiment, it becomes apparent that there are missing values within certain runs and channels.
+Specifically, 14 runs exhibit no signal for this protein.
+This observation underscores the significance of aggregating data at the peptide and protein levels and combining individual runs to construct the final protein expression set.
+The scattering of dots in the plot indicates the inflation of small values and the deflation of large values after transformation.
+The scatter plot of quantile normalized values demonstrates the processing steps taken while preserving outliers and maintaining the biological relevance of the data.
+This processing steps aid in achieving a more balanced distribution of data and avoids the distortion of biological meaning.
 
@@
 
-The statistics module within the application provides users with a
-user-friendly interface for hypothesis testing, even without prior
-expertise in statistics. By leveraging the powerful R-package limma
-\citep{Ritchie2015}, the module offers robust statistical analysis
-methods, empowering users to perform tests and obtain results to
-validate their hypotheses. One key feature of the statistics module is
-its ability to handle multiple factors and therefore correct for batch
-effects. Users can extend the statistical analysis across various
-factors, allowing them to quantify the impact of different variables and
-identify differentially expressed proteins. To enhance the user
-experience, the visualization of statistical results has been improved,
-resulting in visually appealing and interactive displays. These
-visualizations enable users to explore and interpret the results
-effortlessly, aiding in the comprehension of the statistical findings.
-By providing intuitive and engaging visual representations, the module
-facilitates the understanding and communication of complex statistical
-analyses. When comparing the group HIV without metabolic syndrome and
-the healthy control 6 proteins could be found as differentially
-expressed after correcting the p-values by false discovery rate (FDR
-\citep{Benjamini1995}). For example, the protein P68133 (Actin, alpha
-skeletal muscle) exhibits a log fold change (logFC) of -0.25, indicating
-a significant decrease in actin levels within the HIV group. This
-finding aligns with previous publications suggesting that HIV exploits
-cellular skeleton proteins and alters their expression levels
-\citep{Turville2018}. By inducing orchestrated structural changes in
-compartments composed of actin filaments, the virus facilitates its
-spread and infects other lymphocytes \citep{Rodrigues2022}. Another
-protein of interest, P47914 (60S ribosomal protein L29), demonstrates
-the impact of viral infection on protein synthesis. Differential
-expression of this protein further highlights the perturbations in
-cellular processes caused by the virus.
-
-\hypertarget{enrichment}{%
-\subsection{Enrichment}\label{enrichment}}
-
-The enrichment analysis module within the application addresses the
-challenge of selecting protein sets for mapping by incorporating a
-database-fetching functionality. This feature streamlines the enrichment
-analysis process and eliminates the need for users to manually select
-sets to map proteins against. The first variant of the enrichment
-analysis module utilizes the KEGG database, which offers a comprehensive
-collection of pathways and their interactions. By accessing this
-database, users can gain valuable insights into the underlying pathways
-and their relationships. This automated mapping against the KEGG
-database provides a convenient and efficient way to perform enrichment
-analysis. The second variant of the module caters to sophisticated users
-who may have specific protein sets in mind based on their research
-question. This flexibility allows users to select and incorporate their
-own protein sets into the analysis, enabling research in multiple
-contexts and providing a comprehensive insight into the dataset. Both
-variants of the enrichment analysis module present the results in
-interactive plots, offering users a dynamic and immersive experience
-with the data. Users can explore the enriched pathways and interact with
-the visualizations, gaining a deeper understanding of the relationships
-and connections within the dataset. Additionally, if desired, users can
-generate static plots and download them for use in presentations or
-other offline contexts. As indicated by the statistical findings, the
-KEGG enrichment analysis identified proteins associated with the
-ribosome and COVID-19. Both protein sets were connected to two ribosomal
-proteins, namely P47914 (60S ribosomal protein L29) and P46777 (60S
-ribosomal protein L5). These findings align with the fact that HIV and
-COVID-19, like all known viruses, depend entirely on the host cell's
-translation machinery for protein synthesis. This includes utilizing the
-host's ribosomes, tRNAs, amino acids, and all necessary factors for
-protein synthesis initiation, elongation, and termination
-\citep{Ohlmann2014}. Consequently, it is not surprising that these
-ribosomal proteins were found to be abundant in HIV-infected
-individuals.
+The statistics module within the application provides users with a user-friendly interface for hypothesis testing, even without prior expertise in statistics.
+By leveraging the powerful R-package limma \citep{Ritchie2015}, the module offers robust statistical analysis methods, empowering users to perform tests and obtain results to validate their hypotheses.
+One key feature of the statistics module is its ability to handle multiple factors and therefore correct for batch effects.
+Users can extend the statistical analysis across various factors, allowing them to quantify the impact of different variables and identify differentially expressed proteins.
+To enhance the user experience, the visualization of statistical results has been improved, resulting in visually appealing and interactive displays.
+These visualizations enable users to explore and interpret the results effortlessly, aiding in the comprehension of the statistical findings.
+By providing intuitive and engaging visual representations, the module facilitates the understanding and communication of complex statistical analyses.
+When comparing the group HIV without metabolic syndrome and the healthy control 6 proteins could be found as differentially expressed after correcting the p-values by false discovery rate (FDR \citep{Benjamini1995}).
+For example, the protein P68133 (Actin, alpha skeletal muscle) exhibits a log fold change (logFC) of -0.25, indicating a significant decrease in actin levels within the HIV group.
+This finding aligns with previous publications suggesting that HIV exploits cellular skeleton proteins and alters their expression levels \citep{Turville2018}.
+By inducing orchestrated structural changes in compartments composed of actin filaments, the virus facilitates its spread and infects other lymphocytes \citep{Rodrigues2022}.
+Another protein of interest, P47914 (60S ribosomal protein L29), demonstrates the impact of viral infection on protein synthesis.
+Differential expression of this protein further highlights the perturbations in cellular processes caused by the virus.
+
+\hypertarget{enrichment-analysis}{%
+\subsection{Enrichment analysis}\label{enrichment-analysis}}
+
+The enrichment analysis module within the application addresses the challenge of selecting protein sets for mapping by incorporating a database-fetching functionality.
+This feature streamlines the enrichment analysis process and eliminates the need for users to manually select sets to map proteins against.
+The first variant of the enrichment analysis module utilizes the KEGG database, which offers a comprehensive collection of pathways and their interactions.
+By accessing this database, users can gain valuable insights into the underlying pathways and their relationships.
+This automated mapping against the KEGG database provides a convenient and efficient way to perform enrichment analysis.
+The second variant of the module caters to sophisticated users who may have specific protein sets in mind based on their research question.
+This flexibility allows users to select and incorporate their own protein sets into the analysis, enabling research in multiple contexts and providing a comprehensive insight into the dataset.
+Both variants of the enrichment analysis module present the results in interactive plots, offering users a dynamic and immersive experience with the data.
+Users can explore the enriched pathways and interact with the visualizations, gaining a deeper understanding of the relationships and connections within the dataset.
+Additionally, if desired, users can generate static plots and download them for use in presentations or other offline contexts.
+As indicated by the statistical findings, the KEGG enrichment analysis identified proteins associated with the ribosome and COVID-19.
+Both protein sets were connected to two ribosomal proteins, namely P47914 (60S ribosomal protein L29) and P46777 (60S ribosomal protein L5).
+These findings align with the fact that HIV and COVID-19, like all known viruses, depend entirely on the host cell's translation machinery for protein synthesis.
+This includes utilizing the host's ribosomes, tRNAs, amino acids, and all necessary factors for protein synthesis initiation, elongation, and termination \citep{Ohlmann2014}.
+Consequently, it is not surprising that these ribosomal proteins were found to be abundant in HIV-infected individuals.
 
@@
 
-Unfortunately, a substantial number of individual proteins in the data
-set (totaling 658) have missing values. On average, this amounts to
-approximately 53\% of the proteins across all runs and channels. This
-situation leads to the need for missing value imputation using
-techniques such as k-nearest neighbor (K-NN), which can result in
-averaged results. Additionally, it is important to note that this
-imputation process can introduce potential loss of power in subsequent
-statistical analysis. To improve future experiments, it is recommended
-to increase the number of proteins in the carrier proteome. By expanding
-the repertoire of proteins included in the carrier proteome, it can
-enhance the coverage and depth of analysis, providing a more
-comprehensive understanding of the biological system under
-investigation. This can potentially lead to more robust and accurate
-results in downstream analyses.
+Unfortunately, a substantial number of individual proteins in the data set (totaling 651) have missing values.
+On average, this amounts to approximately 53\% of the proteins across all runs and channels (see figure \ref@(fig:missingness)).
+This situation leads to the need for missing value imputation using techniques such as k-nearest neighbor (K-NN), which can sometimes lead to artificially evenly distributed values.
+Additionally, it is important to note that this imputation process can introduce potential loss of power in subsequent statistical analysis.
+To improve future experiments, it is recommended to increase the number of proteins in the carrier proteome.
+By expanding the repertoire of proteins included in the carrier proteome, it can enhance the coverage and depth of analysis, providing a more comprehensive understanding of the biological system under investigation.
+This can potentially lead to more robust and accurate results in downstream analyses.
 
@@
 \hypertarget{clustering-for-stratification-of-cells.}{%
-\subsubsection{Clustering for stratification of
-cells.}\label{clustering-for-stratification-of-cells.}}
-
-The advantage of performing single-cell analysis lies in its ability to
-provide a fine resolution of cellular differences. Unfortunately
-variation causing effects such as cellular differentiation stages, are
-mainly unknown when starting single-cell experiments. The next steps in
-the development of the program aim to enable the stratification of
-single-cell data sets. The analysis typically begins by finding clusters
-within the datasets, identifying differentially expressed proteins, and
-annotating cell types based on their expression profiles. Subsequently,
-the analysis aims to distinguish sample types based on their expression
-profiles.
-
-Here I will explain how the analysis could start and elaborate the
-limitations and challenges for this approach. Additional processing
-steps will be included after the existing pipeline, so ProteoScanR
-provides a starting point for further analysis.
-
-\bgroup  \origfigure[H]
-
-{\centering \includegraphics[width=1\linewidth]{Thesis_files/figure-latex/cluster_analysis_dendrogram-1}
-
-}
-
-\caption{Dendrogram showing the similarity over sample types. Tree was created using Ward`s minimum variance method.}\label{fig:cluster_analysis_dendrogram}
- \endfigure\egroup
-
-As observed in the dendrogram, it is evident that sample type alone does
-not account for the variation observed in the data. The branching
-patterns in the dendrogram suggest that other factors might be
-significant for contributing to the averaging of protein expression
-levels. This can potentially lead to a decrease in the statistical power
-of tests aimed at detecting differential expression between sample
-types. Ideally, clustering approaches should result in individual
-clusters that capture distinct sample types, enabling the testing of
-differential expression within each cluster. This stratification of
-samples within clusters allows for the identification of unique
-expression patterns associated with specific biological variations. The
-dendrogram, as demonstrated above, is an example of hierarchical
-clustering, which provides a visual representation of the clustering
-results. In this context, users should have the flexibility to select
-the desired number of clusters that best explain the underlying
-biological variation in their dataset. This allows for a more targeted
-and precise analysis of differential expression within each cluster,
-facilitating the identification of biologically meaningful patterns and
-insights. To begin with the k-means clustering approach, it is crucial
-to determine the optimal number of clusters before conducting the
-analysis. This can be achieved using various approaches, one of which is
-the gap statistic method \citep{Tibshirani2001}.
-
-\bgroup  \origfigure[H]
-
-{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster_analysis_gap_stat-1}
-
-}
-
-\caption{Optimal numbers of clusters calculated with gap statistics method}\label{fig:cluster_analysis_gap_stat}
- \endfigure\egroup
-
-The gap statistic compares the total intracluster variation across
-various values of k (number of clusters) to the expected variation under
-a null reference distribution. The null reference distribution
-represents a scenario where no clear clustering patterns exist in the
-data. However as seen in the gap statistic a high number of clusters
-will be difficult to distinguish from each other and explore the
-biological meaning such as differentiation in the example macrophage
-dataset. Another approach used to determine the optimal number of
-clusters is the silhouette method.
-
-\bgroup  \origfigure[H]
-
-{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster_analysis_silh-1}
-
-}
-
-\caption{Optimal number of clusters calculated with silhoutte method.}\label{fig:cluster_analysis_silh}
- \endfigure\egroup
-
-The silhouette method calculates silhouette coefficients for each data
-point, which quantify how closely the point resembles its own cluster
-compared to other clusters. If the overall distances between clusters
-are small, indicating overlapping or unclear separation, it is suggested
-to take the minimum k. This suggests that the data points may not be
-well-separated into distinct clusters. As suggested by the silhouette
-method in this case a lower number of clusters to differentiate cell
-types is recommended.
-
-\bgroup  \origfigure[H]
-
-{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster_analysis_dim_red-1}
-
-}
-
-\caption{PCA plot with cluster indication. Shape of the datapoints indicate sample types. Clusters are encircled and color coded.}\label{fig:cluster_analysis_dim_red}
- \endfigure\egroup
-
-As seen in the principal components analysis, clusters are densely
-packed, and the plot shows no clear differentiation between the 8
-suggested clusters. The dense packing and poor resolution indicates that
-other factors may be influencing the lack of distinct separation. One of
-the other factors involves the large number of imputed values due the
-high rate of missing values in the protein expression set. K-NN is not
-able to impute biological differences and therefore averages the
-particular missing value by its k neighbors.
-
-This suggests the utilization of a different data set for developing the
-clustering module of the future program. By employing a data set with a
-lower rate of missing values in the protein expression set, the accuracy
-and reliability of the clustering analysis can be enhanced. Having a
-data set with less missing data will allow for a more precise
-identification of clusters and differentiation between cell types.
+\subsubsection{Clustering for stratification of cells.}\label{clustering-for-stratification-of-cells.}}
+
+The advantage of performing single-cell analysis lies in its ability to provide a fine resolution of cellular differences.
+Unfortunately variation causing effects such as cellular differentiation stages, are mainly unknown when starting single-cell experiments.
+The next steps in the development of the program aim to enable the stratification of single-cell data sets.
+The analysis typically begins by finding clusters within the datasets, identifying differentially expressed proteins, and annotating cell types based on their expression profiles.
+Subsequently, the analysis aims to distinguish sample types based on their expression profiles.
+
+Here I will explain how the analysis could start and elaborate the limitations and challenges for this approach.
+Additional processing steps will be included after the existing pipeline, so the ProteoScanR provides a starting point for further analysis.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=1\linewidth]{Thesis_files/figure-latex/cluster-analysis-dendrogram-1}
+
+}
+
+\caption{Dendrogram showing the similarity over sample types. Tree was created using Ward`s minimum variance method.}\label{fig:cluster-analysis-dendrogram}
+ \endfigure\egroup
+
+As observed in the dendrogram (see figure \ref@(fig:cluster-analysis-dendrogram)), it is evident that sample type alone does not account for the variation observed in the data.
+The branching patterns in the dendrogram suggest that other factors might be significant for contributing to the averaging of protein expression levels.
+This can potentially lead to a decrease in the statistical power of tests aimed at detecting differential expression between sample types.
+Ideally, clustering approaches should result in individual clusters that capture distinct sample types, enabling the testing of differential expression within each cluster.
+This stratification of samples within clusters allows for the identification of unique expression patterns associated with specific biological variations.
+The dendrogram, as demonstrated above, is an example of hierarchical clustering, which provides a visual representation of the clustering results.
+In this context, users should have the flexibility to select the desired number of clusters that best explain the underlying biological variation in their dataset.
+This allows for a more targeted and precise analysis of differential expression within each cluster, facilitating the identification of biologically meaningful patterns and insights.
+To begin with the k-means clustering approach, it is crucial to determine the optimal number of clusters before conducting the analysis.
+This can be achieved using various approaches, one of which is the gap statistic method \citep{Tibshirani2001}.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster-analysis-gap-stat-1}
+
+}
+
+\caption{Optimal numbers of clusters calculated with gap statistics method}\label{fig:cluster-analysis-gap-stat}
+ \endfigure\egroup
+
+The gap statistic (see figure \ref@(fig:cluster-analysis-gap-stat)) compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution.
+The null reference distribution represents a scenario where no clear clustering patterns exist in the data.
+However as seen in the gap statistic a high number of clusters will be difficult to distinguish from each other and explore the biological meaning such as differentiation in the example macrophage dataset.
+Another approach used to determine the optimal number of clusters is the silhouette method.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster-analysis-silh-1}
+
+}
+
+\caption{Optimal number of clusters calculated with silhoutte method.}\label{fig:cluster-analysis-silh}
+ \endfigure\egroup
+
+The silhouette method (see figure \ref@(fig:cluster-analysis-silh)) calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters.
+If the overall distances between clusters are small, indicating overlapping or unclear separation, it is suggested to take the minimum k.
+This suggests that the data points may not be well-separated into distinct clusters.
+As suggested by the silhouette method in this case a lower number of clusters to differentiate cell types is recommended.
+
+\bgroup  \origfigure[H]
+
+{\centering \includegraphics[width=0.9\linewidth]{Thesis_files/figure-latex/cluster-analysis-dim-red-1}
+
+}
+
+\caption{PCA plot with cluster indication. Shape of the datapoints indicate sample types. Clusters are encircled and color coded.}\label{fig:cluster-analysis-dim-red}
+ \endfigure\egroup
+
+As seen in the principal components analysis (see figure \ref@(fig:cluster-analysis-dim-red)), clusters are densely packed, and the plot shows no clear differentiation between the 8 suggested clusters.
+The dense packing and poor resolution indicates that other factors may be influencing the lack of distinct separation.
+One of the other factors involves the large number of imputed values due the high rate of missing values in the protein expression set.
+K-NN is not able to impute biological differences and therefore averages the particular missing value by its k neighbors.
+
+This suggests the utilization of a different data set for developing the clustering module of the future program.
+By employing a data set with a lower rate of missing values in the protein expression set, the accuracy and reliability of the clustering analysis can be enhanced.
+Having a data set with less missing data will allow for a more precise identification of clusters and differentiation between cell types.
 
@@
 
-One potential target for future research could involve leveraging
-artificial intelligence techniques to predict cellular behavior, cell
-types, or even diseases. By running the application on a server and
-creating a database, it would be possible to harness the power of AI
-algorithms such as decision trees, random forests, or support vector
-machines. Applying artificial intelligence to biological data has the
-potential to enhance the classification of cell types, similar to the
-earlier discussed clustering approach. However, it is important to note
-that training and validating these algorithms typically requires
-supervision and can be a time-consuming task.
-
-Another area of focus, not limited to research topics, could be to
-enhance the usability of the application. While a development time of 5
-months for the pipeline and user interface is sufficient to meet the
-analytical needs, there is always room for improvement in terms of
-user-friendliness. It is essential to consider the needs of all users,
-including the least experienced user. To prevent program crashes and
-errors, it may be necessary to implement try-catch around code blocks,
-as well as validate-need statements for the validation of user input.
-Additionally, creating a user-friendly environment involves establishing
-effective channels of communication between users and developers, such
-as utilizing platforms like GitHub for collaboration and feedback.
+One potential target for future research could also involve leveraging artificial intelligence techniques to predict cellular behavior, cell types, or even diseases.
+By running the application on a server and creating a database, it would be possible to harness the power of AI algorithms such as decision trees, random forests, or support vector machines.
+Applying artificial intelligence to biological data has the potential to enhance the classification of cell types, similar to the earlier discussed clustering approach.
+However, it is important to note that training and validating these algorithms typically requires supervision and can be a time-consuming task.
+
+Another area of focus, not limited to research topics, could be to enhance the usability of the application.
+While a development time of 5 months for the pipeline and user interface is sufficient to meet the analytical needs, there is always room for improvement in terms of user-friendliness.
+It is essential to consider the needs of all users, including the least experienced user.
+To prevent program crashes and errors, it may be necessary to implement exception handling around code blocks, as well as validate-need statements for the validation of user input.
+Additionally, creating a user-friendly environment involves establishing effective channels of communication between users and developers, such as utilizing platforms like GitHub for collaboration and feedback.
 
@@
 
-Through this project, I demonstrated that computational analysis can be
-made accessible to a wide range of biologists by providing an
-interactive environment. It is important to recognize that there cannot
-be a one-size-fits-all approach for proteomics datasets. Therefore, the
-application offers easy-to-interpret visualizations that allow users to
-assess the patterns and trends in their observed datasets. The default
-options provided in the application are based on successful analysis
-pipelines such as ScoPE2 \citep{Specht2021}, serving as a starting point
-for customization at multiple levels.
-
-The example dataset provided showcases the inherent challenges in
-proteomic data analysis. It exhibits skewed distributions, batch
-effects, and a substantial number of missing values. Despite these
-complexities, the ProteoScanR pipeline and Proteomics Workbench
-interface offer biologists a convenient means to address their research
-questions and validate the computations performed. The example use case
-also sheds light on the fact that previous proteomics publications, such
-as the ScoPE2 study \citep{Specht2021}, may not always employ the gold
-standard in data processing. This emphasizes the importance of
-critically evaluating and refining data processing steps through trial
-and error to achieve optimal results. The interactive environment
-provided by the application allows users to navigate the trial and error
-process without experiencing frustration. This user-friendly interface
-ensures that researchers can iteratively refine their data processing
-steps and analyze the results in a seamless manner. Furthermore, the
-scalability of the application was initially unknown until the final
-submission of the single-cell dataset. The programming and testing of
-the application primarily focused on bulk datasets, typically consisting
-of up to four runs at once. However, it is noteworthy that the
-application demonstrates remarkable efficiency and performance, even
-when running large datasets. The ability of the application to
-efficiently handle larger datasets, such as single-cell studies, is a
-significant advantage. It allows researchers to analyze complex and
-extensive datasets without sacrificing performance or compromising the
-quality of the results. This scalability makes the application a
-valuable tool for analyzing diverse types of biological data,
-accommodating the needs of researchers working with both smaller and
-larger datasets.
-
-ProteoScanR pipeline and Proteomics Workbench user interface serve as
-valuable tools for individual researchers and research communities,
-enabling them to analyze and visualize their data in meetings, even on
-non-performant computers. However, one of the future perspectives
-involves deploying the application on a server, making it accessible to
-research groups around the world. This would enhance collaboration and
-facilitate the sharing of insights and findings across different
-geographical locations. By providing a centralized platform for
-proteomics analysis, aiming to foster scientific advancements and enable
-researchers to make significant contributions to their respective
-fields. Furthermore by having a central hub for proteomic data analysis
-and the use of artificial intelligence the biological meaning could be
-brought to a wider context. Since proteomic data could be saved without
-violating data protection laws, algorithms could be employed to learn
-from large datasets.
-
-An important aspect of learning from the data involves annotating
-specimens, particularly in the context of single-cell proteomics. The
-clustering approach is a primary focus for future development and
-analysis in the near term. By employing clustering algorithms and
-stratifying the data, individual cells can be labeled and annotated
-based on their phenotypes. This annotation process involves associating
-protein sets collections with specific clusters or cell types, providing
-a more comprehensive understanding of cellular heterogeneity and
-functional characteristics.
-
-In conclusion, the ProteoScanR pipeline and Proteomics Workbench
-interface provide a user-friendly and efficient approach for proteomic
-data analysis, enabling biologists to address their research questions
-and validate computations. Future perspectives include deploying the
-application on a server for global accessibility, leveraging artificial
-intelligence techniques, and implementing clustering approaches to
-further enhance annotation and understanding of cellular heterogeneity
-and functional characteristics.
+Through this project, I demonstrated that computational analysis can be made accessible to a wide range of biologists by providing an interactive environment.
+It is important to recognize that there cannot be a one-size-fits-all approach for proteomics datasets.
+Therefore, the application offers numerous easy-to-interpret visualizations that allow users to assess the patterns and trends in their observed datasets.
+The default options provided in the application are based on successful analysis pipelines such as ScoPE2 \citep{Specht2021}, serving as a starting point for customization at multiple levels.
+
+The example dataset provided showcases the inherent challenges in proteomic data analysis.
+It exhibits skewed distributions, batch effects, and a substantial number of missing values.
+Despite these complexities, the ProteoScanR pipeline and Proteomics Workbench interface offer biologists a convenient means to address their research questions and validate the computations performed.
+The example use case also sheds light on the fact that previous proteomics publications, such as the ScoPE2 study \citep{Specht2021}, may not always employ the gold standard in data processing.
+This emphasizes the importance of critically evaluating and refining data processing steps through trial and error to achieve optimal results.
+The interactive environment provided by the application allows users to navigate the trial and error process without experiencing frustration.
+This user-friendly interface ensures that researchers can iteratively refine their data processing steps and analyze the results in a seamless manner.
+Furthermore, the scalability of the application was initially unknown until the final submission of the single-cell dataset.
+The programming and testing of the application primarily focused on bulk datasets, typically consisting of up to four runs at once.
+However, it is noteworthy that the application demonstrates remarkable efficiency and performance, even when running large datasets.
+The ability of the application to efficiently handle larger datasets, such as single-cell studies, is a significant advantage.
+It allows researchers to analyze complex and extensive datasets without sacrificing performance or compromising the quality of the results.
+This scalability makes the application a valuable tool for analyzing diverse types of biological data, accommodating the needs of researchers working with both smaller and larger datasets.
+
+ProteoScanR pipeline and Proteomics Workbench user interface serve as valuable tools for individual researchers and research communities, enabling them to analyze and visualize their data in meetings, even on non-performant computers.
+However, one of the future perspectives involves deploying the application on a server, making it accessible to research groups around the world.
+This would enhance collaboration and facilitate the sharing of insights and findings across different geographical locations.
+By providing a centralized platform for proteomics analysis, aiming to foster scientific advancements and enable researchers to make significant contributions to their respective fields.
+Furthermore by having a central hub for proteomic data analysis and the use of artificial intelligence the biological meaning could be brought to a wider context.
+Since proteomic data could be saved without violating data protection laws, algorithms could be employed to learn from large datasets.
+
+An important aspect of learning from the data involves annotating specimens, particularly in the context of single-cell proteomics.
+The clustering approach is a primary focus for future development and analysis in the near term.
+By employing clustering algorithms and stratifying the data, individual cells can be labeled and annotated based on their phenotypes.
+This annotation process involves associating protein sets collections with specific clusters or cell types, providing a more comprehensive understanding of cellular heterogeneity and functional characteristics.
+
+In conclusion, the ProteoScanR pipeline and Proteomics Workbench interface provide a user-friendly and efficient approach for proteomic data analysis, enabling biologists to address their research questions and validate computations.
+Future perspectives include deploying the application on a server for global accessibility, leveraging artificial intelligence techniques, and implementing clustering approaches to further enhance annotation and understanding of cellular heterogeneity and functional characteristics.
 
@@
 
-\hypertarget{wet-lab}{%
-\subsection{Wet lab}\label{wet-lab}}
-
-\hypertarget{cells-sorting-protocol}{%
-\subsubsection{Cells sorting protocol}\label{cells-sorting-protocol}}
-
-\begin{enumerate}
-\item 96 single cells from each donor were sorted into a single plate. Each well contained a single cell. As such, 12 different plates containing single cells from 12 different donors were prepared. 
-\item Carrier plates were also prepared. In the carrier plates, 100 cells from one donor were sorted into each well. 24 such wells (3 columns) were prepared for each donor.
-\end{enumerate}
-
-\hypertarget{digestion-protocol}{%
-\subsubsection{Digestion protocol}\label{digestion-protocol}}
-
-\begin{enumerate}
-\item Prepare 25 ng/ $\mu$L trypsin in 100 mmol/L Triethylammonium bicarbonate buffer (TEAB).
-\item Add 25 ng/$\mu$L trypsin: 2 $\mu$L to the carrier samples, 1 $\mu$L to the single cell samples.
-\item Seal the plates.
-\item Centrifuge (2 min at 120 G).
-\item Incubate overnight at 37 °C.
-\end{enumerate}
-
-\hypertarget{labeling-protocol}{%
-\subsubsection{Labeling protocol}\label{labeling-protocol}}
-
-\begin{enumerate}
-\item Let the TMTpro labels come to room temperature so that all the condensation on the tube is gone.
-Dissolve the TMTpro reagents in anhydrous acetonitril (ACN) to obtain a concentration of 10 $\mu$g/$\mu$L.
-\item Add 1 $\mu$L of TMTpro reagents according to the tables. For one set of plates, you will need 16 $\mu$L of each label. It is recommended to aliquot a slightly larger volume, around 20 $\mu$L.
-\item Seal the plates.
-\item Centrifuge for 2 minutes at 120 G.
-\item Incubate for 1 hour at room temperature.
-\item Prepare a 5% hydroxylamine solution in water.
-\item Add 1 $\mu$L of the 5% hydroxylamine solution to each well.
-\item Seal the plates.
-\item Centrifuge for 2 minutes at 120 G.
-\item Incubate for 1 hour at room temperature.
-\item Cover the plates with cap strips.
-\end{enumerate}
-
-\hypertarget{pooling-protocol}{%
-\subsubsection{Pooling protocol}\label{pooling-protocol}}
-
-\begin{enumerate}
-\item Wash a 10 $\mu$L glass microsyringe five times with ACN.
-\item Use the microsyringe to combine TMTpro-labeled samples into 36 LC autosampler vials with built-in glass inserts.
-\item Combine six carrier wells from each sample and add the carrier proteome to the 36 TMTpro sets.
-\item Combine the first set by taking well A1 from the first single-cell plate, well A2 from the second plate, well A3 from the third plate, and so on.
-For the next set, take well A1 from the second plate, well A2 from the third plate, and so on.
-Combine 12 sets following this scheme.
-Combine the next 24 sets by following the same scheme for rows B and C.
-\item Vacuum centrifuge the combined sets to dryness.
-\item Dissolve the dried samples in 7 $\mu$L of buffer A and transfer them to LC-MS/MS for further analysis.
-\end{enumerate}
-
 \hypertarget{dry-lab}{%
@@
 \begin{verbatim}
-R version 4.3.0 (2023-04-21)
-Platform: x86_64-pc-linux-gnu (64-bit)
-Running under: Ubuntu 22.04.2 LTS
-
- [1] visNetwork_2.1.2            snowfall_1.84-6.2          
- [3] snow_0.4-4                  piano_2.16.0               
- [5] org.Hs.eg.db_3.17.0         AnnotationDbi_1.62.1       
- [7] infotheo_1.2.0.1            clusterProfiler_4.8.1      
- [9] plotly_4.10.1               sva_3.48.0                 
-[11] BiocParallel_1.34.2         genefilter_1.82.1          
-[13] mgcv_1.8-42                 nlme_3.1-162               
-[15] impute_1.74.1               CONSTANd_1.8.0             
-[17] limma_3.56.1                scater_1.28.0              
-[19] scuttle_1.10.1              reshape2_1.4.4             
-[21] dplyr_1.1.2                 magrittr_2.0.3             
-[23] ggplot2_3.4.2               SingleCellExperiment_1.22.0
-[25] scp_1.10.0                  QFeatures_1.10.0           
-[27] MultiAssayExperiment_1.26.0 SummarizedExperiment_1.30.1
-[29] Biobase_2.60.0              GenomicRanges_1.52.0       
-[31] GenomeInfoDb_1.36.0         IRanges_2.34.0             
-[33] S4Vectors_0.38.1            BiocGenerics_0.46.0        
-[35] MatrixGenerics_1.12.0       matrixStats_0.63.0         
-[37] spsComps_0.3.2.1            DT_0.28                    
-[39] shinydashboard_0.7.2        shinyWidgets_0.7.6         
-[41] shiny_1.7.4                
+## R version 4.3.1 (2023-06-16)
+## Platform: x86_64-pc-linux-gnu (64-bit)
+## Running under: Ubuntu 23.04
+## 
+## Matrix products: default
+## BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3 
+## LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.21.so;  LAPACK version 3.11.0
+## 
+## locale:
+##  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+##  [3] LC_TIME=de_AT.UTF-8        LC_COLLATE=en_US.UTF-8    
+##  [5] LC_MONETARY=de_AT.UTF-8    LC_MESSAGES=en_US.UTF-8   
+##  [7] LC_PAPER=de_AT.UTF-8       LC_NAME=C                 
+##  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+## [11] LC_MEASUREMENT=de_AT.UTF-8 LC_IDENTIFICATION=C       
+## 
+## time zone: Europe/Vienna
+## tzcode source: system (glibc)
+## 
+## attached base packages:
+##  [1] parallel  stats4    grid      stats     graphics  grDevices utils    
+##  [8] datasets  methods   base     
+## 
+## other attached packages:
+##  [1] visNetwork_2.1.2            snowfall_1.84-6.2          
+##  [3] snow_0.4-4                  piano_2.16.0               
+##  [5] org.Hs.eg.db_3.17.0         AnnotationDbi_1.62.2       
+##  [7] infotheo_1.2.0.1            clusterProfiler_4.8.2      
+##  [9] plotly_4.10.2               spsComps_0.3.3.0           
+## [11] DT_0.28                     shinydashboard_0.7.2       
+## [13] shinyWidgets_0.7.6          shiny_1.7.4.1              
+## [15] cluster_2.1.4               factoextra_1.0.7           
+## [17] tibble_3.2.1                sva_3.48.0                 
+## [19] BiocParallel_1.34.2         genefilter_1.82.1          
+## [21] mgcv_1.8-42                 nlme_3.1-162               
+## [23] impute_1.74.1               CONSTANd_1.8.0             
+## [25] limma_3.56.2                scater_1.28.0              
+## [27] scuttle_1.10.1              reshape2_1.4.4             
+## [29] dplyr_1.1.2                 ggplot2_3.4.2              
+## [31] SingleCellExperiment_1.22.0 scp_1.10.1                 
+## [33] QFeatures_1.10.0            MultiAssayExperiment_1.26.0
+## [35] SummarizedExperiment_1.30.2 Biobase_2.60.0             
+## [37] GenomicRanges_1.52.0        GenomeInfoDb_1.36.1        
+## [39] IRanges_2.34.1              S4Vectors_0.38.1           
+## [41] BiocGenerics_0.46.0         MatrixGenerics_1.12.2      
+## [43] matrixStats_1.0.0           magrittr_2.0.3             
+## [45] glue_1.6.2                  Gmisc_3.0.2                
+## [47] htmlTable_2.4.1             Rcpp_1.0.11                
+## [49] DiagrammeR_1.0.10           tinytex_0.45               
+## 
+## loaded via a namespace (and not attached):
+##   [1] ProtGenerics_1.32.0       bitops_1.0-7             
+##   [3] enrichplot_1.20.0         lubridate_1.9.2          
+##   [5] HDO.db_0.99.1             httr_1.4.6               
+##   [7] RColorBrewer_1.1-3        tools_4.3.1              
+##   [9] backports_1.4.1           utf8_1.2.3               
+##  [11] R6_2.5.1                  lazyeval_0.2.2           
+##  [13] withr_2.5.0               gridExtra_2.3            
+##  [15] cli_3.6.1                 shinyjs_2.1.0            
+##  [17] scatterpie_0.2.1          slam_0.1-50              
+##  [19] labeling_0.4.2            sass_0.4.7               
+##  [21] yulab.utils_0.0.6         gson_0.1.0               
+##  [23] foreign_0.8-84            DOSE_3.26.1              
+##  [25] rstudioapi_0.15.0         RSQLite_2.3.1            
+##  [27] gridGraphics_0.5-1        generics_0.1.3           
+##  [29] gtools_3.9.4              car_3.1-2                
+##  [31] GO.db_3.17.0              Matrix_1.6-0             
+##  [33] ggbeeswarm_0.7.2          fansi_1.0.4              
+##  [35] abind_1.4-5               lifecycle_1.0.3          
+##  [37] yaml_2.3.7                edgeR_3.42.4             
+##  [39] carData_3.0-5             shinytoastr_2.1.1        
+##  [41] gplots_3.1.3              qvalue_2.32.0            
+##  [43] blob_1.2.4                promises_1.2.0.1         
+##  [45] crayon_1.5.2              lattice_0.21-8           
+##  [47] beachmat_2.16.0           cowplot_1.1.1            
+##  [49] annotate_1.78.0           KEGGREST_1.40.0          
+##  [51] pillar_1.9.0              knitr_1.43               
+##  [53] fgsea_1.26.0              marray_1.78.0            
+##  [55] codetools_0.2-19          fastmatch_1.1-3          
+##  [57] downloader_0.4            ggfun_0.1.1              
+##  [59] data.table_1.14.8         treeio_1.24.2            
+##  [61] vctrs_0.6.3               png_0.1-8                
+##  [63] gtable_0.3.3              assertthat_0.2.1         
+##  [65] cachem_1.0.8              xfun_0.39                
+##  [67] S4Arrays_1.0.4            mime_0.12                
+##  [69] tidygraph_1.2.3           survival_3.5-5           
+##  [71] sets_1.0-24               ellipsis_0.3.2           
+##  [73] ggtree_3.8.0              bit64_4.0.5              
+##  [75] bslib_0.5.0               irlba_2.3.5.1            
+##  [77] KernSmooth_2.23-21        vipor_0.4.5              
+##  [79] rpart_4.1.19              colorspace_2.1-0         
+##  [81] DBI_1.1.3                 Hmisc_5.1-0              
+##  [83] nnet_7.3-19               tidyselect_1.2.0         
+##  [85] bit_4.0.5                 compiler_4.3.1           
+##  [87] BiocNeighbors_1.18.0      DelayedArray_0.26.6      
+##  [89] shadowtext_0.1.2          bookdown_0.34            
+##  [91] caTools_1.18.2            checkmate_2.2.0          
+##  [93] scales_1.2.1              relations_0.6-13         
+##  [95] stringr_1.5.0             digest_0.6.33            
+##  [97] rmarkdown_2.23            XVector_0.40.0           
+##  [99] htmltools_0.5.5           pkgconfig_2.0.3          
+## [101] base64enc_0.1-3           sparseMatrixStats_1.12.2 
+## [103] fastmap_1.1.1             rlang_1.1.1              
+## [105] htmlwidgets_1.6.2         DelayedMatrixStats_1.22.1
+## [107] farver_2.1.1              jquerylib_0.1.4          
+## [109] jsonlite_1.8.7            GOSemSim_2.26.1          
+## [111] BiocSingular_1.16.0       RCurl_1.98-1.12          
+## [113] Formula_1.2-5             GenomeInfoDbData_1.2.10  
+## [115] ggplotify_0.1.1           patchwork_1.1.2          
+## [117] munsell_0.5.0             ape_5.7-1                
+## [119] viridis_0.6.3             MsCoreUtils_1.12.0       
+## [121] stringi_1.7.12            ggraph_2.1.0             
+## [123] zlibbioc_1.46.0           MASS_7.3-60              
+## [125] plyr_1.8.8                ggrepel_0.9.3            
+## [127] Biostrings_2.68.1         graphlayouts_1.0.0       
+## [129] splines_4.3.1             locfit_1.5-9.8           
+## [131] igraph_1.5.0              ggpubr_0.6.0             
+## [133] ggsignif_0.6.4            ScaledMatrix_1.8.1       
+## [135] XML_3.99-0.14             evaluate_0.21            
+## [137] tweenr_2.0.2              httpuv_1.6.11            
+## [139] tidyr_1.3.0               purrr_1.0.1              
+## [141] polyclip_1.10-4           clue_0.3-64              
+## [143] BiocBaseUtils_1.2.0       ggforce_0.4.1            
+## [145] rsvd_1.0.5                broom_1.0.5              
+## [147] xtable_1.8-4              AnnotationFilter_1.24.0  
+## [149] tidytree_0.4.4            rstatix_0.7.2            
+## [151] later_1.3.1               viridisLite_0.4.2        
+## [153] aplot_0.1.10              forestplot_3.1.1         
+## [155] memoise_2.0.1             beeswarm_0.4.0           
+## [157] timechange_0.2.0          shinyAce_0.4.2
 \end{verbatim}

### 2023-07-18 markdown/discussion.Rmd
until batch correction done

@@
 With the selection of factors to indicate over multiple plots, biases can be identified on various levels.
-Additionally, the analysis can be examined on an individual protein basis, enabling the identification of suspicious results.
+Additionally, the analysis can be examined on an individual protein basis, enabling the identification of suspicious results and closer validation of significant candidates.
 The statistics module provides users with a simple tool to assess and visualize the results, offering clarity to dense scatter plots through hover functions.
@@
 ## Feature-wise plot
-The feature-wise output plays a crucial role in enhancing the analysis pathway by offering users the ability to select specific proteins and explore their associated peptide spectra, peptide values, and protein values throughout the analysis.
+The feature-wise output plays a crucial role in enhancing the analysis workflow by offering users the ability to select specific proteins and explore their associated peptide spectra, peptide values, and protein values throughout the analysis.
 This feature provides a detailed and comprehensive view of the analysis results, enabling users to identify any missing or erroneous data points, particularly in channels with low or high abundance.
@@
 
-## Enrichment 
+## Enrichment analysis
 The enrichment analysis module within the application addresses the challenge of selecting protein sets for mapping by incorporating a database-fetching functionality.
@@
 
+n_missingness <- sum(!complete.cases(assay(scp[["proteins_norm"]])))
+
 nrows_dataset <- nrow(assay(scp[["proteins_norm"]]))
@@
 
-Unfortunately, a substantial number of individual proteins in the data set (totaling `r nrows_dataset`) have missing values.
-On average, this amounts to approximately `r round(mean_missingness, 2)*100`% of the proteins across all runs and channels.
-This situation leads to the need for missing value imputation using techniques such as k-nearest neighbor (K-NN), which can result in averaged results.
+Unfortunately, a substantial number of individual proteins in the data set (totaling `r n_missingness`) have missing values.
+On average, this amounts to approximately `r round(mean_missingness, 2)*100`% of the proteins across all runs and channels (see figure \ref@(fig:missingness)).
+This situation leads to the need for missing value imputation using techniques such as k-nearest neighbor (K-NN), which can sometimes lead to artificially evenly distributed values.
 Additionally, it is important to note that this imputation process can introduce potential loss of power in subsequent statistical analysis.
@@
 
-As observed in the dendrogram, it is evident that sample type alone does not account for the variation observed in the data.
+As observed in the dendrogram (see figure \ref@(fig:cluster-analysis-dendrogram)), it is evident that sample type alone does not account for the variation observed in the data.
 The branching patterns in the dendrogram suggest that other factors might be significant for contributing to the averaging of protein expression levels.
@@
 
-The gap statistic compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution.
+The gap statistic (see figure \ref@(fig:cluster-analysis-gap-stat)) compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution.
 The null reference distribution represents a scenario where no clear clustering patterns exist in the data.
@@
 
-The silhouette method calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters. 
+The silhouette method (see figure \ref@(fig:cluster-analysis-silh)) calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters.
 If the overall distances between clusters are small, indicating overlapping or unclear separation, it is suggested to take the minimum k.
@@
 
-As seen in the principal components analysis, clusters are densely packed, and the plot shows no clear differentiation between the `r optimal_clusters` suggested clusters.
+As seen in the principal components analysis (see figure \ref@(fig:cluster-analysis-dim-red)), clusters are densely packed, and the plot shows no clear differentiation between the `r optimal_clusters` suggested clusters.
 The dense packing and poor resolution indicates that other factors may be influencing the lack of distinct separation.
@@
 ### Machine learning
-One potential target for future research could involve leveraging artificial intelligence techniques to predict cellular behavior, cell types, or even diseases.
+One potential target for future research could also involve leveraging artificial intelligence techniques to predict cellular behavior, cell types, or even diseases.
 By running the application on a server and creating a database, it would be possible to harness the power of AI algorithms such as decision trees, random forests, or support vector machines.
@@
 It is essential to consider the needs of all users, including the least experienced user.
-To prevent program crashes and errors, it may be necessary to implement try-catch around code blocks, as well as validate-need statements for the validation of user input.
+To prevent program crashes and errors, it may be necessary to implement exception handling around code blocks, as well as validate-need statements for the validation of user input.
 Additionally, creating a user-friendly environment involves establishing effective channels of communication between users and developers, such as utilizing platforms like GitHub for collaboration and feedback.

### 2023-07-18 markdown/methods.Rmd
until batch correction done

@@
 ## Wetlab - Materials and Methods
-For testing and development of the ProteoScanR pipeline the following materials and methods were performed.
+For testing and development of the ProteoScanR pipeline the following materials and methods were applied.
 
 ### Celltypes
-Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome. Cell sorting was performed with FACSAria Fusion (BD Biosciences).
+Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome were isolated.
+Cell sorting was performed with FACSAria Fusion (BD Biosciences) by auto fluorescence.
 
@@
 ### Labeling
-Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.
+Samples of the test data set were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.
 
@@
 ### Hardware
+Data was analysed using commercially available hardware, find details below:
 Lenovo Laptop X360 with:
@@
 
-In order to streamline the analysis of multiple experiments, object oriented programming was applied.
+In order to streamline the analysis of multiple experiments and software development, object oriented programming was applied.
 The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
@@
 ### Filter by q-value
-The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
+The next step for quality control is the exclusion of spectra with a high false discovery rate (FDR).
 When applying multiple statistical testing (e.g.
-t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed.
+t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant result will iteratively increase with each test performed.
 Corrections in statistics are an approach to compensate for the multiplicity of testing.
 There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR.
-In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score.
+Many search engines face that problem by calculating the false discovery rate, by dividing false PSMs (=hit of the decoy database) with the total number of PSMs above the peptide-spectrum matching score.
 The peptides spectrum matching score is defined as $-10 * log(p)$.
@@
 The calculation of the score is highly dependent on the data acquisition method used.
-MaxQuant uses Andromeda, an integrated search engine.
-Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora.
+MaxQuant uses Andromeda \citep{Cox2011}, an integrated search engine.
+Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot \citep{Brosch2009}.
 As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage.
@@
 In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function.
-In the context of processing from PSMs to unique peptides, the desired column is the PSM sequence.
+In the context of processing of PSMs to unique peptides, the desired column is the peptide sequence.
 To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.
@@
 Sample size is often a limiting factor in hypothesis testing.
-A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to an early end of analyses.
+A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to diminished statistical power.
 To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis.
@@
 This can be due two reasons.
-One is the expression level of the given proteins in a cell.
-Meaning, that the expression of the observed cell type is simply lower than the other type.
+One is the expression level of the given proteins in a sample.
+Meaning, that the expression of the observed sample type is simply lower than in the other type.
 Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes.
@@
 Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data.
-The coefficient of variation of a peptide is considered as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins).
+The coefficient of variation of a peptide is defined as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins).
 Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.
@@
 This transformation was used as a reference when comparing methods.
-However, using the logarithm to the base 10 may be an easy way to recalculate expression data in the mind and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
+However, using the logarithm to the base 10 has the same effect and improves the interpretability of the values.
+Furthermore the log10 transformation will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
 When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
@@
 The second method for normalization included is the CONSTANd method, which was proven suitable for relative quantification in the field of proteomics \citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq data.
-CONSTANd uses a technique called matrix raking, which employs the RAS algorithm.
-The expression set reflects the non-negative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $B$ which equals to $diag(x) * A *diag(y)$ .
-Whereas $x \in \mathbb{R}^{m}$ and $y \in \mathbb{R}^{n}$.
-The solution to this problem involves finding the row sum of $B=u_i$ and the column sum of $B=v_j$, where $i$ and $j$ denote the row and column indices of the matrix, respectively \citep{Bacharach1965}.
-In other words, the matrix will be alternatively manipulated on rows and columns until the mean of both equals 1.
+CONSTANd uses a technique called matrix raking, which employs the RAS ($K=RAS$) algorithm.
+The expression set reflects the non-negative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $K$ which equals to $diag(x) * A *diag(y)$, so that $K=diag(x) * A *diag(y)= RAS$.
+Whereas $x \in \mathbb{R}^{m}$, $y \in \mathbb{R}^{n}$ and $R$ denotes for the row multiplier and $S$ for the column multiplier.
+The solution to this problem involves finding the row sum of $K=u_i$ and the column sum of $K=v_j$, where $i$ and $j$ denote the row and column indices of the matrix, respectively \citep{Bacharach1965}.
+In other words, the matrix will be alternatingly manipulated on rows and columns until the mean of both equals 1.
 In order to apply the method and yield true results, one has to consider 3 major assumptions between sample types.
-This assumption can be observed in the MA-plot which indicates the differences of two samples.
+These assumption can be observed in the MA-plot which indicates the differences of two samples.
 Given the intensities of two samples, $R$ and $G$, visualized on a graph with the x-axis as $M=\log_2(\frac{R}{G})$ and the y-axis as $A=\frac{1}{2}\log_2(RG)$.
@@
 When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches.
-From a biological perspective, it is highly unlikely that a cell completely lacks a single protein, as it would imply a complete loss of function for that protein.
-Therefore, an intensity value of 0 is close to impossible for most proteins.
 In order to handle this dark space in the expression set, several methods can be chosen in the application.
-A variety of function employs replacing the missing values with the mean or median of the rest of the matrix.
+A variety of function employs replacing the missing values with the mean or median of the rest of the matrix (total mean/median).
 This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well.
-Even a more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
-A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing.
-$K$ denotes for the count of neighboring values of a gene and can be selected upon preference.
+An even more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
+A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values according to an euclidean metric of the neighboring values in the columns where the feature is not missing.
+$K$ denotes for the count of neighboring values of a protein which are considered and can be selected upon preference.
 The euclidean distance between two points is defined as $d = \sqrt{a^2 * b^2}$, the $K$ nearest neighbors average value will be assigned to the intensity of the missing feature.
@@
 Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data.
-This variance would overshadow the biological differences across sample types when performing dimensionality reduction.
+This variance can overshadow the biological differences across sample types when performing dimensionality reduction.
 A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of ProteoScanR.

### 2023-07-18 markdown/results.Rmd
until batch correction done

@@
 This allows users to observe the direct impact of the applied changes on the dataset.
-In the summary visualization, it may appear cluttered due to the processing of 36 files. In smaller datasets, the nodes of the network graph can be more easily distinguished and understood.
+In the summary visualization, it may appear cluttered due to the processing of 36 files. In smaller data sets, the nodes of the network graph can be more easily distinguished and understood.
 When running the application with SCoPE2 \citep{Specht2021} reference cutoffs, the following results were visualized.
@@
 ```
-In order to get an idea of the magnitude of the data set, the barplot provides an overview (see figure \@ref(fig:ui-numbers-default))
+In order to get an idea of the magnitude of the data set, the barplot provides an overview (see figure \@ref(fig:ui-numbers-default)).
 The default settings of the analysis pipeline resulted in a total number of unique proteins below approximately 1,000, which were aggregated from around 3,000 peptides.
@@
 ```
+Choosing the CONSTANd option for normalization a popup window appears to test for assumptions on the dataset.
 
@@
 ```
-Comparing sample types in regards of reporter ion intensity is recommended to observe any experimental bias such as batch/channel confounding effects which would influence the statistical model afterwards.
+Comparing sample types in regards of reporter ion intensity is recommended to observe any experimental bias such as batch/channel confounding effects which would negatively impact the statistical model afterwards.
 
@@
 ```
-The distribution of median coefficient of variation (CV) shows the uniqueness in peptides for their respective razor peptides.
+The distribution of median coefficient of variation (CV) shows the uniqueness in peptides for their respective razor protein.
 This metric represents the presence or absence of noisy matches in the data set.
@@
 ```
-Razor peptide ambiguity over multiple batches can indicate experimental failures during sample prep or during acquisition.
+Razor peptide ambiguity over multiple batches can indicate experimental failures during sample prep (especially digestion) or during acquisition.
 

### 2023-07-18 markdown/results.Rmd
references improved

@@
 
-The plot (see figure \ref@(fig:ui-cv-rawfile-default)) illustrates that certain batches (=raw files) exhibit a higher number of distinct matches compared to others, as indicated by a thin line depicting low density.
+The plot (see figure \@ref(fig:ui-cv-rawfile-default)) illustrates that certain batches (=raw files) exhibit a higher number of distinct matches compared to others, as indicated by a thin line depicting low density.
 Most interquartile ranges (IQRs) overlap, suggesting similarity among the majority of batches.
@@
 
-By setting the number of observations to 2, the distributions become more equalized (see figure \ref@(fig:ui-cv-rawfile-nobs2)).
+By setting the number of observations to 2, the distributions become more equalized (see figure \@ref(fig:ui-cv-rawfile-nobs2)).
 Additionally, batches that initially exhibited a decreased coefficient of variation (CV) tend to inflate towards the threshold.
@@
 
-When the minimum number of observations for ambiguous peptides is set to 10, the distribution of medians across channels shows a decrease in density (see figure \ref@(fig:ui-cv-rawfile-nobs10)).
+When the minimum number of observations for ambiguous peptides is set to 10, the distribution of medians across channels shows a decrease in density (see figure \@ref(fig:ui-cv-rawfile-nobs10)).
 This can be attributed to the rare occurrence of 10 unique peptides matching to the same protein, which is rarely observed in the dataset.
@@
 ```
-The heatmap (see figure \ref@(fig:ui-heatmap)) reveals the Pearson pairwise correlation coefficients between all proteins.
+The heatmap (see figure \@ref(fig:ui-heatmap)) reveals the Pearson pairwise correlation coefficients between all proteins.
 The dendrogram indicates that there are two main protein clusters, with one of the clusters showing further sub-clustering, indicating a strong relationship among the proteins in the sub groups.
@@
 
-In both the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP) plot (see figure \ref@(fig:ui-dim-red-default)), the second component is plotted against the first component.
+In both the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP) plot (see figure \@ref(fig:ui-dim-red-default)), the second component is plotted against the first component.
 Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch.
@@
 
-The second figure (\ref@(fig:ui-dim-red-changed-k) of the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP), both displaying the second component against the first.
+The second figure (\@ref(fig:ui-dim-red-changed-k) of the Principal Component Analysis (PCA) plot and the Uniform Manifold Approximation and Projection (UMAP), both displaying the second component against the first.
 Surprisingly, neither visualization reveals any discernible clustering patterns based on sample type or batch information.
@@
 
-In the 3D visualization (see figure \ref@(fig:ui-dim-red-c3d)), a subtle clustering pattern can be observed for the first two batches.
+In the 3D visualization (see figure \@ref(fig:ui-dim-red-c3d)), a subtle clustering pattern can be observed for the first two batches.
 Upon switching to the indicator sample type in the 3D visualization, no clustering can be observed whatsoever.
@@
 
-After applying batch correction using ComBat \citep{Leek2012}, neither the 2D (figure \ref@(fig:ui-dim-red-batchC)) nor the 3D (figure \ref@(fig:ui-dim-red-batchC-3d)) visualization demonstrates any noticeable effects related to batches or sample types.
+After applying batch correction using ComBat \citep{Leek2012}, neither the 2D (figure \@ref(fig:ui-dim-red-batchC)) nor the 3D (figure \@ref(fig:ui-dim-red-batchC-3d)) visualization demonstrates any noticeable effects related to batches or sample types.
 The previously observed subtle clustering of the first two batches is now eliminated, indicating successful batch correction.
@@
 
-The scatter plots (see figure \ref@(fig:ui-feature)) indicate the detected values associated with the protein UniProt ID P61313 (= 60S ribosomal protein L15).
+The scatter plots (see figure \@ref(fig:ui-feature)) indicate the detected values associated with the protein UniProt ID P61313 (= 60S ribosomal protein L15).
 Starting from the peptide spectrum level and progressing to peptides aggregated for individual raw files, these plots offer a comprehensive view of the protein's peptide abundance across various batches.
@@
 
-The values for all sample types exhibit a unimodal distribution (see figure \ref@(fig:ui-stat-dependency)).
+The values for all sample types exhibit a unimodal distribution (see figure \@ref(fig:ui-stat-dependency)).
 However, when examining the quantile-quantile (qq) plot, heavy tails can be observed for all three groups.
@@
 
-By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group (see figure \ref@(fig:ui-stat-HC-HIV-noMets)), six differentially expressed proteins were identified.
+By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group (see figure \@ref(fig:ui-stat-HC-HIV-noMets)), six differentially expressed proteins were identified.
 The Venn diagram indicates significantly up/down regulated proteins.
@@
 
-The pathway enrichment mapped and tested with the exact Fisher test \citep{Sprent2011} the differentially expressed proteins P46777 (60S ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the ribosome and coronavirus disease (COVID-19) nodes (see figure \ref@(fig:ui-path-enrich-nw)).
+The pathway enrichment mapped and tested with the exact Fisher test \citep{Sprent2011} the differentially expressed proteins P46777 (60S ribosomal protein L5) and P47914 (60S ribosomal protein L29) to the ribosome and coronavirus disease (COVID-19) nodes (see figure \@ref(fig:ui-path-enrich-nw)).
 
@@
 
-The barplot (see figure \ref@(fig:ui-path-enrich-bar)) shows the significance of the two nodes.
+The barplot (see figure \@ref(fig:ui-path-enrich-bar)) shows the significance of the two nodes.
 The count indicates how many proteins are associated with the node of the network.
@@
 
-The position of the dot (see figure \ref@(fig:ui-go-enrich-bar)) shows the number of matches for the particular protein set.
+The position of the dot (see figure \@ref(fig:ui-go-enrich-bar)) shows the number of matches for the particular protein set.
 Colors indicate for the significance level.

### 2023-07-19 markdown/Thesis.Rmd
thesis finalized

@@
 Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction.
-ProteoScanR and the Proteomics Workbench served as a valuable tool in guiding the identification of significantly expressed proteins in cells under study.
-Subsequently, the pathway enrichment analysis provided additional biological contexts for a comprehensive understanding of their functional implications.
+ProteoScanR and the Proteomics Workbench serves as a valuable tool in guiding the identification of expressed proteins in cells under study.
+Subsequently, the pathway enrichment analysis provides additional biological contexts for a comprehensive understanding of their functional implications.
 The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics.
-Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.
-https://github.com/Lukas67/ProteoScanR
+Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application
+The source code is accessible and can be located at the following URL: https://github.com/Lukas67/ProteoScanR
 

### 2023-07-19 markdown/introduction.Rmd
thesis finalized

@@
 Before the initial exploration of molecular biology, the processes of life were roughly inferred and not well understood.
-The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins.
+Among other principles the central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins.
 This framework consists of two major processes: transcription, which converts DNA into RNA, and translation, which converts RNA into functional proteins.
@@
 Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
-Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as "omics," to comprehensively analyze biomolecules in diverse contexts.
-Omics research has emerged as an approach to study biomolecules and their interactions within biological systems comprehensively.
+Over the past six decades, researchers have explored various domains of this principle, collectively referred to as "omics," to comprehensively analyze biomolecules in diverse contexts.
+Omics research has emerged as an unbiased approach to study biomolecules and their interplay within biological systems comprehensively.
 Omics encompasses various fields, including genomics, transcriptomics, proteomics, and metabolomics, with each field focusing on a specific class of biomolecules.
@@
 Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks.
-Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role in this interactive structure.
+Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role within this interactive structure.
 Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}
@@
 Combining them results in the protein secondary structure such as alpha helix and beta sheet.
-These secondary structures further assemble to create higher-order tertiary structures, which represent specific domains within the protein.
+These secondary structures further assemble to create higher-order tertiary structures, which represent specific structural domains within the protein.
 Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
@@
 Each tissue consists of cells with specific functions, resulting in variations in protein expression.
-Bulk proteomics is a technique used to analyze the protein composition of a tissue sample, which contains all types of cells present in that particular tissue.
+Bulk proteomics is a technique used to analyze the protein composition of a sample, which in case of a tissue contains various cell types.
 This approach finds valuable applications for instance in oncology, where it can provide insights into the protein profile of a tumor.
@@
 In fact the expanded library can be used to validate the performance of single-cell type and single-cell proteomics methods and bioinformatics pipeline \citep{Schoof2021}.
-However, in this type of analysis, the precise properties of individual cell types can only be inferred and remain obscured.
+However, in this type of analysis, the precise properties of individual cell types remain obscured.
 
@@
 Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
-To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed \citep{Liou2015}.
+To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) were employed \citep{Liou2015}.
 These techniques primarily rely on the detection of surface proteins on cells and the specific binding of antibodies to these proteins.
-Auto-fluorescence is observed in certain cells, wherein they possess the capability to emit a light signal when stimulated, and this emission can be used to sort them accordingly.
+Auto-fluorescence is observed in certain cells, wherein they possess the capability to emit a light signal when stimulated, and this emission can also be used to sort them accordingly.
 However auto-fluorescence can lack of accuracy as shown in studies indicating the effect of laboratory gloves on cellular samples \citep{Philpott2017}.
-On the other side antibodies, typically derived from the immune system of other species, bind to antigens in a highly specific lock-and-key manner.
+On the other hand, antibodies typically derived from the immune system of other species, bind to antigens in a highly specific lock-and-key manner.
 Antibodies are widely used in molecular biology and find applications in cell isolation.
@@
 ### Single-cell proteomics
-Cell sorting techniques primarily rely on extracellular proteins to identify and separate different cell types.
+Cell sorting techniques primarily rely on surface proteins to identify and separate different cell types.
 However, the intracellular content of cells can still exhibit variability in experiments.
@@
 These methods are still in use to obtain total concentration for sample preparation.
-Another quantitative technique involved RNA-sequencing.
-RNAseq refers to the task of examining the transcriptome of bulk tissue sample  with deep-sequencing technologies.
+Another quantitative technique involved RNA-sequencing (RNAseq).
+RNAseq refers to the task of examining the transcriptome of a sample  with deep-sequencing technologies.
 RNA in a sample includes mRNAs, non-coding RNAs, and small RNAs.
@@
 However, today these predictions can be done with computational methods such as LASSO and the accuracy of the estimate highly depends on selection and validation of this method \citep{Magnusson2022}.
-When observing the protein ensemble of a cellular environment with a top-down approach  like DNAseq or RNAseq, the characteristics (e.g.
+When observing the protein ensemble of a cellular environment with a top-down approach  like RNAseq, the characteristics (e.g.
 post-translational modifications or charge) and local concentration of the proteins remain hidden.
@@
 Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
-Mass spectrometers measure the mass to charge ratio (m/z) of a particle from a fragmented larger molecule.
+Mass spectrometers measure the mass to charge ratio (m/z) of a charged particle from a larger molecule.
 This is achieved by a physical procedure done with a device which is made of three major components: the ion source, an analyzer, and a detector.
-The ion sources charges molecules and accelerate them through a magnetic field.
+The ion sources charges molecules and accelerate them through a magnetic or electric field.
 The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal \citep{Parker2010}.
@@
 After sample preparation the first step involves tryptic digestion.
-It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the protease enzyme trypsin.
+It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the proteolytic enzyme trypsin.
 The process known as proteolysis takes place in multiple parts of multi cellular organisms and was firstly detected in the small intestine digesting proteins \citep{Wang2008}.
@@
 Afterwards a charge remains on the residues making them detectable for the mass spectrometer.
-It is proven that proteins which are digested to the limit, meaning no further proteolysis can be induced, provide the best results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.
+It is proven that proteins which are completely digested, meaning no further proteolysis can be induced, provide the best results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.
 
 To observe multiple samples in one run, labeling is needed to identify each sample within a batch. 
-Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide.
+Since mass spectrometry is not an absolute quantitative technique by itself, the peak height or area only reflects the relative abundance of a peptide.
 Physicochemical properties of the peptides can change the ionization efficiency and detectability of the target.
-However, when comparing the same analyte between multiple runs of labeled peptides, differences in the reporter ion intensity reflect the abundance of those.
+However, when comparing the same analyte between multiple runs of labeled peptides, differences in the reporter ion intensity reflect the relative abundance of those.
 Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the peptide.
 The two techniques for labeling peptides are metabolic labeling and isobaric labling.
-In metabolic labeling the cells were fed with aminoacids containing heavy isotopes, it is the method of choice in order to label peptides at the earliest possible level.
-This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants \citep{Karen2012}.
+In metabolic labeling the cells are fed with aminoacids containing heavy isotopes, it is the method of choice in order to label peptides at the earliest possible level.
+These atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants \citep{Karen2012}.
 Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage.
-Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s.
+Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s \citep{Ong2006}.
 This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles.
-The limitation of this method is, that it can only be applied to a human cell culture (in vitro) or to model organisms (in vivo).
-Isobaric labeling solely applies to in vitro techniques, with TMT and iTRAQ.
+The limitation of this method is, that it can only be applied to a cell culture (in vitro) or to model organisms (in vivo).
+Isobaric labeling solely applies to in vitro techniques, such as TMT and iTRAQ.
 The data analyzed by ProteoScanR primarily utilized TMT tags.
@@
 This technique allows for the identification, enrichment, and quantification of low-abundance peptide ions, which is particularly valuable in single-cell techniques.
-With this technique it is possible to quantify proteins and differ low abundant proteins from background noise.
+With this technique it is possible to quantify proteins and discriminate low abundant proteins from background noise.
 However, a drawback of isobaric labeling is the presence of co-fragmentation signals in the spectrogram, which requires data normalization to remove unwanted contributions \citep{Marx2019, Budnik2018}.
-Isotopic distribution in TMTs, reflecting the natural distribution, can be corrected during data acquisition by considering a defined spread in other channels.
-Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species.
-Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell \citep{Ye2022}.
+Isotopic distribution in TMTs, reflecting the impurities during production, can be corrected during data processing by considering a defined spread in other channels for each indivdiual batch of TMTpro reagent.
+Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species \citep{Ye2022}.
 
@@
 The principle behind RPLC is a column with a nonpolar stationary phase which interacts with the nonpolar residues of the peptides.
-Peptides were fractionated according to their hydrophobicity \citep{Pitt2009}.
+Peptides are fractionated according to their hydrophobicity \citep{Pitt2009}.
 
@@
 Peptides are further fragmented according to their chemical properties and can be further handled in the mass spectrometer.
-The fragmented peptides are now in charged droplets separated based on their surface charge, splitting further into smaller droplets until they reach the gas phase as ions.
+The fragmented peptides are now located in charged droplets separated based on their surface charge, splitting further into smaller droplets until they reach the gas phase as ions.
 Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM).
@@
 
-Subsequently after the ionization peptides are accelerated through the magnetic field generated by the mass analyzer.
+Subsequently after the ionization peptides are accelerated through the magnetic or electric field generated by the mass analyzer.
 Six general types of mass analyzers are availaible on the market, there are Quadropole, time of flight (TOF), magnetic sector, electrostatic sector, quadropole ion trap and ion cyclotron resonance -mass analyzer.
@@
 Ion trapping enables the technique of coupled mass spectrometers, where ions were further fragmented by collision induced dissociation and analyzed subsequently.
-The detector transforms the impact energy of the ion into an electric signal, called spectrum which can be processed computationally.
+The detector transforms the impact (TOF) or oscillation (Orbitrap) energy of the ion into an electric signal, called spectrum which can be processed computationally.
 
@@
 The availability of gene sequences in databases and the computational matching of peptides against those sequences enable the identification of alterations in a sample at the protein level.
-These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylations or methylations \citep{Aebersold2003}.
+These alterations can rely on the sequence level or could be due to post-translational modifications (PTMs) such as phosphorylations or methylations \citep{Aebersold2003}.
 The peptide identification (P-) score reflects the goodness of fit between the data and the identified sequence in the database, taking into account the length of the peptide.

### 2023-07-19 markdown/methods.Rmd
thesis finalized

@@
 In this case, group differences influence the batch effect reducing statistical power after the correction.
-These both effects are so called interdependent.
+Both of these effects are called interdependent.
 The opposite could happen if the experimental design is heavily unbalanced leading to group differences induced by batch correction.
@@
 After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained.
-Proteomic data consists of multi dimensions (=features) which are overall hard to understand by a glance.
-An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keep a high level of information can be dimensionality reduction.
-After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor.
+Proteomic data consists of multiple dimensions (=features) which are overall hard to understand at a glance.
+An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keeping a high level of information can be dimensionality reduction.
+After the cleaning procedures any additional effect caused by technical or biological replication should be removed in advance (e.g.
+batch correction) or observed in particular with this technique.
 This makes the dimensionality reduction a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.
@@
 #### Principle component analysis (PCA)
-The most common approach in data science for any kind of high dimensional data is the PCA.
+The most common approach for dimensionality reduction in data science for any kind of high dimensional data is the PCA.
 For ProteoScanR the bioconductor package scater \citep{McCarthy2017} was used.
@@
 Since $A v = \lambda v$ the eigenvector $v$ can be obtained by solving the equation for each element of the vector.
-The eigenvectors $v$ never changes is direction and explains the covariance within the dataset and so also the mathematical differences between the samples.
+The eigenvectors $v$ never changes direction and explains the covariance within the dataset and therefore also the mathematical differences between the samples.
 The eigenvectors are sorted according to their eigenvalues in decreasing order and therefore decreasing explained variance.
@@
 Riemann geometry is a mathematical framework that deals with three dimensional functions using volumes, areas and vectors, which are also concept of topological data analysis.
-Before performing UMAP, three assumptions need to be considered:
+Before performing UMAP, three assumptions need to be considered \citep{McInnes2018}:
 
@@
 \end{enumerate}
-\citep{McInnes2018}
 
@@
 One approach is approximating $\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}$ and deriving the p-value by the area under the density curve of the $\chi^2$ distribution $Q = \sum_{i=1}^{k} Z_{i}^{2}$.
-In the distribution $k$ is defined by the degrees of freedom (=df), what can be obtained by $k=(columns-1)*(rows-1)=df$ of the table.
+In the distribution $k$ is defined by the degrees of freedom (=df), which can be obtained by $k=(columns-1)*(rows-1)=df$ of the table.
 Another approach to obtain an exact p-value would be Fisher`s exact test, which states that the margins are distributed according to the hypergeometric distributions.

### 2023-07-27 markdown/article.Rmd
article started

--- 2023-06-26 markdown/Thesis.Rmd
@@
+---
+title: "article"
+author: "Lukas Gamp"
+date: "`r Sys.Date()`"
+output: pdf_document
+---
 ---
@@
 ```
-
-\allsectionsfont{\centering}
-\subsectionfont{\raggedright}
-\subsubsectionfont{\raggedright}
-
-\pagenumbering{gobble}
-
-\begin{centering}
-
-\vspace{3cm}
-```{r uni_logo, echo=F, out.width="20%"}
-knitr::include_graphics("FHCW_logo.png")
-knitr::include_graphics("KI_logo.png")
-```
-
-\vspace{1cm}
-
-\Large
-\doublespacing
-{\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
-\\ presenting ProteoScanR \& Proteomics Workbench
-
-\vspace{1 cm}
-
-\normalsize
-\singlespacing
-
-\vspace{0.5 cm}
-
-\Large
-
-{\bf Lukas Gamp}
-\\ 01630003
-
-\vspace{1.5 cm}
-
-In partial fulfillment of the requirement \\for the degree of MSc \\in Bioinformatics
-
-\vspace{1.5 cm}
-
-\normalsize
-July 2023
-
-\doublespacing
-\doublespacing
-Main Supervisor: Dr. DI(FH) Gerhard Duernberger \\
-External Supervison: Assoc. Prof. Ujjwal Neogi, M.Sc. PhD \\
-Second examiner: FH-Prof. Dr. Alexandra Graf \\
-
-\end{centering}
-
-\newpage
-
-\pagenumbering{gobble}
-
-\begin{centering}
-
-{\bf Abstract}
-
-\end{centering}
-
-\spacing{1.5}
-
-With the growing prominence of single-cell techniques across various omics fields, there is a pressing need to develop a standardized pipeline for proteomics data in the realm of systems biology.
-Unlike DNA sequencing and RNA sequencing, proteomics analysis via mass spectrometry incurs high costs in terms of labor and equipment.
-Additionally, commercially available software solutions often come with hefty price tags and limited transparency regarding the underlying methods employed.
-However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility using the R programming language.
-Introducing ProteoScanR, a state-of-the-art proteomics pipeline integrated into the user-friendly Proteomics Workbench interface. 
-Guided by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of ProteoScanR was thoroughly tested and validated using both bulk and single-cell mass spectrometry data sets.
-The pipeline implementation in the Proteomics Workbench leverages the power of an interactive environment built in R Shiny, empowering users to discover valuable insights for their specific data set.
-Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction.
-ProteoScanR and the Proteomics Workbench served as a valuable tool in guiding the identification of significantly expressed proteins in cells under study.
-Subsequently, the pathway enrichment analysis provided additional biological contexts for a comprehensive understanding of their functional implications.
-The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics.
-Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.
-https://github.com/Lukas67/ProteoScanR
-
-\pagenumbering{roman}
-
-\newpage
-
-\centering
-\raggedright
-\newpage
-\tableofcontents
-
-\newpage
-
-```{r acknowledgments, child='acknowledgments.Rmd'}
-```
-\pagenumbering{arabic}
-
-\newpage
-
-```{r intro, child = 'introduction.Rmd'}
-```
-
-\newpage
-
-```{r methods, child = 'methods.Rmd'}
-```
-
-\newpage
-
-```{r results, child = 'results.Rmd'}
-```
-
-\newpage
-
-```{r discussion, child = 'discussion.Rmd'}
-```
-
-\newpage
-
-```{r conclusion, child = 'conclusion.Rmd'}
-```
-
-\newpage
-
-```{r biblio, child = 'bibliography.Rmd'}
-```
-
-\newpage
-
-```{r appendix, child = 'appendix.Rmd'}
-```
-
-\newpage
-\listoffigures

### 2023-07-27 markdown/article.Rmd
article half finished

--- 2023-07-19 markdown/Thesis.Rmd
@@
     toc: false
-#    number_sections: TRUE
 geometry: "left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm"
@@
 
+```{r header, echo=F, out.width="100%"}
+knitr::include_graphics("fh_header.jpg")
+```
+
 \allsectionsfont{\centering}
@@
 \begin{centering}
-
-\vspace{3cm}
-```{r uni_logo, echo=F, out.width="20%"}
-knitr::include_graphics("FHCW_logo.png")
-knitr::include_graphics("KI_logo.png")
-```
-
-\vspace{1cm}
 
@@
 
+```{r author_foto, echo=F}
+knitr::include_graphics("Lukas_Gamp_grey.png")
+```
+
 {\bf Lukas Gamp}
-\\ 01630003
+\vspace{3 cm}
 
-\vspace{1.5 cm}
-
-In partial fulfillment of the requirement \\for the degree of MSc \\in Bioinformatics
-
-\vspace{1.5 cm}
-
-\normalsize
-July 2023
-
-\doublespacing
-\doublespacing
 Main Supervisor: Dr. DI(FH) Gerhard Duernberger \\
@@
 \newpage
-
-\pagenumbering{gobble}
+\pagenumbering{arabic}
 
@@
 
-\pagenumbering{roman}
+\newpage
+# Introduction
+The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins \citep{Cobb2017}.
+Over the past six decades, researchers have explored various omics fields, such as genomics, transcriptomics, proteomics, and metabolomics, to comprehensively analyze biomolecules in diverse contexts.
+Omics research offers a holistic view of biological processes and regulatory mechanisms, and identify key players in complex biological networks using computational methods.
+Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks.
+Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role within this interactive structure.
+Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}.
 
-\newpage
+The central dogma of molecular biology offers insights not only at the level of individual terms but also provides a hierarchical understanding of biology.
+Hierarchical organization is the underlying principle in all topics of biology.
+Proteins exhibit hierarchical structures that contribute to their functionality.
+At the primary structure level, proteins are composed of smaller building blocks called amino acids.
+Combining them results in the protein secondary structure such as alpha helix and beta sheet.
+These secondary structures further assemble to create higher-order tertiary structures, which represent specific structural domains within the protein.
+Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
+Proteomics is a comprehensive approach that investigates the complete protein composition of a specimen, aiming to understand the intricate biological network it represents.
+For instance if pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
+These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications \citep{Beltran2017}.
 
-\centering
-\raggedright
-\newpage
-\tableofcontents
+Higher organisms are composed of specialized cells organized into tissues, such as skin, muscle, and blood.
+Each tissue consists of cells with specific functions, resulting in variations in protein expression.
+Bulk proteomics is a technique used to analyze the protein composition of a sample, which in case of a tissue contains various cell types.
+Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
+To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) were employed.
+The rise and validation of cell sorting techniques enabled targeted single-cell proteomics (SCP) \citep{Liou2015, Sutermaster2019}.
+Single-cell proteomics reflects the protein ensemble of a specific cell type at a particular time, providing a focused perspective on the studied field compared to bulk methods \citep{Maes2020}.
 
-\newpage
+Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
+Mass spectrometers measure the mass to charge ratio (m/z) of a charged particle from a larger molecule.
+This is achieved by a physical procedure done with a device which is made of three major components: the ion source, an analyzer, and a detector.
+The ion sources charges molecules and accelerate them through a magnetic or electric field.
+The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal \citep{Parker2010}.
+However before acquiring data, sample preparation and pre-processing is required.
+Sample preparation involves cell sorting and protein isolation.
+After sample preparation the first step involves tryptic digestion.
+It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the proteolytic enzyme trypsin.
+The process known as proteolysis takes place in multiple parts of multi cellular organisms and was firstly detected in the small intestine digesting proteins \citep{Wang2008}.
 
-```{r acknowledgments, child='acknowledgments.Rmd'}
-```
-\pagenumbering{arabic}
+To observe multiple samples in one run, labeling is needed to identify each sample within a batch.
+The two techniques for labeling peptides are metabolic labeling and isobaric labling.
+The data analyzed by ProteoScanR primarily utilized TMT tags.
 
-\newpage
+Before ionization peptides need to be further separated according to their chemical properties, size or species by liquid chromatography (LC), because the number of peptide processed simultaneously by the MS device is limited.
 
-```{r intro, child = 'introduction.Rmd'}
+To analyze a biological sample consisting of peptides in solution the liquid needs to be vaporized into gas phase.
+Electrospray ionization (ESI) pushes the analyte through a capillary while applying an electric current to the liquid, vaporizing the sample to a charged aerosol.
+Subsequently after the ionization peptides are accelerated and separated according to their mass to charge ratio (m/z) through the magnetic or electric field generated by the mass analyzer.
+Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their m/z consecutively.
+
+Given the high resolution of MS data, algorithms are employed to convert the raw signal into an interpretable form.
+Software packages like MaxQuant \citep{Cox2008} are commonly used to process the data, providing it for further analysis and statistical testing.
+The data obtained from mass spectrometry has three dimensions: m/z ratio, intensity, and retention time.
+To separate peaks from each other, algorithms are used to identify local minima in the data.
+The centroid of each peak is determined by fitting a Gaussian peak shape, which can be interpreted as locating the peaks of each m/z spectrum as a function of retention time.
+The centroid of a peak corresponds to an isotope.
+
+With advancements in mass spectrometry technology and computational methods, proteomic analysis has emerged as a powerful tool for investigating complex protein samples.
+However, analyzing proteomic data poses challenges in data processing, statistical analysis, and interpretation.
+This thesis aims to address these challenges by exploring computational methods for downstream analysis of proteomics data.
+It is important to acknowledge that downstream analysis in proteomics lacks a standardized approach, and the selection of computational methods depends on the specific dataset and research objectives.
+ProteoScanR comprises a series of steps to streamline the data in an interactive environment, called Proteomics Workbench.
+This user-friendly interface will enable researchers, including those with limited computational expertise, to navigate and comprehend the data effectively.
+
+# Materials and methods
+```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="75%", fig.align='center', fig.cap= "Flowchart presenting the processing by ProteoScanR.
+Gold: techniques which involve explanatory data analysis.
+Yellow: Adjustable features.
+Blue: Inputs"}
+library(Gmisc, quietly = TRUE)
+library(glue)
+library(htmlTable)
+library(grid)
+library(magrittr)
+
+default_txt_setting <- gpar(cex=1.9)
+default_box_setting <- gpar(fill="white")
+
+data_box_setting <- gpar(fill="#56B4E9")
+
+opts_box_setting <- gpar(fill="lightyellow", col="darkblue")
+opts_txt_setting <- gpar(col="darkblue", cex=1.5)
+refinement_arrow_setting <- arrow(angle=30, ends="both", type = "closed")
+
+plot_box_setting <- gpar(fill="#DDCC77")
+# input files
+evidence_file <- boxGrob("evidence file",
+                         box_gp = data_box_setting,
+                         txt_gp = default_txt_setting)
+sample_annotation_file <- boxGrob("sample annotation",
+                                  box_gp = data_box_setting,
+                                  txt_gp = default_txt_setting)
+
+read_data <- boxGrob(glue("read data into q-feature object",.sep = "\n"),
+                     box_gp = default_box_setting,
+                     txt_gp = default_txt_setting)
+
+quality_control_1 <- boxGrob(glue("quality control 1",.sep = "\n"),
+                             box_gp = default_box_setting,
+                             txt_gp = default_txt_setting)
+
+qc1_content <- boxGrob("filter out pool samples\nreplace zeros with NA\nfilter out contaminants\nfilter by q-value cutoff",
+                      box_gp = opts_box_setting,
+                      txt_gp = opts_txt_setting,
+                      just="center")
+
+aggregate_psms <-  boxGrob(glue("aggregate peptide spectrum matches to peptides\n
+                                 join multiple batches", .sep = "\n"),
+                           box_gp = default_box_setting,
+                           txt_gp = default_txt_setting)
+
+quality_control_2 <- boxGrob("quality control 2",
+                             box_gp = plot_box_setting,
+                             txt_gp = default_txt_setting)
+
+qc2_content <- boxGrob("filter by:\nreporter ion intensity cutoff\npepitde covariance to razor proteins\npeptide missing rate",
+                       box_gp = opts_box_setting,
+                       txt_gp = opts_txt_setting,
+                       just="center")
+
+aggregate_peps <- boxGrob(glue("aggregate peptides to proteins", .sep = "\n"),
+                          box_gp = default_box_setting,
+                          txt_gp = default_txt_setting)
+
+transform_data <- boxGrob("transform data",
+                          box_gp = plot_box_setting,
+                          txt_gp = default_txt_setting)
+
+transform_opts <- boxGrob("options:\nlog2\nlog10\nsqrt\nquadratic\nBoxCox\nNone",
+                          box_gp = opts_box_setting,
+                          txt_gp = opts_txt_setting,
+                          just="center")
+
+normalize_data <- boxGrob("normalize data",
+                          box_gp = plot_box_setting,
+                          txt_gp = default_txt_setting)
+
+normalize_opts <- boxGrob("options:\ncol-Median,row_Mean\nCONSTANd\nNone",
+                          box_gp = opts_box_setting,
+                          txt_gp = opts_txt_setting,
+                          just="center")
+
+missing_value_handling <- boxGrob("missing value handling",
+                                  box_gp = plot_box_setting,
+                                  txt_gp = default_txt_setting)
+
+missing_value_opts <- boxGrob("options:\nKNN\ndrop rows\nreplace with mean\nreplace with median\nreplace with 0",
+                              box_gp = opts_box_setting,
+                              txt_gp = opts_txt_setting,
+                              just="center")
+
+batch_correction <- boxGrob("batch correction & dimensionality reduction",
+                            box_gp=plot_box_setting,
+                            txt_gp = default_txt_setting)
+
+batch_correction_opts <- boxGrob("ComBat\nPCA\nUMAP",
+                                 box_gp = opts_box_setting,
+                                 txt_gp = opts_txt_setting,
+                                 just="center")
+
+dim_red <- boxGrob("dimensionality reduction",
+                   box_gp = plot_box_setting,
+                   txt_gp = default_txt_setting)
+
+dim_red_actions <- boxGrob("procedures:\nPCA\nUMAP",
+                       box_gp = opts_box_setting,
+                       txt_gp = opts_txt_setting,
+                       just="center"
+                       )
+
+stat_module <- boxGrob(glue("differential expression\npathway enrichment\nontology enrichment",.sep = "\n"),
+                       box_gp = default_box_setting,
+                       txt_gp = default_txt_setting)
+
+grid.newpage()
+vert <- {spreadVertical(files = evidence_file,
+                       read_data = read_data,
+                       quality_control_1 = quality_control_1,
+                       aggregate_psms,
+                       quality_control_2 = quality_control_2,
+                       aggregate_peps,
+                       transform_data = transform_data,
+                       normalize_data = normalize_data,
+                       missing_value_handling = missing_value_handling,
+#                       batch_correction = batch_correction,
+#                       dim_red = dim_red,
+                       out = stat_module
+)}
+input <- alignVertical(reference = vert$files,
+                      evidence_file, sample_annotation_file) %>%
+  spreadHorizontal()
+vert$files <- NULL
+
+output <- alignVertical(reference = vert$out,
+                      stat_module, batch_correction) %>%
+  spreadHorizontal()
+vert$out <- NULL
+
+# right side
+qc1_content <- moveBox(qc1_content,
+                    x = .8,
+                    y = coords(vert$quality_control_1)$y)
+
+transform_opts <- moveBox(transform_opts,
+                    x = .9,
+                    y = coords(vert$transform_data)$y)
+
+# dim_red <- moveBox(dim_red,
+#                    x = coords(output[[2]])$x,
+#                    y = 0)
+
+batch_correction_opts <- moveBox(batch_correction_opts,
+                    x = coords(output[[2]])$x,
+                    y = coords(vert$missing_value_handling)$y)
+
+# dim_red_actions <- moveBox(dim_red_actions,
+#                     x = (coords(dim_red)$x -.5),
+#                     y = coords(dim_red)$x)
+
+# left side
+qc2_content <- moveBox(qc2_content,
+                    x = .17,
+                    y = coords(vert$quality_control_2)$y)
+
+normalize_opts <- moveBox(normalize_opts,
+                    x = .2,
+                    y = coords(vert$normalize_data)$y)
+
+missing_value_opts <- moveBox(missing_value_opts,
+                    x = .2,
+                    y = coords(vert$missing_value_handling)$y)
+
+for (i in 1:(length(vert) - 1)) {
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+    print
+}
+
+# connect boxes
+connectGrob(input[[1]], vert$read_data, type = "N")
+connectGrob(input[[2]], vert$read_data, type = "N")
+
+connectGrob(vert$quality_control_1, qc1_content, type = "horizontal", arrow_obj = arrow(angle=0))
+
+connectGrob(vert$quality_control_2, qc2_content, type = "horizontal", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$transform_data, transform_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$normalize_data, normalize_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$missing_value_handling, missing_value_opts, type = "horizontal", arrow_obj = refinement_arrow_setting)
+
+connectGrob(vert$missing_value_handling, output[[1]], type = "N")
+connectGrob(vert$missing_value_handling, output[[2]], type = "N")
+
+connectGrob(output[[2]], batch_correction_opts, type = "N", arrow_obj = arrow(angle=0), subelmnt = "r")
+
+# connectGrob(vert$dim_red, dim_red_actions, type = "horizontal", arrow_obj = arrow(angle=0))
+
+# Print boxes
+input
+vert
+qc1_content
+qc2_content
+transform_opts
+normalize_opts
+missing_value_opts
+batch_correction_opts
+output
+# dim_red
+# dim_red_actions
 ```
@@
 \newpage
+The flowchart (figure \@ref(fig:data-processing-pipeline-flowchart-vertical)) describes the processes involved in the ProteoScanR pipeline. 
+ProteoScanR initially filters the data by false discovery rate (FDR) converted to q-value (provided by MaxQuant \citep{Cox2011}), according to precursor ion fraction (PIF) \citep{Tannous2020, Specht2021} and subsequently aggregates peptide spectrum matches to peptides.
+The quality control continues with calculating median reporter ion intensity (RI), median coefficient of variation (CV) and filtering according to custom cutoffs.
+Subsequently peptides with high missing rate are removed and peptides are aggregated to proteins.
+Preparation for hypothesis testing involves data transformation and normalization.
+Missing values for proteins in individual channels are handled according to user selection with methods such as the K-nearest neighbor algorithm (=KNN) \citep{Lan2013}.
+Before statistical analysis, dimensionality reduction techniques can be applied and observed with or without correcting for batch effect with ComBat \citep{Johnson2007}.
+Furthermore data processing can be validated with an entropy based approach for the conservation of information.
+Statistical testing involves differential expression analysis with the R package Limma \citep{Phipson2016}.
+Hypothesizes can be selected by the user and p-value correction can be employed with different approaches to meet conservative as well as liberal study designs.
+After finding differentially expressed proteins, enrichment elucidates the biological context.
+The protein enrichment analysis can be employed in selected contexts such as pathways, cell-types or custom ontologies.
+The pathway enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}.
+The analysis utilizes the entire “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways.
+For custom ontologies the protein enrichment analysis is done using the R bioconductor package piano \citep{Vaeremo2013}.
+In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
+The online database offers pre-built gene set collections for previous research questions in multiple fields of human biology.
 
-```{r methods, child = 'methods.Rmd'}
+\newpage
+# Results
+```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
+knitr::include_graphics("screenshots/numbers_default.png")
+```
+
+The ProteoScanR pipeline is employed in the proteomics workbench.
+Users find settings for the pre-processing on the left hand side of the interface (see figure \@ref(fig:ui-numbers-default)).
+The main panel shows the data in different aspects and helps the user finding a good fit for the methods applied to their individual dataset.
+
+\newpage
+```{r ui-stat-HC-HIV-noMets, echo=F, out.width="70%", fig.align='center', fig.cap="Statistics module"}
+knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
 ```
@@
 \newpage
-
-```{r results, child = 'results.Rmd'}
-```
-
-\newpage
-
-```{r discussion, child = 'discussion.Rmd'}
-```
-
-\newpage
-
-```{r conclusion, child = 'conclusion.Rmd'}
-```
-
-\newpage
-
 ```{r biblio, child = 'bibliography.Rmd'}
 ```
-
-\newpage
-
-```{r appendix, child = 'appendix.Rmd'}
-```
-
-\newpage
-\listoffigures

### 2023-07-31 markdown/article.Rmd
article updated

@@
     toc: false
-geometry: "left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm"
+geometry: "left = 2cm, right = 1.5cm, top = 1.5cm, bottom = 1.5cm"
 fontsize: 11pt
@@
 
+\vspace{1 cm}
+
+\begin{centering}
+```{r uni_logo, echo=F, out.width="20%", fig.align='center'}
+knitr::include_graphics("KI_logo.png")
+```
+This project was performed at the systems virology lab at the Karolinska Institutet in Stockholm.
+\end{centering}
+
 \newpage
@@
 Omics research offers a holistic view of biological processes and regulatory mechanisms, and identify key players in complex biological networks using computational methods.
-Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks.
-Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role within this interactive structure.
 Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}.
-
-The central dogma of molecular biology offers insights not only at the level of individual terms but also provides a hierarchical understanding of biology.
-Hierarchical organization is the underlying principle in all topics of biology.
-Proteins exhibit hierarchical structures that contribute to their functionality.
-At the primary structure level, proteins are composed of smaller building blocks called amino acids.
-Combining them results in the protein secondary structure such as alpha helix and beta sheet.
-These secondary structures further assemble to create higher-order tertiary structures, which represent specific structural domains within the protein.
-Finally, the quaternary structure describes the functional state of the protein as a whole at a given time.
 Proteomics is a comprehensive approach that investigates the complete protein composition of a specimen, aiming to understand the intricate biological network it represents.
-For instance if pathogens interact with cells of an organism a cascade of biochemical reactions takes place.
-These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications \citep{Beltran2017}.
 
@@
 Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types.
-To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) were employed.
-The rise and validation of cell sorting techniques enabled targeted single-cell proteomics (SCP) \citep{Liou2015, Sutermaster2019}.
+To overcome this limitation, cell sorting techniques were employed and enabled targeted single-cell proteomics (SCP) \citep{Liou2015, Sutermaster2019}.
 Single-cell proteomics reflects the protein ensemble of a specific cell type at a particular time, providing a focused perspective on the studied field compared to bulk methods \citep{Maes2020}.
@@
 Mass spectrometers measure the mass to charge ratio (m/z) of a charged particle from a larger molecule.
-This is achieved by a physical procedure done with a device which is made of three major components: the ion source, an analyzer, and a detector.
-The ion sources charges molecules and accelerate them through a magnetic or electric field.
-The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal \citep{Parker2010}.
-However before acquiring data, sample preparation and pre-processing is required.
-Sample preparation involves cell sorting and protein isolation.
-After sample preparation the first step involves tryptic digestion.
-It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the proteolytic enzyme trypsin.
-The process known as proteolysis takes place in multiple parts of multi cellular organisms and was firstly detected in the small intestine digesting proteins \citep{Wang2008}.
-
-To observe multiple samples in one run, labeling is needed to identify each sample within a batch.
-The two techniques for labeling peptides are metabolic labeling and isobaric labling.
-The data analyzed by ProteoScanR primarily utilized TMT tags.
-
-Before ionization peptides need to be further separated according to their chemical properties, size or species by liquid chromatography (LC), because the number of peptide processed simultaneously by the MS device is limited.
-
-To analyze a biological sample consisting of peptides in solution the liquid needs to be vaporized into gas phase.
-Electrospray ionization (ESI) pushes the analyte through a capillary while applying an electric current to the liquid, vaporizing the sample to a charged aerosol.
-Subsequently after the ionization peptides are accelerated and separated according to their mass to charge ratio (m/z) through the magnetic or electric field generated by the mass analyzer.
-Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their m/z consecutively.
-
 Given the high resolution of MS data, algorithms are employed to convert the raw signal into an interpretable form.
 Software packages like MaxQuant \citep{Cox2008} are commonly used to process the data, providing it for further analysis and statistical testing.
-The data obtained from mass spectrometry has three dimensions: m/z ratio, intensity, and retention time.
-To separate peaks from each other, algorithms are used to identify local minima in the data.
-The centroid of each peak is determined by fitting a Gaussian peak shape, which can be interpreted as locating the peaks of each m/z spectrum as a function of retention time.
-The centroid of a peak corresponds to an isotope.
 
@@
 
-qc2_content <- boxGrob("filter by:\nreporter ion intensity cutoff\npepitde covariance to razor proteins\npeptide missing rate",
+qc2_content <- boxGrob("filter by:\nreporter ion intensity cutoff\npepitde coefficient of variation to razor proteins\npeptide missing rate",
                        box_gp = opts_box_setting,
@@
 ProteoScanR initially filters the data by false discovery rate (FDR) converted to q-value (provided by MaxQuant \citep{Cox2011}), according to precursor ion fraction (PIF) \citep{Tannous2020, Specht2021} and subsequently aggregates peptide spectrum matches to peptides.
-The quality control continues with calculating median reporter ion intensity (RI), median coefficient of variation (CV) and filtering according to custom cutoffs.
+
+The quality control continues with calculating median reporter ion intensity (RI), median coefficient of variation (CV) and filtering according to cutoffs.
 Subsequently peptides with high missing rate are removed and peptides are aggregated to proteins.
+
 Preparation for hypothesis testing involves data transformation and normalization.
 Missing values for proteins in individual channels are handled according to user selection with methods such as the K-nearest neighbor algorithm (=KNN) \citep{Lan2013}.
-Before statistical analysis, dimensionality reduction techniques can be applied and observed with or without correcting for batch effect with ComBat \citep{Johnson2007}.
+Before statistical analysis, dimensionality reduction techniques can be applied and observed with or without correcting for batch effect by ComBat \citep{Johnson2007}.
 Furthermore data processing can be validated with an entropy based approach for the conservation of information.
+
 Statistical testing involves differential expression analysis with the R package Limma \citep{Phipson2016}.
 Hypothesizes can be selected by the user and p-value correction can be employed with different approaches to meet conservative as well as liberal study designs.
-After finding differentially expressed proteins, enrichment elucidates the biological context.
+
+After finding differentially expressed proteins, enrichment elucidates them in biological context by finding over represented proteins.
 The protein enrichment analysis can be employed in selected contexts such as pathways, cell-types or custom ontologies.
@@
 In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
-The online database offers pre-built gene set collections for previous research questions in multiple fields of human biology.
 
@@
 
+As seen in figure \@ref(fig:ui-numbers-default) the aspects shown in the main panel include:
+\begin{scriptsize}
+\begin{itemize}
+\item Overview showing the individual processing steps of the analysis pipeline
+\item Barchart showing the number of peptide spectrum matches, peptides and proteins
+\item Mutual information intra- and inter- groupwise
+\item Reporter ion intensity with factor selection to highlight
+\item Median coefficient of variation for razor proteins and correlation heatmap for protein expression
+\item Dimensionality reduction analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP)
+\item Feature wise output for an individual protein over the course of the pipeline.
+\item Statistics module
+\item Pathway enrichment
+\item Protein set ontology enrichment 
+\end{itemize}
+\end{scriptsize}
+
 \newpage
-```{r ui-stat-HC-HIV-noMets, echo=F, out.width="70%", fig.align='center', fig.cap="Statistics module"}
+```{r ui-mutual-info-quant-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with quantile normalization"}
+knitr::include_graphics("screenshots/mutual_info_quant_norm.png")
+```
+
+Beforehand normalization was tested with applying logarithmic transformation to the base 2 and performing column median row mean normalization (not shown in figure).
+The mutual information (MI) within the healthy control (HC) group's sample type exhibited a reduction towards the lower edge of the interquartile range (IQR).
+
+After performing a logarithmic transformation to the base 2 and quantile normalization, the mutual information (MI) within the healthy control (HC) group's sample type exhibits a minor reduction in the interquartile range (IQR) (as seen in figure \@ref(fig:ui-mutual-info-quant-norm)).
+However, it should be noted that there is an increase in the number of outliers compared to the raw data.
+When switching to the quantile normalization the loss of mutual information between sample types decreased (as seen in figure \@ref(fig:ui-mutual-info-quant-norm)).
+Therefore the advised normalization technique for this particular data set would be the quantile normalization compared to the default method.
+
+\newpage
+```{r ui-stat-HC-HIV-noMets, echo=F, out.width="70%", fig.align='center', fig.cap="Statistics module in portrait mode"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 \newpage
+By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group (see figure \@ref(fig:ui-stat-HC-HIV-noMets)), six differentially expressed proteins were identified.
+The Venn diagram indicates significantly up/down regulated proteins.
+The Volcano represents -log10(adjusted p values) against log2 fold changes (logFC).
+These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
+The remaining 652 proteins have been identified as not significantly up/down regulated in the comparison analysis.
+These proteins did not exhibit statistically significant differences in expression between the compared groups or conditions.
+
+# Discussion and conclusion
+The Proteomics Workbench interface and the ProteoScanR pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
+By utilizing an entropy-based visualization approach, the conservation of information can be validated, allowing users to select appropriate methods and adjust thresholds, cutoffs, and techniques accordingly.
+With the selection of factors to indicate over multiple plots, biases can be identified on various levels.
+Additionally, the analysis can be examined on an individual protein basis, enabling the identification of suspicious results and closer validation of significant candidates.
+The statistics module provides users with a simple tool to assess and visualize the results, offering clarity to dense scatter plots through hover functions.
+Enrichment mapping links statistically significant proteins to either the KEGG pathway database or individual research data sets, depending on user selection.
+These final results place the data in a biological context, which can be explored interactively in graphical visualizations.
+
+In conclusion, the ProteoScanR pipeline and Proteomics Workbench interface provide a user-friendly and efficient approach for proteomic data analysis, enabling biologists to address their research questions and validate computations.
+Future perspectives include deploying the application on a server for global accessibility, leveraging artificial intelligence techniques, and implementing clustering approaches to further enhance annotation and understanding of cellular heterogeneity and functional characteristics.
+
+\newpage
 ```{r biblio, child = 'bibliography.Rmd'}
 ```
+
+\newpage
+\listoffigures

### 2023-07-31 markdown/article.Rmd
article first draft finished

@@
 The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics.
-Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application
+Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.
 The source code is accessible and can be located at the following URL: https://github.com/Lukas67/ProteoScanR
@@
 Missing values for proteins in individual channels are handled according to user selection with methods such as the K-nearest neighbor algorithm (=KNN) \citep{Lan2013}.
-Before statistical analysis, dimensionality reduction techniques can be applied and observed with or without correcting for batch effect by ComBat \citep{Johnson2007}.
-Furthermore data processing can be validated with an entropy based approach for the conservation of information.
+Before statistical analysis, dimensionality reduction techniques were applied and observed with and without correcting for batch effect by ComBat \citep{Johnson2007}.
+Furthermore data processing is validated with an entropy based approach for the conservation of information.
 
@@
 After finding differentially expressed proteins, enrichment elucidates them in biological context by finding over represented proteins.
-The protein enrichment analysis can be employed in selected contexts such as pathways, cell-types or custom ontologies.
+The protein enrichment analysis is employed in selected contexts such as pathways, cell-types or custom ontologies.
 The pathway enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}.
 The analysis utilizes the entire “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways.
-For custom ontologies the protein enrichment analysis is done using the R bioconductor package piano \citep{Vaeremo2013}.
-In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
+For custom ontologies the protein enrichment analysis undertaken with the R bioconductor package piano \citep{Vaeremo2013}.
+In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which is either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
 
@@
 As seen in figure \@ref(fig:ui-numbers-default) the aspects shown in the main panel include:
+\spacing{1}
 \begin{scriptsize}
 \begin{itemize}
-\item Overview showing the individual processing steps of the analysis pipeline
-\item Barchart showing the number of peptide spectrum matches, peptides and proteins
-\item Mutual information intra- and inter- groupwise
-\item Reporter ion intensity with factor selection to highlight
-\item Median coefficient of variation for razor proteins and correlation heatmap for protein expression
-\item Dimensionality reduction analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP)
-\item Feature wise output for an individual protein over the course of the pipeline.
-\item Statistics module
-\item Pathway enrichment
-\item Protein set ontology enrichment 
+\item \textbf{Overview:} Summary showing the individual processing steps of the analysis pipeline
+\item \textbf{Summary Barplot:} Barchart showing the number of peptide spectrum matches, peptides and proteins
+\item \textbf{Normalization validation:} Mutual information intra- and inter- groupwise
+\item \textbf{Reporter ion intensity:} Includes factor selection to create boxplots indicating a factor
+\item \textbf{Median coefficient of variation and correlation:} Variation of peptides for razor proteins and correlation heatmap for protein expression
+\item \textbf{Dimensionality reduction:} Analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP)
+\item \textbf{Feature wise output:} Visualization of an individual protein over the course of the pipeline.
+\item \textbf{Statistics module:} Differential expression analysis
+\item \textbf{Pathway enrichment:} Over-representation analysis for differentially expressed proteins in the context of pathways.
+\item \textbf{Protein set ontology enrichment:}  Over-representation analysis for differentially expressed proteins in a user selected context.
 \end{itemize}
 \end{scriptsize}
+\spacing{1.5}
 

### 2023-08-15 markdown/article.Rmd
suggestions from supervisor implemented in article
article shortened significantly

ready to implement beamer blocks

@@
 # Introduction
-The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins \citep{Cobb2017}.
-Over the past six decades, researchers have explored various omics fields, such as genomics, transcriptomics, proteomics, and metabolomics, to comprehensively analyze biomolecules in diverse contexts.
-Omics research offers a holistic view of biological processes and regulatory mechanisms, and identify key players in complex biological networks using computational methods.
+The central dogma underpins genetic information's transformation into functional proteins \citep{Cobb2017}.
 Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}.
-Proteomics is a comprehensive approach that investigates the complete protein composition of a specimen, aiming to understand the intricate biological network it represents.
+Omics fields like genomics, transcriptomics, proteomics, and metabolomics provide an holistic  view of biomolecules.
+These proteins are comprehensively studied through proteomics to unveil intricate biological networks.
 
@@
 
-As seen in figure \@ref(fig:ui-numbers-default) the aspects shown in the main panel include:
-\spacing{1}
-\begin{scriptsize}
-\begin{itemize}
-\item \textbf{Overview:} Summary showing the individual processing steps of the analysis pipeline
-\item \textbf{Summary Barplot:} Barchart showing the number of peptide spectrum matches, peptides and proteins
-\item \textbf{Normalization validation:} Mutual information intra- and inter- groupwise
-\item \textbf{Reporter ion intensity:} Includes factor selection to create boxplots indicating a factor
-\item \textbf{Median coefficient of variation and correlation:} Variation of peptides for razor proteins and correlation heatmap for protein expression
-\item \textbf{Dimensionality reduction:} Analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP)
-\item \textbf{Feature wise output:} Visualization of an individual protein over the course of the pipeline.
-\item \textbf{Statistics module:} Differential expression analysis
-\item \textbf{Pathway enrichment:} Over-representation analysis for differentially expressed proteins in the context of pathways.
-\item \textbf{Protein set ontology enrichment:}  Over-representation analysis for differentially expressed proteins in a user selected context.
-\end{itemize}
-\end{scriptsize}
-\spacing{1.5}
+<!-- As seen in figure \@ref(fig:ui-numbers-default) the aspects shown in the main panel include: -->
+<!-- \spacing{1} -->
+<!-- \begin{scriptsize} -->
+<!-- \begin{itemize} -->
+<!-- \item \textbf{Overview:} Summary showing the individual processing steps of the analysis pipeline -->
+<!-- \item \textbf{Summary Barplot:} Barchart showing the number of peptide spectrum matches, peptides and proteins -->
+<!-- \item \textbf{Normalization validation:} Mutual information intra- and inter- groupwise -->
+<!-- \item \textbf{Reporter ion intensity:} Includes factor selection to create boxplots indicating a factor -->
+<!-- \item \textbf{Median coefficient of variation and correlation:} Variation of peptides for razor proteins and correlation heatmap for protein expression -->
+<!-- \item \textbf{Dimensionality reduction:} Analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP) -->
+<!-- \item \textbf{Feature wise output:} Visualization of an individual protein over the course of the pipeline. -->
+<!-- \item \textbf{Statistics module:} Differential expression analysis -->
+<!-- \item \textbf{Pathway enrichment:} Over-representation analysis for differentially expressed proteins in the context of pathways. -->
+<!-- \item \textbf{Protein set ontology enrichment:}  Over-representation analysis for differentially expressed proteins in a user selected context. -->
+<!-- \end{itemize} -->
+<!-- \end{scriptsize} -->
+<!-- \spacing{1.5} -->
 
-\newpage
-```{r ui-mutual-info-quant-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with quantile normalization"}
-knitr::include_graphics("screenshots/mutual_info_quant_norm.png")
-```
+<!-- \newpage -->
+<!-- ```{r ui-mutual-info-quant-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with quantile normalization"} -->
+<!-- knitr::include_graphics("screenshots/mutual_info_quant_norm.png") -->
+<!-- ``` -->
 
@@
 
-After performing a logarithmic transformation to the base 2 and quantile normalization, the mutual information (MI) within the healthy control (HC) group's sample type exhibits a minor reduction in the interquartile range (IQR) (as seen in figure \@ref(fig:ui-mutual-info-quant-norm)).
+After performing a logarithmic transformation to the base 2 and quantile normalization, the mutual information (MI) within the healthy control (HC) group's sample type exhibits a minor reduction in the interquartile range (IQR).
 However, it should be noted that there is an increase in the number of outliers compared to the raw data.
-When switching to the quantile normalization the loss of mutual information between sample types decreased (as seen in figure \@ref(fig:ui-mutual-info-quant-norm)).
+When switching to the quantile normalization the loss of mutual information between sample types decreased.
 Therefore the advised normalization technique for this particular data set would be the quantile normalization compared to the default method.
 
-\newpage
-```{r ui-stat-HC-HIV-noMets, echo=F, out.width="70%", fig.align='center', fig.cap="Statistics module in portrait mode"}
+```{r ui-stat-HC-HIV-noMets, echo=F, out.width="60%", fig.align='center', fig.cap="Statistics module in portrait mode"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 
-\newpage
 By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group (see figure \@ref(fig:ui-stat-HC-HIV-noMets)), six differentially expressed proteins were identified.
@@
 The Volcano represents -log10(adjusted p values) against log2 fold changes (logFC).
-These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
+<!-- These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
 The remaining 652 proteins have been identified as not significantly up/down regulated in the comparison analysis.
 These proteins did not exhibit statistically significant differences in expression between the compared groups or conditions.
+-->
 
@@
 
-\newpage
-\listoffigures
+<!-- \newpage -->
+<!-- \listoffigures -->

### 2023-08-15 markdown/article.Rmd
landscape introduced

@@
   - \usepackage[nottoc, numbib]{tocbibind}
+  - \usepackage{pdflscape}
+  - \newcommand{\blandscape}{\begin{landscape}}
+  - \newcommand{\elandscape}{\end{landscape}}
 
@@
 
+\newpage
+\blandscape
 # Materials and methods
@@
 
-\newpage
 The flowchart (figure \@ref(fig:data-processing-pipeline-flowchart-vertical)) describes the processes involved in the ProteoScanR pipeline. 
@@
 In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which is either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.
+
+\elandscape
 

### 2023-08-15 markdown/article.Rmd

--- 2023-08-15 markdown/article.Rmd
@@
   - \usepackage[nottoc, numbib]{tocbibind}
+  - \usepackage{pdflscape}
+  - \newcommand{\blandscape}{\begin{landscape}}
+  - \newcommand{\elandscape}{\end{landscape}}
+  - \usepackage{multicol,blindtext}
 
@@
 \newpage
-# Introduction
+
+\begin{centering}
+
+{\bf Introduction}
+
+\end{centering}
+
 The central dogma underpins genetic information's transformation into functional proteins \citep{Cobb2017}.
@@
 
-# Materials and methods
+\newpage
+\blandscape
+\begin{centering}
+
+{\bf Materials and methods}
+
+\end{centering}
+\begin{multicols}{3}
+
 ```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="75%", fig.align='center', fig.cap= "Flowchart presenting the processing by ProteoScanR.
@@
 
-\newpage
-The flowchart (figure \@ref(fig:data-processing-pipeline-flowchart-vertical)) describes the processes involved in the ProteoScanR pipeline. 
+The flowchart describes the processes involved in the ProteoScanR pipeline. 
 ProteoScanR initially filters the data by false discovery rate (FDR) converted to q-value (provided by MaxQuant \citep{Cox2011}), according to precursor ion fraction (PIF) \citep{Tannous2020, Specht2021} and subsequently aggregates peptide spectrum matches to peptides.
@@
 
+\end{multicols}
+\elandscape
+
 \newpage
-# Results
+\begin{centering}
+
+{\bf Results}
+
+\end{centering}
+
 ```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
@@
 The main panel shows the data in different aspects and helps the user finding a good fit for the methods applied to their individual dataset.
-
-<!-- As seen in figure \@ref(fig:ui-numbers-default) the aspects shown in the main panel include: -->
-<!-- \spacing{1} -->
-<!-- \begin{scriptsize} -->
-<!-- \begin{itemize} -->
-<!-- \item \textbf{Overview:} Summary showing the individual processing steps of the analysis pipeline -->
-<!-- \item \textbf{Summary Barplot:} Barchart showing the number of peptide spectrum matches, peptides and proteins -->
-<!-- \item \textbf{Normalization validation:} Mutual information intra- and inter- groupwise -->
-<!-- \item \textbf{Reporter ion intensity:} Includes factor selection to create boxplots indicating a factor -->
-<!-- \item \textbf{Median coefficient of variation and correlation:} Variation of peptides for razor proteins and correlation heatmap for protein expression -->
-<!-- \item \textbf{Dimensionality reduction:} Analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP) -->
-<!-- \item \textbf{Feature wise output:} Visualization of an individual protein over the course of the pipeline. -->
-<!-- \item \textbf{Statistics module:} Differential expression analysis -->
-<!-- \item \textbf{Pathway enrichment:} Over-representation analysis for differentially expressed proteins in the context of pathways. -->
-<!-- \item \textbf{Protein set ontology enrichment:}  Over-representation analysis for differentially expressed proteins in a user selected context. -->
-<!-- \end{itemize} -->
-<!-- \end{scriptsize} -->
-<!-- \spacing{1.5} -->
-
-<!-- \newpage -->
-<!-- ```{r ui-mutual-info-quant-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Mutual information with quantile normalization"} -->
-<!-- knitr::include_graphics("screenshots/mutual_info_quant_norm.png") -->
-<!-- ``` -->
 
@@
 
-```{r ui-stat-HC-HIV-noMets, echo=F, out.width="60%", fig.align='center', fig.cap="Statistics module in portrait mode"}
+\blandscape
+\begin{multicols}{2}
+
+```{r ui-stat-HC-HIV-noMets, echo=F, out.width="80%", fig.align='center', fig.cap="Statistics module in portrait mode"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 The Volcano represents -log10(adjusted p values) against log2 fold changes (logFC).
-<!-- These proteins include P68133 (Actin, alpha skeletal muscle), P47914 (60S ribosomal protein L29), P25311 (Zinc-alpha-2-glycoprotein), O60506 (Heterogeneous nuclear ribonucleoprotein Q), P31153 (S-adenosylmethionine synthase isoform type-2), and P46777 (60S ribosomal protein L5).
-The remaining 652 proteins have been identified as not significantly up/down regulated in the comparison analysis.
-These proteins did not exhibit statistically significant differences in expression between the compared groups or conditions.
--->
 
-# Discussion and conclusion
+\begin{centering}
+
+{\bf Discussion and conclusion}
+
+\end{centering}
+
 The Proteomics Workbench interface and the ProteoScanR pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
@@
 
-In conclusion, the ProteoScanR pipeline and Proteomics Workbench interface provide a user-friendly and efficient approach for proteomic data analysis, enabling biologists to address their research questions and validate computations.
-Future perspectives include deploying the application on a server for global accessibility, leveraging artificial intelligence techniques, and implementing clustering approaches to further enhance annotation and understanding of cellular heterogeneity and functional characteristics.
+\end{multicols}
+\elandscape
 

### 2023-08-17 markdown/article.Rmd
thesis and article upload ready

@@
 \doublespacing
-{\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
+{\bf Design and implementation of an analysis pipeline for single cell proteomics data} 
 \\ presenting ProteoScanR \& Proteomics Workbench
@@
 
-Main Supervisor: Dr. DI(FH) Gerhard Duernberger \\
-External Supervisor: Assoc. Prof. Ujjwal Neogi, M.Sc. PhD \\
+Main supervisor: Dr. DI(FH) Gerhard Duernberger \\
+External supervisor: Assoc. Prof. Ujjwal Neogi, M.Sc. Ph.D. \\
 Second examiner: FH-Prof. Dr. Alexandra Graf \\
+Second external supervisor: Anoop Ambikan, M.Sc. Ph.D. \\
 
@@
 Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample.
-Mass spectrometers measure the mass to charge ratio (m/z) of a charged particle from a larger molecule.
+Mass spectrometers measure the mass to charge ratio (m/z) of charged particles.
 Given the high resolution of MS data, algorithms are employed to convert the raw signal into an interpretable form.
@@
 
-```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
+```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Screenshot of proteomics workbench.
+Summary barplot showing the number of peptide spectrum matches, peptides and proteins"}
 knitr::include_graphics("screenshots/numbers_default.png")
@@
 When switching to the quantile normalization the loss of mutual information between sample types decreased.
-Therefore the advised normalization technique for this particular data set would be the quantile normalization compared to the default method.
+Therefore the advised normalization technique for this particular data set would be the quantile normalization compared to the default method \citep{Specht2021}.
 
@@
 
-```{r ui-stat-HC-HIV-noMets, echo=F, out.width="80%", fig.align='center', fig.cap="Statistics module in portrait mode"}
+```{r ui-stat-HC-HIV-noMets, echo=F, out.width="80%", fig.align='center', fig.cap="Screenshot of statistics module in portrait mode"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 
-By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group (see figure \@ref(fig:ui-stat-HC-HIV-noMets)), six differentially expressed proteins were identified.
+By comparing the log10-transformed and quantile-normalized expression values between the HIV group without metabolic syndrome and the healthy control (HC) group, six differentially expressed proteins were identified.
 The Venn diagram indicates significantly up/down regulated proteins.
@@
 
-The Proteomics Workbench interface and the ProteoScanR pipeline demonstrated how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
+The Proteomics Workbench interface and the ProteoScanR pipeline demonstrates how interactive engagement with the data not only enhances the experience of biologists but also improves the comprehension of the underlying significance of a biological dataset.
 By utilizing an entropy-based visualization approach, the conservation of information can be validated, allowing users to select appropriate methods and adjust thresholds, cutoffs, and techniques accordingly.

### 2023-09-11 markdown/methods.Rmd
update

--- 2023-07-18 markdown/methods.Rmd
@@
 ## Wetlab - Materials and Methods
-For testing and development of the ProteoScanR pipeline the following materials and methods were applied.
+For testing and development of the ProteoScanR pipeline the following materials and methods were performed. For the sample preparation protocol see appendix.
 
 ### Celltypes
-Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome were isolated.
-Cell sorting was performed with FACSAria Fusion (BD Biosciences) by auto fluorescence.
-
-#### Cells sorting protocol
-\begin{enumerate}
-\item 96 single cells from each donor were sorted into a single plate. Each well contained a single cell. As such, 12 different plates containing single cells from 12 different donors were prepared. 
-\item Carrier plates were also prepared. In the carrier plates, 100 cells from one donor were sorted into each well. 24 such wells (3 columns) were prepared for each donor.
-\end{enumerate}
+Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome. Cell sorting was performed with FACSAria Fusion (BD Biosciences).
 
@@
 
-#### Digestion protocol
-\begin{enumerate}
-\item Prepare 25 ng/ $\mu$L trypsin in 100 mmol/L Triethylammonium bicarbonate buffer (TEAB).
-\item Add 25 ng/$\mu$L trypsin: 2 $\mu$L to the carrier samples, 1 $\mu$L to the single cell samples.
-\item Seal the plates.
-\item Centrifuge (2 min at 120 G).
-\item Incubate overnight at 37 °C.
-\end{enumerate}
-
 ### Labeling
-Samples of the test data set were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.
-
-#### Labeling protocol
-\begin{enumerate}
-\item Let the TMTpro labels come to room temperature so that all the condensation on the tube is gone.
-Dissolve the TMTpro reagents in anhydrous acetonitril (ACN) to obtain a concentration of 10 $\mu$g/$\mu$L.
-\item Add 1 $\mu$L of TMTpro reagents according to the tables. For one set of plates, you will need 16 $\mu$L of each label. It is recommended to aliquot a slightly larger volume, of around 20 $\mu$L.
-\item Seal the plates.
-\item Centrifuge for 2 minutes at 120 G.
-\item Incubate for 1 hour at room temperature.
-\item Prepare a 5% hydroxylamine solution in water.
-\item Add 1 $\mu$L of the 5% hydroxylamine solution to each well.
-\item Seal the plates.
-\item Centrifuge for 2 minutes at 120 G.
-\item Incubate for 1 hour at room temperature.
-\item Cover the plates with cap strips.
-\end{enumerate}
+Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.
 
@@
 Carrier proteome was added to the 36 TMTpro sets before pooling sample types for each of the 36 MS-runs.
-
-#### Pooling protocol
-\begin{enumerate}
-\item Wash a 10 $\mu$L glass microsyringe five times with ACN.
-\item Use the microsyringe to combine TMTpro-labeled samples into 36 LC autosampler vials with built-in glass inserts.
-\item Combine six carrier wells from each sample and add the carrier proteome to the 36 TMTpro sets.
-\item Combine the first set by taking well A1 from the first single-cell plate, well A2 from the second plate, well A3 from the third plate, and so on.
-For the next set, take well A1 from the second plate, well A2 from the third plate, and so on.
-Combine 12 sets following this scheme.
-Combine the next 24 sets by following the same scheme for rows B and C.
-\item Vacuum centrifuge the combined sets to dryness.
-\item Dissolve the dried samples in 7 $\mu$L of buffer A and transfer them to LC-MS/MS for further analysis.
-\end{enumerate}
 
@@
 ### Hardware
-Data was analysed using commercially available hardware, find details below:
 Lenovo Laptop X360 with:
@@
 
-```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="80%", fig.align='center', fig.cap= "Flowchart presenting the processing by ProteoScanR"}
+```{r data_processing_pipeline_flowchart_vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="80%", fig.align='center', fig.cap="Flowchart presenting the processing by ProteoScanR"}
 library(Gmisc, quietly = TRUE)
@@
 
-```{r data_processing_pipeline_flowchart_horizontal, eval=FALSE, fig.height=14, fig.width=45, include=FALSE}
+```{r data_processing_pipeline_flowchart_horizontal, eval=FALSE, fig.height=14, fig.width=45, include=F}
 library(Gmisc, quietly = TRUE)
@@
 
-missing_value_opts <- boxGrob("options:\nKNN\ndrop rows\nreplace with mean\nreplace with median\nreplace with 0",
+missing_value_opts <- boxGrob("options:\nKNN\ndrop rows\nreplace with mean\nreplace with median",
                               box_gp = opts_box_setting,
@@
 
-In order to streamline the analysis of multiple experiments and software development, object oriented programming was applied.
+In order to streamline the analysis of multiple experiments, object oriented programming was applied.
 The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure.
@@
 ### Filter by q-value
-The next step for quality control is the exclusion of spectra with a high false discovery rate (FDR).
+The next step for quality control is the exclusion of samples with a high false discovery rate (FDR).
 When applying multiple statistical testing (e.g.
-t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant result will iteratively increase with each test performed.
+t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed.
 Corrections in statistics are an approach to compensate for the multiplicity of testing.
 There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR.
-Many search engines face that problem by calculating the false discovery rate, by dividing false PSMs (=hit of the decoy database) with the total number of PSMs above the peptide-spectrum matching score.
+In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score.
 The peptides spectrum matching score is defined as $-10 * log(p)$.
@@
 The calculation of the score is highly dependent on the data acquisition method used.
-MaxQuant uses Andromeda \citep{Cox2011}, an integrated search engine.
-Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot \citep{Brosch2009}.
+MaxQuant uses Andromeda, an integrated search engine.
+Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora.
 As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage.
@@
 In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function.
-In the context of processing of PSMs to unique peptides, the desired column is the peptide sequence.
+In the context of processing from PSMs to unique peptides, the desired column is the PSM sequence.
 To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.
@@
 Sample size is often a limiting factor in hypothesis testing.
-A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to diminished statistical power.
+A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to an early end of analyses.
 To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis.
@@
 This can be due two reasons.
-One is the expression level of the given proteins in a sample.
-Meaning, that the expression of the observed sample type is simply lower than in the other type.
+One is the expression level of the given proteins in a cell.
+Meaning, that the expression of the observed cell type is simply lower than the other type.
 Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes.
@@
 Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data.
-The coefficient of variation of a peptide is defined as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins).
+The coefficient of variation of a peptide is considered as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins).
 Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.
@@
 This transformation was used as a reference when comparing methods.
-However, using the logarithm to the base 10 has the same effect and improves the interpretability of the values.
-Furthermore the log10 transformation will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
+However, using the logarithm to the base 10 may be an easy way to recalculate expression data in the mind and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
 When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
@@
 The second method for normalization included is the CONSTANd method, which was proven suitable for relative quantification in the field of proteomics \citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq data.
-CONSTANd uses a technique called matrix raking, which employs the RAS ($K=RAS$) algorithm.
-The expression set reflects the non-negative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $K$ which equals to $diag(x) * A *diag(y)$, so that $K=diag(x) * A *diag(y)= RAS$.
-Whereas $x \in \mathbb{R}^{m}$, $y \in \mathbb{R}^{n}$ and $R$ denotes for the row multiplier and $S$ for the column multiplier.
-The solution to this problem involves finding the row sum of $K=u_i$ and the column sum of $K=v_j$, where $i$ and $j$ denote the row and column indices of the matrix, respectively \citep{Bacharach1965}.
-In other words, the matrix will be alternatingly manipulated on rows and columns until the mean of both equals 1.
+CONSTANd uses a technique called matrix raking, which employs the RAS algorithm.
+The expression set reflects the non-negative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $B$ which equals to $diag(x) * A *diag(y)$ .
+Whereas $x \in \mathbb{R}^{m}$ and $y \in \mathbb{R}^{n}$.
+The solution to this problem involves finding the row sum of $B=u_i$ and the column sum of $B=v_j$, where $i$ and $j$ denote the row and column indices of the matrix, respectively \citep{Bacharach1965}.
+In other words, the matrix will be alternatively manipulated on rows and columns until the mean of both equals 1.
 In order to apply the method and yield true results, one has to consider 3 major assumptions between sample types.
-These assumption can be observed in the MA-plot which indicates the differences of two samples.
+This assumption can be observed in the MA-plot which indicates the differences of two samples.
 Given the intensities of two samples, $R$ and $G$, visualized on a graph with the x-axis as $M=\log_2(\frac{R}{G})$ and the y-axis as $A=\frac{1}{2}\log_2(RG)$.
@@
 When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches.
+From a biological perspective, it is highly unlikely that a cell completely lacks a single protein, as it would imply a complete loss of function for that protein.
+Therefore, an intensity value of 0 is close to impossible for most proteins.
 In order to handle this dark space in the expression set, several methods can be chosen in the application.
-A variety of function employs replacing the missing values with the mean or median of the rest of the matrix (total mean/median).
+A variety of function employs replacing the missing values with the mean or median of the rest of the matrix.
 This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well.
-An even more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
-A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values according to an euclidean metric of the neighboring values in the columns where the feature is not missing.
-$K$ denotes for the count of neighboring values of a protein which are considered and can be selected upon preference.
+Even a more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
+A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing.
+$K$ denotes for the count of neighboring values of a gene and can be selected upon preference.
 The euclidean distance between two points is defined as $d = \sqrt{a^2 * b^2}$, the $K$ nearest neighbors average value will be assigned to the intensity of the missing feature.
@@
 Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data.
-This variance can overshadow the biological differences across sample types when performing dimensionality reduction.
+This variance would overshadow the biological differences across sample types when performing dimensionality reduction.
 A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of ProteoScanR.

### 2023-09-11 markdown/acknowledgments.Rmd

--- 2023-02-02 markdown/acknowledgments.Rmd
@@
 
-Thank you for following this tutorial!
+Beside my faithful feline companion, Apollo, who took care of my place in Austria while I worked on this project in Stockholm, I extend my deep gratitude to all the members of Ujjwal Neogi's research group.
+Their consistent guidance and support throughout this project have been truly invaluable, and I am profoundly appreciative of their expertise and assistance.
 
-I hope you'll find it useful to write a very professional dissertation.
+Furthermore, I wish to express my gratitude to Gerhard Dürnberger, my internal supervisor, for his unwavering support and feedback on my master thesis, along with his precious time devoted to our meetings.

### 2023-09-11 markdown/article_doc.Rmd

--- 2023-08-17 markdown/article.Rmd
@@
 output:
-  bookdown::pdf_document2:
+  bookdown::word_document2:
     toc: false
@@
 \begin{centering}
-```{r uni_logo, echo=F, out.width="20%", fig.align='center'}
+```{r uni_logo, echo=F, out.width="20%"}
 knitr::include_graphics("KI_logo.png")
@@
 
-```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="75%", fig.align='center', fig.cap= "Flowchart presenting the processing by ProteoScanR.
+```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="75%", fig.cap= "Flowchart presenting the processing by ProteoScanR.
 Gold: techniques which involve explanatory data analysis.
@@
 
-```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Screenshot of proteomics workbench.
-Summary barplot showing the number of peptide spectrum matches, peptides and proteins"}
+```{r ui-numbers-default, echo=F, out.width="100%", fig.cap="Screenshot of proteomics workbench. Summary barplot showing the number of peptide spectrum matches, peptides and proteins"}
 knitr::include_graphics("screenshots/numbers_default.png")
@@
 
-```{r ui-stat-HC-HIV-noMets, echo=F, out.width="80%", fig.align='center', fig.cap="Screenshot of statistics module in portrait mode"}
+```{r ui-stat-HC-HIV-noMets, echo=F, out.width="80%", fig.cap="Screenshot of statistics module in portrait mode"}
 knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
@@
 ```
-
-<!-- \newpage -->
-<!-- \listoffigures -->

### 2023-09-11 markdown/discussion.Rmd

@@
 Unfortunately, a substantial number of individual proteins in the data set (totaling `r n_missingness`) have missing values.
-On average, this amounts to approximately `r round(mean_missingness, 2)*100`% of the proteins across all runs and channels (see figure \ref@(fig:missingness)).
+On average, this amounts to approximately `r round(mean_missingness, 2)*100`% of the proteins across all runs and channels (see figure \@ref(fig:missingness)).
 This situation leads to the need for missing value imputation using techniques such as k-nearest neighbor (K-NN), which can sometimes lead to artificially evenly distributed values.
@@
 
-As observed in the dendrogram (see figure \ref@(fig:cluster-analysis-dendrogram)), it is evident that sample type alone does not account for the variation observed in the data.
+As observed in the dendrogram (see figure \@ref(fig:cluster-analysis-dendrogram)), it is evident that sample type alone does not account for the variation observed in the data.
 The branching patterns in the dendrogram suggest that other factors might be significant for contributing to the averaging of protein expression levels.
@@
 
-The gap statistic (see figure \ref@(fig:cluster-analysis-gap-stat)) compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution.
+The gap statistic (see figure \@ref(fig:cluster-analysis-gap-stat)) compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution.
 The null reference distribution represents a scenario where no clear clustering patterns exist in the data.
@@
 
-The silhouette method (see figure \ref@(fig:cluster-analysis-silh)) calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters.
+The silhouette method (see figure \@ref(fig:cluster-analysis-silh)) calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters.
 If the overall distances between clusters are small, indicating overlapping or unclear separation, it is suggested to take the minimum k.
@@
 
-As seen in the principal components analysis (see figure \ref@(fig:cluster-analysis-dim-red)), clusters are densely packed, and the plot shows no clear differentiation between the `r optimal_clusters` suggested clusters.
+As seen in the principal components analysis (see figure \@ref(fig:cluster-analysis-dim-red)), clusters are densely packed, and the plot shows no clear differentiation between the `r optimal_clusters` suggested clusters.
 The dense packing and poor resolution indicates that other factors may be influencing the lack of distinct separation.

### 2023-09-11 markdown/introduction.Rmd

@@
 Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role within this interactive structure.
-Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}
+Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}.
 
@@
 
-Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their (m/z) consecutively.
+Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their m/z consecutively.
 Ion trapping enables the technique of coupled mass spectrometers, where ions were further fragmented by collision induced dissociation and analyzed subsequently.
